### 2021-01-13 effects_writeup.Rnw
Might need to rewrite varpred to properly condition

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{natbib}
\usepackage[dvipsnames]{xcolor}
\usepackage{graphicx, subfig}
\usepackage{float}
\usepackage{hyperref}

\newcommand{\JD}[1]{{\color{blue} \emph{#1}}}
\newcommand{\bmb}[1]{{\color{RubineRed!70!} \emph{#1}}}
\newcommand{\pkg}[1]{\textbf{#1}}
\def\code#1{\texttt{#1}}

\bibliographystyle{chicago}
\setlength{\parindent}{0pt}

\title{Marginal predictions}

\begin{document}

\maketitle

\section{Introduction}

In this task, we describe various R machineries for displaying predictions in both simple and complex (involving interaction terms) generalized linear models. We first consider existing approaches for \emph{conditioning} predictions and then describe our proposed approach for \emph{marginalized} predictions. These provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind transformations on the dependent variables whose estimates are usually, but not always, a subject to less clarity of interpretation.

We can derive various quantities from a fitted regression model. The first and most obvious, is the model coefficient estimates. Others include predicted values of the outcome variable---1) predictions at a particular; and 2) mean or median value of the predictors. The first case simply involves evaluation of the fitted model function, say $\hat{f}(X = x)$, at some particular value of $x$. The second case, chooses the values of the predictor based on its distributional properties.

Most importantly, from these predicted values, we can also generate second class quantities of interest -- \emph{conditional} or \emph{marginal} predictions. These quantities describes the change in the predicted value of the dependent variable after changing one independent variable -- either a discrete change in the categorical variable(s) or an instantaneous change in continuous variables, while all other variable are held at specified values.

Suppose we are interested in the \emph{conditional} predictions of a particular predictor (hence forth referred as \emph{focal} predictor otherwise \emph{non-focal}), $x_f$, in the set of predictors. To keep it simple, assume that the model has no interaction terms. Then the idea is to fix the values of \emph{non-focal} predictor(s) at some typical values -- typically determined by averaging in some meaningful way, for example, arithmetic mean and average over the levels of the factors of \emph{non-focal} continuous and categorical predictors, respectively. This is achieved by averaging the columns of \emph{model matrix}, $\mathbf{X}$, except for the column of focal predictor.

Consider a simple linear model with linear predictor $\eta = \mathbf{X\beta}$ and let $g(\mu) = \eta$ be an identity link function (in the case of simple linear model), where $\mu$ is the predicted (expected) value of outcome variable  $y$. Let $\hat{\beta}$ be the estimate of $\beta$, together with the estimated covariance matrix $V(\hat{\beta})$ of $\hat{\beta}$. Let the entries of $\mathbf{X^*}$ include all conditioned (\emph{non-focal}) and \emph{focal} predictors. The model matrix, $\mathbf{X^*}$, inherits most of its key properties, for example transformation (e.g., scaling) on the predictors and interactions from the model matrix, $\mathbf{X}$. Then the predicted values $\hat{\eta}^* = \mathbf{X^*}\hat{\beta}$ represents the \emph{conditional} predictions of the focal predictor. Alternatively, we can transform these predictions to response scale using $g^{-1}(\hat{\eta}^*)$.

Further, we can compute the standard errors (SEs) of the \emph{conditional} predictions, $\hat{\eta}^*$, for constructing confidence intervals, as the $\operatorname{sqrt}(\operatorname{diag}(\mathbf{X}^*V(\hat{\beta})\mathbf{X}^{*T}))$. When computing \emph{marginal} predictions, the \emph{uncertainty} as a result of \emph{non-focal} predictors are removed. This can be achieved in two ways:
\begin{itemize}
\item using variance-covariance matrix, $V(\hat{\beta})$
\item using centered model matrix, $\mathbf{X}^*$
\end{itemize}

\subsection{Variance-covariance}

The computation of $\hat{\eta}^*$ remains the same as described above. However, to compute SEs $V(\hat{\beta})$ is modified by \emph{zeroing-out} (the variance-covariance of all non-focal predictors are assigned zero) entries of \emph{non-focal} terms in $V(\hat{\beta})$. This approach requires \emph{centering} of the predictors in the model matrix, $X$. In other words, the fitted model should have \emph{centered} predictors.

\subsection{Centered model matrix}

Suppose the \emph{non-focal} entries in $X^*$ are computed by some kind of averaging. Consider centered $X^*$, $X^*_{c} = (X^* - \bar{X^*})$. It follows that the \emph{non-focal} entries in $X^*_{c}$ are all zero. As a result, the SEs are computed as  $\operatorname{sqrt}(\operatorname{diag}(\mathbf{X}^*_{c}V(\hat{\beta})\mathbf{X}^{*T}_{c}))$, which actually \emph{zeros-out}. More generally, \emph{centering}, $\mathbf{X}^*_{c} = \mathbf{X}^* - k$ (for example $k = E(\mathbf{X}^*)$) impacts on the estimated value of the intercept and its associated variance. However, the slopes are not affect by this. The implication of this that since \emph{non-focal} terms in $\mathbf{X}^*_c$ are zero, it doesn't matter what their corresponding values are in the variance-covariance matrix. Hence, we can compute \emph{marginal} predictions from non-centered predictors (in other words, fitted models with predictors in their natural scales).

\section{Available packages}
The following R packages \pkg{effects}, \pkg{emmeans} and \pkg{margins} implement various schemes for constructing \emph{conditional} predictions. However, currently, their ability to compute \emph{marginal} prediction is limited to the use of variance-covariance approach which requires the fitted model to be centered. We propose \pkg{varpred} to overcome this limitation.

\section{Yet to figure out}
\begin{enumerate}
\item Handling complex interaction
\item Zero SEs in the case of variance-covariance approach
\end{enumerate}

\section{Illustrations}

\subsection{Simulation}

Consider a simple no-interaction terms simulation:
\begin{itemize}
\item $y = \beta_0 + \beta_1x_1 + \beta_2x_2 + \epsilon$ s.t $\epsilon \sim N(0, 1)$ and $\{\beta_0 = -5, \beta_1 = 0.1, \beta_2 = -0.05\}$
\begin{itemize}
\item $x_1 \sim unif(1, 9)$
\item $x_2$ -- two level categorical variable
\end{itemize}
\end{itemize}

<<set_up, echo=FALSE>>=
library(ggplot2)
library(margins)
library(effects)
library(emmeans)
source("effectsfuns.R")
varpredtheme()
@

<<simple_sim>>=
set.seed(101)
N <- 100
x1_min <- 1
x1_max <- 9
b0 <- 0.3
b1 <- 0.1
b2 <- -0.6
x2_levels <- factor(c("A", "B"))
df <- expand.grid(x1u = runif(n=N, min=x1_min, max=x1_max)
	, x2u = x2_levels
)
X <- model.matrix(~x1u + x2u, df)
betas <- c(b0, b1, b2)
df$y <- rnorm(nrow(df), mean= X %*% betas, sd=1)
df2 <- df
df <- transform(df
	, x1c = drop(scale(x1u, scale=FALSE)) 
	, x1s = drop(scale(x1u, center=FALSE))
	, x1sd = drop(scale(x1u))
)
head(df)
@

\subsection{Model fitting}

No scaling or centering covariates

<<model_lm_u>>=
df_temp <- drop(df) 	# margins doesn't work with transform 
                   	# but we need df to extract attributes :(
lm_u <- lm(y ~ x1u + x2u, data = df_temp)
@

Centered covariates ($x - \bar{x}$) model

<<model_lm_c>>=
lm_c <- lm(y ~ x1c + x2u, data = df_temp)
@

Scaled covariates ($x/sd{x}$) model

<<model_lm_s>>=
lm_s <- lm(y ~ x1s + x2u, data = df_temp)
@

Both scaled and centered covariates

<<model_lm_sd>>=
lm_sd <- lm(y ~ x1sd + x2u, data = df_temp)
@
Coefficients estimates

<<model_summary, echo=FALSE>>=
modsummary <- function(patterns="^lm_", fun, simplify = TRUE
	, combine = c("cbind", "rbind"), match_colnames = NULL){
	mod <- objects(name=1L, pattern=patterns, sorted=TRUE)
	combine <- match.arg(combine)
	out <- sapply(mod, function(m){
		out <- fun(get(m))
		out <- round(out, 3)
		if (combine=="rbind"){
			if (!is.null(match_colnames)){
				colnames(out) <- match_colnames
			}
			out <- cbind.data.frame(out, model = m)
		}
		return(out)
	}, simplify = FALSE)
	if (simplify){
		out <- do.call(combine, out)
		if (combine=="cbind") {
			colnames(out) <- mod
		}
	}
	return(out)
}
@

<<simple_coefs>>=
coef_est <- modsummary(fun=coef)
print(coef_est)
@

Log likelihoods

<<simple_loglik>>=
ll_est <- modsummary(fun=logLik)
print(ll_est)
@

Variance covariance matrix
<<simple_vcov>>=
vcov_est <- modsummary(fun=function(x)vcov(x), simplify=TRUE
	, combine="rbind", match_colnames = names(coef(lm_u)))
print(vcov_est)
@

\subsection{Conditional predictions}

The function will be used downstream to \emph{tidy} conditional predictions from each of the methods highlighted above.

<<condional_summary, cho=FALSE>>=
condsummary <- function(mod_list, fun, simplify = TRUE
	, combine = c("cbind", "rbind"), scale_param = list(), fmethod = NULL){
	focal <- names(mod_list)
	combine <- match.arg(combine)
	out <- sapply(focal, function(f){
		mod <- mod_list[[f]]
		out <- fun(mod, f)
		if (inherits(out, c("emmeans", "emmGrid"))) {
			out <- as.data.frame(out)
			oldn <- c(f, "emmean"
				, grep("\\.CL", colnames(out), value=TRUE)
			)
			newn <- c("xvar", "fit", "lwr", "upr")
			colnames(out)[colnames(out) %in% oldn] <-  newn
		} else if (inherits(out, "eff")) {
			out <- as.data.frame(out)
			oldn <- c(f, "lower", "upper")
			newn <- c("xvar", "lwr", "upr")
			colnames(out)[colnames(out) %in% oldn] <-  newn
		} else if (any(colnames(out) %in% c("xvals", "yvals"))) {
			## Not proper way to handle margins object.
			## Maybe extend cplot to return inheritable object
			oldn <- c("xvals", "yvals", "lower", "upper")
			newn <- c("xvar", "fit", "lwr", "upr")
			colnames(out)[colnames(out) %in% oldn] <-  newn
		} else {
			colnames(out)[colnames(out)%in%f] <- "xvar"
		}
		if (combine=="rbind"){
			out <- cbind.data.frame(out, model = f)
			out <- out[, c("xvar", "fit", "lwr", "upr", "model")]
		}
		pp <- names(scale_param)
		if (!is.null(pp)){
			if(grepl("s$",f)){
				vv <- grep("s$",f, value=TRUE)
				xsd <- scale_param[[vv]]
				out[,"xvar"] <- out[,"xvar"]*xsd
			}
			if(grepl("c$",f)){
				vv <- grep("c$",f, value=TRUE)
				xmu <- scale_param[[vv]]
				out[,"xvar"] <- out[,"xvar"] + xmu
			}
			if(grepl("sd$",f)){
				vv <- grep("sd$",pp,value=TRUE)
				xmu <- scale_param[[vv]][1]
				xsd <- scale_param[[vv]][2] 
				out[,"xvar"] <- out[,"xvar"]*xsd + xmu

			}
		}
		if (!is.null(fmethod)){
			out[,"method"] <- fmethod
		}
		return(out)
	}, simplify = FALSE)
	if (simplify){
		out <- do.call(combine, out)
		if (combine=="cbind") {
			colnames(out) <- mod
		} else {
			rownames(out) <- NULL
		}
	}
	return(out)
}
@

We start by computing the conditional predictions using each of the methods (\pkg{varpred, emmeans, effects, margins}). In the first case, the predictions are displayed on the variable level scale-specific values (original (u), mean centered (c), divided by standard deviation (s), and both mean centered and scaled (sd)). In the second case, the predictions are transformed back to the original scale.

<<simple_model>>=
simple_models <- list(x1u = lm_u, x1c = lm_c, x1s = lm_s, x1sd = lm_sd)
@

\subsubsection{Continuous predictor}

<<simple_conditional_specific, results=hide>>=
## varpred
simple_vpred_all <- condsummary(mod_list=simple_models, fun=function(x, f){
	dd <- varpred(x, f)
}, simplify = TRUE, combine = "rbind", fmethod = "varpred")

## emmeans
simple_empred_all <- condsummary(mod_list=simple_models, fun=function(x, f){
	spec <- as.formula(paste0("~", f))
	dd <- emmeans(x, spec=spec, cov.keep=f)
}, simplify = TRUE, combine = "rbind", fmethod = "emmeans")

## effects
simple_efpred_all <- condsummary(mod_list=simple_models, fun=function(x, f){
	dd <- Effect(f, x, xlevels=100)
}, simplify = TRUE, combine = "rbind", fmethod = "effects")

## margins
simple_margpred_all <- condsummary(mod_list=simple_models, fun=function(x, f){
	dd <- cplot(x, f, what="prediction", n=100, draw=FALSE)
}, simplify = TRUE, combine = "rbind", fmethod = "margins")

## Combine all the estimates
simple_pred_all <- do.call("rbind"
	, list(simple_vpred_all, simple_empred_all, simple_efpred_all, simple_margpred_all)
)
@

<<>>=
head(simple_pred_all)
simple_pred_all_plot <- (plotpreds(simple_pred_all) 
	+ facet_wrap(~method) 
	+ theme(legend.position="bottom")
)
@

\begin{figure}[ht]
\begin{center}
<<simple_pred_all_plot, fig=TRUE, echo=FALSE>>=
simple_pred_all_plot
@
\end{center}
\caption{Conditional predictions on the variable-specific values.}
\end{figure}

Compute predictions and then back transform each predictor to the original scale.

<<simple_conditional_backtrans, results=hide>>=
## Extract scaling parameters
scale_param <- list(x1s = unlist(attributes(df$x1s))
	, x1c = unlist(attributes(df$x1c))
	, x1sd = unlist(attributes(df$x1sd))
)

## varpred
simple_vpred_spec <- condsummary(mod_list=simple_models, fun=function(x, f){
		dd <- varpred(x, f)
	}, simplify = TRUE, combine = "rbind"
	, scale_param = scale_param, fmethod = "varpred"
)

## emmeans
simple_empred_spec <- condsummary(mod_list=simple_models, fun=function(x, f){
		spec <- as.formula(paste0("~", f))
		dd <- emmeans(x, spec=spec, cov.keep=f)
	}, simplify = TRUE, combine = "rbind"
	, scale_param = scale_param, fmethod = "emmeans"
)

## effects
simple_efpred_spec <- condsummary(mod_list=simple_models, fun=function(x, f){
		dd <- Effect(f, x, xlevels=100)
	}, simplify = TRUE, combine = "rbind"
	, scale_param = scale_param, fmethod = "effects"
)

## margins
simple_margpred_spec <- condsummary(mod_list=simple_models, fun=function(x, f){
		dd <- cplot(x, f, what="prediction", n=100, draw=FALSE)
	}, simplify = TRUE, combine = "rbind"
	, scale_param = scale_param, fmethod = "margins"
)

## Combine all the estimates
simple_pred_spec <- do.call("rbind"
	, list(simple_vpred_spec, simple_empred_spec, simple_efpred_spec, simple_margpred_spec)
)
@

<<>>=
head(simple_pred_spec)
simple_pred_spec_plot <- (plotpreds(simple_pred_spec) 
	+ facet_wrap(~method) 
	+ theme(legend.position="bottom")
)
@

\begin{figure}[ht]
\begin{center}
<<simple_pred_spec_plot, fig=TRUE, echo=FALSE>>=
simple_pred_spec_plot
@
\end{center}
\caption{Back-transformed conditional predictions.}
\end{figure}

\subsubsection{Categorical predictors}

Here, we are considering case where we are only interested estimating the conditional predictions. In this case, it doesn't matter which of the four models we choose since they will give the same predictions. Later, we'll show how we can estimate marginalized and/or centered (CI) for categorical predictors.

<<simple_conditional_specific_cat, results=hide>>=
## varpred
simple_models_cat <- list(x2u = lm_u)
simple_vpred_cat <- condsummary(mod_list=simple_models_cat, fun=function(x, f){
	dd <- varpred(x, f)
}, simplify = TRUE, combine = "rbind", fmethod = "varpred")

## emmeans
simple_empred_cat <- condsummary(mod_list=simple_models_cat, fun=function(x, f){
	spec <- as.formula(paste0("~", f))
	dd <- emmeans(x, spec=spec, cov.keep=f)
}, simplify = TRUE, combine = "rbind", fmethod = "emmeans")

## effects
simple_efpred_cat <- condsummary(mod_list=simple_models_cat, fun=function(x, f){
	dd <- Effect(f, x, xlevels=100)
}, simplify = TRUE, combine = "rbind", fmethod = "effects")

## margins
simple_margpred_cat <- condsummary(mod_list=simple_models_cat, fun=function(x, f){
	dd <- cplot(x, f, what="prediction", n=100, draw=FALSE)
}, simplify = TRUE, combine = "rbind", fmethod = "margins")

## Combine all the estimates
simple_pred_cat <- do.call("rbind"
	, list(simple_vpred_cat, simple_empred_cat, simple_efpred_cat, simple_margpred_cat)
)
@

<<>>=
head(simple_pred_cat)
simple_pred_cat_plot <- (plotpreds(simple_pred_cat) 
	+ facet_wrap(~method) 
	+ theme(legend.position="bottom")
)
@

\begin{figure}[ht]
\begin{center}
<<simple_pred_cat_plot, fig=TRUE, echo=FALSE>>=
simple_pred_cat_plot
@
\end{center}
\caption{Conditional predictions.}
\end{figure}

\subsection{Marginal predictions}

We can implement marginalization by \emph{zeroing-out} the variance of non-focal predictors (possible in all four methods) but this requires that the predictors are centered beforehand. On the other hand, in \pkg{varpred} marginal predictions can be obtained by centering the SE estimates.

\subsection{zeroed-out covariance matrix}

We \pkg{zero\_vcov} function in \pkg{varpred} to transform the variances of non-focal predictors to zero. For example:

<<zero_out_ex>>=
lm_u_vcov <- vcov(lm_u)
print(lm_u_vcov)

## x1u is the focal variable
lm_u_vcov_zero <- zero_vcov(lm_u, focal_vars = "x1u")
print(lm_u_vcov_zero)
@

We'll repeat the same procedure in conditional predictions section but modify the functions to incorporate \emph{zeroed-out} covariance matrix.

\subsubsection{Continuous predictor}

Notice that the modification in each of the method function call to pass either covariance matrix or function. In particular, in \pkg{varpred}, it is passed as \pkg{vcmat}, \pkg{emmeans} and \pkg{effects} as \pkg{vcov.} and \pkg{margins} as \pkg{vcov}. However, this seems not to work currently in \pkg{cplot} which computes the predictions in \pkg{margins}.

<<simple_marginal_specific, results=hide>>=
## varpred
simple_vpred_all <- condsummary(mod_list=simple_models, fun=function(x, f){
	dd <- varpred(x, f, vcmat = zero_vcov(x, f))
}, simplify = TRUE, combine = "rbind", fmethod = "varpred")

## emmeans
simple_empred_all <- condsummary(mod_list=simple_models, fun=function(x, f){
	spec <- as.formula(paste0("~", f))
	dd <- emmeans(x, spec=spec, cov.keep=f, vcov. = zero_vcov(x, f))
}, simplify = TRUE, combine = "rbind", fmethod = "emmeans")

## effects
simple_efpred_all <- condsummary(mod_list=simple_models, fun=function(x, f){
	dd <- Effect(f, x, xlevels=100, vcov. = function(x, complete=FALSE)zero_vcov(x, f, complete))
}, simplify = TRUE, combine = "rbind", fmethod = "effects")

## margins
simple_margpred_all <- condsummary(mod_list=simple_models, fun=function(x, f){
	dd <- cplot(x, f, what="prediction", n=100, draw=FALSE, vcov=function(x)zero_vcov(x, f))
}, simplify = TRUE, combine = "rbind", fmethod = "margins")

## Combine all the estimates
simple_pred_all <- do.call("rbind"
	, list(simple_vpred_all, simple_empred_all, simple_efpred_all, simple_margpred_all)
)
@

<<echo=FALSE>>=
simple_pred_all_plot <- (plotpreds(simple_pred_all) 
	+ facet_wrap(~method) 
	+ theme(legend.position="bottom")
)
@

\begin{figure}[H]
\begin{center}
<<simple_pred_all_plot, fig=TRUE, echo=FALSE>>=
simple_pred_all_plot
@
\end{center}
\caption{Marginal predictions with the zeroed-out covariances on the variable-specific scale.}
\end{figure}

Also, we repeat above implementation but back-transform the predictions to original scale (unscale -- uncenter and/or unscale) by modifying the previous code to include scaling parameters. To clearly distinguish between what happens when apply various scaling schemes, we separately do the plots (see \autoref{fig:marginal_transformed}).

\subsection{Centered model matrix}

Marginal predictions in \pkg{effects} and \pkg{emmeans} require models fitted with transformed predictors. However, in \pkg{varpred} we can use model fitted using predictors on their original scale and estimate marginal predictions. In our example, we use the unscaled model \code{lm\_u} and simply set \code{isolate=TRUE}. The predictions (see \autoref{fig:varpred_isolate_num}) are similar to \autoref{fig:marginal_transformed}~b (except for the \pkg{margins}).

<<simple_marginal_backtrans, echo=FALSE, results=hide>>=
## varpred
simple_vpred_spec <- condsummary(mod_list=simple_models, fun=function(x, f){
		dd <- varpred(x, f, vcmat = zero_vcov(x, f))
	}, simplify = TRUE, combine = "rbind"
	, scale_param = scale_param, fmethod = "varpred"
)

## emmeans
simple_empred_spec <- condsummary(mod_list=simple_models, fun=function(x, f){
		spec <- as.formula(paste0("~", f))
		dd <- emmeans(x, spec=spec, cov.keep=f, vcov. = zero_vcov(x, f))
	}, simplify = TRUE, combine = "rbind"
	, scale_param = scale_param, fmethod = "emmeans"
)

## effects
simple_efpred_spec <- condsummary(mod_list=simple_models, fun=function(x, f){
		dd <- Effect(f, x, xlevels=100, vcov. = function(x, complete=FALSE)zero_vcov(x, f, complete))
	}, simplify = TRUE, combine = "rbind"
	, scale_param = scale_param, fmethod = "effects"
)

## margins
simple_margpred_spec <- condsummary(mod_list=simple_models, fun=function(x, f){
		dd <- cplot(x, f, what="prediction", n=100, draw=FALSE, vcov=function(x)zero_vcov(x, f))
	}, simplify = TRUE, combine = "rbind"
	, scale_param = scale_param, fmethod = "margins"
)

## Combine all the estimates
simple_pred_spec <- do.call("rbind"
	, list(simple_vpred_spec, simple_empred_spec, simple_efpred_spec, simple_margpred_spec)
)
@

<<echo=FALSE>>=
## Plot centered and non-centered separately
simple_pred_spec1 <- subset(simple_pred_spec, model=="x1u"|model=="x1s")
simple_pred_spec2 <- subset(simple_pred_spec, model=="x1c"|model=="x1sd")
@

<<label=simple_pred_spec_plot1, fig=TRUE, echo=FALSE, prefix=FALSE, include=FALSE>>=
simple_pred_spec_plot1 <- (plotpreds(simple_pred_spec1) 
	+ facet_wrap(~method) 
	+ theme(legend.position="bottom")
)
simple_pred_spec_plot1
@

<<label=simple_pred_spec_plot2, fig=TRUE, echo=FALSE, prefix=FALSE, include=FALSE>>=
simple_pred_spec_plot2 <- (plotpreds(simple_pred_spec2) 
	+ facet_wrap(~method) 
	+ theme(legend.position="bottom")
)
simple_pred_spec_plot2
@

\begin{figure}[ht]
\begin{center}
\subfloat[Either unscaled or devided by the \emph{sd}.]{
\includegraphics[width=0.55\textwidth, height=0.54\textheight]{simple_pred_spec_plot1.pdf}
}
~~
\subfloat[Centered or centered and/or devided by \emph{sd}.]{
\includegraphics[width=0.55\textwidth, height=0.54\textheight]{simple_pred_spec_plot2.pdf}
}
\end{center}

\caption{Back-transformed marginal predictions.}
\label{fig:marginal_transformed}
\end{figure}

\begin{figure}[H]
\begin{center}
<<vapred_isolate_numeric, fig=TRUE>>=
## Centered predictions from unscaled model
vpred_c <- varpred(lm_u, focal = "x1u", isolate = TRUE)
plotpreds(vpred_c)
@
\caption{Using \pkg{varpred} to obtain marginal predictions from unscaled model.}
\label{fig:varpred_isolate_num}
\end{center}
\end{figure}

\subsubsection{Categorical predictors}

Back to categorical predictor, \code{x2u}. The reference (base) category has not variance and has such we can not use the \emph{zeroed-out} covariance approach to estimate CI for the marginalized predictions. However, if we can figure out how to center model matrix in the computation of CIs, we can actually use model with unscaled categorical predictors and estimate marginal predictions for centered categorical predictors. Currently, only \pkg{varpred} provide this functionality.

<<simple_marginal_specific_cat, results=hide>>=
## varpred
simple_models_cat <- list(x2u = lm_u)
simple_vpred_cat <- condsummary(mod_list=simple_models_cat, fun=function(x, f){
	dd <- varpred(x, f, isolate=TRUE)
}, simplify = TRUE, combine = "rbind", fmethod = "varpred")

## emmeans
simple_empred_cat <- condsummary(mod_list=simple_models_cat, fun=function(x, f){
	spec <- as.formula(paste0("~", f))
	dd <- emmeans(x, spec=spec, cov.keep=f, vcov. = zero_vcov(x, f))
}, simplify = TRUE, combine = "rbind", fmethod = "emmeans")

## effects
simple_efpred_cat <- condsummary(mod_list=simple_models_cat, fun=function(x, f){
	dd <- Effect(f, x, xlevels=100, vcov. = function(x, complete=FALSE)zero_vcov(x, f, complete))
}, simplify = TRUE, combine = "rbind", fmethod = "effects")

## margins
simple_margpred_cat <- condsummary(mod_list=simple_models_cat, fun=function(x, f){
	dd <- cplot(x, f, what="prediction", n=100, draw=FALSE, vcov=function(x)zero_vcov(x, f))
}, simplify = TRUE, combine = "rbind", fmethod = "margins")

## Combine all the estimates
simple_pred_cat <- do.call("rbind"
	, list(simple_vpred_cat, simple_empred_cat, simple_efpred_cat, simple_margpred_cat)
)
@

<<>>=
head(simple_pred_cat)
simple_pred_cat_plot <- (plotpreds(simple_pred_cat) 
	+ facet_wrap(~method) 
	+ theme(legend.position="bottom")
)
@

\begin{figure}[ht]
\begin{center}
<<simple_pred_cat_plot, fig=TRUE, echo=FALSE>>=
simple_pred_cat_plot
@
\end{center}
\caption{Conditional predictions.}
\end{figure}

%\bibliography{effects_writeup}

\end{document}

### 2021-01-13 effects_writeup.R

@@ -1,0 +1,527 @@
+### R code from vignette source '/home/steve/grive/McMaster/PhD/projects/effects/effects_writeup.Rnw'
+
+###################################################
+### code chunk number 1: set_up
+###################################################
+library(ggplot2)
+library(margins)
+library(effects)
+library(emmeans)
+source("effectsfuns.R")
+varpredtheme()
+
+
+###################################################
+### code chunk number 2: simple_sim
+###################################################
+set.seed(101)
+N <- 100
+x1_min <- 1
+x1_max <- 9
+b0 <- 0.3
+b1 <- 0.1
+b2 <- -0.6
+x2_levels <- factor(c("A", "B"))
+df <- expand.grid(x1u = runif(n=N, min=x1_min, max=x1_max)
+	, x2u = x2_levels
+)
+X <- model.matrix(~x1u + x2u, df)
+betas <- c(b0, b1, b2)
+df$y <- rnorm(nrow(df), mean= X %*% betas, sd=1)
+df2 <- df
+df <- transform(df
+	, x1c = drop(scale(x1u, scale=FALSE)) 
+	, x1s = drop(scale(x1u, center=FALSE))
+	, x1sd = drop(scale(x1u))
+)
+head(df)
+
+
+###################################################
+### code chunk number 3: model_lm_u
+###################################################
+df_temp <- drop(df) 	# margins doesn't work with transform 
+                   	# but we need df to extract attributes :(
+lm_u <- lm(y ~ x1u + x2u, data = df_temp)
+
+
+###################################################
+### code chunk number 4: model_lm_c
+###################################################
+lm_c <- lm(y ~ x1c + x2u, data = df_temp)
+
+
+###################################################
+### code chunk number 5: model_lm_s
+###################################################
+lm_s <- lm(y ~ x1s + x2u, data = df_temp)
+
+
+###################################################
+### code chunk number 6: model_lm_sd
+###################################################
+lm_sd <- lm(y ~ x1sd + x2u, data = df_temp)
+
+
+###################################################
+### code chunk number 7: model_summary
+###################################################
+modsummary <- function(patterns="^lm_", fun, simplify = TRUE
+	, combine = c("cbind", "rbind"), match_colnames = NULL){
[…]

### 2021-01-13 effectsfuns.R

@@ -18,41 +18,49 @@ varpred <- function(mod, focal, isolate = FALSE
 	, isoValue = NULL, level = 0.95, steps = 101
 	, at = NULL,  dfspec = 100, vcmat = NULL) {
 	betahat <- extractcoef(mod)
-	mod_objs <- extract_assign(mod)
-	vnames <- mod_objs$assign
-	modMat <- mod_objs$modmat
-	modFrame <- mod_objs$modframe
-	check_vars <- vnames %in% focal
-	if (!any(check_vars)) stop("Specified variable not in the model")
+	mod_objs <- extractvnames(mod)
+	vnames <- mod_objs$vnames
+	termnames <- mod_objs$termnames
+	mm <- mod_objs$mm
+	mf <- recoverdata(mod)
+	Terms <- mod_objs$Terms
+	rTerms <- delete.response(Terms)
+	if (any(vnames %in% focal)) {
+		check_vars <- vnames %in% focal
+	} else {
+		check_vars <- termnames %in% focal
+	}
+	if (!any(check_vars)) stop(paste0(focal, " not in the model"))
 	focalVar <- unique(vnames[check_vars])
 	
-	xlevels <- .getXlevels(terms(mod), modFrame)
+	xlevels <- .getXlevels(Terms, model.frame(mod))
 	contrs <- mod$contrasts
 
 	if(is.null(at)) {
-		vv <- varfun(modFrame[[focalVar]], steps)
+		vv <- varfun(mf[[focalVar]], steps)
 	} else {
 		vv <- at
 	}
   	steps <- length(vv)
 	
-	rowMean <- apply(modMat, 2, mean)
+	rowMean <- apply(mm, 2, mean)
 	modMean <- t(replicate(steps, rowMean))
 
 	# Model matrix with progression of focal variable
-	varFrame <- modFrame[rep(1, steps), ]
+	varFrame <- mf[rep(1, steps), ]
 	varFrame[focalVar] <- vv
-	modTerms <- delete.response(terms(mod))
-	varFrame <- model.frame(modTerms, varFrame
+	varFrame <- model.frame(rTerms, varFrame
 		, xlev = xlevels, drop.unused.levels = TRUE
 	)
-	newMat <- model.matrix(modTerms, varFrame
+	newMat <- model.matrix(rTerms, varFrame
 		, contr = contrs, xlev = xlevels
 	)
 
 	## Better way to do this?
 	modVar <- modMean
 	focal_cols <- grep(focalVar, names(vnames), value = TRUE)
[…]

### 2021-01-13 predict_jdeffect.R

@@ -5,7 +5,7 @@ source("makestuff/makeRfuns.R")
 commandEnvironments()
 sourceFiles()
 
-## Varpre function function
+## Varpred function function
 varpredfun <- function(mod, focal, at, vcmat = NULL, model){
 	if (is.null(vcmat)){
 		pred <- varpred(mod, focal, at = at)

### 2021-01-15 effects_writeup.Rnw
jdeffects

@@
 library(emmeans)
-source("effectsfuns.R")
-varpredtheme()
+library(jdeffects)
+jdtheme()
 @
@@
 <<condional_summary, cho=FALSE>>=
-condsummary <- function(mod_list, fun, simplify = TRUE
+condsummary <- function(mod_list, fun, resp = "y", simplify = TRUE
 	, combine = c("cbind", "rbind"), scale_param = list(), fmethod = NULL){
@@
 		} else {
+			out <- data.frame(out)
 			colnames(out)[colnames(out)%in%f] <- "xvar"
@@
 	}, simplify = FALSE)
+	f <- attr(out, "focal")
 	if (simplify){
@@
 	}
+	attr(out, "response") <- resp
 	return(out)
@@
 head(simple_pred_all)
-simple_pred_all_plot <- (plotpreds(simple_pred_all) 
+class(simple_pred_all) <- c("jdeffects", "data.frame") # plot.effects
+simple_pred_all_plot <- (plot(simple_pred_all) 
 	+ facet_wrap(~method) 
@@
 head(simple_pred_spec)
-simple_pred_spec_plot <- (plotpreds(simple_pred_spec) 
+class(simple_pred_spec) <- c("jdeffects", "data.frame") # plot.effects
+simple_pred_spec_plot <- (plot(simple_pred_spec)
 	+ facet_wrap(~method)
@@
 head(simple_pred_cat)
-simple_pred_cat_plot <- (plotpreds(simple_pred_cat) 
+class(simple_pred_cat) <- c("jdeffects", "data.frame") # plot.effects
+simple_pred_cat_plot <- (plot(simple_pred_cat)
 	+ facet_wrap(~method)
@@
 Notice that the modification in each of the method function call to pass either covariance matrix or function.
-In particular, in \pkg{varpred}, it is passed as \pkg{vcmat}, \pkg{emmeans} and \pkg{effects} as \pkg{vcov.} and \pkg{margins} as \pkg{vcov}.
+In particular, in \pkg{varpred}, it is passed as \pkg{vcov.}, \pkg{emmeans} and \pkg{effects} as \pkg{vcov.} and \pkg{margins} as \pkg{vcov}.
 However, this seems not to work currently in \pkg{cplot} which computes the predictions in \pkg{margins}.
@@
 simple_vpred_all <- condsummary(mod_list=simple_models, fun=function(x, f){
-	dd <- varpred(x, f, vcmat = zero_vcov(x, f))
+	dd <- varpred(x, f, vcov. = zero_vcov(x, f))
 }, simplify = TRUE, combine = "rbind", fmethod = "varpred")
@@
 <<echo=FALSE>>=
-simple_pred_all_plot <- (plotpreds(simple_pred_all) 
+class(simple_pred_all) <- c("jdeffects", "data.frame")
+simple_pred_all_plot <- (plot(simple_pred_all)
 	+ facet_wrap(~method)
@@
 simple_vpred_spec <- condsummary(mod_list=simple_models, fun=function(x, f){
-		dd <- varpred(x, f, vcmat = zero_vcov(x, f))
+		dd <- varpred(x, f, vcov. = zero_vcov(x, f))
 	}, simplify = TRUE, combine = "rbind"
@@
 <<label=simple_pred_spec_plot1, fig=TRUE, echo=FALSE, prefix=FALSE, include=FALSE>>=
-simple_pred_spec_plot1 <- (plotpreds(simple_pred_spec1) 
+class(simple_pred_spec1) <- c("jdeffects", "data.frame")
+simple_pred_spec_plot1 <- (plot(simple_pred_spec1)
 	+ facet_wrap(~method)
@@
 <<label=simple_pred_spec_plot2, fig=TRUE, echo=FALSE, prefix=FALSE, include=FALSE>>=
-simple_pred_spec_plot2 <- (plotpreds(simple_pred_spec2) 
+class(simple_pred_spec2) <- c("jdeffects", "data.frame")
+simple_pred_spec_plot2 <- (plot(simple_pred_spec2)
 	+ facet_wrap(~method)
@@
 vpred_c <- varpred(lm_u, focal = "x1u", isolate = TRUE)
-plotpreds(vpred_c)
+plot(vpred_c)
 @
@@
 head(simple_pred_cat)
-simple_pred_cat_plot <- (plotpreds(simple_pred_cat) 
+class(simple_pred_cat) <- c("jdeffects", "data.frame")
+simple_pred_cat_plot <- (plot(simple_pred_cat)
 	+ facet_wrap(~method)
@@
 
-%\bibliography{effects_writeup}
-
 \end{document}

### 2021-01-15 R/effectsfuns.R

@@ -337,132 +337,3 @@ recoverdata <- function(mod, envir = environment(formula(mod)), ...) {
 	return(df)
 }
 
-#' Plot  predictions 
-#' 
-#' Plots estimated conditional or marginal predictions. 
-#' @param x \code{\link[jdeffects]{varpred}} object
-#' @param ... for future implementations
-#' @param xlabs x-axis label. If \code{NULL}, default, \code{x.var} is used.
-#' @param ylabs y-axis label. If \code{NULL}, default, the response label is used.
-#' @param pos spacing between labels of categorical variable on the plot. 
-#'
-#' @return a \code{\link[ggplot2]{ggplot}} object.
-#'
-#' @seealso
-#'\code{\link[jdeffects]{varpred}}
-#'
-#' @examples
-#' set.seed(4567)
-#' x <- rnorm(100, 3, 5)
-#' y <- 0.4 + 0.7*x + rnorm(100)
-#' df <- data.frame(y = y, x = x)
-#' m1 <- lm(y ~ x, df)
-#'
-#' pred1 <- varpred(m1, "x")
-#' plot(pred1)
-#'
-#' @import ggplot2
-#' @export
-
-plot.jdeffects <- function(x, ..., xlabs = NULL, ylabs = NULL, pos = 0.5){
-
-	lwr <- upr <- NULL
-	df <- x$preds
-	focal <- attr(df, "focal")
-	if (!is.null(focal))	n.focal <- length(focal) else n.focal <- 1L
-	x.var <- attr(df, "x.var")
-	if (!is.null(x.var))	 x.var <- x.var else x.var <- focal[[1]]
-	if (n.focal>1L) {
-		non.focal <- focal[!focal %in% x.var]
-		df[, non.focal] <- sapply(non.focal, function(x){
-			xx <- df[[x]]
-			xx <- if (class(xx) %in% c("numeric", "integer")) round(xx,1) else xx
-			ll <- paste0(x, "==", xx)
-			return(ll)
-		})	
-		gform <- as.formula(paste0(".~", paste0(non.focal, collapse="+")))
-	}
-	if (is.null(xlabs)) xlabs <- x.var
-	if (is.null(ylabs)) ylabs <- attr(df, "response")
-	pos <- position_dodge(pos)
-	if (is.null(df$model)){
-		p1 <- (ggplot(df, aes_string(x = x.var, y = "fit"), colour="black", alpha = 0.2)
-			+ guides(fill = FALSE)
[…]

### 2021-01-15 effects_writeup.R
Autocommit

@@ -226,7 +226,7 @@ simple_pred_all <- do.call("rbind"
 
 
 ###################################################
-### code chunk number 14: effects_writeup.Rnw:297-303
+### code chunk number 14: effects_writeup.Rnw:298-304
 ###################################################
 head(simple_pred_all)
 class(simple_pred_all) <- c("jdeffects", "data.frame") # plot.effects
@@ -287,7 +287,7 @@ simple_pred_spec <- do.call("rbind"
 
 
 ###################################################
-### code chunk number 17: effects_writeup.Rnw:360-366
+### code chunk number 17: effects_writeup.Rnw:361-367
 ###################################################
 head(simple_pred_spec)
 class(simple_pred_spec) <- c("jdeffects", "data.frame") # plot.effects
@@ -336,7 +336,7 @@ simple_pred_cat <- do.call("rbind"
 
 
 ###################################################
-### code chunk number 20: effects_writeup.Rnw:412-418
+### code chunk number 20: effects_writeup.Rnw:413-419
 ###################################################
 head(simple_pred_cat)
 class(simple_pred_cat) <- c("jdeffects", "data.frame") # plot.effects
@@ -395,7 +395,7 @@ simple_pred_all <- do.call("rbind"
 
 
 ###################################################
-### code chunk number 24: effects_writeup.Rnw:484-489
+### code chunk number 24: effects_writeup.Rnw:485-490
 ###################################################
 class(simple_pred_all) <- c("jdeffects", "data.frame")
 simple_pred_all_plot <- (plot(simple_pred_all)
@@ -449,7 +449,7 @@ simple_pred_spec <- do.call("rbind"
 
 
 ###################################################
-### code chunk number 27: effects_writeup.Rnw:546-549
+### code chunk number 27: effects_writeup.Rnw:547-550
 ###################################################
 ## Plot centered and non-centered separately
 simple_pred_spec1 <- subset(simple_pred_spec, model=="x1u"|model=="x1s")
@@ -519,7 +519,7 @@ simple_pred_cat <- do.call("rbind"
 
 
 ###################################################
-### code chunk number 32: effects_writeup.Rnw:633-639
+### code chunk number 32: effects_writeup.Rnw:634-640
 ###################################################
 head(simple_pred_cat)
 class(simple_pred_cat) <- c("jdeffects", "data.frame")

### 2021-01-18 effects_writeup.Rnw
Autocommit

@@
 
-\section{Yet to figure out}
+\section{TODOs}
 \begin{enumerate}
-\item Handling complex interaction
-\item Zero SEs in the case of variance-covariance approach
+\item Models with interactions
+\item GLMs
+\item LMEs
 \end{enumerate}
 
-\section{Illustrations}
+\section{No interactions}
 
@@
 \begin{itemize}
-\item $y = \beta_0 + \beta_1x_1 + \beta_2x_2 + \epsilon$ s.t $\epsilon \sim N(0, 1)$ and $\{\beta_0 = -5, \beta_1 = 0.1, \beta_2 = -0.05\}$
+\item $y = \beta_0 + \beta_1x_1 + \beta_2x_2 + \epsilon$ s.t $\epsilon \sim N(0, 1)$ and $\{\beta_0 = -5, \beta_1 = 0.1, \beta_2 = -0.6\}$
 \begin{itemize}
@@
 
-Log likelihoods
+These four model are equivalent (they all have equal loglikehoods).
 
@@
 
-Variance covariance matrix
+Variance covariance matrix. It is important to note that the slope variance for unscaled and centered models are the same. This makes it easier to center predictions from uncentered models.
 <<simple_vcov>>=
@@
 
-The function will be used downstream to \emph{tidy} conditional predictions from each of the methods highlighted above.
-
-<<condional_summary, cho=FALSE>>=
+<<condional_summary, echo=FALSE>>=
 condsummary <- function(mod_list, fun, resp = "y", simplify = TRUE
@@
 
-We start by computing the conditional predictions using each of the methods (\pkg{varpred, emmeans, effects, margins}).
-In the first case, the predictions are displayed on the variable level scale-specific values (original (u), mean centered (c), divided by standard deviation (s), and both mean centered and scaled (sd)).
-In the second case, the predictions are transformed back to the original scale.
-
-<<simple_model>>=
+<<simple_model, echo=FALSE>>=
 simple_models <- list(x1u = lm_u, x1c = lm_c, x1s = lm_s, x1sd = lm_sd)
@@
 
-<<simple_conditional_specific, results=hide>>=
+<<simple_conditional_specific, echo=FALSE, results=hide>>=
 ## varpred
@@
 
-<<>>=
-head(simple_pred_all)
+<<echo=FALSE>>=
 class(simple_pred_all) <- c("jdeffects", "data.frame") # plot.effects
@@
 
-\begin{figure}[ht]
+\begin{figure}[H]
 \begin{center}
@@
 \caption{Conditional predictions on the variable-specific values.}
+\label{fig:simple_pred_all_plot}
 \end{figure}
 
-Compute predictions and then back transform each predictor to the original scale.
+\autoref{fig:simple_pred_all_plot}, displays the conditional predictions for of the models, i.e., variable-specific values (original (u), mean centered (c), divided by standard deviation (s), and both mean centered and scaled (sd)).
+The predictions looks different but we zoom-in in \autoref{fig:simple_pred_spec_plot} by transforming back all the scaled predictions to the original (unscaled) values -- and actually, conditional predictions are the same for each of the methods.
+This implies that, if we know the scaling or centering parameter, we can always transform back predictions to the untransformed (unscaled) equivalent.
 
-<<simple_conditional_backtrans, results=hide>>=
+<<simple_conditional_backtrans, echo=FALSE, results=hide>>=
 ## Extract scaling parameters
@@
 
-<<>>=
-head(simple_pred_spec)
+<<echo=FALSE>>=
 class(simple_pred_spec) <- c("jdeffects", "data.frame") # plot.effects
@@
 \caption{Back-transformed conditional predictions.}
+\label{fig:simple_pred_spec_plot}
 \end{figure}
@@
 
-Here, we are considering case where we are only interested estimating the conditional predictions.
-In this case, it doesn't matter which of the four models we choose since they will give the same predictions.
+In this case, it doesn't matter which of the four models we choose since they all give the similar conditional predictions.
 Later, we'll show how we can estimate marginalized and/or centered (CI) for categorical predictors.
 
-<<simple_conditional_specific_cat, results=hide>>=
+<<simple_conditional_specific_cat, echo=FALSE, results=hide>>=
 ## varpred
@@
 
-<<>>=
-head(simple_pred_cat)
+<<echo=FALSE>>=
 class(simple_pred_cat) <- c("jdeffects", "data.frame") # plot.effects
@@
 
-We can implement marginalization by \emph{zeroing-out} the variance of non-focal predictors (possible in all four methods) but this requires that the predictors are centered beforehand.
-On the other hand, in \pkg{varpred} marginal predictions can be obtained by centering the SE estimates.
+We can implement marginalization by \emph{zeroing-out} the variance of non-focal predictors (possible in all four methods) but this requires that continuous predictors are centered beforehand or appropriate contrast is applied in the case of categorical predictors.
+On the other hand, in \pkg{varpred} marginal predictions are computed by centering the model matrix (when computing SEs).
+This approach does not require centered predictors and also works with any contrast specification.
 
@@
 
-We \pkg{zero\_vcov} function in \pkg{varpred} to transform the variances of non-focal predictors to zero. For example:
+The \pkg{zero\_vcov} function in \pkg{jdeffects} \emph{zeros-out} variance-covariances of non-focal terms of the predictor in question. For example:
 
-<<zero_out_ex>>=
+\begin{itemize}
+\item full variance-covariance matrix
+<<full_vcov, echo=FALSE>>=
 lm_u_vcov <- vcov(lm_u)
 print(lm_u_vcov)
+@
 
-## x1u is the focal variable
+\item \emph{zeroed-out} variance-covariance matrix with \code{x1u} as the focal variable
+
+<<zero_out_ex, echo=FALSE>>=
 lm_u_vcov_zero <- zero_vcov(lm_u, focal_vars = "x1u")
@@
 @
+\end{itemize}
 
-We'll repeat the same procedure in conditional predictions section but modify the functions to incorporate \emph{zeroed-out} covariance matrix.
+We repeat all the procedures in conditional predictions section but modify the methods to incorporate \emph{zeroed-out} covariance matrix as one of the inputs.
 
@@
 
-Notice that the modification in each of the method function call to pass either covariance matrix or function.
-In particular, in \pkg{varpred}, it is passed as \pkg{vcov.}, \pkg{emmeans} and \pkg{effects} as \pkg{vcov.} and \pkg{margins} as \pkg{vcov}.
+The most modification in each of the methods is to either specify \emph{zeroed-out} covariance matrix or a function to compute the variance-covariance matrix as one of the inputs.
+In particular, in \pkg{varpred}, \pkg{emmeans} and  \pkg{effects} it is specified through \pkg{vcov.} and in \pkg{margins} through \pkg{vcov} argument.
 However, this seems not to work currently in \pkg{cplot} which computes the predictions in \pkg{margins}.
 
-<<simple_marginal_specific, results=hide>>=
+<<simple_marginal_specific, echo=FALSE, results=hide>>=
 ## varpred
@@
 Back to categorical predictor, \code{x2u}.
-The reference (base) category has not variance and has such we can not use the \emph{zeroed-out} covariance approach to estimate CI for the marginalized predictions.
-However, if we can figure out how to center model matrix in the computation of CIs, we can actually use model with unscaled categorical predictors and estimate marginal predictions for centered categorical predictors.
+The default contrast compares the other categories with no-variance reference category (base) and as a result, we can not compute the SEs using \emph{zeroed-out} variance-covariance matrix unless appropriate contrast (\code{"contr.sum"}) was used during model specification.
+Model matrix centering provides a way to overcome this and computes centered independent of the contrasts.
 Currently, only \pkg{varpred} provide this functionality.
 
-<<simple_marginal_specific_cat, results=hide>>=
+<<simple_marginal_specific_cat, echo=FALSE, results=hide>>=
 ## varpred
 simple_models_cat <- list(x2u = lm_u)
-simple_vpred_cat <- condsummary(mod_list=simple_models_cat, fun=function(x, f){
+simple_margvpred_cat <- condsummary(mod_list=simple_models_cat, fun=function(x, f){
 	dd <- varpred(x, f, isolate=TRUE)
@@
 ## emmeans
-simple_empred_cat <- condsummary(mod_list=simple_models_cat, fun=function(x, f){
+simple_margempred_cat <- condsummary(mod_list=simple_models_cat, fun=function(x, f){
 	spec <- as.formula(paste0("~", f))
@@
 ## effects
-simple_efpred_cat <- condsummary(mod_list=simple_models_cat, fun=function(x, f){
+simple_margefpred_cat <- condsummary(mod_list=simple_models_cat, fun=function(x, f){
 	dd <- Effect(f, x, xlevels=100, vcov. = function(x, complete=FALSE)zero_vcov(x, f, complete))
@@
 ## Combine all the estimates
-simple_pred_cat <- do.call("rbind"
-	, list(simple_vpred_cat, simple_empred_cat, simple_efpred_cat, simple_margpred_cat)
+simple_margpred_cat <- do.call("rbind"
+	, list(simple_margvpred_cat, simple_margempred_cat, simple_margefpred_cat, simple_margpred_cat)
 )
@@
 
-<<>>=
-head(simple_pred_cat)
-class(simple_pred_cat) <- c("jdeffects", "data.frame")
-simple_pred_cat_plot <- (plot(simple_pred_cat)
+<<echo=FALSE>>=
+class(simple_margpred_cat) <- c("jdeffects", "data.frame")
+simple_margpred_cat_plot <- (plot(simple_margpred_cat)
 	+ facet_wrap(~method)
@@
 \begin{center}
-<<simple_pred_cat_plot, fig=TRUE, echo=FALSE>>=
-simple_pred_cat_plot
+<<simple_margpred_cat_plot, fig=TRUE, echo=FALSE>>=
+simple_margpred_cat_plot
 @

### 2021-01-18 R/utilities.R
Autocommit

@@ -155,7 +155,7 @@ clean_model <- function(focal.predictors, mod, xlevels = list()
     }
   }
   factor.cols[grep(":", cnames)] <- FALSE   
-  X <- recoverdata(mod)
+  X <- recoverdata(mod, all.predictors)
   
   which.matrices <- sapply(X, function(x) is.matrix(x) && ncol(x) == 1)
   if (any(which.matrices)){

### 2021-01-18 dev/demonstrate.R

@@ -1,6 +1,8 @@
+library(dplyr)
 library(jdeffects)
 library(shellpipes)
 
+jdtheme()
 startGraphics()
 
 set.seed(233)
@@ -10,12 +12,36 @@ beta_dep = 7
 sig = 12
 dmean = 5
 
-dep <- runif(100) + dmean
+dep <- runif(100)
 eps <- runif(100)
 
 ind <- beta_dep*dep + beta_int + sig*eps
 
-m <- lm(ind~dep)
+## Unscaled
+m_un <- lm(ind~dep)
+pvp_un <- varpred(m_un, "dep")
+plot(pvp_un)
 
-pvp <- varpred(m, "dep")
-plot(pvp)
+## Centered dep
+dep_c <- scale(dep, scale=FALSE)
+m_c <- lm(ind~dep_c)
+pvp_c <- varpred(m_c, "dep_c")
+plot(pvp_c)
+
+## Plot them on one axis
+pvp_df <- (bind_rows(
+	(pvp_un$preds
+		%>% data.frame()
+		%>% rename("xvar"=dep)
+		%>% mutate(model="dep_un")) 
+	, (pvp_c$preds
+		%>% data.frame()
+		%>% rename(xvar=dep_c)
+		%>% mutate(model="dep_c"))
+	)
+)
+head(pvp_df)
+
+### Make pvp jdeffects object and use plot method
+class(pvp_df) <- c("jdeffects", "data.frame")
+plot(pvp_df)

### 2021-01-18 effects_writeup.R

@@ -1,4 +1,4 @@
-### R code from vignette source '/home/dushoff/screens/research/effects/effects_writeup.Rnw'
+### R code from vignette source '/home/steve/grive/McMaster/PhD/projects/effects/effects_writeup.Rnw'
 
 ###################################################
 ### code chunk number 1: set_up
@@ -72,7 +72,7 @@ modsummary <- function(patterns="^lm_", fun, simplify = TRUE
 	combine <- match.arg(combine)
 	out <- sapply(mod, function(m){
 		out <- fun(get(m))
-		out <- round(out, 4)
+		out <- round(out, 3)
 		if (combine=="rbind"){
 			if (!is.null(match_colnames)){
 				colnames(out) <- match_colnames
@@ -405,7 +405,7 @@ simple_pred_all_plot <- (plot(simple_pred_all)
 
 
 ###################################################
-### code chunk number 26: simple_pred_all_plot
+### code chunk number 26: simple_margpred_all_plot
 ###################################################
 simple_pred_all_plot
 

### 2021-02-01 effects_writeup.Rnw
Autocommit

@@
 
-\bibliographystyle{chicago}
 \setlength{\parindent}{0pt}
@@
 In this task, we describe various R machineries for displaying predictions in both simple and complex (involving interaction terms) generalized linear models.
-We first consider existing approaches for \emph{conditioning} predictions and then describe our proposed approach for \emph{marginalized} predictions.
-These provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind transformations on the dependent variables whose estimates are usually, but not always, a subject to less clarity of interpretation.
+We first consider existing approaches for constructing model predictions and then describe how our proposed approach is different from the existing ones.
+When visually presented, predictions provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind transformations on the dependent variables whose estimates are usually, but not always, a subject to less clarity of interpretation.
+
+A related term to \emph{marginal} predictions is the \emph{conditional} predictions --- it is not clear whether conditional or marginal predictions is more appropriate for a particular research problem \citep{muff2016marginal}.
+However, we attempt to make the distinction between the two (see \href{https://github.com/mac-theobio/effects/blob/main/glossary.md}{glossary}).
+
+By default, commonly used R packages for constructing predictions do not remove (\emph{zero-out}) the uncertainties resulting from the \emph{non-focal} predictors when computing predictions.
+A non-trivial way to achieve this in some of these packages is to provide a user defined variance-covariance matrix with the covariances of \emph{non-focal} terms set to $0$ -- \emph{zeroing-out} variance-covariance matrix.
+This only works when the predictors are centered prior to model fitting, in case of numerical predictors, and even much complicated when the predictors are categorical.
+We propose a method independent of model scaling to provide more robust \emph{marginal} predictions.
 
@@
 The first and most obvious, is the model coefficient estimates.
-Others include predicted values of the outcome variable---1) predictions at a particular; and 2) mean or median value of the predictors.
+Others include predicted values of the outcome variable --- 1) predictions at a particular; and 2) mean or median value of the predictors.
 The first case simply involves evaluation of the fitted model function, say $\hat{f}(X = x)$, at some particular value of $x$.
@@
 
-Most importantly, from these predicted values, we can also generate second class quantities of interest -- \emph{conditional} or \emph{marginal} predictions.
-These quantities describes the change in the predicted value of the dependent variable after changing one independent variable -- either a discrete change in the categorical variable(s) or an instantaneous change in continuous variables, while all other variable are held at specified values.
+Most importantly, from these predicted values, we can also generate second class quantities of interest --- \emph{marginal} predictions, which describes the change in the predicted value of the dependent variable after changing one independent variable -- either a discrete change in the categorical variable(s) or an instantaneous change in continuous variables, while all other variable are held at specified values.
 
-Suppose we are interested in the \emph{conditional} predictions of a particular predictor (hence forth referred as \emph{focal} predictor otherwise \emph{non-focal}), $x_f$, in the set of predictors.
+Suppose we are interested in the \emph{marginal} predictions of a particular predictor (hence forth referred as \emph{focal} predictor otherwise \emph{non-focal}), $x_f$, in the set of predictors.
 To keep it simple, assume that the model has no interaction terms.
 Then the idea is to fix the values of \emph{non-focal} predictor(s) at some typical values -- typically determined by averaging in some meaningful way, for example, arithmetic mean and average over the levels of the factors of \emph{non-focal} continuous and categorical predictors, respectively.
-This is achieved by averaging the columns of \emph{model matrix}, $\mathbf{X}$, except for the column of focal predictor.
+An alternative to \emph{averaging} is \emph{anchoring} which involves picking a fixed value of the predictor.
+One way to achieve this is using \emph{model matrix},  $\mathbf{X}$, i.e., \emph{anchoring} or \emph{averaging} the columns of \emph{non-focal} terms in $\mathbf{X}$.
+
+\subsection{Simple linear models}
 
@@
 Let $\hat{\beta}$ be the estimate of $\beta$, together with the estimated covariance matrix $V(\hat{\beta})$ of $\hat{\beta}$.
-Let the entries of $\mathbf{X^*}$ include all conditioned (\emph{non-focal}) and \emph{focal} predictors.
+Let the entries of $\mathbf{X^*}$ include all \emph{non-focal} and \emph{focal} predictors.
 The model matrix, $\mathbf{X^*}$, inherits most of its key properties, for example transformation (e.g., scaling) on the predictors and interactions from the model matrix, $\mathbf{X}$.
-Then the predicted values $\hat{\eta}^* = \mathbf{X^*}\hat{\beta}$ represents the \emph{conditional} predictions of the focal predictor.
+Then the predicted values $\hat{\eta}^* = \mathbf{X^*}\hat{\beta}$ represents the \emph{marginal} predictions of the focal predictor.
 Alternatively, we can transform these predictions to response scale using $g^{-1}(\hat{\eta}^*)$.
 
-Further, we can compute the standard errors (SEs) of the \emph{conditional} predictions, $\hat{\eta}^*$, for constructing confidence intervals, as the $\operatorname{sqrt}(\operatorname{diag}(\mathbf{X}^*V(\hat{\beta})\mathbf{X}^{*T}))$.
-When computing \emph{marginal} predictions, the \emph{uncertainty} as a result of \emph{non-focal} predictors are removed.
+Further, we can compute the standard errors (SEs) associated with these predictions, $\hat{\eta}^*$, for constructing confidence intervals, as the $\operatorname{sqrt}(\operatorname{diag}(\mathbf{X}^*V(\hat{\beta})\mathbf{X}^{*T}))$.
+Our proposed method goes one step further of removing the \emph{uncertainty} as a result of \emph{non-focal} predictors are removed.
 This can be achieved in two ways:
@@
 
-\subsection{Variance-covariance}
+\subsubsection{Variance-covariance}
 
@@
 
-\subsection{Centered model matrix}
+\subsubsection{Centered model matrix}
 
-Suppose the \emph{non-focal} entries in $X^*$ are computed by some kind of averaging.
+Suppose the \emph{non-focal} entries in $X^*$ are computed by some kind of averaging or anchoring.
 Consider centered $X^*$, $X^*_{c} = (X^* - \bar{X^*})$.
@@
 
+\subsection{GLMs}
+
+\subsection{LMEs}
+
 \section{Available packages}
-The following R packages \pkg{effects}, \pkg{emmeans} and \pkg{margins} implement various schemes for constructing \emph{conditional} predictions.
-However, currently, their ability to compute \emph{marginal} prediction is limited to the use of variance-covariance approach which requires the fitted model to be centered.
+The following R packages \pkg{effects}, \pkg{emmeans} and \pkg{margins} implement various schemes for constructing \emph{marginal} predictions.
+However, currently, their ability to compute \emph{marginal} predictions with zeroed-out non-focal uncertanities is limited to the use of variance-covariance approach which requires the fitted model to be centered.
 We propose \pkg{varpred} to overcome this limitation.
@@
 
-<<simple_sim>>=
+<<simple_sim, echo=FALSE>>=
 set.seed(101)
@@
 
-\subsection{Conditional predictions}
+\subsection{All uncertainties included}
 
@@
 \end{center}
-\caption{Conditional predictions on the variable-specific values.}
+\caption{Predictions on the variable-specific values.}
 \label{fig:simple_pred_all_plot}
@@
 
-\autoref{fig:simple_pred_all_plot}, displays the conditional predictions for of the models, i.e., variable-specific values (original (u), mean centered (c), divided by standard deviation (s), and both mean centered and scaled (sd)).
-The predictions looks different but we zoom-in in \autoref{fig:simple_pred_spec_plot} by transforming back all the scaled predictions to the original (unscaled) values -- and actually, conditional predictions are the same for each of the methods.
-This implies that, if we know the scaling or centering parameter, we can always transform back predictions to the untransformed (unscaled) equivalent.
+\autoref{fig:simple_pred_all_plot}, displays the simple marginal predictions for each of the four models described above, i.e., variable-specific values (original (u), mean centered (c), divided by standard deviation (s), and both mean centered and scaled (sd)).
+The uncertainties associated with these predictions includes those for non-focal predictors too.
+The predictions looks different but they are actually similar when plotted on the same scale, see \autoref{fig:simple_pred_spec_plot}.
 
@@
 \end{center}
-\caption{Back-transformed conditional predictions.}
+\caption{Back-transformed conditional predictions.
+Remember the model predictors were in different scales -- this implies that if we know the scaling or centering parameter, we can always back-transform model predictions to the original (unscaled) form.}
 \label{fig:simple_pred_spec_plot}
@@
 
-In this case, it doesn't matter which of the four models we choose since they all give the similar conditional predictions.
-Later, we'll show how we can estimate marginalized and/or centered (CI) for categorical predictors.
+In this case, it doesn't matter which of the four models we choose since they all give the similar predictions.
 
@@
 
-\subsection{Marginal predictions}
+\subsection{Zeroed-out uncertainties}
 
-We can implement marginalization by \emph{zeroing-out} the variance of non-focal predictors (possible in all four methods) but this requires that continuous predictors are centered beforehand or appropriate contrast is applied in the case of categorical predictors.
-On the other hand, in \pkg{varpred} marginal predictions are computed by centering the model matrix (when computing SEs).
-This approach does not require centered predictors and also works with any contrast specification.
+We can remove uncertainties associated with non-focal predictors by \emph{zeroing-out} the variance of non-focal terms of the variance-covariance matrix (possible in all four methods) but this requires that continuous predictors are centered or appropriate contrast is applied in the case of categorical predictors before model fitting.
+In \pkg{varpred} marginal predictions are computed by centering the model matrix (when computing SEs), whic does not require centered predictors and is independent of contrast specification.
 
-\subsection{zeroed-out covariance matrix}
+\subsubsection{zeroed-out covariance matrix}
+
+The modification in each of the methods is to either specify \emph{zeroed-out} covariance matrix or a function to compute the variance-covariance matrix as one of the inputs.
+In particular, in \pkg{varpred}, \pkg{emmeans} and  \pkg{effects} it is specified through \pkg{vcov.} and in \pkg{margins} through \pkg{vcov} argument.
+However, this seems not to work currently in \pkg{cplot} which computes the predictions in \pkg{margins}.
 
@@
 
-We repeat all the procedures in conditional predictions section but modify the methods to incorporate \emph{zeroed-out} covariance matrix as one of the inputs.
-
-\subsubsection{Continuous predictor}
-
-The most modification in each of the methods is to either specify \emph{zeroed-out} covariance matrix or a function to compute the variance-covariance matrix as one of the inputs.
-In particular, in \pkg{varpred}, \pkg{emmeans} and  \pkg{effects} it is specified through \pkg{vcov.} and in \pkg{margins} through \pkg{vcov} argument.
-However, this seems not to work currently in \pkg{cplot} which computes the predictions in \pkg{margins}.
+For the continuous predictors, we first compute predictions on predictor-specific scales (\autoref{fig:simple_pred_all_plot_out}) and then back-transform the predictions to original scale (unscale -- uncenter and/or unscale).
+To clearly distinguish between what happens when apply various scaling schemes, we separately do the plots (see \autoref{fig:marginal_transformed}).
 
@@
 class(simple_pred_all) <- c("jdeffects", "data.frame")
-simple_pred_all_plot <- (plot(simple_pred_all)
+simple_pred_all_plot_out <- (plot(simple_pred_all)
 	+ facet_wrap(~method)
@@
 \begin{center}
-<<simple_pred_all_plot, fig=TRUE, echo=FALSE>>=
-simple_pred_all_plot
+<<simple_margpred_all_plot, fig=TRUE, echo=FALSE>>=
+simple_pred_all_plot_out
 @
@@
 \caption{Marginal predictions with the zeroed-out covariances on the variable-specific scale.}
+\label{fig:simple_pred_all_plot_out}
 \end{figure}
 
-Also, we repeat above implementation but back-transform the predictions to original scale (unscale -- uncenter and/or unscale) by modifying the previous code to include scaling parameters.
-To clearly distinguish between what happens when apply various scaling schemes, we separately do the plots (see \autoref{fig:marginal_transformed}).
-
-\subsection{Centered model matrix}
+\subsubsection{Centered model matrix}
 
@@
 In our example, we use the unscaled model \code{lm\_u} and simply set \code{isolate=TRUE}.
-The predictions (see \autoref{fig:varpred_isolate_num}) are similar to \autoref{fig:marginal_transformed}~b (except for the \pkg{margins}).
+The predictions (see \autoref{fig:varpred_isolate_num}) are similar to \autoref{fig:marginal_transformed}~b (except \pkg{margins}).
 
@@
 
-\begin{figure}[H]
+\begin{figure}
 \begin{center}
@@
 
-\begin{figure}[ht]
+\begin{figure}
 \begin{center}
@@
 
+\clearpage
+% References
+\bibliographystyle{plainnat}
+\bibliography{effects_writeup}
+
 \end{document}

### 2021-02-08 effects_writeup.Rnw
Updated glossary
Comparing base R with varpred and emmeans

@@
 
-\title{Marginal predictions}
+\title{Visualizing predictions}
 
@@
 library(emmeans)
-library(jdeffects)
-jdtheme()
+library(jdeffects); jdtheme()
+
+#library(shellpipes)
+#sourceFiles()
+
 @
@@
 simple_vpred_spec <- condsummary(mod_list=simple_models, fun=function(x, f){
-		dd <- varpred(x, f)
+		dd <- varpred(x, f, steps=101)
 	}, simplify = TRUE, combine = "rbind"
@@
 
-\begin{figure}
+\begin{figure}[H]
 \begin{center}
@@
 simple_margpred_cat_plot
+@
+\end{center}
+\caption{Conditional predictions.}
+\end{figure}
+
+\subsubsection{Comparison with base R predict}
+
+<<compare_baseR, echo=FALSE>>=
+load("predict_plots.rda")
+@
+
+Compare prediction estimates from \pkg{base R}, \pkg{emmeans} and \pkg{varpred}
+
+<<>>=
+## Base R vs emmeans
+all.equal(base, emm)
+## Base R vs varpred
+all.equal(base, jd)
+## Implied: emmeans vs varpred
+@
+
+\begin{figure}[H]
+\begin{center}
+<<compare_baseR_plot, fig=TRUE, echo=FALSE>>=
+print(x1_plot)
 @

### 2021-02-08 cpred_baseR.R

@@ -1,3 +1,5 @@
+library(shellpipes)
+
 ## Customize based R predict method to add CI and zero-out non-focal predictors
 ### Manually compute the uncertainty around the estimates
 
@@ -30,3 +32,5 @@ cpredict <- function(mod, newdata, level = 0.95, vvfun = NULL) {
 	)
 	return(pred_df)
 }
+
+saveEnvironment()

### 2021-02-08 predict_baseR.R

@@ -1,7 +1,8 @@
 library(dplyr)
-source("makestuff/makeRfuns.R")
+library(jdeffects)
+library(shellpipes)
+
 commandEnvironments()
-sourceFiles()
 
 ## Predict function. Calls customized base R predict
 predfun <- function(mod, var, newdata, model, level = 0.95, vvfun=NULL){
@@ -48,7 +49,7 @@ head(x1std_base_zero)
 ## Merge the predicitons for each predictor
 predict_x1_base <- bind_rows(list(x1u_base, x1u_base_zero, x1std_base, x1std_base_zero))
 head(predict_x1_base)
-
+rownames(predict_x1_base) <- NULL
 saveVars(predict_x1_base)
 
 

### 2021-02-08 predict_emmeans.R

@@ -1,10 +1,10 @@
 library(dplyr)
 library(data.table)
+library(jdeffects)
 library(emmeans)
+library(shellpipes)
 
-source("makestuff/makeRfuns.R")
 commandEnvironments()
-sourceFiles()
 
 ## emmeans function
 empredfun <- function(mod, spec, at, cov.keep, model, vvfun=NULL){

### 2021-02-08 predict_jdeffect.R

@@ -1,19 +1,20 @@
 library(dplyr)
 library(data.table)
+library(jdeffects)
+library(shellpipes)
 
-source("makestuff/makeRfuns.R")
 commandEnvironments()
-sourceFiles()
 
 ## Varpred function function
 varpredfun <- function(mod, focal, at, vcmat = NULL, model){
 	if (is.null(vcmat)){
 		pred <- varpred(mod, focal, at = at)
 	} else {
-		pred <- varpred(mod, focal, at = at, vcmat = vcmat)
+		pred <- varpred(mod, focal, at = at, vcov. = vcmat)
 	}
+	pred <- pred$preds
 	pred$model <- model
-	pred$x <- at
+	pred$x <- unlist(at)
 	return(pred)
 }
 
@@ -22,14 +23,14 @@ varpredfun <- function(mod, focal, at, vcmat = NULL, model){
 ## Unscaled predictions
 ### Unzeroed vcov
 x1u_jd <- varpredfun(mod_unscaled, focal = "x1"
-	, at = pred_df_x1$x1
+	, at = list(x1 = pred_df_x1$x1)
 	, model = "jd_u"
 )
 head(x1u_jd)
 
 ### Zeroed non-focal vv
 x1u_jd_zero <- varpredfun(mod_unscaled, focal = "x1"
-	, at = pred_df_x1$x1
+	, at = list(x1 = pred_df_x1$x1)
 	, model = "jd_u_zero"
 	, vcmat = zero_vcov(mod_unscaled, "x1")
 )
@@ -38,14 +39,14 @@ head(x1u_jd_zero)
 ## Scaled
 ### Unzeroed vcov
 x1std_jd <- varpredfun(mod_scaled, focal = "x1std"
-	, at = pred_df_x1std$x1std,
+	, at = list(x1std = pred_df_x1std$x1std),
 	, model = "jd_s"
 )
 head(x1std_jd)
 
 ### Zeroed vcov for non-focal predictors
 x1std_jd_zero <- varpredfun(mod_scaled, focal = "x1std"
-	, at = pred_df_x1std$x1std
+	, at = list(x1std = pred_df_x1std$x1std)
 	, model = "jd_s_zero"
 	, vcmat = zero_vcov(mod_scaled, "x1std")
 )

### 2021-02-08 predict_plots.R

@@ -1,10 +1,23 @@
 library(dplyr)
 library(ggplot2); theme_set(theme_bw())
+library(shellpipes)
 
-source("makestuff/makeRfuns.R")
 commandEnvironments()
-sourceFiles()
 makeGraphics()
+
+## Some checks
+base <- (predict_x1_base
+	%>% filter(model=="base_u")
+	%>% select(-model)
+)
+emm <- (predict_x1_em
+	%>% filter(model=="em_u")
+	%>% select(-model)
+)
+jd <- (predict_x1_jd
+	%>% filter(model=="jd_u")
+	%>% select(-model)
+)
 
 # x1 marginal effects
 pred_x1 <- (bind_rows(predict_x1_base, predict_x1_em, predict_x1_jd)
@@ -13,8 +26,6 @@ pred_x1 <- (bind_rows(predict_x1_base, predict_x1_em, predict_x1_jd)
 	)
 )
 head(pred_x1)
-table(pred_x1$trans)
-table(pred_x1$model)
 
 x1_plot <- (ggplot(pred_x1, aes(x = x, group = trans))
 	+ geom_line(aes(y = fit, colour = trans), alpha = 0.7)
@@ -26,5 +37,5 @@ x1_plot <- (ggplot(pred_x1, aes(x = x, group = trans))
 	+ labs(x = "x1", y = "Predictions", fill = "Data", colour = "Data")
 	+ facet_wrap(~model, scales = "free_x")
 )
-print(x1_plot)
 
+saveEnvironment()

### 2021-02-08 sims_data.R

@@ -2,8 +2,7 @@
 
 library(dplyr)
 options(dplyr.width = Inf)
-
-source("makestuff/makeRfuns.R") ## Will eventually be a package
+library(shellpipes)
 
 set.seed(7902)
 

### 2021-02-09 effects_writeup.Rnw
Fiddling
Autocommit

@@
 
+load("predict_plots.rda")
+
+## How to pass arguments??
 #library(shellpipes)
@@
 @
+<<>>=
 
+ls()
+
+@
 <<simple_sim, echo=FALSE>>=
@@
 <<compare_baseR, echo=FALSE>>=
-load("predict_plots.rda")
 @
@@
 <<compare_baseR_plot, fig=TRUE, echo=FALSE>>=
-print(x1_plot)
+print(plot(1:10))
+## print(x1_plot)
 @

### 2021-02-13 effects_writeup.Rnw
Autocommit
Added contrast

--- 2021-02-08 effects_writeup.Rnw
@@
 
+\section{Unbalanced data}
+
+Consider a simple no-interaction terms simulation:
+\begin{itemize}
+\item $y = \beta_0 + \beta_1x_1 + \beta_2x_2 + \epsilon$ s.t $\epsilon \sim N(0, 1)$ and $\{\beta_0 = -5, \beta_1 = 0.1, \beta_2 = \{-0.6, 0.5\}\}$
+\begin{itemize}
+\item $x_1 \sim unif(1, 9)$
+\item $x_2$ -- three level categorical variable s.t $A$, $B$ and $C$ are $10\%$, $30\%$ and $60\%$, respectively.
+\end{itemize}
+\end{itemize}
+
+<<simple_sim, echo=FALSE>>=
+set.seed(101)
+N <- 500
+x1_min <- 1
+x1_max <- 9
+b0 <- -0.5
+b_x1 <- 0.1
+b_x2b <- -0.6
+b_x2c <- 0.5
+
+cat_props <- c(0.1, 0.3, 0.6)
+df <- data.frame(x1u = runif(n=N, min=x1_min, max=x1_max)
+                  , x2u = sample(c("A", "B", "C"), N, prob = cat_props, replace = TRUE)
+)
+X <- model.matrix(~x1u + x2u, df)
+betas <- c(b0, b_x1, b_x2b, b_x2c)
+df$y <- rnorm(nrow(df), mean= X %*% betas, sd=1)
+df <- transform(df
+	, x2c = x2u
+	, x1s = drop(scale(x1u, center=FALSE))
+	, x1sd = drop(scale(x1u))
+)
+prop.table(table(df$x2u))
+head(df)
+@
+
+We fit two models -- the model without contrast and the one with sum-to-contrast (\code{`contr.sum'}):
+
+<<contrast_modes>>=
+## No contrast model
+lm_no_contr <- lm(y ~ x1u + x2u, data = df)
+print(lm_no_contr)
+
+## With contrast
+contrasts(df$x2c) <- "contr.sum"
+lm_contr <- lm(y ~ x1u + x2c, data = df)
+print(lm_contr)
+@
+
+First, we include the intercept variance in the variable effects:
+
+<<simple_model, echo=FALSE>>=
+contr_models <- list(x2u = lm_no_contr, x2c = lm_contr)
+@
+
+<<contr_model_predictions, echo=FALSE, results=hide>>=
+## varpred
+contr_vpred_all <- condsummary(mod_list=contr_models, fun=function(x, f){
+	dd <- varpred(x, f)
+}, simplify = TRUE, combine = "rbind", fmethod = "varpred")
+
+## emmeans
+contr_empred_all <- condsummary(mod_list=contr_models, fun=function(x, f){
+	spec <- as.formula(paste0("~", f))
+	dd <- emmeans(x, spec=spec, cov.keep=f)
+}, simplify = TRUE, combine = "rbind", fmethod = "emmeans")
+
+## effects
+contr_efpred_all <- condsummary(mod_list=contr_models, fun=function(x, f){
+	dd <- Effect(f, x)
+}, simplify = TRUE, combine = "rbind", fmethod = "effects")
+
+## Combine all the estimates
+contr_pred_all <- do.call("rbind"
+	, list(contr_vpred_all, contr_empred_all, contr_efpred_all)
+)
+@
+
+<<echo=FALSE>>=
+class(contr_pred_all) <- c("jdeffects", "data.frame")
+contr_pred_all_plot_out <- (plot(contr_pred_all)
+	+ facet_wrap(~method)
+	+ theme(legend.position="bottom")
+)
+@
+
+\begin{figure}[H]
+\begin{center}
+<<contr_model_predictions_plot, fig=TRUE, echo=FALSE>>=
+contr_pred_all_plot_out
+@
+\end{center}
+\caption{Variable effects with intercept and other non-focal predictor uncertainties included.
+The legend compares the model with non contrast (\code{x2u}) to the one with sum to zero contrast (\code{x2c}).}
+\label{fig:contr_pred_plot}
+\end{figure}
+
+\subsection{Zeroed-out uncertainties}
+
+As explained in the previous section, we again eliminate the uncertainties due to the intercept term and the non-focal variables.
+
+\subsubsection{zeroed-out covariance matrix}
+
+<<contr_model_zeroed_vcov, echo=FALSE, results=hide>>=
+## varpred
+contr_vcov_vpred_all <- condsummary(mod_list=contr_models, fun=function(x, f){
+	dd <- varpred(x, f, vcov. = zero_vcov(x, f))
+}, simplify = TRUE, combine = "rbind", fmethod = "varpred")
+
+## emmeans
+contr_vcov_empred_all <- condsummary(mod_list=contr_models, fun=function(x, f){
+	spec <- as.formula(paste0("~", f))
+	dd <- emmeans(x, spec=spec, cov.keep=f, vcov. = zero_vcov(x, f))
+}, simplify = TRUE, combine = "rbind", fmethod = "emmeans")
+
+## effects
+contr_vcov_efpred_all <- condsummary(mod_list=contr_models, fun=function(x, f){
+	dd <- Effect(f, x, vcov. = function(x, complete=FALSE)zero_vcov(x, f, complete))
+}, simplify = TRUE, combine = "rbind", fmethod = "effects")
+
+## Combine all the estimates
+contr_vcov_pred_all <- do.call("rbind"
+	, list(contr_vcov_vpred_all, contr_vcov_empred_all, contr_vcov_efpred_all)
+)
+@
+
+<<echo=FALSE>>=
+class(contr_vcov_pred_all) <- c("jdeffects", "data.frame")
+contr_pred_zeroed_vcov_plot <- (plot(contr_vcov_pred_all)
+	+ facet_wrap(~method)
+	+ theme(legend.position="bottom")
+)
+@
+
+\begin{figure}[H]
+\begin{center}
+<<contr_zeroed_vcov_plot, fig=TRUE, echo=FALSE>>=
+contr_pred_zeroed_vcov_plot
+@
+\end{center}
+\caption{Variable effects with intercept and other non-focal predictor uncertainties included.
+The legend compares the model with non contrast (\code{x2u}) to the one with sum to zero contrast (\code{x2c}).}
+\label{fig:contr_zeroed_vcov}
+\end{figure}
+
+\subsubsection{Centered model matrix}
+
+<<contr_model_zeroed_mmat, echo=FALSE, results=hide>>=
+## varpred
+contr_mmat_vpred_all <- condsummary(mod_list=contr_models, fun=function(x, f){
+	dd <- varpred(x, f, isolate = TRUE, vcov. = zero_vcov(x, f))
+}, simplify = TRUE, combine = "rbind", fmethod = "varpred")
+
+## emmeans
+contr_mmat_empred_all <- condsummary(mod_list=contr_models, fun=function(x, f){
+	spec <- as.formula(paste0("~", f))
+	dd <- emmeans(x, spec=spec, cov.keep=f, vcov. = zero_vcov(x, f))
+}, simplify = TRUE, combine = "rbind", fmethod = "emmeans")
+
+## effects
+contr_mmat_efpred_all <- condsummary(mod_list=contr_models, fun=function(x, f){
+	dd <- Effect(f, x, vcov. = function(x, complete=FALSE)zero_vcov(x, f, complete))
+}, simplify = TRUE, combine = "rbind", fmethod = "effects")
+
+## Combine all the estimates
+contr_mmat_pred_all <- do.call("rbind"
+	, list(contr_mmat_vpred_all, contr_mmat_empred_all, contr_mmat_efpred_all)
+)
+@
+
+<<echo=FALSE>>=
+class(contr_mmat_pred_all) <- c("jdeffects", "data.frame")
+contr_pred_zeroed_mmat_plot <- (plot(contr_mmat_pred_all)
+	+ facet_wrap(~method)
+	+ theme(legend.position="bottom")
+)
+@
+
+\begin{figure}[H]
+\begin{center}
+<<contr_zeroed_mmat_plot, fig=TRUE, echo=FALSE>>=
+contr_pred_zeroed_mmat_plot
+@
+\end{center}
+\caption{Variable effects with intercept and other non-focal predictor uncertainties included.
+The legend compares the model with non contrast (\code{x2u}) to the one with sum to zero contrast (\code{x2c}).}
+\label{fig:contr_zeroed_mmat}
+\end{figure}
+
 \clearpage

### 2021-02-21 effects_writeup.Rnw
Autocommit

@@
 
-First, we include the intercept variance in the variable effects:
+\autoref{fig:contr_pred_plot} compares variable effects for the three packages when the variance of the non-focal predictors are included.
 
@@
 \end{center}
-\caption{Variable effects with intercept and other non-focal predictor uncertainties included.
-The legend compares the model with non contrast (\code{x2u}) to the one with sum to zero contrast (\code{x2c}).}
+\caption{Variable effects with non-focal predictors' uncertainties included. The legend compares the model with non contrast (\code{x2u}) to the one with sum to zero contrast (\code{x2c}).}
 \label{fig:contr_pred_plot}
@@
 
-As explained in the previous section, we again eliminate the uncertainties due to the intercept term and the non-focal variables.
+As explained in the previous section, we again eliminate the uncertainties due to non-focal predictors.
 
@@
 \end{center}
-\caption{Variable effects with intercept and other non-focal predictor uncertainties included.
-The legend compares the model with non contrast (\code{x2u}) to the one with sum to zero contrast (\code{x2c}).}
+\caption{Here, we eliminate uncertainties using zeroed-variance covariance approach.
+All the three methods give similar estimates for the model with non contrast (\code{x2u}) to the one with sum to zero contrast (\code{x2c}), respectively.
+In comparing, \code{x2u} with \code{x2c}, the uncertainties around the baseline category is completely eliminated.}
 \label{fig:contr_zeroed_vcov}
@@
 \subsubsection{Centered model matrix}
+
+Currently, only \pkg{varpred} provides a way to center the model matrix.
 
@@
 \end{center}
-\caption{Variable effects with intercept and other non-focal predictor uncertainties included.
-The legend compares the model with non contrast (\code{x2u}) to the one with sum to zero contrast (\code{x2c}).}
+\caption{We set \code{isolate = TRUE} in \pkg{varpred} to use centered model matrix in computing the CIs.
+In this case, predictions for both non contrasted (\code{x2u}) and contrasted (\code{x2c}) models are the same in \pkg{varpred}.
+Also, the CIs are slightly different with those in \autoref{fig:contr_zeroed_vcov}.
+\JD{Steve -> JD: Weighted contrast?}.}
 \label{fig:contr_zeroed_mmat}

### 2021-02-21 R/utilities.R

@@ -43,6 +43,18 @@ is.numeric.predictor <- function(predictor, model) {
   is.null(model$xlevels[[predictor]])
 }
 
+# for character and logical predictors
+
+is.factor <- function(x) inherits(x, "factor") || ((is.character(x) || is.logical(x)) && is.vector(x))
+
+levels.character <- function(x) {
+  levs <- unique(x)
+  sort(levs[!is.na(levs)])
+}
+
+levels.logical <- function(x) {
+  c("FALSE", "TRUE")
+}
 
 get_vcov <- function(mod){
 	if (inherits(mod, "glmmTMB")) {
@@ -148,7 +160,19 @@ clean_model <- function(focal.predictors, mod, xlevels = list()
 	, default.levels=NULL, formula.rhs, steps = 101, x.var=NULL, typical=mean){
   if ((!is.null(mod$nan.action)) && inherits(mod$na.action, "exclude"))
     class(mod$na.action) <- "omit"
-  all.predictors <- all.vars(formula.rhs)
+
+  terms <- attr(formula.rhs, "term.labels")
+  all.predictors <- all.vars(parse(text=terms))
+  if (is.numeric(xlevels)){
+    if (length(xlevels) > 1 || round(xlevels != xlevels)) stop("xlevels must be a single whole number or a list")
+    xlevs <- list()
+    for (pred in all.predictors){
+      xlevs[[pred]] <- xlevels
+    }
+    xlevels <- xlevs
+  }
+
+#  all.predictors <- all.vars(formula.rhs) #predictors
   check.vars <- !(focal.predictors %in% all.predictors)
   excluded.predictors <- setdiff(all.predictors, focal.predictors)
   number.bad <- sum(check.vars)
@@ -205,9 +229,11 @@ clean_model <- function(focal.predictors, mod, xlevels = list()
     levs <- xlevels
     for(name in focal.predictors) xlevels[[name]] <- levs
   }
+  oldlevels <- get_xlevels(mod)
   for (name in focal.predictors){
-    levels <- mod$xlevels[[name]]
-    if(is.null(levels)) levels <- mod$xlevels[[paste("factor(",name,")",sep="")]]
+    # levels <- mod$xlevels[[name]] # FIXME: supported models
+    levels <- oldlevels[[name]]
+    if(is.null(levels)) levels <- oldlevels[[paste("factor(",name,")",sep="")]] #FIXME: supported models
     fac <- !is.null(levels)
[…]

### 2021-02-28 effects_writeup.Rnw
Added methods for easy extension

@@
 library(emmeans)
-library(jdeffects); jdtheme()
+library(vareffects); varefftheme()
 
@@
 <<echo=FALSE>>=
-class(simple_pred_all) <- c("jdeffects", "data.frame") # plot.effects
+class(simple_pred_all) <- c("vareffects", "data.frame") # plot.effects
 simple_pred_all_plot <- (plot(simple_pred_all) 
@@
 <<echo=FALSE>>=
-class(simple_pred_spec) <- c("jdeffects", "data.frame") # plot.effects
+class(simple_pred_spec) <- c("vareffects", "data.frame") # plot.effects
 simple_pred_spec_plot <- (plot(simple_pred_spec)
@@
 <<echo=FALSE>>=
-class(simple_pred_cat) <- c("jdeffects", "data.frame") # plot.effects
+class(simple_pred_cat) <- c("vareffects", "data.frame") # plot.effects
 simple_pred_cat_plot <- (plot(simple_pred_cat)
@@
 
-The \pkg{zero\_vcov} function in \pkg{jdeffects} \emph{zeros-out} variance-covariances of non-focal terms of the predictor in question. For example:
+The \pkg{zero\_vcov} function in \pkg{vareffects} \emph{zeros-out} variance-covariances of non-focal terms of the predictor in question. For example:
 
@@
 <<echo=FALSE>>=
-class(simple_pred_all) <- c("jdeffects", "data.frame")
+class(simple_pred_all) <- c("vareffects", "data.frame")
 simple_pred_all_plot_out <- (plot(simple_pred_all)
@@
 <<label=simple_pred_spec_plot1, fig=TRUE, echo=FALSE, prefix=FALSE, include=FALSE>>=
-class(simple_pred_spec1) <- c("jdeffects", "data.frame")
+class(simple_pred_spec1) <- c("vareffects", "data.frame")
 simple_pred_spec_plot1 <- (plot(simple_pred_spec1)
@@
 <<label=simple_pred_spec_plot2, fig=TRUE, echo=FALSE, prefix=FALSE, include=FALSE>>=
-class(simple_pred_spec2) <- c("jdeffects", "data.frame")
+class(simple_pred_spec2) <- c("vareffects", "data.frame")
 simple_pred_spec_plot2 <- (plot(simple_pred_spec2)
@@
 <<echo=FALSE>>=
-class(simple_margpred_cat) <- c("jdeffects", "data.frame")
+class(simple_margpred_cat) <- c("vareffects", "data.frame")
 simple_margpred_cat_plot <- (plot(simple_margpred_cat)
@@
 <<echo=FALSE>>=
-class(contr_pred_all) <- c("jdeffects", "data.frame")
+class(contr_pred_all) <- c("vareffects", "data.frame")
 contr_pred_all_plot_out <- (plot(contr_pred_all)
@@
 <<echo=FALSE>>=
-class(contr_vcov_pred_all) <- c("jdeffects", "data.frame")
+class(contr_vcov_pred_all) <- c("vareffects", "data.frame")
 contr_pred_zeroed_vcov_plot <- (plot(contr_vcov_pred_all)
@@
 <<echo=FALSE>>=
-class(contr_mmat_pred_all) <- c("jdeffects", "data.frame")
+class(contr_mmat_pred_all) <- c("vareffects", "data.frame")
 contr_pred_zeroed_mmat_plot <- (plot(contr_mmat_pred_all)

### 2021-02-28 R/implemented.R

@@ -1,0 +1,53 @@
+# Some of the implemented methods
+
+## default lm
+vareffobj.lm <- function(mod, ...){
+	out <- list()
+	out$coefficients <- coef(mod)
+	out$variance_covariance <- vcov(mod)
+	out$formula <- formula(mod)
+	out$link <- family(mod)
+	class(out) <- "vareffobj"
+	return(out)
+}
+
+get_xlevels.lm <- function(mod) {
+	return(mod$xlevels)
+}
+
+## glmmTMB
+
+vareffobj.glmmTMB <- function(mod, ...) {
+	out <- list()
+	out$coefficients <- glmmTMB::fixef(mod)$cond
+	out$variance_covariance <- vcov(mod)$cond
+	out$formula <- formula(mod, fixed.only=TRUE)
+	out$link <- family(mod)
+	class(out) <- "vareffobj"
+	return(out)
+}
+
+get_xlevels.glmmTMB <- function(mod) {
+	xlevels <- .getXlevels(terms(mod), model.frame(mod))
+	return(xlevels)
+}
+
+## Statistics
+get_stats.glmmTMB <- function(mod, level, dfspec, ...) {
+	mulz <- qnorm(1 - (1 - level)/2)
+	return(mulz)
+}
+
+## vcov
+vcov.vareffobj <- function(x, ...)x$variance_covariance
+
+## Statistics
+get_stats.default <- function(mod, level, dfspec, ...) {
+	df <- ifelse(
+		grepl("df.residual", paste(names(mod), collapse=""))
+		, mod$df.residual, dfspec
+	)
+	mult <- qt(1 - (1 - level)/2, df)
+	return(mult)
+}
+

### 2021-02-28 R/methodfuns.R

@@ -21,8 +21,8 @@ print.varpred <- function(x, ...){
 #' @param x varpred object.
 #' @param ... additional aurguments passed to as.data.frame.
 #'
-#' @method as.data.frame varpred
 #' @export
+#' @method as.data.frame varpred
 #' @export as.data.frame.varpred
 
 as.data.frame.varpred <- function(x, ...) {
@@ -36,8 +36,8 @@ as.data.frame.varpred <- function(x, ...) {
 #' @param x varpred object.
 #' @param ... additional aurguments passed to data.frame.
 #'
-#' @method data.frame varpred
 #' @export
+#' @method data.frame varpred
 #' @export data.frame.varpred
 
 data.frame.varpred <- function(x, ...) {
@@ -45,3 +45,15 @@ data.frame.varpred <- function(x, ...) {
 	if(is.null(df)) df <- x
 	return(data.frame(df, ...))
 }
+
+#'
+#' @keywords internal
+vareffobj <- function(mod, ...)UseMethod("vareffobj")
+
+#'
+#' @keywords internal
+get_xlevels <- function(mod)UseMethod("get_xlevels")
+
+#'
+#' @keywords internal
+get_stats <- function(mod, level, dfspec, ...)UseMethod("get_stats")

### 2021-02-28 R/utilities.R

@@ -4,7 +4,7 @@ get_vnames <- function(mod){
 	mat <- model.matrix(mod)
 	coefnames <- colnames(mat)
 	Terms <- terms(mod)
-	vnames <- all.vars(delete.response(Terms)) 
+	vnames <- all.vars(parse(text=delete.response(Terms)))
 	termnames <- attr(Terms, "term.labels") # Model based names
 	termnames <- gsub(" ", "", termnames)
 	assign <- attr(mat, "assign")
@@ -54,33 +54,6 @@ levels.character <- function(x) {
 
 levels.logical <- function(x) {
   c("FALSE", "TRUE")
-}
-
-get_vcov <- function(mod){
-	if (inherits(mod, "glmmTMB")) {
-		vc <- vcov(mod)$cond
-	} else if (inherits(mod, "clmm")){
-		f <- c(names(mod$alpha)[[1]], names(mod$beta))
-		vc <- vc[f, f]
-	} else {
-		vc <- vcov(mod)
-	}
-	return(vc)
-}
-
-get_degof <- function(mod, dfspec) {
-	df <- ifelse(
-		grepl("df.residual", paste(names(mod), collapse=""))
-		, mod$df.residual, dfspec
-	)
-	return(df)
-}
-
-get_stats <- function(level, df, modfamily=NULL) {
-	if (is.null(modfamily)){
-		mult <- qt(1 - (1 - level)/2, df)
-	}
-	return(mult)
 }
 
 subscripts <- function(index, dims){
@@ -231,9 +204,8 @@ clean_model <- function(focal.predictors, mod, xlevels = list()
   }
   oldlevels <- get_xlevels(mod)
   for (name in focal.predictors){
-    # levels <- mod$xlevels[[name]] # FIXME: supported models
     levels <- oldlevels[[name]]
-    if(is.null(levels)) levels <- oldlevels[[paste("factor(",name,")",sep="")]] #FIXME: supported models
+    if(is.null(levels)) levels <- oldlevels[[paste("factor(",name,")",sep="")]]
     fac <- !is.null(levels)
 	 if (!fac) {
 		levels <- if (is.null(xlevels[[name]])){
@@ -251,7 +223,6 @@ clean_model <- function(focal.predictors, mod, xlevels = list()
 
   x.excluded <- list()
   for (name in excluded.predictors){
-    # levels <- mod$xlevels[[name]] # FIXME: supported models
     levels <- oldlevels[[name]] 
[…]

### 2021-02-28 dev/demonstrate.R

@@ -1,5 +1,5 @@
 library(dplyr)
-library(jdeffects)
+library(vareffects)
 library(shellpipes)
 
 jdtheme()
@@ -42,6 +42,6 @@ pvp_df <- (bind_rows(
 )
 head(pvp_df)
 
-### Make pvp jdeffects object and use plot method
-class(pvp_df) <- c("jdeffects", "data.frame")
+### Make pvp vareffects object and use plot method
+class(pvp_df) <- c("vareffects", "data.frame")
 plot(pvp_df)

### 2021-06-23 bias_correction.rmd
Forgot to push .rmd

---
title: "Bias correction in GLMs"
author: Bicko, Jonathan & Ben
date: "2021 Jun 21 (Mon)"
compact-title: false
output: "pdf_document"
---

```{R setup, include=FALSE, echo=FALSE}
library(dplyr)
library(ggplot2)
library(vareffects); varefftheme()
library(ggpubr)
set.seed(1011)

## Functions
plotEsize <- function(df, pos = 0.5, col_lab = ""){
	pos <- position_dodge(pos)
	p1 <- (ggplot(df, aes(x = reorder(term, -estimate), y = estimate, colour="Estimate"))
		+ geom_pointrange(aes(ymin = conf.low, ymax = conf.high), size=0.2, position = pos)
		+ geom_hline(yintercept=0,lty=2, size=0.2)
		+ coord_flip()
		+ labs(x="", y = "Estimate")
	)
}
```

\section{Introduction}

We intend to investigate our prediction based  on known truth and any bias potentially introduced by non-linear averaging, conditioning or random effect. We'll start with a simple case of a only fixed effect model and then consider a mixed effect model.

\section{Simulation}

We perform a simple simulation for a fixed effect model
\begin{align*}
\mathrm{logit(status} = 1) &= \eta \\ 
\eta &= \beta_0 + \beta_{\mathrm{A}} \mathrm{Age} + \beta_{\mathrm{W}} \mathrm{Wealthindex} \\
\mathrm{Age} &\sim \mathrm{Uniform}(0.2, 1) \\
\mathrm{Wealthindex} &\sim \mathrm{Normal}(0, 1) \\
\beta_0 &= 0.7 \\
\beta_{\mathrm{A}} &= 0.3 \\
\beta_{\mathrm{W}} &= 0.6
\end{align*}

```{R sim_simple}
N <- 10000
beta0 <- 0.7
betaA <- 0.2
betaW <- 0.5

age_max <- 1
age_min <- 0.2
age <- runif(N, age_min, age_max)

wealthindex <- rnorm(N, 0, 1)

eta <- beta0 + betaA * age + betaW * wealthindex
sim_df <- (data.frame(age=age, wealthindex=wealthindex, eta=eta)
	%>% mutate(status = rbinom(N, 1, plogis(eta)))
	%>% select(-eta)
)
true_prop <- mean(sim_df$status)
print(true_prop)
head(sim_df)
```

\subsection{Simple logistic model}

```{R simple_logistic}
simple_mod <- glm(status ~ age + wealthindex, data = sim_df, family="binomial")
```

Coefficient plots

```{R simple_coef}
## True beta
true_beta_df <- data.frame(term=c("Intercept", "age", "wealthindex")
	, estimate=c(beta0, betaA, betaW)
)

## Tidy coef estimates
coef_df <- (broom::tidy(simple_mod, conf.int=TRUE)
#	%>% dotwhisker::by_2sd(sim_df)
	%>% mutate(term = gsub("\\(|\\)", "", term))
)
print(coef_df)

simple_coef_plot <- (plotEsize(coef_df) 
	+ geom_point(data=true_beta_df, aes(x=term, y=estimate, colour="Truth"))
	+ labs(colour="Type")
	+ scale_colour_brewer(palette="Dark2")
)
print(simple_coef_plot)
```

Variable effect plots -- with and without bias adjustment

```{R simple_vareffects, fig.cap = "Variable prediction plots comparing bias unadjusted and adjusted predictions. The bias adjusted predictions are slighly lower but close to what we expect from effect size estimates (and their confidence intervals)?"}
# Age
## Not bias adjusted
simple_vareff_age <- varpred(simple_mod, "age", isolate=TRUE, modelname="Not adjusted")
print(sigma(simple_mod))

## Bias adjusted
simple_vareff_age_adjust <- varpred(simple_mod, "age", isolate=TRUE, bias.adjust=TRUE, modelname="Bias adjusted")

vareff_age <- simple_vareff_age
vareff_age$preds <- do.call("rbind", list(vareff_age$preds, simple_vareff_age_adjust$preds))
age_plot <- (plot(vareff_age)
 	+ labs(y="Prob. of improved \n service", colour="Model")
	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
	+ theme(legend.position="bottom")
)

# Wealth index
## Not bias adjusted
simple_vareff_wealthindex <- varpred(simple_mod, "wealthindex", isolate=TRUE, modelname="Not adjusted")

## Bias adjusted
simple_vareff_wealthindex_adjust <- varpred(simple_mod, "wealthindex", isolate=TRUE, bias.adjust=TRUE, modelname="Bias adjusted")

vareff_wealthindex <- simple_vareff_wealthindex
vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, simple_vareff_wealthindex_adjust$preds))
wealthindex_plot <- (plot(vareff_wealthindex)
 	+ labs(y="", colour="Model")
	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
	+ theme(legend.position="bottom")
)
ggarrange(age_plot, wealthindex_plot, common.legend=TRUE)
```

Effect sizes on logit scale:

```{R simple_logit}
coef_df_logit <- (coef_df
	%>% select(term, estimate, conf.low, conf.high)
	%>% group_by(term)
	%>% summarise_all(plogis)
)
print(coef_df_logit)
```

\subsection{MRP approach}

```{R simple_mrp}
pred <- predict(simple_mod, type="link", se.fit=TRUE)
z.val <- qnorm(1 - (1 - 0.95)/2)
pred_df <- (data.frame(age=sim_df$age, wealthindex=sim_df$wealthindex, estimate=pred$fit, se=pred$se.fit)
	%>% mutate(lwr = plogis(estimate-z.val*se)
		, upr = plogis(estimate+z.val*se)
		, estimate = plogis(estimate)
	)
)
head(pred_df)

## Age plot
age_mrp <- (pred_df
	%>% select(age, estimate, lwr, upr)
	%>% group_by(age)
	%>% summarise_all(mean)
)
age_mrp_plot <- (ggplot(age_mrp, aes(x=age, y=estimate))
	+ geom_line()
	+ geom_line(aes(y=lwr), lty=2)
	+ geom_line(aes(y=upr), lty=2)
#	+ geom_smooth(aes(ymin=lwr, ymax=upr), stat="identity")
)
print(age_mrp_plot)
```

### 2021-06-26 bias_correction.rmd
thinksync
Not clear yet!

@@
 
-```{R simple_vareffects, fig.cap = "Variable prediction plots comparing bias unadjusted and adjusted predictions.
-The bias adjusted predictions are slighly lower but close to what we expect from effect size estimates (and their confidence intervals)?"}
+```{R simple_vareffects} 
 # Age
-## Not bias adjusted
-simple_vareff_age <- varpred(simple_mod, "age", isolate=TRUE, modelname="Not adjusted")
-print(sigma(simple_mod))
-
-## Bias adjusted
-simple_vareff_age_adjust <- varpred(simple_mod, "age", isolate=TRUE, bias.adjust=TRUE, modelname="Bias adjusted")
-
-vareff_age <- simple_vareff_age
-vareff_age$preds <- do.call("rbind", list(vareff_age$preds, simple_vareff_age_adjust$preds))
-age_plot <- (plot(vareff_age)
- 	+ labs(y="Prob. of improved \n service", colour="Model")
-	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
-	+ theme(legend.position="bottom")
-)
+## varpred way
+simple_vareff_age <- varpred(simple_mod, "age", isolate=FALSE, modelname="varpred")
 
@@
 ## Not bias adjusted
-simple_vareff_wealthindex <- varpred(simple_mod, "wealthindex", isolate=TRUE, modelname="Not adjusted")
+simple_vareff_wealthindex <- varpred(simple_mod, "wealthindex", isolate=TRUE, modelname="varpred")
 
 ## Bias adjusted
-simple_vareff_wealthindex_adjust <- varpred(simple_mod, "wealthindex", isolate=TRUE, bias.adjust=TRUE, modelname="Bias adjusted")
+simple_vareff_wealthindex_adjust <- varpred(simple_mod, "wealthindex", isolate=TRUE, pop.ave=TRUE, modelname="Pop. ave")
 
@@
 )
-ggarrange(age_plot, wealthindex_plot, common.legend=TRUE)
 ```
@@
 
-\subsection{MRP approach}
+\subsection{Population averaging}
 
-```{R simple_mrp}
-pred <- predict(simple_mod, type="link", se.fit=TRUE)
-z.val <- qnorm(1 - (1 - 0.95)/2)
-pred_df <- (data.frame(age=sim_df$age, wealthindex=sim_df$wealthindex, estimate=pred$fit, se=pred$se.fit)
-	%>% mutate(lwr = plogis(estimate-z.val*se)
-		, upr = plogis(estimate+z.val*se)
-		, estimate = plogis(estimate)
+- Averages of the entire population of the non-focal predictor
+
+```{R pop_ave, fig.cap = "A comparison of population averaged and varpred-based predictions.
+For `age`, we implement the naive approach to compute the predictions in `popavefun` function and then implement the same in `vareffects` so as to use the centering machineries.
+In both cases, the population averaging and varpred gives similar estimates.
+The estimated population average is very close to the observed in the case of `wealthindex` but slightly higher in the case of `age`, see previous paragraph."}
+
+popavefun <- function(mod, focal, non.focal, level=0.95, modelname="Pop. ave", ...) {
+	mf <- model.matrix(mod)
+	mm <- (mf
+		%>% data.frame()
+		%>% mutate_at(non.focal, mean)
+		%>% as.matrix()
+	)
+	vc <- vcov(mod)
+	linpred <- as.vector(mm %*% coef(mod))
+	pse_var <- sqrt(rowSums(mm * t(tcrossprod(data.matrix(vc), mm))))
+	z.val <- qnorm(1 - (1 - level)/2)
+	pred_df <- (mf
+		%>% data.frame()
+		%>% select_at(focal)
+		%>% mutate(fit = linpred
+			, lwr = plogis(fit - z.val*pse_var)
+			, upr = plogis(fit + z.val*pse_var)
+			, fit = plogis(fit)
+			, model = modelname
+			, se = NA
 		)
 	)
-head(pred_df)
+	return(pred_df)
+}
 
-## Age plot
-age_mrp <- (pred_df
-	%>% select(age, estimate, lwr, upr)
-	%>% group_by(age)
-	%>% summarise_all(mean)
+# simple_vareff_age_pop <- varpred(simple_mod, "age", isolate=TRUE, modelname="Pop. ave")
+simple_vareff_age_pop <- popavefun(simple_mod, "age", "wealthindex", modelname = "Pop. ave")
+
+vareff_age <- simple_vareff_age
+vareff_age$preds <- do.call("rbind", list(vareff_age$preds, simple_vareff_age_pop))
+age_plot <- (plot(vareff_age)
+ 	+ labs(y="Prob. of improved \n service", colour="Model")
+	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
+	+ theme(legend.position="bottom")
 )
-age_mrp_plot <- (ggplot(age_mrp, aes(x=age, y=estimate))
-	+ geom_line()
-	+ geom_line(aes(y=lwr), lty=2)
-	+ geom_line(aes(y=upr), lty=2)
-#	+ geom_smooth(aes(ymin=lwr, ymax=upr), stat="identity")
-)
-print(age_mrp_plot)
+
+ggarrange(age_plot, wealthindex_plot, common.legend=TRUE)
 ```
+
+The observed population average is `r true_prop` while the estimated population averages are:
+
+- age: `r mean(simple_vareff_age_pop$fit)`
+- wealthindex: `r mean(simple_vareff_wealthindex_adjust$preds$fit)`

### 2021-06-26 R/utilities.R

@@ -130,8 +130,7 @@ matrix.to.df <- function(matrix, colclasses){
 }
 
 clean_model <- function(focal.predictors, mod, xlevels = list()
-	, default.levels=NULL, formula.rhs, steps = 101, x.var=NULL, typical=mean
-	, vnames, which.interaction){
+	, default.levels=NULL, formula.rhs, steps = 101, x.var=NULL, typical=mean, vnames, which.interaction, pop.ave=FALSE){
   if ((!is.null(mod$nan.action)) && inherits(mod$na.action, "exclude"))
     class(mod$na.action) <- "omit"
 
@@ -231,6 +230,9 @@ clean_model <- function(focal.predictors, mod, xlevels = list()
 		}
 	 }
 	 else factor.levels[[name]] <- levels
+	 if (pop.ave) {
+	 	levels <- as.vector(X[[name]])
+	 }
 	 x[[name]] <- list(name=name, is.factor=is.factor(X[, name]), levels=levels)
   }
 

### 2021-06-30 bias_correction.rmd
Simple mixed effect model bias adjustment seems to work

@@
 ```{R setup, include=FALSE, echo=FALSE}
+library(glmmTMB)
 library(dplyr)
@@
 ```{R sim_simple}
-N <- 10000
+N <- 1e5
 beta0 <- 0.7
@@
 	+ labs(colour="Type")
-	+ scale_colour_brewer(palette="Dark2")
+	+ scale_colour_manual(values=c("black", "blue"))
 )
@@
 
-Variable effect plots -- with and without bias adjustment
+Variable effect plots -- varpred and population averaging approach
 
@@
 	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
+	+ scale_colour_manual(values=c("black", "blue"))
 	+ theme(legend.position="bottom")
@@
 - wealthindex: `r mean(simple_vareff_wealthindex_adjust$preds$fit)`
+
+\section{Random effect model}
+
+```{r sim_randeff}
+# Simulation parameters
+nHH <- 1000		# Number of HH (primary units) per year
+
+nyrs <- 10	# Number of years to simulate
+yrs <- 2000 + c(1:nyrs) # Years to simulate
+N <- nyrs * nHH
+
+## HH random effect sd
+hhSD <- 0.5
+
+# Generate dataset template
+temp_df <- (data.frame(hhid = rep(c(1:nHH), each = nyrs)
+		, years = rep(yrs, nHH)
+		, age = runif(n=N, age_min, age_max)
+		, wealthindex = rnorm(n = N, 0, 1)
+	)
+)
+
+# Simulate HH-level random effects (residual error)
+hhRE <- rnorm(nHH, hhSD)
+temp_df$hhRE <- hhRE[temp_df$hhid]
+
+sim_df <- (temp_df
+	%>% mutate(eta = beta0 + betaA * age + betaW * wealthindex + hhRE
+		, status = rbinom(N, 1, plogis(eta))
+	)
+	%>% select(-eta)
+)
+true_prop_reff <- mean(sim_df$status)
+print(true_prop)
+print(head(sim_df, 50))
+```
+
+\subsection{Fit model}
+
+```{R model_reff}
+reff_mod <- glmmTMB(status ~ age + wealthindex + (1|hhid)
+	, data = sim_df
+	, family = binomial(link = "logit")
+)
+```
+
+```{R reff_coef}
+## Tidy coef estimates
+reff_coef_df <- (broom.mixed::tidy(reff_mod, conf.int=TRUE)
+	%>% mutate(term = gsub("\\(|\\)", "", term))
+	%>% filter(effect=="fixed")
+)
+print(reff_coef_df)
+
+reff_coef_plot <- (plotEsize(reff_coef_df) 
+	+ geom_point(data=true_beta_df, aes(x=term, y=estimate, colour="Truth"))
+	+ labs(colour="Type")
+	+ scale_colour_manual(values=c("black", "blue"))
+)
+print(reff_coef_plot)
+```
+
+Variable effect plots
+
+- Age
+
+```{R reff_vareffects_age} 
+## varpred way
+reff_vareff_age <- varpred(reff_mod, "age", isolate=FALSE, modelname="varpred")
+
+## Pop. average
+reff_vareff_age_pop <- varpred(reff_mod, "age", isolate=TRUE, pop.ave=TRUE, modelname = "Pop. ave")
+
+## Bias adjust
+reff_vareff_age_adjust <- varpred(reff_mod, "age", isolate=TRUE, bias.adjust=TRUE, modelname = "Bias adjust")
+
+vareff_age <- reff_vareff_age
+vareff_age$preds <- do.call("rbind", list(vareff_age$preds, reff_vareff_age_pop$preds, reff_vareff_age_adjust$preds))
+age_plot <- (plot(vareff_age)
+ 	+ labs(y="Prob. of improved \n service", colour="Model")
+	+ geom_hline(yintercept=true_prop_reff, lty=2, colour="grey")
+	+ scale_colour_manual(values=c("black", "blue", "red"))
+	+ theme(legend.position="bottom")
+)
+
+```
+
+- Wealth index
+
+```{R reff_vareff_wealthindex, fig.cap = "A comparison of population averaged, varpred-based and bias-adjusted predictions.
+For each of the predictors, the population averaging and varpred gives similar estimates, and slightly off the truth.
+However, when we apply bias-adjustment, the estimates are very precise (i.e., close to the truth)."}
+# Wealth index
+## varpred 
+reff_vareff_wealthindex <- varpred(reff_mod, "wealthindex", isolate=TRUE, modelname="varpred")
+
+## Pop. average
+reff_vareff_wealthindex_pop <- varpred(reff_mod, "wealthindex", isolate=TRUE, pop.ave=TRUE, modelname="Pop. ave")
+
+## Bias adjust
+reff_vareff_wealthindex_adjust <- varpred(reff_mod, "wealthindex", isolate=TRUE, bias.adjust=TRUE, modelname="Bias adjust")
+
+vareff_wealthindex <- reff_vareff_wealthindex
+vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, reff_vareff_wealthindex_pop$preds, reff_vareff_wealthindex_adjust$preds))
+wealthindex_plot <- (plot(vareff_wealthindex)
+ 	+ labs(y="", colour="Model")
+	+ geom_hline(yintercept=true_prop_reff, lty=2, colour="grey")
+	+ scale_colour_manual(values=c("black", "blue", "red"))
+	+ theme(legend.position="bottom")
+)
+
+ggarrange(age_plot, wealthindex_plot, common.legend=TRUE)
+```

### 2021-07-06 bias_correction.rmd
Bias-adjustment for simple simulation

@@
 compact-title: false
-output: "pdf_document"
+output:
+  pdf_document:
+    extra_dependencies: ["float"]
 ---
 
-```{R setup, include=FALSE, echo=FALSE}
+```{r setup, include=FALSE, echo=FALSE}
+knitr::opts_chunk$set(fig.pos = "H", out.extra = "")
+
 library(glmmTMB)
@@
 
-```{R sim_simple}
+```{r sim_simple}
 N <- 1e5
@@
 
-```{R simple_logistic}
+```{r simple_logistic}
 simple_mod <- glm(status ~ age + wealthindex, data = sim_df, family="binomial")
@@
 
-```{R simple_coef}
+```{r simple_coef}
 ## True beta
@@
 
-```{R simple_vareffects} 
+```{r simple_vareffects} 
 # Age
@@
 # Wealth index
-## Not bias adjusted
+## varpred - no bias adjustment
 simple_vareff_wealthindex <- varpred(simple_mod, "wealthindex", isolate=TRUE, modelname="varpred")
@@
 
-```{R simple_logit}
+```{r simple_logit}
 coef_df_logit <- (coef_df
@@
 
-\subsection{Population averaging}
+\subsection{Variable predictions}
 
-- Averages of the entire population of the non-focal predictor
+We consider both \textbf{varpred} and population averaging approach; and then introduce bias correction to \textbf{varpred} predictions.
 
-```{R pop_ave, fig.cap = "A comparison of population averaged and varpred-based predictions.
-For `age`, we implement the naive approach to compute the predictions in `popavefun` function and then implement the same in `vareffects` so as to use the centering machineries.
-In both cases, the population averaging and varpred gives similar estimates.
-The estimated population average is very close to the observed in the case of `wealthindex` but slightly higher in the case of `age`, see previous paragraph."}
-
+```{r pop_ave_fun}
 popavefun <- function(mod, focal, non.focal, level=0.95, modelname="Pop. ave", ...) {
@@
 
-# simple_vareff_age_pop <- varpred(simple_mod, "age", isolate=TRUE, modelname="Pop. ave")
+```
+
+- Age
+
+```{r simple_vareffects_age} 
+## varpred way
+simple_vareff_age <- varpred(simple_mod, "age", isolate=FALSE, modelname="varpred")
+
+## Pop. average
 simple_vareff_age_pop <- popavefun(simple_mod, "age", "wealthindex", modelname = "Pop. ave")
 
+## Bias adjust
+simple_vareff_age_adjust <- varpred(simple_mod, "age", isolate=TRUE, bias.adjust=TRUE, modelname = "Bias adjust")
+
 vareff_age <- simple_vareff_age
-vareff_age$preds <- do.call("rbind", list(vareff_age$preds, simple_vareff_age_pop))
+vareff_age$preds <- do.call("rbind", list(vareff_age$preds, simple_vareff_age_pop, simple_vareff_age_adjust$preds))
 age_plot <- (plot(vareff_age)
@@
 	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
-	+ scale_colour_manual(values=c("black", "blue"))
+	+ scale_colour_manual(values=c("black", "blue", "red"))
+	+ theme(legend.position="bottom")
+)
+```
+
+- Wealth index
+
+```{r simple_vareff_wealthindex, fig.cap = "A comparison of population averaged and varpred-based predictions.
+We also apply bias-adjustemnt to the `vapred` predictions.
+For `age`, we implement the naive approach to compute the predictions in `popavefun` function and then implement the same in `vareffects` so as to use the centering machineries.
+In both cases, the population averaging and varpred gives similar estimates.
+The estimated population average is very close to the observed in the case of `wealthindex` but slightly higher in the case of `age` for the unadjusted estimates (but very low for bias-adjusted).
+, see previous paragraph."}
+# Wealth index
+## varpred 
+simple_vareff_wealthindex <- varpred(simple_mod, "wealthindex", isolate=TRUE, modelname="varpred")
+
+## Pop. average
+simple_vareff_wealthindex_pop <- varpred(simple_mod, "wealthindex", isolate=TRUE, pop.ave=TRUE, modelname="Pop. ave")
+
+## Bias adjust
+simple_vareff_wealthindex_adjust <- varpred(simple_mod, "wealthindex", isolate=TRUE, bias.adjust=TRUE, modelname="Bias adjust")
+
+vareff_wealthindex <- simple_vareff_wealthindex
+vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, simple_vareff_wealthindex_pop$preds, simple_vareff_wealthindex_adjust$preds))
+wealthindex_plot <- (plot(vareff_wealthindex)
+ 	+ labs(y="", colour="Model")
+	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
+	+ scale_colour_manual(values=c("black", "blue", "red"))
 	+ theme(legend.position="bottom")
@@
 
-The observed population average is `r true_prop` while the estimated population averages are:
+The observed population average is `r true_prop` while the estimated population averages (similar estimates with `varpred`) are:
 
-- age: `r mean(simple_vareff_age_pop$fit)`
-- wealthindex: `r mean(simple_vareff_wealthindex_adjust$preds$fit)`
+- age: unadjusted `r mean(simple_vareff_age_pop$fit)`; adjusted `r mean(simple_vareff_age_adjust$preds$fit)`
+- wealthindex: unadjusted `r mean(simple_vareff_wealthindex_pop$preds$fit)`; adjusted `r mean(simple_vareff_wealthindex_adjust$preds$fit)`
 
@@
 
-```{R model_reff}
+```{r model_reff}
 reff_mod <- glmmTMB(status ~ age + wealthindex + (1|hhid)
@@
 
-```{R reff_coef}
+```{r reff_coef}
 ## Tidy coef estimates
@@
 
-```{R reff_vareffects_age} 
+```{r reff_vareffects_age} 
 ## varpred way
@@
 
-```{R reff_vareff_wealthindex, fig.cap = "A comparison of population averaged, varpred-based and bias-adjusted predictions.
+```{r reff_vareff_wealthindex, fig.cap = "A comparison of population averaged, varpred-based and bias-adjusted predictions.
 For each of the predictors, the population averaging and varpred gives similar estimates, and slightly off the truth.

### 2021-07-15 bias_correction.rmd
Normal for age
Population averaging seems to work

@@
 \eta &= \beta_0 + \beta_{\mathrm{A}} \mathrm{Age} + \beta_{\mathrm{W}} \mathrm{Wealthindex} \\
-\mathrm{Age} &\sim \mathrm{Uniform}(0.2, 1) \\
+\mathrm{Age} &\sim \mathrm{Normal}(0.2, 1) \\
 \mathrm{Wealthindex} &\sim \mathrm{Normal}(0, 1) \\
@@
 age_min <- 0.2
-age <- runif(N, age_min, age_max)
+# age <- runif(N, age_min, age_max)
+age <- rnorm(N, age_max, age_max)
 
@@
 
-Variable effect plots -- varpred and population averaging approach
-
-```{r simple_vareffects} 
-# Age
-## varpred way
-simple_vareff_age <- varpred(simple_mod, "age", isolate=FALSE, modelname="varpred")
-
-# Wealth index
-## varpred - no bias adjustment
-simple_vareff_wealthindex <- varpred(simple_mod, "wealthindex", isolate=TRUE, modelname="varpred")
-
-## Bias adjusted
-simple_vareff_wealthindex_adjust <- varpred(simple_mod, "wealthindex", isolate=TRUE, pop.ave=TRUE, modelname="Pop. ave")
-
-vareff_wealthindex <- simple_vareff_wealthindex
-vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, simple_vareff_wealthindex_adjust$preds))
-wealthindex_plot <- (plot(vareff_wealthindex)
- 	+ labs(y="", colour="Model")
-	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
-	+ scale_colour_manual(values=c("black", "blue"))
-	+ theme(legend.position="bottom")
-)
-```
-
 Effect sizes on logit scale:
@@
 ```{r pop_ave_fun}
-popavefun <- function(mod, focal, non.focal, level=0.95, modelname="Pop. ave", ...) {
-	mf <- model.matrix(mod)
+popavefun <- function(mod, focal, non.focal, level=0.95, steps=100, bins=50, modelname="Pop. ave", ...) {
+	mf <- model.frame(mod)
 	mm <- (mf
-		%>% data.frame()
-		%>% mutate_at(non.focal, mean)
-		%>% as.matrix()
+		%>% select_at(c(focal, non.focal))
 	)
+	check_df <- (mf
+		%>% arrange_at(focal)
+		%>% mutate(bin=ceiling(row_number()*bins/nrow(.)))
+		%>% group_by(bin)
+		%>% summarise_all(mean)
+		%>% mutate(model="binned")
+	)
+	quant <- seq(0, 1, length.out=steps)
+	mm <- sapply(c(focal, non.focal), function(x)quantile(mm[,x], quant), simplify = FALSE)
+	mm <- do.call("expand.grid", mm)
 	vc <- vcov(mod)
-	linpred <- as.vector(mm %*% coef(mod))
-	pse_var <- sqrt(rowSums(mm * t(tcrossprod(data.matrix(vc), mm))))
+	mm2 <- model.matrix(formula(mod)[c(1,3)], mm)
+	linpred <- as.vector(mm2 %*% coef(mod))
+	pse_var <- sqrt(rowSums(mm2 * t(tcrossprod(data.matrix(vc), mm2))))
 	z.val <- qnorm(1 - (1 - level)/2)
-	pred_df <- (mf
-		%>% data.frame()
+	pred_df <- (mm
 		%>% select_at(focal)
@@
 			, fit = plogis(fit)
-			, model = modelname
-			, se = NA
 		)
+		%>% group_by_at(focal)
+		%>% summarise_at(c("fit", "lwr", "upr"), mean)
+		%>% mutate(model = modelname, se=NA)
 	)
-	return(pred_df)
+	return(list(preds=pred_df, check=check_df))
 }
@@
 simple_vareff_age_pop <- popavefun(simple_mod, "age", "wealthindex", modelname = "Pop. ave") 
-
-## Bias adjust
-simple_vareff_age_adjust <- varpred(simple_mod, "age", isolate=TRUE, bias.adjust=TRUE, modelname = "Bias adjust")
+binned_df <- simple_vareff_age_pop$check
 
 vareff_age <- simple_vareff_age
-vareff_age$preds <- do.call("rbind", list(vareff_age$preds, simple_vareff_age_pop, simple_vareff_age_adjust$preds))
+vareff_age$preds <- do.call("rbind", list(vareff_age$preds, simple_vareff_age_pop$preds))
 age_plot <- (plot(vareff_age)
@@
 	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
+	+ geom_point(data=binned_df, aes(x=age, y=status, color="binned"))
 	+ scale_colour_manual(values=c("black", "blue", "red"))
@@
 ## varpred 
-simple_vareff_wealthindex <- varpred(simple_mod, "wealthindex", isolate=TRUE, modelname="varpred")
+simple_vareff_wealthindex <- varpred(simple_mod, "wealthindex", isolate=FALSE, modelname="varpred")
 
 ## Pop. average
-simple_vareff_wealthindex_pop <- varpred(simple_mod, "wealthindex", isolate=TRUE, pop.ave=TRUE, modelname="Pop. ave")
-
-## Bias adjust
-simple_vareff_wealthindex_adjust <- varpred(simple_mod, "wealthindex", isolate=TRUE, bias.adjust=TRUE, modelname="Bias adjust")
+simple_vareff_wealthindex_pop <- popavefun(simple_mod, "wealthindex", "age", modelname="Pop. ave")
+binned_df <- simple_vareff_wealthindex_pop$check
 
 vareff_wealthindex <- simple_vareff_wealthindex
-vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, simple_vareff_wealthindex_pop$preds, simple_vareff_wealthindex_adjust$preds))
+vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, simple_vareff_wealthindex_pop$preds))
 wealthindex_plot <- (plot(vareff_wealthindex)
@@
 	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
+	+ geom_point(data=binned_df, aes(x=wealthindex, y=status, color="binned"))
 	+ scale_colour_manual(values=c("black", "blue", "red"))
@@
 
-- age: unadjusted `r mean(simple_vareff_age_pop$fit)`; adjusted `r mean(simple_vareff_age_adjust$preds$fit)`
-- wealthindex: unadjusted `r mean(simple_vareff_wealthindex_pop$preds$fit)`; adjusted `r mean(simple_vareff_wealthindex_adjust$preds$fit)`
+- age: pop. average `r mean(simple_vareff_age_pop$preds$fit)`; varpred `r mean(simple_vareff_age$preds$fit)`
+- wealthindex: pop. average `r mean(simple_vareff_wealthindex_pop$preds$fit)`; varpred `r mean(simple_vareff_wealthindex$preds$fit)`
 

### 2021-07-20 bias_correction.rmd
Population averaging seems to work in a simple case

@@
 ```{r sim_simple}
-N <- 1e5
+N <- 1e4
 beta0 <- 0.7
@@
 age_min <- 0.2
-# age <- runif(N, age_min, age_max)
-age <- rnorm(N, age_max, age_max)
+age <- runif(N, age_min, age_max)
+# age <- rnorm(N, age_max, age_max)
 
@@
 ## Pop. average
-simple_vareff_age_pop <- popavefun(simple_mod, "age", "wealthindex", modelname = "Pop. ave") 
-binned_df <- simple_vareff_age_pop$check
+simple_vareff_age_pop <- varpred(simple_mod, "age", isolate=FALSE, pop.ave=TRUE, modelname="Pop. ave")
+# simple_vareff_age_pop <- popavefun(simple_mod, "age", "wealthindex", modelname = "Pop. ave") 
+# binned_df <- simple_vareff_age_pop$check
 
@@
 	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
-	+ geom_point(data=binned_df, aes(x=age, y=status, color="binned"))
-	+ scale_colour_manual(values=c("black", "blue", "red"))
+	+ scale_colour_manual(values=c("black", "blue", "red", "green"))
 	+ theme(legend.position="bottom")
@@
 ```{r simple_vareff_wealthindex, fig.cap = "A comparison of population averaged and varpred-based predictions.
-We also apply bias-adjustemnt to the `vapred` predictions.
-For `age`, we implement the naive approach to compute the predictions in `popavefun` function and then implement the same in `vareffects` so as to use the centering machineries.
-In both cases, the population averaging and varpred gives similar estimates.
-The estimated population average is very close to the observed in the case of `wealthindex` but slightly higher in the case of `age` for the unadjusted estimates (but very low for bias-adjusted).
-, see previous paragraph."}
+In the case of biased predictions (age), population averaging gives better estimates.
+On the other hand, for wealthindex, the estimates varpred and population estimates are somehow similar."}
 # Wealth index
@@
 ## Pop. average
-simple_vareff_wealthindex_pop <- popavefun(simple_mod, "wealthindex", "age", modelname="Pop. ave")
-binned_df <- simple_vareff_wealthindex_pop$check
+simple_vareff_wealthindex_pop <- varpred(simple_mod, "wealthindex", isolate=FALSE, pop.ave=TRUE, modelname="Pop. ave")
+# binned_df <- simple_vareff_wealthindex_pop$check
 
@@
 	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
-	+ geom_point(data=binned_df, aes(x=wealthindex, y=status, color="binned"))
+#	+ geom_point(data=binned_df, aes(x=wealthindex, y=status, color="binned"))
 	+ scale_colour_manual(values=c("black", "blue", "red"))
@@
 - wealthindex: pop. average `r mean(simple_vareff_wealthindex_pop$preds$fit)`; varpred `r mean(simple_vareff_wealthindex$preds$fit)`
-
-\section{Random effect model}
-
-```{r sim_randeff}
-# Simulation parameters
-nHH <- 1000		# Number of HH (primary units) per year
-
-nyrs <- 10	# Number of years to simulate
-yrs <- 2000 + c(1:nyrs) # Years to simulate
-N <- nyrs * nHH
-
-## HH random effect sd
-hhSD <- 0.5
-
-# Generate dataset template
-temp_df <- (data.frame(hhid = rep(c(1:nHH), each = nyrs)
-		, years = rep(yrs, nHH)
-		, age = runif(n=N, age_min, age_max)
-		, wealthindex = rnorm(n = N, 0, 1)
-	)
-)
-
-# Simulate HH-level random effects (residual error)
-hhRE <- rnorm(nHH, hhSD)
-temp_df$hhRE <- hhRE[temp_df$hhid]
-
-sim_df <- (temp_df
-	%>% mutate(eta = beta0 + betaA * age + betaW * wealthindex + hhRE
-		, status = rbinom(N, 1, plogis(eta))
-	)
-	%>% select(-eta)
-)
-true_prop_reff <- mean(sim_df$status)
-print(true_prop)
-print(head(sim_df, 50))
-```
-
-\subsection{Fit model}
-
-```{r model_reff}
-reff_mod <- glmmTMB(status ~ age + wealthindex + (1|hhid)
-	, data = sim_df
-	, family = binomial(link = "logit")
-)
-```
-
-```{r reff_coef}
-## Tidy coef estimates
-reff_coef_df <- (broom.mixed::tidy(reff_mod, conf.int=TRUE)
-	%>% mutate(term = gsub("\\(|\\)", "", term))
-	%>% filter(effect=="fixed")
-)
-print(reff_coef_df)
-
-reff_coef_plot <- (plotEsize(reff_coef_df) 
-	+ geom_point(data=true_beta_df, aes(x=term, y=estimate, colour="Truth"))
-	+ labs(colour="Type")
-	+ scale_colour_manual(values=c("black", "blue"))
-)
-print(reff_coef_plot)
-```
-
-Variable effect plots
-
-- Age
-
-```{r reff_vareffects_age} 
-## varpred way
-reff_vareff_age <- varpred(reff_mod, "age", isolate=FALSE, modelname="varpred")
-
-## Pop. average
-reff_vareff_age_pop <- varpred(reff_mod, "age", isolate=TRUE, pop.ave=TRUE, modelname = "Pop. ave")
-
-## Bias adjust
-reff_vareff_age_adjust <- varpred(reff_mod, "age", isolate=TRUE, bias.adjust=TRUE, modelname = "Bias adjust")
-
-vareff_age <- reff_vareff_age
-vareff_age$preds <- do.call("rbind", list(vareff_age$preds, reff_vareff_age_pop$preds, reff_vareff_age_adjust$preds))
-age_plot <- (plot(vareff_age)
- 	+ labs(y="Prob. of improved \n service", colour="Model")
-	+ geom_hline(yintercept=true_prop_reff, lty=2, colour="grey")
-	+ scale_colour_manual(values=c("black", "blue", "red"))
-	+ theme(legend.position="bottom")
-)
-
-```
-
-- Wealth index
-
-```{r reff_vareff_wealthindex, fig.cap = "A comparison of population averaged, varpred-based and bias-adjusted predictions.
-For each of the predictors, the population averaging and varpred gives similar estimates, and slightly off the truth.
-However, when we apply bias-adjustment, the estimates are very precise (i.e., close to the truth)."}
-# Wealth index
-## varpred 
-reff_vareff_wealthindex <- varpred(reff_mod, "wealthindex", isolate=TRUE, modelname="varpred")
-
-## Pop. average
-reff_vareff_wealthindex_pop <- varpred(reff_mod, "wealthindex", isolate=TRUE, pop.ave=TRUE, modelname="Pop. ave")
-
-## Bias adjust
-reff_vareff_wealthindex_adjust <- varpred(reff_mod, "wealthindex", isolate=TRUE, bias.adjust=TRUE, modelname="Bias adjust")
-
-vareff_wealthindex <- reff_vareff_wealthindex
-vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, reff_vareff_wealthindex_pop$preds, reff_vareff_wealthindex_adjust$preds))
-wealthindex_plot <- (plot(vareff_wealthindex)
- 	+ labs(y="", colour="Model")
-	+ geom_hline(yintercept=true_prop_reff, lty=2, colour="grey")
-	+ scale_colour_manual(values=c("black", "blue", "red"))
-	+ theme(legend.position="bottom")
-)
-
-ggarrange(age_plot, wealthindex_plot, common.legend=TRUE)
-```

### 2021-07-20 R/effectsfuns.R

@@ -130,38 +130,9 @@ varpred <- function(mod, focal_predictors, x.var = NULL
 	X <- model_frame_objs$X
 	x.var <- model_frame_objs$x.var
 	mf <- model.frame(rTerms, predict.data, xlev = factor.levels, na.action=NULL)
-	mod.matrix <- model.matrix(formula.rhs, data = mf, contrasts.arg = vareff_objects$contrasts)
-	mod.matrix.all <- model.matrix(mod)
 	typical <- avefun
-
-	mm <- get_model_matrix(mod, mod.matrix, mod.matrix.all, X.mod
-		, factor.cols, cnames, focal.predictors, excluded.predictors
-		, typical, apply.typical.to.factors = TRUE
-	)
+	mod.matrix.all <- model.matrix(mod)
 	
-	if (which.interaction=="emmeans") {
-		ff_cols <- factor.cols[factor.cols==TRUE]
-		if (length(focal.predictors)>1L) {
-			ncats <- 0
-			for (f in focal.predictors) {
-				name <- names(grep(f, vnames, value=TRUE))
-				if (!any(grepl("\\:", name)) & length(name)>1L) {
-					ncats <- ncats + length(name)
-				}	
-			}
-			ncats <- ifelse(ncats==0, 1, ncats)
-			ff_cols <- factor.cols[factor.cols==TRUE]
-			for (name in names(ff_cols)) {
-				components <- unlist(strsplit(name, ':'))
-				component <- components[1]
-				if (length(components) > 1) {
-					if (any(vnames[names(vnames) %in% name] %in% focal.predictors)) ncats <- 1
-	  				mm[,name] <- apply(mm[,c(component, name)], 1, prod) * ncats
-				}
-			}
-		}
-	}
-
 	if (is.null(x.var) & n.focal>1L) {
 		x.var <- focal.predictors[[2]]
 		message(paste0("x.var was not specified, ", x.var, " is used instead."))
@@ -169,6 +140,46 @@ varpred <- function(mod, focal_predictors, x.var = NULL
 		x.var <- focal.predictors[[1]]
 	}
 	
+	if (pop.ave) {
+		mf <- mf[, c(x.var, colnames(mf)[!colnames(mf) %in% x.var])]
+		mf <- do.call("expand.grid", mf)
+		mod.matrix <- model.matrix(formula.rhs, data = mf, contrasts.arg = vareff_objects$contrasts)
[…]

### 2021-07-21 bias_correction.rmd
Population averaging in mixed models

@@
 - wealthindex: pop. average `r mean(simple_vareff_wealthindex_pop$preds$fit)`; varpred `r mean(simple_vareff_wealthindex$preds$fit)`
+
+\section{Random effect model}
+
+```{r sim_randeff}
+# Simulation parameters
+nHH <- 1000	# Number of HH (primary units) per year
+
+nyrs <- 50	# Number of years to simulate
+yrs <- 2000 + c(1:nyrs) # Years to simulate
+N <- nyrs * nHH
+
+## HH random effect sd
+hhSD <- 0.5
+
+# Generate dataset template
+temp_df <- (data.frame(hhid = rep(c(1:nHH), each = nyrs)
+		, years = rep(yrs, nHH)
+		, age = runif(n=N, age_min, age_max)
+		, wealthindex = rnorm(n = N, 0, 1)
+	)
+)
+
+# Simulate HH-level random effects (residual error)
+hhRE <- rnorm(nHH, hhSD)
+temp_df$hhRE <- hhRE[temp_df$hhid]
+
+sim_df <- (temp_df
+	%>% mutate(eta = beta0 + betaA * age + betaW * wealthindex + hhRE
+		, status = rbinom(N, 1, plogis(eta))
+	)
+	%>% select(-eta)
+)
+true_prop_reff <- mean(sim_df$status)
+print(true_prop)
+print(head(sim_df, 10))
+```
+
+\subsection{Fit model}
+
+```{r model_reff}
+reff_mod <- glmmTMB(status ~ age + wealthindex + (1|hhid)
+	, data = sim_df
+	, family = binomial(link = "logit")
+)
+```
+
+```{r reff_coef}
+## Tidy coef estimates
+reff_coef_df <- (broom.mixed::tidy(reff_mod, conf.int=TRUE)
+	%>% mutate(term = gsub("\\(|\\)", "", term))
+	%>% filter(effect=="fixed")
+)
+print(reff_coef_df)
+
+reff_coef_plot <- (plotEsize(reff_coef_df) 
+	+ geom_point(data=true_beta_df, aes(x=term, y=estimate, colour="Truth"))
+	+ labs(colour="Type")
+	+ scale_colour_manual(values=c("black", "blue"))
+)
+print(reff_coef_plot)
+```
+
+Variable effect plots
+
+- Age
+
+```{r reff_vareffects_age}
+## varpred way
+reff_vareff_age <- varpred(reff_mod, "age", isolate=FALSE, modelname="varpred")
+
+## Pop. average
+reff_vareff_age_pop <- varpred(reff_mod, "age", isolate=FALSE, pop.ave=TRUE, modelname = "Pop. ave")
+
+vareff_age <- reff_vareff_age
+vareff_age$preds <- do.call("rbind", list(vareff_age$preds, reff_vareff_age_pop$preds))
+age_plot <- (plot(vareff_age)
+ 	+ labs(y="Prob. of improved \n service", colour="Model")
+	+ geom_hline(yintercept=true_prop_reff, lty=2, colour="grey")
+	+ scale_colour_manual(values=c("black", "blue", "red"))
+	+ theme(legend.position="bottom")
+)
+```
+
+- Wealth index
+
+```{r reff_vareff_wealthindex, fig.cap = "A comparison of population average and varpred based predictions. In the case of age, both methods seems to over-approximate the predictions."} 
+# Wealth index
+## varpred
+reff_vareff_wealthindex <- varpred(reff_mod, "wealthindex", isolate=FALSE, modelname="varpred")
+
+## Pop. average
+reff_vareff_wealthindex_pop <- varpred(reff_mod, "wealthindex", isolate=FALSE, pop.ave=TRUE, modelname="Pop. ave")
+
+vareff_wealthindex <- reff_vareff_wealthindex
+vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, reff_vareff_wealthindex_pop$preds))
+wealthindex_plot <- (plot(vareff_wealthindex)
+ 	+ labs(y="", colour="Model")
+	+ geom_hline(yintercept=true_prop_reff, lty=2, colour="grey")
+	+ scale_colour_manual(values=c("black", "blue", "red"))
+	+ theme(legend.position="bottom")
+)
+
+ggarrange(age_plot, wealthindex_plot, common.legend=TRUE)
+```

### 2021-07-28 bias_correction.rmd
Some progess with quantiles

@@
 ```{r pop_ave_fun}
-popavefun <- function(mod, focal, non.focal, level=0.95, steps=100, bins=50, modelname="Pop. ave", ...) {
+popavefun <- function(mod, focal, non.focal, level=0.95, steps=100, modelname="quantiles", ...) {
 	mf <- model.frame(mod)
@@
 		%>% select_at(c(focal, non.focal))
-	)
-	check_df <- (mf
-		%>% arrange_at(focal)
-		%>% mutate(bin=ceiling(row_number()*bins/nrow(.)))
-		%>% group_by(bin)
-		%>% summarise_all(mean)
-		%>% mutate(model="binned")
 	)
@@
 	)
-	return(list(preds=pred_df, check=check_df))
+	return(pred_df)
+}
+
+binfun <- function(mod, focal, non.focal, bins=50, ...) {
+	mf <- model.frame(mod)
+	mm <- (mf
+		%>% select_at(c(focal, non.focal))
+	)
+	check_df <- (mf
+		%>% arrange_at(focal)
+		%>% mutate(bin=ceiling(row_number()*bins/nrow(.)))
+		%>% group_by(bin)
+		%>% summarise_all(mean)
+		%>% mutate(model="binned")
+	)
+	return(check_df)
 }
@@
 ## Pop. average
-simple_vareff_age_pop <- varpred(simple_mod, "age", isolate=FALSE, pop.ave=TRUE, modelname="Pop. ave")
-# simple_vareff_age_pop <- popavefun(simple_mod, "age", "wealthindex", modelname = "Pop. ave") 
-# binned_df <- simple_vareff_age_pop$check
+simple_vareff_age_pop <- varpred(simple_mod, "age", isolate=FALSE, pop.ave="quantile", modelname="quantiles")
+binned_df <- binfun(simple_mod, "age", "wealthindex")
 
@@
 	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
+	+ geom_point(data=binned_df, aes(x=age, y=status, color="binned"))
 	+ scale_colour_manual(values=c("black", "blue", "red", "green"))
@@
 ## Pop. average
-simple_vareff_wealthindex_pop <- varpred(simple_mod, "wealthindex", isolate=FALSE, pop.ave=TRUE, modelname="Pop. ave")
-# binned_df <- simple_vareff_wealthindex_pop$check
+simple_vareff_wealthindex_pop <- varpred(simple_mod, "wealthindex", isolate=FALSE, pop.ave="quantile", modelname="quantiles")
+binned_df <- binfun(simple_mod, "wealthindex", "age")
 
@@
 	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
-#	+ geom_point(data=binned_df, aes(x=wealthindex, y=status, color="binned"))
+	+ geom_point(data=binned_df, aes(x=wealthindex, y=status, color="binned"))
 	+ scale_colour_manual(values=c("black", "blue", "red"))
@@
 ## Pop. average
-reff_vareff_age_pop <- varpred(reff_mod, "age", isolate=FALSE, pop.ave=TRUE, modelname = "Pop. ave")
+reff_vareff_age_pop <- varpred(reff_mod, "age", isolate=FALSE, pop.ave="quantile", include.re=TRUE, modelname = "quantiles")
+binned_df <- binfun(reff_mod, "age", "wealthindex")
 
@@
 	+ geom_hline(yintercept=true_prop_reff, lty=2, colour="grey")
+	+ geom_point(data=binned_df, aes(x=age, y=status, color="binned"))
 	+ scale_colour_manual(values=c("black", "blue", "red"))
@@
 
-```{r reff_vareff_wealthindex, fig.cap = "A comparison of population average and varpred based predictions. In the case of age, both methods seems to over-approximate the predictions."} 
+```{r reff_vareff_wealthindex, fig.cap = "A comparison of quantile and varpred based predictions. In both cases, varpred seems to over-approximate the predictions."} 
 # Wealth index
@@
 reff_vareff_wealthindex <- varpred(reff_mod, "wealthindex", isolate=FALSE, modelname="varpred")
+binned_df <- binfun(reff_mod, "wealthindex", "age")
 
 ## Pop. average
-reff_vareff_wealthindex_pop <- varpred(reff_mod, "wealthindex", isolate=FALSE, pop.ave=TRUE, modelname="Pop. ave")
+reff_vareff_wealthindex_pop <- varpred(reff_mod, "wealthindex", isolate=FALSE, pop.ave="quantile", include.re=TRUE, modelname="quantiles")
 
@@
 	+ geom_hline(yintercept=true_prop_reff, lty=2, colour="grey")
+	+ geom_point(data=binned_df, aes(x=wealthindex, y=status, color="binned"))
 	+ scale_colour_manual(values=c("black", "blue", "red"))

### 2021-07-28 logist_normal_bias.Rmd

Closed form integral of Gaussian over logistic curve?

We are interested in the mean of logistic(gaussian)

Tried it in Python (and Wolfram Alpha). This is just to reiterate what we knew already, which is that the logistic-normal doesn't have a closed form solution (WA can actually do the non-generic case (standard Normal), which sympy can't, but even WA breaks down for non-standard Normal case.

```{python}
from sympy import *
x = Symbol('x')
integrate(exp(-(x**2))/(1+exp(-x)), (x, -oo, oo))
```

A [CrossValidated question](https://stats.stackexchange.com/questions/45267/expected-value-of-a-gaussian-random-variable-transformed-with-a-logistic-functio) about calculating this transformation (and various interesting series approximations etc., although these are very likely not worth the trouble).  Turns out that if we don't want to use the delta method there is an existing package that codes up the mean & variance of the logistic-normal:

```{r mln_ex, eval=FALSE}
logitnorm::momentsLogitnorm(mu= 2, sigma = 2)
##       mean        var 
## 0.77520025 0.06186495 
```

although the entire function definition is not very complicated, we don't really need a package for this ...

```{r mln}
function (mu, sigma, abs.tol = 0, ...) {
    fExp <- function(x) plogis(x) * dnorm(x, mean = mu, sd = sigma)
    .exp <- integrate(fExp, -Inf, Inf, abs.tol = abs.tol, ...)$value
    fVar <- function(x) (plogis(x) - .exp)^2 * dnorm(x, mean = mu, 
        sd = sigma)
    .var <- integrate(fVar, -Inf, Inf, abs.tol = abs.tol, ...)$value
    c(mean = .exp, var = .var)
}
```

Further thoughts about my scribbles: the covariances of the design matrix do almost certainly differ if we condition on values/ranges of a focal predictor, e.g. if vars A and B are strongly correlated and we select a narrow range for A (the focal var), then the variance of B in that slice will inevitably be reduced.  Don't know if there are closed-form solutions for this (even assuming multivariate normality).

### 2021-08-10 bias_correction.rmd
Population averaging seems to work!

--- 2021-07-20 bias_correction.rmd
@@
 
-We intend to investigate our prediction based  on known truth and any bias potentially introduced by non-linear averaging, conditioning or random effect.
-We'll start with a simple case of a only fixed effect model and then consider a mixed effect model.
+Prediction of the variation in the response variable depends on whether the relationship between the response variable and the predictors is linear or nonlinear.
+For example, when response, $Y$, changes nonlinearly with the predictor variable, $X$, averaged response variable with respect to the predictor, $E(Y(X))$, does not necessarily equal the response at the mean of the predictor, $Y(E(X)$.
+This can be understood in relation to Jensen's inequality which states that, for a nonlinear function, $Y(X)$, then $E(Y(X)) > Y(E(X))$, if $Y(X)$ is positive second derivative; and $E(Y(X)) < Y(E(X))$ if $Y(X)$ is negative second derivative.
+Most packages for predicting responses make distribution assumptions about the predictors, (for example, conditioning at the mean value of the predictor), leading to potential biasness if the particular predictor is not well represented, or as a result of nonlinear averaging.
+We consider the following approaches for bias correction:
 
-\section{Simulation}
+- Population averaging
 
-We perform a simple simulation for a fixed effect model
+	- Whole population
+	- Quantiles
+
+- Distributional conditioning
+- Second-order correction
+
+We implement and apply these methods in the context of both simple generalized linear and mixed effect models, using simulated data sets.
+
+\section{Simple fixed effect model}
+
 \begin{align*}
@@
 \beta_0 &= 0.7 \\
-\beta_{\mathrm{A}} &= 0.3 \\
-\beta_{\mathrm{W}} &= 0.6
+\beta_{\mathrm{A}} &= 0.2 \\
+\beta_{\mathrm{W}} &= 0.5
 \end{align*}
@@
 
-```{r simple_coef}
+```{r simple_coef, fig.cap="Coefficient plot for simple logistic model."}
 ## True beta
@@
 
-```{r pop_ave_fun}
-popavefun <- function(mod, focal, non.focal, level=0.95, steps=100, bins=50, modelname="Pop. ave", ...) {
+```{r pop_ave_fun, echo=FALSE}
+popavefun <- function(mod, focal, non.focal, level=0.95, steps=100, modelname="quantiles", ...) {
 	mf <- model.frame(mod)
@@
 ## varpred way
-simple_vareff_age <- varpred(simple_mod, "age", isolate=FALSE, modelname="varpred")
+simple_vareff_age <- varpred(simple_mod, "age", isolate=FALSE, modelname="Distribution")
 
 ## Pop. average
-simple_vareff_age_pop <- varpred(simple_mod, "age", isolate=FALSE, pop.ave=TRUE, modelname="Pop. ave")
-# simple_vareff_age_pop <- popavefun(simple_mod, "age", "wealthindex", modelname = "Pop. ave") 
-# binned_df <- simple_vareff_age_pop$check
+### Quantiles
+simple_vareff_age_quant <- varpred(simple_mod, "age", isolate=FALSE, pop.ave="quantile", modelname="quantiles")
+### Population
+simple_vareff_age_pop <- varpred(simple_mod, "age", isolate=FALSE, pop.ave="population", modelname="Population")
+binned_df <- binfun(simple_mod, "age", "wealthindex", bins=15)
 
 vareff_age <- simple_vareff_age
-vareff_age$preds <- do.call("rbind", list(vareff_age$preds, simple_vareff_age_pop$preds))
+vareff_age$preds <- do.call("rbind", list(vareff_age$preds, simple_vareff_age_quant$preds, simple_vareff_age_pop$preds))
 age_plot <- (plot(vareff_age)
@@
 	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
-	+ scale_colour_manual(values=c("black", "blue", "red", "green"))
+	+ geom_point(data=binned_df, aes(x=age, y=status, color="binned"))
+	+ scale_colour_manual(values=c("grey", "blue", "red", "black"))
 	+ theme(legend.position="bottom")
@@
 
-```{r simple_vareff_wealthindex, fig.cap = "A comparison of population averaged and varpred-based predictions.
-In the case of biased predictions (age), population averaging gives better estimates.
-On the other hand, for wealthindex, the estimates varpred and population estimates are somehow similar."}
+```{r simple_vareff_wealthindex, fig.cap = "A comparison of various bias-correction approaches -- Distribution approach is based on averaging of non-focal predictors; Population approach involves using the observed values of non-focal predictors together with the quantiles of the focal predictors; while quantiles approach involves sampling (or picking quantiles of) the observed values of both focal and non-focal predictors.
+Except for the distribution approach, population and quantiles approaches give very close estimates."}
 # Wealth index
 ## varpred 
-simple_vareff_wealthindex <- varpred(simple_mod, "wealthindex", isolate=FALSE, modelname="varpred")
+simple_vareff_wealthindex <- varpred(simple_mod, "wealthindex", isolate=FALSE, modelname="Distribution")
 
 ## Pop. average
-simple_vareff_wealthindex_pop <- varpred(simple_mod, "wealthindex", isolate=FALSE, pop.ave=TRUE, modelname="Pop. ave")
-# binned_df <- simple_vareff_wealthindex_pop$check
+### Quantiles
+simple_vareff_wealthindex_quant <- varpred(simple_mod, "wealthindex", isolate=FALSE, pop.ave="quantile", modelname="quantiles")
+### Population
+simple_vareff_wealthindex_pop <- varpred(simple_mod, "wealthindex", isolate=FALSE, pop.ave="population", modelname="Population")
+binned_df <- binfun(simple_mod, "wealthindex", "age")
 
 vareff_wealthindex <- simple_vareff_wealthindex
-vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, simple_vareff_wealthindex_pop$preds))
+vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, simple_vareff_wealthindex_quant$preds, simple_vareff_wealthindex_pop$preds))
 wealthindex_plot <- (plot(vareff_wealthindex)
@@
 	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
-#	+ geom_point(data=binned_df, aes(x=wealthindex, y=status, color="binned"))
-	+ scale_colour_manual(values=c("black", "blue", "red"))
+	+ geom_point(data=binned_df, aes(x=wealthindex, y=status, color="binned"))
+	+ scale_colour_manual(values=c("grey", "blue", "red", "black"))
 	+ theme(legend.position="bottom")
@@
 
-- age: pop. average `r mean(simple_vareff_age_pop$preds$fit)`; varpred `r mean(simple_vareff_age$preds$fit)`
-- wealthindex: pop. average `r mean(simple_vareff_wealthindex_pop$preds$fit)`; varpred `r mean(simple_vareff_wealthindex$preds$fit)`
+- age: quantiles `r mean(simple_vareff_age_quant$preds$fit)`; population `r mean(simple_vareff_age_pop$preds$fit)`; distribution `r mean(simple_vareff_age$preds$fit)`
+- wealthindex: quantiles `r mean(simple_vareff_wealthindex_quant$preds$fit)`; population `r mean(simple_vareff_wealthindex_pop$preds$fit)`; distribution `r mean(simple_vareff_wealthindex$preds$fit)`
+
+\section{Random effect model}
+
+```{r sim_randeff}
+# Simulation parameters
+nHH <- 100	# Number of HH (primary units) per year
+
+nyrs <- 50	# Number of years to simulate
+yrs <- 2000 + c(1:nyrs) # Years to simulate
+N <- nyrs * nHH
+
+## HH random effect sd
+hhSD <- 0.5
+
+# Generate dataset template
+temp_df <- (data.frame(hhid = rep(c(1:nHH), each = nyrs)
+		, years = rep(yrs, nHH)
+		, age = runif(n=N, age_min, age_max)
+		, wealthindex = rnorm(n = N, 0, 1)
+	)
+)
+
+# Simulate HH-level random effects (residual error)
+hhRE <- rnorm(nHH, hhSD)
+temp_df$hhRE <- hhRE[temp_df$hhid]
+
+sim_df <- (temp_df
+	%>% mutate(eta = beta0 + betaA * age + betaW * wealthindex + hhRE
+		, status = rbinom(N, 1, plogis(eta))
+	)
+	%>% select(-eta)
+)
+true_prop_reff <- mean(sim_df$status)
+print(true_prop)
+print(head(sim_df, 10))
+```
+
+\subsection{Fit model}
+
+```{r model_reff}
+reff_mod <- glmmTMB(status ~ age + wealthindex + (1|hhid)
+	, data = sim_df
+	, family = binomial(link = "logit")
+)
+```
+
+```{r reff_coef, fig.cap="Mixed model coefficient estimates."}
+## Tidy coef estimates
+reff_coef_df <- (broom.mixed::tidy(reff_mod, conf.int=TRUE)
+	%>% mutate(term = gsub("\\(|\\)", "", term))
+	%>% filter(effect=="fixed")
+)
+print(reff_coef_df)
+
+reff_coef_plot <- (plotEsize(reff_coef_df) 
+	+ geom_point(data=true_beta_df, aes(x=term, y=estimate, colour="Truth"))
+	+ labs(colour="Type")
+	+ scale_colour_manual(values=c("black", "blue"))
+)
+print(reff_coef_plot)
+```
+
+Variable effect plots
+
+- Age
+
+```{r reff_vareffects_age}
+## varpred way
+reff_vareff_age <- varpred(reff_mod, "age", isolate=FALSE, modelname="Distribution")
+
+## Pop. average
+### Quantiles
+reff_vareff_age_quant <- varpred(reff_mod, "age", isolate=FALSE, pop.ave="quantile", include.re=TRUE, modelname = "quantiles")
+### Population
+reff_vareff_age_pop <- varpred(reff_mod, "age", isolate=FALSE, pop.ave="population", include.re=TRUE, modelname = "Population")
+binned_df <- binfun(reff_mod, "age", "wealthindex", bins=15)
+
+vareff_age <- reff_vareff_age
+vareff_age$preds <- do.call("rbind", list(vareff_age$preds, reff_vareff_age_quant$preds, reff_vareff_age_pop$preds))
+age_plot <- (plot(vareff_age)
+ 	+ labs(y="Prob. of improved \n service", colour="Model")
+	+ geom_hline(yintercept=true_prop_reff, lty=2, colour="grey")
+	+ geom_point(data=binned_df, aes(x=age, y=status, color="binned"))
+	+ scale_colour_manual(values=c("grey", "blue", "red", "black"))
+	+ theme(legend.position="bottom")
+)
+```
+
+- Wealth index
+
+```{r reff_vareff_wealthindex, fig.cap = "Distribution based approach over-estimates the predictions in age and wealth index."} 
+# Wealth index
+## varpred
+reff_vareff_wealthindex <- varpred(reff_mod, "wealthindex", isolate=FALSE, modelname="Distribution")
+binned_df <- binfun(reff_mod, "wealthindex", "age")
+
+## Pop. average
+### Quantiles
+reff_vareff_wealthindex_quant <- varpred(reff_mod, "wealthindex", isolate=FALSE, pop.ave="quantile", include.re=TRUE, modelname="quantiles")
+### Population
+reff_vareff_wealthindex_pop <- varpred(reff_mod, "wealthindex", isolate=FALSE, pop.ave="population", include.re=TRUE, modelname="Population")
+
+vareff_wealthindex <- reff_vareff_wealthindex
+vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, reff_vareff_wealthindex_quant$preds, reff_vareff_wealthindex_pop$preds))
+vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, reff_vareff_wealthindex_quant$preds, reff_vareff_wealthindex_pop$preds))
+wealthindex_plot <- (plot(vareff_wealthindex)
+ 	+ labs(y="", colour="Model")
+	+ geom_hline(yintercept=true_prop_reff, lty=2, colour="grey")
+	+ geom_point(data=binned_df, aes(x=wealthindex, y=status, color="binned"))
+	+ scale_colour_manual(values=c("grey", "blue", "red", "black"))
+	+ theme(legend.position="bottom")
+)
+
+ggarrange(age_plot, wealthindex_plot, common.legend=TRUE)
+```

### 2021-08-11 second_order_correction.rmd
Taylor approximation

---
title: "Second order bias correction"
author: Bicko, Jonathan & Ben
date: "2021 Aug 11 (Mon)"
compact-title: false
output:
  pdf_document:
    extra_dependencies: ["float"]
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(fig.pos = "H", out.extra = "")

library(glmmTMB)
library(dplyr)
library(ggplot2)
library(vareffects); varefftheme()
library(ggpubr)
set.seed(1011)

```

\section{Introduction}

We consider Taylor series approximation approach for bias correction. Suppose $f(.)$ represents a back-transformation function, $\mu$ the mean and $\sigma^2_\mu$ on the linear predictor scale (**univariate case** // **what happens in multivariate case?**) predictor. Then using Taylor expansion around $\mu$, the approximation on the original scale is given by

\[
f(\mu) + \frac{1}{2}\sigma^2_{\mu}f''(\mu)
\]

To start with, we consider logistic function

\[
g(\mu) = \frac{1}{1 + \exp(-\mu)},
\]
with the first and second derivatives give by

\[
g'(\mu) = g(\mu)(1 - g(\mu))
\]
and

\[
g''(\mu) = g(\mu)g(-\mu)(1 - 2g(\mu))
\]
respectively.

```{r taylor_fun}
taylorapprox <- function(fun, mu, sigma, ...) {
	untrans <- fun(mu)
	der2 <- fun(mu)*fun(-mu)*(1 - 2*fun(mu))
	corrected <- untrans + der2*sigma^2/2
	return(corrected)
}
```

Let us consider a simple univariate case

```{r simple_univariate}
b0 <- 6
b1 <- 5
N <- 100
x <- scale(rnorm(N))
eta <- b0 + b1*x

corrected <- taylorapprox(plogis, mean(eta), sd(eta))
out <- c(NULL
	, mean = mean(plogis(eta))
	, uncorrected = plogis(mean(eta))
	, corrected = corrected
)
print(out)
logitnorm::momentsLogitnorm(mu= mean(eta), sigma = sd(eta))
```

**??**

- How can we verify this?
- Does anything change in multivariate case?
- How does it work in _focal_--_non-focal_ variables case?

### 2021-08-11 logist_normal_bias.rmd
working on bias document; rename to rmd

---
title: "Nonlinear transformations and bias correction"
---

Generating predictions and confidence intervals of predictions for regression models is surprisingly hard. A short list of the challenges:

- dealing with multiple predictor variables: choosing representative values for one or more 'focal variables' and choosing an appropriate 'model center' for non-focal variables
- Jensen's inequality and bias in the mean induced by nonlinear transformations of the response variable (e.g., link functions in GLM(M)s)
- propagation of uncertainty ([how] can we incorporate uncertainty in nonlinear components of (G)LMMs? should we exclude variation in non-focal parameters (JD-style effect plots)? Can we avoid assuming a Gaussian sampling distribution of the uncertainty, e.g. by importance sampling?) (Many but not necessarily all of these issues are resolved by moving to Bayesian sampling-based approaches, at the cost of additional (explicit) assumptions and computational effort.)

More broadly:

- what questions do we want to ask (this implicitly determines answers to questions about where the model center should be)?
- what assumptions are we willing to make (multivariate normality, weak nonlinearity [for delta-method approximations], independence of non-focal variables?)

Note that we will be interested in/concerned about the distribution of the predictors (we have the distribution represented by our sample; we may want to treat this as a sample from an underlying distribution, e.g. multivariate Normal or something else, both for computational convenience and in order to get a smoother prediction) as well as the sampling distribution of the coefficients (which we will often assume is multivariate Normal for convenience). Random-effects coefficients are an interesting case; they are coefficients, but we could also treat them as part of a population sample. (We usually assume that random-effects coefficients are independent of everything else.)

In what follows let's say we have a focal predictor $x_f$ and a set of non-focal predictors $x_\{n\}$ (and without loss of generality, the focal predictor is first in the vector of predictors).

## Population-based approach

\newcommand{\bX}{{\mathbf X}}
\newcommand{\bbeta}{{\boldsymbol \beta}}
\newcommand{\boldeta}{{\boldsymbol \eta}}

The most precise (although not necessarily accurate!) way to predict is to condition on a value $F$ of the focal predictor and make predictions for all observations (members of the population) for which $x_f = F$ (or in a small range around $F$ ...). The linear predictor values are $\boldeta = (F,\bX_\{-f\}) \bbeta$.

Closed form integral of Gaussian over logistic curve?

We are interested in the mean of logistic(gaussian)

Tried it in Python (and Wolfram Alpha). This is just to reiterate what we knew already, which is that the logistic-normal doesn't have a closed form solution (WA can actually do the non-generic case (standard Normal), which sympy can't, but even WA breaks down for non-standard Normal case.

```{python}
from sympy import *
x = Symbol('x')
integrate(exp(-(x**2))/(1+exp(-x)), (x, -oo, oo))
```

A [CrossValidated question](https://stats.stackexchange.com/questions/45267/expected-value-of-a-gaussian-random-variable-transformed-with-a-logistic-functio) about calculating this transformation (and various interesting series approximations etc., although these are very likely not worth the trouble).  Turns out that if we don't want to use the delta method there is an existing package that codes up the mean & variance of the logistic-normal:

```{r mln_ex, eval=FALSE}
logitnorm::momentsLogitnorm(mu= 2, sigma = 2)
##       mean        var 
## 0.77520025 0.06186495
x <- rnorm(1e5, 2, 2)
x <- scale(x)*2 + 2 ## exact moments
mean(x)
mean(plogis(x))
plogis(mean(x))
deltamethod(1/(1+exp(-x)), x, max.order=2)
```

although the entire function definition is not very complicated, we don't really need a package for this ...

```{r mln}
function (mu, sigma, abs.tol = 0, ...) {
    fExp <- function(x) plogis(x) * dnorm(x, mean = mu, sd = sigma)
    .exp <- integrate(fExp, -Inf, Inf, abs.tol = abs.tol, ...)$value
    fVar <- function(x) (plogis(x) - .exp)^2 * dnorm(x, mean = mu, 
        sd = sigma)
    .var <- integrate(fVar, -Inf, Inf, abs.tol = abs.tol, ...)$value
    c(mean = .exp, var = .var)
}
```

Further thoughts about my scribbles: the covariances of the design matrix do almost certainly differ if we condition on values/ranges of a focal predictor, e.g. if vars A and B are strongly correlated and we select a narrow range for A (the focal var), then the variance of B in that slice will inevitably be reduced.  Don't know if there are closed-form solutions for this (even assuming multivariate normality).

### 2021-08-11 logist_normal_bias.rmd
more logistic-normal stuff

@@
 The most precise (although not necessarily accurate!) way to predict is to condition on a value $F$ of the focal predictor and make predictions for all observations (members of the population) for which $x_f = F$ (or in a small range around $F$ ...).
-The linear predictor values are $\boldeta = (F,\bX_\{-f\}) \bbeta$.
+The linear predictor values are $\boldeta = (F,\bX_{\{-f\}}) \bbeta$.
+A key point is that the nonlinear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population (e.g.
+all individuals with $x_f = F$) to a one-dimensional distribution of $\eta$.
+
+Once we've got our vector of $\boldeta$ (which is essentially a set of samples from the distribution over $\eta$ for the conditional set), we want a mean and confidence intervals on the mean on the response (data) scale, i.e.
+after back-transforming.
+Most precisely the mean is $\int P(\eta') g^{-1}(\eta') d\,\eta$.
+
+- if we use the observations themselves then we just compute the individual values of $g^{-1}(\eta)$ and compute their mean
+- we could compute the quantiles of the distribution of $\eta$ and use this to construct an approximate Riemann sum over the distribution ($\sum g^{-1}(Q_i(\eta))/n_Q$ ?)
+- compute the mean and variance of the $\eta$ values, assume Normality (or some other distribution we like), and integrate numerically 
+- for some inverse-link functions like the exponential (but not logistic), there's a closed-form solution
+- assume Normality, use Gaussian quadrature instead of brute-force numerical integration
+- assume Normality, use the delta-method (second-order Taylor) approximation
+
+What about the confidence intervals?
+The limits of the confidence intervals are points, not mean values.
+In principle, every observation/set of non-focal predictors has a different CI.
+The predictions are $\bX \bbeta$; the variances of the predictions are $\textrm{Diag}(\bX \Sigma \bX^\top)$ (I think [this can be computed more efficiently since we only want the diagonal: see [here](https://stackoverflow.com/questions/56962209/compact-efficient-replacement-for-diagx-v-xt)]).
+I'm not sure how to combine these: do we take the distributions of the upper/lower CIs (e.g.
+Wald intervals using $\pm q \sigma$ where $q$ is an appropriate quantile of the Normal or t distribution) and find their transformed mean values in the same way?
+???
+If we consider only the uncertainty of the focal predictor, so that the confidence intervals are $\eta \pm q \sigma_f$, we still have the same question of how to condense the set of endpoints.
+
+If we compute the individual back-transformed predictions for a poorly sampled/finely spaced set of focal values, we will get a noisy prediction line as the values of the non-focal predictors shift across the focal values.
+(Simple example: suppose everyone below the median age has wealth index $w_1$, everyone above the median has $w_2$.
+Then the predicted value will have a discontinuity at the median age.) We can deal with this by taking bigger bins (a form of smoothing), or by post-smoothing the results (by loess or something).
+The principled form of this would be to assume/recognize that our uneven distribution of observed non-focal predictors actually represents a sample of a distribution that will vary \emph{smoothly} as a function of the focal predictor.
+
+## Full-normal approximation
+
+Suppose we assume that the full distribution of all of the predictors is multivariate Normal.
+Then for a fixed value of $x_f = F$ the conditional distribution of the non-focal predictors is also MVN with mean and covariance matrix we can get from [Wikipedia](https://en.wikipedia.org/wiki/Multivariate_normal_distribution#Conditional_distributions) [!!] (see also lots of answers from [this twitter thread](https://twitter.com/bolkerb/status/1425546539833995266), including [this blog post](https://fabiandablander.com/statistics/Two-Properties.html)).
+Once we know the mean and variance of the predictors we can directly compute the mean and variance of the linear predictor (mean is $\bX' \bbeta$ as before, variance is $\bbeta \Sigma' \bbeta\top$), and proceed as above.
+
+## Links to MRP
+
+What does MRP do about this?
+Does the Bayesian averaging over uncertainty somehow smooth over the population distribution?
+Or do they directly construct an estimate of the population-level distributions of predictors?
+
+## junk
 
@@
 
-```{python}
+```{python integrate}
 from sympy import *
@@
 A [CrossValidated question](https://stats.stackexchange.com/questions/45267/expected-value-of-a-gaussian-random-variable-transformed-with-a-logistic-functio) about calculating this transformation (and various interesting series approximations etc., although these are very likely not worth the trouble).
-Turns out that if we don't want to use the delta method there is an existing package that codes up the mean & variance of the logistic-normal:
+Turns out that if we don't want to use the delta method there is an existing package that codes up the mean & variance of the logistic-normal (exact, up to integration accuracy):
 
-```{r mln_ex, eval=FALSE}
+```{r mln_ex}
 logitnorm::momentsLogitnorm(mu= 2, sigma = 2)
-##       mean        var 
-## 0.77520025 0.06186495
 x <- rnorm(1e5, 2, 2)
@@
 mean(x)  
-mean(plogis(x))
-plogis(mean(x))
-deltamethod(1/(1+exp(-x)), x, max.order=2)
+mean(plogis(x))  ## correct value
+plogis(mean(x))  ## naive value
+emdbook::deltamethod(1/(1+exp(-x)), x)
+## The `delta2` value is supposedly a higher-order Taylor expansion/correction, but a little bit of poking around suggests these are bogus/possibly wrong ...
 ```
 
-although the entire function definition is not very complicated, we don't really need a package for this ...
+The entire `mln()` function definition is not very complicated, we don't really need a package for this ...
 
@@
 ```
-
-Further thoughts about my scribbles: the covariances of the design matrix do almost certainly differ if we condition on values/ranges of a focal predictor, e.g.
-if vars A and B are strongly correlated and we select a narrow range for A (the focal var), then the variance of B in that slice will inevitably be reduced.
-Don't know if there are closed-form solutions for this (even assuming multivariate normality).

### 2021-08-13 second_order_correction.rmd
Autocommit

@@
 logitnorm::momentsLogitnorm(mu= mean(eta), sigma = sd(eta))
+emdbook::deltamethod(1/(1+exp(-x)), eta)
 ```

### 2021-08-17 aprroximation_bias_correction.rmd
More examples on delta and taylor approximation

---
title: "Taylor approximation - based methods for bias correction"
author: Bicko, Jonathan & Ben
date: "2021 Aug 11 (Mon)"
compact-title: false
output:
  pdf_document:
    extra_dependencies: ["float"]
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(fig.pos = "H", out.extra = "")

library(glmmTMB)
library(dplyr)
library(ggplot2)
library(vareffects); varefftheme()
library(ggpubr)
set.seed(1011)

prob2logit <- function(x)log(x / (1 - x))
```

\section{Introduction}

We consider two approaches:

\begin{itemize}
\item Second order Taylor approximation
\item Delta method
\end{itemize}

and then compare our approximations to _closed form integral of Gaussian over logistic curve_.

\subsection{Second order Taylor approximation}

We consider Taylor series approximation approach for bias correction. Suppose $f(.)$ represents a back-transformation function, $\mu$ the mean and $\sigma^2_\mu$ on the linear predictor scale (**univariate case** // **what happens in multivariate case?**) predictor. Then using Taylor expansion around $\mu$, the approximation on the original scale is given by

\[
f(\mu) + \frac{1}{2}\sigma^2_{\mu}f''(\mu)
\]

To start with, we consider logistic function

\[
g(\mu) = \frac{1}{1 + \exp(-\mu)},
\]
with the first and second derivatives give by

\[
g'(\mu) = g(\mu)(1 - g(\mu))
\]
and

\[
g''(\mu) = g(\mu)g(-\mu)(1 - 2g(\mu))
\]
respectively.

```{r taylor_fun}
taylorapprox <- function(fun, var="x", mu, sigma, ...) {
	ff <- as.expression(substitute(fun))
	x <- mu
	untrans <- eval(ff, list(x))
	der <- D(ff, var)		# 1s derivative
	der2 <- D(der, var)	# 2nd derivative
	der2 <- eval(der2, list(x))
	corrected <- untrans + der2*sigma^2/2
	return(corrected)
}
```

\subsection{Delta method}

Consider a first order Taylor approximation of $f(.)$ about the mean $\mu$, evaluated at the random variable $x_i$ defined as:

\[
f(x_i) \approx f(\mu) + f'(x_i)(x_i - \mu)
\]
with a variance of

\[
\mbox{Var}\left(f(x_i)\right) = f'(\mu)\mbox{Var}(x_i)f'(\mu)
\]

```{r deltaapprox}
deltaapprox <- function(fun, var="x", z, ...) {
	ff <- as.expression(substitute(fun))
	der <- D(ff, var)
	x <- z
	ymu <- eval(ff, list(x))
	x <- mean(z)
	der <- eval(der, list(x))
	y <- ymu + der * (z - x)
	yvar <- der * var(z) * der		# Check with BB
	out <- list(fx = y, var = yvar)
	return(out)
}
```

Let us consider a simple univariate case

```{r simple_univariate}
b0 <- 6
b1 <- 5
N <- 100
x <- scale(rnorm(N))
eta <- b0 + b1*x

taylor <- taylorapprox(fun=1/(1+exp(-x)), mu=mean(eta), sigma=sd(eta))
delta <- deltaapprox(fun=1/(1+exp(-x)), z=eta)
normal_approx <- logitnorm::momentsLogitnorm(mu= mean(eta), sigma = sd(eta))
out <- c(NULL
	, truth = mean(plogis(eta))
	, naive = plogis(mean(eta))
	, taylor = taylor
	, delta = mean(delta$fx)
	, normal = normal_approx[[1]]
)
print(out)
delta$var
normal_approx[[2]]
```

Simple logistic model example

```{r simple_sim}
N <- 1e4
beta0 <- 0.7
betaA <- 0.2
betaW <- 0.5

age_max <- 1
age_min <- 0.2
age <- runif(N, age_min, age_max)
# age <- rnorm(N, age_max, age_max)

wealthindex <- rnorm(N, 0, 1)

eta <- beta0 + betaA * age + betaW * wealthindex
sim_df <- (data.frame(age=age, wealthindex=wealthindex, eta=eta)
	%>% mutate(status = rbinom(N, 1, plogis(eta)))
	%>% select(-eta)
)
true_prop <- mean(sim_df$status)
print(true_prop)
```

```{r simple_logistic}
simple_mod <- glm(status ~ age + wealthindex, data = sim_df, family="binomial")
```

Applying `taylor`, `delta` and `normal approximation` method

```{r generate_pred}
genpred <- function(mod, focal, non.focal, level=0.95, steps=100, pop=FALSE, bias.adjust=c("none", "delta", "normal", "taylor"), modelname="none", ...) {

	bias.adjust <- match.arg(bias.adjust)
	mf <- model.frame(mod)
	mm <- (mf
		%>% select_at(c(focal, non.focal))
	)
	quant <- seq(0, 1, length.out=steps)
	mm1 <- sapply(focal, function(x)as.vector(quantile(mm[,x], quant)), simplify = FALSE)
	if (pop) {
		mm1[[non.focal]] <- mm[[non.focal]]
		mm <- do.call("expand.grid", mm1)
		mm <- model.matrix(formula(mod)[c(1,3)], mm)
	} else {
		mm2 <- sapply(non.focal, function(x)mean(mm[,x]), simplify = FALSE)
		mm <- do.call("data.frame", c(mm1, mm2))
		mm <- model.matrix(formula(mod)[c(1,3)], mm)
	}
	linpred <- as.vector(mm %*% coef(mod))
	out <- (mm
		%>% data.frame()
		%>% select_at(focal)
		%>% mutate(lp = linpred, model=modelname)
	)

	if (bias.adjust=="delta") {
		out <- (out
			%>% mutate(fit=deltaapprox(fun=1/(1+exp(-x)), z=lp)$fx)
		)
	} else if (bias.adjust=="taylor") {
		vc <- vcov(mod)
		pse_var <- sqrt(rowSums(mm * t(tcrossprod(data.matrix(vc), mm))))
		lp <- out$lp
		fit <- unlist(lapply(1:length(lp), function(i){
			taylorapprox(fun=1/(1+exp(-x)), mu=lp[[i]], sigma=pse_var[[i]])
		}))
		out <- (out
			%>% select(-lp)
			%>% mutate(fit = fit) 
		)
	}	else if (bias.adjust=="normal") {
		vc <- vcov(mod)
		pse_var <- sqrt(rowSums(mm * t(tcrossprod(data.matrix(vc), mm))))
		lp <- out$lp
		fit <- unlist(lapply(1:length(lp), function(i){
			logitnorm::momentsLogitnorm(mu = lp[[i]], sigma = pse_var[[i]])[[1]]
		}))
		out <- (out
			%>% select(-lp)
			%>% mutate(fit = fit) 
		)
	} else if (bias.adjust=="none") {
		out <- (out
			%>% mutate(fit = plogis(lp)) 
		)
	}
	if (pop) {
		out <- (out
			%>% group_by_at(focal)
			%>% summarise_at("fit", mean)
			%>% mutate(model=modelname)
		)
	}
	return(out)
}

## Population averaging
pop_none <- genpred(simple_mod, "age", "wealthindex", pop=TRUE, bias.adjust="none", modelname="pop-none")
pop_delta <- genpred(simple_mod, "age", "wealthindex", pop=TRUE, bias.adjust="delta", modelname="pop-delta")

## Averaged // centered??
centered_none <- genpred(simple_mod, "age", "wealthindex", pop=FALSE, bias.adjust="none", modelname="centered-none")
centered_delta <- genpred(simple_mod, "age", "wealthindex", pop=FALSE, bias.adjust="delta", modelname="centered-delta")
centered_taylor <- genpred(simple_mod, "age", "wealthindex", pop=FALSE, bias.adjust="taylor", modelname="centered-taylor")

```

```{r simple_preds}
p1 <- (ggplot(pop_none, aes(x=age, y=fit, colour=model))
	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
	+ geom_line()
	+ geom_line(data=pop_delta, aes(x=age, y=fit, colour=model))
	+ geom_line(data=centered_none, aes(x=age, y=fit, colour=model))
	+ geom_line(data=centered_delta, aes(x=age, y=fit, colour=model))
	+ geom_line(data=centered_taylor, aes(x=age, y=fit, colour=model))
	+ scale_colour_manual(values=c("blue", "red", "black", "orange", "purple"))
	+ theme(legend.position="right")
)
print(p1)
```

### 2021-08-21 logist_normal_bias.rmd
Delta method implementation
more on prediction vs effects plots

@@
 ```
+
+## Prediction vs effects plots
+
+The purpose and goal of a *prediction plot* seems fairly straightforward; for specified values of (a) focal predictor(s), we want to give a point estimate and confidence intervals for the prediction of the model for a "typical" (= random sample over the multivariate distribution of non-focal parameters) individual with those values of the predictors ("what is the expected probability of having clean water for a 25-year-old male"?).
+These are most of the problems we have actually been addressing with our bias correction stuff above.
+To do these computations, we need to take the values of the non-focal predictors (or their means and covariances) from the *conditional distribution*.
+(If the focal predictors are discrete, we condition on exact values; if they are continuous/have mostly unique values, we condition on appropriately sized bins.)
+
+In contrast, an *effects plot* is supposed to show the *marginal effect* (in the economics/differential sense) of varying the focal parameter(s).
+This means, technically that we want the *derivative* (and its uncertainty) with respect to the focal parameters to match that predicted by the model.
+
+* Although it might not matter, we do need pick a reference point to start from (e.g.
+the prediction for the model center).
+Then, *on the linear predictor scale*, we can compute the derivative (which is just the coefficient for the focal parameter unless there are interactions, in which case we need to incorporate the conditional values/distribution of the non-focal parameters in the calculation).
+If there are no interactions, then the derivative on the response scale is the same for the entire distribution and by the chain rule is the product of the derivative ($\beta_f = d\eta/dx_f$) and the derivative of the inverse link function at the point ($dg^{-1}(\eta)/d\eta$, which is the `mu.eta` component of a `family` object in R).
+How does variability (due to variation in the non-focal parameters + interactions) affect this result??
+Suppose $\hat \beta_f(x_n)$ is the derivative for a particular set of non-focal parameters (i.e.
+$\beta_f + \sum \beta_{fi} x_i$ where the second term is the sum of the relevant interaction coefficients)
+
+$$
+\overline{\frac{d \mu}{dx}} = \int \frac{dg^{-1}(\eta)}{d\eta} \hat \beta_f \, d x_n
+$$
+
+This seems to be linear in $x_n$, so I think we get to *not* worry about averaging effects here?
+
+It seems that the appropriate computed value (y-axis height) of the effects plot at a given value of $x$ is 
+$$
+\begin{split}
+ & y_{\textrm{ref}} + \int \frac{dg^{-1}(\eta)}{d\eta} \hat \beta_f \, dx_f \\
+ = & y_{\textrm{ref}} + \frac{dg^{-1}(\eta)}{d\eta} \overline{\hat \beta_f} \\
+ \approx & y_{\textrm{ref}} + \sum \frac{dg^{-1}(\eta(x_f,\bar x_n))}{d\eta} \hat \beta_f(\bar x_n) \, \Delta x_f
+\end{split}
+$$
+
+in other words, start from the appropriately bias-corrected value on the response scale at the central value of $x_f$ ($y_{\textrm{ref}}$).
+For every step $\Delta x_f$, calculate the value of the slope (including interactions) at the *mean value of the non-focal parameters conditional on the current value of $x_f$*.
+(If there are no interactions, this reduces to $\beta_f$.) Calculate the linear predictor, again based on the conditional mean value of the non-focal predictors; use it to compute the expansion factor (derivative of the inverse-link function).
+Take the product of the slope, the expansion factor, and $\Delta x_f$; add to the integrand and repeat.
+
+The good news is that we don't have to fart around with logistic-normal integrals etc.
+except when we are computing $y_{\textrm{\ref}}$.
+The bad news is that the effect plot computed this way will only agree with the prediction plot in special cases ...

### 2021-08-25 aprroximation_bias_correction.rmd
Autocommit

@@
 
+```{r generate_pred2}
+genpred2 <- function(mod, focal, level=0.95, steps=100, modelname="none", ...) {
+
+	mm <- model.matrix(mod)
+	quant <- seq(0, 1, length.out=steps)
+	xvals <- as.vector(quantile(mm[, focal], quant))
+	vnames <- vareffects:::get_vnames(mod)$termnames
+	check <- !vnames %in% focal
+	non.focal.terms <- vnames[check]
+	betas <- coef(mod)
+	non.focal.coefs <- betas[check]
+	focal.coefs <- betas[!check]
+	lp.non.focal <- as.vector(as.matrix(mm[, non.focal.terms, drop=FALSE]) %*% non.focal.coefs)
+	lp.all <- unlist(lapply(xvals, function(x){
+		mu <- mean(plogis(as.vector(c(x %*% focal.coefs) + lp.non.focal)))
+		return(mu)
+	}))
+
+	out <- list(xx = xvals, xx2 = lp.all)
+	return(out)
+}
+
+x1 <- genpred2(simple_mod, "wealthindex")
+mean(x1$xx2)
+
+```
+
 ```{r simple_preds}

### 2021-08-25 bias_correction.rmd
Autocommit

@@
 ```{r setup, include=FALSE, echo=FALSE}
-knitr::opts_chunk$set(fig.pos = "H", out.extra = "")
+knitr::opts_chunk$set(fig.pos = "H", out.extra = "", message=FALSE, warning=FALSE)
 
@@
 This can be understood in relation to Jensen's inequality which states that, for a nonlinear function, $Y(X)$, then $E(Y(X)) > Y(E(X))$, if $Y(X)$ is positive second derivative; and $E(Y(X)) < Y(E(X))$ if $Y(X)$ is negative second derivative.
-Most packages for predicting responses make distribution assumptions about the predictors, (for example, conditioning at the mean value of the predictor), leading to potential biasness if the particular predictor is not well represented, or as a result of nonlinear averaging.
+Most packages for predicting responses make distribution assumptions about the non-focal predictors, (for example, conditioning at the mean value of the predictor), leading to potential biasness if the particular predictor is not well represented, or as a result of nonlinear averaging.
 We consider the following approaches for bias correction:
@@
 	- Quantiles
-
-- Distributional conditioning
 - Second-order correction
 
-We implement and apply these methods in the context of both simple generalized linear and mixed effect models, using simulated data sets.
+and provide a comparison with the uncorrected _Distributional conditioning_ (varpred or emmeans approach).
+
+We implement and apply these methods in the context of both simple generalized linear and mixed effect models, using simulated data sets -- for univariate and multivariate models.
+
+We start with a univariate case where we have only one predictor.
 
@@
 
-Coefficient plots
-
-```{r simple_coef, fig.cap="Coefficient plot for simple logistic model."}
-## True beta
-true_beta_df <- data.frame(term=c("Intercept", "age", "wealthindex")
-	, estimate=c(beta0, betaA, betaW)
-)
-
-## Tidy coef estimates
-coef_df <- (broom::tidy(simple_mod, conf.int=TRUE)
-#	%>% dotwhisker::by_2sd(sim_df)
-	%>% mutate(term = gsub("\\(|\\)", "", term))
-)
-print(coef_df)
-
-simple_coef_plot <- (plotEsize(coef_df) 
-	+ geom_point(data=true_beta_df, aes(x=term, y=estimate, colour="Truth"))
-	+ labs(colour="Type")
-	+ scale_colour_manual(values=c("black", "blue"))
-)
-print(simple_coef_plot)
-```
-
-Effect sizes on logit scale:
-
-```{r simple_logit}
-coef_df_logit <- (coef_df
-	%>% select(term, estimate, conf.low, conf.high)
-	%>% group_by(term)
-	%>% summarise_all(plogis)
-)
-print(coef_df_logit)
-```
-
 \subsection{Variable predictions}
-
-We consider both \textbf{varpred} and population averaging approach; and then introduce bias correction to \textbf{varpred} predictions.
 
@@
 
-- Age
-
-```{r simple_vareffects_age} 
+```{r simple_vareffects_age, echo=FALSE} 
 ## varpred way
@@
 
-- Wealth index
-
-```{r simple_vareff_wealthindex, fig.cap = "A comparison of various bias-correction approaches -- Distribution approach is based on averaging of non-focal predictors; Population approach involves using the observed values of non-focal predictors together with the quantiles of the focal predictors; while quantiles approach involves sampling (or picking quantiles of) the observed values of both focal and non-focal predictors.
-Except for the distribution approach, population and quantiles approaches give very close estimates."}
+```{r simple_vareff_wealthindex, echo=FALSE, fig.cap = "A comparison of variable predictions with uncorrected and corrected estimates.
+The uncorrected Distribution approach is based on averaging of non-focal predictors; Population approach involves using the observed values of non-focal predictors together with the quantiles of the focal predictors; while quantiles approach involves sampling (or picking quantiles of) the observed values of both focal and non-focal predictors.
+The uncorrected Distribution approach seems to over-predict and is slightly higher then the marginal estimates."}
 # Wealth index
@@
 
-The observed population average is `r true_prop` while the estimated population averages (similar estimates with `varpred`) are:
+The marginal mean is `r true_prop` while the estimated are:
 
@@
 
-```{r reff_coef, fig.cap="Mixed model coefficient estimates."}
-## Tidy coef estimates
-reff_coef_df <- (broom.mixed::tidy(reff_mod, conf.int=TRUE)
-	%>% mutate(term = gsub("\\(|\\)", "", term))
-	%>% filter(effect=="fixed")
-)
-print(reff_coef_df)
+\subsection{Variable predictions}
 
-reff_coef_plot <- (plotEsize(reff_coef_df) 
-	+ geom_point(data=true_beta_df, aes(x=term, y=estimate, colour="Truth"))
-	+ labs(colour="Type")
-	+ scale_colour_manual(values=c("black", "blue"))
-)
-print(reff_coef_plot)
-```
-
-Variable effect plots
-
-- Age
-
-```{r reff_vareffects_age}
+```{r reff_vareffects_age, echo=FALSE}
 ## varpred way
@@
 
-- Wealth index
-
-```{r reff_vareff_wealthindex, fig.cap = "Distribution based approach over-estimates the predictions in age and wealth index."} 
+```{r reff_vareff_wealthindex, echo=FALSE, fig.cap = "Similar to the previous simple model, the quantile and population-based averaging gives very close estimates as opposed to the uncorrected distribution-based which appears to over-predict."}
 # Wealth index
@@
 ```
+
+We can also add centered predictions as in the case of _varpred_:
+
+```{r reff_vareffects_age_centered, echo=FALSE}
+## varpred way
+reff_vareff_age_cen <- varpred(reff_mod, "age", isolate=TRUE, modelname="Distribution")
+
+## Pop. average
+### Quantiles
+reff_vareff_age_quant_cen <- varpred(reff_mod, "age", isolate=TRUE, pop.ave="quantile", include.re=TRUE, modelname = "quantiles")
+### Population
+reff_vareff_age_pop_cen <- varpred(reff_mod, "age", isolate=TRUE, pop.ave="population", include.re=TRUE, modelname = "Population")
+binned_df <- binfun(reff_mod, "age", "wealthindex", bins=15)
+
+vareff_age <- reff_vareff_age_cen
+vareff_age$preds <- do.call("rbind", list(vareff_age$preds, reff_vareff_age_quant_cen$preds, reff_vareff_age_pop_cen$preds))
+age_plot_cen <- (plot(vareff_age)
+ 	+ labs(y="Prob. of improved \n service", colour="Model")
+	+ geom_hline(yintercept=true_prop_reff, lty=2, colour="grey")
+	+ geom_point(data=binned_df, aes(x=age, y=status, color="binned"))
+	+ scale_colour_manual(values=c("grey", "blue", "red", "black"))
+	+ theme(legend.position="bottom")
+)
+```
+
+```{r reff_vareff_wealthindex_cen, echo=FALSE, fig.cap = "Same plots above but with centered confidence bands..."} 
+# Wealth index
+## varpred
+reff_vareff_wealthindex_cen <- varpred(reff_mod, "wealthindex", isolate=TRUE, modelname="Distribution")
+binned_df <- binfun(reff_mod, "wealthindex", "age")
+
+## Pop. average
+### Quantiles
+reff_vareff_wealthindex_quant_cen <- varpred(reff_mod, "wealthindex", isolate=TRUE, pop.ave="quantile", include.re=TRUE, modelname="quantiles")
+### Population
+reff_vareff_wealthindex_pop_cen <- varpred(reff_mod, "wealthindex", isolate=TRUE, pop.ave="population", include.re=TRUE, modelname="Population")
+
+vareff_wealthindex <- reff_vareff_wealthindex_cen
+vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, reff_vareff_wealthindex_quant_cen$preds, reff_vareff_wealthindex_pop_cen$preds))
+vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, reff_vareff_wealthindex_quant_cen$preds, reff_vareff_wealthindex_pop_cen$preds))
+wealthindex_plot_cen <- (plot(vareff_wealthindex)
+ 	+ labs(y="", colour="Model")
+	+ geom_hline(yintercept=true_prop_reff, lty=2, colour="grey")
+	+ geom_point(data=binned_df, aes(x=wealthindex, y=status, color="binned"))
+	+ scale_colour_manual(values=c("grey", "blue", "red", "black"))
+	+ theme(legend.position="bottom")
+)
+
+ggarrange(age_plot_cen, wealthindex_plot_cen, common.legend=TRUE)
+```
+
+\section{Multivariate model}
+
+We now consider a model with multiple predictors.
+
+\subsection{Simple logistic model}
+
+```{r simple_sim_multi}
+N <- 1e4
+beta0 <- 0.7
+betaA <- 0.2
+betaE <- -0.7
+betaW <- 0.5
+age_max <- 1
+age_min <- 0.2
+expend_mean <- 0.2
+expend_sd <- 1
+age <- runif(N, age_min, age_max)
+expenditure <- rnorm(N, expend_mean, expend_sd)
+wealthindex <- rnorm(N, 0, 1)
+eta <- beta0 + betaA * age + betaW * wealthindex + betaE*expenditure
+sim_df <- (data.frame(age=age, expenditure=expenditure, wealthindex=wealthindex, eta=eta)
+	%>% mutate(status = rbinom(N, 1, plogis(eta)))
+	%>% select(-eta)
+)
+true_prop <- mean(sim_df$status)
+print(true_prop)
+```
+
+```{r simple_mod_multi}
+simple_mod_multi <- glm(status ~ age + expenditure + wealthindex, data = sim_df, family="binomial")
+```
+
+```{r simple_vareffects_age_multi, echo=FALSE}
+## varpred way
+simple_vareff_age_multi <- varpred(simple_mod_multi, "age", isolate=TRUE, modelname="Distribution")
+
+## Pop. average
+### Quantiles
+simple_vareff_age_quant_multi <- varpred(simple_mod_multi, "age", isolate=TRUE, pop.ave="quantile", modelname="quantiles")
+### Population
+simple_vareff_age_pop_multi <- varpred(simple_mod_multi, "age", isolate=TRUE, pop.ave="population", modelname="Population")
+binned_df <- binfun(simple_mod_multi, focal = "age", non.focal = c("expenditure", "wealthindex"), bins=50)
+
+vareff_age <- simple_vareff_age_multi
+vareff_age$preds <- do.call("rbind", list(vareff_age$preds, simple_vareff_age_quant_multi$preds, simple_vareff_age_pop_multi$preds))
+age_plot_multi <- (plot(vareff_age)
+ 	+ labs(y="Prob. of improved \n service", colour="Model")
+	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
+	+ geom_point(data=binned_df, aes(x=age, y=status, color="binned"))
+	+ scale_colour_manual(values=c("grey", "blue", "red", "black"))
+	+ theme(legend.position="bottom")
+)
+```
+
+```{r simple_vareffects_expenditure_multi, echo=FALSE}
+## varpred way
+simple_vareff_expenditure_multi <- varpred(simple_mod_multi, "expenditure", isolate=TRUE, modelname="Distribution")
+
+## Pop. average
+### Quantiles
+simple_vareff_expenditure_quant_multi <- varpred(simple_mod_multi, "expenditure", isolate=TRUE, pop.ave="quantile", modelname="quantiles")
+### Population
+simple_vareff_expenditure_pop_multi <- varpred(simple_mod_multi, "expenditure", isolate=TRUE, pop.ave="population", modelname="Population")
+binned_df <- binfun(simple_mod_multi, focal = "expenditure", non.focal = c("age", "wealthindex"), bins=50)
+
+vareff_expenditure <- simple_vareff_expenditure_multi
+vareff_expenditure$preds <- do.call("rbind", list(vareff_expenditure$preds, simple_vareff_expenditure_quant_multi$preds, simple_vareff_expenditure_pop_multi$preds))
+expenditure_plot_multi <- (plot(vareff_expenditure)
+ 	+ labs(y="Prob. of improved \n service", colour="Model")
+	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
+	+ geom_point(data=binned_df, aes(x=expenditure, y=status, color="binned"))
+	+ scale_colour_manual(values=c("grey", "blue", "red", "black"))
+	+ theme(legend.position="bottom")
+)
+```
+
+```{r simple_vareffects_wealthindex_multi, echo=FALSE, fig.cap="A comparison of bias correction approaches for a simple multivariate logistic model."}
+## varpred way
+simple_vareff_wealthindex_multi <- varpred(simple_mod_multi, "wealthindex", isolate=TRUE, modelname="Distribution")
+
+## Pop. average
+### Quantiles
+simple_vareff_wealthindex_quant_multi <- varpred(simple_mod_multi, "wealthindex", isolate=TRUE, pop.ave="quantile", modelname="quantiles")
+### Population
+simple_vareff_wealthindex_pop_multi <- varpred(simple_mod_multi, "wealthindex", isolate=TRUE, pop.ave="population", modelname="Population")
+binned_df <- binfun(simple_mod_multi, focal = "wealthindex", non.focal = c("age", "expenditure"), bins=50)
+
+vareff_wealthindex <- simple_vareff_wealthindex_multi
+vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, simple_vareff_wealthindex_quant_multi$preds, simple_vareff_wealthindex_pop_multi$preds))
+wealthindex_plot_multi <- (plot(vareff_wealthindex)
+ 	+ labs(y="Prob. of improved \n service", colour="Model")
+	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
+	+ geom_point(data=binned_df, aes(x=wealthindex, y=status, color="binned"))
+	+ scale_colour_manual(values=c("grey", "blue", "red", "black"))
+	+ theme(legend.position="bottom")
+)
+ggarrange(age_plot_multi + rremove("ylab")
+	, expenditure_plot_multi + rremove("ylab")
+	, wealthindex_plot_multi + rremove("ylab")
+	, nrow=1, common.legend=TRUE
+)
+```
+
+\subsection{Mixed model}
+
+```{r sim_mixed_multi}
+beta0 <- 0.7
+betaA <- 0.2
+betaW <- 0.5
+betaE <- -0.7
+age_max <- 1
+age_min <- 0.2
+expend_mean <- 0.2
+expend_sd <- 1
+
+# Simulation parameters
+nHH <- 100 # Number of HH (primary units) per year
+nyrs <- 50 # Number of years to simulate
+yrs <- 2000 + c(1:nyrs) # Years to simulate
+N <- nyrs * nHH
+## HH random effect sd
+hhSD <- 0.5
+# Generate dataset template
+temp_df <- (data.frame(hhid = rep(c(1:nHH), each = nyrs)
+		, years = rep(yrs, nHH)
+		, age = runif(n=N, age_min, age_max)
+		, expenditure = rnorm(N, expend_mean, expend_sd)
+		, wealthindex = rnorm(n = N, 0, 1)
+	)
+)
+# Simulate HH-level random effects (residual error)
+hhRE <- rnorm(nHH, hhSD)
+temp_df$hhRE <- hhRE[temp_df$hhid]
+sim_df <- (temp_df
+	%>% mutate(eta = beta0 + betaA * age + betaE * expenditure + betaW * wealthindex + hhRE
+	, status = rbinom(N, 1, plogis(eta))
+	)
+	%>% select(-eta)
+)
+true_prop_reff <- mean(sim_df$status)
+print(true_prop_reff)
+```
+
+```{r mixed_model_multi}
+reff_mod_multi <- glmmTMB(status ~ age + expenditure + wealthindex + (1|hhid)
+	, data = sim_df
+	, family = binomial(link = "logit")
+)
+```
+
+```{r reff_vareffects_age_multi, echo=FALSE}
+## varpred way
+reff_vareff_age_multi <- varpred(reff_mod_multi, "age", isolate=TRUE, modelname="Distribution")
+
+## Pop. average
+### Quantiles
+reff_vareff_age_quant_multi <- varpred(reff_mod_multi, "age", isolate=TRUE, pop.ave="quantile", modelname="quantiles")
+### Population
+reff_vareff_age_pop_multi <- varpred(reff_mod_multi, "age", isolate=TRUE, pop.ave="population", modelname="Population")
+binned_df <- binfun(reff_mod_multi, focal = "age", non.focal = c("expenditure", "wealthindex"), bins=50)
+
+vareff_age <- reff_vareff_age_multi
+vareff_age$preds <- do.call("rbind", list(vareff_age$preds, reff_vareff_age_quant_multi$preds, reff_vareff_age_pop_multi$preds))
+age_plot_multi <- (plot(vareff_age)
+ 	+ labs(y="Prob. of improved \n service", colour="Model")
+	+ geom_hline(yintercept=true_prop_reff, lty=2, colour="grey")
+	+ geom_point(data=binned_df, aes(x=age, y=status, color="binned"))
+	+ scale_colour_manual(values=c("grey", "blue", "red", "black"))
+	+ theme(legend.position="bottom")
+)
+```
+
+```{r reff_vareffects_expenditure_multi, echo=FALSE}
+## varpred way
+reff_vareff_expenditure_multi <- varpred(reff_mod_multi, "expenditure", isolate=TRUE, modelname="Distribution")
+
+## Pop. average
+### Quantiles
+reff_vareff_expenditure_quant_multi <- varpred(reff_mod_multi, "expenditure", isolate=TRUE, pop.ave="quantile", modelname="quantiles")
+### Population
+reff_vareff_expenditure_pop_multi <- varpred(reff_mod_multi, "expenditure", isolate=TRUE, pop.ave="population", modelname="Population")
+binned_df <- binfun(reff_mod_multi, focal = "expenditure", non.focal = c("age", "wealthindex"), bins=50)
+
+vareff_expenditure <- reff_vareff_expenditure_multi
+vareff_expenditure$preds <- do.call("rbind", list(vareff_expenditure$preds, reff_vareff_expenditure_quant_multi$preds, reff_vareff_expenditure_pop_multi$preds))
+expenditure_plot_multi <- (plot(vareff_expenditure)
+ 	+ labs(y="Prob. of improved \n service", colour="Model")
+	+ geom_hline(yintercept=true_prop_reff, lty=2, colour="grey")
+	+ geom_point(data=binned_df, aes(x=expenditure, y=status, color="binned"))
+	+ scale_colour_manual(values=c("grey", "blue", "red", "black"))
+	+ theme(legend.position="bottom")
+)
+```
+
+```{r reff_vareffects_wealthindex_multi, echo=FALSE, fig.cap="A comparison of bias correction approaches for a reff multivariate logistic model."}
+## varpred way
+reff_vareff_wealthindex_multi <- varpred(reff_mod_multi, "wealthindex", isolate=TRUE, modelname="Distribution")
+
+## Pop. average
+### Quantiles
+reff_vareff_wealthindex_quant_multi <- varpred(reff_mod_multi, "wealthindex", isolate=TRUE, pop.ave="quantile", modelname="quantiles")
+### Population
+reff_vareff_wealthindex_pop_multi <- varpred(reff_mod_multi, "wealthindex", isolate=TRUE, pop.ave="population", modelname="Population")
+binned_df <- binfun(reff_mod_multi, focal = "wealthindex", non.focal = c("age", "expenditure"), bins=50)
+
+vareff_wealthindex <- reff_vareff_wealthindex_multi
+vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, reff_vareff_wealthindex_quant_multi$preds, reff_vareff_wealthindex_pop_multi$preds))
+wealthindex_plot_multi <- (plot(vareff_wealthindex)
+ 	+ labs(y="Prob. of improved \n service", colour="Model")
+	+ geom_hline(yintercept=true_prop_reff, lty=2, colour="grey")
+	+ geom_point(data=binned_df, aes(x=wealthindex, y=status, color="binned"))
+	+ scale_colour_manual(values=c("grey", "blue", "red", "black"))
+	+ theme(legend.position="bottom")
+)
+ggarrange(age_plot_multi + rremove("ylab")
+	, expenditure_plot_multi + rremove("ylab")
+	, wealthindex_plot_multi + rremove("ylab")
+	, nrow=1, common.legend=TRUE
+)
+```

### 2021-08-25 bias_correction.rmd
Population averaging seems to work!

@@
 ### Quantiles
-reff_vareff_age_quant_multi <- varpred(reff_mod_multi, "age", isolate=TRUE, pop.ave="quantile", modelname="quantiles")
+reff_vareff_age_quant_multi <- varpred(reff_mod_multi, "age", isolate=TRUE, include.re=TRUE, pop.ave="quantile", modelname="quantiles")
 ### Population
-reff_vareff_age_pop_multi <- varpred(reff_mod_multi, "age", isolate=TRUE, pop.ave="population", modelname="Population")
+reff_vareff_age_pop_multi <- varpred(reff_mod_multi, "age", isolate=TRUE, include.re=TRUE, pop.ave="population", modelname="Population")
 binned_df <- binfun(reff_mod_multi, focal = "age", non.focal = c("expenditure", "wealthindex"), bins=50)
@@
 ### Quantiles
-reff_vareff_expenditure_quant_multi <- varpred(reff_mod_multi, "expenditure", isolate=TRUE, pop.ave="quantile", modelname="quantiles")
+reff_vareff_expenditure_quant_multi <- varpred(reff_mod_multi, "expenditure", isolate=TRUE, pop.ave="quantile", include.re=TRUE, modelname="quantiles")
 ### Population
-reff_vareff_expenditure_pop_multi <- varpred(reff_mod_multi, "expenditure", isolate=TRUE, pop.ave="population", modelname="Population")
+reff_vareff_expenditure_pop_multi <- varpred(reff_mod_multi, "expenditure", isolate=TRUE, pop.ave="population", include.re=TRUE, modelname="Population")
 binned_df <- binfun(reff_mod_multi, focal = "expenditure", non.focal = c("age", "wealthindex"), bins=50)

### 2021-08-26 logist_normal_bias.rmd
working on notation

@@
 ---
+
+```{r setup,include=FALSE}
+```
 
@@
 
-```{python integrate}
+```{python integrate, eval=FALSE}
 from sympy import *
@@
 
+In which I try to figure out what the hell we're actually trying to do.
+
+Notation:
+
+\newcommand{\nset}[1]{#1_{\{n\}}}
+
+- $x_f$: a value of the focal predictor
+- $\nset{x}$: a vector of values of the non-focal predictors for a particular observation
+- $\eta(x_f, \nset{x}) = \beta_f x_f + \sum \nset{\beta} \nset{x}$ = linear predictor (e.g. prediction on the log-odds scale)
+- $g^{-1}()$: inverse-link function (e.g. logistic)
+- $D(\nset{x}|x_f)$: distribution of the non-focal predictors conditional on a particular value of the focal predictor
+- $\beta_{fi}$: the coefficient describing the interaction(s) of the focal and non-focal parameters
+
 The purpose and goal of a *prediction plot* seems fairly straightforward; for specified values of (a) focal predictor(s), we want to give a point estimate and confidence intervals for the prediction of the model for a "typical" (= random sample over the multivariate distribution of non-focal parameters) individual with those values of the predictors ("what is the expected probability of having clean water for a 25-year-old male"?).
@@
 To do these computations, we need to take the values of the non-focal predictors (or their means and covariances) from the *conditional distribution*.
-(If the focal predictors are discrete, we condition on exact values; if they are continuous/have mostly unique values, we condition on appropriately sized bins.)
+(If the focal predictors are discrete, we condition on exact values; if they are continuous/have mostly unique values, we condition on appropriately sized bins.) In other words,
+$$
+\textrm{mean} \left( g^{-1} \eta(D(\nset{x}),x_f) \right)
+$$
 
@@
 Suppose $\hat \beta_f(x_n)$ is the derivative for a particular set of non-focal parameters (i.e.
-$\beta_f + \sum \beta_{fi} x_i$ where the second term is the sum of the relevant interaction coefficients)
+$\beta_f + \sum \beta_{f\nset{i}} \nset{x}$ where the second term is the sum of the relevant interaction coefficients)
 
 $$
-\overline{\frac{d \mu}{dx}} = \int \frac{dg^{-1}(\eta)}{d\eta} \hat \beta_f \, d x_n
+\overline{\frac{d \mu}{dx}} = \int \frac{dg^{-1}(\eta)}{d\eta} \hat \beta_f \, D(\nset{x}|x_f) d \nset{x}
 $$

### 2021-08-30 logist_normal_bias.rmd
fix notation etc.

@@
 \newcommand{\nset}[1]{#1_{\{n\}}}
+\newcommand{\yref}{y_{\textrm{ref}}}
+\newcommand{\cdist}{{D(\nset{x}|x_f)}}
+\newcommand{\cdistprime}{{D(\nset{x}|x_{f'})}}
+\newcommand{\xfprime}{x_{f'}}
 
@@
 - $g^{-1}()$: inverse-link function (e.g. logistic)
-- $D(\nset{x}|x_f)$: distribution of the non-focal predictors conditional on a particular value of the focal predictor
+- $\cdist$: distribution of the non-focal predictors conditional on a particular value of the focal predictor
 - $\beta_{fi}$: the coefficient describing the interaction(s) of the focal and non-focal parameters
@@
 $$
-\textrm{mean} \left( g^{-1} \eta(D(\nset{x}),x_f) \right)
+\underset{\cdist}{\textrm{mean}}  g^{-1} \left(\eta(\nset{x},x_f) \right)
 $$
@@
 Suppose $\hat \beta_f(x_n)$ is the derivative for a particular set of non-focal parameters (i.e.
-$\beta_f + \sum \beta_{f\nset{i}} \nset{x}$ where the second term is the sum of the relevant interaction coefficients)
+$\hat \beta f(x_n) \equiv \beta_f + \sum \beta_{f\nset{i}} \nset{x}$ where the second term is the sum of the relevant interaction coefficients).
+Then I claim that the marginal effect at a particular value of the focal parameter $\xfprime$ is
 
 $$
-\overline{\frac{d \mu}{dx}} = \int \frac{dg^{-1}(\eta)}{d\eta} \hat \beta_f \, D(\nset{x}|x_f) d \nset{x}
+\overline{\frac{d \mu}{dx_f}}(\xfprime) = \int_\cdistprime \frac{dg^{-1}(\eta)}{d\eta} \hat \beta_f \,  d \nset{x}
 $$
 
-This seems to be linear in $x_n$, so I think we get to *not* worry about averaging effects here?
+$\hat \beta_f$ is linear in $\nset{x}$, so nonlinear averaging doesn't affect it, but we do have to worry about weighting the different values of $\hat \beta f$ by the corresponding values of $dg^{-1}(\eta)/d\eta$ (since $\eta$ *also depends on $\nset{x}$*).
 
 It seems that the appropriate computed value (y-axis height) of the effects plot at a given value of $x$ is
+
 $$
-\begin{split}
- & y_{\textrm{ref}} + \int \frac{dg^{-1}(\eta)}{d\eta} \hat \beta_f \, dx_f \\
- = & y_{\textrm{ref}} + \frac{dg^{-1}(\eta)}{d\eta} \overline{\hat \beta_f} \\
- \approx & y_{\textrm{ref}} + \sum \frac{dg^{-1}(\eta(x_f,\bar x_n))}{d\eta} \hat \beta_f(\bar x_n) \, \Delta x_f
-\end{split}
+\yref + \iint_{D(\nset{x}|x_f')} \frac{dg^{-1}(\eta(\xfprime, \nset{x})}{d\eta} \hat \beta_f \, d \nset{x} \, d x_f
 $$
 
-in other words, start from the appropriately bias-corrected value on the response scale at the central value of $x_f$ ($y_{\textrm{ref}}$).
-For every step $\Delta x_f$, calculate the value of the slope (including interactions) at the *mean value of the non-focal parameters conditional on the current value of $x_f$*.
-(If there are no interactions, this reduces to $\beta_f$.) Calculate the linear predictor, again based on the conditional mean value of the non-focal predictors; use it to compute the expansion factor (derivative of the inverse-link function).
-Take the product of the slope, the expansion factor, and $\Delta x_f$; add to the integrand and repeat.
+In the case without interactions $\hat \beta f = \beta_f$ is a constant and we can pull it out of the integrals, but unless the distribution of the non-focal parameters is independent of the focal parameter(s) it looks like we might have to do some relatively complicated stuff.
 
-The good news is that we don't have to fart around with logistic-normal integrals etc.
-except when we are computing $y_{\textrm{\ref}}$.
-The bad news is that the effect plot computed this way will only agree with the prediction plot in special cases ...
+In practice to implement this I would suggest:
+
+- compute $\yref$ by taking $\cdist$ at the center value of the focal parameter and finding the mean (either directly or by computing the mean and variance of $\eta$ and using the logistic-normal calculation
+- for each step away from the center
+   - find $\cdist$ for the new value of $x_f$;
+   - if necessary (i.e. if the model has interactions with the focal parameter) compute $\hat \beta f$ across $\cdist$;
+   - compute the derivative across $\cdist$
+   - compute the mean value of $\phi = dg^{-1}/d\eta \hat \beta f$
+   - add $\phi \Delta x_f$ to the effects-plot value

### 2021-08-30 logist_normal_bias.rmd
more math fixing & description

@@
    - add $\phi \Delta x_f$ to the effects-plot value
+
+## Alternatives
+
+- Do everything by computing means of $g^{-1}$ applied to different values of the focal predictor and the *unconditional* distribution of non-focal parameters (I think this is what Steve is currently doing, at least in some cases, and what JD would prefer?)
+   - if the unconditional distribution is different from (e.g.
+much wider than) the conditional distribution at the central value of $x_f$, then the computed value of $\yref$ won't match the observed marginal probability computed from the data
+- Compute $\yref$ using the conditional distribution for the central value of $x_f$, then compute the increments of $dg^{-1}(\eta)/d\eta$ using the unconditional (population-wide) distribution of $\nset{x}$
+   - you can certainly do this, but it seems incoherent to me based on the discussion above
+- Compute $\yref$ using the conditional distribution for the central value of $x_f$, then compute the increments of $dg^{-1}(\eta)/d\eta$ using the conditional distribution of $\nset{x}$ for the central value of $x_f$
+   - this is coherent but answers an odd question ("what would the predicted response be if we took a typical (central) individual and modified only their focal characteristic, leaving all of their non-focal values the same?").
+We would end up (for example) with a prediction of what we would get if we took a 30-year-old and let them age to 90 without ever changing their wealth or education status ...
+
+I think the key is to start by asking "what is the effects plot supposed to represent?"; I claim it's supposed to represent (in some way) the marginal (economic-sense) effect of the focal parameter on typical individuals at a given value of the focal parameter.
+It might be less confusing to plot the effect itself (e.g.
+put d(probability)/d(age) on the y-axis, rather than probability) &mdash; it would eliminate one of the integrals &mdash; but that's not the convention.

### 2021-08-31 logist_normal_bias.rmd
annotate preferences

--- 2021-08-30 logist_normal_bias.rmd
@@
    - add $\phi \Delta x_f$ to the effects-plot value
+
+(**BB prefers this, JD says let's not do it now**)
+
+## Alternatives
+
+- Do everything by computing means of $g^{-1}$ applied to different values of the focal predictor and the *unconditional* distribution of non-focal parameters (I think this is what Steve is currently doing, at least in some cases, and what JD would prefer?) (**JD's preference, BB is OK with it for now**)
+   - if the unconditional distribution is different from (e.g.
+much wider than) the conditional distribution at the central value of $x_f$, then the computed value of $\yref$ won't match the observed marginal probability computed from the data
+- Compute $\yref$ using the conditional distribution for the central value of $x_f$, then compute the increments of $dg^{-1}(\eta)/d\eta$ using the unconditional (population-wide) distribution of $\nset{x}$ (**JD and BB agree we shouldn't do this**)
+   - you can certainly do this, but it seems incoherent to me based on the discussion above
+- Compute $\yref$ using the conditional distribution for the central value of $x_f$, then compute the increments of $dg^{-1}(\eta)/d\eta$ using the conditional distribution of $\nset{x}$ for the central value of $x_f$ (**JD and BB agree we shouldn't do this**)
+   - this is coherent but answers an odd question ("what would the predicted response be if we took a typical (central) individual and modified only their focal characteristic, leaving all of their non-focal values the same?").
+We would end up (for example) with a prediction of what we would get if we took a 30-year-old and let them age to 90 without ever changing their wealth or education status ...
+
+I think the key is to start by asking "what is the effects plot supposed to represent?"; I claim it's supposed to represent (in some way) the marginal (economic-sense) effect of the focal parameter on typical individuals at a given value of the focal parameter.
+It might be less confusing to plot the effect itself (e.g.
+put d(probability)/d(age) on the y-axis, rather than probability) &mdash; it would eliminate one of the integrals &mdash; but that's not the convention.

### 2021-09-03 aprroximation_bias_correction.rmd
Autocommit

@@
 
-## Population averaging
-pop_none <- genpred(simple_mod, "age", "wealthindex", pop=TRUE, bias.adjust="none", modelname="pop-none")
-pop_delta <- genpred(simple_mod, "age", "wealthindex", pop=TRUE, bias.adjust="delta", modelname="pop-delta")
-
-## Averaged // centered??
-centered_none <- genpred(simple_mod, "age", "wealthindex", pop=FALSE, bias.adjust="none", modelname="centered-none")
-centered_delta <- genpred(simple_mod, "age", "wealthindex", pop=FALSE, bias.adjust="delta", modelname="centered-delta")
-centered_taylor <- genpred(simple_mod, "age", "wealthindex", pop=FALSE, bias.adjust="taylor", modelname="centered-taylor")
-
+### Population averaging
+#pop_none <- genpred(simple_mod, "age", "wealthindex", pop=TRUE, bias.adjust="none", modelname="pop-none")
+#pop_delta <- genpred(simple_mod, "age", "wealthindex", pop=TRUE, bias.adjust="delta", modelname="pop-delta")
+#
+### Averaged // centered??
+#centered_none <- genpred(simple_mod, "age", "wealthindex", pop=FALSE, bias.adjust="none", modelname="centered-none")
+#centered_delta <- genpred(simple_mod, "age", "wealthindex", pop=FALSE, bias.adjust="delta", modelname="centered-delta")
+#centered_taylor <- genpred(simple_mod, "age", "wealthindex", pop=FALSE, bias.adjust="taylor", modelname="centered-taylor")
+#
 ```
@@
 
-```{r simple_preds}
-p1 <- (ggplot(pop_none, aes(x=age, y=fit, colour=model))
-	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
-	+ geom_line()
-	+ geom_line(data=pop_delta, aes(x=age, y=fit, colour=model))
-	+ geom_line(data=centered_none, aes(x=age, y=fit, colour=model))
-	+ geom_line(data=centered_delta, aes(x=age, y=fit, colour=model))
-	+ geom_line(data=centered_taylor, aes(x=age, y=fit, colour=model))
-	+ scale_colour_manual(values=c("blue", "red", "black", "orange", "purple"))
-	+ theme(legend.position="right")
-)
-print(p1)
+```{r simple_preds, echo=FALSE}
+#p1 <- (ggplot(pop_none, aes(x=age, y=fit, colour=model))
+#	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
+#	+ geom_line()
+#	+ geom_line(data=pop_delta, aes(x=age, y=fit, colour=model))
+#	+ geom_line(data=centered_none, aes(x=age, y=fit, colour=model))
+#	+ geom_line(data=centered_delta, aes(x=age, y=fit, colour=model))
+#	+ geom_line(data=centered_taylor, aes(x=age, y=fit, colour=model))
+#	+ scale_colour_manual(values=c("blue", "red", "black", "orange", "purple"))
+#	+ theme(legend.position="right")
+#)
+# print(p1)
 ```

### 2021-09-03 logist_normal_bias.rmd

@@
 Suppose $\hat \beta_f(x_n)$ is the derivative for a particular set of non-focal parameters (i.e.
-$\hat \beta f(x_n) \equiv \beta_f + \sum \beta_{f\nset{i}} \nset{x}$ where the second term is the sum of the relevant interaction coefficients).
+$\hat \beta_f(x_n) \equiv \beta_f + \sum \beta_{f\nset{i}} \nset{x}$ where the second term is the sum of the relevant interaction coefficients).
 Then I claim that the marginal effect at a particular value of the focal parameter $\xfprime$ is
@@
    - find $\cdist$ for the new value of $x_f$;
-   - if necessary (i.e. if the model has interactions with the focal parameter) compute $\hat \beta f$ across $\cdist$;
+   - if necessary (i.e. if the model has interactions with the focal parameter) compute $\hat \beta_f$ across $\cdist$;
    - compute the derivative across $\cdist$
-   - compute the mean value of $\phi = dg^{-1}/d\eta \hat \beta f$
+   - compute the mean value of $\phi = dg^{-1}/d\eta \hat \beta_f$
    - add $\phi \Delta x_f$ to the effects-plot value

### 2021-09-07 bias_correction.rmd
Comparing BB's and JD's approaches

--- 2021-08-10 bias_correction.rmd
@@
 ```{r setup, include=FALSE, echo=FALSE}
-knitr::opts_chunk$set(fig.pos = "H", out.extra = "")
+knitr::opts_chunk$set(fig.pos = "H", out.extra = "", message=FALSE, warning=FALSE)
 
+library(shellpipes)
 library(glmmTMB)
@@
 library(ggpubr)
-set.seed(1011)
 
+commandEnvironments()
+makeGraphics()
+
+set.seed(2121)
 ## Functions
@@
 
-Prediction of the variation in the response variable depends on whether the relationship between the response variable and the predictors is linear or nonlinear.
-For example, when response, $Y$, changes nonlinearly with the predictor variable, $X$, averaged response variable with respect to the predictor, $E(Y(X))$, does not necessarily equal the response at the mean of the predictor, $Y(E(X)$.
+Generating predictions from regression models can be challenging and depends on whether the relationship between the response variable and the predictors is linear or nonlinear.
+For example, if response, $Y$, changes nonlinearly with the predictor variable, $X$, averaged response variable with respect to the predictor, $E(Y(X))$, does not necessarily equal the response at the mean of the predictor, $Y(E(X)$.
 This can be understood in relation to Jensen's inequality which states that, for a nonlinear function, $Y(X)$, then $E(Y(X)) > Y(E(X))$, if $Y(X)$ is positive second derivative; and $E(Y(X)) < Y(E(X))$ if $Y(X)$ is negative second derivative.
-Most packages for predicting responses make distribution assumptions about the predictors, (for example, conditioning at the mean value of the predictor), leading to potential biasness if the particular predictor is not well represented, or as a result of nonlinear averaging.
-We consider the following approaches for bias correction:
+Currently, existing R packages for predicting responses make distribution assumptions about the non-focal predictors, (for example, averaging the non-focal predictors at various levels of focal predictors), leading to potential biasness if the particular predictor is not well represented, or as a result of nonlinear averaging.
+To understand what we are trying to achieve, we first define the following:
 
-- Population averaging
+\subsection{Prediction plots}
 
-	- Whole population
-	- Quantiles
+Notation :
 
-- Distributional conditioning
-- Second-order correction
+\newcommand{\nset}[1]{#1_{\{n\}}}
+\newcommand{\yref}{y_{\textrm{ref}}}
+\newcommand{\cdist}{{D(\nset{x}|x_f)}}
+\newcommand{\cdistprime}{{D(\nset{x}|x_{f'})}}
+\newcommand{\xfprime}{x_{f'}}
 
-We implement and apply these methods in the context of both simple generalized linear and mixed effect models, using simulated data sets.
+- $x_f$: a value of the focal predictor
+- $\nset{x}$: a vector of values of the non-focal predictors for a particular observation
+- $\eta(x_f, \nset{x}) = \beta_f x_f + \sum \nset{\beta} \nset{x}$ = linear predictor (e.g. prediction on the log-odds scale)
+- $g^{-1}()$: inverse-link function (e.g. logistic)
+- $\cdist$: distribution of the non-focal predictors conditional on a particular value of the focal predictor
+- $\beta_{fi}$: the coefficient describing the interaction(s) of the focal and non-focal parameters
+
+The purpose and goal of a *prediction plot* seems fairly straightforward; for specified values of (a) focal predictor(s), we want to give a point estimate and confidence intervals for the prediction of the model for a "typical" (= random sample over the multivariate distribution of non-focal parameters) individual with those values of the predictors.
+
+**BB**'s
+
+("what is the expected probability of having clean water for a 25-year-old male"?).
+These are most of the problems we have actually been addressing with our bias correction stuff above.
+To do these computations, we need to take the values of the non-focal predictors (or their means and covariances) from the *conditional distribution*.
+(If the focal predictors are discrete, we condition on exact values; if they are continuous/have mostly unique values, we condition on appropriately sized bins.) In other words,
+$$
+\underset{\cdist}{\textrm{mean}}  g^{-1} \left(\eta(\nset{x},x_f) \right)
+$$
+
+Suppose we consider the values of non-focal predictors, to implement this:
+
+- compute linear predictor of the non-focal predictors, $\nset{\eta} = \sum \nset{\beta} \nset{x}$
+- find a list of vectors of observations of $\nset{\eta}$ associated with each value (bin) of the focal predictor, $\nset{{\eta_j}}$, $j = 1, 2, \cdots$
+- for each $\nset{{\eta_j}}$:
+	- compute $\hat{y}_j  = \textrm{mean} ~ g^{-1} \left(\beta_f x_{j_f} + \nset{{\eta_j}}\right)$
+
+_What's the expected probability of having clean water for all $x$-year-old with the **corresponding** wealth index??_
+
+**JD**'s and **SC**'s
+
+- compute linear predictor of the non-focal predictors, $\nset{\eta} = \sum \nset{\beta} \nset{x}$
+- for every value of the focal predictor, $x_{j_f}$:
+	- compute $\hat{y}_j  = \textrm{mean} ~ g^{-1} \left(\beta_f x_{j_f} + \nset{\eta}\right)$
+
+The major difference between **BB**'s and **JD**'s approach is that in the former case, we consider non-focal linear predictor corresponding to specific value (bin) of the focal predictor, while in the later case, for each value of the focal predictor, we consider linear predictor associated with entire population.
+
+**Questions**
+
+- What does the second case (JD's) represent?
+- Which is the correct approach?
+
+\subsection{Simulation}
+
+We implement and apply these methods in the context of both simple generalized linear and mixed effect models, using simulated data sets -- for univariate and multivariate models.
+
+We start with a univariate case where we have only one predictor.
 
@@
 \mathrm{Wealthindex} &\sim \mathrm{Normal}(0, 1) \\
-\beta_0 &= 0.7 \\
-\beta_{\mathrm{A}} &= 0.2 \\
-\beta_{\mathrm{W}} &= 0.5
+\beta_0 &= 1.5 \\
+\beta_{\mathrm{A}} &= 1.0 \\
+\beta_{\mathrm{W}} &= 2
 \end{align*}
 
+Considering $N=1e4$
+
 ```{r sim_simple}
-N <- 1e4
-beta0 <- 0.7
-betaA <- 0.2
-betaW <- 0.5
 
-age_max <- 1
-age_min <- 0.2
-age <- runif(N, age_min, age_max)
-# age <- rnorm(N, age_max, age_max)
-
-wealthindex <- rnorm(N, 0, 1)
-
+simplesim <- function(N=1e4, beta0=1.5, betaA=1.0, betaW=2
+		, age_sd=1, age_mean=0.2
+		, wealth_sd=1, wealth_mean=0
+	) {
+	age <- rnorm(N, age_mean, age_sd)
+	wealthindex <- rnorm(N, wealth_mean, wealth_sd)
 	eta <- beta0 + betaA * age + betaW * wealthindex
@@
 	)
+	return(sim_df)
+}
+sim_df <- simplesim()
 true_prop <- mean(sim_df$status)
@@
 
-Coefficient plots
-
-```{r simple_coef, fig.cap="Coefficient plot for simple logistic model."}
-## True beta
-true_beta_df <- data.frame(term=c("Intercept", "age", "wealthindex")
-	, estimate=c(beta0, betaA, betaW)
-)
-
-## Tidy coef estimates
-coef_df <- (broom::tidy(simple_mod, conf.int=TRUE)
-#	%>% dotwhisker::by_2sd(sim_df)
-	%>% mutate(term = gsub("\\(|\\)", "", term))
-)
-print(coef_df)
-
-simple_coef_plot <- (plotEsize(coef_df) 
-	+ geom_point(data=true_beta_df, aes(x=term, y=estimate, colour="Truth"))
-	+ labs(colour="Type")
-	+ scale_colour_manual(values=c("black", "blue"))
-)
-print(simple_coef_plot)
-```
-
-Effect sizes on logit scale:
-
-```{r simple_logit}
-coef_df_logit <- (coef_df
-	%>% select(term, estimate, conf.low, conf.high)
-	%>% group_by(term)
-	%>% summarise_all(plogis)
-)
-print(coef_df_logit)
-```
-
 \subsection{Variable predictions}
 
-We consider both \textbf{varpred} and population averaging approach; and then introduce bias correction to \textbf{varpred} predictions.
-
 ```{r pop_ave_fun, echo=FALSE}
-popavefun <- function(mod, focal, non.focal, level=0.95, steps=100, modelname="quantiles", ...) {
-	mf <- model.frame(mod)
-	mm <- (mf
-		%>% select_at(c(focal, non.focal))
-	)
-	quant <- seq(0, 1, length.out=steps)
-	mm <- sapply(c(focal, non.focal), function(x)quantile(mm[,x], quant), simplify = FALSE)
-	mm <- do.call("expand.grid", mm)
-	vc <- vcov(mod)
-	mm2 <- model.matrix(formula(mod)[c(1,3)], mm)
-	linpred <- as.vector(mm2 %*% coef(mod))
-	pse_var <- sqrt(rowSums(mm2 * t(tcrossprod(data.matrix(vc), mm2))))
-	z.val <- qnorm(1 - (1 - level)/2)
-	pred_df <- (mm
-		%>% select_at(focal)
-		%>% mutate(fit = linpred
-			, lwr = plogis(fit - z.val*pse_var)
-			, upr = plogis(fit + z.val*pse_var)
-			, fit = plogis(fit)
-		)
-		%>% group_by_at(focal)
-		%>% summarise_at(c("fit", "lwr", "upr"), mean)
-		%>% mutate(model = modelname, se=NA)
-	)
-	return(pred_df)
-}
-
 binfun <- function(mod, focal, non.focal, bins=50, ...) {
@@
 
-- Age
+```{r simple_pred_age, echo=FALSE, fig.cap="Prediction plots with N=1e4: JD's approach on the left; BB's on the right."} 
+binned_df <- binfun(simple_mod, "age", "wealthindex")
 
-```{r simple_vareffects_age} 
-## varpred way
-simple_vareff_age <- varpred(simple_mod, "age", isolate=FALSE, modelname="Distribution")
-
-## Pop. average
-### Quantiles
-simple_vareff_age_quant <- varpred(simple_mod, "age", isolate=FALSE, pop.ave="quantile", modelname="quantiles")
-### Population
-simple_vareff_age_pop <- varpred(simple_mod, "age", isolate=FALSE, pop.ave="population", modelname="Population")
-binned_df <- binfun(simple_mod, "age", "wealthindex", bins=15)
-
-vareff_age <- simple_vareff_age
-vareff_age$preds <- do.call("rbind", list(vareff_age$preds, simple_vareff_age_quant$preds, simple_vareff_age_pop$preds))
-age_plot <- (plot(vareff_age)
- 	+ labs(y="Prob. of improved \n service", colour="Model")
+## JD's approach
+pred_age_jd <- predictionfun(simple_mod, "age", isolate=TRUE, include.nonfocal="jd", modelname="JD")
+age_plot_jd <- (ggplot(pred_age_jd, aes(x=age, y=fit))
+	+ geom_line()
+	+ geom_line(aes(y=lwr), lty=2)
+	+ geom_line(aes(y=upr), lty=2)
 	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
 	+ geom_point(data=binned_df, aes(x=age, y=status, color="binned"))
-	+ scale_colour_manual(values=c("grey", "blue", "red", "black"))
-	+ theme(legend.position="bottom")
+	+ labs(y="Probability")
+	+ theme(legend.position="none")
 )
+
+## BB's approach
+pred_age_bb <- predictionfun(simple_mod, "age", isolate=TRUE, include.nonfocal="bb", modelname="BB")
+age_plot_bb <- (ggplot(pred_age_bb, aes(x=age, y=fit))
+	+ geom_line()
+	+ geom_line(aes(y=lwr), lty=2)
+	+ geom_line(aes(y=upr), lty=2)
+	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
+	+ geom_point(data=binned_df, aes(x=age, y=status, color="binned"))
+	+ geom_smooth(colour="black", size=0.5)
+	+ labs(y="Probability")
+	+ theme(legend.position="none")
+)
+
+ggarrange(age_plot_jd, age_plot_bb + rremove("ylab"), common.legend=FALSE)
+
 ```
 
-- Wealth index
+Considering $N=1e5$ and then refit the model
 
-```{r simple_vareff_wealthindex, fig.cap = "A comparison of various bias-correction approaches -- Distribution approach is based on averaging of non-focal predictors; Population approach involves using the observed values of non-focal predictors together with the quantiles of the focal predictors; while quantiles approach involves sampling (or picking quantiles of) the observed values of both focal and non-focal predictors.
-Except for the distribution approach, population and quantiles approaches give very close estimates."}
-# Wealth index
-## varpred 
-simple_vareff_wealthindex <- varpred(simple_mod, "wealthindex", isolate=FALSE, modelname="Distribution")
+```{r sim_simple_ne5}
+sim_df <- simplesim(N=1e5)
+true_prop <- mean(sim_df$status)
+simple_mod <- glm(status ~ age + wealthindex, data = sim_df, family="binomial")
+```
 
-## Pop. average
-### Quantiles
-simple_vareff_wealthindex_quant <- varpred(simple_mod, "wealthindex", isolate=FALSE, pop.ave="quantile", modelname="quantiles")
-### Population
-simple_vareff_wealthindex_pop <- varpred(simple_mod, "wealthindex", isolate=FALSE, pop.ave="population", modelname="Population")
-binned_df <- binfun(simple_mod, "wealthindex", "age")
+```{r simple_pred_age_ne5, echo=FALSE, fig.cap="Prediction plots with N=1e5: JD's approach on the left; BB's on the right."} 
+binned_df <- binfun(simple_mod, "age", "wealthindex")
 
-vareff_wealthindex <- simple_vareff_wealthindex
-vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, simple_vareff_wealthindex_quant$preds, simple_vareff_wealthindex_pop$preds))
-wealthindex_plot <- (plot(vareff_wealthindex)
- 	+ labs(y="", colour="Model")
+## JD's approach
+pred_age_jd <- predictionfun(simple_mod, "age", isolate=TRUE, include.nonfocal="jd", modelname="JD")
+age_plot_jd <- (ggplot(pred_age_jd, aes(x=age, y=fit))
+	+ geom_line()
+	+ geom_line(aes(y=lwr), lty=2)
+	+ geom_line(aes(y=upr), lty=2)
 	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
-	+ geom_point(data=binned_df, aes(x=wealthindex, y=status, color="binned"))
-	+ scale_colour_manual(values=c("grey", "blue", "red", "black"))
-	+ theme(legend.position="bottom")
+	+ geom_point(data=binned_df, aes(x=age, y=status, color="binned"))
+	+ labs(y="Probability")
+	+ theme(legend.position="none")
 )
 
-ggarrange(age_plot, wealthindex_plot, common.legend=TRUE)
-```
-
-The observed population average is `r true_prop` while the estimated population averages (similar estimates with `varpred`) are:
-
-- age: quantiles `r mean(simple_vareff_age_quant$preds$fit)`; population `r mean(simple_vareff_age_pop$preds$fit)`; distribution `r mean(simple_vareff_age$preds$fit)`
-- wealthindex: quantiles `r mean(simple_vareff_wealthindex_quant$preds$fit)`; population `r mean(simple_vareff_wealthindex_pop$preds$fit)`; distribution `r mean(simple_vareff_wealthindex$preds$fit)`
-
-\section{Random effect model}
-
-```{r sim_randeff}
-# Simulation parameters
-nHH <- 100	# Number of HH (primary units) per year
-
-nyrs <- 50	# Number of years to simulate
-yrs <- 2000 + c(1:nyrs) # Years to simulate
-N <- nyrs * nHH
-
-## HH random effect sd
-hhSD <- 0.5
-
-# Generate dataset template
-temp_df <- (data.frame(hhid = rep(c(1:nHH), each = nyrs)
-		, years = rep(yrs, nHH)
-		, age = runif(n=N, age_min, age_max)
-		, wealthindex = rnorm(n = N, 0, 1)
-	)
+## BB's approach
+pred_age_bb <- predictionfun(simple_mod, "age", isolate=TRUE, include.nonfocal="bb", modelname="BB")
+age_plot_bb <- (ggplot(pred_age_bb, aes(x=age, y=fit))
+	+ geom_line()
+	+ geom_line(aes(y=lwr), lty=2)
+	+ geom_line(aes(y=upr), lty=2)
+	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
+	+ geom_point(data=binned_df, aes(x=age, y=status, color="binned"))
+	+ geom_smooth(colour="black", size=0.5)
+	+ labs(y="Probability")
+	+ theme(legend.position="none")
 )
 
-# Simulate HH-level random effects (residual error)
-hhRE <- rnorm(nHH, hhSD)
-temp_df$hhRE <- hhRE[temp_df$hhid]
+ggarrange(age_plot_jd, age_plot_bb + rremove("ylab"), common.legend=FALSE)
 
-sim_df <- (temp_df
-	%>% mutate(eta = beta0 + betaA * age + betaW * wealthindex + hhRE
-		, status = rbinom(N, 1, plogis(eta))
-	)
-	%>% select(-eta)
-)
-true_prop_reff <- mean(sim_df$status)
-print(true_prop)
-print(head(sim_df, 10))
 ```
-
-\subsection{Fit model}
-
-```{r model_reff}
-reff_mod <- glmmTMB(status ~ age + wealthindex + (1|hhid)
-	, data = sim_df
-	, family = binomial(link = "logit")
-)
-```
-
-```{r reff_coef, fig.cap="Mixed model coefficient estimates."}
-## Tidy coef estimates
-reff_coef_df <- (broom.mixed::tidy(reff_mod, conf.int=TRUE)
-	%>% mutate(term = gsub("\\(|\\)", "", term))
-	%>% filter(effect=="fixed")
-)
-print(reff_coef_df)
-
-reff_coef_plot <- (plotEsize(reff_coef_df) 
-	+ geom_point(data=true_beta_df, aes(x=term, y=estimate, colour="Truth"))
-	+ labs(colour="Type")
-	+ scale_colour_manual(values=c("black", "blue"))
-)
-print(reff_coef_plot)
-```
-
-Variable effect plots
-
-- Age
-
-```{r reff_vareffects_age}
-## varpred way
-reff_vareff_age <- varpred(reff_mod, "age", isolate=FALSE, modelname="Distribution")
-
-## Pop. average
-### Quantiles
-reff_vareff_age_quant <- varpred(reff_mod, "age", isolate=FALSE, pop.ave="quantile", include.re=TRUE, modelname = "quantiles")
-### Population
-reff_vareff_age_pop <- varpred(reff_mod, "age", isolate=FALSE, pop.ave="population", include.re=TRUE, modelname = "Population")
-binned_df <- binfun(reff_mod, "age", "wealthindex", bins=15)
-
-vareff_age <- reff_vareff_age
-vareff_age$preds <- do.call("rbind", list(vareff_age$preds, reff_vareff_age_quant$preds, reff_vareff_age_pop$preds))
-age_plot <- (plot(vareff_age)
- 	+ labs(y="Prob. of improved \n service", colour="Model")
-	+ geom_hline(yintercept=true_prop_reff, lty=2, colour="grey")
-	+ geom_point(data=binned_df, aes(x=age, y=status, color="binned"))
-	+ scale_colour_manual(values=c("grey", "blue", "red", "black"))
-	+ theme(legend.position="bottom")
-)
-```
-
-- Wealth index
-
-```{r reff_vareff_wealthindex, fig.cap = "Distribution based approach over-estimates the predictions in age and wealth index."} 
-# Wealth index
-## varpred
-reff_vareff_wealthindex <- varpred(reff_mod, "wealthindex", isolate=FALSE, modelname="Distribution")
-binned_df <- binfun(reff_mod, "wealthindex", "age")
-
-## Pop. average
-### Quantiles
-reff_vareff_wealthindex_quant <- varpred(reff_mod, "wealthindex", isolate=FALSE, pop.ave="quantile", include.re=TRUE, modelname="quantiles")
-### Population
-reff_vareff_wealthindex_pop <- varpred(reff_mod, "wealthindex", isolate=FALSE, pop.ave="population", include.re=TRUE, modelname="Population")
-
-vareff_wealthindex <- reff_vareff_wealthindex
-vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, reff_vareff_wealthindex_quant$preds, reff_vareff_wealthindex_pop$preds))
-vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, reff_vareff_wealthindex_quant$preds, reff_vareff_wealthindex_pop$preds))
-wealthindex_plot <- (plot(vareff_wealthindex)
- 	+ labs(y="", colour="Model")
-	+ geom_hline(yintercept=true_prop_reff, lty=2, colour="grey")
-	+ geom_point(data=binned_df, aes(x=wealthindex, y=status, color="binned"))
-	+ scale_colour_manual(values=c("grey", "blue", "red", "black"))
-	+ theme(legend.position="bottom")
-)
-
-ggarrange(age_plot, wealthindex_plot, common.legend=TRUE)
-```

### 2021-09-14 bias_correction.rmd
Autocommit

@@
     extra_dependencies: ["float"]
+bibliography: bias_correction.bib
+link-citations: yes
 ---
@@
 }
+
+simplesim <- function(N=1e4, beta0=1.5, betaA=1.0, betaW=2
+		, age_sd=1, age_mean=0.2
+		, wealth_sd=1, wealth_mean=0
+		, link_scale = FALSE
+	) {
+	age <- rnorm(N, age_mean, age_sd)
+	wealthindex <- rnorm(N, wealth_mean, wealth_sd)
+	eta <- beta0 + betaA * age + betaW * wealthindex
+	if (link_scale) {
+		eta <- eta + rnorm(N) 
+	}
+	sim_df <- data.frame(age=age, wealthindex=wealthindex, eta=eta)
+	if (!link_scale) {
+		sim_df <- (sim_df
+			%>% mutate(status = rbinom(N, 1, plogis(eta)))
+			%>% select(-eta)
+		)
+	}
+	return(sim_df)
+}
 ```
-
-\section{Introduction}
-
-Generating predictions from regression models can be challenging and depends on whether the relationship between the response variable and the predictors is linear or nonlinear.
-For example, if response, $Y$, changes nonlinearly with the predictor variable, $X$, averaged response variable with respect to the predictor, $E(Y(X))$, does not necessarily equal the response at the mean of the predictor, $Y(E(X)$.
-This can be understood in relation to Jensen's inequality which states that, for a nonlinear function, $Y(X)$, then $E(Y(X)) > Y(E(X))$, if $Y(X)$ is positive second derivative; and $E(Y(X)) < Y(E(X))$ if $Y(X)$ is negative second derivative.
-Currently, existing R packages for predicting responses make distribution assumptions about the non-focal predictors, (for example, averaging the non-focal predictors at various levels of focal predictors), leading to potential biasness if the particular predictor is not well represented, or as a result of nonlinear averaging.
-To understand what we are trying to achieve, we first define the following:
-
-\subsection{Prediction plots}
-
-Notation :
 
@@
 \newcommand{\xfprime}{x_{f'}}
+
+\newcommand{\bX}{{\mathbf X}}
+\newcommand{\bbeta}{{\boldsymbol \beta}}
+\newcommand{\boldeta}{{\boldsymbol \eta}}
+
+Notation :
 
@@
 
-The purpose and goal of a *prediction plot* seems fairly straightforward; for specified values of (a) focal predictor(s), we want to give a point estimate and confidence intervals for the prediction of the model for a "typical" (= random sample over the multivariate distribution of non-focal parameters) individual with those values of the predictors.
+_\
 
-**BB**'s
+\section{Introduction}
 
-("what is the expected probability of having clean water for a 25-year-old male"?).
-These are most of the problems we have actually been addressing with our bias correction stuff above.
+Generating variable predictions from regression models can be challenging and depends on whether the relationship between the response variable and the predictors is linear or nonlinear.
+For example, predictions from generalized linear (mixed) models (GLM(M)s) may be inaccurate due to a number of reasons:
+\begin{enumerate}
+\item choice of representative values of \emph{focal} variable(s) and appropriate 'model center' for \emph{non-focal} variables in the case of multiple predictors
+\item Jensen’s inequality and bias in the mean induced by nonlinear transformations of the response variable (e.g., link functions in GLM(M)s)
+\item  propagation of uncertainty ([how] can we incorporate uncertainty in nonlinear components of (G)LMMs? Should we exclude variation in non-focal parameters?)
+\end{enumerate}
+
+Most common way of dealing with the first challenge is taking the taking unique levels of the _focal_ variable(s) if they are discrete or taking appropriately sized quantiles (or bins) if they are continuous, and then conditioning these values on the mean of the non-focal predictors.
+We propose an alternative approach, the population-based approach.
+The two approaches result in predictions corresponding to each level of the focal predictor i.e., \emph{variable prediction}.
+Second and third challenges are the focus of this article.
+
+To get an intuition of how conditioning on the mean values of the non-focal predictors work, suppose we are interested in the variable predictions of a particular predictor (hence forth referred as \emph{focal} predictor otherwise \emph{non-focal}), $x_f$, from the set of predictors.
+To keep it simple, assume that the model has no interaction terms.
+Then the idea is to fix the values of \emph{non-focal} predictor(s) at some typical values -- typically determined by averaging (for now) in some meaningful way, for example, arithmetic mean and average over the levels of the factors of \emph{non-focal} continuous and categorical predictors, respectively.
+An alternative to \emph{averaging} is \emph{anchoring} which involves picking a fixed value of the non-focal predictor.
+One way to achieve this is by averaging the columns of non-focal terms in model matrix, $\bX$.
+For a simple linear model with identity link function, the estimated variable predictions corresponding to focal predictor, $x_f$, is $\eta(x_f, \nset{{\bar{x}}}) = \beta_f x_f + \sum \nset{\beta} \nset{{\bar{x}}}$, where $\nset{{\bar{x}}}$ are the appropriately averaged entries of non-focal predictors.
+Almost all existing \textbf{R} packages for constructing predictions employ this kind of averaging of the non-focal predictors across the levels of the focal predictors [@lenth2018package; @leeper2017package; @fox2009effect].
+More complicated models and link functions are described in the subsequent sections.
+
+\subsection{Variable prediction and prediction plot}
+
+Variable predictions are individual predictions corresponding to the levels (or unique values) of focal predictor(s) conditioned on the "typical" values of the non-focal predictors.
+Prediction plot provides a way to describe variable predictions.
+In particular, the purpose and goal of a *prediction plot* seems fairly straightforward; for specified values of (a) focal predictor(s), we want to give a point estimate and confidence intervals for the prediction of the model for a "typical" (= random sample over the multivariate distribution of non-focal parameters) individual with those values of the predictors.
+
+\subsection{Jensen’s inequality and bias in the expected mean}
+
+Nonlinear relationship between the response and independent variables can either increase or decrease the predicted outcome, depending on the shape of the link function.
+A key challenge is understanding how the variability in the focal predictor(s) affect the (mean) predictions of the outcome.
+Classically, Jensen's inequality provides a way to examine the nature of the link function and deducing its effect on the prediction.
+Specifically, for a random variable $x$ with mean of $\bar{x}$; a nonlinear function, $f(x)$, the mean of $f(x)$, $\bar{f(x)}$, does not equal the the nonlinear function applied to mean of $x$, $f(\bar{x})$.
+When $f(x)$ is accelerating ($f''(x) > 0$), $\bar{f(x)} > f(\bar{x})$; and when $f(x)$ is decreasing ($f''(x) < 0$), $\bar{f(x)} < f(\bar{x})$.
+In other words, the sign of the difference between $\bar{f(x)}$ and $f(\bar{x})$ depends on the nature of the link function.
+
+For accelerating functions, the Jensen's inequality describes how the changes in the predictors elevate the predicted outcome and describes how these changes depress the predictions.
+**SC -> JD: graphical illustration?**
+
+In many applications, it usually important to report the estimates that reflect the expected values of the untransformed respense.
+In such cases, bias-adjustment is needed when back-transforming the predictions to the original scales, due to the biases induced by the nonlinear transformation on the expected mean.
+More specifically, suppose the transformed response is $\eta$, the back-transformed response is $y = g^{-1}(\eta)$.
+Most common approach for bias-adjustment is second-order Taylor approximation [@lenth2018package; @duursma2003bias].
+Here, we describe and implement a different approach, \emph{population-based} approach for bias correction.
+
+\subsubsection{Population-based approach for bias correction}
+
+The most precise (although not necessarily accurate!) way to predict is to condition on a value $F$ of the focal predictor and make predictions for all observations (members of the population) for which $x_f = F$ (or in a small range around $F$ ...).
+A key point is that the nonlinear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population (e.g.
+all individuals with $x_f = F$) to a one-dimensional distribution of $\eta(x_f, \nset{x})$.
+
+Once we have got our vector of $\boldeta$ (which is essentially a set of samples from the distribution over $\eta$ for the conditional set), we want a mean and confidence intervals on the mean on the response (data) scale, i.e.
+after back-transforming.
+Most precisely the mean is $\int P(\eta') g^{-1}(\eta') d\,\eta$:
+\begin{enumerate}
+\item if we use the observations themselves then we just compute the individual values of $g^{-1}(\eta)$ and compute their mean
+\item we could compute the quantiles of the distribution of $\eta$ and use this to construct an approximate Riemann sum over the distribution
+\end{enumerate}
+
+In principle, the first approach is more applicable, hence our focus, in our case since  we are interested in individual (at each level of the focal predictor) predictions i.e., the \emph{variable prediction}, not necessarily the mean.
+In addition, we can conveniently compute the mean of $g^{-1}(\eta)$.
+
+There are two ways to compute $\eta$ when we use the individuals observations, i.e., the first case. In particular for each level of the focal predictor, we can take
+\begin{enumerate}
+\item corresponding values of the non-focal linear (hence forth referred to as binned non-focal linear predictor // \textbf{BB})
+\item entire population of the non-focal linear predictor (hence forth referred to as population-based non-focal linear predictor // \textbf{JD})
+\end{enumerate}
+
+\subsubsection{Binned non-focal linear predictor}
+
 To do these computations, we need to take the values of the non-focal predictors (or their means and covariances) from the *conditional distribution*.
-(If the focal predictors are discrete, we condition on exact values; if they are continuous/have mostly unique values, we condition on appropriately sized bins.) In other words,
+If the focal predictors are discrete, we condition on exact values; if they are continuous/have mostly unique values, we condition on appropriately sized bins.
+In other words,
 $$
@@
 
-Suppose we consider the values of non-focal predictors, to implement this:
+To implement this:
+\begin{itemize}
+\item compute linear predictor of the non-focal predictors, $\nset{\eta} = \sum \nset{\beta} \nset{x}$
+\item find a list of vectors of observations of $\nset{\eta}$ associated with each value (bin) of the focal predictor, $\nset{{\eta_j}}$, $j = 1, 2, \cdots$
+\item for each $\nset{{\eta_j}}$:
+\begin{itemize}
+\item compute $\hat{y}_j  = \textrm{mean} ~ g^{-1} \left(\beta_f x_{j_f} + \nset{{\eta_j}}\right)$
+\end{itemize}
+\end{itemize}
 
-- compute linear predictor of the non-focal predictors, $\nset{\eta} = \sum \nset{\beta} \nset{x}$
-- find a list of vectors of observations of $\nset{\eta}$ associated with each value (bin) of the focal predictor, $\nset{{\eta_j}}$, $j = 1, 2, \cdots$
-- for each $\nset{{\eta_j}}$:
-	- compute $\hat{y}_j  = \textrm{mean} ~ g^{-1} \left(\beta_f x_{j_f} + \nset{{\eta_j}}\right)$
+Consider an example where we want to predict the probability of having clean water based on age and gender of household head, the prediction $\hat{y}_j$ would represent the expected probability of clean water for a 25-year-old, for example.
 
-_What's the expected probability of having clean water for all $x$-year-old with the **corresponding** wealth index??_
+If we compute the individual back-transformed predictions for a poorly sampled/finely spaced set of focal values, we will get a noisy prediction line as the values of the non-focal predictors shift across the focal values.
+Simple example: suppose everyone below the median age has wealth index $w_1$, everyone above the median has $w_2$.
+Then the predicted value will have a discontinuity at the median age.
+We can deal with this by taking bigger bins (a form of smoothing), or by post-smoothing the results (by loess, for example).
+The principled form of this would be to assume/recognize that our uneven distribution of observed non-focal predictors actually represents a sample of a distribution that will vary \emph{smoothly} as a function of the focal predictor.
 
-**JD**'s and **SC**'s
+\subsubsection{Whole population non-focal linear predictor}
 
-- compute linear predictor of the non-focal predictors, $\nset{\eta} = \sum \nset{\beta} \nset{x}$
-- for every value of the focal predictor, $x_{j_f}$:
-	- compute $\hat{y}_j  = \textrm{mean} ~ g^{-1} \left(\beta_f x_{j_f} + \nset{\eta}\right)$
-
-The major difference between **BB**'s and **JD**'s approach is that in the former case, we consider non-focal linear predictor corresponding to specific value (bin) of the focal predictor, while in the later case, for each value of the focal predictor, we consider linear predictor associated with entire population.
+Suppose we are now interested in \emph{expected probability of having clean water for all $x$-year-old across the levels of gender??}, then at each level of the focal predictor, we want to add overall contribution of non-focal predictors.
+In particular:
+\begin{itemize}
+\item compute linear predictor of the non-focal predictors, $\nset{\eta} = \sum \nset{\beta} \nset{x}$
+\item for every value of the focal predictor, $x_{j_f}$:
+\begin{itemize}
+\item compute $\hat{y}_j  = \textrm{mean} ~ g^{-1} \left(\beta_f x_{j_f} + \nset{\eta}\right)$
+\end{itemize}
+\end{itemize}
 
 **Questions**
+\begin{itemize}
+\item How to conceptualize \textbf{JD} approach. What does it represent in reality?
+\item Which approach is the most appropriate?
+\end{itemize}
 
-- What does the second case (JD's) represent?
-- Which is the correct approach?
+\subsection{Propagation of uncertainty}
 
-\subsection{Simulation}
+What about the confidence intervals (CI)?
+The limits of the confidence intervals are points, not mean values.
+In principle, every observation/set of non-focal predictors has a different CI.
+The predictions are $\eta(x_f, \nset{x})$ or $\eta(x_f, \nset{{\bar{x}}})$, depending on how we treat non-focal predictors; the variances of the predictions are $\sigma^2 = \textrm{Diag}(\bX^\star \Sigma \bX^{\star\top})$, where the entries in $\bX^\star$ constitutes appropriately constructed (as per $\eta$) non-focal predictors (or corresponding terms) together with appropriate values of the non-focal predictors; and $\Sigma = V(\bbeta)$ is variance-covariance matrix of $\bbeta$.
+If the non-focal predictors are averaged, the variances are directly computed from $\bX^\star$.
+However, for the population-based approaches, we follow the same steps described above to compute the distributions of lower/upper CI, $\eta \pm q\sigma$, and compute the transformed mean values, i.e.,
+$$
+\underset{\cdist}{\textrm{mean}}  g^{-1} \left(\eta(\nset{x},x_f) \pm q\sigma \right),
+$$
+where $q$ is an appropriate quantile of the Normal or t distribution.
 
-We implement and apply these methods in the context of both simple generalized linear and mixed effect models, using simulated data sets -- for univariate and multivariate models.
+If we consider only the uncertainty of the focal predictor, so that the confidence intervals are $\eta \pm q \sigma_f$, we can construct \emph{centered} CI by removing the uncertainty associated with the non-focal predictor(s) using either variance-covariance matrix, $\Sigma$ or centered model matrix, $\bX^{\star}$.
 
-We start with a univariate case where we have only one predictor.
+\subsubsection{Variance-covariance}
 
-\section{Simple fixed effect model}
+The computation of $\hat{\eta}$ remains the same as described above.
+However, to compute $\sigma$, $\Sigma$ is modified by \emph{zeroing-out} (the variance-covariance of all non-focal predictors are assigned zero) entries of non-focal terms$.
+This approach requires \emph{centering} of the predictors in the model matrix, $X$.
+In other words, the fitted model should have centered predictors i.e., $\bX_{c} = \bX - \bar{\bX}$.
 
+_\
+
+\subsubsection{Centered model matrix}
+
+Consider centered $\bX^{\star}_{c}$.
+It follows that the \emph{non-focal} terms in $\bX^{\star}_{c}$ are all zero.
+Consequently the uncertainty due to non-focal predictors are all zeroed-out in the computation of $\sigma$.
+More generally, centered design matrix, $\bX^{\star}_{c}$, impacts on the estimated value of the intercept and its associated variance but not the slopes.
+Thus since non-focal terms in $\bX^{\star}_{c}$ are all zero, it does not matter what their corresponding values are in the variance-covariance matrix.
+Hence, we can compute variable predictions from non-centered predictors (in other words, fitted models with predictors in their natural scales).
+
+\subsection{Simulation examples}
+
+In this section, we illustrate bias in the context of linear and generalized (mixed) linear models.
+We compare predictions when non-focal predictors averaging and when population-based approaches are used.
+Later on, we illustrate and compare our approaches for describing uncertainty in the predictions to existing major \textbf{R} packages for prediction.
+
+\subsection{Simple linear model}
+
+Consider a simple simulation
 \begin{align*}
-\mathrm{logit(status} = 1) &= \eta \\ 
-\eta &= \beta_0 + \beta_{\mathrm{A}} \mathrm{Age} + \beta_{\mathrm{W}} \mathrm{Wealthindex} \\
-\mathrm{Age} &\sim \mathrm{Normal}(0.2, 1) \\
-\mathrm{Wealthindex} &\sim \mathrm{Normal}(0, 1) \\
+y &= \beta_0 + \beta_1\mathrm{x_1} + \beta_2\mathrm{x_2} + \epsilon \\
+\mathrm{x_1} &\sim \mathrm{Normal}(0.2, 1) \\
+\mathrm{x_2} &\sim \mathrm{Normal}(0, 1) \\
+\epsilon &\sim \mathrm{Normal}(0, 1) \\
 \beta_0 &= 1.5 \\
-\beta_{\mathrm{A}} &= 1.0 \\
-\beta_{\mathrm{W}} &= 2
+\beta_{\mathrm{x_1}} &= 1.0 \\
+\beta_{\mathrm{x_2}} &= 2
 \end{align*}
+for $10000$ observations.
 
-Considering $N=1e4$
-
-```{r sim_simple}
-
-simplesim <- function(N=1e4, beta0=1.5, betaA=1.0, betaW=2
-		, age_sd=1, age_mean=0.2
-		, wealth_sd=1, wealth_mean=0
-	) {
-	age <- rnorm(N, age_mean, age_sd)
-	wealthindex <- rnorm(N, wealth_mean, wealth_sd)
-	eta <- beta0 + betaA * age + betaW * wealthindex
-	sim_df <- (data.frame(age=age, wealthindex=wealthindex, eta=eta)
-		%>% mutate(status = rbinom(N, 1, plogis(eta)))
-		%>% select(-eta)
-	)
-	return(sim_df)
-}
-sim_df <- simplesim()
-true_prop <- mean(sim_df$status)
-print(true_prop)
-head(sim_df)
+```{r simple_linear_sim, echo=FALSE}
+sim_linear_df <- simplesim(N=10000, link_scale=TRUE)
+colnames(sim_linear_df) <- c("x1", "x2", "y")
+true_prop_linear <- mean(sim_linear_df$y)
+str(sim_linear_df)
 ```
 
-\subsection{Simple logistic model}
+We want to compare the model predicted mean with "true" (marginal) mean, i.e., `r round(true_prop_linear, 2)`. We first check the coefficient estimates which are relatively close to our truth.
 
-```{r simple_logistic}
-simple_mod <- glm(status ~ age + wealthindex, data = sim_df, family="binomial")
+```{r simple_linear_model}
+simple_lm_mod <- lm(y ~ x1 + x2, sim_linear_df)
+summary(simple_lm_mod)
 ```
 
-\subsection{Variable predictions}
+The next step is to construct variable predictions -- non-focal predictor averaging (hence forth referred to as "average") and population-based, i.e., binned non-focal linear predictor (hence referred to as "binned-nlp") and whole population non-focal linear predictor (hence referred to as "whole-nlp").
 
-```{r pop_ave_fun, echo=FALSE}
-binfun <- function(mod, focal, non.focal, bins=50, ...) {
-	mf <- model.frame(mod)
-	mm <- (mf
-		%>% select_at(c(focal, non.focal))
-	)
-	check_df <- (mf
-		%>% arrange_at(focal)
-		%>% mutate(bin=ceiling(row_number()*bins/nrow(.)))
-		%>% group_by(bin)
-		%>% summarise_all(mean)
-		%>% mutate(model="binned")
-	)
-	return(check_df)
-}
-
+```{r simple_linear_pred, exho=FALSE}
+pred_average <- varpred(simple_lm_mod, "x1", pop.ave="population", nlp.type="binned")
+plot(pred_average)
 ```
 
-```{r simple_pred_age, echo=FALSE, fig.cap="Prediction plots with N=1e4: JD's approach on the left; BB's on the right."} 
-binned_df <- binfun(simple_mod, "age", "wealthindex")
-
-## JD's approach
-pred_age_jd <- predictionfun(simple_mod, "age", isolate=TRUE, include.nonfocal="jd", modelname="JD")
-age_plot_jd <- (ggplot(pred_age_jd, aes(x=age, y=fit))
-	+ geom_line()
-	+ geom_line(aes(y=lwr), lty=2)
-	+ geom_line(aes(y=upr), lty=2)
-	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
-	+ geom_point(data=binned_df, aes(x=age, y=status, color="binned"))
-	+ labs(y="Probability")
-	+ theme(legend.position="none")
-)
-
-## BB's approach
-pred_age_bb <- predictionfun(simple_mod, "age", isolate=TRUE, include.nonfocal="bb", modelname="BB")
-age_plot_bb <- (ggplot(pred_age_bb, aes(x=age, y=fit))
-	+ geom_line()
-	+ geom_line(aes(y=lwr), lty=2)
-	+ geom_line(aes(y=upr), lty=2)
-	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
-	+ geom_point(data=binned_df, aes(x=age, y=status, color="binned"))
-	+ geom_smooth(colour="black", size=0.5)
-	+ labs(y="Probability")
-	+ theme(legend.position="none")
-)
-
-ggarrange(age_plot_jd, age_plot_bb + rremove("ylab"), common.legend=FALSE)
-
-```
-
-Considering $N=1e5$ and then refit the model
-
-```{r sim_simple_ne5}
-sim_df <- simplesim(N=1e5)
-true_prop <- mean(sim_df$status)
-simple_mod <- glm(status ~ age + wealthindex, data = sim_df, family="binomial")
-```
-
-```{r simple_pred_age_ne5, echo=FALSE, fig.cap="Prediction plots with N=1e5: JD's approach on the left; BB's on the right."} 
-binned_df <- binfun(simple_mod, "age", "wealthindex")
-
-## JD's approach
-pred_age_jd <- predictionfun(simple_mod, "age", isolate=TRUE, include.nonfocal="jd", modelname="JD")
-age_plot_jd <- (ggplot(pred_age_jd, aes(x=age, y=fit))
-	+ geom_line()
-	+ geom_line(aes(y=lwr), lty=2)
-	+ geom_line(aes(y=upr), lty=2)
-	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
-	+ geom_point(data=binned_df, aes(x=age, y=status, color="binned"))
-	+ labs(y="Probability")
-	+ theme(legend.position="none")
-)
-
-## BB's approach
-pred_age_bb <- predictionfun(simple_mod, "age", isolate=TRUE, include.nonfocal="bb", modelname="BB")
-age_plot_bb <- (ggplot(pred_age_bb, aes(x=age, y=fit))
-	+ geom_line()
-	+ geom_line(aes(y=lwr), lty=2)
-	+ geom_line(aes(y=upr), lty=2)
-	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
-	+ geom_point(data=binned_df, aes(x=age, y=status, color="binned"))
-	+ geom_smooth(colour="black", size=0.5)
-	+ labs(y="Probability")
-	+ theme(legend.position="none")
-)
-
-ggarrange(age_plot_jd, age_plot_bb + rremove("ylab"), common.legend=FALSE)
-
-```
+\newpage
+\section{References}

### 2021-09-16 bias_correction.rmd
Comparing variable predictions

@@
 library(shellpipes)
+library(lme4)
 library(glmmTMB)
@@
 
+## Simple linear models simulation
 simplesim <- function(N=1e4, beta0=1.5, betaA=1.0, betaW=2
@@
 }
+
+## Random effect simulation
+randsim <- function(nHH = 10, perHH = 5, hh_sd = 2
+	, age_mean = 0.2, age_sd = 1, beta0 = 1.5, betaA = 1
+	, link_scale = FALSE){
+	N <- nHH * perHH
+	hhID <- rep(1:perHH, each=nHH)
+	age <- rnorm(N, age_mean, age_sd)
+	if (link_scale) {
+		eps <- rnorm(N, 0, 1)
+	} else {
+		eps <- 0
+	}
+	sim_df <- (data.frame(hhid=hhID, age=age)
+		%>% group_by(hhid)
+		%>% mutate(hhRE = rnorm(1, 0, hh_sd))
+		%>% ungroup()
+		%>% mutate(eta = beta0 + hhRE + betaA*age + eps)
+		%>% select(-hhRE)
+	)
+	if (link_scale) {
+		sim_df <- (sim_df
+			%>% mutate(y = eta)
+			%>% select(-eta)
+		)
+	} else {
+		sim_df <- (sim_df
+			%>% mutate(y = rbinom(N, 1, plogis(eta)))
+			%>% select(-eta)
+		)
+	}
+	return(sim_df)
+}
+
+binfun <- function(mod, focal, non.focal, bins=50, ...) {
+	mf <- model.frame(mod)
+	mm <- (mf
+		%>% select_at(c(focal, non.focal))
+	)
+	check_df <- (mf
+		%>% arrange_at(focal)
+		%>% mutate(bin=ceiling(row_number()*bins/nrow(.)))
+		%>% group_by(bin)
+		%>% summarise_all(mean)
+		%>% mutate(model="binned")
+	)
+	return(check_df)
+}
+
 ```
@@
 ```{r simple_linear_sim, echo=FALSE}
-sim_linear_df <- simplesim(N=10000, link_scale=TRUE)
+sim_linear_df <- simplesim(N=1e4, link_scale=TRUE)
 colnames(sim_linear_df) <- c("x1", "x2", "y")
 true_prop_linear <- mean(sim_linear_df$y)
-str(sim_linear_df)
 ```
 
-We want to compare the model predicted mean with "true" (marginal) mean, i.e., `r round(true_prop_linear, 2)`. We first check the coefficient estimates which are relatively close to our truth.
+We want to compare the model predicted mean with "true" (marginal) mean, i.e., `r round(true_prop_linear, 2)`.
+The first step is to fit the model and then construct variable predictions -- non-focal predictor averaging (hence forth referred to as "average") and population-based, i.e., binned non-focal linear predictor (hence referred to as "binned-nlp") and whole population non-focal linear predictor (hence referred to as "whole-nlp").
 
-```{r simple_linear_model}
-simple_lm_mod <- lm(y ~ x1 + x2, sim_linear_df)
-summary(simple_lm_mod)
+```{r simple_linear_model, echo=FALSE}
+simple_linear_mod <- lm(y ~ x1 + x2, sim_linear_df)
 ```
 
-The next step is to construct variable predictions -- non-focal predictor averaging (hence forth referred to as "average") and population-based, i.e., binned non-focal linear predictor (hence referred to as "binned-nlp") and whole population non-focal linear predictor (hence referred to as "whole-nlp").
+```{r simple_linear_pred_ave, echo=FALSE}
+## Binned data
+binned_df <- binfun(simple_linear_mod, "x1", "x2", bins=15)
 
-```{r simple_linear_pred, exho=FALSE}
-pred_average <- varpred(simple_lm_mod, "x1", pop.ave="population", nlp.type="binned")
-plot(pred_average)
+## Averaged non-focal predictors
+pred_average <- varpred(simple_linear_mod, "x1", isolate=TRUE, pop.ave="none")
+pred_average_plot <- (plot(pred_average) 
+	+ geom_hline(yintercept=true_prop_linear, lty=2, colour="grey")
+	+ geom_hline(yintercept=mean(pred_average$preds$fit), lty=2, colour="blue")
+	+ geom_vline(xintercept=mean(sim_linear_df$x1), lty=2, colour="grey")
+	+ geom_point(data=binned_df, aes(x=x1, y=y), colour="grey")
+)
+```
+
+```{r simple_linear_pred_binned, echo=FALSE}
+## Binned non-focal linear predictor
+pred_binned <- varpred(simple_linear_mod, "x1", isolate=TRUE
+	, pop.ave="population", nlp.type="binned"
+)
+pred_binned_plot <- (plot(pred_binned) 
+	+ geom_smooth(colour="red", size=0.5)
+	+ geom_hline(yintercept=true_prop_linear, lty=2, colour="grey")
+	+ geom_hline(yintercept=mean(pred_binned$preds$fit), lty=2, colour="blue")
+	+ geom_vline(xintercept=mean(sim_linear_df$x1), lty=2, colour="grey")
+	+ geom_point(data=binned_df, aes(x=x1, y=y), colour="grey")
+)
+```
+
+Figure \ref{fig:simple_linear_pred_plots} compares variable predictions based on the three approaches highlighted above.
+All the three approaches give similar estimates (`r round(mean(pred_average$preds$fit), 3)`) of expected mean, $\bar{f(x)}$, and very close to the observed mean, $f(\bar{x})$.
+In other words, for the simple linear model, $\bar{f(x)} \approx f(\bar{x}) = `r round(true_prop_linear, 2)`$.
+For perfect predictions, we expect the blue, grey and the black (or red) to intersect at the same (or very close) point.
+Additional smoothing step is needed for binned-nlp, Figure \ref{fig:simple_linear_pred_plots}c, in order to generate smooth trend lines from noisy predictions.
+
+```{r simple_linear_pred_whole, echo=FALSE}
+## Whole population non-focal linear predictor
+pred_whole <- varpred(simple_linear_mod, "x1", isolate=TRUE
+	, pop.ave="population", nlp.type="whole"
+)
+pred_whole_plot <- (plot(pred_whole) 
+	+ geom_hline(yintercept=true_prop_linear, lty=2, colour="grey")
+	+ geom_hline(yintercept=mean(pred_whole$preds$fit), lty=2, colour="blue")
+	+ geom_vline(xintercept=mean(sim_linear_df$x1), lty=2, colour="grey")
+	+ geom_point(data=binned_df, aes(x=x1, y=y), colour="grey")
+)
+```
+
+```{r simple_linear_pred_plots, echo=FALSE, fig.cap=" \\label{fig:simple_linear_pred_plots}A comparison of variable predictions.
+The dotted blue and grey horizontal lines are the expected and observed/true means, respectively.
+The grey dots are the binned observations.
+The vertical grey line is the mean of the focal predictor (model center).
+In a) non-focal predictors averaged across the levels of focal predictor; b) non-focal linear predictor is binned across the levels of the focal predictor; and c) whole population non-focal linear predictor is used for each level of the focal predictor.
+In c, the noisy predictions, i.e., black lines, are smoothened resulting to the red trend-line with the corresponding CI represented by the grey shading."}
+ggarrange(pred_average_plot
+	, pred_binned_plot + rremove("ylab")
+	, pred_whole_plot
+	, labels = c("a)", "b)", "c)")
+	, common.legend=FALSE
+)
+```
+
+\subsection{Simple generalized linear model}
+
+Consider binary outcome simulation for improved (1) or unimproved (0) water services status, together with two socio-economic variables (age and wealth index), such that:
+\begin{align*}
+\mathrm{logit(status} = 1) &= \eta \\
+\eta &= \beta_0 + \beta_{\mathrm{A}} \mathrm{Age} + \beta_{\mathrm{W}} \mathrm{Wealthindex} \\
+\mathrm{Age} &\sim \mathrm{Normal}(0.2, 1) \\
+\mathrm{Wealthindex} &\sim \mathrm{Normal}(0, 1) \\
+\beta_0 &= 1.5 \\
+\beta_{\mathrm{A}} &= 1.0 \\
+\beta_{\mathrm{W}} &= 1
+\end{align*}
+
+```{r simple_glm_sim, echo=FALSE}
+sim_glm_df <- simplesim(N=1e4)
+true_prop_glm <- mean(sim_glm_df$status)
+```
+
+```{r simple_glm_model, echo=FALSE}
+simple_glm_mod <- glm(status ~ age + wealthindex, sim_glm_df, family="binomial")
+```
+
+```{r simple_glm_pred_ave, echo=FALSE}
+## Binned data
+binned_df_glm <- binfun(simple_glm_mod, "age", "wealthindex", bins=15)
+
+## Averaged non-focal predictors
+pred_average_glm <- varpred(simple_glm_mod, "age", isolate=TRUE, pop.ave="none")
+pred_average_glm_plot <- (plot(pred_average_glm) 
+	+ geom_hline(yintercept=true_prop_glm, lty=2, colour="grey")
+	+ geom_hline(yintercept=mean(pred_average_glm$preds$fit), lty=2, colour="blue")
+	+ geom_vline(xintercept=mean(sim_glm_df$age), lty=2, colour="grey")
+	+ geom_point(data=binned_df_glm, aes(x=age, y=status), colour="grey")
+	+ labs(y="Prob. of improved \n service")
+)
+```
+
+```{r simple_glm_pred_binned, echo=FALSE}
+## Binned non-focal linear predictor
+pred_binned_glm <- varpred(simple_glm_mod, "age", isolate=TRUE
+	, pop.ave="population", nlp.type="binned"
+)
+pred_binned_glm_plot <- (plot(pred_binned_glm) 
+	+ geom_smooth(colour="red", size=0.5)
+	+ geom_hline(yintercept=true_prop_glm, lty=2, colour="grey")
+	+ geom_hline(yintercept=mean(pred_binned_glm$preds$fit), lty=2, colour="blue")
+	+ geom_vline(xintercept=mean(sim_glm_df$age), lty=2, colour="grey")
+	+ geom_point(data=binned_df_glm, aes(x=age, y=status), colour="grey")
+	+ labs(y="Prob. of improved \n service")
+)
+```
+
+```{r simple_glm_pred_whole, echo=FALSE}
+## Whole non-focal linear predictor
+pred_whole_glm <- varpred(simple_glm_mod, "age", isolate=TRUE
+	, pop.ave="population", nlp.type="whole"
+)
+pred_whole_glm_plot <- (plot(pred_whole_glm) 
+	+ geom_hline(yintercept=true_prop_glm, lty=2, colour="grey")
+	+ geom_hline(yintercept=mean(pred_whole_glm$preds$fit), lty=2, colour="blue")
+	+ geom_vline(xintercept=mean(sim_glm_df$age), lty=2, colour="grey")
+	+ geom_point(data=binned_df_glm, aes(x=age, y=status), colour="grey")
+	+ labs(y="Prob. of improved \n service")
+)
+```
+
+Figure \ref{fig:simple_glm_pred_plots} shows predicted probability of improved services for a particular year old household head.
+In comparison to binned-nlp ($\bar{f(x)} = `r round(mean(pred_binned_glm$preds$fit), 3)`$) and whole-nlp ($\bar{f(x)} = `r round(mean(pred_whole_glm$preds$fit), 3)`$) which closely estimate the expected marginal probability ($\bar{f(x)} \approx f(\bar{x}) = `r round(true_prop_glm, 2)`$), averaged non-focal predictors approach (Figure \ref{fig:simple_glm_pred_plots}a) over-predicts the expected probabilities with $\bar{f(x)} = `r round(mean(pred_average_glm$preds$fit), 2)` > f(\bar{x})$.
+Because of the nonlinear link function in the model, our prediction are likely to be biased.
+However since population-based approaches (binned-nlp and whole-nlp) incorporate bias correction in their construction we see better estimates as opposed to averaged non-focal predictors approach.
+
+```{r simple_glm_pred_plots, echo=FALSE, fig.cap=" \\label{fig:simple_glm_pred_plots}A comparison of variable predictions.
+The dotted blue and grey horizontal lines are the expected and observed/true means, respectively.
+The grey dots are the binned observations.
+The vertical grey line is the mean of the focal predictor (model center).
+In a) non-focal predictors averaged across the levels of focal predictor; b) non-focal linear predictor is binned across the levels of the focal predictor; and c) whole population non-focal linear predictor is used for each level of the focal predictor.
+In c, the noisy predictions, i.e., black lines, are smoothened resulting to the red trend-line with the corresponding CI represented by the grey shading."}
+
+ggarrange(pred_average_glm_plot
+	, pred_binned_glm_plot + rremove("ylab")
+	, pred_whole_glm_plot
+	, labels = c("a)", "b)", "c)")
+	, common.legend=FALSE
+)
+```
+
+\subsection{Mixed models}
+
+We now consider more complex model involving fixed and random effects. We start with simple ones and then evolve to more complicated models.
+
+\subsubsection{One grouping factor, linear random effect model}
+
+Suppose in the first simulation (but now only a single predictor), the observations are recorded more than once from a number.
+In particular, let $H$ be the number of households indexed by the grouping factor, and $h[i]$ be household of the $i$th observation, so that:
+\begin{align*}
+y_i &= \beta_0 + \alpha_{h[i]}+ \beta_{\mathrm{x}} \mathrm{x}_i + \epsilon_i\\
+\alpha_h &\sim \mathrm{Normal}(0, 2), \quad \mathrm{h = 1, \cdots, H}\\
+\epsilon_i &\sim \mathrm{Normal}(0, 1) \\
+\mathrm{x}_i &\sim \mathrm{Normal}(0.2, 1),  \quad \mathrm{i = 1, \cdots, n}\\
+\beta_0 &= 1.5 \\
+\beta_{\mathrm{x}} &= 1.0 \\
+\end{align*}
+
+```{r lme_sim, echo=FALSE}
+sim_lme_df <- randsim(nHH=1000, perHH=10, hh_sd=2, link_scale=TRUE)
+colnames(sim_lme_df) <- c("hhid", "x", "y")
+true_prop_lme <- mean(sim_lme_df$y)
+str(sim_lme_df)
+```
+
+```{r raneff_linear_model, echo=FALSE}
+lme_mod <- glmmTMB(y ~ x + (1|hhid)
+	, data = sim_lme_df
+	, family = gaussian()
+)
+summary(lme_mod)
+```
+
+```{r lme_pred_ave, echo=FALSE}
+## Binned data
+binned_df_lme <- binfun(lme_mod, "x", "hhid", bins=15)
+
+## Averaged non-focal predictors
+pred_average_lme <- varpred(lme_mod, "x", isolate=TRUE, pop.ave="none")
+pred_average_lme_plot <- (plot(pred_average_lme) 
+	+ geom_hline(yintercept=true_prop_lme, lty=2, colour="grey")
+	+ geom_hline(yintercept=mean(pred_average_lme$preds$fit), lty=2, colour="blue")
+	+ geom_vline(xintercept=mean(sim_lme_df$x), lty=2, colour="grey")
+	+ geom_point(data=binned_df_lme, aes(x=x, y=y), colour="grey")
+	+ labs(y="Predicted y")
+)
+```
+
+```{r lme_pred_binned, echo=FALSE}
+## Binned non-focal linear predictor
+pred_binned_lme <- varpred(lme_mod, "x", isolate=TRUE, pop.ave="population", nlp.type="binned")
+pred_binned_lme_plot <- (plot(pred_binned_lme) 
+	+ geom_smooth(colour="red", size=0.5)
+	+ geom_hline(yintercept=true_prop_lme, lty=2, colour="grey")
+	+ geom_hline(yintercept=mean(pred_binned_lme$preds$fit), lty=2, colour="blue")
+	+ geom_vline(xintercept=mean(sim_lme_df$x), lty=2, colour="grey")
+	+ geom_point(data=binned_df_lme, aes(x=x, y=y), colour="grey")
+	+ labs(y="Predicted y")
+)
+```
+
+```{r lme_pred_whole, echo=FALSE}
+## Whole non-focal linear predictor
+pred_whole_lme <- varpred(lme_mod, "x", isolate=TRUE, pop.ave="population", nlp.type="whole")
+pred_whole_lme_plot <- (plot(pred_whole_lme) 
+	+ geom_hline(yintercept=true_prop_lme, lty=2, colour="grey")
+	+ geom_hline(yintercept=mean(pred_whole_lme$preds$fit), lty=2, colour="blue")
+	+ geom_vline(xintercept=mean(sim_lme_df$x), lty=2, colour="grey")
+	+ geom_point(data=binned_df_lme, aes(x=x, y=y), colour="grey")
+	+ labs(y="Predicted y")
+)
+```
+
+In \ref{fig:simple_lme_pred_plots}, we investigate the contribution of random effects to bias in predictions in the case of a simple linear mixed model with only one predictor and random intercept effect.
+Similar to the case of simple linear model, the expected mean is very close to the observed in all the three approaches.
+
+```{r simple_lme_pred_plots, echo=FALSE, fig.cap=" \\label{fig:simple_lme_pred_plots}A comparison of variable predictions for simple linear mixed effect model.
+The dotted blue and grey horizontal lines are the expected and observed/true means, respectively.
+The grey dots are the binned observations.
+The vertical grey line is the mean of the focal predictor (model center).
+In a) non-focal predictors averaged across the levels of focal predictor; b) non-focal linear predictor is binned across the levels of the focal predictor; and c) whole population non-focal linear predictor is used for each level of the focal predictor.
+In c, the noisy predictions, i.e., black lines, are smoothened resulting to the red trend-line with the corresponding CI represented by the grey shading."}
+
+ggarrange(pred_average_lme_plot
+	, pred_binned_lme_plot + rremove("ylab")
+	, pred_whole_lme_plot
+	, labels = c("a)", "b)", "c)")
+	, common.legend=FALSE
+)
 ```

### 2021-09-16 R/effectsfuns.R

@@ -86,7 +86,7 @@
 
 varpred <- function(mod, focal_predictors, x.var = NULL
 	, type = c("response", "link"), isolate = FALSE, isolate.value = NULL, level = 0.95
-	, steps = 101, at = list(),  dfspec = 100, vcov. = NULL, internal = FALSE, avefun = mean
+	, steps = 100, at = list(),  dfspec = 100, vcov. = NULL, internal = FALSE, avefun = mean
 	, zero_out_interaction = FALSE, which.interaction = c("emmeans", "effects")
 	, pop.ave = c("none", "quantile", "population"), nlp.type=c("whole", "binned")
 	, include.re = FALSE, bias.adjust = FALSE, sigma = NULL
@@ -194,7 +194,7 @@ varpred <- function(mod, focal_predictors, x.var = NULL
 				} 
 			}
 			mm_non.focal <- mm[, terms_non.focal, drop=FALSE]
-			pred_non.focal <- mm_non.focal %*% betahat_non.focal
+			pred_non.focal <- as.vector(mm_non.focal %*% betahat_non.focal)
 			if (pop.ave=="quantile") {
 				mm_x.var <- as.vector(mm[, terms_focal])
 			}

### 2021-09-22 variable_predictions.Rnw
Variable effects or predictions?

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{natbib}
\usepackage[dvipsnames]{xcolor}
\usepackage{graphicx, subfig}
\usepackage{float}
\usepackage{hyperref}

\def\code#1{\texttt{#1}}

\newcommand{\JD}[1]{{\color{blue} \emph{#1}}}
\newcommand{\bmb}[1]{{\color{RubineRed!70!} \emph{#1}}}
\newcommand{\pkg}[1]{\textbf{#1}}

\newcommand{\bX}{{\mathbf X}}
\newcommand{\bbeta}{{\boldsymbol \beta}}
\newcommand{\boldeta}{{\boldsymbol \eta}}
\newcommand{\boldmu}{{\boldsymbol \mu}}

\newcommand{\nset}[1]{#1_{\{n\}}}
\newcommand{\yref}{y_{\textrm{ref}}}
\newcommand{\cdist}{{D(\nset{x}|x_f)}}
\newcommand{\cdistprime}{{D(\nset{x}|x_{f'})}}
\newcommand{\xfprime}{x_{f'}}

\setlength{\parindent}{0pt}

\title{Generating and visualizing predictor effects???}

\begin{document}

\maketitle
\setkeys{Gin}{width=0.8\textwidth}

\section{Introduction}

<<setup, echo=FALSE, message=FALSE, warning=FALSE>>=
library(ggplot2)
library(stargazer)
library(vareffects); varefftheme()

set.seed(9991)
load("variable_predictions_objs.rda")
@

In many applications, predictions of expected responses, or response probabilities, are often of interest. Predictions provide a great way to summarize what the regression model is telling us and are very useful for interpreting and visualizing models estimates using graphs. For example, in logistic regression models, the coefficient estimates are usually not easy to interpret, and we may want to explore the "effects" of covariates on predicted probabilities, or we may require predictions of responses at various levels of covariates. For example, as a way to plan, public health officials may want to know the predicted probability of improved water services in slums on the basis of the dwellers' household size, income, wealth index, etc.

Both simple and generalized linear (mixed) models (GL(M)Ms) can examine very complex relationships, including nonlinear relationships between response and predictors, interactions between predictors (and via splines), and nonlinear transformations via link functions due to their flexibility. This flexibility comes at a cost, for example, complex multivariate models may risk misinterpretation, and miscalculation of quantities of interest. Also, coefficient estimates of models involving nonlinear link functions or interactions lose their direct interpretation \citep{leeper2017interpreting}, meaning that interpretation of derived quantities from these estimates requires some understanding of the specified model.

When visually presented, predictions provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind transformations on the dependent variables whose estimates are usually, but not always, a subject to less clarity of interpretation. However, generating predictions and associated confidence intervals for regression models has a number of challenges. In particular: 
\begin{enumerate}
\item choice of representative values of \emph{focal} variable(s) and appropriate 'model center' for \emph{non-focal} variables in the case of multiple predictors
\item  propagation of uncertainty ([how] can we incorporate uncertainty in nonlinear components of (G)LMMs? Should we exclude variation in non-focal parameters?)
\end{enumerate}

Most common way of dealing with the first challenge is taking unique levels of the focal predictor if discrete or taking appropriately sized quantiles (or bins) if continuous, and then calculating the predictions while holding non-focal predictors at their mean values. There a number of variations on description of this quantity -- \emph{predictor effect} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} and \emph{estimated marginal means} \citep{lenth2018package}. In this article, we refer to this quantity as \emph{variable (effect?) predictions} since this quantity should, for example, tell us what we would expect the outcome to be at a particular value or level of the variable. An alternative, is the population-based approach which involves computing the prediction over the population of the non-focal predictors and then averaging across the values of the focal predictor. We give special focus to the second challenge later in the article.

When dealing with nonlinear link functions, the correct predictions for example, are even much harder to estimate. One approach is to make predictions on the transformed scale (linear predictor scale), and then back-transform back to the original scale. However, the back-transformation may either result in biased predictions or requires some approximation. In particular, Jensen's inequality and bias in expected mean prediction induced by nonlinear transformation of the response variable can lead inaccurate predictions.

The main purpose of this article is to discuss and implement various approaches for computing variable predictions and how to correctly compute the associated confidence intervals. We further explore approaches for correcting bias in predictions for GL(M)Ms involving nonlinear link functions.

The outline of this article is as follows: Section \ref{sec:qoi} provides formal definition various quantities of interest, Section \ref{sec:stats_background} provides statistical background of estimation of model predictions, Section \ref{sec:uncertainity_propagation} describes mathematical and computation implementation of the proposed method for uncertainty propagation, and lastly Section \ref{sec:bias_correction} describes methods for bias correction in GL(M)M together with the computational implementation.

\section{Quantities of interest}\label{sec:qoi}

Several quantities of interest may be derived from regression models. The first one being the coefficient estimates. Others being predictors, model matrix, predicted values and marginal effects. \emph{predictors} are model input variables and is associated with one or more \emph{variables} (in polynomials, splines and models with interaction). Model matrix refers to the design matrix whose rows include all combination of variables appearing in the interaction terms, along with the ``typical" values of the focal and non-focal predictors.

To illustrate some of the concepts, consider results from, hypothetical  simulated example, regression of household size as function of household wealth index and age of household head, and the second model involving these variables plus the interaction between them.

<<qoi_coefs, echo=FALSE, results=tex, message=FALSE, warning=FALSE>>=
stargazer(qoi_mod1, qoi_mod2
	, title = "Example of simple linear model output"
	, single.row=TRUE
	, table.placement="H"
	, label = "tab:qoi_coefs"
)
@

In simple linear models with no interaction terms, the default out for the coefficient estimates in Table ~\ref{tab:qoi_coefs} model (1) is simple and directly interpretable as the expected change in household size for a unit change in age. In particular, a unit change in age is associated with a household that is \Sexpr{round(coef(qoi_mod1)[[2]], 3)} bigger. This is the \emph{unconditional marginal effect} \citep{leeper2017interpreting} and it is constant across all the observations and levels of all other variables.

Figure ~\ref{fig:qoi_pred_plot} shows the comparison of \emph{variable prediction} and unconditional marginal effect of effect of age on household size, based on model estimates in Table ~\ref{tab:qoi_coefs}. Since the model has no interactions, the relationship between the predicted household size and age is linear, hence the marginal effect of age is the slope ($\frac{\Delta \mathrm{hh size}}{\Delta \mathrm{age}}$) of the variable effect line and can be calculated irrespective of the values of wealth index. 
%
\begin{figure}[H]
\begin{center}
<<qoi_pred_plot, echo=FALSE, results=tex, fig=TRUE, message=FALSE>>=
qoi_age_pred_plot
@
\end{center}
\caption{Variable effect plot and marginal effect of age of household head on household size. For a linear model with no interaction, the marginal effect is the slope of the variable effect line.}
\label{fig:qoi_pred_plot}
\end{figure}
%
Variable effect, on the other hand, is the expected household size for a particular age, holding wealth index at its mean. In particular, the purpose and goal of a variable effect seems fairly straightforward; for specified values of (a) focal predictor(s), we want to give a point estimate and confidence intervals for the prediction of the model for a ``typical" (= random sample over the multivariate distribution of non-focal parameters) individual with those values of the predictors.

On the other hand, when model specification involves other kinds of terms such as variable interactions, log and power transformations, etc., the coefficient estimates cannot easily and directly communicate the relationship between the outcome and the independent variable of interest because of the multiple coefficients of the same variable. For instance, in model (2), the interaction between age and wealth index is added, hence the coefficient estimate for age (\Sexpr{round(coef(qoi_mod2)[[2]], 3)}) can only be regarded as unconditional marginal effect when effect of wealth index (and thus \code{age:wealthindex}) is zero, which is never the case. In other words, for models involving interactions terms, looking at a single estimated coefficient in isolation and failing to take into account the estimates involving the interaction terms may lead to inaccurate interpretations \citep{brambor2006understanding, leeper2017interpreting}.

\section{Statistical background}\label{sec:stats_background}

To get an intuition of how conditioning on the mean values of the non-focal predictors work, suppose we are interested in the variable predictions of a particular predictor (hence forth referred as \emph{focal} predictor otherwise \emph{non-focal}), $x_f$, from the set of predictors. To keep it simple, assume that the model has no interaction terms. Then the idea is to \emph{anchor} the values of non-focal predictors to some particular values. For example fixing the values of \emph{non-focal} predictor(s) at some typical values -- typically determined by averaging (for now) in some meaningful way, for example, arithmetic mean  for continuous and average over the levels of the factors for categorical non-focal predictors. The easiest way to achieve this is by constructing $\bX^\star$ by averaging the columns of non-focal variables in model matrix $\bX$, and together with appropriately chosen values of focal predictor(s).

Consider a simple linear model with linear predictor $\eta = \bX\bbeta$ and let $g(\boldmu) = \boldeta$ be an identity link function (in the case of simple linear model), where $\boldmu$ is the expected value of response variable $y$. Let $\hat{\bbeta}$ be the estimate of $\bbeta$, together with the estimated covariance matrix $V(\hat{\bbeta})$ of $\hat{\bbeta}$. Let $\mathbf{X^*}$ be the model matrix, inheriting most of its key properties, for example transformations on predictors and interactions from the model matrix, $\mathbf{X}$. Then the prediction $\hat{\boldeta}^\star = \bX^\star\hat{\bbeta}$ is the variable effect for the focal predictor in question.

An alternative, save for later, formulation of variable effect involves, expressing the linear predictor as the sum of the focal and non-focal predictor linear predictor. In particular, $\eta^\star(x_f, \nset{{\bar{x}}}) = \beta_f x_f + \sum \nset{\beta} \nset{{\bar{x}^\star}}$, where $\nset{{\bar{x}^\star}}$ are the appropriately averaged entries of non-focal predictors and $\nset{x}$ is a vector of values of the non-focal predictors for a particular observation.

Generally, in generalized linear models with other link functions, the predictions are first generated in linear predictor scale, $\hat{\boldeta}^*$, and then transformed to the original response scale using $g^{-1}(\hat{\boldeta}^*)$.

\section{Uncertainty propagation}\label{sec:uncertainity_propagation}

What about the confidence intervals (CI)? The limits of the confidence intervals are points, not mean values. In principle, every observation/set of non-focal predictors has a different CI.

\section{Bias correction}\label{sec:bias_correction}

\clearpage
% References
\bibliographystyle{plainnat}
\bibliography{variable_predictions}

\end{document}

### 2021-09-22 variable_predictions_funs.R

@@ -1,0 +1,24 @@
+
+## Simulate data for simple lm or glm
+### If link_scale=TRUE -> lm otherwise glm
+simplesim <- function(N=1e4, beta0=1.5, beta1=1.0, beta2=2
+		, x1_sd=1, x1_mean=0.2
+		, x2_sd=1, x2_mean=0
+		, link_scale = FALSE
+	) {
+	x1 <- rnorm(N, x1_mean, x1_sd)
+	x2 <- rnorm(N, x2_mean, x2_sd)
+	eta <- beta0 + beta1 * x1 + beta2 * x2
+	if (link_scale) {
+		eta <- eta + rnorm(N) 
+	}
+	sim_df <- data.frame(x1=x1, x2=x2, y=eta)
+	if (!link_scale) {
+		sim_df <- (sim_df
+			%>% mutate(y = rbinom(N, 1, plogis(y)))
+		)
+	}
+	return(sim_df)
+}
+
+saveEnvironment()

### 2021-09-22 variable_predictions.Rnw
Autocommit

@@
 \def\code#1{\texttt{#1}}
+\let\proglang=\textsf
 
@@
 Consider a simple linear model with linear predictor $\eta = \bX\bbeta$ and let $g(\boldmu) = \boldeta$ be an identity link function (in the case of simple linear model), where $\boldmu$ is the expected value of response variable $y$.
-Let $\hat{\bbeta}$ be the estimate of $\bbeta$, together with the estimated covariance matrix $V(\hat{\bbeta})$ of $\hat{\bbeta}$.
+Let $\hat{\bbeta}$ be the estimate of $\bbeta$, together with the estimated covariance matrix $\Sigma = V(\hat{\bbeta})$ of $\hat{\bbeta}$.
 Let $\mathbf{X^*}$ be the model matrix, inheriting most of its key properties, for example transformations on predictors and interactions from the model matrix, $\mathbf{X}$.
@@
 
-What about the confidence intervals (CI)? The limits of the confidence intervals are points, not mean values. In principle, every observation/set of non-focal predictors has a different CI.
+What about the confidence intervals (CI)?
+The limits of the confidence intervals are points, not mean values.
+In principle, every observation/value of focal predictors has a different CI.
+The traditional way to compute variances for predictions is $\sigma^2 = \textrm{Diag}(\bX^\star \Sigma \bX^{\star\top})$ \citep{lenth2018package, fox2009effect}, so that the confidence intervals are $\eta \pm q\sigma$, where $q$ is an appropriate quantile of Normal or t distribution.
+This approach incorporates all the uncertainties -- including the uncertainty due to non-focal predictors.
+But what if we are only interested in the uncertainty as a result of the focal predictor, so that the confidence intervals are $\eta \pm q \sigma_f$?
+
+Currently, commonly used \proglang{R} packages for constructing predictions do not exclude the uncertainties resulting from the \emph{non-focal} predictors when computing the CIs.
+A non-trivial way to exclude uncertainties associated with non-focal predictors in some of these packages is to provide a user defined variance-covariance matrix with the covariances of \emph{non-focal} terms set to $0$ -- \emph{zeroing-out} variance-covariance matrix.
+This only works when the input predictors are \emph{centered} prior to model fitting, in case of numerical predictors, and even much complicated when the predictors are categorical.
+We first describe this variance-covariance based approach and then discuss our proposed method which is based on \emph{model centering} and does not require input predictors to be scaled prior to model fitting.
+Model center is found by averaging the variables over the model matrix, $\bX^\star$.
+
+\subsection{Variance-covariance}
+
+The computation of $\hat{\eta}^\star$ remains the same as described above.
+However, to compute $\sigma$, $\Sigma$ is modified by \emph{zeroing-out} (the variance-covariance of all non-focal predictors are assigned zero) variances of non-focal terms.
+Although this is the simplest approach, it requires centering of continuous predictors prior to model fitting and proper handling of categorical predictors.
+One way to center predictors is to create a design matrix, $\bX$, and then compute $\bX_{c} = \bX - \bar{\bX}$.
+
+\subsection{Centered model matrix}
+
+Consider centered model matrix $\bX^{\star}_{c} = \{\bX_f^\star, \nset{{\bX}^\star} - \bar{\nset{{\bX^\star}}}\}$.
+It follows that the non-focal terms in $\bX^{\star}_{c}$ are all zero.
+Consequently the uncertainty due to non-focal predictors are all zeroed-out in the computation of $\sigma$.
+More generally, centered design matrix, $\bX^{\star}_{c}$, impacts on the estimated value of the intercept and its associated variance but not the slopes.
+Thus since non-focal terms in $\bX^{\star}_{c}$ are all zero, it does not matter what their corresponding values are in the variance-covariance matrix.
+Hence, we can compute CIs from non-centered predictors (in other words, fitted models with predictors in their natural scales).
+
+\subsection{Simulation examples}
+
+In this section, we illustrate variable effects in the context of linear models involving continuous and categorical predictors with or without interactions.
+We compare the expected to the known simulation "truth".
+We compare predictions when non-focal predictors averaging and when population-based approaches are used.
+Later on, we illustrate and compare our approaches for describing uncertainty in the predictions to existing major \textbf{R} packages for prediction.
+
+\subsection{Simple linear model}
+
+Consider a simple simulation
+\begin{align*}
+y &= \beta_0 + \beta_1\mathrm{x_1} + \beta_2\mathrm{x_2} + \epsilon \\
+\mathrm{x_1} &\sim \mathrm{Normal}(0.2, 1) \\
+\mathrm{x_2} &\sim \mathrm{Normal}(0, 1) \\
+\epsilon &\sim \mathrm{Normal}(0, 1) \\
+\beta_0 &= 1.5 \\
+\beta_{\mathrm{x_1}} &= 1.0 \\
+\beta_{\mathrm{x_2}} &= 2
+\end{align*}
+for $10000$ observations.
+
+```{r simple_linear_sim, echo=FALSE}
+sim_linear_df <- simplesim(N=1e4, link_scale=TRUE)
+colnames(sim_linear_df) <- c("x1", "x2", "y")
+true_prop_linear <- mean(sim_linear_df$y)
+```
 

### 2021-09-23 variable_predictions.Rnw
Can we talk about interactions?

@@
 \maketitle
-\setkeys{Gin}{width=0.8\textwidth}
+%\setkeys{Gin}{width=0.8\textwidth}
 
@@
 library(ggplot2)
+library(ggpubr)
+
 library(stargazer)
+
 library(vareffects); varefftheme()
@@
 
-Consider centered model matrix $\bX^{\star}_{c} = \{\bX_f^\star, \nset{{\bX}^\star} - \bar{\nset{{\bX^\star}}}\}$.
+Consider centered model matrix $\bX^{\star}_{c} = \{\bX_f^\star, \nset{{\bX}^\star} - \nset{{\bar{\bX}}^\star}\}$.
 It follows that the non-focal terms in $\bX^{\star}_{c}$ are all zero.
@@
 
+\subsection{Handling higher order terms}
+
+In models containing higher order terms such as interactions, splines, polynomials, etc., we adopt the method implemented in \pkg{effects} package \citep{fox2009effect} which involves combining the interactions with the main effects marginal to the interaction.
+
 \subsection{Simulation examples}
 
-In this section, we illustrate variable effects in the context of linear models involving continuous and categorical predictors with or without interactions.
+In this section, we illustrate variable effects (together with the associated CI) in the context of linear models involving continuous and categorical predictors with or without interactions.
 We compare the expected to the known simulation "truth".
-We compare predictions when non-focal predictors averaging and when population-based approaches are used.
-Later on, we illustrate and compare our approaches for describing uncertainty in the predictions to existing major \textbf{R} packages for prediction.
 
-\subsection{Simple linear model}
+\subsection{Continuous predictors}
 
-Consider a simple simulation
+Consider simulation example described in Section \ref{sec:qoi}, with slight modifications to incorporate interaction between predictors. The first case, is the model with no interaction: 
+%
 \begin{align*}
-y &= \beta_0 + \beta_1\mathrm{x_1} + \beta_2\mathrm{x_2} + \epsilon \\
-\mathrm{x_1} &\sim \mathrm{Normal}(0.2, 1) \\
-\mathrm{x_2} &\sim \mathrm{Normal}(0, 1) \\
-\epsilon &\sim \mathrm{Normal}(0, 1) \\
+\mathrm{hh~size} &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \epsilon_i \\
+\mathrm{Age}_i &\sim \mathrm{Normal}(0.2, 1) \\
+\mathrm{Wealthindex}_i &\sim \mathrm{Normal}(0, 1) \\
+\epsilon_i &\sim \mathrm{Normal}(0, 1) \\
 \beta_0 &= 1.5 \\
-\beta_{\mathrm{x_1}} &= 1.0 \\
-\beta_{\mathrm{x_2}} &= 2
+\beta_{\mathrm{A}} &= 0.2 \\
+\beta_{\mathrm{W}} &= 2 \\
+i &= 1,\cdots, 1000
 \end{align*}
-for $10000$ observations.
+%
+Secondly, we add interaction terms between the non-focal predictors (\code{wealth index} and \code{household expenditure}) but not with the focal predictor \code{age}:
+%
+\begin{align*}
+\mathrm{hh~size} &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \beta_{\mathrm{E}}\mathrm{Expenditure}_i\\ 
+&+ \beta_{\mathrm{WE}}\mathrm{Wealthindex}_i*\mathrm{Expenditure}_i + \epsilon_i \\
+\mathrm{Age}_i &\sim \mathrm{Normal}(0.2, 1) \\
+\mathrm{Wealthindex}_i &\sim \mathrm{Normal}(0, 1) \\
+\mathrm{Expenditure}_i &\sim \mathrm{Normal}(0, 1) \\
+\epsilon_i &\sim \mathrm{Normal}(0, 1) \\
+\beta_0 &= 1.5 \\
+\beta_{\mathrm{A}} &= 0.8 \\
+\beta_{\mathrm{W}} &= 2 \\
+\beta_{\mathrm{E}} &= 1.5 \\
+\beta_{\mathrm{WE}} &= 1 \\
+i &= 1,\cdots, 1000
+\end{align*}
+%
+And lastly, we now consider model with interaction terms between the focal predictor (\code{age}) and one of the non-focal predictor (\code{wealth index}):
+%
+\begin{align*}
+\mathrm{hh~size} &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \beta_{\mathrm{E}}\mathrm{Expenditure}_i\\ 
+&+ \beta_{\mathrm{AW}}\mathrm{Age}_i*\mathrm{Wealthindex}_i + \epsilon_i \\
+\mathrm{Age}_i &\sim \mathrm{Normal}(0.2, 1) \\
+\mathrm{Wealthindex}_i &\sim \mathrm{Normal}(0, 1) \\
+\mathrm{Expenditure}_i &\sim \mathrm{Normal}(0, 1) \\
+\epsilon_i &\sim \mathrm{Normal}(0, 1) \\
+\beta_0 &= 1.5 \\
+\beta_{\mathrm{A}} &= 0.2 \\
+\beta_{\mathrm{W}} &= 2 \\
+\beta_{\mathrm{E}} &= 1.5 \\
+\beta_{\mathrm{AW}} &= 1 \\
+i &= 1,\cdots, 1000
+\end{align*}
+%
 
-```{r simple_linear_sim, echo=FALSE}
-sim_linear_df <- simplesim(N=1e4, link_scale=TRUE)
-colnames(sim_linear_df) <- c("x1", "x2", "y")
-true_prop_linear <- mean(sim_linear_df$y)
-```
+We first fitted all the three models and in each case, compared three different variable effects together with the corresponding CIs -- 1) uncertainty due to non-focal predictors included (everything), 2) uncertainty due to non-focal predictors removed using variance-covariance matrix (zero-vcov), and 3) uncertainty due to non-focal predictors removed using centered model matrix (centered mm).
+The results are summarised in Figure ~\ref{fig:pred_age_cont_plots}.
+
+Figure ~\ref{fig:pred_age_cont_plots}a represents variable effects for the model with no interaction, Figure ~\ref{fig:pred_age_cont_plots}b shows variable effects for model with non-focal predictors interacting with each other, and Figure ~\ref{fig:pred_age_cont_plots}c shows variable effects for the model with focal and non-focal predictors interacting.
+
+For perfect predictions, we expect the yellow and grey dotted lines to overlay each other and they should intersect with the black solid line and the vertical dotted grey line at the same point (or very close), Figure ~\ref{fig:pred_age_cont_plots}a.
+In other words, at the model center (mean of the focal predictor), generally, we expect the narrower CIs.
+If we properly anchor the model at its center (or any other appropriate value, assuming no interactions), we would expect the variance at the anchor to be zero, i.e., the CIs bands crosses at that point (see zero-cov and centered mm in Figure ~~\ref{fig:pred_age_cont_plots}).
+In Figure ~\ref{fig:pred_age_cont_plots}b, the predicted overall mean is slightly higher than the observed mean (yellow horizontal line is slightly above the horizontal grey line), and vice versa in Figure ~\ref{fig:pred_age_cont_plots}c.
+%
+\begin{figure}[H]
+\begin{center}
+<<pred_age_cont_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
+print(pred_age_cont_plots)
+@
+\end{center}
+\caption{A comparison of variable effects and their corresponding CIs.
+Figure a) shows model with no interaction, Figure b) shows model with non-focal predictors interacting with each other, and Figure c) shows model with focal and non-focal predictors interacting.
+The dotted yellow and grey horizontal lines are the expected and observed/true means, respectively.
+The vertical grey line is the mean of the focal predictor (model center).
+The solid black, red (non-focal uncertainty excluded) and blue lines (all uncertainty included) are the variable effects while their corresponding CIs are the dotted black, red and blue lines, respectively.}
+\label{fig:pred_age_cont_plots}
+\end{figure}
+%
 

### 2021-09-23 R/utilities.R

@@ -1,6 +1,6 @@
 # Internal functions used to condition the model. Some of the condition code is based on R package effects (clean_model). Transfered and modified here because they are not exported in effects.
 	
-get_sderror <- function(mod, vcov., mm, col_mean, isolate, isolate.value, vareff_objects, x.var, typical, formula.rhs, zero_out_interaction, ...) {
+get_sderror <- function(mod, vcov., mm, col_mean, isolate, isolate.value, internal, vareff_objects, x.var, typical, formula.rhs, zero_out_interaction, ...) {
 	
 	if (is.null(vcov.)){
 		vc <- vcov(vareff_objects)

### 2021-09-27 variable_predictions.Rnw
Averaging over variables?

@@
 
-\title{Generating and visualizing predictor effects???}
+\title{Uncertainty propagation for variable (predictor) effects and bias correction in generalized linear (mixed) models???}
 
@@
 Predictions provide a great way to summarize what the regression model is telling us and are very useful for interpreting and visualizing models estimates using graphs.
-For example, in logistic regression models, the coefficient estimates are usually not easy to interpret, and we may want to explore the "effects" of covariates on predicted probabilities, or we may require predictions of responses at various levels of covariates.
+For example, in logistic regression models, the coefficient estimates are usually not easy to interpret, and we may want to explore the ``effects" of covariates on predicted probabilities, or predictions of responses at various levels of covariates.
 For example, as a way to plan, public health officials may want to know the predicted probability of improved water services in slums on the basis of the dwellers' household size, income, wealth index, etc.
 
-Both simple and generalized linear (mixed) models (GL(M)Ms) can examine very complex relationships, including nonlinear relationships between response and predictors, interactions between predictors (and via splines), and nonlinear transformations via link functions due to their flexibility.
+Both simple and generalized linear (mixed) models (GL(M)Ms) can examine very complex relationships, including nonlinear relationships between response and predictors, interactions between predictors (and via splines for example), and nonlinear transformations via link functions due to their flexibility.
 This flexibility comes at a cost, for example, complex multivariate models may risk misinterpretation, and miscalculation of quantities of interest.
@@
 
-When visually presented, predictions provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind transformations on the dependent variables whose estimates are usually, but not always, a subject to less clarity of interpretation.
-However, generating predictions and associated confidence intervals for regression models has a number of challenges.
+When visually presented, predictions provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind of transformations on the dependent variables whose estimates are usually, but not always, a subject to less clarity of interpretation.
+Further, generating predictions and associated confidence intervals for regression models has a number of challenges.
 In particular:
 \begin{enumerate}
-\item choice of representative values of \emph{focal} variable(s) and appropriate 'model center' for \emph{non-focal} variables in the case of multiple predictors
+\item choice of representative values of \emph{focal} variable(s) and appropriate ``model center" for \emph{non-focal} variables especially in multivariate models
 \item  propagation of uncertainty ([how] can we incorporate uncertainty in nonlinear components of (G)LMMs? Should we exclude variation in non-focal parameters?)
@@
 Most common way of dealing with the first challenge is taking unique levels of the focal predictor if discrete or taking appropriately sized quantiles (or bins) if continuous, and then calculating the predictions while holding non-focal predictors at their mean values.
-There a number of variations on description of this quantity -- \emph{predictor effect} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} and \emph{estimated marginal means} \citep{lenth2018package}.
-In this article, we refer to this quantity as \emph{variable (effect?) predictions} since this quantity should, for example, tell us what we would expect the outcome to be at a particular value or level of the variable.
-An alternative, is the population-based approach which involves computing the prediction over the population of the non-focal predictors and then averaging across the values of the focal predictor.
+This generates -- \emph{predictor effect} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} and \emph{estimated marginal means} \citep{lenth2018package}.
+In this article, we refer to this quantity as \emph{predictor effect} since this quantity should, for example, tell us what we would expect the presponse to be at a particular value or level of the predictor.
+An alternative to averaging the non-focal predictors, is the population-based approach which involves computing the prediction over the population of the non-focal predictors and then averaging across the values of the focal predictor.
 We give special focus to the second challenge later in the article.
@@
 When dealing with nonlinear link functions, the correct predictions for example, are even much harder to estimate.
-One approach is to make predictions on the transformed scale (linear predictor scale), and then back-transform back to the original scale.
+One approach is to make predictions on the transformed scale (linear predictor scale), and then back-transform to the original scale.
 However, the back-transformation may either result in biased predictions or requires some approximation.
@@
 
-The main purpose of this article is to discuss and implement various approaches for computing variable predictions and how to correctly compute the associated confidence intervals.
+The main purpose of this article is to discuss and implement various approaches for computing variable predictions and provide an alternative method for computing the associated confidence intervals.
 We further explore approaches for correcting bias in predictions for GL(M)Ms involving nonlinear link functions.
@@
 	, label = "tab:qoi_coefs"
+	, keep.stat="n"
 )

### 2021-09-28 variable_predictions.Rnw
Some edits on manuscript

@@
 The first one being the coefficient estimates.
-Others being predictors, model matrix, predicted values and marginal effects.
-\emph{predictors} are model input variables and is associated with one or more \emph{variables} (in polynomials, splines and models with interaction).
+Others being \emph{predictors}, \emph{model matrix}, \emph{predicted values} and \emph{marginal effects}.
+Predictors are model input variables and is associated with one or more \emph{variables} (in polynomials, splines and models with interaction).
 Model matrix refers to the design matrix whose rows include all combination of variables appearing in the interaction terms, along with the ``typical" values of the focal and non-focal predictors.
@@
 
-Figure ~\ref{fig:qoi_pred_plot} shows the comparison of \emph{variable prediction} and unconditional marginal effect of effect of age on household size, based on model estimates in Table ~\ref{tab:qoi_coefs}.
+Figure ~\ref{fig:qoi_pred_plot} shows the comparison of predictor effect and unconditional marginal effect of age on household size, based on model estimates in Table ~\ref{tab:qoi_coefs}.
 Since the model has no interactions, the relationship between the predicted household size and age is linear, hence the marginal effect of age is the slope ($\frac{\Delta \mathrm{hh size}}{\Delta \mathrm{age}}$) of the variable effect line and can be calculated irrespective of the values of wealth index.
@@
 %
-Variable effect, on the other hand, is the expected household size for a particular age, holding wealth index at its mean.
+Predictor effect, on the other hand, is the expected household size for a particular age, holding wealth index at its mean.
 In particular, the purpose and goal of a variable effect seems fairly straightforward; for specified values of (a) focal predictor(s), we want to give a point estimate and confidence intervals for the prediction of the model for a ``typical" (= random sample over the multivariate distribution of non-focal parameters) individual with those values of the predictors.
@@
 
-To get an intuition of how conditioning on the mean values of the non-focal predictors work, suppose we are interested in the variable predictions of a particular predictor (hence forth referred as \emph{focal} predictor otherwise \emph{non-focal}), $x_f$, from the set of predictors.
+To get an intuition of how conditioning on the mean values of the non-focal predictors work, suppose we are interested in predictor effects of a particular predictor (hence forth referred as \emph{focal} predictor otherwise \emph{non-focal}), $x_f$, from the set of predictors.
 To keep it simple, assume that the model has no interaction terms.
@@
 An alternative, save for later, formulation of variable effect involves, expressing the linear predictor as the sum of the focal and non-focal predictor linear predictor.
-In particular, $\eta^\star(x_f, \nset{{\bar{x}}}) = \beta_f x_f + \sum \nset{\beta} \nset{{\bar{x}^\star}}$, where $\nset{{\bar{x}^\star}}$ are the appropriately averaged entries of non-focal predictors and $\nset{x}$ is a vector of values of the non-focal predictors for a particular observation.
+In particular, $\eta^\star(x_f, \nset{{\bar{x}}}) = \beta_f x_f + \sum \nset{\beta} \nset{{\bar{x}^\star}}$, where $\nset{{\bar{x}^\star}}$ are the appropriately averaged entries of non-focal predictors and $x_f$ is a vector of values of the focal predictors for a particular observation.
 
@@
 However, to compute $\sigma$, $\Sigma$ is modified by \emph{zeroing-out} (the variance-covariance of all non-focal predictors are assigned zero) variances of non-focal terms.
-Although this is the simplest approach, it requires centering of continuous predictors prior to model fitting and proper handling of categorical predictors.
+Although this is the simplest approach, it requires centering of continuous predictors prior to model fitting and proper way to average categorical predictors.
 One way to center predictors is to create a design matrix, $\bX$, and then compute $\bX_{c} = \bX - \bar{\bX}$.

### 2021-09-28 variable_predictions_objs.R

@@ -59,7 +59,9 @@ focal_prop_cni <- mean(sim_df_cni$age)
 ##### Non-centered
 mod_cni <- lm(hhsize~age+wealthindex, sim_df_cni)
 ##### Centered predictor
-mod_cen_cni <- lm(hhsize~scale(age, scale=FALSE)+wealthindex, sim_df_cni)
+sim_df_cni_cen <- sim_df_cni
+sim_df_cni_cen$age <- drop(scale(sim_df_cni$age, scale=FALSE))
+mod_cen_cni <- lm(hhsize~age+wealthindex, sim_df_cni_cen)
 
 #### Binned data
 binned_df_cni <- binfun(mod_cni, "age", "wealthindex", bins=10)
@@ -79,10 +81,10 @@ print(pred_age_trad_cni_plot)
 ##### Centered
 ###### Variance-covariance
 pred_age_vcov_cni <- varpred(mod_cen_cni, "age", isolate=FALSE
-	, vcov. = zero_vcov(mod_cni, "age"), pop.ave="none"
+	, vcov. = zero_vcov(mod_cen_cni, "age"), pop.ave="none"
 	, modelname = "zero-vcov"
 )
-
+pred_age_vcov_cni$preds$age <- pred_age_vcov_cni$preds$age + focal_prop_cni
 ##### Centered model matrix
 pred_age_centered_cni <- varpred(mod_cni, "age", isolate=TRUE
 	, pop.ave="none", modelname = "centered mm"
@@ -112,7 +114,9 @@ focal_prop_wi_nf <- mean(sim_df_wi_nf$age)
 ##### Non-centered
 mod_wi_nf <- lm(hhsize~age+wealthindex+expenditure+wealthindex:expenditure, sim_df_wi_nf)
 ##### Centered predictor
-mod_cen_wi_nf <- lm(hhsize~scale(age, scale=FALSE)+wealthindex+expenditure+wealthindex:expenditure, sim_df_wi_nf)
+sim_df_wi_nf_cen <- sim_df_wi_nf
+sim_df_wi_nf_cen$age <- drop(scale(sim_df_wi_nf$age, scale=FALSE))
+mod_cen_wi_nf <- lm(hhsize~age+wealthindex+expenditure+wealthindex:expenditure, sim_df_wi_nf_cen)
 
 #### Binned data
[…]

### 2021-09-28 variable_predictions.Rnw
Simple lm seems to work...

@@
 This is the \emph{unconditional marginal effect} \citep{leeper2017interpreting} and it is constant across all the observations and levels of all other variables.
+As a result, the interpretation of interaction models differs in an important way from linear-additive regression models.
+To see this more clearly, compare the marginal effect of $x_1$ in the following two models:
+%
+\begin{align}
+y_1 &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \epsilon_1 \label{eq:simple_lm}\\
+y_2 &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_{12}x_1x_2 + \epsilon_2 \label{eq:simple_inter_lm}
+\end{align}
+%
+The marginal effect of $x_1$ in Equation ~\ref{eq:simple_lm} is $\frac{\partial y_1}{\partial x_1} = \beta_1$.
+On the other hand, the marginal effect of $x_1$ in Equation ~\ref{eq:simple_inter_lm} is given by $\frac{\partial y_2}{\partial x_1} = \beta_1 + \beta_{12}x_2$.
+In other words, if there are no interactions, the marginal effect of $x_1$ on $y_1$ is constant, while, if there are interactions in the model, the marginal effect of a change in $x_1$ on $y_1$ depends on the value of the \emph{conditioning} predictor $x_2$.
 

### 2021-09-29 variable_predictions.Rnw
Not everything seems to work (:

@@
 \begin{enumerate}
-\item choice of representative values of \emph{focal} variable(s) and appropriate ``model center" for \emph{non-focal} variables especially in multivariate models
+\item choice of representative values of \emph{focal} predictor(s) and appropriate ``model center" for \emph{non-focal} predictors especially in multivariate models
 \item  propagation of uncertainty ([how] can we incorporate uncertainty in nonlinear components of (G)LMMs? Should we exclude variation in non-focal parameters?)
@@
 
-The main purpose of this article is to discuss and implement various approaches for computing variable predictions and provide an alternative method for computing the associated confidence intervals.
+The main purpose of this article is to discuss and implement various approaches for computing predictor effect and provide an alternative method for computing the associated confidence intervals.
 We further explore approaches for correcting bias in predictions for GL(M)Ms involving nonlinear link functions.
@@
 In particular, a unit change in age is associated with a household that is \Sexpr{round(coef(qoi_mod1)[[2]], 3)} bigger.
-This is the \emph{unconditional marginal effect} \citep{leeper2017interpreting} and it is constant across all the observations and levels of all other variables.
+This is the \emph{unconditional marginal effect} \citep{leeper2017interpreting} and it is constant across all the observations and levels of all other predictors.
 As a result, the interpretation of interaction models differs in an important way from linear-additive regression models.
@@
 Figure ~\ref{fig:qoi_pred_plot} shows the comparison of predictor effect and unconditional marginal effect of age on household size, based on model estimates in Table ~\ref{tab:qoi_coefs}.
-Since the model has no interactions, the relationship between the predicted household size and age is linear, hence the marginal effect of age is the slope ($\frac{\Delta \mathrm{hh size}}{\Delta \mathrm{age}}$) of the variable effect line and can be calculated irrespective of the values of wealth index.
+Since the model has no interactions, the relationship between the predicted household size and age is linear, hence the marginal effect of age is the slope ($\frac{\Delta \mathrm{hh size}}{\Delta \mathrm{age}}$) of the predictor effect line and can be calculated irrespective of the values of wealth index.
 %
@@
 \end{center}
-\caption{Variable effect plot and marginal effect of age of household head on household size. For a linear model with no interaction, the marginal effect is the slope of the variable effect line.}
+\caption{Variable effect plot and marginal effect of age of household head on household size. For a linear model with no interaction, the marginal effect is the slope of the predictor effect line.}
 \label{fig:qoi_pred_plot}
@@
 Predictor effect, on the other hand, is the expected household size for a particular age, holding wealth index at its mean.
-In particular, the purpose and goal of a variable effect seems fairly straightforward; for specified values of (a) focal predictor(s), we want to give a point estimate and confidence intervals for the prediction of the model for a ``typical" (= random sample over the multivariate distribution of non-focal parameters) individual with those values of the predictors.
+In particular, the purpose and goal of a predictor effect seems fairly straightforward; for specified values of (a) focal predictor(s), we want to give a point estimate and confidence intervals for the prediction of the model for a ``typical" (= random sample over the multivariate distribution of non-focal parameters) individual with those values of the predictors.
 
@@
 For example fixing the values of \emph{non-focal} predictor(s) at some typical values -- typically determined by averaging (for now) in some meaningful way, for example, arithmetic mean  for continuous and average over the levels of the factors for categorical non-focal predictors.
-The easiest way to achieve this is by constructing $\bX^\star$ by averaging the columns of non-focal variables in model matrix $\bX$, and together with appropriately chosen values of focal predictor(s).
+The easiest way to achieve this is by constructing $\bX^\star$ by averaging the columns of non-focal predictors in model matrix $\bX$, and together with appropriately chosen values of focal predictor(s).
 
@@
 Let $\mathbf{X^*}$ be the model matrix, inheriting most of its key properties, for example transformations on predictors and interactions from the model matrix, $\mathbf{X}$.
-Then the prediction $\hat{\boldeta}^\star = \bX^\star\hat{\bbeta}$ is the variable effect for the focal predictor in question.
+Then the prediction $\hat{\boldeta}^\star = \bX^\star\hat{\bbeta}$ is the predictor effect for the focal predictor in question.
 
-An alternative, save for later, formulation of variable effect involves, expressing the linear predictor as the sum of the focal and non-focal predictor linear predictor.
+An alternative, save for later, formulation of predictor effect involves, expressing the linear predictor as the sum of the focal and non-focal predictor linear predictor.
 In particular, $\eta^\star(x_f, \nset{{\bar{x}}}) = \beta_f x_f + \sum \nset{\beta} \nset{{\bar{x}^\star}}$, where $\nset{{\bar{x}^\star}}$ are the appropriately averaged entries of non-focal predictors and $x_f$ is a vector of values of the focal predictors for a particular observation.
@@
 
-In models containing higher order terms such as interactions, splines, polynomials, etc., we adopt the method implemented in \pkg{effects} package \citep{fox2009effect} which involves combining the interactions with the main effects marginal to the interaction.
+Higher order terms such as interactions, splines, polynomials, etc., can be between the non-focal predictors or focal and non-focal predictor(s).
+In the former case, we treat the interactions as just another column in the variable space (of the model matrix).
+\textbf{Still figuring out the later case but currently treating everything as just additional column???}
 
@@
 
-In this section, we illustrate variable effects (together with the associated CI) in the context of linear models involving continuous and categorical predictors with or without interactions.
+In this section, we illustrate predictor effects (together with the associated CI) in the context of linear models involving continuous and categorical predictors with or without interactions.
 We compare the expected to the known simulation "truth".
 
-\subsection{Continuous predictors}
+\subsubsection{Continuous predictors}\label{sec:continuous_predictors}
 
@@
 \beta_0 &= 1.5 \\
-\beta_{\mathrm{A}} &= 0.2 \\
+\beta_{\mathrm{A}} &= 0.1 \\
 \beta_{\mathrm{W}} &= 2 \\
-i &= 1,\cdots, 1000
+i &= 1,\cdots, 100
 \end{align*}
@@
 \beta_0 &= 1.5 \\
-\beta_{\mathrm{A}} &= 0.8 \\
+\beta_{\mathrm{A}} &= 0.1 \\
 \beta_{\mathrm{W}} &= 2 \\
@@
 \beta_{\mathrm{WE}} &= 1 \\
-i &= 1,\cdots, 1000
+i &= 1,\cdots, 100
 \end{align*}
@@
 \beta_0 &= 1.5 \\
-\beta_{\mathrm{A}} &= 0.2 \\
+\beta_{\mathrm{A}} &= 0.1 \\
 \beta_{\mathrm{W}} &= 2 \\
@@
 \beta_{\mathrm{AW}} &= 1 \\
-i &= 1,\cdots, 1000
+i &= 1,\cdots, 100
 \end{align*}
@@
 
-We first fitted all the three models and in each case, compared three different variable effects together with the corresponding CIs -- 1) uncertainty due to non-focal predictors included (everything), 2) uncertainty due to non-focal predictors removed using variance-covariance matrix (zero-vcov), and 3) uncertainty due to non-focal predictors removed using centered model matrix (centered mm).
-The results are summarised in Figure ~\ref{fig:pred_age_cont_plots}.
+We first fitted all the three models and in each case, compared three different predictor effects together with the corresponding CIs -- 1) uncertainty due to non-focal predictors included (everything), 2) uncertainty due to non-focal predictors removed using variance-covariance matrix (zero-vcov), and 3) uncertainty due to non-focal predictors removed using centered model matrix (centered mm).
+The results are shown in Figure ~\ref{fig:pred_age_cont_plots}.
 
-Figure ~\ref{fig:pred_age_cont_plots}a represents variable effects for the model with no interaction, Figure ~\ref{fig:pred_age_cont_plots}b shows variable effects for model with non-focal predictors interacting with each other, and Figure ~\ref{fig:pred_age_cont_plots}c shows variable effects for the model with focal and non-focal predictors interacting.
-
-For perfect predictions, we expect the yellow and grey dotted lines to overlay each other and they should intersect with the black solid line and the vertical dotted grey line at the same point (or very close), Figure ~\ref{fig:pred_age_cont_plots}a.
-In other words, at the model center (mean of the focal predictor), generally, we expect the narrower CIs.
+Figure ~\ref{fig:pred_age_cont_plots}a represents predictor effects for the model with no interaction, Figure ~\ref{fig:pred_age_cont_plots}b shows predictor effects for model with non-focal predictors interacting with each other, and Figure ~\ref{fig:pred_age_cont_plots}c shows predictor effects for the model with focal and non-focal predictors interacting.
+For perfect predictions, we expect the yellow and grey dotted lines to overlay each other and they should intersect with the black solid line and the vertical dotted grey line at the same point (or very close).
+In other words, at the model center (mean of the focal predictor), generally, we expect the model to predict the average of the response variable and narrower CIs.
 If we properly anchor the model at its center (or any other appropriate value, assuming no interactions), we would expect the variance at the anchor to be zero, i.e., the CIs bands crosses at that point (see zero-cov and centered mm in Figure ~~\ref{fig:pred_age_cont_plots}).
-In Figure ~\ref{fig:pred_age_cont_plots}b, the predicted overall mean is slightly higher than the observed mean (yellow horizontal line is slightly above the horizontal grey line), and vice versa in Figure ~\ref{fig:pred_age_cont_plots}c.
+Although the overall predictor effects trend lines are different in each of the simulation, the expected household size for a household head with average age is not only the same in all the three simulations but also very close to the ``truth''.
 %
@@
 \end{center}
-\caption{A comparison of variable effects and their corresponding CIs.
+\caption{A comparison of predictor effects and their corresponding CIs.
 Figure a) shows model with no interaction, Figure b) shows model with non-focal predictors interacting with each other, and Figure c) shows model with focal and non-focal predictors interacting.
@@
 The vertical grey line is the mean of the focal predictor (model center).
-The solid black, red (non-focal uncertainty excluded) and blue lines (all uncertainty included) are the variable effects while their corresponding CIs are the dotted black, red and blue lines, respectively.}
+The solid black, red (non-focal uncertainty excluded) and blue lines (all uncertainty included) are the predictor effects while their corresponding CIs are the dotted black, red and blue lines, respectively.
+For properly generated predictor effects, we expect the yellow and the grey horizontal dotted lines to overlay each other and intersect with the black, blue, red solid trend lines, and the vertical dotted grey line at the same point (model center).}
 \label{fig:pred_age_cont_plots}
 \end{figure}
+%
+
+\subsubsection{Categorical predictors}
+
+Suppose, in Section ~\ref{sec:continuous_predictors}, in addition to the predictors already defined, we add gender and level of education of the household head.
+For example, we may want to know whether men (M) and females (F) differ in the relationship between age of the household head and the size of the household?
+How does households with male as their heads differ in size as compare to those with female heads?
+
+The first model in this case, slightly modifies the first model in Section ~\ref{sec:continuous_predictors} by adding gender of the household head.
+The proportion of households with males as their household heads is $60\%$:
+%
+\begin{align*}
+\mathrm{hh~size} &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{M}}\mathbf{I}_{\mathrm{Gender}_i=Male} + \epsilon_i \\
+\mathrm{Age}_i &\sim \mathrm{Normal}(0.2, 1) \\
+\mathbf{I}_{\mathrm{Gender}_i=Male} &\sim \mathrm{Sample}(\mathrm{Male}, 0.6) \\
+\epsilon_i &\sim \mathrm{Normal}(0, 1) \\
+\beta_0 &= 1.5 \\
+\beta_{\mathrm{A}} &= 0.8 \\
+\beta_{\mathrm{M}} &= 0.2 \\
+i &= 1,\cdots, 100
+\end{align*}
 %

### 2021-09-29 categorical_predictors.R

@@ -1,0 +1,75 @@
+library(shellpipes)
+library(vareffects); varefftheme()
+library(ggpubr)
+library(ggplot2)
+library(dplyr)
+
+commandEnvironments()
+makeGraphics()
+ 
+set.seed(9991)
+
+############################################################################################
+# Global variables
+N_obs <- 500
+gender_prop <- c(0.3, 0.7)
+gender_levels <- c("Female", "Male")
+
+############################################################################################
+# Section: 4.4.2 Categorical predictors
+############################################################################################
+
+### No interaction model
+
+#### Simulation
+sim_catni <- linearsim(N=N_obs
+	, form=~x1+x2
+	, betas=c(1.5,0.8,0.2)
+	, link_scale = TRUE
+)
+sim_df_catni <- sim_catni$data
+print(sim_catni$betas)
+head(sim_df_catni)
+colnames(sim_df_catni) <- c("age", "gender", "hhsize")
+focal_prop_catni <- (sim_df_catni
+	%>% group_by(gender)
+	%>% summarize(hhsize=mean(hhsize))
+)
+focal_prop_catni
+#### Model
+##### Non-centered
+mod_catni <- lm(hhsize~age+gender, sim_df_catni)
+summary(mod_catni)
+
+#### Variable effect
+##### Traditional CI
+pred_age_trad_catni <- varpred(mod_catni, "gender", isolate=FALSE, pop.ave="none", modelname="everything")
+pred_age_trad_catni_plot <- (plot(pred_age_trad_catni) 
+#	+ geom_hline(yintercept=true_prop_catni, lty=2, colour="grey")
+#	+ geom_hline(yintercept=mean(pred_age_trad_catni$preds$fit), lty=2, colour="blue")
+	+ geom_point(data=focal_prop_catni, aes(x=gender, y=hhsize), colour="red")
+	+ labs(y="Predicted household size")
+)
+print(pred_age_trad_catni_plot)
+
+quit()
+##### Centered model matrix
+pred_age_centered_catni <- varpred(mod_catni, "age", isolate=TRUE
+	, pop.ave="none", modelname = "centered mm"
+)
+pred_age_all_catni <- pred_age_centered_catni
+pred_age_all_catni$preds <- do.call("rbind", list(pred_age_trad_catni$preds, pred_age_centered_catni$preds, pred_age_vcov_catni$preds))
[…]

### 2021-09-30 variable_predictions.Rnw
Categorical variables!

@@
 load("variable_predictions_objs.rda")
+load("categorical_predictors.rda")
 @
@@
 Suppose, in Section ~\ref{sec:continuous_predictors}, in addition to the predictors already defined, we add gender and level of education of the household head.
-For example, we may want to know whether men (M) and females (F) differ in the relationship between age of the household head and the size of the household?
-How does households with male as their heads differ in size as compare to those with female heads?
+For example, we may want to know whether men (M) and females (F) differ in the relationship between age of the household head and the size of the household -- how does households with male as their heads differ in size as compare to those with female heads?
 
@@
 %
+The second model has interaction between non-focal predictors (\code{age} and \code{wealth index}), while the third model has interaction between focal (\code{gender} and \code{age}).
+The results are shown in Figure ~\ref{fig:pred_gender_cat_plots}, which compares the predictors effects (either centered or all uncertainty included -- everything) for no interaction model (a), interaction between non-focal predictors (b) and interaction between focal and non-focal predictor (c).
+\textbf{For each gender, we expect the observed average household size ("truth" -- red dots), to be very close to the estimated predictor effects (blue and black dots and lines)??}
+%
+\begin{figure}[H]
+\begin{center}
+<<pred_gender_cat_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
+print(pred_gender_cat_plots)
+@
+\end{center}
+\caption{A comparison of predictor effects and their corresponding CIs.
+Figure a) shows model with no interaction, Figure b) shows model with non-focal predictors interacting with each other, and Figure c) shows model with focal and non-focal predictors interacting.
+The red dots are the observed observed/true means.
+The black (non-focal uncertainty excluded) and blue solid dots (all uncertainty included) are the predictor effects with their corresponding CIs, respectively.
+\textbf{For properly generated predictor effects, we expect the black, blue and red dots to be as close as possible???}.}
+\label{fig:pred_gender_cat_plots}
+\end{figure}
+%
 

### 2021-09-30 categorical_predictors.R

@@ -11,9 +11,7 @@ set.seed(9991)
 
 ############################################################################################
 # Global variables
-N_obs <- 500
-gender_prop <- c(0.3, 0.7)
-gender_levels <- c("Female", "Male")
+N_obs <- 100
 
 ############################################################################################
 # Section: 4.4.2 Categorical predictors
@@ -23,14 +21,20 @@ gender_levels <- c("Female", "Male")
 
 #### Simulation
 sim_catni <- linearsim(N=N_obs
-	, form=~x1+x2
-	, betas=c(1.5,0.8,0.2)
-	, link_scale = TRUE
+	, form=~1+x1+x2
+	, betas=c(1.5, 0.5, 0.2)
+	, pgausian=list(p=1, fun=rnorm, mean=0, sd=1)
+	, pcat=list(p=1, fun=sample, nlevels=2
+		, labels=c("Male", "Female")
+		, prob=c(0.4, 0.6)
+		, replace=TRUE
+	)
+	, link_scale=TRUE
+	, vnames=c("age", "gender", "hhsize")
 )
 sim_df_catni <- sim_catni$data
 print(sim_catni$betas)
 head(sim_df_catni)
-colnames(sim_df_catni) <- c("age", "gender", "hhsize")
 focal_prop_catni <- (sim_df_catni
 	%>% group_by(gender)
 	%>% summarize(hhsize=mean(hhsize))
@@ -43,33 +47,166 @@ summary(mod_catni)
 
 #### Variable effect
 ##### Traditional CI
-pred_age_trad_catni <- varpred(mod_catni, "gender", isolate=FALSE, pop.ave="none", modelname="everything")
-pred_age_trad_catni_plot <- (plot(pred_age_trad_catni) 
-#	+ geom_hline(yintercept=true_prop_catni, lty=2, colour="grey")
-#	+ geom_hline(yintercept=mean(pred_age_trad_catni$preds$fit), lty=2, colour="blue")
+pred_gender_trad_catni <- varpred(mod_catni, "gender", isolate=FALSE
+	, pop.ave="none", modelname="everything"
+)
+pred_gender_trad_catni_plot <- (plot(pred_gender_trad_catni) 
 	+ geom_point(data=focal_prop_catni, aes(x=gender, y=hhsize), colour="red")
 	+ labs(y="Predicted household size")
 )
-print(pred_age_trad_catni_plot)
[…]

### 2021-09-30 variable_predictions_objs.R

@@ -19,11 +19,15 @@ beta_age <- 0.1 # Focal predictor effect
 ############################################################################################
 
 ## Coefficient estimate
-qoi_df <- simplesim(N=1000, link_scale=TRUE)
+qoi_df <- linearsim(N=1000, form=~1+x1+x2
+	, betas=c(1.5, 1, 2)
+	, pgausian=list(p=2, fun=rnorm, mean=c(0.2,0), sd=1)
+	, pcat=list(p=0)
+	, link_scale = TRUE
+)$data
 colnames(qoi_df) <- c("age", "wealthindex", "hhsize")
 qoi_mod1 <- lm(hhsize~age+wealthindex, qoi_df)
 qoi_mod2 <- lm(hhsize~age*wealthindex, qoi_df)
-
 
 ## Prediction plot
 qoi_age_pred <- varpred(qoi_mod1, "age")
@@ -51,7 +55,12 @@ print(qoi_age_pred_plot)
 ### No interaction model
 
 #### Simulation
-sim_df_cni <- simplesim(N=N_obs, beta1=beta_age, link_scale=TRUE)
+sim_df_cni <- linearsim(N=N_obs, form=~1+x1+x2
+	, betas = c(1.5, 0.1, 2)
+	, pgausian=list(p=2,fun=rnorm, mean=c(0.2,0), sd=c(1,1))
+	, pcat=list(p=0)
+	, link_scale=TRUE
+)$data
 colnames(sim_df_cni) <- c("age", "wealthindex", "hhsize")
 true_prop_cni <- mean(sim_df_cni$hhsize)
 focal_prop_cni <- mean(sim_df_cni$age)
@@ -106,7 +115,12 @@ print(pred_age_all_cni_plot)
 ### With interaction between non-focal
 
 #### Simulation
-sim_df_wi_nf <- interactionsim(N=N_obs, beta1=beta_age, link_scale=TRUE)
+sim_df_wi_nf <- linearsim(N=N_obs, form=~1+x1+x2+x3+x2:x3
+	, betas=c(1.5,0.1,2,1.5,1)
+	, pgausian=list(p=3,fun=rnorm, mean=c(0.2,0,0), sd=c(1,1,1))
+	, pcat=list(p=0)
+	, link_scale=TRUE
+)$data
 colnames(sim_df_wi_nf) <- c("age", "wealthindex", "expenditure", "hhsize")
[…]

### 2021-10-06 variable_predictions.Rnw
More simulation examples

@@
 load("categorical_predictors.rda")
+load("lme_random_intercept.rda")
+load("multiple_outcomes_preds.rda")
 @
@@
 %
-\begin{align*}
-\mathrm{hh~size} &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \epsilon_i \\
-\mathrm{Age}_i &\sim \mathrm{Normal}(0.2, 1) \\
-\mathrm{Wealthindex}_i &\sim \mathrm{Normal}(0, 1) \\
-\epsilon_i &\sim \mathrm{Normal}(0, 1) \\
-\beta_0 &= 1.5 \\
-\beta_{\mathrm{A}} &= 0.1 \\
-\beta_{\mathrm{W}} &= 2 \\
+\begin{align}\label{sim:lm_no_interaction}
+\mathrm{hh~size} &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \epsilon_i \nonumber\\
+\mathrm{Age}_i &\sim \mathrm{Normal}(0.2, 1) \nonumber\\
+\mathrm{Wealthindex}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
+\epsilon_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
+\beta_0 &= 1.5 \nonumber\\
+\beta_{\mathrm{A}} &= 0.1 \nonumber\\
+\beta_{\mathrm{W}} &= 2 \nonumber\\
 i &= 1,\cdots, 100
-\end{align*}
+\end{align}
 %
@@
 %
-\begin{align*}
-\mathrm{hh~size} &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \beta_{\mathrm{E}}\mathrm{Expenditure}_i\\ 
-&+ \beta_{\mathrm{WE}}\mathrm{Wealthindex}_i*\mathrm{Expenditure}_i + \epsilon_i \\
-\mathrm{Age}_i &\sim \mathrm{Normal}(0.2, 1) \\
-\mathrm{Wealthindex}_i &\sim \mathrm{Normal}(0, 1) \\
-\mathrm{Expenditure}_i &\sim \mathrm{Normal}(0, 1) \\
-\epsilon_i &\sim \mathrm{Normal}(0, 1) \\
-\beta_0 &= 1.5 \\
-\beta_{\mathrm{A}} &= 0.1 \\
-\beta_{\mathrm{W}} &= 2 \\
-\beta_{\mathrm{E}} &= 1.5 \\
-\beta_{\mathrm{WE}} &= 1 \\
+\begin{align}\label{sim:lm_nf_interaction}
+\mathrm{hh~size} &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \beta_{\mathrm{E}}\mathrm{Expenditure}_i \nonumber\\ 
+&+ \beta_{\mathrm{WE}}\mathrm{Wealthindex}_i*\mathrm{Expenditure}_i + \epsilon_i \nonumber\\
+\mathrm{Age}_i &\sim \mathrm{Normal}(0.2, 1) \nonumber\\
+\mathrm{Wealthindex}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
+\mathrm{Expenditure}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
+\epsilon_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
+\beta_0 &= 1.5 \nonumber\\
+\beta_{\mathrm{A}} &= 0.1 \nonumber\\
+\beta_{\mathrm{W}} &= 2 \nonumber\\
+\beta_{\mathrm{E}} &= 1.5 \nonumber\\
+\beta_{\mathrm{WE}} &= 1 \nonumber\\
 i &= 1,\cdots, 100
-\end{align*}
+\end{align}
 %
@@
 %
-\begin{align*}
-\mathrm{hh~size} &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \beta_{\mathrm{E}}\mathrm{Expenditure}_i\\ 
-&+ \beta_{\mathrm{AW}}\mathrm{Age}_i*\mathrm{Wealthindex}_i + \epsilon_i \\
-\mathrm{Age}_i &\sim \mathrm{Normal}(0.2, 1) \\
-\mathrm{Wealthindex}_i &\sim \mathrm{Normal}(0, 1) \\
-\mathrm{Expenditure}_i &\sim \mathrm{Normal}(0, 1) \\
-\epsilon_i &\sim \mathrm{Normal}(0, 1) \\
-\beta_0 &= 1.5 \\
-\beta_{\mathrm{A}} &= 0.1 \\
-\beta_{\mathrm{W}} &= 2 \\
-\beta_{\mathrm{E}} &= 1.5 \\
-\beta_{\mathrm{AW}} &= 1 \\
+\begin{align}\label{sim:lm_f_interaction}
+\mathrm{hh~size} &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \beta_{\mathrm{E}}\mathrm{Expenditure}_i \nonumber\\ 
+&+ \beta_{\mathrm{AW}}\mathrm{Age}_i*\mathrm{Wealthindex}_i + \epsilon_i \nonumber\\
+\mathrm{Age}_i &\sim \mathrm{Normal}(0.2, 1) \nonumber\\
+\mathrm{Wealthindex}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
+\mathrm{Expenditure}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
+\epsilon_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
+\beta_0 &= 1.5 \nonumber\\
+\beta_{\mathrm{A}} &= 0.1 \nonumber\\
+\beta_{\mathrm{W}} &= 2 \nonumber\\
+\beta_{\mathrm{E}} &= 1.5 \nonumber\\
+\beta_{\mathrm{AW}} &= 1 \nonumber\\
 i &= 1,\cdots, 100
-\end{align*}
+\end{align}
 %
@@
 
-\subsubsection{Categorical predictors}
+\subsubsection{Categorical predictors}\label{sec:categorical_predictors}
 
@@
 %
-\begin{align*}
-\mathrm{hh~size} &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{M}}\mathbf{I}_{\mathrm{Gender}_i=Male} + \epsilon_i \\
-\mathrm{Age}_i &\sim \mathrm{Normal}(0.2, 1) \\
-\mathbf{I}_{\mathrm{Gender}_i=Male} &\sim \mathrm{Sample}(\mathrm{Male}, 0.6) \\
-\epsilon_i &\sim \mathrm{Normal}(0, 1) \\
-\beta_0 &= 1.5 \\
-\beta_{\mathrm{A}} &= 0.8 \\
-\beta_{\mathrm{M}} &= 0.2 \\
+\begin{align}\label{sim:lm_no_interaction_cat}
+\mathrm{hh~size} &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{M}}{\mathrm{Gender}_i} + \epsilon_i \nonumber\\
+\mathrm{Age}_i &\sim \mathrm{Normal}(0.2, 1) \nonumber\\
+\mathrm{Gender}_i &\sim \mathrm{Sample}(\mathrm{Gender_i}, p) \nonumber\\
+p &= \begin{cases}
+   0.4 & \mathrm{Gender_{i=Female}} \nonumber\\
+   0.6 & \mathrm{Gender_{i=Male}} \nonumber\\
+  \end{cases} \nonumber\\
+\epsilon_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
+\beta_0 &= 1.5 \nonumber\\
+\beta_{\mathrm{A}} &= 0.8 \nonumber\\
+\beta_{\mathrm{M}} &= 0.2 \nonumber\\
 i &= 1,\cdots, 100
-\end{align*}
+\end{align}
 %
@@
 
+\subsubsection{Linear mixed effect model}
+
+We now consider linear models with additional random effects components. In the simplest case, we start with a single grouping factor as the random effect components.
+
+Suppose, in the simulation described in Section ~\ref{sec:stats_background}, the observations are  recorded at least once per household.
+In particular, let $\mathrm{H}$ be the number of households indexed by the grouping factor, and $\mathrm{h}[i]$ be the household of the $i$th observation.
+We simulate three different models incorporating this kind of formulation -- first, one grouping factor-one predictor-linear random effect:
+%
+\begin{align}\label{sim:lme_one_pred}
+\mathrm{hh~size} &= \beta_0 + \alpha_{0,\mathrm{h}[i]} + \beta_{\mathrm{A}}\mathrm{Age}_i + \epsilon_i \nonumber\\
+\alpha_{0,\mathrm{h}} &\sim \mathrm{Normal}(0, 1), ~~ h=1, \cdots, \mathrm{H} \nonumber\\
+\mathrm{Age}_i &\sim \mathrm{Normal}(0.2, 1) \nonumber\\
+\epsilon_i &\sim \mathrm{Normal}(0, 2) \nonumber\\
+\beta_0 &= 1.5 \nonumber\\
+\beta_{\mathrm{A}} &= 0.1 \nonumber\\
+i &= 1,\cdots, 100
+\end{align}
+%
+The second and third models add the random effects, $\alpha_{\mathrm{h}[i]}$, to models ~\ref{sim:lm_nf_interaction} and ~\ref{sim:lm_f_interaction}, respectively.
+%
+\begin{figure}[H]
+\begin{center}
+<<pred_age_cont_lme_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
+print(pred_age_cont_lme_plots)
+@
+\end{center}
+\caption{A comparison of predictor effects and their corresponding CIs.
+Figure a) shows model with no interaction, Figure b) shows model with non-focal predictors interacting with each other, and Figure c) shows model with focal and non-focal predictors interacting.
+The dotted yellow and grey horizontal lines are the expected and observed/true means, respectively.
+The vertical grey line is the mean of the focal predictor (model center).
+The solid black, red (non-focal uncertainty excluded) and blue lines (all uncertainty included) are the predictor effects while their corresponding CIs are the dotted black, red and blue lines, respectively.
+For properly generated predictor effects, we expect the yellow and the grey horizontal dotted lines to overlay each other and intersect with the black, blue, red solid trend lines, and the vertical dotted grey line at the same point (model center).}
+\label{fig:pred_age_cont_lme_plots}
+\end{figure}
+%
+
+\subsubsection{Multiple outcomes}
+
+We now consider a bit more complicated model involving more two outcome variables (services) -- household size and rental price.
+In particular, we simulate an intercept only multivariate random effect model to, \emph{jointly}, predict household size and rental price based on age and wealth index.
+Each household has its own correlated effect on the intercepts (random-intercept).
+In addition, we add latent variable to account for unobserved correlations.
+%
+\begin{align}\label{sim:lme_multi_outcome}
+\mathrm{y}_{\mathrm{h}[i]\mathrm{s}} &= \beta_{0,s} + \alpha_{0,\mathrm{h}[i]\mathrm{s}} + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \beta_{\mathrm{L}}\mathrm{Latent}_i + \epsilon_{i\mathrm{s}} \nonumber\\
+\alpha_{0,\mathrm{h}\mathrm{s}} &\sim \mathrm{Normal}(0, 2), ~~ h=1, \cdots, \mathrm{H} \nonumber\\
+\mathrm{Age}_i &\sim \mathrm{Normal}(0.2, 1) \nonumber\\
+\mathrm{Wealthindex}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
+\mathrm{Latent}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
+\epsilon_{i\mathrm{s}} &\sim \mathrm{Normal}(0, 1) \nonumber\\
+\beta_0 &= 1.5 \nonumber\\
+\beta_{\mathrm{A}} &= 0.1 \nonumber\\
+\beta_{\mathrm{W}} &= 2 \nonumber\\
+\beta_{\mathrm{L}} &= 1 \nonumber\\
+i &= 1,\cdots, 100
+\end{align}
+%
+Where:
+\begin{itemize}
+\item $\beta_{0,s}$ \textbf{is the service specific effect???}, i.e., baseline household size and rental price
+\item $\alpha_{0,\mathrm{h}\mathrm{s}}$ household specific random intercepts, i.e., captures the offset from the overall baseline (mean)
+\end{itemize}
+
+%
+\begin{figure}[H]
+\begin{center}
+<<pred_cont_joint_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
+print(pred_cont_joint_plots)
+@
+\end{center}
+\caption{A comparison of predictor effects and their corresponding CIs.
+Figure a) shows model with no interaction, Figure b) shows model with non-focal predictors interacting with each other, and Figure c) shows model with focal and non-focal predictors interacting.
+The dotted yellow and grey horizontal lines are the expected and observed/true means, respectively.
+The vertical grey line is the mean of the focal predictor (model center).
+The solid black, red (non-focal uncertainty excluded) and blue lines (all uncertainty included) are the predictor effects while their corresponding CIs are the dotted black, red and blue lines, respectively.
+For properly generated predictor effects, we expect the yellow and the grey horizontal dotted lines to overlay each other and intersect with the black, blue, red solid trend lines, and the vertical dotted grey line at the same point (model center).}
+\label{fig:pred_cont_joint_plots}
+\end{figure}
+%
+
 \section{Bias correction}\label{sec:bias_correction}

### 2021-10-06 R/plotsfuns.R

@@ -34,6 +34,7 @@ plot.vareffects <- function(x, ..., xlabs = NULL, ylabs = NULL, pos = 0.5, ci = 
 	if (!is.null(focal))	n.focal <- length(focal) else {n.focal <- 1L; focal <- "xvar"}
 	x.var <- attr(df, "x.var")
 	if (is.null(x.var)) x.var <- focal[[1]]
+	add_col <- df$.varpred
 	if (n.focal>1L) {
 		non.focal <- focal[!focal %in% x.var]
 		df[, non.focal] <- sapply(non.focal, function(x){
@@ -42,7 +43,13 @@ plot.vareffects <- function(x, ..., xlabs = NULL, ylabs = NULL, pos = 0.5, ci = 
 			ll <- paste0(x, ": ", xx)
 			return(ll)
 		})	
-		gform <- as.formula(paste0(".~", paste0(non.focal, collapse="+")))
+		if (is.null(add_col)) ff <- ".~" else ff <- ".varpred~"
+		gform <- as.formula(paste0(ff, paste0(non.focal, collapse="+")))
+	} else {
+		if (!is.null(add_col)) {
+			ff <- "~.varpred"
+			gform <- as.formula(ff)
+		}
 	}
 	if (is.null(xlabs)) xlabs <- x.var
 	if (is.null(ylabs)) ylabs <- attr(df, "response")
@@ -87,7 +94,7 @@ plot.vareffects <- function(x, ..., xlabs = NULL, ylabs = NULL, pos = 0.5, ci = 
 		}
 	}
 
-	if (n.focal>1L) {
+	if (n.focal>1L || !is.null(add_col)) {
 		p2 <- p2 + facet_wrap(gform, scales=facet_scales, ncol=facet_ncol)#, labeller = label_parsed)
 	}
 	return(p2)

### 2021-10-06 multiple_outcomes_model.R

@@ -1,0 +1,29 @@
+library(shellpipes)
+library(dplyr)
+library(glmmTMB)
+
+commandEnvironments()
+makeGraphics()
+ 
+set.seed(9991)
+
+############################################################################################
+## Joint model
+############################################################################################
+
+## Observed averages
+### Outcomes
+true_prop_df <- (sim_df_cont_joint_long
+	%>% group_by(services)
+	%>% summarize(fit=mean(values))
+)
+print(true_prop_df)
+
+## Models
+mod_cont_joint <- glmmTMB(values~-1+services+services:(age+wealthindex)+(services-1|hhid)
+	, data=sim_df_cont_joint_long
+	, family=gaussian()
+)
+summary(mod_cont_joint)
+
+saveEnvironment()

### 2021-10-06 variable_predictions_funs.R

@@ -2,10 +2,10 @@
 ### Default: Continuous outcome 2 and 1 categorical predictors
 ### form : specify model using formula
 
-linearsim <- function(N=1000, form=~1+x1+x2+x3
-	, betas=NULL, pgausian=list(p=2, fun=rnorm, mean=0, sd=1)
+linearsim <- function(nHH=1000, perHH=1, form=~1+x1+x2+x3
+	, hhSD=2, betas=NULL, pgausian=list(p=2, fun=rnorm, mean=0, sd=1)
 	, pcat=list(p=1, fun=sample, nlevels=2, labels=NULL, prob=NULL, replace=TRUE)
-	, vnames=NULL, link_scale=FALSE) {
+	, noutcomes=1, vnames=NULL, link_scale=FALSE) {
 	
 	Terms <- attr(terms(form), "term.labels")
 	nbetas <- length(Terms)
@@ -19,11 +19,19 @@ linearsim <- function(N=1000, form=~1+x1+x2+x3
 		stop(paste0("\nNumber of predictors in form should be same as sum of p in pgausian and pcat: ", nvars, " != ", p))
 	}
 
-	mu <- pgausian$mean
-	.sd <- pgausian$sd
+	if (noutcomes<1)stop("Number of outcome variables should be  >= 1")
+	if (perHH < 1)stop("perHH >= 1")
+	N <- nHH * perHH
+
 	pg <- pgausian$p
+	mu <- pgausian$mean
+	if (is.null(mu)) mu <- rep(0, pg)
+	.sd <- pgausian$sd
+	if(is.null(.sd)) sd <- 1
+
 	pgausian$p <- NULL
 	gfun <- pgausian$fun
+	if (is.null(gfun)) gfun <- rnorm
 	pgausian$fun <- NULL
 	pgausian$n <- N
 	if (length(mu)==1) {
@@ -44,10 +52,17 @@ linearsim <- function(N=1000, form=~1+x1+x2+x3
 	pcat$p <- NULL
 	if (cp > 0) {
 		nlevels <- pcat$nlevels
+		if (is.null(nlevels)) {
+			nlevels <- rep(2, cp)
+			pcat$labels <- NULL
+			pcat$prob <- NULL
+			pcat$replace <- TRUE
+		}
 		if (length(nlevels)!=cp)
 			stop("\n p in pcat must be equal to the length of nlevels")
 		Xcat <- list()
 		cfun <- pcat$fun
+		if (is.null(cfun)) cfun <- sample
[…]

### 2021-10-06 variable_predictions.Rnw
Autocommit

@@
 Predictions provide a great way to summarize what the regression model is telling us and are very useful for interpreting and visualizing models estimates using graphs.
-For example, in logistic regression models, the coefficient estimates are usually not easy to interpret, and we may want to explore the ``effects" of covariates on predicted probabilities, or predictions of responses at various levels of covariates.
-For example, as a way to plan, public health officials may want to know the predicted probability of improved water services in slums on the basis of the dwellers' household size, income, wealth index, etc.
+For example, in logistic regression models, the coefficient estimates are usually not easy to interpret, and we may want to explore the ``effects" of covariates on predicted probabilities, or predictions of responses at various levels of covariates -- \emph{predictor effects}.
+For example, as a way to plan, public health officials may want to know the ``effects" of household income, wealth index, etc, on the predicted probability of having improved water services among slum dwellers.
 
@@
 
-When visually presented, predictions provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind of transformations on the dependent variables whose estimates are usually, but not always, a subject to less clarity of interpretation.
-Further, generating predictions and associated confidence intervals for regression models has a number of challenges.
+When visually presented, predictor effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind of transformations on the predictors whose estimates are usually, but not always, a subject to less clarity of interpretation.
+Further, generating predictor effects together with the associated confidence intervals for regression models has a number of challenges.
 In particular:
@@
 
-Most common way of dealing with the first challenge is taking unique levels of the focal predictor if discrete or taking appropriately sized quantiles (or bins) if continuous, and then calculating the predictions while holding non-focal predictors at their mean values.
+Most common way of dealing with the first challenge is taking unique levels of the focal predictor if discrete or taking appropriately sized quantiles (or bins) if continuous, and then calculating the predictions while holding non-focal predictors at their typical values (e.g., averages).
 This generates -- \emph{predictor effect} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} and \emph{estimated marginal means} \citep{lenth2018package}.
-In this article, we refer to this quantity as \emph{predictor effect} since this quantity should, for example, tell us what we would expect the presponse to be at a particular value or level of the predictor.
+In this article, we refer to this quantity as \emph{predictor effect} since it should, for example, tell us what we would expect the presponse to be at a particular value or level of the predictor.
 An alternative to averaging the non-focal predictors, is the population-based approach which involves computing the prediction over the population of the non-focal predictors and then averaging across the values of the focal predictor.
-We give special focus to the second challenge later in the article.
+Dealing with second challenge is discussed in Section ~\ref{sec:uncertainity_propagation}
 
@@
 However, the back-transformation may either result in biased predictions or requires some approximation.
-In particular, Jensen's inequality and bias in expected mean prediction induced by nonlinear transformation of the response variable can lead inaccurate predictions.
+In particular, bias in expected mean prediction induced by nonlinear transformation of the response variable can lead inaccurate predictions.
 
-The main purpose of this article is to discuss and implement various approaches for computing predictor effect and provide an alternative method for computing the associated confidence intervals.
+The main purpose of this article is to discuss and implement various approaches for computing predictor effects and provide an alternative method for computing the associated confidence intervals.
 We further explore approaches for correcting bias in predictions for GL(M)Ms involving nonlinear link functions.
@@
 Several quantities of interest may be derived from regression models.
-The first one being the coefficient estimates.
-Others being \emph{predictors}, \emph{model matrix}, \emph{predicted values} and \emph{marginal effects}.
+The first one is the coefficient estimates.
+Others are \emph{predictors}, \emph{model matrix}, \emph{predicted values} and \emph{marginal effects}.
 Predictors are model input variables and is associated with one or more \emph{variables} (in polynomials, splines and models with interaction).
@@
 
-To illustrate some of the concepts, consider results from, hypothetical  simulated example, regression of household size as function of household wealth index and age of household head, and the second model involving these variables plus the interaction between them.
+To illustrate some of the concepts, consider results from, hypothetical  simulated example, regression of household size as function of household wealth index and age of household head, and the second model involving these predictors plus the interaction between them.
 

### 2021-10-06 variable_predictions_objs.R

@@ -19,8 +19,8 @@ beta_age <- 0.1 # Focal predictor effect
 ############################################################################################
 
 ## Coefficient estimate
-qoi_df <- linearsim(nHH=1000, form=~1+x1+x2
-	, betas=c(1.5, 1, 2)
+qoi_df <- linearsim(nHH=1000, form=~1+x1*x2
+	, betas=c(1.5, 1, 2, 0.5)
 	, pgausian=list(p=2, fun=rnorm, mean=c(0.2,0), sd=1)
 	, pcat=list(p=0)
 	, link_scale = TRUE

### 2021-10-07 compare_emmeans_varpred.rmd
Emmeans vs varpreds

---
title: "Bias correction in GLMs"
author: Bicko, Jonathan & Ben
date: "2021 Oct 06 (Wed)"
compact-title: false
output:
  pdf_document:
    extra_dependencies: ["float"]
bibliography: bias_correction.bib
link-citations: yes
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(fig.pos = "H", out.extra = "", message=FALSE, warning=FALSE, echo=FALSE)

library(shellpipes)
library(glmmTMB)
library(dplyr)
library(ggplot2)
library(vareffects); varefftheme()
library(emmeans)

commandEnvironments()
makeGraphics()

set.seed(2121)

vplots <- vareffects:::plot.vareffects

```

\section{Continuous predictors}

\subsection{No interaction}

```{r cont_no_interaction}
quants <- seq(0, 1, length.out=100)
focal_levels <- quantile(sim_df_cni$age, quants)
truth = c(truth=mean(sim_df_cni$hhsize))
print(truth)
compare_cni_plots <- (combinepreds(mod_cni, focal="age"
		, funs=c("emmeans", "varpred")
		, at=list(age=focal_levels)
		, x.var="age"
		, plotit=TRUE
	)
	+ geom_vline(aes(xintercept=mean(focal_levels)), lty=2, colour="grey")
)
print(compare_cni_plots)
```

\subsection{Interaction between non-focal predictors}

```{r cont_nwi_nf_interaction}
quants <- seq(0, 1, length.out=100)
focal_levels <- quantile(sim_df_wi_nf$age, quants)
truth = c(truth=mean(sim_df_wi_nf$hhsize))
print(truth)
compare_wi_nf_plots <- (combinepreds(mod_wi_nf, focal="age"
		, funs=c("emmeans", "varpred")
		, at=list(age=focal_levels)
		, x.var="age"
		, plotit=TRUE
	)
	+ geom_vline(aes(xintercept=mean(focal_levels)), lty=2, colour="grey")
)
print(compare_wi_nf_plots)
```

### 2021-10-07 variable_predictions.Rnw
Emmeans vs varpreds

@@
 An alternative to averaging the non-focal predictors, is the population-based approach which involves computing the prediction over the population of the non-focal predictors and then averaging across the values of the focal predictor.
-Dealing with second challenge is discussed in Section ~\ref{sec:uncertainity_propagation}
+Dealing with second challenge is discussed in Section~\ref{sec:uncertainity_propagation}
 
@@
 
-In simple linear models with no interaction terms, the default out for the coefficient estimates in Table ~\ref{tab:qoi_coefs} model (1) is simple and directly interpretable as the expected change in household size for a unit change in age.
+In simple linear models with no interaction terms, the default out for the coefficient estimates in Table~\ref{tab:qoi_coefs} model (1) is simple and directly interpretable as the expected change in household size for a unit change in age.
 In particular, a unit change in age is associated with a household that is \Sexpr{round(coef(qoi_mod1)[[2]], 3)} bigger.
@@
 %
-The marginal effect of $x_1$ in Equation ~\ref{eq:simple_lm} is $\frac{\partial y_1}{\partial x_1} = \beta_1$.
-On the other hand, the marginal effect of $x_1$ in Equation ~\ref{eq:simple_inter_lm} is given by $\frac{\partial y_2}{\partial x_1} = \beta_1 + \beta_{12}x_2$.
+The marginal effect of $x_1$ in Equation~\ref{eq:simple_lm} is $\frac{\partial y_1}{\partial x_1} = \beta_1$.
+On the other hand, the marginal effect of $x_1$ in Equation~\ref{eq:simple_inter_lm} is given by $\frac{\partial y_2}{\partial x_1} = \beta_1 + \beta_{12}x_2$.
 In other words, if there are no interactions, the marginal effect of $x_1$ on $y_1$ is constant, while, if there are interactions in the model, the marginal effect of a change in $x_1$ on $y_1$ depends on the value of the \emph{conditioning} predictor $x_2$.
 
-Figure ~\ref{fig:qoi_pred_plot} shows the comparison of predictor effect and unconditional marginal effect of age on household size, based on model estimates in Table ~\ref{tab:qoi_coefs}.
+Figure~\ref{fig:qoi_pred_plot} shows the comparison of predictor effect and unconditional marginal effect of age on household size, based on model estimates in Table~\ref{tab:qoi_coefs}.
 Since the model has no interactions, the relationship between the predicted household size and age is linear, hence the marginal effect of age is the slope ($\frac{\Delta \mathrm{hh size}}{\Delta \mathrm{age}}$) of the predictor effect line and can be calculated irrespective of the values of wealth index.
@@
 This approach incorporates all the uncertainties -- including the uncertainty due to non-focal predictors.
-But what if we are only interested in the uncertainty as a result of the focal predictor, so that the confidence intervals are $\eta \pm q \sigma_f$?
+But what if we are only interested in the uncertainty as a result of the focal predictor, so that the confidence intervals are $\eta \pm q \sigma_f$.
 
-Currently, commonly used \proglang{R} packages for constructing predictions do not exclude the uncertainties resulting from the \emph{non-focal} predictors when computing the CIs.
-A non-trivial way to exclude uncertainties associated with non-focal predictors in some of these packages is to provide a user defined variance-covariance matrix with the covariances of \emph{non-focal} terms set to $0$ -- \emph{zeroing-out} variance-covariance matrix.
+Currently, commonly used \proglang{R} packages for constructing predictions do not exclude the uncertainties resulting from the non-focal predictors when computing the CIs.
+A non-trivial way to exclude uncertainties associated with non-focal predictors in some of these packages is to provide a user defined variance-covariance matrix with the covariances of non-focal terms set to $0$ -- \emph{zeroing-out} variance-covariance matrix.
 This only works when the input predictors are \emph{centered} prior to model fitting, in case of numerical predictors, and even much complicated when the predictors are categorical.
@@
 In this section, we illustrate predictor effects (together with the associated CI) in the context of linear models involving continuous and categorical predictors with or without interactions.
-We compare the expected to the known simulation "truth".
+We compare the expected to the known simulation ``truth".
 
@@
 %
-Secondly, we add interaction terms between the non-focal predictors (\code{wealth index} and \code{household expenditure}) but not with the focal predictor \code{age}:
+Secondly, we add interaction between the non-focal predictors (\code{wealth index} and \code{household expenditure}) but not with the focal predictor \code{age}:
 %
@@
 %
-And lastly, we now consider model with interaction terms between the focal predictor (\code{age}) and one of the non-focal predictor (\code{wealth index}):
+And lastly, the model with interaction terms between the focal predictor (\code{age}) and one of the non-focal predictor (\code{wealth index}):
 %
@@
 We first fitted all the three models and in each case, compared three different predictor effects together with the corresponding CIs -- 1) uncertainty due to non-focal predictors included (everything), 2) uncertainty due to non-focal predictors removed using variance-covariance matrix (zero-vcov), and 3) uncertainty due to non-focal predictors removed using centered model matrix (centered mm).
-The results are shown in Figure ~\ref{fig:pred_age_cont_plots}.
+The results are shown in Figure~\ref{fig:pred_age_cont_plots}.
 
-Figure ~\ref{fig:pred_age_cont_plots}a represents predictor effects for the model with no interaction, Figure ~\ref{fig:pred_age_cont_plots}b shows predictor effects for model with non-focal predictors interacting with each other, and Figure ~\ref{fig:pred_age_cont_plots}c shows predictor effects for the model with focal and non-focal predictors interacting.
+Figure~\ref{fig:pred_age_cont_plots}a represents predictor effects for the model with no interaction, Figure~\ref{fig:pred_age_cont_plots}b shows predictor effects for the model with interaction between non-focal predictors, and Figure~\ref{fig:pred_age_cont_plots}c shows predictor effects for the model with interaction between focal and non-focal predictors.
 For perfect predictions, we expect the yellow and grey dotted lines to overlay each other and they should intersect with the black solid line and the vertical dotted grey line at the same point (or very close).
 In other words, at the model center (mean of the focal predictor), generally, we expect the model to predict the average of the response variable and narrower CIs.
-If we properly anchor the model at its center (or any other appropriate value, assuming no interactions), we would expect the variance at the anchor to be zero, i.e., the CIs bands crosses at that point (see zero-cov and centered mm in Figure ~~\ref{fig:pred_age_cont_plots}).
+If we properly anchor the model at its center (or any other appropriate value, assuming no interactions), we would expect the variance at the anchor to be zero, i.e., the CIs bands crosses at that point (see zero-cov and centered mm in Figure~~\ref{fig:pred_age_cont_plots}).
 Although the overall predictor effects trend lines are different in each of the simulation, the expected household size for a household head with average age is not only the same in all the three simulations but also very close to the ``truth''.
@@
 
-Suppose, in Section ~\ref{sec:continuous_predictors}, in addition to the predictors already defined, we add gender and level of education of the household head.
+Suppose, in Section~\ref{sec:continuous_predictors}, in addition to the predictors already defined, we add gender of the household head.
 For example, we may want to know whether men (M) and females (F) differ in the relationship between age of the household head and the size of the household -- how does households with male as their heads differ in size as compare to those with female heads?
 
-The first model in this case, slightly modifies the first model in Section ~\ref{sec:continuous_predictors} by adding gender of the household head.
+The first model in this case, slightly modifies the first model in Section~\ref{sec:continuous_predictors} by adding gender of the household head.
 The proportion of households with males as their household heads is $60\%$:
@@
 The second model has interaction between non-focal predictors (\code{age} and \code{wealth index}), while the third model has interaction between focal (\code{gender} and \code{age}).
-The results are shown in Figure ~\ref{fig:pred_gender_cat_plots}, which compares the predictors effects (either centered or all uncertainty included -- everything) for no interaction model (a), interaction between non-focal predictors (b) and interaction between focal and non-focal predictor (c).
-\textbf{For each gender, we expect the observed average household size ("truth" -- red dots), to be very close to the estimated predictor effects (blue and black dots and lines)??}
+The results are shown in Figure~\ref{fig:pred_gender_cat_plots}, which compares the predictors effects (either centered or all uncertainty included -- everything) for no interaction model (a), interaction between non-focal predictors (b) and interaction between focal and non-focal predictor (c).
+For each gender, we expect the observed average household size (``truth" -- red dots), to be very close to the estimated predictor effects (blue and black dots).
 %
@@
 The black (non-focal uncertainty excluded) and blue solid dots (all uncertainty included) are the predictor effects with their corresponding CIs, respectively.
-\textbf{For properly generated predictor effects, we expect the black, blue and red dots to be as close as possible???}.}
+As espected, the observed average household size (red) is very similar to the predicted (red and blue).}
 \label{fig:pred_gender_cat_plots}
@@
 
-We now consider linear models with additional random effects components. In the simplest case, we start with a single grouping factor as the random effect components.
+We now consider linear models with additional random effects components.
+In the simplest case, we start with a single grouping factor as the random effect component, i.e., random intercept model.
+This way, we can investigate the contribution of random effects overall expected mean.
 
-Suppose, in the simulation described in Section ~\ref{sec:stats_background}, the observations are  recorded at least once per household.
+Suppose, in the simulation described in Section~\ref{sec:stats_background}, the observations are  recorded at least once per household.
 In particular, let $\mathrm{H}$ be the number of households indexed by the grouping factor, and $\mathrm{h}[i]$ be the household of the $i$th observation.
-We simulate three different models incorporating this kind of formulation -- first, one grouping factor-one predictor-linear random effect:
+We simulate three different models incorporating this kind of formulation -- first, one grouping factor-one predictor-linear random effect model:
 %
@@
 \mathrm{hh~size} &= \beta_0 + \alpha_{0,\mathrm{h}[i]} + \beta_{\mathrm{A}}\mathrm{Age}_i + \epsilon_i \nonumber\\
-\alpha_{0,\mathrm{h}} &\sim \mathrm{Normal}(0, 1), ~~ h=1, \cdots, \mathrm{H} \nonumber\\
+\alpha_{0,\mathrm{h}} &\sim \mathrm{Normal}(0, 1),~~ h=1, \cdots, \mathrm{10} \nonumber\\
 \mathrm{Age}_i &\sim \mathrm{Normal}(0.2, 1) \nonumber\\
@@
 \beta_{\mathrm{A}} &= 0.1 \nonumber\\
-i &= 1,\cdots, 100
+i &= 1,\cdots, 50
 \end{align}
 %
-The second and third models add the random effects, $\alpha_{\mathrm{h}[i]}$, to models ~\ref{sim:lm_nf_interaction} and ~\ref{sim:lm_f_interaction}, respectively.
+The second and third models add the random effects, $\alpha_{\mathrm{h}[i]}$, to models~\ref{sim:lm_nf_interaction} and~\ref{sim:lm_f_interaction}, respectively.
+Figre~\ref{fig:pred_age_cont_lme_plots} shows and compares predictor efffects for the three models.
 %
@@
 \begin{align}\label{sim:lme_multi_outcome}
-\mathrm{y}_{\mathrm{h}[i]\mathrm{s}} &= \beta_{0,s} + \alpha_{0,\mathrm{h}[i]\mathrm{s}} + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \beta_{\mathrm{L}}\mathrm{Latent}_i + \epsilon_{i\mathrm{s}} \nonumber\\
-\alpha_{0,\mathrm{h}\mathrm{s}} &\sim \mathrm{Normal}(0, 2), ~~ h=1, \cdots, \mathrm{H} \nonumber\\
+\mathrm{y}_{\mathrm{h}[i]\mathrm{s}} &= \beta_{0,s} + \alpha_{0,\mathrm{h}[i]\mathrm{s}} + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \beta_{\mathrm{Ls}}\mathrm{Latent}_{i\mathrm{s}} + \epsilon_{i\mathrm{s}} \nonumber\\
+\alpha_{0,\mathrm{h}\mathrm{s}} &\sim \mathrm{Normal}(0, 2),~~ h=1, \cdots, \mathrm{H} \nonumber\\
 \mathrm{Age}_i &\sim \mathrm{Normal}(0.2, 1) \nonumber\\
 \mathrm{Wealthindex}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
-\mathrm{Latent}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
+\mathrm{Latent}_{i\mathrm{s}} &\sim \mathrm{Normal}(0, 1) \nonumber\\
 \epsilon_{i\mathrm{s}} &\sim \mathrm{Normal}(0, 1) \nonumber\\
@@
 \beta_{\mathrm{W}} &= 2 \nonumber\\
-\beta_{\mathrm{L}} &= 1 \nonumber\\
+\beta_{\mathrm{L}1} &= 1 \nonumber\\
+\beta_{\mathrm{L}2} &= -5 \nonumber\\
 i &= 1,\cdots, 100
@@
 
+Predictor effects for the above model are show in Figure~\ref{fig:pred_cont_joint_plots}.
+The top plot compares centered and model centered predictor effects with the observed average household size and rental price.
+The point estimates (black and blue dots) are the expected baseline household size and the rental price.
+The two bottom plots compares the predictor effect (of age) on the predicted household size and rental price.
+Again, in the case of joint model, at the model center, the expected predictions are very similar to the observed marginal averages for the outcomes.
 %
@@
 \end{center}
-\caption{A comparison of predictor effects and their corresponding CIs.
-Figure a) shows model with no interaction, Figure b) shows model with non-focal predictors interacting with each other, and Figure c) shows model with focal and non-focal predictors interacting.
+\caption{Predictor effect of age on the predicted household size and rental price.
+The top plots shows the baseline (mean) predicted household size and rental price.
+The bottom plots compares the predictor effect of age on the predicted household size and rental price.
 The dotted yellow and grey horizontal lines are the expected and observed/true means, respectively.
@@
 The solid black, red (non-focal uncertainty excluded) and blue lines (all uncertainty included) are the predictor effects while their corresponding CIs are the dotted black, red and blue lines, respectively.
-For properly generated predictor effects, we expect the yellow and the grey horizontal dotted lines to overlay each other and intersect with the black, blue, red solid trend lines, and the vertical dotted grey line at the same point (model center).}
+For properly generated predictor effects, we expect the yellow and the grey horizontal dotted lines to overlay each other and intersect with the black, blue, red solid trend lines, and the vertical dotted grey line at the same point (model center).
+Without removing the variations due to non-focal predictors, the confidence bands are very wide ("everything" in the figure).
+However, when CIs are centered, we are only focussing on the uncertainty due to the focal predictor.}
 \label{fig:pred_cont_joint_plots}

### 2021-10-07 compare_emmeans_varpred.rmd
Comparison plots

@@
 ---
-title: "Bias correction in GLMs"
+title: "emmeans and varpred comparison"
 author: Bicko, Jonathan & Ben
@@
 
-```{r cont_no_interaction}
+```{r cont_no_interaction, fig.cap="Continuous: no interaction"}
 quants <- seq(0, 1, length.out=100)
 focal_levels <- quantile(sim_df_cni$age, quants)
-truth = c(truth=mean(sim_df_cni$hhsize))
-print(truth)
+print("Truth:")
+print(true_prop_cni)
 compare_cni_plots <- (combinepreds(mod_cni, focal="age"
@@
 
-```{r cont_nwi_nf_interaction}
-quants <- seq(0, 1, length.out=100)
+```{r cont_wi_nf_interaction, fig.cap="Continuous: non-focal predictors interaction"}
 focal_levels <- quantile(sim_df_wi_nf$age, quants)
-truth = c(truth=mean(sim_df_wi_nf$hhsize))
-print(truth)
+print("Truth:")
+print(true_prop_wi_nf)
 compare_wi_nf_plots <- (combinepreds(mod_wi_nf, focal="age"
@@
 ```
+
+\subsection{Interaction between focal and non-focal predictors}
+
+```{r cont_wi_f_interaction, fig.cap="Continuous: interaction betweem focal and non-focal predictors"}
+focal_levels <- quantile(sim_df_wi_f$age, quants)
+print("Truth:")
+print(true_prop_wi_f)
+compare_wi_f_plots <- (combinepreds(mod_wi_f, focal="age"
+		, funs=c("emmeans", "varpred")
+		, at=list(age=focal_levels)
+		, x.var="age"
+		, nesting="wealthindex %in% age"
+		, plotit=TRUE
+	)
+	+ geom_vline(aes(xintercept=mean(focal_levels)), lty=2, colour="grey")
+)
+print(compare_wi_f_plots)
+```
+
+\section{Categorical predictors}
+
+\subsection{No interaction}
+
+```{r catni, fig.cap="Categorical: no interaction"}
+compare_catni_plots <- (combinepreds(mod_catni, focal="gender"
+		, funs=c("emmeans", "varpred")
+		, x.var="gender"
+		, x.var.factor=TRUE
+		, plotit=TRUE
+	)
+	+ geom_point(data=focal_prop_catni, aes(x=gender, y=hhsize), colour="black")
+)
+print(compare_catni_plots)
+```

### 2021-10-07 categorical_predictors.R

@@ -139,7 +139,7 @@ focal_prop_catni$.varpred <- "a) No interaction"
 focal_prop_catwi_nf$.varpred <- "b) Non-focal interaction"
 focal_prop_catwi_f$.varpred <- "c) Focal-non-focal interaction"
 
-true_prop_cont_df <- do.call("rbind"
+true_prop_cat_df <- do.call("rbind"
 	, list(focal_prop_catni, focal_prop_catwi_nf, focal_prop_catwi_f)
 )
 lnames <- c("a) No interaction"
@@ -160,7 +160,7 @@ pred_gender_cat_plots <- (comparevarpred(vlist=vlist, lnames=lnames
 		, plotit=TRUE, addmarginals=FALSE
 		, facet_ncol=2
 	)
-	+ geom_point(data=true_prop_cont_df, aes(x=gender, y=hhsize, colour="truth"))
+	+ geom_point(data=true_prop_cat_df, aes(x=gender, y=hhsize, colour="truth"))
 	+ scale_colour_manual(breaks = c("truth", "everything", "centered mm")
 		, values=c(truth="red", everything="blue", "centered mm"="black")
 	)

### 2021-10-08 compare_emmeans_varpred.rmd
why the difference: emmeans vs varpred

@@
 ```
+
+\section{Why the differences?}
+
+We actually don't know which method is correct. Let us consider a simple simulation:
+
+- Outcome: hhsize
+- Predictors:
+	- Gender (40\% Males)
+	- Age: Females are slightly older
+
+```{r sim_compare, echo=TRUE}
+## Simulation
+N <- 100
+extraAge <- 0.4
+meanAge <- 0.5
+sdAge <- 1
+prop <- c(0.6, 0.4)
+betas <- c(1.5, 0.5, 0.5)
+gender <- sample(c("Female", "Male"), N, replace = TRUE, prob = prop)
+df <- (data.frame(gender=gender)
+	%>% mutate(muAge=ifelse(gender=="Female", meanAge+extraAge, meanAge)
+		, age=rnorm(N, muAge, sdAge)
+	)
+	%>% select(-muAge)
+)
+
+mm <- model.matrix(~gender+age, df)
+df$hhsize <- rnorm(N, mean=as.vector(mm %*% betas), sd=1)
+
+## Model
+mod <- lm(hhsize~gender+age, data=df)
+print(mod)
+```
+
+Let us take a look at the observed marginals:
+
+```{r observed_margins, echo=TRUE}
+observed_margins <- (df
+	%>% group_by(gender)
+	%>% summarise_all(mean)
+)
+observed_margins
+```
+
+- **varpred** constructs model matrix by averaging _age_ within the levels of _gender_. The population average is the weighted (by the observed proportions) average of these averages:
+
+```{r varpred_mm, echo=TRUE}
+varpred_pred <- varpred(mod, "gender", within.category=TRUE, returnall=TRUE)
+varpred_mm <- varpred_pred$raw$model.matrix
+print(varpred_mm)
+```
+
+Estimates:
+```{r varpred_pred}
+print(varpred_pred)
+```
+
+By hand calculation:
+
+```{r varpred_hand, echo=TRUE}
+varpred_mm %*% coef(mod)
+```
+
+- **emmeans** uses the population average and seems not apply the weights:
+
+```{r emmeans_mm, echo=TRUE}
+emmeans_grid <- ref_grid(mod)
+emmeans_mm <- emmeans_grid@grid
+print(emmeans_mm)
+```
+
+Estimates:
+```{r emmeans_pred}
+emmeans_pred <- emmeans(emmeans_grid, spec=~gender)
+print(emmeans_pred)
+```
+
+By hand calculation:
+
+```{r emmeans_hand, echo=TRUE}
+emmeans_mm <- emmeans_pred@linfct
+print(emmeans_mm)
+
+as.matrix(emmeans_mm) %*% coef(mod)
+```
+
+Mathematically, in **varpred**
+
+\begin{align}
+\mathbb{E}(Y|X) =
+\begin{cases}
+(\beta_0 + \alpha) + \beta_1 \bar{{Age_{M}}},& \text{ if } gender = Male\\\\
+\beta_0 + \beta_1 \bar{{Age_{F}}},&  \text{ if } gender = Female\
+\end{cases}
+\end{align}
+
+**emmeans**
+
+\begin{align}
+\mathbb{E}(Y|X) =
+\begin{cases}
+(\beta_0 + \alpha) + \beta_1 \bar{{Age}},& \text{ if } gender = Male\\\\
+\beta_0 + \beta_1 \bar{{Age}},&  \text{ if } gender = Female\
+\end{cases}
+\end{align}

### 2021-10-11 compare_emmeans_varpred.rmd
Package documentation

@@
 
-vplots <- vareffects:::plot.vareffects
+within.category <- TRUE
 
@@
 		, x.var="gender"
+		, within.category=within.category
 		, x.var.factor=TRUE

### 2021-10-11 R/utilities.R

@@ -163,7 +163,7 @@ matrix.to.df <- function(matrix, colclasses){
 
 clean_model <- function(focal.predictors, mod, xlevels
 	, default.levels, formula.rhs, steps, x.var, typical
-	, vnames, which.interaction, within.category, pop.ave){
+	, vnames, within.category, bias.adjust){
   if ((!is.null(mod$nan.action)) && inherits(mod$na.action, "exclude"))
     class(mod$na.action) <- "omit"
 
@@ -207,9 +207,7 @@ clean_model <- function(focal.predictors, mod, xlevels
       factor.cols[names(grep(name, vnames, value=TRUE))] <- TRUE
     }
   }
-  if (which.interaction=="effects") {
-     factor.cols[grep(":", cnames)] <- FALSE   
-  }
+  factor.cols[grep(":", cnames)] <- FALSE   
 
   X <- recoverdata(mod=mod, extras=all.predictors)
   
@@ -258,9 +256,9 @@ clean_model <- function(focal.predictors, mod, xlevels
 	 if (!fac) {
 		levels <- if (is.null(xlevels[[name]])){
 			 quant <- seq(0, 1, length.out=steps)
-			 if (pop.ave=="quantile") {
+			 if (bias.adjust=="quantile") {
 			 	as.vector(quantile(X[,name], quant))
-			 } else if (pop.ave=="population") {
+			 } else if (bias.adjust=="population") {
 			 	as.vector(X[, name])
 			 } else {
 			 #	seq(min(X[, name]), max(X[, name]), length.out=steps)
@@ -274,7 +272,7 @@ clean_model <- function(focal.predictors, mod, xlevels
 		}
 	 } else {
 		factor.levels[[name]] <- levels
-	 	if (pop.ave=="population" || (within.category & name==x.var)) {
+	 	if (bias.adjust=="population" || (within.category & name==x.var)) {
 			levels <- as.vector(X[, name])
 		}
 	 }
@@ -293,18 +291,18 @@ clean_model <- function(focal.predictors, mod, xlevels
 	 		levels[1]
 		}
 	 } else {
-	 	if (pop.ave=="quantile") {
+	 	if (bias.adjust=="quantile") {
 			quant <- seq(0, 1, length.out=steps)
 			as.vector(quantile(X[,name], quant))	
-		} else if (pop.ave=="population" || (within.category & !is.factor(X[, name]))) {
[…]

### 2021-10-14 variable_predictions.Rnw

@@
 load("multiple_outcomes_preds.rda")
+load("cubic_predictors_preds.rda")
 @
@@
 
+\subsection{Polynomial interaction}
+
+We consider more complex interactions in which the focal predictor is modelled as a cubic polynomial. Specifically:
+%
+\begin{align}\label{sim:lm_cubic}
+\mathrm{hh~size} &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}^3_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \epsilon_i \nonumber\\
+\mathrm{Age}_i &\sim \mathrm{Uniform}(-1, 1) \nonumber\\
+\mathrm{Wealthindex}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
+\epsilon_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
+\beta_0 &= 1.5 \nonumber\\
+\beta_{\mathrm{A}} &= 0.3 \nonumber\\
+\beta_{\mathrm{W}} &= 2 \nonumber\\
+i &= 1,\cdots, 100
+\end{align}
+%
+
+%
+\begin{figure}[H]
+\begin{center}
+<<pred_cubic_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
+print(pred_cubic_plots)
+@
+\end{center}
+\caption{A comparison of predictor effects and their corresponding CIs.
+Figure a) shows model with no interaction, Figure b) shows model with non-focal predictors interacting with each other, and Figure c) shows model with focal and non-focal predictors interacting.
+The dotted yellow and grey horizontal lines are the expected and observed/true means, respectively.
+The vertical grey line is the mean of the focal predictor (model center).
+The solid black, red (non-focal uncertainty excluded) and blue lines (all uncertainty included) are the predictor effects while their corresponding CIs are the dotted black, red and blue lines, respectively.
+For properly generated predictor effects, we expect the yellow and the grey horizontal dotted lines to overlay each other and intersect with the black, blue, red solid trend lines, and the vertical dotted grey line at the same point (model center).}
+\label{fig:pred_cubic_plots}
+\end{figure}
+%
+
 \subsubsection{Categorical predictors}\label{sec:categorical_predictors}

### 2021-10-20 variable_predictions.Rnw
Autocommit
Temp for focal_pop

@@
 \item  propagation of uncertainty ([how] can we incorporate uncertainty in nonlinear components of (G)LMMs? Should we exclude variation in non-focal parameters?)
+\item bias in the expected mean prediction induced by the nonlinear transformation of the response variable (especially in GL(M)M)
 \end{enumerate}
@@
 In this article, we refer to this quantity as \emph{predictor effect} since it should, for example, tell us what we would expect the presponse to be at a particular value or level of the predictor.
+Formerly, predictor effects computes the expected outcome by meaningfully holding the non-focal predictors constant (or averaged in some meaningful way) while varying the focal predictor, with the goal that the outcome expected prediction represents how the model responds to the changes in the focal predictor.
+
 An alternative to averaging the non-focal predictors, is the population-based approach which involves computing the prediction over the population of the non-focal predictors and then averaging across the values of the focal predictor.
@@
 \begin{align}\label{sim:lm_cubic}
-\mathrm{hh~size} &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}^3_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \epsilon_i \nonumber\\
+\mathrm{hh~size} &= \beta_0 + \beta_{\mathrm{A_1}}\mathrm{Age}_i + \beta_{\mathrm{A_2}}\mathrm{Age}^2_i + \beta_{\mathrm{A_3}}\mathrm{Age}^3_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \epsilon_i \nonumber\\
 \mathrm{Age}_i &\sim \mathrm{Uniform}(-1, 1) \nonumber\\

### 2021-10-21 R/utilities.R
Population-based bias

@@ -1,77 +1,21 @@
 # Internal functions used to condition the model. Some of the condition code is based on R package effects (clean_model). Transfered and modified here because they are not exported in effects.
 
 
-focal_pop <- function(mf, betahat, formula.rhs, rTerms, contr
-	, x, x.var, factor.levels, steps, stats, ...) {
+focal_pop <- function(x.focal, x.excluded, betahat, formula.rhs
+	, rTerms, factor.levels, contr, x.var, steps, stats, off, ...) {
 	
-	.focal <- names(x)
-
-	quant <- seq(0, 1, length.out=steps)
-	for (f in 1:length(.focal)) {
-		.x.focal <- .focal[f]
-		if (!(x[[.x.focal]]$is.factor)) {
-			x[[.x.focal]]$levels <- quantile(x[[.x.focal]]$levels, quant, names=FALSE)
-		}
+	pred_list <- list()
+	for (i in 1:NROW(x.focal)) {
+		focal_i <- x.focal[i, ,drop=FALSE]
+		focal_i[1:NROW(x.excluded), ] <- focal_i
+		mf_i <- cbind(focal_i, x.excluded)
+		mf_i <- model.frame(rTerms, mf_i, xlev=factor.levels, na.action=NULL)
+		mm_i <- model.matrix(formula.rhs, data = mf_i, contrasts.arg = contr)
+		focal_i$fit <- off + as.vector(mm_i %*% betahat)
+		pred_list[[i]] <- focal_i
 	}
-	
-	if (length(.focal) > 1) {
-		for (f in 1:length(.focal)) {
-			.x.focal <- .focal[f]
-			if (x[[.x.focal]]$is.factor) {
-				.new_xf <- x[[.x.focal]]$levels
-			} else {
-				.new_xf <- x[[.x.focal]]$levels
-				.new_xf <- quantile(.new_xf, quant, names=FALSE)
-			}
-
-			pred_focal <- list()
-			for (i in 1:length(.new_x)) {
-				mf_focal[[x.var]] <- .new_x[i]
-				pred_j <- list()
-				for (j in 1:length(.new_xf)) {
-					mf_focal[[.x.focal]] <- .new_xf[j]
-					mf_focal <- model.frame(rTerms, mf_focal, xlev=factor.levels)
-					mm <- model.matrix(formula.rhs, data=mf_focal, contrasts.arg=contr)
-					pred <- as.vector(mm %*% betahat)
[…]

### 2021-10-21 multiple_outcomes_preds.R

@@ -9,12 +9,9 @@ makeGraphics()
 
 ## Services
 
-within.category <- FALSE
-
 ### All uncertainties 
 pred_services_trad_cont_joint <- varpred(mod_cont_joint
 	, "services"
-	, within.category=within.category
 	, modelname="everything"
 )
 
@@ -22,7 +19,6 @@ pred_services_trad_cont_joint <- varpred(mod_cont_joint
 pred_services_centered_cont_joint <- varpred(mod_cont_joint
 	, "services"
 	, isolate=TRUE
-	, within.category=within.category
 	, zero_out_interaction=FALSE
 	, modelname="centered mm"
 )
@@ -49,7 +45,6 @@ pred_services_cont_plots <- (comparevarpred(vlist=vlist
 pred_age_trad_cont_joint <- varpred(mod_cont_joint
 	, c("services", "age")
 	, x.var="age"
-	, within.category=within.category
 	, modelname="everything"
 )
 
@@ -58,7 +53,6 @@ pred_age_centered_cont_joint <- varpred(mod_cont_joint
 	, c("services", "age")
 	, x.var="age"
 	, isolate=TRUE
-	, within.category=within.category
 	, zero_out_interaction=FALSE
 	, modelname="centered mm"
 )

### 2021-10-21 variable_predictions_funs.R

@@ -267,7 +267,7 @@ combinepreds <- function(mod, funs, focal, x.var, x.var.factor=FALSE, plotit=TRU
 		if (inherits(est, c("emmeans", "emmGrid"))) {
 			type <- est@misc$predict.type
 			est <- as.data.frame(est)
-			if (type=="response") {
+			if (length(type) && type=="response") {
 				.nn <- "prob"
 			} else {
 				.nn <- "emmean"

### 2021-10-28 compare_emmeans_varpred.rmd
Categorical interactions

@@
 library(emmeans)
+library(effects)
 
@@
 \end{align}
+
+\section{Dealing with categorical iterations}
+
+```{r categorical_toy, echo=TRUE}
+df <- (expand.grid(services=c("water", "garbage", "toilet")
+		, gender=c("Male", "Female")
+		, statusP=c("Base year", "Not observed", "Unimproved", "Improved")
+	)
+	%>% mutate(age = rnorm(n())
+		, status=rbinom(n(), 1, plogis(rnorm(n(), 3, 5)))
+	)
+)
+head(df)
+```
+
+Model:
+
+```{r toy_model, echo=TRUE}
+toy_mod <- glm(status~-1+services+(age+gender+statusP):services
+	, data=df
+	, family="binomial"
+	, contrasts=list(services="contr.sum")
+)
+```
+
+\subsection{varpred}
+
+- Averaging all the columns of model matrix?
+
+```{r varpred_all_av}
+varpred_all_av <- varpred(toy_mod, "services", handle.inter="emmeans", returnall=TRUE)
+print(varpred_all_av$raw$model.matrix)
+```
+
+- On baseline level?
+
+```{r varpred_no_av}
+varpred_no_av <- varpred(toy_mod, "services", handle.inter="effects", returnall=TRUE)
+print(varpred_no_av$raw$model.matrix)
+```
+
+\subsection{emmeans}
+
+```{r emmeans_toy}
+emmeans_toy <- emmeans(toy_mod, specs=~services, nesting=NULL, weighting="equal")
+print(emmeans_toy@linfct)
+```
+
+\subsection{Effects}
+
+```{r effects_toy}
+effects_toy <- Effect("services", toy_mod)
+print(effects_toy$model.matrix)
+```

### 2021-10-28 R/methodfuns.R

@@ -69,3 +69,12 @@ includeRE <- function(mod, ...)UseMethod("includeRE")
 #'
 #' @keywords internal
 check_intercept <- function(mod, ...)UseMethod("check_intercept")
+
+#'
+#' @keywords internal
+get_contrasts <- function(mod, ...)UseMethod("get_contrasts")
+
+
+#'
+#' @keywords internal
+get_model.mm <- function(mod, ...)UseMethod("get_model.mm")

### 2021-11-02 compare_emmeans_varpred.rmd
Autocommit

@@
 ```{r toy_model, echo=TRUE}
-toy_mod <- glm(status~-1+services+(age+gender+statusP):services
+toy_mod <- glm(status~services*age
 	, data=df
 	, family="binomial"
-	, contrasts=list(services="contr.sum")
+#	, contrasts=list(services="contr.sum")
 )

### 2021-11-02 short_lab_report.rmd
Autocommit

--- 2021-10-07 variable_predictions.Rnw
@@
-\documentclass[12pt]{article}
-\usepackage{amsmath}
-\usepackage{natbib}
-\usepackage[dvipsnames]{xcolor}
-\usepackage{graphicx, subfig}
-\usepackage{float}
-\usepackage{hyperref}
+---
+title: "Predictor effects and Bias correction in GL(M)M"
+author: Bicko, Ben & Jonathan
+date: "2021 Oct 29 (Fri)"
+compact-title: false
+output:
+  pdf_document:
+    extra_dependencies: ["float"]
+bibliography: bias_correction.bib
+link-citations: yes
+---
 
-\def\code#1{\texttt{#1}}
-\let\proglang=\textsf
-
-\newcommand{\JD}[1]{{\color{blue} \emph{#1}}}
-\newcommand{\bmb}[1]{{\color{RubineRed!70!} \emph{#1}}}
-\newcommand{\pkg}[1]{\textbf{#1}}
-
-\newcommand{\bX}{{\mathbf X}}
-\newcommand{\bbeta}{{\boldsymbol \beta}}
-\newcommand{\boldeta}{{\boldsymbol \eta}}
-\newcommand{\boldmu}{{\boldsymbol \mu}}
-
-\newcommand{\nset}[1]{#1_{\{n\}}}
-\newcommand{\yref}{y_{\textrm{ref}}}
-\newcommand{\cdist}{{D(\nset{x}|x_f)}}
-\newcommand{\cdistprime}{{D(\nset{x}|x_{f'})}}
-\newcommand{\xfprime}{x_{f'}}
-
-\setlength{\parindent}{0pt}
-
-\title{Uncertainty propagation for variable (predictor) effects and bias correction in generalized linear (mixed) models???}
-
-\begin{document}
-
-\maketitle
-%\setkeys{Gin}{width=0.8\textwidth}
-
-\section{Introduction}
-
-<<setup, echo=FALSE, message=FALSE, warning=FALSE>>=
+```{r, echo=FALSE}
 library(ggplot2)
-library(ggpubr)
-
-library(stargazer)
 
@@
 load("variable_predictions_objs.rda")
-load("categorical_predictors.rda")
-load("lme_random_intercept.rda")
-load("multiple_outcomes_preds.rda")
-@
+load("binom_correlated_preds.rda")
+```
 
-In many applications, predictions of expected responses, or response probabilities, are often of interest.
-Predictions provide a great way to summarize what the regression model is telling us and are very useful for interpreting and visualizing models estimates using graphs.
-For example, in logistic regression models, the coefficient estimates are usually not easy to interpret, and we may want to explore the ``effects" of covariates on predicted probabilities, or predictions of responses at various levels of covariates -- \emph{predictor effects}.
-For example, as a way to plan, public health officials may want to know the ``effects" of household income, wealth index, etc, on the predicted probability of having improved water services among slum dwellers.
+We want:
 
-Both simple and generalized linear (mixed) models (GL(M)Ms) can examine very complex relationships, including nonlinear relationships between response and predictors, interactions between predictors (and via splines for example), and nonlinear transformations via link functions due to their flexibility.
-This flexibility comes at a cost, for example, complex multivariate models may risk misinterpretation, and miscalculation of quantities of interest.
-Also, coefficient estimates of models involving nonlinear link functions or interactions lose their direct interpretation \citep{leeper2017interpreting}, meaning that interpretation of derived quantities from these estimates requires some understanding of the specified model.
+\[
+E(y|\{X_f, \bar{X}_{\{n\}}\})
+\]
 
-When visually presented, predictor effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind of transformations on the predictors whose estimates are usually, but not always, a subject to less clarity of interpretation.
-Further, generating predictor effects together with the associated confidence intervals for regression models has a number of challenges.
-In particular:
+Challenges:
+
 \begin{enumerate}
@@
 \item  propagation of uncertainty ([how] can we incorporate uncertainty in nonlinear components of (G)LMMs? Should we exclude variation in non-focal parameters?)
+\item bias in the expected mean prediction induced by the nonlinear transformation of the response variable (especially in GL(M)M)
 \end{enumerate}
 
-Most common way of dealing with the first challenge is taking unique levels of the focal predictor if discrete or taking appropriately sized quantiles (or bins) if continuous, and then calculating the predictions while holding non-focal predictors at their typical values (e.g., averages).
-This generates -- \emph{predictor effect} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} and \emph{estimated marginal means} \citep{lenth2018package}.
-In this article, we refer to this quantity as \emph{predictor effect} since it should, for example, tell us what we would expect the presponse to be at a particular value or level of the predictor.
-An alternative to averaging the non-focal predictors, is the population-based approach which involves computing the prediction over the population of the non-focal predictors and then averaging across the values of the focal predictor.
-Dealing with second challenge is discussed in Section~\ref{sec:uncertainity_propagation}
+```{r echo=FALSE, fig.height=10}
+print(pred_age_cont_plots)
+```
 
-When dealing with nonlinear link functions, the correct predictions for example, are even much harder to estimate.
-One approach is to make predictions on the transformed scale (linear predictor scale), and then back-transform to the original scale.
-However, the back-transformation may either result in biased predictions or requires some approximation.
-In particular, bias in expected mean prediction induced by nonlinear transformation of the response variable can lead inaccurate predictions.
-
-The main purpose of this article is to discuss and implement various approaches for computing predictor effects and provide an alternative method for computing the associated confidence intervals.
-We further explore approaches for correcting bias in predictions for GL(M)Ms involving nonlinear link functions.
-
-The outline of this article is as follows: Section \ref{sec:qoi} provides formal definition various quantities of interest, Section \ref{sec:stats_background} provides statistical background of estimation of model predictions, Section \ref{sec:uncertainity_propagation} describes mathematical and computation implementation of the proposed method for uncertainty propagation, and lastly Section \ref{sec:bias_correction} describes methods for bias correction in GL(M)M together with the computational implementation.
-
-\section{Quantities of interest}\label{sec:qoi}
-
-Several quantities of interest may be derived from regression models.
-The first one is the coefficient estimates.
-Others are \emph{predictors}, \emph{model matrix}, \emph{predicted values} and \emph{marginal effects}.
-Predictors are model input variables and is associated with one or more \emph{variables} (in polynomials, splines and models with interaction).
-Model matrix refers to the design matrix whose rows include all combination of variables appearing in the interaction terms, along with the ``typical" values of the focal and non-focal predictors.
-
-To illustrate some of the concepts, consider results from, hypothetical  simulated example, regression of household size as function of household wealth index and age of household head, and the second model involving these predictors plus the interaction between them.
-
-<<qoi_coefs, echo=FALSE, results=tex, message=FALSE, warning=FALSE>>=
-stargazer(qoi_mod1, qoi_mod2
-	, title = "Example of simple linear model output"
-	, single.row=TRUE
-	, table.placement="H"
-	, label = "tab:qoi_coefs"
-	, keep.stat="n"
-)
-@
-
-In simple linear models with no interaction terms, the default out for the coefficient estimates in Table~\ref{tab:qoi_coefs} model (1) is simple and directly interpretable as the expected change in household size for a unit change in age.
-In particular, a unit change in age is associated with a household that is \Sexpr{round(coef(qoi_mod1)[[2]], 3)} bigger.
-This is the \emph{unconditional marginal effect} \citep{leeper2017interpreting} and it is constant across all the observations and levels of all other predictors.
-As a result, the interpretation of interaction models differs in an important way from linear-additive regression models.
-To see this more clearly, compare the marginal effect of $x_1$ in the following two models:
-%
-\begin{align}
-y_1 &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \epsilon_1 \label{eq:simple_lm}\\
-y_2 &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_{12}x_1x_2 + \epsilon_2 \label{eq:simple_inter_lm}
-\end{align}
-%
-The marginal effect of $x_1$ in Equation~\ref{eq:simple_lm} is $\frac{\partial y_1}{\partial x_1} = \beta_1$.
-On the other hand, the marginal effect of $x_1$ in Equation~\ref{eq:simple_inter_lm} is given by $\frac{\partial y_2}{\partial x_1} = \beta_1 + \beta_{12}x_2$.
-In other words, if there are no interactions, the marginal effect of $x_1$ on $y_1$ is constant, while, if there are interactions in the model, the marginal effect of a change in $x_1$ on $y_1$ depends on the value of the \emph{conditioning} predictor $x_2$.
-
-Figure~\ref{fig:qoi_pred_plot} shows the comparison of predictor effect and unconditional marginal effect of age on household size, based on model estimates in Table~\ref{tab:qoi_coefs}.
-Since the model has no interactions, the relationship between the predicted household size and age is linear, hence the marginal effect of age is the slope ($\frac{\Delta \mathrm{hh size}}{\Delta \mathrm{age}}$) of the predictor effect line and can be calculated irrespective of the values of wealth index.
-%
-\begin{figure}[H]
-\begin{center}
-<<qoi_pred_plot, echo=FALSE, results=tex, fig=TRUE, message=FALSE>>=
-qoi_age_pred_plot
-@
-\end{center}
-\caption{Variable effect plot and marginal effect of age of household head on household size. For a linear model with no interaction, the marginal effect is the slope of the predictor effect line.}
-\label{fig:qoi_pred_plot}
-\end{figure}
-%
-Predictor effect, on the other hand, is the expected household size for a particular age, holding wealth index at its mean.
-In particular, the purpose and goal of a predictor effect seems fairly straightforward; for specified values of (a) focal predictor(s), we want to give a point estimate and confidence intervals for the prediction of the model for a ``typical" (= random sample over the multivariate distribution of non-focal parameters) individual with those values of the predictors.
-
-On the other hand, when model specification involves other kinds of terms such as variable interactions, log and power transformations, etc., the coefficient estimates cannot easily and directly communicate the relationship between the outcome and the independent variable of interest because of the multiple coefficients of the same variable.
-For instance, in model (2), the interaction between age and wealth index is added, hence the coefficient estimate for age (\Sexpr{round(coef(qoi_mod2)[[2]], 3)}) can only be regarded as unconditional marginal effect when effect of wealth index (and thus \code{age:wealthindex}) is zero, which is never the case.
-In other words, for models involving interactions terms, looking at a single estimated coefficient in isolation and failing to take into account the estimates involving the interaction terms may lead to inaccurate interpretations \citep{brambor2006understanding, leeper2017interpreting}.
-
-\section{Statistical background}\label{sec:stats_background}
-
-To get an intuition of how conditioning on the mean values of the non-focal predictors work, suppose we are interested in predictor effects of a particular predictor (hence forth referred as \emph{focal} predictor otherwise \emph{non-focal}), $x_f$, from the set of predictors.
-To keep it simple, assume that the model has no interaction terms.
-Then the idea is to \emph{anchor} the values of non-focal predictors to some particular values.
-For example fixing the values of \emph{non-focal} predictor(s) at some typical values -- typically determined by averaging (for now) in some meaningful way, for example, arithmetic mean  for continuous and average over the levels of the factors for categorical non-focal predictors.
-The easiest way to achieve this is by constructing $\bX^\star$ by averaging the columns of non-focal predictors in model matrix $\bX$, and together with appropriately chosen values of focal predictor(s).
-
-Consider a simple linear model with linear predictor $\eta = \bX\bbeta$ and let $g(\boldmu) = \boldeta$ be an identity link function (in the case of simple linear model), where $\boldmu$ is the expected value of response variable $y$.
-Let $\hat{\bbeta}$ be the estimate of $\bbeta$, together with the estimated covariance matrix $\Sigma = V(\hat{\bbeta})$ of $\hat{\bbeta}$.
-Let $\mathbf{X^*}$ be the model matrix, inheriting most of its key properties, for example transformations on predictors and interactions from the model matrix, $\mathbf{X}$.
-Then the prediction $\hat{\boldeta}^\star = \bX^\star\hat{\bbeta}$ is the predictor effect for the focal predictor in question.
-
-An alternative, save for later, formulation of predictor effect involves, expressing the linear predictor as the sum of the focal and non-focal predictor linear predictor.
-In particular, $\eta^\star(x_f, \nset{{\bar{x}}}) = \beta_f x_f + \sum \nset{\beta} \nset{{\bar{x}^\star}}$, where $\nset{{\bar{x}^\star}}$ are the appropriately averaged entries of non-focal predictors and $x_f$ is a vector of values of the focal predictors for a particular observation.
-
-Generally, in generalized linear models with other link functions, the predictions are first generated in linear predictor scale, $\hat{\boldeta}^*$, and then transformed to the original response scale using $g^{-1}(\hat{\boldeta}^*)$.
-
-\section{Uncertainty propagation}\label{sec:uncertainity_propagation}
-
-What about the confidence intervals (CI)?
-The limits of the confidence intervals are points, not mean values.
-In principle, every observation/value of focal predictors has a different CI.
-The traditional way to compute variances for predictions is $\sigma^2 = \textrm{Diag}(\bX^\star \Sigma \bX^{\star\top})$ \citep{lenth2018package, fox2009effect}, so that the confidence intervals are $\eta \pm q\sigma$, where $q$ is an appropriate quantile of Normal or t distribution.
-This approach incorporates all the uncertainties -- including the uncertainty due to non-focal predictors.
-But what if we are only interested in the uncertainty as a result of the focal predictor, so that the confidence intervals are $\eta \pm q \sigma_f$.
-
-Currently, commonly used \proglang{R} packages for constructing predictions do not exclude the uncertainties resulting from the non-focal predictors when computing the CIs.
-A non-trivial way to exclude uncertainties associated with non-focal predictors in some of these packages is to provide a user defined variance-covariance matrix with the covariances of non-focal terms set to $0$ -- \emph{zeroing-out} variance-covariance matrix.
-This only works when the input predictors are \emph{centered} prior to model fitting, in case of numerical predictors, and even much complicated when the predictors are categorical.
-We first describe this variance-covariance based approach and then discuss our proposed method which is based on \emph{model centering} and does not require input predictors to be scaled prior to model fitting.
-Model center is found by averaging the variables over the model matrix, $\bX^\star$.
-
-\subsection{Variance-covariance}
-
-The computation of $\hat{\eta}^\star$ remains the same as described above.
-However, to compute $\sigma$, $\Sigma$ is modified by \emph{zeroing-out} (the variance-covariance of all non-focal predictors are assigned zero) variances of non-focal terms.
-Although this is the simplest approach, it requires centering of continuous predictors prior to model fitting and proper way to average categorical predictors.
-One way to center predictors is to create a design matrix, $\bX$, and then compute $\bX_{c} = \bX - \bar{\bX}$.
-
-\subsection{Centered model matrix}
-
-Consider centered model matrix $\bX^{\star}_{c} = \{\bX_f^\star, \nset{{\bX}^\star} - \nset{{\bar{\bX}}^\star}\}$.
-It follows that the non-focal terms in $\bX^{\star}_{c}$ are all zero.
-Consequently the uncertainty due to non-focal predictors are all zeroed-out in the computation of $\sigma$.
-More generally, centered design matrix, $\bX^{\star}_{c}$, impacts on the estimated value of the intercept and its associated variance but not the slopes.
-Thus since non-focal terms in $\bX^{\star}_{c}$ are all zero, it does not matter what their corresponding values are in the variance-covariance matrix.
-Hence, we can compute CIs from non-centered predictors (in other words, fitted models with predictors in their natural scales).
-
-\subsection{Handling higher order terms}
-
-Higher order terms such as interactions, splines, polynomials, etc., can be between the non-focal predictors or focal and non-focal predictor(s).
-In the former case, we treat the interactions as just another column in the variable space (of the model matrix).
-\textbf{Still figuring out the later case but currently treating everything as just additional column???}
-
-\subsection{Simulation examples}
-
-In this section, we illustrate predictor effects (together with the associated CI) in the context of linear models involving continuous and categorical predictors with or without interactions.
-We compare the expected to the known simulation ``truth".
-
-\subsubsection{Continuous predictors}\label{sec:continuous_predictors}
-
-Consider simulation example described in Section \ref{sec:qoi}, with slight modifications to incorporate interaction between predictors. The first case, is the model with no interaction: 
-%
-\begin{align}\label{sim:lm_no_interaction}
-\mathrm{hh~size} &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \epsilon_i \nonumber\\
-\mathrm{Age}_i &\sim \mathrm{Normal}(0.2, 1) \nonumber\\
-\mathrm{Wealthindex}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
-\epsilon_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
-\beta_0 &= 1.5 \nonumber\\
-\beta_{\mathrm{A}} &= 0.1 \nonumber\\
-\beta_{\mathrm{W}} &= 2 \nonumber\\
-i &= 1,\cdots, 100
-\end{align}
-%
-Secondly, we add interaction between the non-focal predictors (\code{wealth index} and \code{household expenditure}) but not with the focal predictor \code{age}:
-%
-\begin{align}\label{sim:lm_nf_interaction}
-\mathrm{hh~size} &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \beta_{\mathrm{E}}\mathrm{Expenditure}_i \nonumber\\ 
-&+ \beta_{\mathrm{WE}}\mathrm{Wealthindex}_i*\mathrm{Expenditure}_i + \epsilon_i \nonumber\\
-\mathrm{Age}_i &\sim \mathrm{Normal}(0.2, 1) \nonumber\\
-\mathrm{Wealthindex}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
-\mathrm{Expenditure}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
-\epsilon_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
-\beta_0 &= 1.5 \nonumber\\
-\beta_{\mathrm{A}} &= 0.1 \nonumber\\
-\beta_{\mathrm{W}} &= 2 \nonumber\\
-\beta_{\mathrm{E}} &= 1.5 \nonumber\\
-\beta_{\mathrm{WE}} &= 1 \nonumber\\
-i &= 1,\cdots, 100
-\end{align}
-%
-And lastly, the model with interaction terms between the focal predictor (\code{age}) and one of the non-focal predictor (\code{wealth index}):
-%
-\begin{align}\label{sim:lm_f_interaction}
-\mathrm{hh~size} &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \beta_{\mathrm{E}}\mathrm{Expenditure}_i \nonumber\\ 
-&+ \beta_{\mathrm{AW}}\mathrm{Age}_i*\mathrm{Wealthindex}_i + \epsilon_i \nonumber\\
-\mathrm{Age}_i &\sim \mathrm{Normal}(0.2, 1) \nonumber\\
-\mathrm{Wealthindex}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
-\mathrm{Expenditure}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
-\epsilon_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
-\beta_0 &= 1.5 \nonumber\\
-\beta_{\mathrm{A}} &= 0.1 \nonumber\\
-\beta_{\mathrm{W}} &= 2 \nonumber\\
-\beta_{\mathrm{E}} &= 1.5 \nonumber\\
-\beta_{\mathrm{AW}} &= 1 \nonumber\\
-i &= 1,\cdots, 100
-\end{align}
-%
-
-We first fitted all the three models and in each case, compared three different predictor effects together with the corresponding CIs -- 1) uncertainty due to non-focal predictors included (everything), 2) uncertainty due to non-focal predictors removed using variance-covariance matrix (zero-vcov), and 3) uncertainty due to non-focal predictors removed using centered model matrix (centered mm).
-The results are shown in Figure~\ref{fig:pred_age_cont_plots}.
-
-Figure~\ref{fig:pred_age_cont_plots}a represents predictor effects for the model with no interaction, Figure~\ref{fig:pred_age_cont_plots}b shows predictor effects for the model with interaction between non-focal predictors, and Figure~\ref{fig:pred_age_cont_plots}c shows predictor effects for the model with interaction between focal and non-focal predictors.
-For perfect predictions, we expect the yellow and grey dotted lines to overlay each other and they should intersect with the black solid line and the vertical dotted grey line at the same point (or very close).
-In other words, at the model center (mean of the focal predictor), generally, we expect the model to predict the average of the response variable and narrower CIs.
-If we properly anchor the model at its center (or any other appropriate value, assuming no interactions), we would expect the variance at the anchor to be zero, i.e., the CIs bands crosses at that point (see zero-cov and centered mm in Figure~~\ref{fig:pred_age_cont_plots}).
-Although the overall predictor effects trend lines are different in each of the simulation, the expected household size for a household head with average age is not only the same in all the three simulations but also very close to the ``truth''.
-%
-\begin{figure}[H]
-\begin{center}
-<<pred_age_cont_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
-print(pred_age_cont_plots)
-@
-\end{center}
-\caption{A comparison of predictor effects and their corresponding CIs.
-Figure a) shows model with no interaction, Figure b) shows model with non-focal predictors interacting with each other, and Figure c) shows model with focal and non-focal predictors interacting.
-The dotted yellow and grey horizontal lines are the expected and observed/true means, respectively.
-The vertical grey line is the mean of the focal predictor (model center).
-The solid black, red (non-focal uncertainty excluded) and blue lines (all uncertainty included) are the predictor effects while their corresponding CIs are the dotted black, red and blue lines, respectively.
-For properly generated predictor effects, we expect the yellow and the grey horizontal dotted lines to overlay each other and intersect with the black, blue, red solid trend lines, and the vertical dotted grey line at the same point (model center).}
-\label{fig:pred_age_cont_plots}
-\end{figure}
-%
-
-\subsubsection{Categorical predictors}\label{sec:categorical_predictors}
-
-Suppose, in Section~\ref{sec:continuous_predictors}, in addition to the predictors already defined, we add gender of the household head.
-For example, we may want to know whether men (M) and females (F) differ in the relationship between age of the household head and the size of the household -- how does households with male as their heads differ in size as compare to those with female heads?
-
-The first model in this case, slightly modifies the first model in Section~\ref{sec:continuous_predictors} by adding gender of the household head.
-The proportion of households with males as their household heads is $60\%$:
-%
-\begin{align}\label{sim:lm_no_interaction_cat}
-\mathrm{hh~size} &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{M}}{\mathrm{Gender}_i} + \epsilon_i \nonumber\\
-\mathrm{Age}_i &\sim \mathrm{Normal}(0.2, 1) \nonumber\\
-\mathrm{Gender}_i &\sim \mathrm{Sample}(\mathrm{Gender_i}, p) \nonumber\\
-p &= \begin{cases}
-   0.4 & \mathrm{Gender_{i=Female}} \nonumber\\
-   0.6 & \mathrm{Gender_{i=Male}} \nonumber\\
-  \end{cases} \nonumber\\
-\epsilon_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
-\beta_0 &= 1.5 \nonumber\\
-\beta_{\mathrm{A}} &= 0.8 \nonumber\\
-\beta_{\mathrm{M}} &= 0.2 \nonumber\\
-i &= 1,\cdots, 100
-\end{align}
-%
-The second model has interaction between non-focal predictors (\code{age} and \code{wealth index}), while the third model has interaction between focal (\code{gender} and \code{age}).
-The results are shown in Figure~\ref{fig:pred_gender_cat_plots}, which compares the predictors effects (either centered or all uncertainty included -- everything) for no interaction model (a), interaction between non-focal predictors (b) and interaction between focal and non-focal predictor (c).
-For each gender, we expect the observed average household size (``truth" -- red dots), to be very close to the estimated predictor effects (blue and black dots).
-%
-\begin{figure}[H]
-\begin{center}
-<<pred_gender_cat_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
-print(pred_gender_cat_plots)
-@
-\end{center}
-\caption{A comparison of predictor effects and their corresponding CIs.
-Figure a) shows model with no interaction, Figure b) shows model with non-focal predictors interacting with each other, and Figure c) shows model with focal and non-focal predictors interacting.
-The red dots are the observed observed/true means.
-The black (non-focal uncertainty excluded) and blue solid dots (all uncertainty included) are the predictor effects with their corresponding CIs, respectively.
-As espected, the observed average household size (red) is very similar to the predicted (red and blue).}
-\label{fig:pred_gender_cat_plots}
-\end{figure}
-%
-
-\subsubsection{Linear mixed effect model}
-
-We now consider linear models with additional random effects components.
-In the simplest case, we start with a single grouping factor as the random effect component, i.e., random intercept model.
-This way, we can investigate the contribution of random effects overall expected mean.
-
-Suppose, in the simulation described in Section~\ref{sec:stats_background}, the observations are  recorded at least once per household.
-In particular, let $\mathrm{H}$ be the number of households indexed by the grouping factor, and $\mathrm{h}[i]$ be the household of the $i$th observation.
-We simulate three different models incorporating this kind of formulation -- first, one grouping factor-one predictor-linear random effect model:
-%
-\begin{align}\label{sim:lme_one_pred}
-\mathrm{hh~size} &= \beta_0 + \alpha_{0,\mathrm{h}[i]} + \beta_{\mathrm{A}}\mathrm{Age}_i + \epsilon_i \nonumber\\
-\alpha_{0,\mathrm{h}} &\sim \mathrm{Normal}(0, 1),~~ h=1, \cdots, \mathrm{10} \nonumber\\
-\mathrm{Age}_i &\sim \mathrm{Normal}(0.2, 1) \nonumber\\
-\epsilon_i &\sim \mathrm{Normal}(0, 2) \nonumber\\
-\beta_0 &= 1.5 \nonumber\\
-\beta_{\mathrm{A}} &= 0.1 \nonumber\\
-i &= 1,\cdots, 50
-\end{align}
-%
-The second and third models add the random effects, $\alpha_{\mathrm{h}[i]}$, to models~\ref{sim:lm_nf_interaction} and~\ref{sim:lm_f_interaction}, respectively.
-Figre~\ref{fig:pred_age_cont_lme_plots} shows and compares predictor efffects for the three models.
-%
-\begin{figure}[H]
-\begin{center}
-<<pred_age_cont_lme_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
-print(pred_age_cont_lme_plots)
-@
-\end{center}
-\caption{A comparison of predictor effects and their corresponding CIs.
-Figure a) shows model with no interaction, Figure b) shows model with non-focal predictors interacting with each other, and Figure c) shows model with focal and non-focal predictors interacting.
-The dotted yellow and grey horizontal lines are the expected and observed/true means, respectively.
-The vertical grey line is the mean of the focal predictor (model center).
-The solid black, red (non-focal uncertainty excluded) and blue lines (all uncertainty included) are the predictor effects while their corresponding CIs are the dotted black, red and blue lines, respectively.
-For properly generated predictor effects, we expect the yellow and the grey horizontal dotted lines to overlay each other and intersect with the black, blue, red solid trend lines, and the vertical dotted grey line at the same point (model center).}
-\label{fig:pred_age_cont_lme_plots}
-\end{figure}
-%
-
-\subsubsection{Multiple outcomes}
-
-We now consider a bit more complicated model involving more two outcome variables (services) -- household size and rental price.
-In particular, we simulate an intercept only multivariate random effect model to, \emph{jointly}, predict household size and rental price based on age and wealth index.
-Each household has its own correlated effect on the intercepts (random-intercept).
-In addition, we add latent variable to account for unobserved correlations.
-%
-\begin{align}\label{sim:lme_multi_outcome}
-\mathrm{y}_{\mathrm{h}[i]\mathrm{s}} &= \beta_{0,s} + \alpha_{0,\mathrm{h}[i]\mathrm{s}} + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \beta_{\mathrm{Ls}}\mathrm{Latent}_{i\mathrm{s}} + \epsilon_{i\mathrm{s}} \nonumber\\
-\alpha_{0,\mathrm{h}\mathrm{s}} &\sim \mathrm{Normal}(0, 2),~~ h=1, \cdots, \mathrm{H} \nonumber\\
-\mathrm{Age}_i &\sim \mathrm{Normal}(0.2, 1) \nonumber\\
-\mathrm{Wealthindex}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
-\mathrm{Latent}_{i\mathrm{s}} &\sim \mathrm{Normal}(0, 1) \nonumber\\
-\epsilon_{i\mathrm{s}} &\sim \mathrm{Normal}(0, 1) \nonumber\\
-\beta_0 &= 1.5 \nonumber\\
-\beta_{\mathrm{A}} &= 0.1 \nonumber\\
-\beta_{\mathrm{W}} &= 2 \nonumber\\
-\beta_{\mathrm{L}1} &= 1 \nonumber\\
-\beta_{\mathrm{L}2} &= -5 \nonumber\\
-i &= 1,\cdots, 100
-\end{align}
-%
-Where:
-\begin{itemize}
-\item $\beta_{0,s}$ \textbf{is the service specific effect???}, i.e., baseline household size and rental price
-\item $\alpha_{0,\mathrm{h}\mathrm{s}}$ household specific random intercepts, i.e., captures the offset from the overall baseline (mean)
-\end{itemize}
-
-Predictor effects for the above model are show in Figure~\ref{fig:pred_cont_joint_plots}.
-The top plot compares centered and model centered predictor effects with the observed average household size and rental price.
-The point estimates (black and blue dots) are the expected baseline household size and the rental price.
-The two bottom plots compares the predictor effect (of age) on the predicted household size and rental price.
-Again, in the case of joint model, at the model center, the expected predictions are very similar to the observed marginal averages for the outcomes.
-%
-\begin{figure}[H]
-\begin{center}
-<<pred_cont_joint_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
-print(pred_cont_joint_plots)
-@
-\end{center}
-\caption{Predictor effect of age on the predicted household size and rental price.
-The top plots shows the baseline (mean) predicted household size and rental price.
-The bottom plots compares the predictor effect of age on the predicted household size and rental price.
-The dotted yellow and grey horizontal lines are the expected and observed/true means, respectively.
-The vertical grey line is the mean of the focal predictor (model center).
-The solid black, red (non-focal uncertainty excluded) and blue lines (all uncertainty included) are the predictor effects while their corresponding CIs are the dotted black, red and blue lines, respectively.
-For properly generated predictor effects, we expect the yellow and the grey horizontal dotted lines to overlay each other and intersect with the black, blue, red solid trend lines, and the vertical dotted grey line at the same point (model center).
-Without removing the variations due to non-focal predictors, the confidence bands are very wide ("everything" in the figure).
-However, when CIs are centered, we are only focussing on the uncertainty due to the focal predictor.}
-\label{fig:pred_cont_joint_plots}
-\end{figure}
-%
-
-\section{Bias correction}\label{sec:bias_correction}
-
-\clearpage
-% References
-\bibliographystyle{plainnat}
-\bibliography{variable_predictions}
-
-\end{document}
+```{r echo=FALSE, fig.height=10}
+print(pred_bin_corr_plots)
+```

### 2021-11-02 understanding_mm.rmd
Autocommit

---
title: "How to we handle the categorical predictors?"
author: Bicko, Jonathan & Ben
date: "2021 Nov 01 (Mon)"
compact-title: false
output:
  pdf_document:
    extra_dependencies: ["float"]
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(fig.pos = "H"
	, out.extra = ""
	, message=FALSE
	, warning=FALSE
	, echo=FALSE
	, comment=""
)

library(shellpipes)
library(dplyr)
library(ggplot2)
library(vareffects); varefftheme()
library(emmeans)
library(effects)

commandEnvironments()
makeGraphics()

set.seed(2121)

load("variable_predictions_funs.rda")
```

\section{Single categorical, single continuous predictors}

```{r global_vars}
N <- 100
```

```{r single_cat_single_cont}
single_cat_single_cont_sim <- linearsim(nHH=N
	, form=~1+x1+x2
	, betas=NULL
	, pgausian=list(p=1)
	, pcat=list(p=1
		, fun=sample
		, nlevels=3
		, labels=c("Christian","Muslim","Jews")
		, prob=c(0.4, 0.35, 0.25)
		, replace=TRUE
	)
	, link_scale=TRUE
	, vnames=c("age", "rel", "hhsize")
)
single_cat_single_cont_sim_df <- single_cat_single_cont_sim$data
head(single_cat_single_cont_sim_df)
```

```{r single_cat_single_cont_mod}
single_cat_single_cont_mod <- lm(hhsize ~ age + rel, data=single_cat_single_cont_sim_df)
```

\subsection{religion}

\subsubsection{varpred}

```{r single_cat_single_cont_vpred1}
single_cat_single_cont_vpred1 <- varpred(single_cat_single_cont_mod
	, "rel"
	, handle.inter="emmeans"
	, returnall=TRUE
)
single_cat_single_cont_vpred1$raw$model.matrix
```

\subsubsection{emmeans}

```{r single_cat_single_cont_em1}
single_cat_single_cont_em1 <- emmeans(single_cat_single_cont_mod
	, specs=~rel
)
single_cat_single_cont_em1@linfct
```

\subsubsection{effects}

```{r single_cat_single_cont_eff1}
single_cat_single_cont_eff1 <- Effect("rel"
	, single_cat_single_cont_mod
)
eff_mm <- single_cat_single_cont_eff1$model.matrix
attr(eff_mm, "contrasts") <- NULL
attr(eff_mm, "assign") <- NULL
eff_mm
```

\subsection{age}

\subsubsection{varpred}

```{r single_cat_single_cont_vpred_age}
quant <- seq(0,1,length.out=4)
age_at <- quantile(single_cat_single_cont_sim_df$age, quant)
single_cat_single_cont_vpred_age <- varpred(single_cat_single_cont_mod
	, "age"
	, at=list(age=age_at)
	, handle.inter="emmeans"
	, returnall=TRUE
)
single_cat_single_cont_vpred_age$raw$model.matrix
```

\subsubsection{emmeans}

```{r single_cat_single_cont_em_age}
single_cat_single_cont_em_age <- emmeans(single_cat_single_cont_mod
	, specs=~age
	, at=list(age=age_at)
	, weights="proportional"
)
single_cat_single_cont_em_age@linfct
```

\subsubsection{effects}

```{r single_cat_single_cont_eff_age}
single_cat_single_cont_eff_age <- Effect("age"
	, single_cat_single_cont_mod
	, xlevels=list(age=age_at)
)
eff_mm <- single_cat_single_cont_eff_age$model.matrix
attr(eff_mm, "contrasts") <- NULL
attr(eff_mm, "assign") <- NULL
eff_mm
```

\section{age and religion interaction}

```{r single_cat_single_cont_mod_inter}
single_cat_single_cont_mod_inter <- lm(hhsize ~ age*rel, data=single_cat_single_cont_sim_df)
```

\subsection{religion$\star$age}

\subsubsection{varpred}

```{r single_cat_single_cont_vpred1_inter}
age_mean <- mean(single_cat_single_cont_sim_df$age)
single_cat_single_cont_vpred1_inter <- varpred(single_cat_single_cont_mod_inter
	, c("rel", "age")
	, x.var="rel"
	, at=list(age=age_at)
	, handle.inter="emmeans"
	, returnall=TRUE
)
single_cat_single_cont_vpred1_inter$raw$model.matrix
```

\subsubsection{emmeans}

```{r single_cat_single_cont_em1_inter}
single_cat_single_cont_em1_inter <- emmeans(single_cat_single_cont_mod_inter
	, specs=~rel*age
	, at=list(age=age_at)
)
single_cat_single_cont_em1_inter@linfct
```

\subsubsection{effects}

```{r single_cat_single_cont_eff1_inter}
single_cat_single_cont_eff1_inter <- Effect(c("rel", "age")
	, single_cat_single_cont_mod_inter
	, x.var="rel"
	, xlevels=list(age=age_at)
)
eff_mm <- single_cat_single_cont_eff1_inter$model.matrix
attr(eff_mm, "contrasts") <- NULL
attr(eff_mm, "assign") <- NULL
eff_mm
```

\subsection{Ignore interaction}

\subsection{religion}

\subsubsection{varpred}

```{r single_cat_single_cont_vpred1_inter_ign}
single_cat_single_cont_vpred1_inter_ign <- varpred(single_cat_single_cont_mod_inter
	, "rel"
	, handle.inter="emmeans"
	, returnall=TRUE
)
single_cat_single_cont_vpred1_inter_ign$raw$model.matrix
```

\subsubsection{emmeans}

```{r single_cat_single_cont_em1_inter_ign}
single_cat_single_cont_em1_inter_ign <- emmeans(single_cat_single_cont_mod_inter
	, specs=~rel
	, nesting=NULL
)
single_cat_single_cont_em1_inter_ign@linfct
```

\subsubsection{effects}

```{r single_cat_single_cont_eff1_inter_ign}
single_cat_single_cont_eff1_inter_ign <- Effect("rel"
	, single_cat_single_cont_mod_inter
)
eff_mm <- single_cat_single_cont_eff1_inter_ign$model.matrix
attr(eff_mm, "contrasts") <- NULL
attr(eff_mm, "assign") <- NULL
eff_mm
```

\subsection{age}

\subsubsection{varpred}

```{r single_cat_single_cont_vpred_age_inter_ign}
single_cat_single_cont_vpred_age_inter_ign <- varpred(single_cat_single_cont_mod_inter
	, "age"
	, at=list(age=age_at)
	, handle.inter="emmeans"
	, returnall=TRUE
)
single_cat_single_cont_vpred_age_inter_ign$raw$model.matrix
```

\subsubsection{emmeans}

```{r single_cat_single_cont_em_age_inter_ign}
single_cat_single_cont_em_age_inter_ign <- emmeans(single_cat_single_cont_mod_inter
	, specs=~age
#	, cov.keep="age"
	, at=list(age=age_at)
	, nesting=NULL
	, weights="proportional"
)
single_cat_single_cont_em_age_inter_ign@linfct
```

\subsubsection{effects}

```{r single_cat_single_cont_eff_age_inter_ign}
single_cat_single_cont_eff_age_inter_ign <- Effect("age"
	, single_cat_single_cont_mod_inter
	, xlevels=list(age=age_at)
)
eff_mm <- single_cat_single_cont_eff_age_inter_ign$model.matrix
attr(eff_mm, "contrasts") <- NULL
attr(eff_mm, "assign") <- NULL
eff_mm
```

\section{Two categorical, single continuous predictors}

```{r two_cat_single_cont}
two_cat_sim <- linearsim(nHH=N
	, form=~1+x1+x2+x3
	, betas=NULL
	, pgausian=list(p=1)
	, pcat=list(p=2
		, fun=sample
		, nlevels=c(3,2)
		, labels=list(c("Christian","Muslim","Jews"), c("Male", "Female"))
		, prob=list(c(0.4, 0.35, 0.25), c(0.6, 0.4))
		, replace=TRUE
	)
	, link_scale=TRUE
	, vnames=c("age", "rel", "gender", "hhsize")
)
two_cat_sim_df <- two_cat_sim$data
age_at <- quantile(two_cat_sim_df$age, quant)
head(two_cat_sim_df)
```

```{r two_cat_mod}
two_cat_mod <- lm(hhsize~rel*gender+age, data=two_cat_sim_df)
```

\subsection{religion$\star$gender}

\subsubsection{varpred}

```{r two_cat_vpred_inter}
two_cat_vpred_inter <- varpred(two_cat_mod
	, c("rel", "gender")
	, x.var="rel"
	, returnall=TRUE
)
two_cat_vpred_inter$raw$model.matrix
```

\subsubsection{emmeans}

```{r two_cat_em_inter}
two_cat_em_inter <- emmeans(two_cat_mod
	, specs=~rel*gender
	, weights="proportional"
)
two_cat_em_inter@linfct
```

\subsubsection{effects}

```{r two_cat_eff_inter}
two_cat_eff_inter <- Effect(c("rel", "gender")
	, two_cat_mod
	, x.var="rel"
)
eff_mm <- two_cat_eff_inter$model.matrix
attr(eff_mm, "contrasts") <- NULL
attr(eff_mm, "assign") <- NULL
eff_mm
```

\subsection{age}

\subsubsection{varpred}

```{r two_cat_vpred_inter_age}
two_cat_vpred_inter_age <- varpred(two_cat_mod
	, "age"
	, at=list(age=age_at)
	, returnall=TRUE
)
two_cat_vpred_inter_age$raw$model.matrix
```

\subsubsection{emmeans}

```{r two_cat_em_inter_age}
two_cat_em_inter_age <- emmeans(two_cat_mod
	, specs=~age
	, at=list(age=age_at)
	, weights="proportional"
	, nesting="rel %in% gender"
)
two_cat_em_inter_age@linfct
```

\subsubsection{effects}

```{r two_cat_eff_inter_age}
two_cat_eff_inter_age <- Effect("age"
	, two_cat_mod
	, xlevels=list(age=age_at)
)
eff_mm <- two_cat_eff_inter_age$model.matrix
attr(eff_mm, "contrasts") <- NULL
attr(eff_mm, "assign") <- NULL
eff_mm
```

\subsection{religion}

\subsubsection{varpred}

```{r two_cat_vpred_inter_gender}
two_cat_vpred_inter_rel <- varpred(two_cat_mod
	, "rel"
	, returnall=TRUE
)
two_cat_vpred_inter_rel$raw$model.matrix
```

\subsubsection{emmeans}

```{r two_cat_em_inter_rel}
two_cat_em_inter_rel <- emmeans(two_cat_mod
	, specs=~rel
	, weights="proportional"
	, nesting=NULL
)
two_cat_em_inter_rel@linfct
```

**Check**

```{r}
colMeans(model.matrix(two_cat_mod))
```

### 2021-11-02 R/effectsfuns.R

@@ -291,7 +291,8 @@ varpred <- function(mod
 		mm <- NULL
 	}
 	
-
+	attr(mm,"contrasts") <- NULL
+	attr(mm, "assign") <- NULL
 	out <- list(term = paste(focal.predictors, collapse="*")
 		, formula = formula(mod)
 		, response = get_response(mod)

### 2021-11-02 binom_correlated_preds.R

@@ -3,6 +3,7 @@ library(vareffects); varefftheme()
 library(ggpubr)
 library(ggplot2)
 library(dplyr)
+library(emmeans)
 
 commandEnvironments()
 makeGraphics()
@@ -19,6 +20,11 @@ pred_x_centered_bin_corr <- varpred(mod_bin_corr
 	, isolate=TRUE
 	, modelname="centered mm"
 )
+true_corr <- mean(pred_x_centered_bin_corr$preds$fit)
+print(true_corr)
+em_preds <- emmeans(mod_bin_corr, "x", at=list(x=pred_x_centered_bin_corr$preds$x), type="response")
+em_preds <- as.data.frame(em_preds)
+mean(em_preds$prob)
 
 ### Combine all predictions
 vlist <- list(pred_x_trad_bin_corr, pred_x_centered_bin_corr)
@@ -31,7 +37,8 @@ pred_x_bin_corr_plots <- (comparevarpred(vlist=vlist
 	+ scale_colour_manual(breaks = c("everything", "centered mm")
 		, values=c(everything="blue", "centered mm"="black")
 	)
-	+ geom_hline(data=observed_df_corr, aes(yintercept=z), lty=2)
+	+ geom_hline(data=observed_df_corr, aes(yintercept=z), lty=2, col="red")
+	+ geom_hline(yintercept=true_corr, lty=2)
 	+ labs(y="Predictions", colour="Method", title="Correlated")
 	+ theme(legend.position="bottom")
 )
@@ -49,6 +56,7 @@ pred_x_centered_bin_noncorr <- varpred(mod_bin_noncorr
 	, isolate=TRUE
 	, modelname="centered mm"
 )
+true_noncorr <- mean(pred_x_centered_bin_noncorr$preds$fit)
 
 ### Combine all predictions
 vlist <- list(pred_x_trad_bin_noncorr, pred_x_centered_bin_noncorr)
@@ -61,7 +69,8 @@ pred_x_bin_noncorr_plots <- (comparevarpred(vlist=vlist
 	+ scale_colour_manual(breaks = c("everything", "centered mm")
 		, values=c(everything="blue", "centered mm"="black")
 	)
-	+ geom_hline(data=observed_df_noncorr, aes(yintercept=z), lty=2)
+	+ geom_hline(data=observed_df_noncorr, aes(yintercept=z), lty=2, col="red")
+	+ geom_hline(yintercept=true_noncorr, lty=2)
[…]

### 2021-11-02 variable_predictions_funs.R

@@ -64,20 +64,27 @@ linearsim <- function(nHH=1000, perHH=1, form=~1+x1+x2+x3
 		Xcat <- list()
 		cfun <- pcat$fun
 		if (is.null(cfun)) cfun <- sample
+		labels_temp <- pcat$labels
+		prob_temp <- pcat$prob
 		for (nl in 1:cp) {
+			if(is.list(labels)) {
+				labels <- labels_temp[[nl]]
+				prob <- prob_temp[[nl]]
+			} else {
+				labels <- labels_temp
+				prob <- prob_temp
+			}
 			if (nlevels[[nl]]<2)stop("Levels of categorical variable must be >1")
-			labels <- pcat$labels
-			if(is.list(labels)) labels <- labels[[nl]]
 			if (is.null(labels)) {
 				levs <- sample(LETTERS, nlevels[[nl]], replace=FALSE)
 			} else {
 				levs <- labels
+				if (is.list(levs)) levs <- levs[[nl]]
 			}
 			if (length(levs) != nlevels[[nl]])
 				stop("nlevels must be equal to the length of labels")
 			pcat$labels <- NULL
 			pcat$x <- levs
-			prob <- pcat$prob
 			if(is.list(prob)) prob <- prob[[nl]]
 			if (is.null(prob)) {
 				prob <- rep(1/nlevels[[nl]], nlevels[[nl]])

### 2021-11-03 R/effectsfuns.R
Consistent with emmeans

@@ -353,7 +353,7 @@ varpred <- function(mod
 	attr(result, "response") <- out$response
 	attr(result, "x.var") <- out$x.var 
 	if (returnall) {
-		res <- list(preds = result, offset=out$offset, bias.adjust.sigma=out$bias.adjust.sigma, raw = out, factor.cols=factor.cols)
+		res <- list(preds = result, offset=out$offset, bias.adjust.sigma=out$bias.adjust.sigma, raw = out, factor.cols=factor.cols, ff=factor.type)
 	} else {
 		res <- list(preds = result, offset=out$offset, bias.adjust.sigma=out$bias.adjust.sigma)
 	}

### 2021-11-03 R/utilities.R

@@ -291,12 +291,13 @@ clean_model <- function(focal.predictors, mod, xlevels
   names(factor.cols) <- cnames
 
   for (name in all.predictors){
-    if (check_factor(name, mod)) {
+    ff_check <- check_factor(name, mod)
+    if (ff_check) {
       factor.cols[grep(paste("^", name, sep=""), cnames)] <- TRUE
-    } 
-#	 else {
-#      factor.cols[grep(paste0(":", name), cnames)] <- FALSE 
-#	 }
+    }
+	 if (!ff_check) {
+	 	factor.cols[grep(paste0(":", name, "|", name, ":"), cnames)] <- FALSE
+	 }
   }
 #  factor.cols[grep(":", cnames)] <- FALSE   
 
@@ -487,10 +488,15 @@ get_model_matrix <- function(mod, mod.matrix, X.mod, factor.cols, cnames
     for (name in cnames){
       components <- unlist(strsplit(name, ':'))
       components <- components[components %in% cnames]
+		if (length(components)==1) components <- unique(c(components, name))
       if (length(components) > 1) {
-#			if (!factor.cols[[name]]) {
+			if (factor.type[[x.var]]) {
 				mod.matrix[,name] <- apply(mod.matrix[,components], 1, prod)
-#			}
+			} else {
+				if (grepl(x.var, name)) {
+					mod.matrix[,name] <- apply(mod.matrix[,components], 1, prod)
+				}
+			}
       }
     }
   }

### 2021-11-09 bias_correction_methods.rmd
A lot of pieces

--- 2021-09-07 bias_correction.rmd
@@
 library(shellpipes)
-library(glmmTMB)
 library(dplyr)
@@
 library(vareffects); varefftheme()
-library(ggpubr)
 
@@
 
-set.seed(2121)
-## Functions
-plotEsize <- function(df, pos = 0.5, col_lab = ""){
-	pos <- position_dodge(pos)
-	p1 <- (ggplot(df, aes(x = reorder(term, -estimate), y = estimate, colour="Estimate"))
-		+ geom_pointrange(aes(ymin = conf.low, ymax = conf.high), size=0.2, position = pos)
-		+ geom_hline(yintercept=0,lty=2, size=0.2)
-		+ coord_flip()
-		+ labs(x="", y = "Estimate")
-	)
-}
+set.seed(9991)
 ```
-
-\section{Introduction}
-
-Generating predictions from regression models can be challenging and depends on whether the relationship between the response variable and the predictors is linear or nonlinear.
-For example, if response, $Y$, changes nonlinearly with the predictor variable, $X$, averaged response variable with respect to the predictor, $E(Y(X))$, does not necessarily equal the response at the mean of the predictor, $Y(E(X)$.
-This can be understood in relation to Jensen's inequality which states that, for a nonlinear function, $Y(X)$, then $E(Y(X)) > Y(E(X))$, if $Y(X)$ is positive second derivative; and $E(Y(X)) < Y(E(X))$ if $Y(X)$ is negative second derivative.
-Currently, existing R packages for predicting responses make distribution assumptions about the non-focal predictors, (for example, averaging the non-focal predictors at various levels of focal predictors), leading to potential biasness if the particular predictor is not well represented, or as a result of nonlinear averaging.
-To understand what we are trying to achieve, we first define the following:
-
-\subsection{Prediction plots}
-
-Notation :
 
@@
 
-The purpose and goal of a *prediction plot* seems fairly straightforward; for specified values of (a) focal predictor(s), we want to give a point estimate and confidence intervals for the prediction of the model for a "typical" (= random sample over the multivariate distribution of non-focal parameters) individual with those values of the predictors.
+_\
 
-**BB**'s
+\section{Population-based approach for bias correction}
 
-("what is the expected probability of having clean water for a 25-year-old male"?).
-These are most of the problems we have actually been addressing with our bias correction stuff above.
-To do these computations, we need to take the values of the non-focal predictors (or their means and covariances) from the *conditional distribution*.
-(If the focal predictors are discrete, we condition on exact values; if they are continuous/have mostly unique values, we condition on appropriately sized bins.) In other words,
-$$
-\underset{\cdist}{\textrm{mean}}  g^{-1} \left(\eta(\nset{x},x_f) \right)
-$$
+\subsection{Binned non-focal linear predictor}
 
-Suppose we consider the values of non-focal predictors, to implement this:
+To implement this:
+\begin{itemize}
+\item compute linear predictor of the non-focal predictors, $\nset{\eta} = \sum \nset{\beta} \nset{x}$
+\item find a list of vectors of observations of $\nset{\eta}$ associated with each value (bin) of the focal predictor, $\nset{{\eta_j}}$, $j = 1, 2, \cdots$
+\item for each $\nset{{\eta_j}}$:
+\begin{itemize}
+\item compute $\hat{y}_j  = \textrm{mean} ~ g^{-1} \left(\beta_f x_{j_f} + \nset{{\eta_j}}\right)$
+\end{itemize}
+\end{itemize}
 
-- compute linear predictor of the non-focal predictors, $\nset{\eta} = \sum \nset{\beta} \nset{x}$
-- find a list of vectors of observations of $\nset{\eta}$ associated with each value (bin) of the focal predictor, $\nset{{\eta_j}}$, $j = 1, 2, \cdots$
-- for each $\nset{{\eta_j}}$:
-	- compute $\hat{y}_j  = \textrm{mean} ~ g^{-1} \left(\beta_f x_{j_f} + \nset{{\eta_j}}\right)$
+If we compute the individual back-transformed predictions for a poorly sampled/finely spaced set of focal values, we will get a noisy prediction line as the values of the non-focal predictors shift across the focal values.
+Simple example: suppose everyone below the median age has wealth index $w_1$, everyone above the median has $w_2$.
+Then the predicted value will have a discontinuity at the median age.
+We can deal with this by taking bigger bins (a form of smoothing), or by post-smoothing the results (by loess, for example).
+The principled form of this would be to assume/recognize that our uneven distribution of observed non-focal predictors actually represents a sample of a distribution that will vary \emph{smoothly} as a function of the focal predictor.
 
-_What's the expected probability of having clean water for all $x$-year-old with the **corresponding** wealth index??_
+\subsection{Whole population non-focal linear predictor}
 
-**JD**'s and **SC**'s
+Instead of binning non-focal associated with particular level of focal predictor, we add the overall contribution of the non-focal predictor to the corresponding value of the focal linear predictor.
+In particular:
 
-- compute linear predictor of the non-focal predictors, $\nset{\eta} = \sum \nset{\beta} \nset{x}$
-- for every value of the focal predictor, $x_{j_f}$:
-	- compute $\hat{y}_j  = \textrm{mean} ~ g^{-1} \left(\beta_f x_{j_f} + \nset{\eta}\right)$
+\begin{itemize}
+\item compute linear predictor of the non-focal predictors, $\nset{\eta} = \sum \nset{\beta} \nset{x}$
+\item for every value of the focal predictor, $x_{j_f}$:
+\begin{itemize}
+\item compute $\hat{y}_j  = \textrm{mean} ~ g^{-1} \left(\beta_f x_{j_f} + \nset{\eta}\right)$
+\end{itemize}
+\end{itemize}
 
-The major difference between **BB**'s and **JD**'s approach is that in the former case, we consider non-focal linear predictor corresponding to specific value (bin) of the focal predictor, while in the later case, for each value of the focal predictor, we consider linear predictor associated with entire population.
+\section{Delta method}
 
-**Questions**
-
-- What does the second case (JD's) represent?
-- Which is the correct approach?
-
-\subsection{Simulation}
-
-We implement and apply these methods in the context of both simple generalized linear and mixed effect models, using simulated data sets -- for univariate and multivariate models.
-
-We start with a univariate case where we have only one predictor.
-
-\section{Simple fixed effect model}
+Suppose that we have a response $Y$ and the transformed response on the link scale, $\eta$ and an inverse-link function $h(.) = g^{-1}(.)$.
+We consider $\eta$ as a one-dimensional random variable with mean ($\mu_{\eta}$) and standard deviation ($\sigma_{\eta}$).
+We are interested in the expected value of $h(\eta)$, i.e., the back-transformed values so that $Y = h(\eta)$.
+More specifically, based on a second-order Taylor expansion
 
 \begin{align*}
-\mathrm{logit(status} = 1) &= \eta \\ 
-\eta &= \beta_0 + \beta_{\mathrm{A}} \mathrm{Age} + \beta_{\mathrm{W}} \mathrm{Wealthindex} \\
-\mathrm{Age} &\sim \mathrm{Normal}(0.2, 1) \\
-\mathrm{Wealthindex} &\sim \mathrm{Normal}(0, 1) \\
-\beta_0 &= 1.5 \\
-\beta_{\mathrm{A}} &= 1.0 \\
-\beta_{\mathrm{W}} &= 2
+Y \approx h(\mu_{\eta}) + h'(\eta)(\eta - \mu_{\eta}) + \frac{1}{2}h''(\eta)(\eta - \mu_{\eta})^2
 \end{align*}
 
-Considering $N=1e4$
+so that
 
-```{r sim_simple}
+\begin{align*}
+E(Y) = h(\eta) + \frac{1}{2}h''(\mu_{\eta})\sigma^2_{\eta}
+\end{align*}
 
-simplesim <- function(N=1e4, beta0=1.5, betaA=1.0, betaW=2
-		, age_sd=1, age_mean=0.2
-		, wealth_sd=1, wealth_mean=0
-	) {
-	age <- rnorm(N, age_mean, age_sd)
-	wealthindex <- rnorm(N, wealth_mean, wealth_sd)
-	eta <- beta0 + betaA * age + betaW * wealthindex
-	sim_df <- (data.frame(age=age, wealthindex=wealthindex, eta=eta)
-		%>% mutate(status = rbinom(N, 1, plogis(eta)))
-		%>% select(-eta)
-	)
-	return(sim_df)
-}
-sim_df <- simplesim()
-true_prop <- mean(sim_df$status)
-print(true_prop)
-head(sim_df)
-```
+\section{McCulloch and Searle (2001)}
+\[
+E(Y) \approx h\left(\eta-\frac{0.5\sigma^2\tanh(\eta(1 + 2\exp(-0.5\sigma^2)))}{6}\right)
+\]
 
-\subsection{Simple logistic model}
-
-```{r simple_logistic}
-simple_mod <- glm(status ~ age + wealthindex, data = sim_df, family="binomial")
-```
-
-\subsection{Variable predictions}
-
-```{r pop_ave_fun, echo=FALSE}
-binfun <- function(mod, focal, non.focal, bins=50, ...) {
-	mf <- model.frame(mod)
-	mm <- (mf
-		%>% select_at(c(focal, non.focal))
-	)
-	check_df <- (mf
-		%>% arrange_at(focal)
-		%>% mutate(bin=ceiling(row_number()*bins/nrow(.)))
-		%>% group_by(bin)
-		%>% summarise_all(mean)
-		%>% mutate(model="binned")
-	)
-	return(check_df)
-}
-
-```
-
-```{r simple_pred_age, echo=FALSE, fig.cap="Prediction plots with N=1e4: JD's approach on the left; BB's on the right."} 
-binned_df <- binfun(simple_mod, "age", "wealthindex")
-
-## JD's approach
-pred_age_jd <- predictionfun(simple_mod, "age", isolate=TRUE, include.nonfocal="jd", modelname="JD")
-age_plot_jd <- (ggplot(pred_age_jd, aes(x=age, y=fit))
-	+ geom_line()
-	+ geom_line(aes(y=lwr), lty=2)
-	+ geom_line(aes(y=upr), lty=2)
-	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
-	+ geom_point(data=binned_df, aes(x=age, y=status, color="binned"))
-	+ labs(y="Probability")
-	+ theme(legend.position="none")
-)
-
-## BB's approach
-pred_age_bb <- predictionfun(simple_mod, "age", isolate=TRUE, include.nonfocal="bb", modelname="BB")
-age_plot_bb <- (ggplot(pred_age_bb, aes(x=age, y=fit))
-	+ geom_line()
-	+ geom_line(aes(y=lwr), lty=2)
-	+ geom_line(aes(y=upr), lty=2)
-	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
-	+ geom_point(data=binned_df, aes(x=age, y=status, color="binned"))
-	+ geom_smooth(colour="black", size=0.5)
-	+ labs(y="Probability")
-	+ theme(legend.position="none")
-)
-
-ggarrange(age_plot_jd, age_plot_bb + rremove("ylab"), common.legend=FALSE)
-
-```
-
-Considering $N=1e5$ and then refit the model
-
-```{r sim_simple_ne5}
-sim_df <- simplesim(N=1e5)
-true_prop <- mean(sim_df$status)
-simple_mod <- glm(status ~ age + wealthindex, data = sim_df, family="binomial")
-```
-
-```{r simple_pred_age_ne5, echo=FALSE, fig.cap="Prediction plots with N=1e5: JD's approach on the left; BB's on the right."} 
-binned_df <- binfun(simple_mod, "age", "wealthindex")
-
-## JD's approach
-pred_age_jd <- predictionfun(simple_mod, "age", isolate=TRUE, include.nonfocal="jd", modelname="JD")
-age_plot_jd <- (ggplot(pred_age_jd, aes(x=age, y=fit))
-	+ geom_line()
-	+ geom_line(aes(y=lwr), lty=2)
-	+ geom_line(aes(y=upr), lty=2)
-	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
-	+ geom_point(data=binned_df, aes(x=age, y=status, color="binned"))
-	+ labs(y="Probability")
-	+ theme(legend.position="none")
-)
-
-## BB's approach
-pred_age_bb <- predictionfun(simple_mod, "age", isolate=TRUE, include.nonfocal="bb", modelname="BB")
-age_plot_bb <- (ggplot(pred_age_bb, aes(x=age, y=fit))
-	+ geom_line()
-	+ geom_line(aes(y=lwr), lty=2)
-	+ geom_line(aes(y=upr), lty=2)
-	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
-	+ geom_point(data=binned_df, aes(x=age, y=status, color="binned"))
-	+ geom_smooth(colour="black", size=0.5)
-	+ labs(y="Probability")
-	+ theme(legend.position="none")
-)
-
-ggarrange(age_plot_jd, age_plot_bb + rremove("ylab"), common.legend=FALSE)
-
-```
+\section{Diggle et al. (2004)}
+\[
+E(Y) \approx h\left(\frac{\eta}{\sqrt{1 + \left(\frac{16\sqrt{3}}{15\pi}\right)^2\sigma^2}}\right)
+\]

### 2021-11-09 understanding_mm_special.rmd
A lot of pieces

---
title: "How to we handle the categorical predictors?"
author: Bicko, Jonathan & Ben
date: "2021 Nov 06 (Sat)"
compact-title: false
output:
  pdf_document:
    extra_dependencies: ["float"]
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(fig.pos = "H"
	, out.extra = ""
	, message=FALSE
	, warning=FALSE
	, echo=FALSE
	, comment=""
)

library(shellpipes)
library(dplyr)
library(ggplot2)
library(vareffects); varefftheme()
library(emmeans)
library(effects)

commandEnvironments()
makeGraphics()

set.seed(2121)

load("variable_predictions_funs.rda")

cols <- c(Effect="black", varpred="red", emmeans="blue")
```

\section{Two categorical, single continuous predictors}

\begin{align}\label{sim:lm_no_interaction_cat}
\mathrm{hh~size} &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{[\mathrm{r}]}\mathrm{Religion_i} + \beta_{[\mathrm{g}]}{\mathrm{Gender}_i} + \beta_{[\mathrm{rg}]}\mathrm{Gender_i
\times Religion_i} + \epsilon_i \nonumber\\
\mathrm{Age}_i &\sim \mathrm{Normal}(0, 3) \nonumber\\
\mathrm{Prob(Gender_{i})} &= \begin{cases}
   0.4 & \mathrm{i=Female} \nonumber\\
   0.6 & \mathrm{i=Male} \nonumber\\
  \end{cases} \nonumber\\
\mathrm{Prob(Religion_{i})} &= \begin{cases}
   0.6 & \mathrm{i=Christian} \nonumber\\
   0.3 & \mathrm{i=Muslim} \nonumber\\
   0.1 & \mathrm{i=Jew} \nonumber\\
  \end{cases} \nonumber\\
\epsilon_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
\beta_0 &= 1.5 \nonumber\\
\beta_{\mathrm{A}} &= 0.8 \nonumber\\
\vdots \nonumber \\
i &= 1,\cdots, 100
\end{align}

```{r depend_eff, echo=TRUE}
N <- 100
quant <- seq(0,1,length.out=50)
## Betas
b0 <- 10
b_age <- 0.8
b_jew <- 1
b_muslim <- 0.4
b_male <- 2
b_male_jew <- 0.7
b_male_muslim <- 0.2
betas <- c(b0, b_age, b_jew, b_muslim, b_male, b_male_jew, b_male_muslim)

## Proportions
rel_prob <- c(Christian=0.6, Muslim=0.3, Jew=0.1)
gender_prob <- c(Male=0.6, Female=0.4)

dat <- tibble(age=rnorm(N, 0, 3)
	, rel=sample(c("Christian", "Muslim", "Jew"), size=N, prob=rel_prob, replace=TRUE)
	, gender=sample(c("Male", "Female"), size=N, prob=gender_prob, replace=TRUE)
)
mm <- model.matrix(~1+age+rel*gender, dat)
eta <- as.vector(mm %*% betas)

df <- (dat
	%>% mutate(hhsize=rnorm(N, mean=eta, sd=1))
)
age_at <- quantile(df$age, quant)
head(df)
```

The model:

```{r depend_eff_mod, echo=TRUE}
depend_eff_mod <- lm(hhsize~rel*gender+age, df)
```

\newpage

\subsubsection{gender$\star$religion}

- All the three methods seems to be doing the same thing:
	- product of interacting terms

```{r depend_eff_pred}
depend_eff_pred <- combinepreds(depend_eff_mod, c("emmeans", "varpred", "Effect"), c("gender","rel")
	, x.var = "gender"
	, weights="proportional"
	, x.var.factor = TRUE, plotit = TRUE
)
print(depend_eff_pred 
	+ facet_wrap(~rel, scales = "free_y")
	+ scale_colour_manual(values=cols)
)
```

\newpage

\subsubsection{Main effect of non-interacting predictor}

- If we ignore the interactions:
	- **varpred** and **emmeans** averages the columns of the interaction terms
	- **effects** computes the product of the interacting terms
	- The predictions are similar but slightly different

```{r depend_eff_age}
depend_eff_age <- combinepreds(depend_eff_mod, c("emmeans", "varpred", "Effect"), "age"
	, x.var = "age"
	, at=list(age=age_at)
	, xlevel=list(age=age_at)
	, nesting=NULL
	, weights="proportional"
	, plotit = TRUE
)
cat("Truth: \n", mean(df$hhsize))
print(depend_eff_age
	+ scale_colour_manual(values=cols)
)
```

\section{Unbalanced data}

```{r unbalanced_sim, echo=TRUE}
N <- 100
quant <- seq(0,1,length.out=50)
## Betas
b0 <- 10
b_age <- 0.8
b_jew <- 1
b_muslim <- 0.4
b_male <- 2
b_male_jew <- 0.7
b_male_muslim <- 0.2
betas <- c(b0, b_age, b_jew, b_muslim, b_male, b_male_jew, b_male_muslim)

## Proportions
rel_prob <- c(Christian=0.6, Muslim=0.3, Jew=0.1)
gender_prob <- list(Christian=c(0.97, 0.03)
	, Muslim=c(0.01, 0.99)
	, Jew=c(0.5, 0.5)
)

samplefun <- function(prob, n) {
	out <- sample(c("Male", "Female"), size=n, prob=prob, replace=TRUE)
}

dat <- tibble(age=rnorm(N, 0, 3)
	, rel=sample(c("Christian", "Muslim", "Jew"), size=N, prob=rel_prob, replace=TRUE)
)
dat <- (dat
	%>% group_by(rel)
	%>% mutate(gender=ifelse(rel=="Christian"
			, samplefun(gender_prob[["Christian"]], n())
				, ifelse(rel=="Muslim", samplefun(gender_prob[["Muslim"]], n())
				, samplefun(gender_prob[["Jew"]], n())
			)
		)
	)
	%>% ungroup()
)
head(dat)
mm <- model.matrix(~1+age+rel*gender, dat)
eta <- as.vector(mm %*% betas)

unbalanced_df <- (dat
	%>% mutate(hhsize=rnorm(N, mean=eta, sd=1))
)
age_at <- quantile(unbalanced_df$age, quant)
head(unbalanced_df)

```

The model:

```{r unbalanced_mod, echo=TRUE}
unbalanced_mod <- lm(hhsize~rel*gender+age, unbalanced_df)
```

\newpage

\subsubsection{gender$\star$religion}

- All the three methods seems to be doing the same thing:
	- product of interacting terms

```{r unbalanced_pred}
unbalanced_pred <- combinepreds(unbalanced_mod, c("emmeans", "varpred", "Effect"), c("gender","rel")
	, x.var = "gender"
	, weights="proportional"
	, x.var.factor = TRUE, plotit = TRUE
)
print(unbalanced_pred 
	+ facet_wrap(~rel, scales = "free_y")
	+ scale_colour_manual(values=cols)
)
table(unbalanced_df$rel, unbalanced_df$gender)
```

\newpage

\subsubsection{Main effect of non-interacting predictor}

- If we ignore the interactions:
	- **varpred** and **emmeans** averages the columns of the interaction terms
	- **effects** computes the product of the interacting terms
	- The predictions are similar but slightly different

- CIs excluded for clarity

```{r unbalanced_age}
unbalanced_age <- combinepreds(unbalanced_mod, c("emmeans", "varpred", "Effect"), "age"
	, x.var = "age"
	, at=list(age=age_at)
	, xlevel=list(age=age_at)
	, nesting=NULL
	, weights="proportional"
	, plotit = TRUE
	, ci=FALSE
)
cat("Truth: \n", mean(unbalanced_df$hhsize))
print(unbalanced_age
	+ scale_colour_manual(values=cols)
)
```

### 2021-11-09 R/plotsfuns.R

@@ -57,12 +57,12 @@ plot.vareffects <- function(x, ..., xlabs = NULL, ylabs = NULL, pos = 0.5, ci = 
 	nn <- unique(df$model)
 	if (!(length(nn)>1)){
 		p1 <- (ggplot(df, aes_string(x = x.var, y = "fit"), colour="black", alpha = 0.2)
-			+ guides(fill = FALSE)
+			+ guides(fill = "none")
 			+ theme(legend.position = "none")
 		)
 	} else {
 		p1 <- (ggplot(df, aes_string(x = x.var, y = "fit", colour ="model"), alpha = 0.2)
-			+ guides(fill = FALSE)
+			+ guides(fill = "none")
 			+ theme(legend.position = "right")
 		)
 	}

### 2021-11-09 mediate.R

@@ -9,7 +9,7 @@ set.seed(991)
 ## Simulate mediated variables such that:
 ### x -> y -> z
 
-N <- 100
+N <- 10000
 beta_xy <- 0.8
 beta_xz <- 0	# Set to 0 -> x: no direct of x on z
 beta_yz <- 0.8

### 2021-11-09 mediate_bin_preds_adjust.R

@@ -1,0 +1,173 @@
+library(shellpipes)
+library(vareffects); varefftheme()
+library(ggpubr)
+library(ggplot2)
+library(dplyr)
+
+commandEnvironments()
+makeGraphics()
+
+## Not mediated
+### No bias adjustment
+pred_none_notmediated <- varpred(mod_notmediated_bin
+	, "x"
+	, isolate=TRUE
+	, bias.adjust="none"
+	, modelname="none"
+)
+
+### Delta-sigma
+pred_delta_mod_notmediated <- varpred(mod_notmediated_bin
+	, "x"
+	, isolate=TRUE
+	, bias.adjust="delta"
+	, sigma="mod"
+	, modelname="delta-mod"
+)
+
+### Delta-lp
+pred_delta_lp_notmediated <- varpred(mod_notmediated_bin
+	, "x"
+	, isolate=TRUE
+	, bias.adjust="delta"
+	, sigma="lp"
+	, modelname="delta-lp"
+)
+
+### mcculloch-sigma
+pred_mc_mod_notmediated <- varpred(mod_notmediated_bin
+	, "x"
+	, isolate=TRUE
+	, bias.adjust="mcculloch"
+	, sigma="mod"
+	, modelname="mc-mod"
+)
+
+### mcculloch-sigma
+pred_mc_lp_notmediated <- varpred(mod_notmediated_bin
+	, "x"
+	, isolate=TRUE
+	, bias.adjust="mcculloch"
+	, sigma="lp"
+	, modelname="mc-lp"
+)
+
+### population
+pred_pop_notmediated <- varpred(mod_notmediated_bin
+	, "x"
+	, steps=50
+	, isolate=TRUE
+	, bias.adjust="population"
+	, modelname="pop"
+)
+
+### Combine all predictions
+vlist <- list(pred_none_notmediated
+	, pred_delta_mod_notmediated
+	, pred_delta_lp_notmediated
+	, pred_mc_mod_notmediated
+	, pred_mc_lp_notmediated
+	, pred_pop_notmediated
+)
+
+pred_notmediated_plots <- (comparevarpred(vlist=vlist
+		, lnames=NULL
+		, plotit=TRUE
+		, addmarginals=FALSE
+		, ci=FALSE
+	)
+#	+ geom_hline(yintercept=pred_prop_notmed, lty=2, col="black")
+	+ geom_hline(data=observed_df_med, aes(yintercept=zbin), lty=2, col="red")
[…]

### 2021-11-09 variable_predictions_funs.R

@@ -265,7 +265,11 @@ combinepreds <- function(mod, funs, focal, x.var, x.var.factor=FALSE, plotit=TRU
 		spec <- as.formula(paste0("~", focal))
 		focal_temp <- "xvar"
 	}
-	args <- list(mod, spec=spec, focal_predictors=focal, x.var=x.var)
+	args <- list(mod, spec=spec
+		, focal.predictors=focal
+		, focal_predictors=focal
+		, x.var=x.var
+	)
 	add_args <- list(...)
 	if (length(add_args)) args[names(add_args)] <- add_args
 
@@ -305,14 +309,18 @@ combinepreds <- function(mod, funs, focal, x.var, x.var.factor=FALSE, plotit=TRU
 	}
 	if (plotit) {
 		cols <- rainbow(length(funs))
+#		colfun <- colorRampPalette(c("black", "red"))
+#		cols <- colfun(length(funs))
 		names(cols) <- funs
-		p1 <- (vareffects:::plot.vareffects(out)
+		out <- list(out)
+		out[names(add_args)] <- add_args
+		p1 <- (do.call(vareffects:::plot.vareffects, out)
 			+ scale_colour_manual(breaks = funs, values=cols)
 			+ labs(y="Predictions", x=x.var, colour="Method")
 			+ theme(legend.position="bottom")
 		)
 		if (!x.var.factor) {
-			pred_prop_df <- (out
+			pred_prop_df <- (out[[1]]
 				%>% group_by_at(c("model", focal[!focal %in% x.var]))
 				%>% summarize(fit=mean(fit))
 				%>% data.frame()

### 2021-11-15 understanding_mm_special.rmd
Autocommit

@@
 N <- 100
-quant <- seq(0,1,length.out=50)
+quant <- seq(0,1,length.out=100)
 ## Betas
@@
 ```
+
+```{r}
+varpred(unbalanced_mod, "age", at=list(age=mean(unbalanced_df$age)))
+emmeans(unbalanced_mod, specs=~age, weights="proportional", nesting=NULL)
+```

### 2021-11-17 understanding_mm_special.rmd
Population-based...

--- 2021-11-09 understanding_mm_special.rmd
@@
 )
+
+# varpred(depend_eff_mod, "gender")
+# varpred(depend_eff_mod, "gender", bias.adjust="population")
 ```
@@
 ```{r unbalanced_sim, echo=TRUE}
-N <- 100
-quant <- seq(0,1,length.out=50)
+N <- 1000
+quant <- seq(0,1,length.out=100)
 ## Betas
@@
 
+We are interested in predicted _hhsize_ at the model center. We can explore this in two different ways:
+
+1. **Center of the focal predictor**
+$$\mathrm{E(hhsize|\{\bar{x}_{\{f\}}, \bar{x}_{\{n\}}\})}$$
+
+2. **Averaged over levels of focal predictor**
+$$\mathrm{mean ~ E(hhsize_i|\{x_{i\{f\}}, \bar{x}_{\{n\}}\})}$$
+
+A few challenges:
+
+- How do we get the representative samples of $x_{i\{f\}}$?
+	- `quantiles`
+	- `seq`
+	- `entire population`
+
+In our simulation example:
+\begin{align}
+\mathrm{E(hhsize|age)} &= \mathrm{\beta_0 + \beta_A Age + \beta_J Jew + \beta_M Muslim + \beta_M Male}\nonumber\\
+&+ \mathrm{\beta_{MJ} Male \times Jew + \beta_{MM} Male \times Muslim}
+\end{align}
+
+The non-focal categorical interactions can be handled in two ways:
+
+- product of interacting variables (**effects**)
+- model center, i.e., average of all columns of the variables in model matrix (**varpred** and **emmeans**)
+
 - If we ignore the interactions:
@@
 
-- CIs excluded for clarity
+Using Eqn. 2 with approach 1 (center of the focal predictor)
+
+<br>
 
 ```{r unbalanced_age}
-unbalanced_age <- combinepreds(unbalanced_mod, c("emmeans", "varpred", "Effect"), "age"
+mean_age <- mean(unbalanced_df$age)
+mean_hhsize <- mean(unbalanced_df$hhsize)
+unbalanced_age <- combinepreds(unbalanced_mod, c("emmeans", "varpred", "Effect")
+	, focal = "age"
+	, x.var = "age"
+	, at=list(age=mean_age)
+	, xlevel=list(age=rep(mean_age,5))
+	, nesting=NULL
+	, weights="proportional"
+	, plotit = FALSE
+	, ci=FALSE
+)
+unbalanced_age <- (unbalanced_age
+	%>% group_by(model)
+	%>% summarize(fit=mean(fit))
+	%>% mutate(age=mean_age
+		, dev=abs(fit-mean_hhsize)
+	)
+	%>% data.frame()
+)
+unbalanced_age
+```
+
+An alternative is using Eqn 2 with approach 2
+
+<br>
+
+```{r unbalanced_age_levels}
+mean_age <- mean(age_at)
+unbalanced_age_levels <- combinepreds(unbalanced_mod, c("emmeans", "varpred", "Effect")
+	, focal = "age"
 	, x.var = "age"
@@
 	, weights="proportional"
-	, plotit = TRUE
+	, plotit = FALSE
 	, ci=FALSE
 )
-cat("Truth: \n", mean(unbalanced_df$hhsize))
-print(unbalanced_age
-	+ scale_colour_manual(values=cols)
+unbalanced_age_levels <- (unbalanced_age_levels
+	%>% group_by(model)
+	%>% summarize(fit=mean(fit))
+	%>% mutate(age=mean_age
+		, dev=abs(fit-mean_hhsize)
 	)
+	%>% data.frame()
+)
+unbalanced_age_levels
 ```
+
+**The later heavily depends on the choice of the levels of the focal predictor**

### 2021-11-17 R/effectsfuns.R

@@ -280,6 +280,7 @@ varpred <- function(mod
 			, x.var=x.var
 			, typical=typical
 			, zero_out_interaction=zero_out_interaction
+			, include.re=include.re
 		)
 		pred_obj <- pred_obj_all$pred_df
 		predict.data <- pred_obj[, colnames(x.focal), drop=FALSE]

### 2021-11-17 R/utilities.R

@@ -4,14 +4,15 @@
 pop.bias.adjust <- function(x.focal, x.excluded, betahat, formula.rhs
 	, rTerms, factor.levels, contr, offset, mult, vnames, ...
 	, mod, vcov., isolate, isolate.value, internal, vareff_objects, x.var
-	, typical, zero_out_interaction) {
+	, typical, zero_out_interaction, include.re) {
 
 	non_focal_terms <- names(vnames)[!vnames %in% colnames(x.focal)]
 	focal_terms <- names(vnames)[vnames %in% colnames(x.focal)]
 
 	mm <- get_model.mm(mod)
 	col_mean <- colMeans(mm)
-	
+
+	mm_temp <- mm
 	if (isolate) {
 		# FIXME: assuming pop of non-focal -> no variance
 		mm[,non_focal_terms] <- 0
@@ -22,6 +23,12 @@ pop.bias.adjust <- function(x.focal, x.excluded, betahat, formula.rhs
 		nM <- NROW(x.excluded)
 	} else {
 		nM <- NROW(mm)
+	}
+
+	if (include.re) {
+		re <- includeRE(mod)	
+	} else {
+		re <- 0
 	}
 
 	pred_list <- list()
@@ -42,7 +49,7 @@ pop.bias.adjust <- function(x.focal, x.excluded, betahat, formula.rhs
 		
 		off <- get_offset(offset, mf_i)
 		offs[i] <- off
-		pred <- off + as.vector(mm_i %*% betahat)
+		pred <- off + as.vector(mm_i %*% betahat) + re
 		
 		pse_var <- mult*get_sderror(mod=mod
 			, vcov.=vcov.

### 2021-11-19 variable_predictions.Rnw
Population-based bias
Autocommit

@@
 
-Generally, in generalized linear models with other link functions, the predictions are first generated in linear predictor scale, $\hat{\boldeta}^*$, and then transformed to the original response scale using $g^{-1}(\hat{\boldeta}^*)$.
+\subsection{Higher order interactions}
+
+Higher order terms such as interactions, splines, polynomials, etc., can be between the non-focal predictors or focal and non-focal predictor(s).
+In the former case, we treat the interactions as just another column in the variable space (of the model matrix).
+In the later case, the non-focal variables in the model matrix are averaged as before.
+However, for the interacting focal predictors, a combination of each unique levels (or quantiles) are first generated and then the interaction terms are generated by multiplying these combinations.
+For example, consider the model~\ref{eq:simple_inter_higher} below:
+%
+\begin{align}
+y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \beta_{23}x_2x_3 + \epsilon \label{eq:simple_inter_higher}
+\end{align}
+%
+For the first case, if we consider $x_1$ as the focal predictor, the interaction ($x_2x_3$) is between the non-focal predictors, $x_2$ and $x_3$, the linear predictor is 
+%
+\begin{align*}
+\eta^\star(x_{1i}, \nset{{\bar{x}^\star}}) = \beta_0\mathbf{1} + \beta_1 x_{1i} + \beta_2\bar{x}_2 + \beta_3\bar{x}_3 + \beta_{23}\bar{x}_2\bar{x}_3
+\end{align*}
+%
+where $x_{1i}$ are the carefully chosen levels of the focal predictor.
+On the other hand, for the second case, if we consider $x_2$ as the focal predictor, then the interaction ($x_2x_3$) is between focal and non-focal predictor.
+In this case, the linear predictor is given by
+%
+\begin{align*}
+\eta^\star(x_{2ij}, \nset{{\bar{x}^\star}}) = \beta_0\mathbf{1} + \beta_1 \bar{x}_1 + \beta_2x_{2ij} + \beta_3x_{3j} + \beta_{23}x_{2ij}x_{3j}
+\end{align*}
+%
+In general, our formulation, even for more complicated interactions, follow these two basic principles -- interaction between non-focal predictors and interaction between focal and non-focal predictors.
 
@@
 This approach incorporates all the uncertainties -- including the uncertainty due to non-focal predictors.
-But what if we are only interested in the uncertainty as a result of the focal predictor, so that the confidence intervals are $\eta \pm q \sigma_f$.
+But what if we are only interested in the uncertainty as a result of the focal predictor, so that the confidence intervals are $\eta \pm q \sigma_f$, i.e., \emph{isolated} confidence intervals?
 
@@
 
-\subsection{Handling higher order terms}
+\subsubsection{Numeric interactions}
 
-Higher order terms such as interactions, splines, polynomials, etc., can be between the non-focal predictors or focal and non-focal predictor(s).
-In the former case, we treat the interactions as just another column in the variable space (of the model matrix).
-\textbf{Still figuring out the later case but currently treating everything as just additional column???}
+Consider a multiple linear regression model such that
 

### 2021-11-21 variable_predictions.Rnw
Trying to write

@@
 
-\title{Uncertainty propagation for variable (predictor) effects and bias correction in generalized linear (mixed) models???}
+\title{Uncertainty propagation for predictor effects and bias correction in generalized linear (mixed) models}
 
@@
 In many applications, predictions of expected responses, or response probabilities, are often of interest.
-Predictions provide a great way to summarize what the regression model is telling us and are very useful for interpreting and visualizing models estimates using graphs.
+Predictions provide a great way to summarize what the regression model is telling us and are very useful for interpreting and visualizing model estimates using graphs.
 For example, in logistic regression models, the coefficient estimates are usually not easy to interpret, and we may want to explore the ``effects" of covariates on predicted probabilities, or predictions of responses at various levels of covariates -- \emph{predictor effects}.
@@
 \item choice of representative values of \emph{focal} predictor(s) and appropriate ``model center" for \emph{non-focal} predictors especially in multivariate models
-\item  propagation of uncertainty ([how] can we incorporate uncertainty in nonlinear components of (G)LMMs? Should we exclude variation in non-focal parameters?)
-\item bias in the expected mean prediction induced by the nonlinear transformation of the response variable (especially in GL(M)M)
+\item  propagation of uncertainty ([how] can we incorporate uncertainty in nonlinear components of (G)L(M)Ms? Should we exclude variation in non-focal parameters?)
+\item bias in the expected mean prediction induced by the nonlinear transformation of the response variable (especially in GL(M)Ms)
 \end{enumerate}
@@
 
-The outline of this article is as follows: Section \ref{sec:qoi} provides formal definition various quantities of interest, Section \ref{sec:stats_background} provides statistical background of estimation of model predictions, Section \ref{sec:uncertainity_propagation} describes mathematical and computation implementation of the proposed method for uncertainty propagation, and lastly Section \ref{sec:bias_correction} describes methods for bias correction in GL(M)M together with the computational implementation.
+The outline of this article is as follows: Section \ref{sec:qoi} provides formal definition various quantities of interest, Section \ref{sec:stats_background} provides statistical background of estimation of model predictions, Section \ref{sec:uncertainity_propagation} describes mathematical and computation implementation of the proposed method for uncertainty propagation, Section \ref{sec:bias_correction} describes methods for bias correction in GL(M)M together with the computational implementation, and lastly, in Section \ref{sec:simulation_example} we consider some simulation examples.
 
@@
 We first describe this variance-covariance based approach and then discuss our proposed method which is based on \emph{model centering} and does not require input predictors to be scaled prior to model fitting.
-Model center is found by averaging the variables over the model matrix, $\bX^\star$.
 
@@
 The computation of $\hat{\eta}^\star$ remains the same as described above.
-However, to compute $\sigma$, $\Sigma$ is modified by \emph{zeroing-out} (the variance-covariance of all non-focal predictors are assigned zero) variances of non-focal terms.
-Although this is the simplest approach, it requires centering of continuous predictors prior to model fitting and proper way to average categorical predictors.
-One way to center predictors is to create a design matrix, $\bX$, and then compute $\bX_{c} = \bX - \bar{\bX}$.
+However, to compute $\sigma$, $\Sigma$ is modified by \emph{zeroing-out} (the variance-covariance of all non-focal predictors are set to zero) variances of non-focal terms.
+Although this is the simplest approach, it requires centering of continuous, i.e., $x_c = x - \bar{x}$, predictors prior to model fitting and proper way to average categorical predictors.
 
@@
 Consider centered model matrix $\bX^{\star}_{c} = \{\bX_f^\star, \nset{{\bX}^\star} - \nset{{\bar{\bX}}^\star}\}$.
-It follows that the non-focal terms in $\bX^{\star}_{c}$ are all zero.
-Consequently the uncertainty due to non-focal predictors are all zeroed-out in the computation of $\sigma$.
-More generally, centered design matrix, $\bX^{\star}_{c}$, impacts on the estimated value of the intercept and its associated variance but not the slopes.
-Thus since non-focal terms in $\bX^{\star}_{c}$ are all zero, it does not matter what their corresponding values are in the variance-covariance matrix.
-Hence, we can compute CIs from non-centered predictors (in other words, fitted models with predictors in their natural scales).
+It follows that the non-focal terms in $\bX^{\star}_{c}$ are all zero in simple models without interactions but are isolated (narrower) around the model center.
+Consequently the uncertainty due to non-focal predictors are isolated in the computation of $\sigma^2 = \textrm{Diag}(\bX^\star_c \Sigma \bX^{\star\top}_c)$.
+In addition, the computation of $\bX^{\star}_c$ impacts only on the intercepts and the non-focal terms, i.e., the slopes and variance of the focal predictors are not affected.
+This means that we can still generate isolated CIs without necessarily centering the predictors prior to model fitting.
 
-\subsubsection{Numeric interactions}
+\section{Bias correction}\label{sec:bias_correction}
 
-Consider a multiple linear regression model such that
+In many applications, it usually important to report the estimates that reflect the expected values of the untransformed response.
+However, when dealing with nonlinear link functions, it is even harder to generate correct predictions that reflect the untransformed response due to the bias in the expected mean induced by the nonlinear transformation of the response variable.
+In such cases, bias correction is needed when back-transforming the predictions to the original scales.
+Most common approach for bias-adjustment is second-order Taylor approximation \citep{lenth2018package, duursma2003bias}.
+Here, we describe and implement a different approach, \emph{population-based} approach for bias correction.
 
-\subsection{Simulation examples}
+\subsection{Population-based approach for bias correction}
+
+The most precise (although not necessarily accurate!) way to predict is to condition on a value $F$ of the focal predictor and make predictions for all observations (members of the population) for which $x_f = F$ (or in a small range around $F$ ...).
+A key point is that the nonlinear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population (e.g.
+all individuals with $x_f = F$) to a one-dimensional distribution of $\eta^\star$.
+
+Once we have got our vector of $\eta^\star$ (which is essentially a set of samples from the distribution over $\eta^\star(x_f, \nset{x})$ for the conditional set), we may want a mean and confidence intervals on the mean on the response (data) scale, i.e.
+after back-transforming.
+In other words we can use the observations themselves then we just compute the individual values of $g^{-1}(\eta^\star)$ and compute their mean.
+
+In population-based approach, $\eta^\star$ is constructed as a function of properly constituted levels of focal predictors, $x_f$, and entire population of the non-focal predictors, $\nset{x}$, as opposed to averaged non-focal predictors, $\nset{{\bar{x}}}$, described in the previous section.
+More specifically:
+%
+\begin{itemize}
+\item compute linear predictor associated with the non-focal predictors, $\nset{\eta} = \sum \nset{\beta} \nset{x}$
+\item compute linear predictor associated with the focal predictors, $\eta_{jf} = \sum{\beta_f x_{jf}}$
+\item for every value of the focal predictor, $\eta_{jf}$:
+\begin{itemize}
+\item $\eta_j^\star  = \eta_{jf} + \nset{\eta}$
+\item $\hat{y}_j  = \textrm{mean} ~ g^{-1} \left(\eta_j^\star\right)$
+\end{itemize}
+\end{itemize}
+%
+We make similar adjustments to compute the variances of the predictions at every level of the focal predictor:
+%
+\begin{align}
+\sigma_{jf}^2 = \textrm{Diag}(\bX^\star_{jc} \Sigma \bX^{\star\top}_{jc})
+\end{align}
+%
+where $\bX^{\star}_{jc} = \{\bX_{jf}^\star, \nset{{\bX}^\star} - \nset{{\bar{\bX}}^\star}\}$ and 
+%
+\begin{align}
+\mathrm{CI}_j = \mathrm{mean} ~ \eta_j^\star \pm q\sigma_{jf}
+\end{align}
+%
+
+\section{Simulation examples}\label{sec:simulation_example}
 
@@
 
-\subsubsection{Continuous predictors}\label{sec:continuous_predictors}
+\subsection{Continuous predictors}\label{sec:continuous_predictors}
 
@@
 
-\subsubsection{Categorical predictors}\label{sec:categorical_predictors}
+\subsection{Categorical predictors}\label{sec:categorical_predictors}
 
@@
 
-\subsubsection{Linear mixed effect model}
+\subsection{Linear mixed effect model}
 
@@
 
-\subsubsection{Multiple outcomes}
+\subsection{Multiple outcomes}
 
@@
 
-\section{Bias correction}\label{sec:bias_correction}
-
 \clearpage

### 2021-11-23 variable_predictions.Rnw
Manuscript...?

@@
 
+\author{Bicko Cygu, Ben Bolker, Jonathan Dushoff}
+
 \begin{document}
@@
 load("variable_predictions_objs.rda")
-load("categorical_predictors.rda")
-load("lme_random_intercept.rda")
-load("multiple_outcomes_preds.rda")
 load("cubic_predictors_preds.rda")
+load("cubic_predictors_preds_adjust.rda")
+load("glme_random_intercept_preds.rda")
 @
@@
 \epsilon_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
-\beta_0 &= 1.5 \nonumber\\
+\beta_0 &= 5 \nonumber\\
 \beta_{\mathrm{A}} &= 0.1 \nonumber\\
@@
 
-We first fitted all the three models and in each case, compared three different predictor effects together with the corresponding CIs -- 1) uncertainty due to non-focal predictors included (everything), 2) uncertainty due to non-focal predictors removed using variance-covariance matrix (zero-vcov), and 3) uncertainty due to non-focal predictors removed using centered model matrix (centered mm).
+We first fitted all the three models and in each case, compared three different predictor effects together with the corresponding CIs -- 1) uncertainty due to non-focal predictors included (everything), 2) uncertainty due to non-focal predictors removed using variance-covariance matrix (isolated (vcov)), and 3) uncertainty due to non-focal predictors removed using centered model matrix (isolated (mm)).
 The results are shown in Figure~\ref{fig:pred_age_cont_plots}.
@@
 Figure~\ref{fig:pred_age_cont_plots}a represents predictor effects for the model with no interaction, Figure~\ref{fig:pred_age_cont_plots}b shows predictor effects for the model with interaction between non-focal predictors, and Figure~\ref{fig:pred_age_cont_plots}c shows predictor effects for the model with interaction between focal and non-focal predictors.
-For perfect predictions, we expect the yellow and grey dotted lines to overlay each other and they should intersect with the black solid line and the vertical dotted grey line at the same point (or very close).
+For perfect predictions, we expect the horizontal yellow and grey dotted lines to overlay each other and they should intersect with the black solid line and the vertical dotted grey line at the same point (or very close), i.e., model center.
 In other words, at the model center (mean of the focal predictor), generally, we expect the model to predict the average of the response variable and narrower CIs.
-If we properly anchor the model at its center (or any other appropriate value, assuming no interactions), we would expect the variance at the anchor to be zero, i.e., the CIs bands crosses at that point (see zero-cov and centered mm in Figure~~\ref{fig:pred_age_cont_plots}).
+If we properly anchor the model at its center (or any other appropriate value, assuming no interactions), we would expect the variance at the anchor to be zero, i.e., the CIs bands crosses at that point (see isolated (mm) and isolated (vcov) in Figure~~\ref{fig:pred_age_cont_plots}).
 Although the overall predictor effects trend lines are different in each of the simulation, the expected household size for a household head with average age is not only the same in all the three simulations but also very close to the ``truth''.
@@
 The vertical grey line is the mean of the focal predictor (model center).
-The solid black, red (non-focal uncertainty excluded) and blue lines (all uncertainty included) are the predictor effects while their corresponding CIs are the dotted black, red and blue lines, respectively.
-For properly generated predictor effects, we expect the yellow and the grey horizontal dotted lines to overlay each other and intersect with the black, blue, red solid trend lines, and the vertical dotted grey line at the same point (model center).}
+The dotted black, red (non-focal uncertainty excluded) and the solid blue lines (all uncertainty included) are the predictor effects together with their corresponding CIs.
+For properly generated predictor effects or simple models with no interactions, we expect the yellow and the grey horizontal dotted lines to overlay each other and intersect with the black, blue, red solid trend lines, and the vertical dotted grey line at the same point (model center).}
 \label{fig:pred_age_cont_plots}
@@
 \mathrm{hh~size} &= \beta_0 + \beta_{\mathrm{A_1}}\mathrm{Age}_i + \beta_{\mathrm{A_2}}\mathrm{Age}^2_i + \beta_{\mathrm{A_3}}\mathrm{Age}^3_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \epsilon_i \nonumber\\
-\mathrm{Age}_i &\sim \mathrm{Uniform}(-1, 1) \nonumber\\
+\mathrm{Age}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
 \mathrm{Wealthindex}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
-\epsilon_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
-\beta_0 &= 1.5 \nonumber\\
-\beta_{\mathrm{A}} &= 0.3 \nonumber\\
-\beta_{\mathrm{W}} &= 2 \nonumber\\
+\epsilon_i &\sim \mathrm{Normal}(0, 10) \nonumber\\
+\beta_0 &= 20 \nonumber\\
+\beta_{\mathrm{A}_1} &= 0.1 \nonumber\\
+\beta_{\mathrm{A}_2} &= 0.8 \nonumber\\
+\beta_{\mathrm{A}_3} &= 0.3 \nonumber\\
+\beta_{\mathrm{W}} &= 0.8 \nonumber\\
 i &= 1,\cdots, 100
@@
 
+In Figure~\ref{fig:pred_cubic_plots} we compare the non-isolated (everything) and the isolated predictor effects.
+In the first case (Figure~\ref{fig:pred_cubic_plots}a), the focal predictor, \code{age}, has a higher order interaction in the form of cubic polynomial while in the second case (Figure~\ref{fig:pred_cubic_plots}b), the focal predictor, \code{wealthindex}, is simple but the non-focal predictor is a cubic polynomial.
+In Figure~\ref{fig:pred_cubic_plots}a, the average prediction and the ``truth'' are very similar but in this case, cubic polynomial, the model center is not a point in the predictor space.
+On the hand, in Figure~\ref{fig:pred_cubic_plots}b, the average prediction is higher than the observed marginal, ``truth'' but the model center is a point in the predictor space.
 %
@@
 \caption{A comparison of predictor effects and their corresponding CIs.
-Figure a) shows model with no interaction, Figure b) shows model with non-focal predictors interacting with each other, and Figure c) shows model with focal and non-focal predictors interacting.
+In Figure a) the focal predictor has a higher cubic polynomial interaction while in Figure b) the focal predictor has no interaction but the non-focal predictor is a cubic polynomial.
 The dotted yellow and grey horizontal lines are the expected and observed/true means, respectively.
 The vertical grey line is the mean of the focal predictor (model center).
-The solid black, red (non-focal uncertainty excluded) and blue lines (all uncertainty included) are the predictor effects while their corresponding CIs are the dotted black, red and blue lines, respectively.
-For properly generated predictor effects, we expect the yellow and the grey horizontal dotted lines to overlay each other and intersect with the black, blue, red solid trend lines, and the vertical dotted grey line at the same point (model center).}
+The dotted black (non-focal uncertainty excluded) and blue lines (all uncertainty included) are the predictor effects together with their corresponding CIs.}
 \label{fig:pred_cubic_plots}
@@
 
-\subsection{Categorical predictors}\label{sec:categorical_predictors}
-
-Suppose, in Section~\ref{sec:continuous_predictors}, in addition to the predictors already defined, we add gender of the household head.
-For example, we may want to know whether men (M) and females (F) differ in the relationship between age of the household head and the size of the household -- how does households with male as their heads differ in size as compare to those with female heads?
-
-The first model in this case, slightly modifies the first model in Section~\ref{sec:continuous_predictors} by adding gender of the household head.
-The proportion of households with males as their household heads is $60\%$:
-%
-\begin{align}\label{sim:lm_no_interaction_cat}
-\mathrm{hh~size} &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{M}}{\mathrm{Gender}_i} + \epsilon_i \nonumber\\
-\mathrm{Age}_i &\sim \mathrm{Normal}(0.2, 1) \nonumber\\
-\mathrm{Gender}_i &\sim \mathrm{Sample}(\mathrm{Gender_i}, p) \nonumber\\
-p &= \begin{cases}
-   0.4 & \mathrm{Gender_{i=Female}} \nonumber\\
-   0.6 & \mathrm{Gender_{i=Male}} \nonumber\\
-  \end{cases} \nonumber\\
-\epsilon_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
-\beta_0 &= 1.5 \nonumber\\
-\beta_{\mathrm{A}} &= 0.8 \nonumber\\
-\beta_{\mathrm{M}} &= 0.2 \nonumber\\
-i &= 1,\cdots, 100
-\end{align}
-%
-The second model has interaction between non-focal predictors (\code{age} and \code{wealth index}), while the third model has interaction between focal (\code{gender} and \code{age}).
-The results are shown in Figure~\ref{fig:pred_gender_cat_plots}, which compares the predictors effects (either centered or all uncertainty included -- everything) for no interaction model (a), interaction between non-focal predictors (b) and interaction between focal and non-focal predictor (c).
-For each gender, we expect the observed average household size (``truth" -- red dots), to be very close to the estimated predictor effects (blue and black dots).
+Even in simple interaction model in Figure~\ref{fig:pred_age_cont_plots}b and Figure~\ref{fig:pred_cubic_plots}b, we start to observe bias in the predicted and observed means.
+To illustrate the bias correction, we regenerate the predictor effects in Figure~\ref{fig:pred_cubic_plots}b but incorporate bias adjustment.
 %
@@
 \begin{center}
-<<pred_gender_cat_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
-print(pred_gender_cat_plots)
+<<pred_cubic_plots_adjust, echo=FALSE, fig=TRUE, message=FALSE>>=
+print(pred_cubic_plots_adjust)
 @
 \end{center}
-\caption{A comparison of predictor effects and their corresponding CIs.
-Figure a) shows model with no interaction, Figure b) shows model with non-focal predictors interacting with each other, and Figure c) shows model with focal and non-focal predictors interacting.
-The red dots are the observed observed/true means.
-The black (non-focal uncertainty excluded) and blue solid dots (all uncertainty included) are the predictor effects with their corresponding CIs, respectively.
-As espected, the observed average household size (red) is very similar to the predicted (red and blue).}
-\label{fig:pred_gender_cat_plots}
+\caption{A comparison of bias corrected and uncorrected predictor effects.
+The red solid and dotted lines are not bias corrected (none), the black ones are corrected for bias, the horizontal grey and yellow lines are the observed and predicted average household size, respectively, while the vertical dotted line is the average wealth index.
+In the absence of bias, the two approaches give similar estimates (Figure a).
+On the other hand, in Figure b, without correcting for the bias, the predictions are slightly higher than the corrected ones which matches the observed relatively well.}
+\label{fig:pred_cubic_plots_adjust}
 \end{figure}
@@
 
-\subsection{Linear mixed effect model}
+\subsection{Generalized linear mixed effect model}
 
-We now consider linear models with additional random effects components.
+Here we consider generalized linear models with random effects components.
 In the simplest case, we start with a single grouping factor as the random effect component, i.e., random intercept model.
-This way, we can investigate the contribution of random effects overall expected mean.
+This way, we can investigate the contribution of random effects overall predicted probability.
 
-Suppose, in the simulation described in Section~\ref{sec:stats_background}, the observations are  recorded at least once per household.
-In particular, let $\mathrm{H}$ be the number of households indexed by the grouping factor, and $\mathrm{h}[i]$ be the household of the $i$th observation.
-We simulate three different models incorporating this kind of formulation -- first, one grouping factor-one predictor-linear random effect model:
+Suppose, in the simulation described in Section~\ref{sec:stats_background}, the observations are  recorded at least once per household and we are interested in the status improved water services (improved or unimproved).
+In particular, let $\mathrm{H}$ be the number of households indexed by the grouping factor, and $\mathrm{h}[i]$ be the household of the $i$th observation such that
 %
 \begin{align}\label{sim:lme_one_pred}
-\mathrm{hh~size} &= \beta_0 + \alpha_{0,\mathrm{h}[i]} + \beta_{\mathrm{A}}\mathrm{Age}_i + \epsilon_i \nonumber\\
-\alpha_{0,\mathrm{h}} &\sim \mathrm{Normal}(0, 1),~~ h=1, \cdots, \mathrm{10} \nonumber\\
-\mathrm{Age}_i &\sim \mathrm{Normal}(0.2, 1) \nonumber\\
-\epsilon_i &\sim \mathrm{Normal}(0, 2) \nonumber\\
-\beta_0 &= 1.5 \nonumber\\
-\beta_{\mathrm{A}} &= 0.1 \nonumber\\
-i &= 1,\cdots, 50
+\mathrm{logit(status} = 1) &= \eta \nonumber\\
+\mathrm{\eta} &= \beta_0 + \alpha_{0,\mathrm{h}[i]} + \beta_{\mathrm{A}}\mathrm{Age}_i + \epsilon_i \nonumber\\
+\alpha_{0,\mathrm{h}} &\sim \mathrm{Normal}(0, 5),~~ h=1, \cdots, \mathrm{10} \nonumber\\
+\mathrm{Age}_i &\sim \mathrm{Normal}(0, 3) \nonumber\\
+\epsilon_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
+\beta_0 &= 5 \nonumber\\
+\beta_{\mathrm{A}} &= 0.8 \nonumber\\
+i &= 1,\cdots, 500
 \end{align}
 %
-The second and third models add the random effects, $\alpha_{\mathrm{h}[i]}$, to models~\ref{sim:lm_nf_interaction} and~\ref{sim:lm_f_interaction}, respectively.
-Figre~\ref{fig:pred_age_cont_lme_plots} shows and compares predictor efffects for the three models.
+We compare predictor effects generated without any bias correction to those corrected for the bias, and then compare the average predicted probability to the observed.
+The results are shown in Figure~\ref{fig:glme_plots}.
+The uncorrected approach overestimated the predicted probabilities as opposed to the bias corrected estimates which closely match the observations and the observed average probability.
 %
@@
 \begin{center}
-<<pred_age_cont_lme_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
-print(pred_age_cont_lme_plots)
+<<glme_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
+print(glme_plots)
 @
 \end{center}
-\caption{A comparison of predictor effects and their corresponding CIs.
-Figure a) shows model with no interaction, Figure b) shows model with non-focal predictors interacting with each other, and Figure c) shows model with focal and non-focal predictors interacting.
-The dotted yellow and grey horizontal lines are the expected and observed/true means, respectively.
-The vertical grey line is the mean of the focal predictor (model center).
-The solid black, red (non-focal uncertainty excluded) and blue lines (all uncertainty included) are the predictor effects while their corresponding CIs are the dotted black, red and blue lines, respectively.
-For properly generated predictor effects, we expect the yellow and the grey horizontal dotted lines to overlay each other and intersect with the black, blue, red solid trend lines, and the vertical dotted grey line at the same point (model center).}
-\label{fig:pred_age_cont_lme_plots}
-\end{figure}
-%
-
-\subsection{Multiple outcomes}
-
-We now consider a bit more complicated model involving more two outcome variables (services) -- household size and rental price.
-In particular, we simulate an intercept only multivariate random effect model to, \emph{jointly}, predict household size and rental price based on age and wealth index.
-Each household has its own correlated effect on the intercepts (random-intercept).
-In addition, we add latent variable to account for unobserved correlations.
-%
-\begin{align}\label{sim:lme_multi_outcome}
-\mathrm{y}_{\mathrm{h}[i]\mathrm{s}} &= \beta_{0,s} + \alpha_{0,\mathrm{h}[i]\mathrm{s}} + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \beta_{\mathrm{Ls}}\mathrm{Latent}_{i\mathrm{s}} + \epsilon_{i\mathrm{s}} \nonumber\\
-\alpha_{0,\mathrm{h}\mathrm{s}} &\sim \mathrm{Normal}(0, 2),~~ h=1, \cdots, \mathrm{H} \nonumber\\
-\mathrm{Age}_i &\sim \mathrm{Normal}(0.2, 1) \nonumber\\
-\mathrm{Wealthindex}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
-\mathrm{Latent}_{i\mathrm{s}} &\sim \mathrm{Normal}(0, 1) \nonumber\\
-\epsilon_{i\mathrm{s}} &\sim \mathrm{Normal}(0, 1) \nonumber\\
-\beta_0 &= 1.5 \nonumber\\
-\beta_{\mathrm{A}} &= 0.1 \nonumber\\
-\beta_{\mathrm{W}} &= 2 \nonumber\\
-\beta_{\mathrm{L}1} &= 1 \nonumber\\
-\beta_{\mathrm{L}2} &= -5 \nonumber\\
-i &= 1,\cdots, 100
-\end{align}
-%
-Where:
-\begin{itemize}
-\item $\beta_{0,s}$ \textbf{is the service specific effect???}, i.e., baseline household size and rental price
-\item $\alpha_{0,\mathrm{h}\mathrm{s}}$ household specific random intercepts, i.e., captures the offset from the overall baseline (mean)
-\end{itemize}
-
-Predictor effects for the above model are show in Figure~\ref{fig:pred_cont_joint_plots}.
-The top plot compares centered and model centered predictor effects with the observed average household size and rental price.
-The point estimates (black and blue dots) are the expected baseline household size and the rental price.
-The two bottom plots compares the predictor effect (of age) on the predicted household size and rental price.
-Again, in the case of joint model, at the model center, the expected predictions are very similar to the observed marginal averages for the outcomes.
-%
-\begin{figure}[H]
-\begin{center}
-<<pred_cont_joint_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
-print(pred_cont_joint_plots)
-@
-\end{center}
-\caption{Predictor effect of age on the predicted household size and rental price.
-The top plots shows the baseline (mean) predicted household size and rental price.
-The bottom plots compares the predictor effect of age on the predicted household size and rental price.
-The dotted yellow and grey horizontal lines are the expected and observed/true means, respectively.
-The vertical grey line is the mean of the focal predictor (model center).
-The solid black, red (non-focal uncertainty excluded) and blue lines (all uncertainty included) are the predictor effects while their corresponding CIs are the dotted black, red and blue lines, respectively.
-For properly generated predictor effects, we expect the yellow and the grey horizontal dotted lines to overlay each other and intersect with the black, blue, red solid trend lines, and the vertical dotted grey line at the same point (model center).
-Without removing the variations due to non-focal predictors, the confidence bands are very wide ("everything" in the figure).
-However, when CIs are centered, we are only focussing on the uncertainty due to the focal predictor.}
-\label{fig:pred_cont_joint_plots}
+\caption{A compare of bias corrected and uncorrected predictor effects.
+The red lines are the bias uncorrected estimates while the black ones are the bias corrected.
+The horizontal red and black lines are average predicted probability of improved water.
+The horizontal grey (overlaid by the black) dotted line is the observed average probability of improved water.
+The vertical dotted grey line is the average age while the grey points are the binned observations.}
+\label{fig:glme_plots}
 \end{figure}

### 2021-11-23 R/plotsfuns.R

@@ -68,13 +68,25 @@ plot.vareffects <- function(x, ..., xlabs = NULL, ylabs = NULL, pos = 0.5, ci = 
 	}
 	p1 <- p1 + labs(x = xlabs, y = ylabs)
 	if (class(df[[x.var]]) %in% c("numeric", "integer")) {
-		p2 <- p1 + geom_line()
+		if (length(nn)>1) {
+			p2 <- p1 + geom_line(aes(linetype=model))
+		} else {
+			p2 <- p1 + geom_line()
+		}
 		if (ci) {
-			p2 <- (p2
-				+ geom_line(aes(y = lwr), lty = 2)
-				+ geom_line(aes(y = upr), lty = 2)
-				+ scale_colour_viridis_d(option = "plasma")
-			)
+#			if (length(nn)>1) {
+#				p2 <- (p2
+#					+ geom_line(aes(y = lwr))
+#					+ geom_line(aes(y = upr))
+#					+ scale_colour_viridis_d(option = "plasma")
+#				)
+#			} else {
+				p2 <- (p2
+					+ geom_line(aes(y = lwr), lty=2)
+					+ geom_line(aes(y = upr), lty=2)
+					+ scale_colour_viridis_d(option = "plasma")
+				)
+#			}
 		}
 	} else {
 		if (ci) {

### 2021-11-23 cubic_predictors_preds_adjust.R

@@ -14,15 +14,16 @@ makeGraphics()
 ### All uncertainties 
 pred_age_trad_cubic <- varpred(mod_cubic
 	, "age"
-	, isolate=FALSE
-	, modelname="everything"
+	, isolate=TRUE
+	, modelname="none"
 )
 
 ### Centered predictions
 pred_age_centered_cubic <- varpred(mod_cubic
 	, "age"
 	, isolate=TRUE
-	, modelname="isolated (mm)"
+	, bias.adjust="population"
+	, modelname="bias corrected"
 )
 pred_age_mean <- mean(pred_age_centered_cubic$preds$fit)
 
@@ -34,12 +35,13 @@ pred_age_cubic_plots <- (comparevarpred(vlist=vlist
 		, plotit=TRUE
 		, addmarginals=FALSE
 	)
-	+ geom_hline(data=true_prop_df, aes(yintercept=hhsize), colour="grey", lty=2)
+	+ geom_hline(data=true_prop_df, aes(yintercept=hhsize), colour="grey", lty=1)
 	+ geom_hline(yintercept=pred_age_mean, colour="yellow", lty=2)
 #	+ geom_vline(aes(xintercept=mean(age)), colour="grey", lty=2)
-	+ scale_colour_manual(breaks = c("everything", "isolated (mm)")
-		, values=c(everything="blue", "isolated (mm)"="black")
+	+ scale_colour_manual(breaks = c("none", "bias corrected")
+		, values=c("none"="red", "bias corrected"="black")
 	)
+	+ scale_linetype_manual(values=c("none"="dotted", "bias corrected"="solid"))
 	+ labs(title="a) Focal polynomial", y="Predictions", linetype="Method", colour="Method")
 	+ theme(legend.position="bottom")
 )
@@ -50,15 +52,16 @@ print(pred_age_cubic_plots)
 ### All uncertainties 
 pred_wealthindex_trad_cubic <- varpred(mod_cubic
 	, "wealthindex"
-	, isolate=FALSE
-	, modelname="everything"
+	, isolate=TRUE
+	, modelname="none"
 )
 
 ### Centered predictions
 pred_wealthindex_centered_cubic <- varpred(mod_cubic
 	, "wealthindex"
 	, isolate=TRUE
-	, modelname="isolated (mm)"
+	, bias.adjust="population"
+	, modelname="bias corrected"
 )
[…]

### 2021-12-05 variable_predictions.Rnw
Figure 2 analytical soln

@@
 \newcommand{\xfprime}{x_{f'}}
+
+\newcommand{\yE}{{\mathrm E}}
 
@@
 %
+We may be interested in comparing the observed average household size ($\yE(\mathrm{hh~size})$) and the predicted average household size ($\yE(\widehat{\mathrm{hh~size}})$). 
+%
+\begin{align}\label{sim:lm_cubic_mean}
+\yE(\mathrm{hh~size}) &\approx \beta_0 + \beta_{\mathrm{A_1}}\yE(\mathrm{Age}_i) + \beta_{\mathrm{A_2}}\yE(\mathrm{Age}^2_i) + \beta_{\mathrm{A_3}}\yE(\mathrm{Age}^3_i) \nonumber\\ 
+	& + \beta_{\mathrm{W}}\yE(\mathrm{Wealthindex}_i)
+\end{align}
+%
+
+We provide analytical comparison in two cases, i.e., when the interaction is in the focal predictor ($\mathrm{Age}$ is the focal predictor)
+%
+\begin{align}\label{sim:lm_cubic_analytic_age}
+\eta^\star(\mathrm{Age}_i, \nset{{\overline{\mathrm{Wealthindex}}}}) &= \beta_0 + \beta_{\mathrm{A_1}}\mathrm{Age}_i + \beta_{\mathrm{A_2}}\mathrm{Age}^2_i + \beta_{\mathrm{A_3}}\mathrm{Age}^3_i \nonumber\\
+	& + \beta_{\mathrm{W}}\overline{\mathrm{Wealthindex}} \nonumber\\
+\yE({\eta^\star(\mathrm{Age}_i, \nset{{\overline{\mathrm{Wealthindex}}}})}) &= \beta_0 + \beta_{\mathrm{A_1}}\yE({\mathrm{Age}_i}) + \beta_{\mathrm{A_2}}\yE({\mathrm{Age}^2_i}) + \beta_{\mathrm{A_3}}\yE({\mathrm{Age}^3_i}) \nonumber\\
+	& + \beta_{\mathrm{W}}\overline{\mathrm{Wealthindex}}
+\end{align}
+%
+and when the interaction is in the non-focal predictor ($\mathrm{Wealthindex}$ is the focal predictor)
+%
+\begin{align}\label{sim:lm_cubic_analytic_wealthindex}
+\eta^\star(\mathrm{Wealthindex}_i, \nset{{\{\overline{\mathrm{Age}}, \overline{\mathrm{Age}}^2, \overline{\mathrm{Age}}^3\}}}) &= \beta_0 + \beta_{\mathrm{A_1}}\overline{\mathrm{Age}} + \beta_{\mathrm{A_2}}\overline{\mathrm{Age}}^2 + \beta_{\mathrm{A_3}}\overline{\mathrm{Age}}^3 \nonumber\\
+	& + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i \nonumber\\
+\yE(\eta^\star(\mathrm{Wealthindex}_i, \nset{{\{\overline{\mathrm{Age}}, \overline{\mathrm{Age}}^2, \overline{\mathrm{Age}}^3\}}})) &= \beta_0 + \beta_{\mathrm{A_1}}\overline{\mathrm{Age}} + \beta_{\mathrm{A_2}}\overline{\mathrm{Age}}^2 + \beta_{\mathrm{A_3}}\overline{\mathrm{Age}}^3 \nonumber\\
+	& + \beta_{\mathrm{W}}\yE(\mathrm{Wealthindex}_i)
+\end{align}
+%
+From Equation~\ref{sim:lm_cubic_mean} and ~\ref{sim:lm_cubic_analytic_age}
+%
+\begin{align*}
+\yE(\mathrm{hh~size}) &\approx \yE({\eta^\star(\mathrm{Age}_i, \nset{{\overline{\mathrm{Wealthindex}}}})})
+\end{align*}
+%
+since $\overline{\mathrm{Wealthindex}} = \yE(\mathrm{Wealthindex}_i)$. On the other hand
+%
+\begin{align*}
+\yE(\mathrm{hh~size}) &\approx \yE(\eta^\star(\mathrm{Wealthindex}_i, \nset{{\{\overline{\mathrm{Age}}, \overline{\mathrm{Age}}^2, \overline{\mathrm{Age}}^3\}}}))
+\end{align*}
+%
+if and only if $\yE(\mathrm{Age^2}) = \yE(\mathrm{Age}) \yE(\mathrm{Age})$ and $\yE(\mathrm{Age^3}) = \yE(\mathrm{Age}) \yE(\mathrm{Age}) \yE(\mathrm{Age})$, but, without loss of generality, for example
+%
+\begin{align}
+\yE(\mathrm{Age^2}) &= \yE(\mathrm{Age}) \yE(\mathrm{Age}) + \sigma^2_{\mathrm{Age}} \nonumber\\
+&= \overline{\mathrm{Age}}^2 + \sigma^2_{\mathrm{Age}}
+\end{align}
+%
 

### 2021-12-13 variable_predictions.Rnw
Autocommit

--- 2021-11-23 variable_predictions.Rnw
@@
 \newcommand{\xfprime}{x_{f'}}
+
+\newcommand{\yE}{{\mathrm E}}
+\let\over=\overline
 
@@
 %
+We may be interested in comparing the observed average household size ($\over{\mathrm{hh~size}}$) and the predicted average household size ($\over{\widehat{\mathrm{hh~size}}}$). 
+%
+\begin{align}\label{sim:lm_cubic_mean}
+\over{\mathrm{hh~size}} &\approx \beta_0 + \beta_{\mathrm{A_1}}\over{\mathrm{Age}_i} + \beta_{\mathrm{A_2}}\over{\mathrm{Age}^2_i} + \beta_{\mathrm{A_3}}\over{\mathrm{Age}^3_i} \nonumber\\ 
+	& + \beta_{\mathrm{W}}\over{\mathrm{Wealthindex}_i}
+\end{align}
+%
+
+We provide analytical comparison in two cases, i.e., when the interaction is in the focal predictor ($\mathrm{Age}$ is the focal predictor)
+%
+\begin{align}\label{sim:lm_cubic_analytic_age}
+\eta^\star(\mathrm{Age}_i, \nset{{\over{\mathrm{Wealthindex}}}}) &= \beta_0 + \beta_{\mathrm{A_1}}\mathrm{Age}_i + \beta_{\mathrm{A_2}}\mathrm{Age}^2_i + \beta_{\mathrm{A_3}}\mathrm{Age}^3_i \nonumber\\
+	& + \beta_{\mathrm{W}}\over{\mathrm{Wealthindex}} \nonumber\\
+\over{{\eta^\star(\mathrm{Age}_i, \nset{{\over{\mathrm{Wealthindex}}}})}} &= \beta_0 + \beta_{\mathrm{A_1}}\over{{\mathrm{Age}_i}} + \beta_{\mathrm{A_2}}\over{{\mathrm{Age}^2_i}} + \beta_{\mathrm{A_3}}\over{{\mathrm{Age}^3_i}} \nonumber\\
+	& + \beta_{\mathrm{W}}\over{\mathrm{Wealthindex}}
+\end{align}
+%
+and when the interaction is in the non-focal predictor ($\mathrm{Wealthindex}$ is the focal predictor)
+%
+\begin{align}\label{sim:lm_cubic_analytic_wealthindex}
+\eta^\star(\mathrm{Wealthindex}_i, \nset{{\{\over{\mathrm{Age}}, \over{\mathrm{Age}^2}, \over{\mathrm{Age}^2}\}}}) &= \beta_0 + \beta_{\mathrm{A_1}}\over{\mathrm{Age}} + \beta_{\mathrm{A_2}}\over{\mathrm{Age}^2} + \beta_{\mathrm{A_3}}\over{\mathrm{Age}^3} \nonumber\\
+	& + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i \nonumber\\
+\yE(\eta^\star(\mathrm{Wealthindex}_i, \nset{{\{\over{\mathrm{Age}}, \over{\mathrm{Age}^2}, \over{\mathrm{Age}^3}\}}})) &= \beta_0 + \beta_{\mathrm{A_1}}\over{\mathrm{Age}} + \beta_{\mathrm{A_2}}\over{\mathrm{Age}^2} + \beta_{\mathrm{A_3}}\over{\mathrm{Age}^3} \nonumber\\
+	& + \beta_{\mathrm{W}}\over{\mathrm{Wealthindex}_i}
+\end{align}
+%
+Depending on the choice of the values of the focal predictors in each case, Equation~\ref{sim:lm_cubic_analytic_age} and \ref{sim:lm_cubic_analytic_wealthindex} approximates \ref{sim:lm_cubic_mean} i.e., mean household size at the model center.
+In Equation~\ref{sim:lm_cubic_analytic_age}, the model center is not a point in the predictor space but rather described by the terms of the cubic polynomial, as opposed to \ref{sim:lm_cubic_analytic_wealthindex} which has the model center as point.
 

### 2021-12-13 R/effectsfuns.R

@@ -310,8 +310,9 @@ varpred <- function(mod
 			pred_obj <- pred_obj_all$pred_df
 			predict.data <- x.focal[rep(1:NROW(x.focal), each=pred_obj_all$dim), 1:n.focal, drop=FALSE]
 			pse_var <- pred_obj$pse_var
+			off <- pred_obj_all$off
 		} else if (bias.adjust=="population2") {
-			pred_obj_all <- pop2.bias.adjust(x.focal=x.focal
+			pred_obj <- pop2.bias.adjust(x.focal=x.focal
 				, x.excluded=x.excluded
 				, betahat=betahat
 				, formula.rhs=formula.rhs
@@ -332,14 +333,15 @@ varpred <- function(mod
 				, zero_out_interaction=zero_out_interaction
 				, include.re=include.re
 			)
-			pred_obj <- pred_obj_all$pred_df
-			predict.data <- x.focal[rep(1:NROW(x.focal), each=NROW(x.excluded)), 1:n.focal, drop=FALSE]
-			pse_var <- unname(pred_obj_all$pse_var)[rep(1:NROW(x.focal), each=NROW(x.excluded))]
+#			pred_obj <- pred_obj_all$pred_df
+			predict.data <- x.focal[rep(1:NROW(x.focal), each=NROW(X.mod)), 1:n.focal, drop=FALSE]
+#			pse_var <- unname(pred_obj_all$pse_var)[rep(1:NROW(x.focal), each=NROW(x.excluded))]
+			pse_var <- pred_obj$pse_var
+			off <- pred_obj$off
 		}
 		pred <- pred_obj$pred
 		lwr <- pred_obj$lwr
 		upr <- pred_obj$upr
-		off <- pred_obj_all$off
 		mm <- NULL
 	}
 	

### 2021-12-13 cubic_predictors_preds.R

@@ -100,6 +100,8 @@ compare_cubic_plots <- (combinepreds(mod_cubic, focal="wealthindex"
 		, plotit=TRUE
 	)
 	+ geom_vline(aes(xintercept=mean(focal_levels)), lty=2, colour="grey")
+	+ geom_hline(yintercept=pred_wealthindex_mean, colour="yellow", lty=1, alpha=0.5)
+	+ labs(colour="Method", linetype="Method")
 )
 print(compare_cubic_plots)
 

### 2021-12-13 variable_predictions_funs.R

@@ -329,7 +329,7 @@ combinepreds <- function(mod, funs, focal, x.var, x.var.factor=FALSE, plotit=TRU
 				%>% data.frame()
 			)
 			print(pred_prop_df)
-			out <- p1 + geom_hline(data=pred_prop_df, aes(yintercept=fit, colour=model))
+			out <- p1 + geom_hline(data=pred_prop_df, aes(yintercept=fit, colour=model, linetype=model))
 		} else {
 			out <- p1
 		}

### 2021-12-15 variable_predictions.Rnw
Autocommit

@@
 load("cubic_predictors_preds_adjust.rda")
+load("glm_one_predictor_preds.rda")
 load("glme_random_intercept_preds.rda")
+load("binom_multiple_outcomes_preds_plots.rda")
 @
@@
 \begin{align*}
-\eta^\star(x_{1i}, \nset{{\bar{x}^\star}}) = \beta_0\mathbf{1} + \beta_1 x_{1i} + \beta_2\bar{x}_2 + \beta_3\bar{x}_3 + \beta_{23}\bar{x}_2\bar{x}_3
+\eta^\star(x_{1i}, \nset{{\bar{x}^\star}}) = \beta_0\mathbf{1} + \beta_1 x_{1i} + \beta_2\bar{x}_2 + \beta_3\bar{x}_3 + \beta_{23}\over{x_2x_3}
 \end{align*}
@@
 In the first case (Figure~\ref{fig:pred_cubic_plots}a), the focal predictor, \code{age}, has a higher order interaction in the form of cubic polynomial while in the second case (Figure~\ref{fig:pred_cubic_plots}b), the focal predictor, \code{wealthindex}, is simple but the non-focal predictor is a cubic polynomial.
-In Figure~\ref{fig:pred_cubic_plots}a, the average prediction and the ``truth'' are very similar but in this case, cubic polynomial, the model center is not a point in the predictor space.
-On the hand, in Figure~\ref{fig:pred_cubic_plots}b, the average prediction is higher than the observed marginal, ``truth'' but the model center is a point in the predictor space.
+In both cases, the average prediction and the ``truth'' are very similar but in Figure~\ref{fig:pred_cubic_plots}a, cubic polynomial focal predictor, the model center is not a point in the predictor space.
+On the hand, in Figure~\ref{fig:pred_cubic_plots}b, the model center is a point in the predictor space.
 %
@@
 
-Even in simple interaction model in Figure~\ref{fig:pred_age_cont_plots}b and Figure~\ref{fig:pred_cubic_plots}b, we start to observe bias in the predicted and observed means.
-To illustrate the bias correction, we regenerate the predictor effects in Figure~\ref{fig:pred_cubic_plots}b but incorporate bias adjustment.
+\subsection{Generalized linear model}
+
+Consider model described in \ref{sim:lm_no_interaction} with only \code{Age} as the predictor such that 
+%
+\begin{align}\label{sim:glm_one_pred}
+\mathrm{logit(status} = 1) &= \eta \nonumber\\
+\mathrm{\eta} &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i + \epsilon_i
+\end{align}
+%
+where \code{status} is a binary outcome with 0 for unimproved and 1 for improved water quality in the slums.
+By having only a single predictor, we may be able to investigate the contribution of nonlinear link function on the predicted probability of improved water quality.
+
+Figure~\ref{fig:glm_plots} shows the predictor effect of \code{age} on predicted probability of improved water quality.
+For such a simple generalized linear model, there is no much bias induced by the nonlinear averaging induced by the nonlinear link function.
+In particular, the predicted average probability of improved water quality, horizontal black line, closely match the observed proportion of households with improved water quality, horizontal solid black line.
 %
@@
 \begin{center}
-<<pred_cubic_plots_adjust, echo=FALSE, fig=TRUE, message=FALSE>>=
-print(pred_cubic_plots_adjust)
+<<glm_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
+print(glm_plots)
 @
 \end{center}
-\caption{A comparison of bias corrected and uncorrected predictor effects.
-The red solid and dotted lines are not bias corrected (none), the black ones are corrected for bias, the horizontal grey and yellow lines are the observed and predicted average household size, respectively, while the vertical dotted line is the average wealth index.
-In the absence of bias, the two approaches give similar estimates (Figure a).
-On the other hand, in Figure b, without correcting for the bias, the predictions are slightly higher than the corrected ones which matches the observed relatively well.}
-\label{fig:pred_cubic_plots_adjust}
+\caption{Predictor effect of age on improved water quality.
+The horizontal black and grey lines are the average predicted probability and observed proportion of improved water quality, respectively.
+The vertical dotted grey line is the mean age while the red points are the observed bines of proportion of improved water quality.}
+\label{fig:glm_plots}
 \end{figure}
@@
 The results are shown in Figure~\ref{fig:glme_plots}.
-The uncorrected approach overestimated the predicted probabilities as opposed to the bias corrected estimates which closely match the observations and the observed average probability.
+The uncorrected approach overestimates the predicted probabilities as opposed to the bias corrected estimates which closely match the observations and the observed average probability.
 %
@@
 \end{center}
-\caption{A compare of bias corrected and uncorrected predictor effects.
+\caption{A comparison of bias corrected and uncorrected predictor effects.
 The red lines are the bias uncorrected estimates while the black ones are the bias corrected.
@@
 \label{fig:glme_plots}
+\end{figure}
+%
+
+\subsubsection{Multivariate binary outcomes}
+
+We now consider a bit more complicated model involving two outcome variables (services) -- status (improved or unimproved) of water and garbage services.
+In particular, we simulate an intercept only multivariate random effect model to, \emph{jointly}, predict probability of improved water and garbage services based on age and wealth index.
+Each household has its own correlated effect on the intercepts (random-intercept).
+In addition, we add latent variable to account for unobserved correlations.
+%
+\begin{align}\label{sim:lme_multi_outcome}
+\mathrm{services}_{{\mathrm{h}[i]\mathrm{s}}} &= \mathrm{Binomial}(P_{{\mathrm{h}[i]\mathrm{s}}}) \nonumber\\
+\mathrm{logit}(P_{{\mathrm{h}[i]\mathrm{s}}}) &= \beta_{0,s} + \alpha_{0,\mathrm{h}[i]\mathrm{s}} + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \beta_{\mathrm{Ls}}\mathrm{Latent}_{i\mathrm{s}} + \epsilon_{i\mathrm{s}} \nonumber\\
+\alpha_{0,\mathrm{h}\mathrm{s}} &\sim \mathrm{Normal}(0, 2),~~ h=1, \cdots, \mathrm{H} \nonumber\\
+\mathrm{Age}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
+\mathrm{Wealthindex}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
+\mathrm{Latent}_{[i]\mathrm{s}} &\sim \mathrm{Normal}(0, 1) \nonumber\\
+\epsilon_{[i]\mathrm{\text{\tiny{water}}}} &\sim \mathrm{Normal}(0, 2) \nonumber\\
+\epsilon_{[i]\mathrm{\text{\tiny{garbage}}}} &\sim \mathrm{Normal}(0, 3) \nonumber\\
+\beta_0 &= 5 \nonumber\\
+\beta_{\mathrm{A}} &= 2 \nonumber\\
+\beta_{\mathrm{W}} &= -3 \nonumber\\
+\beta_{\mathrm{L\text{\tiny{water}}}} &= 1 \nonumber\\
+\beta_{\mathrm{L\text{\tiny{garbage}}}} &= -5 \nonumber\\
+i &= 1,\cdots, 100
+\end{align}
+%
+Where:
+\begin{itemize}
+\item $\beta_{0,s}$ \textbf{is the service specific effect???}, i.e., baseline log odds of a household having improved service
+\item $\alpha_{0,\mathrm{h}\mathrm{s}}$ household specific random intercepts, i.e., captures the offset from the overall baseline (mean)
+\end{itemize}
+
+%
+\begin{figure}[H]
+\begin{center}
+<<binom_outcomes_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
+print(binom_outcomes_plots)
+@
+\end{center}
+\caption{TODO}
+\label{fig:binom_outcomes_plots}
 \end{figure}

### 2021-12-15 R/utilities.R

@@ -460,7 +460,7 @@ clean_model <- function(focal.predictors, mod, xlevels
 	 	factor.cols[grep(paste0(":", name, "|", name, ":"), cnames)] <- FALSE
 	 }
 	 if (!ff_check) {
-	 	factor.cols[grep(paste0(":", name, "|", name, ":"), cnames)] <- FALSE
+	 	factor.cols[grep(paste0(":", name, "|", name, ":"), cnames)] <- TRUE # FALSE for product of means as opposed to mean of products
       ## FIXME: Best way to identify all polynomial/function terms in a model
 		factor.cols[grep(paste("^[aA-zZ]+\\(", name, sep=""), cnames)] <- TRUE
 	 } 

### 2021-12-15 binom_multiple_outcomes.R

@@ -25,16 +25,16 @@ perHH <- 10
 #### Simulation
 sim_df_cont_joint <- linearsim(nHH=nHH_obs
 	, perHH=perHH
-	, form=~1+x1+x2+x3
-	, betas = c(5, 2, -3, 1)
+	, form=~1+x1+x2
+	, betas = c(5, 2, -3)
 	, hhSD = c(2, 3)
-	, pgausian=list(p=3,fun=rnorm, mean=0, sd=1)
+	, pgausian=list(p=2,fun=rnorm, mean=0, sd=1)
 	, pcat=list(p=0)
 	, noutcomes=2
 	, separatelatent=TRUE
 	, link_scale=FALSE
 	, blatent=c(1, -5)
-	, vnames=c("age", "wealthindex", "latent", "hhsize", "rent")
+	, vnames=c("age", "wealthindex", "water", "garbage")
 )$data
 head(sim_df_cont_joint)
 
@@ -45,7 +45,7 @@ head(sim_df_cont_joint)
 
 sim_df_cont_joint_long <- (sim_df_cont_joint
 	%>% mutate(iid=1:n())
-	%>% gather(services, values, c("hhsize", "rent"))
+	%>% gather(services, values, c("water", "garbage"))
 	%>% data.frame()
 )
 head(sim_df_cont_joint_long)

### 2021-12-15 glm_one_predictor_preds.R

@@ -1,0 +1,46 @@
+library(shellpipes)
+library(vareffects); varefftheme()
+library(ggpubr)
+library(ggplot2)
+library(dplyr)
+library(emmeans)
+
+loadEnvironments()
+startGraphics()
+
+## No bias adjustment
+glm_pred_none <- varpred(glm_mod
+	, "age"
+	, isolate=TRUE
+)
+
+## Bins 
+binned_df <- binfun(glm_mod, focal="age", bins=50, groups=NULL)
+
+glm_plots <- (plot(glm_pred_none)
+	+ geom_point(data=binned_df, aes(x=age, y=status), colour="red")
+	+ geom_hline(data=true_prop_df, aes(yintercept=status), lty=2, col="red")
+	+ geom_hline(aes(yintercept=mean(fit)), lty=1, col="black")
+	+ geom_vline(aes(xintercept=mean(age)), lty=2, col="grey")
+	+ labs(y="Probability of improved water quality", colour="Method", linetype="Method")
+)
+
+print(glm_plots)
+
+## Compare with emmeans
+age_at <- quantile(glm_sim_df$age, seq(0,1,length.out=100))
+glm_varpred_emmeans_plots <- combinepreds(glm_mod, c("emmeans", "varpred")
+	, focal = "age"
+	, x.var = "age"
+	, at=list(age=age_at)
+	, xlevel=list(age=age_at)
+	, nesting=NULL
+	, type="response"
+	, weights="proportional"
+	, plotit = TRUE
+	, ci=FALSE
+)
+glm_varpred_emmeans_plots
+
+saveVars(glm_plots)
+

### 2021-12-16 variable_predictions.Rnw
Meeting tomorrow

@@
 load("variable_predictions_objs.rda")
+load("justify_preds.rda")
 load("cubic_predictors_preds.rda")
@@
 Formerly, predictor effects computes the expected outcome by meaningfully holding the non-focal predictors constant (or averaged in some meaningful way) while varying the focal predictor, with the goal that the outcome expected prediction represents how the model responds to the changes in the focal predictor.
+
+The go to \proglang{R} packages (\pkg{emmeans} and \pkg{effects}) for generating predictor effects on the basic level, averages the non-focal predictors.
+However, there are a number of choices which needs to be made when averaging the non-focal predictors.
+For example, in the presence of interaction, averaging the interactions (averaging product of interacting predictors) or multiplying the averages of the interacting non-focal predictors (default for \pkg{emmeans} and \pkg{effects}).
+We claim that neither of these two approaches is the most appropriate but we show that averaging the interaction closely matches the observed ``truth''.
+To illustrate this, consider models~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher}  below, with $x_1$ as the focal predictor:
+%
+\begin{align}
+y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \epsilon \label{eq:simple_inter_higher_no_interaction}\\
+y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \beta_{23}x_2x_3 + \epsilon \label{eq:simple_inter_higher}
+\end{align}
+%
+We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -0.3$, $\beta_2 = 1$, $\beta_3 = 2$ and $\beta_{23} = 0.8$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the known ``truth'' i.e., $\bar{y}$.
+
+In the absence of interaction (model~\ref{eq:simple_inter_higher_no_interaction}), the three approaches give the similar estimates, which match the observed ``truth'', Figure~\ref{fig:justify_plots}a.
+However, in the presence of interaction, even as simple as the one in model~\ref{eq:simple_inter_higher}, the estimates starts to differ.
+In particular, \pkg{emmeans} and \pkg{effects} give similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is similar to the observed truth ($\bar{y}$), Figure~\ref{fig:justify_plots}a.
+To generate Figure~\ref{fig:justify_plots}a, all the three packages averages the non-focal predictors $x_2$ and $x_3$.
+On other hand, to generate Figure~\ref{fig:justify_plots}b, the difference in the estimates lies on how each of the packages average the interaction term ($x_2x_3$).
+In particular \pkg{emmeans} and \pkg{effects} computes $\bar{x_2}\bar{x_3}$ while \pkg{varpred} computes $\over{x_2x_3}$.
+%
+\begin{figure}[H]
+\begin{center}
+<<justify_plots, echo=FALSE, results=tex, fig=TRUE, message=FALSE>>=
+justify_plots
+@
+\end{center}
+\caption{A comparison of predictor effect of $x_1$ on predicted $y$ from \pkg{emmeans}, \pkg{effects} and \pkg{varpred} to the observed.
+The horizontal blue, green and black lines are the mean of the predictions, i.e., $\bar{\hat{y}}$ while the red one is the mean of the observed $y$, i.e., $\bar{y}$.
+The trend lines represents the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other predictors at their mean.
+In the absence of interaction, Figure a, $\bar{\hat{y}}$ closely matches the observed in all the three packages.
+However, even with the simple interaction between the non-focal predictors, we start seeing larger deviation of $\bar{hat{y}}$ from $\bar{y}$ for the two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not, the proposed \pkg{varpred}.
+}
+\label{fig:justify_plots}
+\end{figure}
+%
 
@@
 However, for the interacting focal predictors, a combination of each unique levels (or quantiles) are first generated and then the interaction terms are generated by multiplying these combinations.
-For example, consider the model~\ref{eq:simple_inter_higher} below:
-%
-\begin{align}
-y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \beta_{23}x_2x_3 + \epsilon \label{eq:simple_inter_higher}
-\end{align}
-%
-For the first case, if we consider $x_1$ as the focal predictor, the interaction ($x_2x_3$) is between the non-focal predictors, $x_2$ and $x_3$, the linear predictor is 
+For example, consider model~\ref{eq:simple_inter_higher} above.
+In the first case, if we consider $x_1$ as the focal predictor, the interaction ($x_2x_3$) is between the non-focal predictors, $x_2$ and $x_3$, the linear predictor is
 %
@@
 The results are shown in Figure~\ref{fig:glme_plots}.
-The uncorrected approach overestimates the predicted probabilities as opposed to the bias corrected estimates which closely match the observations and the observed average probability.
+The uncorrected approach overestimates the predicted probabilities as opposed to the bias corrected (population-based approach) estimates which closely match the observations and the observed average probability.
+In the latter approach, the random effects are incorporated (as additional non-focal predictor) in computation of $\nset{\eta}$ to take into account the bias induced by the random effect predictor.
 %
@@
 
+Figure~\ref{fig:binom_outcomes_plots} compares bias uncorrected and bias corrected (population-based approach).
+The bias corrected estimates (black points and curves) closely match the observed proportions (grey points).
+The top figure shows the expected probability of improved service for an ``average'' household at the baseline.
+While the two bottom plots shows the predictor effect of age and wealth index on the predicted probability of improved service.
 %
@@
 \end{center}
-\caption{TODO}
+\caption{Predictor effect plot for multivariate response model.
+The top plot shows the predicted probability of improved service at the baseline.
+The bottom plots compares the predictor effect of age and wealth index on the expected probability of improved service.
+The grey points are the observed proportions (or bins in bottom plots) of improved service.
+The red points and curves are the bias uncorrected estimates while the black points and curves are the population-based (bias corrected) estimates.
+Generally, without taking into account the potential bias, we risk overestimating the predicted probabilities.}
 \label{fig:binom_outcomes_plots}

### 2021-12-16 justify_preds.R

@@ -1,0 +1,73 @@
+library(shellpipes)
+library(vareffects); varefftheme()
+library(ggplot2)
+library(dplyr)
+library(ggpubr)
+library(emmeans)
+library(effects)
+
+loadEnvironments()
+startGraphics()
+
+## No interactions
+quants <- seq(0,1,length.out=100)
+x1_focal <- quantile(justify_sim_df$x1, quants, names=FALSE)
+
+justify_plots <- (combinepreds(justify_mod
+		, c("varpred", "emmeans", "Effect")
+		, focal="x1"
+		, x.var="x1"
+		, at=list(x1=x1_focal)
+		, type="response"
+		, nesting=NULL
+		, ci=FALSE
+	)
+	+ geom_hline(data=true_prop_df, aes(yintercept=y, colour="truth"), lty=4)
+	+ scale_colour_manual(breaks = c("truth", "emmeans", "Effect", "varpred")
+		, values=c("truth"="red", "emmeans"="blue", "Effect"="green", "varpred"="black")
+		, labels=c("truth", "emmeans", "effects", "varpred")
+	)
+	+ scale_linetype_manual(values=c("truth"=4, "emmeans"=2, "Effect"=3, "varpred"=1)
+		, labels=c("truth", "emmeans", "effects", "varpred")
+	)
+	+ labs(title="a) Model 1", colour="Method", linetype="Method")
+	+ theme(legend.position="bottom")
+)
+justify_plots
+
+## With interactions
+quants <- seq(0,1,length.out=100)
+x1_focal <- quantile(justify_inter_sim_df$x1, quants, names=FALSE)
+
+justify_inter_plots <- (combinepreds(justify_inter_mod
+		, c("varpred", "emmeans", "Effect")
+		, focal="x1"
+		, x.var="x1"
+		, at=list(x1=x1_focal)
+		, type="response"
+		, nesting=NULL
+		, ci=FALSE
+	)
+	+ geom_hline(data=true_prop_inter_df, aes(yintercept=y, colour="truth"), lty=4)
+	+ scale_colour_manual(breaks = c("truth", "emmeans", "Effect", "varpred")
+		, values=c("truth"="red", "emmeans"="blue", "Effect"="green", "varpred"="black")
+		, labels=c("truth", "emmeans", "effects", "varpred")
+	)
+	+ scale_linetype_manual(values=c("truth"=4, "emmeans"=2, "Effect"=3, "varpred"=1)
[…]

### 2021-12-16 variable_predictions_funs.R

@@ -276,17 +276,15 @@ combinepreds <- function(mod, funs, focal, x.var, x.var.factor=FALSE, plotit=TRU
 	)
 	add_args <- list(...)
 	if (length(add_args)) args[names(add_args)] <- add_args
+	
+	if (!is.null(add_args$at) & ("Effect" %in% funs)) args$xlevels <- add_args$at
 
 	out <- sapply(funs, function(f){
 		est <- do.call(f, args)
 		if (inherits(est, c("emmeans", "emmGrid"))) {
-			type <- est@misc$predict.type
+			.nn <- est@misc$inv.lbl
+			if (is.null(.nn)) .nn <- est@misc$estName
 			est <- as.data.frame(est)
-			if (length(type) && type=="response") {
-				.nn <- "prob"
-			} else {
-				.nn <- "emmean"
-			}
 			oldn <- c(focal, .nn
 				, grep("\\.CL|\\.LCL|\\.UCL", colnames(est), value=TRUE)
 			)

### 2021-12-17 variable_predictions.Rnw
Autocommit

@@
 \begin{enumerate}
-\item choice of representative values of \emph{focal} predictor(s) and appropriate ``model center" for \emph{non-focal} predictors especially in multivariate models
+\item choice of representative values of \emph{focal} predictor(s) and appropriate \emph{model center} for \emph{non-focal} predictors especially in multivariate models
 \item  propagation of uncertainty ([how] can we incorporate uncertainty in nonlinear components of (G)L(M)Ms? Should we exclude variation in non-focal parameters?)
@@
 
-The go to \proglang{R} packages (\pkg{emmeans} and \pkg{effects}) for generating predictor effects on the basic level, averages the non-focal predictors.
+The go to \proglang{R} packages (\pkg{emmeans} and \pkg{effects}) for generating predictor effects, by default, averages the non-focal predictors.
 However, there are a number of choices which needs to be made when averaging the non-focal predictors.
-For example, in the presence of interaction, averaging the interactions (averaging product of interacting predictors) or multiplying the averages of the interacting non-focal predictors (default for \pkg{emmeans} and \pkg{effects}).
+For example, in the presence of interaction, averaging the interactions (averaging product of interacting predictors) versus product of the averages of the interacting predictors (default for \pkg{emmeans} and \pkg{effects}).
 We claim that neither of these two approaches is the most appropriate but we show that averaging the interaction closely matches the observed ``truth''.
@@
 %
-We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -0.3$, $\beta_2 = 1$, $\beta_3 = 2$ and $\beta_{23} = 0.8$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the known ``truth'' i.e., $\bar{y}$.
+We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -0.5$, $\beta_2 = 1$, $\beta_3 = 2$ and $\beta_{23} = 5$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the known ``truth'' i.e., $\bar{y}$.
 
-In the absence of interaction (model~\ref{eq:simple_inter_higher_no_interaction}), the three approaches give the similar estimates, which match the observed ``truth'', Figure~\ref{fig:justify_plots}a.
+In the absence of interaction (model~\ref{eq:simple_inter_higher_no_interaction}), the three approaches have similar estimates, which match the observed ``truth'', Figure~\ref{fig:justify_plots}a.
 However, in the presence of interaction, even as simple as the one in model~\ref{eq:simple_inter_higher}, the estimates starts to differ.
-In particular, \pkg{emmeans} and \pkg{effects} give similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is similar to the observed truth ($\bar{y}$), Figure~\ref{fig:justify_plots}a.
+In particular, \pkg{emmeans} and \pkg{effects} give similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is similar to the observed truth ($\bar{y}$), Figure~\ref{fig:justify_plots}b.
 To generate Figure~\ref{fig:justify_plots}a, all the three packages averages the non-focal predictors $x_2$ and $x_3$.
@@
 In the absence of interaction, Figure a, $\bar{\hat{y}}$ closely matches the observed in all the three packages.
-However, even with the simple interaction between the non-focal predictors, we start seeing larger deviation of $\bar{hat{y}}$ from $\bar{y}$ for the two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not, the proposed \pkg{varpred}.
+However, even with the simple interaction between the non-focal predictors, we start seeing larger deviation of $\bar{\hat{y}}$ from $\bar{y}$ for the two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not, the proposed \pkg{varpred}.
 }
@@
 An alternative to averaging the non-focal predictors, is the population-based approach which involves computing the prediction over the population of the non-focal predictors and then averaging across the values of the focal predictor.
-Dealing with second challenge is discussed in Section~\ref{sec:uncertainity_propagation}
 
-When dealing with nonlinear link functions, the correct predictions for example, are even much harder to estimate.
+In some applications, we may only be interested in the uncertainties associated with the focal predictor -- \emph{isolated} confidence intervals.
+However, currently, mostly used \proglang{R} packages do not provide straightforward way for \emph{isolating} uncertainties associated with the particular focal predictor.
+In addition to prediction lines shown in Figure~\ref{fig:justify_plots}a above, we include the confidence bands associated with the predictions for all three methods as shown in Figure~\ref{fig:justify_ci_plots}.
+For \pkg{emmeans} and \pkg{effects}, the confidence bands are much wider because they include uncertainties associated with the non-focal predictors, but narrower and crosses at the mean of the focal predictor in \pkg{varpred}.
+In other words, with \pkg{varpred}, we are able to generate isolated confidence bands indicating zero uncertainty at the value of the focal predictor we are more certain about, i.e., mean of the focal predictor.
+For simple models, the point where the confidence bands crosses is the model center and it corresponds to typical value we choose, in this case, the mean of the focal predictor.
+%
+\begin{figure}[H]
+\begin{center}
+<<justify_ci_plots, echo=FALSE, results=tex, fig=TRUE, message=FALSE>>=
+justify_ci_plots
+@
+\end{center}
+\caption{A comparison of regular and isolated confidence bands.
+The description of horizontal and trend lines remain the same as in Figure 1a.
+The wider dotted blue curves overlaying the green curves are the regular confidence bands for \pkg{emmeans} and \pkg{effects}, while the narrower black curves crossing at the mean of the focal predictor are the isolated confidence bands for \pkg{varpred}.
+For simple models, the isolated confidence bands crosses at the model center.}
+\label{fig:justify_ci_plots}
+\end{figure}
+%
+
+When dealing with nonlinear link functions, the correct predictions, for example, are even much harder to estimate.
 One approach is to make predictions on the transformed scale (linear predictor scale), and then back-transform to the original scale.
@@
 
-In simple linear models with no interaction terms, the default out for the coefficient estimates in Table~\ref{tab:qoi_coefs} model (1) is simple and directly interpretable as the expected change in household size for a unit change in age.
+In simple linear models with no interaction terms, the default output for the coefficient estimates in Table~\ref{tab:qoi_coefs} model (1) is simple and directly interpretable as the expected change in household size for a unit change in age.
 In particular, a unit change in age is associated with a household that is \Sexpr{round(coef(qoi_mod1)[[2]], 3)} bigger.
@@
 \end{center}
-\caption{Variable effect plot and marginal effect of age of household head on household size. For a linear model with no interaction, the marginal effect is the slope of the predictor effect line.}
+\caption{Predictor effect plot and marginal effect of age of household head on household size. For a linear model with no interaction, the marginal effect is the slope of the predictor effect line.}
 \label{fig:qoi_pred_plot}
@@
 %
-Predictor effect, on the other hand, is the expected household size for a particular age, holding wealth index at its mean.
-In particular, the purpose and goal of a predictor effect seems fairly straightforward; for specified values of (a) focal predictor(s), we want to give a point estimate and confidence intervals for the prediction of the model for a ``typical" (= random sample over the multivariate distribution of non-focal parameters) individual with those values of the predictors.
-
 On the other hand, when model specification involves other kinds of terms such as variable interactions, log and power transformations, etc., the coefficient estimates cannot easily and directly communicate the relationship between the outcome and the independent variable of interest because of the multiple coefficients of the same variable.
@@
 In other words, for models involving interactions terms, looking at a single estimated coefficient in isolation and failing to take into account the estimates involving the interaction terms may lead to inaccurate interpretations \citep{brambor2006understanding, leeper2017interpreting}.
+
+Predictor effect, on the other hand, is the expected household size for a particular age, holding wealth index at its mean.
+In particular, the purpose and goal of a predictor effect seems fairly straightforward; for specified values of (a) focal predictor(s), we want to give a point estimate and confidence intervals for the prediction of the model for a ``typical" (= random sample over the multivariate distribution of non-focal parameters) individual with those values of the predictors.
 
@@
 
-\subsubsection{Multivariate binary outcomes}
+\subsection{Multivariate binary outcomes}
 

### 2021-12-21 variable_predictions.Rnw
Autocommit

@@
 The first one is the coefficient estimates.
-Others are \emph{predictors}, \emph{model matrix}, \emph{predicted values} and \emph{marginal effects}.
-Predictors are model input variables and is associated with one or more \emph{variables} (in polynomials, splines and models with interaction).
+Others are \emph{input variables},  \emph{predictors}, \emph{model matrix}, \emph{predicted values} and \emph{marginal effects}.
+Input variables are the variables that were measured (may be transformed), while predictors are the terms that are entered in the model.
+Hence, predictors encompass the main effects, but also polynomials of input variables and interaction terms (for example in polynomials and splines) \citep{schielzeth2010simple}.
 Model matrix refers to the design matrix whose rows include all combination of variables appearing in the interaction terms, along with the ``typical" values of the focal and non-focal predictors.
@@
 Consider centered model matrix $\bX^{\star}_{c} = \{\bX_f^\star, \nset{{\bX}^\star} - \nset{{\bar{\bX}}^\star}\}$.
-It follows that the non-focal terms in $\bX^{\star}_{c}$ are all zero in simple models without interactions but are isolated (narrower) around the model center.
+It follows that the non-focal terms in $\bX^{\star}_{c}$ are all zero in simple models without interactions but are isolated (narrower) around the model center in models involving complex interactions.
 Consequently the uncertainty due to non-focal predictors are isolated in the computation of $\sigma^2 = \textrm{Diag}(\bX^\star_c \Sigma \bX^{\star\top}_c)$.

### 2021-12-27 R/implemented.R
Autocommit

@@ -1,6 +1,36 @@
 # Some of the implemented methods
 
-## default lm
+## Some models require prep
+prepmod.default <- function(mod, ...) {
+	return(mod)
+}
+
+## brms
+prepmod.brmsfit <- function(mod, ...) {
+	check_terms <- attr(mod, "terms")
+	if (is.null(check_terms)) {
+		bform <- formula(mod)
+		bform <- brms:::update_re_terms(bform, re_formula=NA)
+		bterms <- brms::brmsterms(bform, resp_rhs_all = FALSE)
+		bterms <- attr(model.frame(bterms$allvars, data = mod$data), "terms")
+		mod$terms <- bterms
+	}
+	return(mod)
+}
+
+## default
+vareffobj.default <- function(mod, ...){
+	out <- list()
+	out$coefficients <- coef(mod)
+	out$variance_covariance <- vcov(mod)
+	out$formula <- formula(mod)
+	out$link <- family(mod)
+	out$contrasts <- mod$contrasts
+	class(out) <- "vareffobj"
+	return(out)
+}
+
+## lm
 vareffobj.lm <- function(mod, ...){
 	out <- list()
 	out$coefficients <- coef(mod)
@@ -33,8 +63,40 @@ get_xlevels.lm <- function(mod) {
 }
 
 check_intercept.default <- function(mod, ...) {
-	any(names(coefficients(mod))=="(Intercept)")
+	any(grepl("Intercept", names(coefficients(mod)), ignore.case=TRUE))
 }
+
+## brms
+vareffobj.brmsfit <- function(mod, ...) {
+	out <- list()
+	out$coefficients <- fixef(mod)[,"Estimate"]
+	out$variance_covariance <- vcov(mod)
+	out$formula <- as.formula(brms:::update_re_terms(formula(mod), re_formula=NA))
+	out$link <- family(mod)
+	out$contrasts <- get_contrasts(mod, ...)
+	class(out) <- "vareffobj"
+	return(out)
+}
+
+get_model.mm.brmsfit <- function(mod, ...) {
+	mm <- model.matrix(terms(mod), mod$data)
+	return(mm)
+}
+
+get_contrasts.brmsfit <- function(mod, ...) {
+	contr <- attr(get_model.mm(mod, ...), "contrasts")
+	return(contr)
+}
+
[…]

### 2022-01-05 variable_predictions.Rnw
Autocommit

@@
 \newcommand{\bX}{{\mathbf X}}
+\newcommand{\bZ}{{\mathbf Z}}
 \newcommand{\bbeta}{{\boldsymbol \beta}}
@@
 load("variable_predictions_objs.rda")
+load("mediate_bin_preds.rda")
+load("mediate_preds.rda")
 load("justify_preds.rda")
@@
 \item for every value of the focal predictor, $\eta_{jf}$:
-\begin{itemize}
-\item $\eta_j^\star  = \eta_{jf} + \nset{\eta}$
-\item $\hat{y}_j  = \textrm{mean} ~ g^{-1} \left(\eta_j^\star\right)$
-\end{itemize}
+%
+\begin{align}\label{eq:pop_eta} 
+\eta_j^\star  &= \eta_{jf} + \nset{\eta} \\
+\hat{y}_j  &= \textrm{mean} ~ g^{-1} \left(\eta_j^\star\right)
+\end{align}
+%
 \end{itemize}
@@
 %
+
+For models with random effects components, another of adjustments can be made to correct for bias induced by the random effects terms.
+In the population approach, we treat the random effects terms as additional non-focal predictors and simply make adjustment to Equation~\ref{eq:pop_eta}.
+In particular
+%
+\begin{align}\label{eq:pop_eta_re} 
+\tau &= \bZ b \nonumber \\
+\eta_j^\star  &= \eta_{jf} + \nset{\eta} + \tau_j
+\end{align}
+%
+where $\bZ$ and $b$ are the design matrix and a vector of random effects, respectively.
 
@@
 \begin{align}\label{sim:glm_one_pred}
-\mathrm{logit(status} = 1) &= \eta \nonumber\\
-\mathrm{\eta} &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i + \epsilon_i
+\mathrm{status}_i &\sim \mathrm{Binomial}(1, \mathrm{P(status}_i = 1)) \nonumber\\
+\mathrm{logit(P(status}_i = 1)) &= \eta_i \nonumber\\
+\mathrm{\eta}_i &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i \nonumber\\
+\mathrm{Age}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
+\beta_0 &= 5 \nonumber\\
+\beta_{\mathrm{A}} &= 0.8 \nonumber\\
+i &= 1,\cdots, 500
 \end{align}
@@
 \begin{align}\label{sim:lme_one_pred}
-\mathrm{logit(status} = 1) &= \eta \nonumber\\
-\mathrm{\eta} &= \beta_0 + \alpha_{0,\mathrm{h}[i]} + \beta_{\mathrm{A}}\mathrm{Age}_i + \epsilon_i \nonumber\\
+\mathrm{status}_i &\sim \mathrm{Binomial}(1, \mathrm{P(status}_i = 1)) \nonumber\\
+\mathrm{logit(P(status}_i = 1)) &= \eta_i \nonumber\\
+\mathrm{\eta}_i &= \beta_0 + \alpha_{0,\mathrm{h}[i]} + \beta_{\mathrm{A}}\mathrm{Age}_i \nonumber\\
 \alpha_{0,\mathrm{h}} &\sim \mathrm{Normal}(0, 5),~~ h=1, \cdots, \mathrm{10} \nonumber\\
-\mathrm{Age}_i &\sim \mathrm{Normal}(0, 3) \nonumber\\
-\epsilon_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
+\mathrm{Age}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
 \beta_0 &= 5 \nonumber\\
@@
 \begin{align}\label{sim:lme_multi_outcome}
-\mathrm{services}_{{\mathrm{h}[i]\mathrm{s}}} &= \mathrm{Binomial}(P_{{\mathrm{h}[i]\mathrm{s}}}) \nonumber\\
-\mathrm{logit}(P_{{\mathrm{h}[i]\mathrm{s}}}) &= \beta_{0,s} + \alpha_{0,\mathrm{h}[i]\mathrm{s}} + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \beta_{\mathrm{Ls}}\mathrm{Latent}_{i\mathrm{s}} + \epsilon_{i\mathrm{s}} \nonumber\\
+\mathrm{status}_{{\mathrm{h}[i]\mathrm{s}}} &\sim \mathrm{Binomial}(1, P(\mathrm{status}_{{\mathrm{h}[i]\mathrm{s}}} = 1)) \nonumber\\
+\mathrm{logit}(P(\mathrm{status}_{{\mathrm{h}[i]\mathrm{s}}} = 1)) &= \beta_{0,s} + \alpha_{0,\mathrm{h}[i]\mathrm{s}} + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \beta_{\mathrm{Ls}}\mathrm{Latent}_{i\mathrm{s}} \nonumber\\
 \alpha_{0,\mathrm{h}\mathrm{s}} &\sim \mathrm{Normal}(0, 2),~~ h=1, \cdots, \mathrm{H} \nonumber\\
@@
 \mathrm{Latent}_{[i]\mathrm{s}} &\sim \mathrm{Normal}(0, 1) \nonumber\\
-\epsilon_{[i]\mathrm{\text{\tiny{water}}}} &\sim \mathrm{Normal}(0, 2) \nonumber\\
-\epsilon_{[i]\mathrm{\text{\tiny{garbage}}}} &\sim \mathrm{Normal}(0, 3) \nonumber\\
 \beta_0 &= 5 \nonumber\\
@@
 
+\begin{appendix}
+
+\section{Appendix A: Additional simulations}
+
+\subsection{Mediated variables}
+
+We consider simple indirect mediation example such that $x$ has direct effect on $y$ which in turn has effect on $z$ but $x$ has no direct effect on $y$, i.e., $x \rightarrow y \rightarrow z$.
+In particular:
+%
+\begin{align}
+z &= \beta_0 + \beta_{yz} y + \epsilon_z \nonumber\\
+y &= \beta_{xy} x + \epsilon_y \nonumber\\
+x &\sim \mathrm{Normal(0, 1)} \nonumber\\
+\epsilon_y &\sim \mathrm{Normal(0, 1)} \nonumber\\
+\epsilon_z &\sim \mathrm{Normal(0, 1)} \nonumber\\
+\beta_0 &= 5 \nonumber\\
+\beta_{xy} &= 1 \nonumber\\
+\beta_{yz} &= 1.5
+\end{align}
+
+%
+\begin{figure}[H]
+\begin{center}
+<<pred_mediate_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
+print(pred_mediate_plots)
+@
+\end{center}
+\caption{Predictor effect plot for multivariate response model.
+The top plot shows the predicted probability of improved service at the baseline.
+The bottom plots compares the predictor effect of age and wealth index on the expected probability of improved service.
+The grey points are the observed proportions (or bins in bottom plots) of improved service.
+The red points and curves are the bias uncorrected estimates while the black points and curves are the population-based (bias corrected) estimates.
+Generally, without taking into account the potential bias, we risk overestimating the predicted probabilities.}
+\label{fig:pred_mediate_plots}
+\end{figure}
+%
+
+%
+\begin{figure}[H]
+\begin{center}
+<<pred_mediate_bin_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
+print(pred_mediate_bin_plots)
+@
+\end{center}
+\caption{Predictor effect plot for multivariate response model.
+The top plot shows the predicted probability of improved service at the baseline.
+The bottom plots compares the predictor effect of age and wealth index on the expected probability of improved service.
+The grey points are the observed proportions (or bins in bottom plots) of improved service.
+The red points and curves are the bias uncorrected estimates while the black points and curves are the population-based (bias corrected) estimates.
+Generally, without taking into account the potential bias, we risk overestimating the predicted probabilities.}
+\label{fig:pred_mediate_bin_plots}
+\end{figure}
+%
+
+\end{appendix}
+
 \clearpage

### 2022-01-05 R/implemented.R

@@ -326,3 +326,18 @@ includeRE.stanreg <- function(mod, ...){
 	re <- as.vector(rstanarm:::get_z(mod) %*% as.matrix(ran_eff))
 	return(re)	
 }
+
+includeRE.brmsfit <- function(mod, ...){
+	Z <- prepare_predictions(mod, ...)$dpars$mu$re$Z
+	ran_eff <- ranef(mod)
+	out <- sapply(names(ran_eff), function(x){
+		z <- Z[[x]]
+		reff <- as.data.frame(ran_eff[[x]])
+		reff <-as.list( reff[, grep("Estimate\\.", colnames(reff), value=TRUE)])
+		reff <- as.matrix(do.call("c", reff))
+		re <- as.vector(z %*% reff)
+	}, simplify = FALSE)
+	out <- do.call("cbind", out)
+	out <- rowSums(out)
+	return(out)
+}

### 2022-01-05 mediate.R

@@ -10,15 +10,16 @@ set.seed(991)
 ### x -> y -> z
 
 N <- 1e4
-beta_0 <- 2
-beta_xy <- 0
-beta_xz <- 1	# Set to 0 -> x: no direct of x on z
-beta_yz <- 0.8
+
+b0 <- 5
+bxz <- 0
+bxy <- 1
+byz <- 1.5
 
 df <- data.frame(x=rnorm(N))
 sim_df_mediate <- (df
-	%>% mutate(y = rnorm(N) + beta_xy*x
-		, z = rnorm(N) + beta_xz*x + beta_yz*y + beta_0
+	%>% mutate(y = rnorm(N) + bxy*x
+		, z = b0 + rnorm(N) + bxz*x + bxy*y
 		, zbin = rbinom(n(), 1, plogis(z))
 	)
 )

### 2022-01-05 mediate_bin_preds.R

@@ -8,79 +8,97 @@ loadEnvironments()
 startGraphics()
 
 ## Not mediated
-pred_x_trad_notmediated <- varpred(mod_notmediated_bin
-	, "x"
-	, isolate=FALSE
-	, modelname="everything"
-)
 
-pred_x_centered_notmediated <- varpred(mod_notmediated_bin
+### Not corrected
+pred_notmediated_none <- varpred(mod_notmediated_bin
 	, "x"
-	, isolate=TRUE
-	, modelname="centered mm"
+	, bias.adjust="none"
+	, modelname="none"
 )
-pred_prop_notmed <- mean(pred_x_centered_notmediated$preds$fit)
+pred_notmediated_none_mean <- getmeans(pred_notmediated_none, what="estimate")
+
+### Bias corrected
+pred_notmediated_pop <- varpred(mod_notmediated_bin
+	, "x"
+	, bias.adjust="population"
+	, modelname="bias corrected"
+)
+pred_notmediated_pop_mean <- getmeans(pred_notmediated_pop, what="estimate")
+
+### Binned obs
+binned_df <- binfun(mod_notmediated_bin, focal="x", bins=50, groups=NULL)
 
 ### Combine all predictions
-vlist <- list(pred_x_trad_notmediated, pred_x_centered_notmediated)
+vlist <- list(pred_notmediated_none, pred_notmediated_pop)
 
-pred_x_notmediated_plots <- (comparevarpred(vlist=vlist
+pred_notmediated_plots <- (comparevarpred(vlist=vlist
 		, lnames=NULL
 		, plotit=TRUE
 		, addmarginals=FALSE
+		, ci=FALSE
 	)
-	+ scale_colour_manual(breaks = c("everything", "centered mm")
-		, values=c(everything="blue", "centered mm"="black")
+	+ geom_hline(data=pred_notmediated_none_mean, aes(yintercept=fit, colour=model, lty=model))
+	+ geom_hline(data=pred_notmediated_pop_mean, aes(yintercept=fit, colour=model, lty=model))
+	+ geom_hline(data=observed_df_med, aes(yintercept=zbin, colour="observed", lty="observed"))
+	+ geom_vline(data=observed_df_med, aes(xintercept=x), lty=2, col="grey")
+#	+ geom_point(data=binned_df, aes(x=x, y=zbin), colour="grey")
+	+ scale_colour_manual(breaks = c("observed", "none", "bias corrected")
[…]

### 2022-01-05 mediate_preds.R

@@ -8,79 +8,98 @@ loadEnvironments()
 startGraphics()
 
 ## Not mediated
-pred_x_trad_notmediated <- varpred(mod_notmediated
-	, "x"
-	, isolate=FALSE
-	, modelname="everything"
-)
 
-pred_x_centered_notmediated <- varpred(mod_notmediated
+### Not corrected
+pred_notmediated_none <- varpred(mod_notmediated
 	, "x"
-	, isolate=TRUE
-	, modelname="centered mm"
+	, bias.adjust="none"
+	, modelname="none"
 )
-pred_prop_notmed <- mean(pred_x_centered_notmediated$preds$fit)
+pred_notmediated_none_mean <- getmeans(pred_notmediated_none, what="estimate")
+
+### Bias corrected
+pred_notmediated_pop <- varpred(mod_notmediated
+	, "x"
+	, bias.adjust="population"
+	, modelname="bias corrected"
+)
+pred_notmediated_pop_mean <- getmeans(pred_notmediated_pop, what="estimate")
+
+### Binned obs
+binned_df <- binfun(mod_notmediated, focal="x", bins=50, groups=NULL)
 
 ### Combine all predictions
-vlist <- list(pred_x_trad_notmediated, pred_x_centered_notmediated)
+vlist <- list(pred_notmediated_none, pred_notmediated_pop)
 
-pred_x_notmediated_plots <- (comparevarpred(vlist=vlist
+pred_notmediated_plots <- (comparevarpred(vlist=vlist
 		, lnames=NULL
 		, plotit=TRUE
 		, addmarginals=FALSE
+		, ci=FALSE
 	)
-	+ scale_colour_manual(breaks = c("everything", "centered mm")
-		, values=c(everything="blue", "centered mm"="black")
+	+ geom_hline(data=pred_notmediated_none_mean, aes(yintercept=fit, colour=model, lty=model))
+	+ geom_hline(data=pred_notmediated_pop_mean, aes(yintercept=fit, colour=model, lty=model))
+	+ geom_hline(data=observed_df_med, aes(yintercept=z, colour="observed", lty="observed"))
+	+ geom_vline(data=observed_df_med, aes(xintercept=x), lty=2, col="grey")
+#	+ geom_point(data=binned_df, aes(x=x, y=z), colour="grey")
+	+ scale_colour_manual(breaks = c("observed", "none", "bias corrected")
[…]

### 2022-01-05 variable_predictions.Rnw
For the meeting

@@
 %
-\begin{align}
+\begin{align}\label{sim:simple_mediate}
 z &= \beta_0 + \beta_{yz} y + \epsilon_z \nonumber\\
@@
 \end{align}
-
+%
+We fit two models -- \emph{non-mediated} which models \emph{z} as a function of \emph{x} and the \emph{mediated} which models \emph{z} as a function of \emph{x} and \emph{y}.
+In both cases, we compare predictions without bias correction (none) with those corrected for bias (bias corrected).
+Figure~\ref{fig:pred_mediate_plots} compares the predictor effect of \emph{x} on \emph{z} for the two models.
+For simple linear model, the predicted averages (model center estimates) are similar to the observed average ($\bar{z}$) and none bias corrected and bias corrected predictions are very similar.
 %
@@
 \end{center}
-\caption{Predictor effect plot for multivariate response model.
-The top plot shows the predicted probability of improved service at the baseline.
-The bottom plots compares the predictor effect of age and wealth index on the expected probability of improved service.
-The grey points are the observed proportions (or bins in bottom plots) of improved service.
-The red points and curves are the bias uncorrected estimates while the black points and curves are the population-based (bias corrected) estimates.
-Generally, without taking into account the potential bias, we risk overestimating the predicted probabilities.}
+\caption{The non-mediated models z as a function of x while the mediated models z as a function of both x and y.
+The horizontal dotted red, blue and black lines are the observed, none bias corrected and bias corrected prediction averages, respectively.
+The vertical dotted grey lines are the average of the focal predictor, x.
+The dotted blue and black dotted lines are predictor effects of x on z.}
 \label{fig:pred_mediate_plots}
@@
 
+We extend the model described in \ref{sim:simple_mediate} to generate binary outcome 
+%
+%
+\begin{align}\label{sim:simple_mediate_bin}
+\mathrm{zbin} &\sim \mathrm{Binomial(1, P(zbin = 1))} \nonumber\\
+\mathrm{logit(P(zbin = 1))} &= z
+\end{align}
+%
+and then fit two generalized linear models together with the comparisons similar to the ones described above.
+The results are shown in Figure~\ref{fig:pred_mediate_bin_plots}.
+For non-mediated model, the bias corrected and non-corrected  predictions are very similar and their averages are very similar to the observed average.
+However, for the mediated model, the two approaches give very different estimates but bias corrected predicted average is relatively close to the observed average.
 %
@@
 \end{center}
-\caption{Predictor effect plot for multivariate response model.
-The top plot shows the predicted probability of improved service at the baseline.
-The bottom plots compares the predictor effect of age and wealth index on the expected probability of improved service.
-The grey points are the observed proportions (or bins in bottom plots) of improved service.
-The red points and curves are the bias uncorrected estimates while the black points and curves are the population-based (bias corrected) estimates.
-Generally, without taking into account the potential bias, we risk overestimating the predicted probabilities.}
+\caption{The non-mediated models zbin as a function of x while the mediated models zbin as a function of both x and y.
+The horizontal dotted red, blue and black lines are the observed, none bias corrected and bias corrected prediction averages, respectively.
+The vertical dotted grey lines are the average of the focal predictor, x.
+The dotted blue and black dotted lines are predictor effects of x on zbin.}
 \label{fig:pred_mediate_bin_plots}

### 2022-01-05 glm_one_predictor.R

@@ -11,7 +11,7 @@ set.seed(9991)
 
 ############################################################################################
 # Global variables
-nHH_obs <- 500
+nHH_obs <- 5000
 perHH <- 1
 
 ############################################################################################

### 2022-01-19 manuscript/vareffects_plosone.tex
Adding manuscript

% Template for PLoS
% Version 3.5 March 2018
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
%\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
%\setlength{\headheight}{27.023pt}
%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\today}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}
\newcommand{\fix}{{\bf FIXME}}
\newcommand{\JD}[1]{{\color{blue} \emph{#1}}}
\newcommand{\bmb}[1]{{\color{RubineRed!70!} \emph{#1}}}
\newcommand{\pkg}[1]{\textbf{#1}}

\def\code#1{\texttt{#1}}
\let\proglang=\textsf

\newcommand{\bX}{{\mathbf X}}
\newcommand{\bZ}{{\mathbf Z}}
\newcommand{\bbeta}{{\boldsymbol \beta}}
\newcommand{\boldeta}{{\boldsymbol \eta}}
\newcommand{\boldmu}{{\boldsymbol \mu}}

\newcommand{\nset}[1]{#1_{\{n\}}}
\newcommand{\yref}{y_{\textrm{ref}}}
\newcommand{\cdist}{{D(\nset{x}|x_f)}}
\newcommand{\cdistprime}{{D(\nset{x}|x_{f'})}}
\newcommand{\xfprime}{x_{f'}}

\newcommand{\yE}{{\mathrm E}}
\let\over=\overline

%% END MACROS SECTION

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Describing the curves: uncertainty propagation and bias correction for predictor effects in simple and generalized linear (mixed) models} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Name1 Surname\textsuperscript{1,2\Yinyang},
Name2 Surname\textsuperscript{2\Yinyang},
Name3 Surname\textsuperscript{2,3\textcurrency},
Name4 Surname\textsuperscript{2},
Name5 Surname\textsuperscript{2\ddag},
Name6 Surname\textsuperscript{2\ddag},
Name7 Surname\textsuperscript{1,2,3*},
with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{2} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{3} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
\Yinyang These authors contributed equally to this work.

% Additional Equal Contribution Note
% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
\ddag These authors also contributed equally to this work.

% Current address notes
\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
% \textcurrency b Insert second current address 
% \textcurrency c Insert third current address

% Deceased author note
\dag Deceased

% Group/Consortium Author Note
\textpilcrow Membership list can be found in the Acknowledgments section.

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* correspondingauthor@institute.edu

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}

\fix

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}

\fix

\linenumbers

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}

In many applications, outcome predictions or probabilities, are often of interest. Predictions provide a great way to summarize what the regression model is telling us and very useful for interpreting and visualizing model estimates. For example, in logistic regression models, the coefficient estimates are usually not easy to interpret and less informative. Since logistic models are nonlinear (due to the nonlinear link function), in multivariate models with interactions, the magnitude of the effect of change in the outcome depends on the values of the predictor of interest and other predictors. Hence, the conclusions one can make about the estimated effects greatly depends on how well we choose the values of the other variables.

Both simple and generalized linear (mixed) models (GL(M)Ms) can examine very complex relationships, including nonlinear relationships between response and predictors, interactions between predictors (and via splines for example), and nonlinear transformations via link functions due to their flexibility. This flexibility comes at a cost, for example, complex multivariate models may risk misinterpretation, and miscalculation of quantities of interest. Also, coefficient estimates of models involving nonlinear link functions or interactions lose their direct interpretation \cite{leeper2017interpreting}, meaning that interpretation of derived quantities from these estimates requires some understanding of the specified model. An alternative is to explore the \emph{effects} of predictors on the predictions or probabilities of the outcome at various level of the predictor of interest -- \emph{predictor effects} \cite{fox2009effect, leeper2017package, lenth2018package}. For example, as a way to plan, public health officials may want to know the effects of household income, wealth index, etc, on the predicted probability of having improved water services among slum dwellers.

When visually presented, predictor effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind of transformations on the predictors whose estimates are usually, but not always, a subject to less clarity of interpretation. Further, generating predictor effects together with the associated confidence intervals for regression models has a number of challenges. In particular:
\begin{enumerate}
\item choice of representative values of \emph{focal} predictor(s) and appropriate \emph{model center} for \emph{non-focal} predictors especially in multivariate models
\item  propagation of uncertainty -- can we incorporate uncertainty in nonlinear components of GL(M)Ms? Should we exclude variation in non-focal parameters?
\item bias in the expected mean prediction induced by the nonlinear transformation of the response variable (especially in GL(M)Ms)
\end{enumerate}

The most common way of dealing with the first challenge is taking unique levels of the focal (predictor of interest) predictor if discrete or taking appropriately sized quantiles (or bins) if continuous, and then calculating the predictions while holding non-focal (other predictors other than the focal predictor) predictors at their typical values (e.g., averages) \cite{hanmer2013behind}. This generates -- \emph{predictor effects}\cite{fox2009effect}, \emph{marginal predictions} \cite{leeper2017package} and \emph{estimated marginal means} \cite{lenth2018package}. In this article, we refer to this quantity as predictor effects since it should, for example, tell us what we would expect the presponse to be at a particular value or level of the predictor. Formerly, predictor effects computes the expected outcome by meaningfully holding the non-focal predictors constant (or averaged in some meaningful way) while varying the focal predictor, with the goal that the outcome expected prediction represents how the model responds to the changes in the focal predictor.

The commonly used \proglang{R} software packages (\pkg{emmeans} and \pkg{effects}) for generating predictor effects, by default, averages the non-focal predictors. However, there are a number of choices a number of choices which one has to make when considering this approach -- for example, in the presence of interaction, averaging the interactions (averaging product of interacting predictors) versus product of the averages of the interacting predictors (default for \pkg{emmeans} and \pkg{effects}). We claim that neither of these two approaches is the most appropriate but we show that averaging the interaction closely matches the observed values. To illustrate this, consider models~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher} below, with $x_1$ as the focal predictor:

%
\begin{align}
y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \epsilon \label{eq:simple_inter_higher_no_interaction}\\
y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \beta_{23}x_2x_3 + \epsilon \label{eq:simple_inter_higher}
\end{align}
%

We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -0.5$, $\beta_2 = 1$, $\beta_3 = 2$ and $\beta_{23} = 5$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the observed average i.e., $\bar{y}$. Fig~\ref{fig:justify_plots} compares the predictors effects of the two models.

\begin{figure}[!h]
\centering
\includegraphics[width=0.8\textwidth]{justify_plots-figure1.pdf}
\caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} predictor effects for $x_1$ on $y$ for models with and without interactions.}
The horizontal blue, green and black lines are the mean predictions, i.e., $\bar{\hat{y}}$, the red ones are the mean observed $y$, i.e., $\bar{y}$, while the vertical dotted grey lines are the means of the corresponding focal predictors. The trend lines represents the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other predictors at their average. A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the observed in all the three approaches. B: Even with the simple interaction between the non-focal predictors, we start seeing larger deviation of the predicted mean, $\bar{\hat{y}}$, from the observed, $\bar{y}$, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not, the proposed \pkg{varpred}.}
\label{fig:justify_plots}
\end{figure}

In the absence of interaction (model~\ref{eq:simple_inter_higher_no_interaction}), the three approaches produce similar estimates, which match the observed values, Fig~\ref{fig:justify_plots}A. However, in the presence of interaction, even as simple as the one in model~\ref{eq:simple_inter_higher}, the estimates starts to differ. In particular, \pkg{emmeans} and \pkg{effects} give similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is similar to the observed average ($\bar{y}$), Fig~\ref{fig:justify_plots}B. To generate Fig~\ref{fig:justify_plots}A, all the three packages averages the non-focal predictors $x_2$ and $x_3$. On other hand, to generate Fig~\ref{fig:justify_plots}B, the difference in the estimates lies on how each of the packages average the interaction term ($x_2x_3$). In particular \pkg{emmeans} and \pkg{effects} computes $\bar{x_2}\bar{x_3}$ while \pkg{varpred} computes $\over{x_2x_3}$.

An alternative to averaging the non-focal predictors, is the \emph{population-based} approach which involves computing the prediction over the population of the non-focal predictors and then averaging across the values of the focal predictor \cite{hanmer2013behind}.

In some applications, we may only be interested in the uncertainties associated with a particular focal predictor -- \emph{isolated} confidence intervals. However, currently, the two packages do not provide straightforward way do achieve this. In addition to prediction lines shown in Fig~\ref{fig:justify_plots}A above, we may want to include the confidence bands associated with the predictions from  all the three methods, as shown in Fig~\ref{fig:justify_ci_plots}.

\begin{figure}[!h]
\centering
\includegraphics[width=0.8\textwidth]{justify_ci_plots-figure2.pdf}
\caption{{\bf A comparison of regular and isolated confidence bands.} The description of horizontal, vertical and trend lines remain the same as above. The wider dotted blue curves overlaying the green curves are the regular confidence bands from \pkg{emmeans} and \pkg{effects}, while the narrower black curves crossing at the mean of the focal predictor are the isolated confidence bands from \pkg{varpred}. For simple models, the isolated confidence bands crosses at the model center.}
\label{fig:justify_ci_plots}
\end{figure}

For \pkg{emmeans} and \pkg{effects}, the confidence bands are much wider because they include uncertainties associated with the non-focal predictors, but narrower and crosses at the mean of the focal predictor, i.e., model center, in \pkg{varpred}. In other words, with \pkg{varpred}, we are able to generate isolated confidence bands indicating zero uncertainty at the value of the focal predictor we are more certain about, i.e., mean of the focal predictor. For simple models, the point where the confidence bands crosses is the model center and it corresponds to typical value we choose, in this case, the mean of the focal predictor.

When dealing with nonlinear link functions, the correct predictions, for example, are even much harder to estimate. One approach is to make predictions on the transformed scale (linear predictor scale), and then back-transform to the original scale. However, the back-transformation may either result in biased predictions or requires some approximation. In particular, bias in expected mean prediction induced by nonlinear transformation of the response variable can lead inaccurate predictions.

The main purpose of this article is to discuss and implement various approaches for computing predictor effects and provide an alternative method for computing the associated confidence intervals. We further explore approaches for correcting bias in predictions for GL(M)Ms involving nonlinear link functions.

\section*{Quantities of interest}

Several quantities of interest may be derived from regression models. The first one is the coefficient estimates. Others are \emph{input variables},  \emph{predictors}, \emph{model matrix}, \emph{predicted values} and \emph{marginal effects}. Input variables are the variables that were measured (may be transformed), while predictors are the terms that are entered in the model. Hence, predictors encompass the main effects, but also polynomials of input variables and interaction terms (for example in polynomials and splines) \cite{schielzeth2010simple}. Model matrix refers to the design matrix whose rows include all combination of variables appearing in the interaction terms, along with the typical values of the focal and non-focal predictors.

In simple linear models with no interaction terms, the default output for the coefficient estimates are simple and directly interpretable as the expected change in outcome for a unit change in focal predictor. This is the \emph{unconditional marginal effect} \cite{leeper2017interpreting} and it is constant across all the observations and levels of all other predictors. Consider models ~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher}. The marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher_no_interaction} is $\frac{\partial y}{\partial x_2} = \beta_2$. On the other hand, the marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher} is given by $\frac{\partial y}{\partial x_2} = \beta_2 + \beta_{23}x_3$. In other words, if there are no interactions, the marginal effect of $x_2$ on $y$ is constant, while, if there are interactions in the model, the marginal effect of a change in $x_2$ on $y$ depends on the value of the other \emph{conditioning} predictor $x_3$. As a result, the interpretation of interaction models differs in an important way from linear-additive regression models.

To distinguish between predictor effects and marginal effects, consider result from a hypothetical  simulated example -- regression of household size as function of household wealth index and age of household head, as shown in Fig~\ref{fig:qoi_age_pred_plot}. Since the model has no interactions, the relationship between the predicted household size and age is linear, hence the marginal effect of age is the slope ($\frac{\Delta \mathrm{hh size}}{\Delta \mathrm{age}}$) of the predictor effect line and can be calculated irrespective of the values of wealth index.

\begin{figure}[!h]
\centering
\includegraphics[width=0.8\textwidth]{qoi_age_pred_plot-figure3.pdf}
\caption{{\bf A comparison of predictor effect and unconditional marginal effect of age on household size.} The black line is the predictor effect trend line line, while the dotted lines indicate the changes along the \code{age} and \code{household size} axis. For a linear model with no interaction, the marginal effect is the slope of the predictor effect line.}
\label{fig:qoi_age_pred_plot}
\end{figure}

Predictor effect, on the other hand, is the expected household size for a particular age, holding wealth index at its mean. In particular, the purpose and goal of a predictor effect seems fairly straightforward; for specified values of a focal predictor, we want to give a point estimate and confidence intervals for the prediction of the model for a typical individual with those values of the predictors.

\section*{Statistical background}

To get an intuition of how conditioning on the mean values of the non-focal predictors work, suppose we are interested in predictor effects of a particular predictor, i.e., focal, $x_f$, from the set of predictors. To keep it simple, assume that the model has no interaction terms. The idea is to \emph{anchor} the values of non-focal predictors to some particular values. For example fixing the values of non-focal predictor(s) at some typical values -- typically determined by averaging (for now) in some meaningful way, for example, arithmetic mean  for continuous and average over the levels of the factors for categorical non-focal predictors. The easiest way to achieve this is by constructing $\bX^\star$ by averaging the columns of non-focal predictors in model matrix $\bX$, and together with appropriately chosen values of focal predictor(s).

Consider a simple linear model with linear predictor $\eta = \bX\bbeta$ and let $g(\boldmu) = \boldeta$ be an identity link function (in the case of simple linear model), where $\boldmu$ is the expected value of response variable $y$. Let $\hat{\bbeta}$ be the estimate of $\bbeta$, together with the estimated covariance matrix $\Sigma = V(\hat{\bbeta})$ of $\hat{\bbeta}$. Let $\mathbf{X^*}$ be the model matrix, inheriting most of its key properties, for example transformations on predictors and interactions from the model matrix, $\mathbf{X}$. Then the prediction $\hat{\boldeta}^\star = \bX^\star\hat{\bbeta}$ is the predictor effect for the focal predictor in question \cite{fox2009effect}.

An alternative, save for later, formulation of predictor effect involves, expressing the linear predictor as the sum of the focal and non-focal predictor linear predictor. In particular, $\eta^\star(x_f, \nset{{\bar{x}}}) = \beta_f x_f + \sum \nset{\beta} \nset{{\bar{x}^\star}}$, where $\nset{{\bar{x}^\star}}$ are the appropriately averaged entries of non-focal predictors and $x_f$ is a vector of values of the focal predictors for a particular observation.

\subsection*{Dealing with higher order interactions}

Higher order terms such as interactions, splines, polynomials, etc., can be between the non-focal predictors or focal and non-focal predictor(s). In the former case, we treat the interactions as just another column in the variable space (of the model matrix). In the later case, the non-focal variables in the model matrix are averaged as before. However, for the interacting focal predictors, a combination of each unique levels (or quantiles) are first generated and then the interaction terms are generated by multiplying these combinations. For example, consider model~\ref{eq:simple_inter_higher} above. In the first case, if we consider $x_1$ as the focal predictor, the interaction ($x_2x_3$) is between the non-focal predictors, $x_2$ and $x_3$, the linear predictor is

\begin{align*}
\eta^\star(x_{1i}, \nset{{\bar{x}^\star}}) = \beta_0\mathbf{1} + \beta_1 x_{1i} + \beta_2\bar{x}_2 + \beta_3\bar{x}_3 + \beta_{23}\over{x_2x_3}
\end{align*}
where $x_{1i}$ are the carefully chosen levels of the focal predictor. On the other hand, for the second case, if we consider $x_2$ as the focal predictor, then the interaction ($x_2x_3$) is between focal and non-focal predictor. In this case, the linear predictor is given by

\begin{align*}
\eta^\star(x_{2ij}, \nset{{\bar{x}^\star}}) = \beta_0\mathbf{1} + \beta_1 \bar{x}_1 + \beta_2x_{2ij} + \beta_3x_{3j} + \beta_{23}x_{2ij}x_{3j}
\end{align*}

In general, our formulation, even for more complicated interactions, follow these two basic principles -- interaction between non-focal predictors and interaction between focal and non-focal predictors.

\section*{Uncertainty propagation}

What about the confidence intervals (CI)? The limits of the confidence intervals are points, not mean values. In principle, every value of focal predictor has a different CI. The traditional way to compute variances for predictions is $\sigma^2 = \textrm{Diag}(\bX^\star \Sigma \bX^{\star\top})$ \cite{lenth2018package, fox2009effect}, so that the confidence intervals are $\eta \pm q\sigma$, where $q$ is an appropriate quantile of Normal or t distribution. This approach incorporates all the uncertainties -- including the uncertainty due to non-focal predictors.  But what if we are only interested in the uncertainty as a result of the focal predictor, so that the confidence intervals are $\eta \pm q \sigma_f$, i.e., isolated confidence intervals?

Currently, commonly used \proglang{R} packages for constructing predictions do not exclude the uncertainties resulting from the non-focal predictors when computing the CIs. A non-trivial way to exclude uncertainties associated with non-focal predictors in some of these packages is to provide a user defined variance-covariance matrix with the covariances of non-focal terms set to $0$ -- \emph{zeroing-out} variance-covariance matrix. This only works when the input predictors are \emph{centered} prior to model fitting, in case of numerical predictors, and even much complicated when the predictors are categorical. We first describe the variance-covariance based approach and then discuss our proposed method which is based on \emph{centering model matrix} and does not require input predictors to be scaled prior to model fitting.

\subsection*{Variance-covariance}

The computation of $\hat{\eta}^\star$ remains the same as described above. However, to compute $\sigma$, $\Sigma$ is modified by \emph{zeroing-out} (the variance-covariance of all non-focal predictors are set to zero) variances of non-focal terms. Although this is the simplest approach, it requires centering of continuous, i.e., $x_c = x - \bar{x}$, predictors prior to model fitting and proper way to average categorical predictors.

\subsection*{Centered model matrix}

Consider centered model matrix $\bX^{\star}_{c} = \{\bX_f^\star, \nset{{\bX}^\star} - \nset{{\bar{\bX}}^\star}\}$. It follows that the non-focal terms in $\bX^{\star}_{c}$ are all zero in simple models without interactions but are isolated (narrower) around the model center in models involving complex interactions. Consequently the uncertainty due to non-focal predictors are isolated in the computation of $\sigma^2 = \textrm{Diag}(\bX^\star_c \Sigma \bX^{\star\top}_c)$. In addition, the computation of $\bX^{\star}_c$ impacts only on the intercepts and the non-focal terms, i.e., the slopes and variance of the focal predictors are not affected. This means that we can still generate isolated CIs without necessarily centering the predictors prior to model fitting.

\section*{Bias correction}

In many applications, it usually important to report the estimates that reflect the expected values of the untransformed response. However, when dealing with nonlinear link functions, it is even harder to generate correct predictions that reflect the untransformed response due to the bias in the expected mean induced by the nonlinear transformation of the response variable. In such cases, bias correction is needed when back-transforming the predictions to the original scales. Most common approach for bias-adjustment is second-order Taylor approximation \cite{lenth2018package, duursma2003bias}. Here, we describe and implement a different approach, population-based approach for bias correction.

\subsection*{Population-based approach for bias correction}

The most precise (although not necessarily accurate!) way to predict is to condition on values of the focal predictor and make predictions for all observations (members of the population) \cite{hanmer2013behind}. A key point is that the nonlinear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population to a one-dimensional distribution of $\eta^\star$.

Once we have got our vector of $\eta^\star$ (which is essentially a set of samples from the distribution over $\eta^\star(x_f, \nset{x})$ for the conditional set), we may want a mean and confidence intervals of the mean on the response (data) scale, i.e. after back-transforming. In other words we can use the observations themselves then we just compute the individual values of $g^{-1}(\eta^\star)$ and compute their mean.

In population-based approach, $\eta^\star$ is constructed as a function of properly constituted levels of focal predictors, $x_f$, and entire population of the non-focal predictors, $\nset{x}$, as opposed to averaged non-focal predictors, $\nset{{\bar{x}}}$, described in the previous section. More specifically:

\begin{itemize}
\item compute linear predictor associated with the non-focal predictors, $\nset{\eta} = \sum \nset{\beta} \nset{x}$
\item compute linear predictor associated with the focal predictors, $\eta_{jf} = \sum{\beta_f x_{jf}}$
\item for every value of the focal predictor, $\eta_{jf}$:

\begin{align}\label{eq:pop_eta} 
\eta_j^\star  &= \eta_{jf} + \nset{\eta} \\
\hat{y}_j  &= \textrm{mean} ~ g^{-1} \left(\eta_j^\star\right)
\end{align}

\end{itemize}
We make similar adjustments to compute the variances of the predictions at every level of the focal predictor:

\begin{align}
\sigma_{jf}^2 = \textrm{Diag}(\bX^\star_{jc} \Sigma \bX^{\star\top}_{jc})
\end{align}
where $\bX^{\star}_{jc} = \{\bX_{jf}^\star, \nset{{\bX}^\star} - \nset{{\bar{\bX}}^\star}\}$ and

\begin{align}
\mathrm{CI}_j = \mathrm{mean} ~ \eta_j^\star \pm q\sigma_{jf}
\end{align}

For models with random effects components, we make further adjustment to correct for bias induced by the random effects terms. In the population approach, we treat the random effects terms as additional non-focal predictors and simply make adjustment to Equation~\ref{eq:pop_eta}. In particular

\begin{align}\label{eq:pop_eta_re} 
\tau &= \bZ b \nonumber \\
\eta_j^\star  &= \eta_{jf} + \nset{\eta} + \tau_j
\end{align}
where $\bZ$ and $b$ are the design matrix and a vector of random effects, respectively.

\nolinenumbers

% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% or
%
% Compile your BiBTeX database using our plos2015.bst
% style file and paste the contents of your .bbl file
% here. See http://journals.plos.org/plosone/s/latex for 
% step-by-step instructions.
%

\bibliography{vareffects_plosone}

\end{document}

### 2022-01-19 manuscript/quantity_of_interest.R

@@ -1,0 +1,52 @@
+library(shellpipes)
+library(vareffects); varefftheme()
+library(ggpubr)
+library(ggplot2)
+library(dplyr)
+
+loadEnvironments()
+startGraphics()
+ 
+set.seed(9991)
+
+############################################################################################
+# Global variables
+nHH_obs <- 100
+beta_age <- 0.1 # Focal predictor effect
+
+############################################################################################
+# Section: 2 Quantities of interest
+## Figure 3
+############################################################################################
+
+## Coefficient estimate
+qoi_df <- linearsim(nHH=1000, form=~1+x1*x2
+	, betas=c(5, 1, 2, 0.5)
+	, pgausian=list(p=2, fun=rnorm, mean=c(0.2,0), sd=1)
+	, pcat=list(p=0)
+	, link_scale = TRUE
+)$data
+colnames(qoi_df) <- c("age", "wealthindex", "hhsize")
+qoi_mod1 <- lm(hhsize~age+wealthindex, qoi_df)
+qoi_mod2 <- lm(hhsize~age*wealthindex, qoi_df)
+
+## Prediction plot
+qoi_age_pred <- varpred(qoi_mod1, "age")
+qoi_age_pred_plot <- (plot(qoi_age_pred, ci=FALSE)
+	+ geom_segment(aes(x=0,y=coef(qoi_mod1)[[1]],xend=1,yend=coef(qoi_mod1)[[1]]), colour="blue")
+	+ geom_segment(aes(x=min(qoi_age_pred$preds$age), y=coef(qoi_mod1)[[1]]+coef(qoi_mod1)[[2]], xend=1, yend=coef(qoi_mod1)[[1]]+coef(qoi_mod1)[[2]]), linetype=2)
+	+ geom_segment(aes(x=min(qoi_age_pred$preds$age), y=coef(qoi_mod1)[[1]], xend=0, yend=coef(qoi_mod1)[[1]]), linetype=2)
+	+ geom_segment(aes(x=1,y=coef(qoi_mod1)[[1]],xend=1,yend=coef(qoi_mod1)[[1]]+coef(qoi_mod1)[[2]]), colour="blue")
+	+ geom_segment(aes(x=1,y=min(qoi_age_pred$preds$fit),xend=1,yend=coef(qoi_mod1)[[1]]), linetype=2)
[…]

### 2022-01-19 manuscript/variable_predictions_funs.R

@@ -1,0 +1,365 @@
+## More general simulation function
+### Default: Continuous outcome 2 and 1 categorical predictors
+### form : specify model using formula
+
+linearsim <- function(nHH=1000, perHH=1, form=~1+x1+x2+x3
+	, hhSD=2, noiseSD=1, addnoiseGLM=FALSE, betas=NULL
+	, pgausian=list(p=2, fun=rnorm, mean=0, sd=1)
+	, pcat=list(p=1, fun=sample, nlevels=2, labels=NULL, prob=NULL, replace=TRUE)
+	, noutcomes=1
+	, beta0=log(runif(noutcomes))
+	, separatelatent=FALSE
+	, blatent=1
+	, mulatent=0
+	, sdlatent=1
+	, vnames=NULL
+	, link_scale=FALSE) {
+	
+	Terms <- attr(terms(form), "term.labels")
+	nbetas <- length(Terms)
+	if (noutcomes==1) {
+		if (!is.null(betas) & length(betas) != (nbetas+1))
+			stop(paste0("betas should a vector of length ", nbetas+1))
+	} else {
+		if (!is.null(betas) & length(betas) != (nbetas))
+			stop(paste0("betas should a vector of length ", nbetas+1))
+	}
+
+	p <- pgausian$p + pcat$p
+	nvars <- length(all.vars(form))
+
+	if (p != nvars) {
+		stop(paste0("\nNumber of predictors in form should be same as sum of p in pgausian and pcat: ", nvars, " != ", p))
+	}
+	
+	# TODO: warn users if there's intercept in the multiple outcome
+	check_intercept <- attr(terms(form), "intercept")
+	if (noutcomes>1) {
+		if (check_intercept>=1) {
+			form <- reformulate(c(all.vars(form), -1))
+			betas <- betas[-1]
+		}
+		if (length(beta0) != noutcomes) stop("length of beta0 should be the same as noutcomes")
+	}
+
+	if (noutcomes<1)stop("Number of outcome variables should be  >= 1")
+	if (perHH < 1)stop("perHH >= 1")
+	N <- nHH * perHH
+
+	pg <- pgausian$p
+	mu <- pgausian$mean
+	if (is.null(mu)) mu <- rep(0, pg)
+	.sd <- pgausian$sd
+	if(is.null(.sd)) sd <- 1
+
+	pgausian$p <- NULL
[…]

### 2022-01-23 manuscript/vareffects_plosone.tex
Autocommit

@@
 \newcommand{\ipsum}{{\bf IPSUM}}
-\newcommand{\fix}{{\bf FIXME}}
+\newcommand{\fix}{{\bf \textcolor{red}{FIXME}}}
 \newcommand{\JD}[1]{{\color{blue} \emph{#1}}}
@@
 \\
-Name1 Surname\textsuperscript{1,2\Yinyang},
-Name2 Surname\textsuperscript{2\Yinyang},
-Name3 Surname\textsuperscript{2,3\textcurrency},
-Name4 Surname\textsuperscript{2},
-Name5 Surname\textsuperscript{2\ddag},
-Name6 Surname\textsuperscript{2\ddag},
-Name7 Surname\textsuperscript{1,2,3*},
-with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
+Steve Cygu\textsuperscript{1},
+Benjamin M. Bolker\textsuperscript{1,2},
+Jonathan Dushoff\textsuperscript{1,2}
 \\
 \bigskip
-\textbf{1} Affiliation Dept/Program/Center, Institution Name, City, State, Country
+\textbf{1} School of Computational Science and Engineering, McMaster University, Hamilton, Ontario, Canada
 \\
-\textbf{2} Affiliation Dept/Program/Center, Institution Name, City, State, Country
-\\
-\textbf{3} Affiliation Dept/Program/Center, Institution Name, City, State, Country
+\textbf{2} Department of Biology, McMaster University, Hamilton, Ontario, Canada
 \\
@@
 
-% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
-% 
-% Remove or comment out the author notes below if they aren't used.
-%
-% Primary Equal Contribution Note
-\Yinyang These authors contributed equally to this work.
-
-% Additional Equal Contribution Note
-% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
-\ddag These authors also contributed equally to this work.
-
-% Current address notes
-\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
-% \textcurrency b Insert second current address 
-% \textcurrency c Insert third current address
-
-% Deceased author note
-\dag Deceased
-
-% Group/Consortium Author Note
-\textpilcrow Membership list can be found in the Acknowledgments section.
-
 % Use the asterisk to denote corresponding authorship and provide email address in note below.
-* correspondingauthor@institute.edu
+* cygu@aims.ac.za
 
@@
 
-\fix
+In many application, outcome predictions or predicted probabilities is of interest.
+For models that involve complex multiplicative interactions, additional non-focal predictors or nonlinear link functions, the estimated predictor effects are sensitive to how the predictions are generated.
+The most common approach is generating predictor effects at a ``typical value'', usually the mean, of non-focal predictor, i.e., mean-anchored which is the effect of an ``average case'' in the population.
+In the presence of sources of bias such as additional non-focal predictors, nonlinear link functions, random effect terms, etc., mean-anchored approach generate predictor effects that are biased and not consistent with the observed quantities.
+An alternative is the population-based approach which estimates the average effect in the population.
+Moreover, isolated confidence intervals provide an alternative and a more clear way to describe uncertainty associated with the focal predictor of interest.
+In addition to theoretical and methodical comparison, using simulation, we illustrate the two approaches and show that they can produce substantially different results and that population-based approach can not only produce estimates consistent with the observed values, but also appropriate for bias correction.
 
@@
 In other words, if there are no interactions, the marginal effect of $x_2$ on $y$ is constant, while, if there are interactions in the model, the marginal effect of a change in $x_2$ on $y$ depends on the value of the other \emph{conditioning} predictor $x_3$.
-As a result, the interpretation of interaction models differs in an important way from linear-additive regression models.
+In general, for a continuous variable, $x_k$, the marginal effect is $\frac{\partial y}{\partial x_k} = g(x\beta)\beta_k$, where $g()$ is the probability density function.
+This implies that, in addition to coefficient estimates, the marginal effect depends on the nature of $g()$ and the values of the other non-focal predictors.
+For nonlinear link functions, $g()$, the effect is not constant and instead depends on the function and the values over which the curve is evaluated.
 
@@
 For example fixing the values of non-focal predictor(s) at some typical values -- typically determined by averaging (for now) in some meaningful way, for example, arithmetic mean  for continuous and average over the levels of the factors for categorical non-focal predictors.
-The easiest way to achieve this is by constructing $\bX^\star$ by averaging the columns of non-focal predictors in model matrix $\bX$, and together with appropriately chosen values of focal predictor(s).
+We refer to this as mean-anchored approach.
+The most convenient way to achieve this is by constructing $\bX^\star$ by averaging the columns of non-focal predictors in model matrix $\bX$, and together with appropriately chosen values of focal predictor(s).
 
@@
 
-An alternative, save for later, formulation of predictor effect involves, expressing the linear predictor as the sum of the focal and non-focal predictor linear predictor.
-In particular, $\eta^\star(x_f, \nset{{\bar{x}}}) = \beta_f x_f + \sum \nset{\beta} \nset{{\bar{x}^\star}}$, where $\nset{{\bar{x}^\star}}$ are the appropriately averaged entries of non-focal predictors and $x_f$ is a vector of values of the focal predictors for a particular observation.
+An alternative, save for later, formulation of predictor effect involves, expressing the linear predictor as the sum of the focal and non-focal predictor linear predictor. In particular,
+
+\begin{align}\label{eq:eta_mean}
+\eta^\star(x_f, \nset{{\bar{x}}}) &= \beta_f x_f + \sum \nset{\beta} \nset{{\bar{x}^\star}} \\
+\hat{y}_f  &= g^{-1} \left(\eta^\star(x_f, \nset{{\bar{x}}})\right)
+\end{align}
+where $\nset{{\bar{x}^\star}}$ are the appropriately averaged entries of non-focal predictors and $x_f$ is a vector of values of the focal predictors for a particular observation.
 
@@
 The most precise (although not necessarily accurate!) way to predict is to condition on values of the focal predictor and make predictions for all observations (members of the population) \cite{hanmer2013behind}.
-A key point is that the nonlinear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population to a one-dimensional distribution of $\eta^\star$.
-
-Once we have got our vector of $\eta^\star$ (which is essentially a set of samples from the distribution over $\eta^\star(x_f, \nset{x})$ for the conditional set), we may want a mean and confidence intervals of the mean on the response (data) scale, i.e.
-after back-transforming.
-In other words we can use the observations themselves then we just compute the individual values of $g^{-1}(\eta^\star)$ and compute their mean.
-
-In population-based approach, $\eta^\star$ is constructed as a function of properly constituted levels of focal predictors, $x_f$, and entire population of the non-focal predictors, $\nset{x}$, as opposed to averaged non-focal predictors, $\nset{{\bar{x}}}$, described in the previous section.
+A key point is that the nonlinear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population to a one-dimensional distribution of $\eta^\star(x_f, \nset{x})$, which is a function of focal predictor and the observed values of the non-focal predictors, as oposed to the definition in Equation~\ref{eq:eta_mean}.
 More specifically:
@@
 \begin{align}\label{eq:pop_eta} 
-\eta_j^\star  &= \eta_{jf} + \nset{\eta} \\
-\hat{y}_j  &= \textrm{mean} ~ g^{-1} \left(\eta_j^\star\right)
+\eta_j^\star(\eta_{jf}, \nset{\eta})  &= \eta_{jf} + \nset{\eta} \nonumber \\
+&= \eta_j^\star(x_f, \nset{x})
+\end{align}
+\end{itemize}
+
+Once Equation~\ref{eq:pop_eta} is computed, one can back-transform the estimates to the original scale and the average over the levels of the focal predictors, $j$:
+
+\begin{align}\label{eq:pop_response} 
+\hat{y}_f  &= \textrm{mean} ~ g^{-1} \left(\eta_j^\star(x_f, \nset{x})\right)
 \end{align}
 
-\end{itemize}
 We make similar adjustments to compute the variances of the predictions at every level of the focal predictor:
@@
 \begin{align}
-\mathrm{CI}_j = \mathrm{mean} ~ \eta_j^\star \pm q\sigma_{jf}
+\mathrm{CI}_f = \mathrm{mean} ~ g^{-1} \left(\eta_j^\star(x_f, \nset{x}) \pm q\sigma_{jf}\right)
 \end{align}
@@
 \tau &= \bZ b \nonumber \\
-\eta_j^\star  &= \eta_{jf} + \nset{\eta} + \tau_j
+\eta_j^\star(x_f, \nset{x}, \tau)  &= \eta_j^\star(x_f, \nset{x}) + \tau
 \end{align}
 where $\bZ$ and $b$ are the design matrix and a vector of random effects, respectively.
+
+\section*{Mean-anchored vs population-based}
+
+As mentioned above, the two common choices to generate predictor effects are: 1) setting the non-focal predictors to their mean -- mean-anchored; and 2) using the entire population of the non-focal predictors -- population-based.
+If the link function is nonlinear and/or the model has complex higher order interactions, the average of the predictions evaluated at the mean of the non-focal predictors and the average of predictions evaluated at the population level (and then averaged) of the non-focal predictor are not equivalent, i.e.,
+
+\begin{align}\label{eq:compare_anchored_pop_based}
+g^{-1} \left(\eta_j^\star(\bar{x}_f, \nset{\bar{x}})\right) \neq \frac{1}{n} \sum_{i=1}^n{ g^{-1} \left(\eta_j^\star(\bar{x}_f, \nset{x})\right)}.
+\end{align}
+
+If $g()$ is strictly concave or strictly convex, one can use Jensen's inequality to determine which side of Equation~\ref{eq:compare_anchored_pop_based} is smaller or larger.
+However, for non-fully convex or concave link function, one can use Taylor series expansion to approximate the range over which the two are smaller or smaller than the other \cite{hanmer2013behind}.
+
+As stated above, in simple linear models without interactions, the effect is constant, so both approaches yield similar results.
+However, picking a single value, e.g., mean of the predictor, on which to draw conclusions about the effect can be problematic, unrealistic or not contained in or representative of the population.
+In addition, the mean-anchored approach fails to use the every values of the non-focal predictors hence not utilizing the full potential of the information contained in the data.
+This may limit the inferences we can make about the entire population.
+In general, the mean-anchored approach provides the predictor effect of an average case, whereas, the population-based approach, summarizes the predictor effect over the entire population -- and in some applications, the effect of an average case might not be generalizable to the entire population, especially, if the average does not represent the population.
+The population-based focuses on specific observations since the prediction is first obtained for each observation and then averaged across the levels of the focal predictor.
+
+Another potential concern with the mean-anchored approach arises in situations where direct naive use leads to rare or meaningless basis for generalization.
+For example, setting categorical dummy categorical variables in the model matrix to their means, which, by default, sets them to their sample means or observed proportions.
+For example, a dummy variable for christian household heads may be set to $0.2$, translating to prediction for an household head who is $20\%$ christian.
+This problem may extend to models with other complex interaction terms \cite{hanmer2013behind}.
+
+Population-based approach is not entirely foolproof.
+For instance, similar to the mean-anchored approach, in the case of continuous focal predictors, choosing the representative values of the focal predictors can be very challenging especially if the cases are not evenly distributed around the minimum and the maximum values or within some subgroups defined in the population.
+In addition, population-based approach can be very computationally intensive for large datasets.
+
+\section*{Simulation examples}
+
+We now illustrate the construction of predictor effects with isolated confidence intervals and also demonstrate that the mean-anchored and population approaches produce different results in models with nonlinear link functions and/or additional source of potential bias.
+In addition, we demonstrate that population-based approach can be used to correct bias in the predictions.
+
+\subsection*{Isolated confidence intervals}
+
+As mentioned above, in some applications, one may be interested in uncertainties associated with the focal predictor only.
+In that case, it is important to exclude all the uncertainties as a result of the non-focal predictors and only show confidence intervals describing the predictor of interest.
+To illustrate this, we consider the simulation model described in \nameref{S1_Appendix}.
+
+We start by fitting two models from the simulated data, i.e., consider the polynomial predictor, age, as a focal predictor and also, consider the non-polynomial predictor, wealth index as the focal predictor.
+After fitting the model, we constructed predictor effects and then compared isolated vs non-isolated confidence intervals associated with these predictions (see Fig~\ref{fig:pred_cubic_plots}).
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.8\textwidth]{pred_cubic_plots-figure4.pdf}
+\caption{{\bf Isolated and non-isolated predictor effects $95\%$ confidence bands.} These figures compare the predictor effects together with their corresponding $95\%$ confidence bands.
+A: The focal predictor is a cubic polynomial.
+B: The focal predictor is not a polynomial function, implying that the complex interaction is in the non-focal predictor.
+In both cases, the confidence bands are narrower in the isolated case, and crosses at the mean of the focal predictor (model center) in the case of simple (non-polynomial) focal predictor -- B.
+In the case of cubic polynomial focal predictor, the model center is not a point in the predictor space, i.e., mean of the focal predictor, but a combination of all the terms in the cubic polynomial, consequently, the isolated bands do not cross at a point -- A.
+The horizontal black and red dotted lines are the observed and predicted average household size, i.e., $\over{hh~size}$ and $\over{\widehat{hh~size}}$, respectively; the inner red dotted curves and crossing line are the isolated confidence bands for A and B, respectively, while the outer blue curves are the non-isolated confidence bands in both cases.
+The central curve and trend line, in A and B, respectively, are the predictor effect of age and wealth index on the predicted household size, respectively.}
+\label{fig:pred_cubic_plots}
+\end{figure}
+
+In the absence of complex higher order interaction terms or transformations on the focal predictor in simple linear models, we would expect, at the model center, the observed average, $\over{hh~size}$, and the average of the predictions, $\over{\widehat{hh~size}}$, to be identical and cross, in the case of isolated confidence bands, at the model center (see Fig~\ref{fig:pred_cubic_plots}B).
+On the other hand, if the focal predictor is characterized by interactions or any other form of transformation, the model center is not necessarily a point in the predictor space.
+In this case, the isolated confidence bands will be narrower then the non-isolated but not necessarily crossing (see Fig~\ref{fig:pred_cubic_plots}A).
+
+\subsection*{Bias correction}
+
+Mean-anchored and population-based approaches can produce very different results in models with nonlinear link functions with additional sources of potential bias such as additional non-focal predictor(s), random effect terms or complex interactions.
+To demonstrate this, we considered a two predictor binary outcome simulation described in \nameref{S2_Appendix}, such that the age effect is way smaller than that of the wealth index, and compared their predictor effect on the predicted probability of improved water quality as shown in Fig~\ref{fig:pred_bin_plots}.
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.8\textwidth]{pred_bin_plots-figure5.pdf}
+\caption{{\bf Mean-anchored and population-based predictor effects.} The two approaches produce relatively different predictor effects for each of the predictors but the difference is more pronounced when the effect of the particular focal predictor has no strong effect -- \emph{on the left}.
+In both cases, age and wealth index effect, the average predicted probability of improved water based on population-based approach is very close to the observed proportion in the simulated data.
+The horizontal blue, black and the red dotted lines are the average mean-anchored, population-based predicted probability and observed proportion of improved water quality, respectively.
+The vertical dotted black line represents the mean of the focal predictor, and the point at which it crosses the red dotted line represents the expected ``perfect'' prediction at the model center.
+The dotted grey points are binned observations -- observed proportions of improved water quality in each bin.
+We would expect the all the curves, horizontal, and vertical lines to cross at the model center but due the nonlinear averaging (at least in the bias corrected population-based approach) of the predictions on the response scale, this may not always be case.}
+\label{fig:pred_bin_plots}
+\end{figure}
+
+If there was no effect of nonlinear averaging, then we would expect the observed proportion and the average of the predicted probabilities to cross at the model center as we see in Fig~\ref{fig:pred_cubic_plots}B.
+However, the differences we see in Fig~\ref{fig:pred_bin_plots} are due to nonlinear averaging since both observed status and predicted probabilities are averaged on the response scale as opposed to link scale.
+In particular, if the range of values are bigger than $0.5$ (seemingly the case here), then we would expected the averages to be slightly higher than what we would expect at the model center, and vice versa.
+
+\subsection*{Mediated effect}
+
+Consider the simple indirect mediation simulation described in \nameref{S3_Appendix}.
+We then fitted two models -- \emph{non-mediated} which models \emph{z} as a function of \emph{x} and the \emph{mediated} which models \emph{z} as a function of both \emph{x} and \emph{y}.
+In both cases, we compared mean-anchored and the population-based predictions, as shown in Fig~\ref{fig:pred_mediated_plots}.
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.8\textwidth]{pred_mediated_plots-figure6.pdf}
+\caption{{\bf Mean-anchored and population-based predictor effects for mediated effect.} These figures compare the mean-anchored and population-based predictor effect of $x$ from a mediated, $z \sim x + y$, and mediated, $z \sim x$, models.
+A: In the absence of mediator variable, both mean-anchored and population-based approaches predictors are identical since there is no any other additional (non-focal predictors) sources of bias.
+Similarly, in both approaches, the average predicted probability is very close to the observed proportion.
+B: When the mediator variable is included, there is no direct effect of $x$ on $z$ and as a result the predictor effect curve does not align with the observations.
+However, the population-based approach still closely approximates the average proportion in the simulated data.
+The simulated data.
+The horizontal blue, black and the red dotted lines are the average mean-anchored, population-based predicted probability and observed proportion, respectively.
+The vertical dotted black line represents the mean of the focal predictor, and the point at which it crosses the red dotted line represents the expected ``perfect'' prediction at the model center.
+The dotted grep points are the binned observations.}
+\label{fig:pred_mediated_plots}
+\end{figure}
+
+From Fig~\ref{fig:pred_mediated_plots}A, we see what we would expect in the absence of additional sources of bias, even though in the simulation, the effect of $x$ on $z$ is mediated through $y$.
+By ignoring $y$ in the model, we are still able to capture the effect of $x$ and closely match the observed values using both approaches.
+However, if there were additional non-focal predictors, we would expect to see the differences similar to those in Fig~\ref{fig:pred_bin_plots}.
+Including both $y$ and $x$ dilutes the direct effect of $x$ on $y$ and as a result, our prediction do not necessarily match the observed binned observations (see Fig~\ref{fig:pred_mediated_plots}B).
+
+\section*{Discussion}
+
+Our simulations examples majorly focused on simple linear and logistic models due their wide range of usage and application.
+In addition, these models act as a starting point for building other complex models, including mixed effect models and models with categorical predictors.
+However, the logic for extension of approaches to more complex models, including other forms of nonlinear link functions is very straightforward.
+In addition, our \proglang{R} package implementation already extends to and supports most of the nonlinear link functions and mixed model framework.
+
+Although commonly used \proglang{R} software packages, by default, implements mean-anchored approach, our simulation results demonstrated that the population-based approach has a potential of yielding results which are more consistent with the observed data.
+We would, therefore, argue that the use of mean-anchored approach should have some theoretical justification, especially in complex models.
+
+\section*{Conclusion}
+
+Generating outcome predictions or predicted probabilities from simple and generalized linear (mixed) models is not only important but also, generating quantities which are consistent with the observed values should be of interest.
+However, many studies still report coefficient estimates from generalized models like probit, logistic, etc.,\cite{hanmer2013behind}, which are subject of less clarity due to lack of direct link to the outcome of interest.
+
+The argument and results we present in this paper supports a greater need for a shift on focus on what and how to present predictions from generalized models.
+For example, we believe that isolated confidence intervals could provide more clarity concerning the uncertainty due to the predictor of interests as opposed to the traditional way of incorporating everything.
+
+From our theoretical, methodological and simulation results, researchers these kind of models should, in the absence of theoretical justification, report predictions based on population-based approach or at least attempt to do a comparison of the two approaches before settling on the most appropriate in answering their research question.
+Moreover, we provide \proglang{R} package, \pkg{vareffects}, which implements these methods and is available on github (\href{https://github.com/mac-theobio/effects}{https://github.com/mac-theobio/effects}).
+
+\section*{Supporting information}
+
+% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
+\paragraph*{S1 Appendix.}
+\label{S1_Appendix}
+{\bf Cubic polynomial interaction simulation.} Consider an hypothetical simulation which simulates household size as a function of household wealth index and cubic function of the age of the household head, specified as follows:
+
+\begin{align}\label{sim:lm_cubic}
+\mathrm{hh~size}_i &= \beta_0 + \beta_{\mathrm{A_1}}\mathrm{Age}_i + \beta_{\mathrm{A_2}}\mathrm{Age}^2_i + \beta_{\mathrm{A_3}}\mathrm{Age}^3_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \epsilon_i \nonumber\\
+\mathrm{Age}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
+\mathrm{Wealthindex}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
+\epsilon_i &\sim \mathrm{Normal}(0, 10) \nonumber\\
+\beta_0 &= 20 \nonumber\\
+\beta_{\mathrm{A}_1} &= 0.1 \nonumber\\
+\beta_{\mathrm{A}_2} &= 0.8 \nonumber\\
+\beta_{\mathrm{A}_3} &= 0.3 \nonumber\\
+\beta_{\mathrm{W}} &= -0.5 \nonumber\\
+i &= 1,\cdots, 100
+\end{align}
+
+\paragraph*{S2 Appendix.}
+\label{S2_Appendix}
+{\bf Binary outcome simulation.} Consider a simple simulation for improved water quality in Nairobi slums, such that the status is $1$ for improved and $0$ for unimproved water quality.
+In additional to the focal predictor, age of the household head, we add wealth index.
+In particular:
+
+\begin{align}\label{sim:glm_two_pred}
+\mathrm{status}_i &\sim \mathrm{Binomial}(1, \mathrm{P(status}_i = 1)) \nonumber\\
+\mathrm{logit(P(status}_i = 1)) &= \eta_i \nonumber\\
+\mathrm{\eta}_i &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i \nonumber\\
+\mathrm{Age}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
+\mathrm{Wealthindex}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
+\beta_0 &= 5 \nonumber\\
+\beta_{\mathrm{A}} &= 0.5 \nonumber\\
+\beta_{\mathrm{W}} &= 1.5 \nonumber\\
+i &= 1,\cdots, 10000
+\end{align}
+
+\paragraph*{S3 Appendix.}
+\label{S3_Appendix}
+{\bf Mediated effect simulation.} Next, we consider a simple indirect mediation simulation such that $x$ has direct effect on $y$ which in turn has effect on $z$ but $x$ has no direct effect on $y$, i.e., $x \rightarrow y \rightarrow z$.
+In particular:
+
+\begin{align}\label{sim:simple_mediate}
+z_i &\sim \mathrm{Binomial(1, P(z_i = 1))} \nonumber\\
+\mathrm{logit(P}(z_i = 1)) &= \eta_i \nonumber\\
+\eta_i &= \beta_0 + \beta_{yz} y_i \nonumber\\
+y_i &= \beta_{xy} x_i + \epsilon_y \nonumber\\
+x_i &\sim \mathrm{Normal(0, 1)} \nonumber\\
+\epsilon_y &\sim \mathrm{Normal(0, 1)} \nonumber\\
+\beta_0 &= 5 \nonumber\\
+\beta_{xy} &= 1 \nonumber\\
+\beta_{yz} &= 1.5 \nonumber\\
+i &= 1,\cdots, 10000
+\end{align}
 

### 2022-01-23 manuscript/cubic_predictors.R

@@ -1,0 +1,41 @@
+library(shellpipes)
+library(tidyr)
+library(dplyr)
+
+loadEnvironments()
+startGraphics()
+ 
+set.seed(9991)
+
+############################################################################################
+# Global variables
+N <- 100
+b0 <- 20
+age_b11 <- 0.1
+age_b12 <- 0.8
+age_b13 <- 0.3
+b2 <- -0.5
+age_sd <- 1
+wealthindex_sd <- 1
+hhsize_sd <- 10
+
+
+############################################################################################
+# Cubic polynomial
+############################################################################################
+
+### 1 polynomial (focal)
+### 1 non-focal predictor
+
+#### Simulation
+sim_df_cubic <- (
+	data.frame(age=rnorm(N, 0, age_sd)
+		, wealthindex=rnorm(N, 0, wealthindex_sd)
+	)
+	%>% mutate(eta = b0 + age_b11*age + age_b12*age^2 + age_b13*age^3 + b2*wealthindex
+		, hhsize=rnorm(N, mean=eta, sd=hhsize_sd)
+	)
+)
+head(sim_df_cubic)
+
+saveVars(sim_df_cubic, comparevarpred)

### 2022-01-23 manuscript/cubic_predictors_preds.R

@@ -1,0 +1,98 @@
+library(shellpipes)
+library(vareffects); varefftheme()
+library(ggpubr)
+library(ggplot2)
+library(dplyr)
+library(emmeans)
+
+loadEnvironments()
+startGraphics()
+
+## Focal polynomial
+
+### Non-isolated CIs 
+pred_age_non_cubic <- varpred(mod_cubic
+	, "age"
+	, isolate=FALSE
+	, modelname="Non-isolated"
+)
+
+### Isolated CIs
+pred_age_isolated_cubic <- varpred(mod_cubic
+	, "age"
+	, isolate=TRUE
+	, modelname="Isolated"
+)
+pred_fit_mean <- getmeans(pred_age_isolated_cubic, what="estimate", modelname="Predicted mean")
+print(pred_fit_mean)
+
+### Combine
+vlist <- list(pred_age_non_cubic, pred_age_isolated_cubic)
+
+pred_age_cubic_plots <- (comparevarpred(vlist=vlist
+		, lnames=NULL
+		, plotit=TRUE
+		, addmarginals=FALSE
+	)
+	+ geom_hline(data=pred_fit_mean, aes(yintercept=fit, colour=model), lty=1)
+	+ geom_hline(data=true_prop_df, aes(yintercept=fit, colour=model), lty=2)
+	+ scale_colour_manual(breaks = c("Observed mean", "Predicted mean", "Isolated", "Non-isolated")
+		, values=c("Observed mean"="black", "Predicted mean"="red", "Isolated"="red", "Non-isolated"="blue")
+	)
+	+ scale_linetype_manual(values=c("Observed mean"=2, "Predicted mean"=1, "Isolated"=1, "Non-isolated"=3))
+	+ labs(title="A) Focal polynomial", y="Predicted household size", linetype="Method", colour="Method")
+	+ theme(legend.position="bottom")
+)
+
+## Non-focal polynomial
+
+### Non-isolated CIs 
+pred_wealthindex_non_cubic <- varpred(mod_cubic
+	, "wealthindex"
+	, isolate=FALSE
+	, modelname="Non-isolated"
+)
+
+### Isolated CIs
+pred_wealthindex_isolated_cubic <- varpred(mod_cubic
+	, "wealthindex"
+	, isolate=TRUE
+	, modelname="Isolated"
+)
+pred_fit_mean <- getmeans(pred_wealthindex_isolated_cubic, what="estimate", modelname="Predicted mean")
[…]

### 2022-01-24 manuscript/vareffects_plosone.tex
varpred manuscript

@@
 
-In many application, outcome predictions or predicted probabilities is of interest.
-For models that involve complex multiplicative interactions, additional non-focal predictors or nonlinear link functions, the estimated predictor effects are sensitive to how the predictions are generated.
+In many applications using genralized linear (mixed) models, outcome predictions or predicted probabilities are often of interest.
+For models that involve complex multiplicative interactions, additional non-focal predictors or nonlinear link functions, the estimated coefficient are not readily interpretable.
+A general way to summarize these kind of models is through predictor effects, which are also sensitive to which values of non-focal predictors are chosen.
 The most common approach is generating predictor effects at a ``typical value'', usually the mean, of non-focal predictor, i.e., mean-anchored which is the effect of an ``average case'' in the population.
@@
 
-In many applications, outcome predictions or probabilities, are often of interest.
+In many applications using simple or generalized linear (mixed) models, outcome predictions or probabilities, are often of interest.
 Predictions provide a great way to summarize what the regression model is telling us and very useful for interpreting and visualizing model estimates.
@@
 Since logistic models are nonlinear (due to the nonlinear link function), in multivariate models with interactions, the magnitude of the effect of change in the outcome depends on the values of the predictor of interest and other predictors.
-Hence, the conclusions one can make about the estimated effects greatly depends on how well we choose the values of the other variables.
+Hence, the conclusions one can make about the estimated effects greatly depends on how well we choose the values of the other predictors, i.e., \emph{non-focal} predictors.
 
@@
 \begin{enumerate}
-\item choice of representative values of \emph{focal} predictor(s) and appropriate \emph{model center} for \emph{non-focal} predictors especially in multivariate models
+\item choice of representative values of \emph{focal} predictor(s) and appropriate \emph{anchor} for non-focal predictors especially in multivariate models
 \item  propagation of uncertainty -- can we incorporate uncertainty in nonlinear components of GL(M)Ms? Should we exclude variation in non-focal parameters?
@@
 This generates -- \emph{predictor effects}\cite{fox2009effect}, \emph{marginal predictions} \cite{leeper2017package} and \emph{estimated marginal means} \cite{lenth2018package}.
-In this article, we refer to this quantity as predictor effects since it should, for example, tell us what we would expect the presponse to be at a particular value or level of the predictor.
+In this article, we refer to this quantity as predictor effects since it should, for example, tell us what we would expect the presponse to be at a particular value or level of the predictor, for an ``average case''.
 Formerly, predictor effects computes the expected outcome by meaningfully holding the non-focal predictors constant (or averaged in some meaningful way) while varying the focal predictor, with the goal that the outcome expected prediction represents how the model responds to the changes in the focal predictor.
@@
 The commonly used \proglang{R} software packages (\pkg{emmeans} and \pkg{effects}) for generating predictor effects, by default, averages the non-focal predictors.
-However, there are a number of choices a number of choices which one has to make when considering this approach -- for example, in the presence of interaction, averaging the interactions (averaging product of interacting predictors) versus product of the averages of the interacting predictors (default for \pkg{emmeans} and \pkg{effects}).
+However, there are a number of choices one can make when considering this approach -- for example, in the presence of interaction, averaging the interactions (averaging product of interacting predictors) versus product of the averages of the interacting predictors (default for \pkg{emmeans} and \pkg{effects}).
 We claim that neither of these two approaches is the most appropriate but we show that averaging the interaction closely matches the observed values.
@@
 
-We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -0.5$, $\beta_2 = 1$, $\beta_3 = 2$ and $\beta_{23} = 5$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the observed average i.e., $\bar{y}$.
-Fig~\ref{fig:justify_plots} compares the predictors effects of the two models.
+We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$ and $\beta_{23} = 5$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.
 
@@
 The horizontal blue, green and black lines are the mean predictions, i.e., $\bar{\hat{y}}$, the red ones are the mean observed $y$, i.e., $\bar{y}$, while the vertical dotted grey lines are the means of the corresponding focal predictors.
+The grey points are the binned simulated $y$.
 The trend lines represents the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other predictors at their average.
-A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the observed in all the three approaches.
-B: Even with the simple interaction between the non-focal predictors, we start seeing larger deviation of the predicted mean, $\bar{\hat{y}}$, from the observed, $\bar{y}$, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not, the proposed \pkg{varpred}.}
+A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the observed in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$.
+B: Even with the simple interaction between the non-focal predictors, we start seeing deviation of the predicted mean, $\bar{\hat{y}}$, from the observed, $\bar{y}$, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not, the proposed \pkg{varpred}.}
 \label{fig:justify_plots}
@@
 
-An alternative to averaging the non-focal predictors, is the \emph{population-based} approach which involves computing the prediction over the population of the non-focal predictors and then averaging across the values of the focal predictor \cite{hanmer2013behind}.
-
 In some applications, we may only be interested in the uncertainties associated with a particular focal predictor -- \emph{isolated} confidence intervals.
 However, currently, the two packages do not provide straightforward way do achieve this.
-In addition to prediction lines shown in Fig~\ref{fig:justify_plots}A above, we may want to include the confidence bands associated with the predictions from  all the three methods, as shown in Fig~\ref{fig:justify_ci_plots}.
+To illustrate this, we generate predictor effect of $x_2$ from model~\ref{eq:simple_inter_higher_no_interaction} together with the associated $95\%$ confidence bands, as shown in Fig~\ref{fig:justify_ci_plots}.
 
@@
 \includegraphics[width=0.8\textwidth]{justify_ci_plots-figure2.pdf}
-\caption{{\bf A comparison of regular and isolated confidence bands.} The description of horizontal, vertical and trend lines remain the same as above.
-The wider dotted blue curves overlaying the green curves are the regular confidence bands from \pkg{emmeans} and \pkg{effects}, while the narrower black curves crossing at the mean of the focal predictor are the isolated confidence bands from \pkg{varpred}.
+\caption{{\bf A comparison of traditional and isolated $95\%$ confidence bands.} The description of horizontal, vertical and trend lines remain the same as above.
+The wider dotted blue curves overlaying the green curves are the traditional confidence bands from \pkg{emmeans} and \pkg{effects}, while the narrower black curves crossing at the mean of the focal predictor, i.e., the model center, are the isolated confidence bands from \pkg{varpred}.
 For simple models, the isolated confidence bands crosses at the model center.}
@@
 
-When dealing with nonlinear link functions, the correct predictions, for example, are even much harder to estimate.
+When dealing with nonlinear link functions or models with multiple non-focal predictors, the correct predictions, for example, are even much harder to estimate.
 One approach is to make predictions on the transformed scale (linear predictor scale), and then back-transform to the original scale.
@@
 In particular, bias in expected mean prediction induced by nonlinear transformation of the response variable can lead inaccurate predictions.
+An alternative to averaging the non-focal predictors, is the \emph{population-based} approach, discussed in detail later, which involves computing the prediction over the population of the non-focal predictors and then averaging across the values of the focal predictor \cite{hanmer2013behind}.
 
 The main purpose of this article is to discuss and implement various approaches for computing predictor effects and provide an alternative method for computing the associated confidence intervals.
-We further explore approaches for correcting bias in predictions for GL(M)Ms involving nonlinear link functions.
+We further explore and demonstrate, using simulated data, approaches for correcting bias in predictions for GL(M)Ms involving nonlinear link functions.
 
@@
 \includegraphics[width=0.8\textwidth]{qoi_age_pred_plot-figure3.pdf}
-\caption{{\bf A comparison of predictor effect and unconditional marginal effect of age on household size.} The black line is the predictor effect trend line line, while the dotted lines indicate the changes along the \code{age} and \code{household size} axis.
+\caption{{\bf A comparison of predictor effect and unconditional marginal effect of age on household size.} The black line is the predictor effect trend line, while the dotted lines indicate the changes along the \code{age} and \code{household size} axis.
 For a linear model with no interaction, the marginal effect is the slope of the predictor effect line.}
@@
 The idea is to \emph{anchor} the values of non-focal predictors to some particular values.
-For example fixing the values of non-focal predictor(s) at some typical values -- typically determined by averaging (for now) in some meaningful way, for example, arithmetic mean  for continuous and average over the levels of the factors for categorical non-focal predictors.
-We refer to this as mean-anchored approach.
-The most convenient way to achieve this is by constructing $\bX^\star$ by averaging the columns of non-focal predictors in model matrix $\bX$, and together with appropriately chosen values of focal predictor(s).
+For example, fixing the values of non-focal predictor(s) at some typical values -- typically determined by averaging (for now) in some meaningful way, for example, arithmetic mean  for continuous and average over the levels of the factors for categorical non-focal predictors.
+We refer to this as \emph{mean-anchored} approach.
+The most convenient way to achieve this is by constructing $\bX^\star$ by averaging the columns of non-focal predictors in model matrix $\bX$, and together with appropriately chosen values of focal predictor.
 
@@
 This approach incorporates all the uncertainties -- including the uncertainty due to non-focal predictors.
-But what if we are only interested in the uncertainty as a result of the focal predictor, so that the confidence intervals are $\eta \pm q \sigma_f$, i.e., isolated confidence intervals?
+But what if we are only interested in the uncertainty as a result of the focal predictor, so that the confidence intervals are $\eta \pm q \sigma_f$?
+We call this isolated confidence intervals.
 
@@
 Most common approach for bias-adjustment is second-order Taylor approximation \cite{lenth2018package, duursma2003bias}.
+Another potential source of bias comes from additional non-focal predictors, especially in models with nonlinear link functions.
 Here, we describe and implement a different approach, population-based approach for bias correction.
@@
 
-We start by fitting two models from the simulated data, i.e., consider the polynomial predictor, age, as a focal predictor and also, consider the non-polynomial predictor, wealth index as the focal predictor.
+We start by fitting two models from the simulated data, i.e., consider a cubic polynomial predictor, age, as a focal predictor and also, consider the non-polynomial predictor, wealth index as the focal predictor.
 After fitting the model, we constructed predictor effects and then compared isolated vs non-isolated confidence intervals associated with these predictions (see Fig~\ref{fig:pred_cubic_plots}).
@@
 On the other hand, if the focal predictor is characterized by interactions or any other form of transformation, the model center is not necessarily a point in the predictor space.
-In this case, the isolated confidence bands will be narrower then the non-isolated but not necessarily crossing (see Fig~\ref{fig:pred_cubic_plots}A).
+In this case, the isolated confidence bands will be narrower than the non-isolated but not necessarily crossing at the model center (see Fig~\ref{fig:pred_cubic_plots}A).
 
@@
 The vertical dotted black line represents the mean of the focal predictor, and the point at which it crosses the red dotted line represents the expected ``perfect'' prediction at the model center.
-The dotted grey points are binned observations -- observed proportions of improved water quality in each bin.
-We would expect the all the curves, horizontal, and vertical lines to cross at the model center but due the nonlinear averaging (at least in the bias corrected population-based approach) of the predictions on the response scale, this may not always be case.}
+The grey points are binned observations -- observed proportions of improved water quality in each bin.
+We would expect all the curves, horizontal, and vertical lines to cross at the model center but due the nonlinear averaging (at least in the bias corrected population-based approach) of the predictions on the response scale, this may not always be the case.}
 \label{fig:pred_bin_plots}
@@
 \includegraphics[width=0.8\textwidth]{pred_mediated_plots-figure6.pdf}
-\caption{{\bf Mean-anchored and population-based predictor effects for mediated effect.} These figures compare the mean-anchored and population-based predictor effect of $x$ from a mediated, $z \sim x + y$, and mediated, $z \sim x$, models.
+\caption{{\bf Mean-anchored and population-based predictor effects for mediated effect.} These figures compare the mean-anchored and population-based predictor effect of $x$ from a mediated, $z \sim x + y$, and non-mediated, $z \sim x$, models.
 A: In the absence of mediator variable, both mean-anchored and population-based approaches predictors are identical since there is no any other additional (non-focal predictors) sources of bias.
-Similarly, in both approaches, the average predicted probability is very close to the observed proportion.
+Consequently, in both approaches, the average predicted probability is very close to the observed proportion.
 B: When the mediator variable is included, there is no direct effect of $x$ on $z$ and as a result the predictor effect curve does not align with the observations.
 However, the population-based approach still closely approximates the average proportion in the simulated data.
-The simulated data.
 The horizontal blue, black and the red dotted lines are the average mean-anchored, population-based predicted probability and observed proportion, respectively.
 The vertical dotted black line represents the mean of the focal predictor, and the point at which it crosses the red dotted line represents the expected ``perfect'' prediction at the model center.
-The dotted grep points are the binned observations.}
+The grey points are the binned observations.}
 \label{fig:pred_mediated_plots}
@@
 However, if there were additional non-focal predictors, we would expect to see the differences similar to those in Fig~\ref{fig:pred_bin_plots}.
-Including both $y$ and $x$ dilutes the direct effect of $x$ on $y$ and as a result, our prediction do not necessarily match the observed binned observations (see Fig~\ref{fig:pred_mediated_plots}B).
+Including both $y$ and $x$ ``dilutes'' the direct effect of $x$ on $y$ and as a result, our prediction do not necessarily match the observed binned observations (see Fig~\ref{fig:pred_mediated_plots}B).
 
@@
 However, the logic for extension of approaches to more complex models, including other forms of nonlinear link functions is very straightforward.
-In addition, our \proglang{R} package implementation already extends to and supports most of the nonlinear link functions and mixed model framework.
+In addition, our \proglang{R} package implementation already extends to and supports most of the nonlinear link functions and mixed model framework, including multivariate binary outcome models.
 
@@
 
-From our theoretical, methodological and simulation results, researchers these kind of models should, in the absence of theoretical justification, report predictions based on population-based approach or at least attempt to do a comparison of the two approaches before settling on the most appropriate in answering their research question.
+From our theoretical, methodological and simulation results, researchers using these kind of models should, in the absence of theoretical justification, report predictions based on population-based approach or at least attempt to do a comparison of the two approaches before settling on the most appropriate in answering their research question.
 Moreover, we provide \proglang{R} package, \pkg{vareffects}, which implements these methods and is available on github (\href{https://github.com/mac-theobio/effects}{https://github.com/mac-theobio/effects}).
@@
 
+\section*{Acknowledgments}
+
+This work was supported by a grant to Jonathan Dushoff from the Natural Sciences and Engineering Research Council of Canada (NSERC) Discovery.
+
+\section*{Author Contributions}
+
+\textbf{Conceptualization:} Jonathan Dushoff
+
+\noindent\textbf{Software:} Steve Cygu, Benjamin M. Bolker
+
+\noindent\textbf{Supervision:} Jonathan Dushoff, Benjamin M. Bolker
+
+\noindent\textbf{Writing – original draft:} Steve Cygu
+
 \nolinenumbers

### 2022-02-08 manuscript/vareffects_plosone.tex
Minor
Autocommit

@@
 
-\newcommand{\lorem}{{\bf LOREM}}
-\newcommand{\ipsum}{{\bf IPSUM}}
 \newcommand{\fix}{{\bf \textcolor{red}{FIXME}}}
@@
 
-In many applications using genralized linear (mixed) models, outcome predictions or predicted probabilities are often of interest.
+In many applications which use generalized linear (mixed) models, outcome predictions or predicted probabilities are often of interest.
 For models that involve complex multiplicative interactions, additional non-focal predictors or nonlinear link functions, the estimated coefficient are not readily interpretable.
-A general way to summarize these kind of models is through predictor effects, which are also sensitive to which values of non-focal predictors are chosen.
-The most common approach is generating predictor effects at a ``typical value'', usually the mean, of non-focal predictor, i.e., mean-anchored which is the effect of an ``average case'' in the population.
-In the presence of sources of bias such as additional non-focal predictors, nonlinear link functions, random effect terms, etc., mean-anchored approach generate predictor effects that are biased and not consistent with the observed quantities.
+A general way to summarize these kind of models is through predictor effects, which are not only dependent on choice of representative values of focal predictor but also sensitive to which values of non-focal predictors are chosen.
+The most common approach is generating predictor effects at a ``reference point'', usually the mean, of non-focal predictor, i.e., mean-based which could be considered as the effect of an ``average case'' in the population.
+In the presence of sources of bias such as additional non-focal predictors, nonlinear link functions, random effect terms, etc., mean-based approach generate predictor effects that are biased and may not be consistent with the observed quantities.
 An alternative is the population-based approach which estimates the average effect in the population.
@@
 In addition to theoretical and methodical comparison, using simulation, we illustrate the two approaches and show that they can produce substantially different results and that population-based approach can not only produce estimates consistent with the observed values, but also appropriate for bias correction.
+We also present an alternative way, isolated confidence intervals, to describe uncertainties associated with these predictions.
 
@@
 
-In many applications using simple or generalized linear (mixed) models, outcome predictions or probabilities, are often of interest.
+In many applications which use simple or generalized linear (mixed) models, outcome predictions or probabilities, are often of interest.
 Predictions provide a great way to summarize what the regression model is telling us and very useful for interpreting and visualizing model estimates.
@@
 \begin{enumerate}
-\item choice of representative values of \emph{focal} predictor(s) and appropriate \emph{anchor} for non-focal predictors especially in multivariate models
-\item  propagation of uncertainty -- can we incorporate uncertainty in nonlinear components of GL(M)Ms? Should we exclude variation in non-focal parameters?
+\item choice of representative values of \emph{focal} predictor and the \emph{reference point} for the non-focal predictors especially in multivariate models
+\item  propagation of uncertainty -- appropriate choice of \emph{anchor} for computing confidence intervals; and how to incorporate the uncertainty due to non-focal predictors 
 \item bias in the expected mean prediction induced by the nonlinear transformation of the response variable (especially in GL(M)Ms)
@@
 
-The most common way of dealing with the first challenge is taking unique levels of the focal (predictor of interest) predictor if discrete or taking appropriately sized quantiles (or bins) if continuous, and then calculating the predictions while holding non-focal (other predictors other than the focal predictor) predictors at their typical values (e.g., averages) \cite{hanmer2013behind}.
-This generates -- \emph{predictor effects}\cite{fox2009effect}, \emph{marginal predictions} \cite{leeper2017package} and \emph{estimated marginal means} \cite{lenth2018package}.
+The most common way of dealing with the first challenge is taking unique levels of the focal (predictor of interest) predictor if discrete or taking appropriately sized quantiles (or bins) if continuous, and then calculating the predictions while holding non-focal (other predictors other than the focal predictor) predictors at their reference point (e.g., means) \cite{hanmer2013behind}.
+This generates -- \emph{predictor effects}\cite{fox2009effect}, \emph{marginal predictions} \cite{leeper2017package} or \emph{estimated marginal means} \cite{lenth2018package}.
 In this article, we refer to this quantity as predictor effects since it should, for example, tell us what we would expect the presponse to be at a particular value or level of the predictor, for an ``average case''.
@@
 
-The commonly used \proglang{R} software packages (\pkg{emmeans} and \pkg{effects}) for generating predictor effects, by default, averages the non-focal predictors.
+The commonly used \proglang{R} software packages (\pkg{emmeans} and \pkg{effects}) for generating predictor effects, by default, use the average of the non-focal predictors as the reference point.
 However, there are a number of choices one can make when considering this approach -- for example, in the presence of interaction, averaging the interactions (averaging product of interacting predictors) versus product of the averages of the interacting predictors (default for \pkg{emmeans} and \pkg{effects}).
@@
 \caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} predictor effects for $x_1$ on $y$ for models with and without interactions.}
-The horizontal blue, green and black lines are the mean predictions, i.e., $\bar{\hat{y}}$, the red ones are the mean observed $y$, i.e., $\bar{y}$, while the vertical dotted grey lines are the means of the corresponding focal predictors.
+The horizontal blue, green and black lines are the mean predictions, i.e., $\bar{\hat{y}}$, the red ones are the means of the simulated $y$, i.e., $\bar{y}$, while the vertical dotted grey lines are the means of the corresponding focal predictors.
 The grey points are the binned simulated $y$.
 The trend lines represents the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other predictors at their average.
-A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the observed in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$.
-B: Even with the simple interaction between the non-focal predictors, we start seeing deviation of the predicted mean, $\bar{\hat{y}}$, from the observed, $\bar{y}$, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not, the proposed \pkg{varpred}.}
+A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the simulated in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$.
+B: Even with the simple interaction between the non-focal predictors, we start seeing deviation of the predicted mean, $\bar{\hat{y}}$, from the simulated, $\bar{y}$, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not, the proposed \pkg{varpred}.}
 \label{fig:justify_plots}
@@
 
-In the absence of interaction (model~\ref{eq:simple_inter_higher_no_interaction}), the three approaches produce similar estimates, which match the observed values, Fig~\ref{fig:justify_plots}A.
+In the absence of interaction (model~\ref{eq:simple_inter_higher_no_interaction}), the three approaches produce similar estimates, which match the simulated values, Fig~\ref{fig:justify_plots}A.
 However, in the presence of interaction, even as simple as the one in model~\ref{eq:simple_inter_higher}, the estimates starts to differ.
-In particular, \pkg{emmeans} and \pkg{effects} give similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is similar to the observed average ($\bar{y}$), Fig~\ref{fig:justify_plots}B.
+In particular, \pkg{emmeans} and \pkg{effects} give similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is very close to the simulated average ($\bar{y}$), Fig~\ref{fig:justify_plots}B.
 To generate Fig~\ref{fig:justify_plots}A, all the three packages averages the non-focal predictors $x_2$ and $x_3$.
@@
 
-In some applications, we may only be interested in the uncertainties associated with a particular focal predictor -- \emph{isolated} confidence intervals.
+Sometimes, one may be interested in the uncertainties associated with the focal predictor of interest only, excluding other uncertainties due to other non-focal predictors -- \emph{isolated} confidence intervals.
 However, currently, the two packages do not provide straightforward way do achieve this.
@@
 In other words, with \pkg{varpred}, we are able to generate isolated confidence bands indicating zero uncertainty at the value of the focal predictor we are more certain about, i.e., mean of the focal predictor.
-For simple models, the point where the confidence bands crosses is the model center and it corresponds to typical value we choose, in this case, the mean of the focal predictor.
+For simple models, the point where the confidence bands crosses is the model center and it corresponds to anchor we choose, in this case, the mean of the focal predictor.
 
@@
 In particular, bias in expected mean prediction induced by nonlinear transformation of the response variable can lead inaccurate predictions.
-An alternative to averaging the non-focal predictors, is the \emph{population-based} approach, discussed in detail later, which involves computing the prediction over the population of the non-focal predictors and then averaging across the values of the focal predictor \cite{hanmer2013behind}.
+An alternative to mean-based reference point, i.e., averaging the non-focal predictors, is the \emph{population-based} approach, discussed in detail later, which involves computing the prediction over the population of the non-focal predictors and then averaging across the values of the focal predictor \cite{hanmer2013behind}.
 
@@
 On the other hand, the marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher} is given by $\frac{\partial y}{\partial x_2} = \beta_2 + \beta_{23}x_3$.
-In other words, if there are no interactions, the marginal effect of $x_2$ on $y$ is constant, while, if there are interactions in the model, the marginal effect of a change in $x_2$ on $y$ depends on the value of the other \emph{conditioning} predictor $x_3$.
-In general, for a continuous variable, $x_k$, the marginal effect is $\frac{\partial y}{\partial x_k} = g(x\beta)\beta_k$, where $g()$ is the probability density function.
+In other words, if there are no interactions, the marginal effect of $x_2$ on $y$ is constant, while, if there are interactions in the model, the marginal effect of a change in $x_2$ on $y$ depends on the value of the other \emph{conditioning} predictor, $x_3$.
+
+% In general, for a continuous variable, $x_k$, the marginal effect is $\frac{\partial y}{\partial x_k} = g(x\beta)\beta_k$, where $g()$ is the probability density function.
 This implies that, in addition to coefficient estimates, the marginal effect depends on the nature of $g()$ and the values of the other non-focal predictors.
@@
 Predictor effect, on the other hand, is the expected household size for a particular age, holding wealth index at its mean.
-In particular, the purpose and goal of a predictor effect seems fairly straightforward; for specified values of a focal predictor, we want to give a point estimate and confidence intervals for the prediction of the model for a typical individual with those values of the predictors.
+In particular, the purpose and goal of a predictor effect seems fairly straightforward; for specified values of a focal predictor, we want to give a point estimate and confidence intervals for the prediction of the model for a ``reference point'' individual with those values of the predictors.
 
@@
 To keep it simple, assume that the model has no interaction terms.
-The idea is to \emph{anchor} the values of non-focal predictors to some particular values.
+The idea is to choose a reference point for the values of non-focal predictors.
 For example, fixing the values of non-focal predictor(s) at some typical values -- typically determined by averaging (for now) in some meaningful way, for example, arithmetic mean  for continuous and average over the levels of the factors for categorical non-focal predictors.
-We refer to this as \emph{mean-anchored} approach.
+We refer to this as \emph{mean-based} approach.
 The most convenient way to achieve this is by constructing $\bX^\star$ by averaging the columns of non-focal predictors in model matrix $\bX$, and together with appropriately chosen values of focal predictor.
@@
 
-An alternative, save for later, formulation of predictor effect involves, expressing the linear predictor as the sum of the focal and non-focal predictor linear predictor. In particular,
+An alternative formulation of predictor effect involves, expressing the linear predictor as the sum of the focal and non-focal predictors' linear predictor. In particular,
 
 \begin{align}\label{eq:eta_mean}
-\eta^\star(x_f, \nset{{\bar{x}}}) &= \beta_f x_f + \sum \nset{\beta} \nset{{\bar{x}^\star}} \\
-\hat{y}_f  &= g^{-1} \left(\eta^\star(x_f, \nset{{\bar{x}}})\right)
+\eta^\star(x_f, \nset{{\bar{x}^\star}}) &= \beta_f x_f + \sum \nset{\beta} \nset{{\bar{x}^\star}} \\
+\hat{y}_f  &= g^{-1} \left(\eta^\star(x_f, \nset{{\bar{x}^\star}})\right)
 \end{align}
@@
 
-Higher order terms such as interactions, splines, polynomials, etc., can be between the non-focal predictors or focal and non-focal predictor(s).
-In the former case, we treat the interactions as just another column in the variable space (of the model matrix).
-In the later case, the non-focal variables in the model matrix are averaged as before.
-However, for the interacting focal predictors, a combination of each unique levels (or quantiles) are first generated and then the interaction terms are generated by multiplying these combinations.
-For example, consider model~\ref{eq:simple_inter_higher} above.
-In the first case, if we consider $x_1$ as the focal predictor, the interaction ($x_2x_3$) is between the non-focal predictors, $x_2$ and $x_3$, the linear predictor is
+Higher order terms such as interactions, splines, polynomials, etc., can be within the focal predictor, within (between) non-focal predictor(s), between the focal and non-focal predictor(s).
+To distinguish the three, suppose the model which describes the hypothetical household size simulation described above is
+
+\begin{align}\label{sim:lm_cubic}
+\mathrm{hh~size}_i &= \beta_0 + \beta_{\mathrm{A_1}}\mathrm{Age}_i + \beta_{\mathrm{A_2}}\mathrm{Age}^2_i + \beta_{\mathrm{A_3}}\mathrm{Age}^3_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \epsilon_i \nonumber\\
+\end{align}
+
+In the first case, with $Age$ as the focal predictor, the interaction is within its polynomial terms.
+In this case, each of the polynomial terms of the focal predictor are evaluated independently across the chosen levels of the focal predictor, $Age_i$.
+Specifically, the higher order terms are treated as additional columns of the model matrix evaluated with the same values of the focal predictor, while the non-focal predictors are fixed at their reference point as discussed in the previous section.
+In this case, the predictor effect of $Age$ on the linear scale becomes
 
 \begin{align*}
-\eta^\star(x_{1i}, \nset{{\bar{x}^\star}}) = \beta_0\mathbf{1} + \beta_1 x_{1i} + \beta_2\bar{x}_2 + \beta_3\bar{x}_3 + \beta_{23}\over{x_2x_3}
+\eta^\star(\mathrm{Age}_i, \nset{{\over{\mathrm{Wealthindex}}}}) &= \beta_0 + \beta_{\mathrm{A_1}}\mathrm{Age}_i + \beta_{\mathrm{A_2}}\mathrm{Age}^2_i + \beta_{\mathrm{A_3}}\mathrm{Age}^3_i\\
+	& + \beta_{\mathrm{W}}\over{\mathrm{Wealthindex}}
 \end{align*}
-where $x_{1i}$ are the carefully chosen levels of the focal predictor.
-On the other hand, for the second case, if we consider $x_2$ as the focal predictor, then the interaction ($x_2x_3$) is between focal and non-focal predictor.
-In this case, the linear predictor is given by
+
+In the second case, with $Wealthindex$ as the focal predictor, the higher order interactions in the non-focal predictor, $Age$, are treated simply as additional columns in the variable space (of model matrix) and appropriate choice of reference point applies just like in the models without interactions.
+For instance, in mean-based approach, we simply average all the non-focal interaction terms.
+In this setup, the predictor effect of the previous model, on link scale, becomes
 
 \begin{align*}
-\eta^\star(x_{2ij}, \nset{{\bar{x}^\star}}) = \beta_0\mathbf{1} + \beta_1 \bar{x}_1 + \beta_2x_{2ij} + \beta_3x_{3j} + \beta_{23}x_{2ij}x_{3j}
+\eta^\star(\mathrm{Wealthindex}_i, \nset{{\{\over{\mathrm{Age}}, \over{\mathrm{Age}^2}, \over{\mathrm{Age}^2}\}}}) &= \beta_0 + \beta_{\mathrm{A_1}}\over{\mathrm{Age}} + \beta_{\mathrm{A_2}}\over{\mathrm{Age}^2} + \beta_{\mathrm{A_3}}\over{\mathrm{Age}^3}\\
+	& + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i
 \end{align*}
 
-In general, our formulation, even for more complicated interactions, follow these two basic principles -- interaction between non-focal predictors and interaction between focal and non-focal predictors.
+Lastly, consider model~\ref{eq:simple_inter_higher}, with the interactions between the focal predictor, $x_2$, and the non-focal predictor, $x_3$.
+In this case, the non-interacting predictors are treated as non-focal predictors, with appropriate reference points.
+However, for the interacting predictors, we first choose representative values for the focal predictor and some particular values of the interacting predictor similar to the way we choose values for the focal predictor or predetermined.
+The compute the possible combination of all the chosen levels of the focal predictor with chose or predetermined levels of the interacting predictor.
+In our example, suppose we pick $k$ unique values of the focal predictor, $x_2$, then
+
+\begin{align*}
+\eta^\star(x_{2[1:k,j]}, \nset{{\bar{x}^\star}}) = \beta_0 + \beta_1 \bar{x}_1 + \beta_2x_{2[1:k,j]} + \beta_3x_{3[j,1:k]} + \beta_{23}x_{2[1:k,j]}x_{3[1:k,j]}
+\end{align*}
+where $j = 1, 2, \cdots$ are the indices of the unique values of the interacting predictor, $x_3$, $x_{2[1:k,j]}$, for example, are the values of the focal predictor corresponding to the $j$th interacting predictor.
+
+In general, our formulation, even for more complicated interactions, follow these three basic principles -- interaction within focal predictor, interaction between non-focal predictors and interaction between focal and non-focal predictors.
 
@@
 
-Currently, commonly used \proglang{R} packages for constructing predictions do not exclude the uncertainties resulting from the non-focal predictors when computing the CIs.
+Currently, commonly used \proglang{R} packages for constructing predictions, by default, do not exclude the uncertainties resulting from the non-focal predictors when computing the CIs.
 A non-trivial way to exclude uncertainties associated with non-focal predictors in some of these packages is to provide a user defined variance-covariance matrix with the covariances of non-focal terms set to $0$ -- \emph{zeroing-out} variance-covariance matrix.
@@
 The most precise (although not necessarily accurate!) way to predict is to condition on values of the focal predictor and make predictions for all observations (members of the population) \cite{hanmer2013behind}.
-A key point is that the nonlinear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population to a one-dimensional distribution of $\eta^\star(x_f, \nset{x})$, which is a function of focal predictor and the observed values of the non-focal predictors, as oposed to the definition in Equation~\ref{eq:eta_mean}.
+A key point is that the nonlinear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population to a one-dimensional distribution of $\eta^\star(x_f, \nset{x})$, which is a function of focal predictor and the observed values of the non-focal predictors, as opposed to the definition in Equation~\ref{eq:eta_mean}.
 More specifically:
@@
 
-\section*{Mean-anchored vs population-based}
+\section*{Mean-based vs population-based}
 
-As mentioned above, the two common choices to generate predictor effects are: 1) setting the non-focal predictors to their mean -- mean-anchored; and 2) using the entire population of the non-focal predictors -- population-based.
+As mentioned above, the two common choices to generate predictor effects are: 1) setting the non-focal predictors to their mean -- mean-based; and 2) using the entire population of the non-focal predictors -- population-based.
 If the link function is nonlinear and/or the model has complex higher order interactions, the average of the predictions evaluated at the mean of the non-focal predictors and the average of predictions evaluated at the population level (and then averaged) of the non-focal predictor are not equivalent, i.e.,
@@
 However, picking a single value, e.g., mean of the predictor, on which to draw conclusions about the effect can be problematic, unrealistic or not contained in or representative of the population.
-In addition, the mean-anchored approach fails to use the every values of the non-focal predictors hence not utilizing the full potential of the information contained in the data.
+In addition, the mean-based approach fails to use every values of the non-focal predictors hence not utilizing the full potential of the information contained in the data.
 This may limit the inferences we can make about the entire population.
-In general, the mean-anchored approach provides the predictor effect of an average case, whereas, the population-based approach, summarizes the predictor effect over the entire population -- and in some applications, the effect of an average case might not be generalizable to the entire population, especially, if the average does not represent the population.
+In general, the mean-based approach provides the predictor effect of an average case, whereas, the population-based approach, summarizes the predictor effect over the entire population -- and in some applications, the effect of an average case might not be generalizable to the entire population, especially, if the average does not represent the population.
 The population-based focuses on specific observations since the prediction is first obtained for each observation and then averaged across the levels of the focal predictor.
 
-Another potential concern with the mean-anchored approach arises in situations where direct naive use leads to rare or meaningless basis for generalization.
+Another potential concern with the mean-based approach arises in situations where direct naive use leads to rare or meaningless basis for generalization.
 For example, setting categorical dummy categorical variables in the model matrix to their means, which, by default, sets them to their sample means or observed proportions.
@@
 Population-based approach is not entirely foolproof.
-For instance, similar to the mean-anchored approach, in the case of continuous focal predictors, choosing the representative values of the focal predictors can be very challenging especially if the cases are not evenly distributed around the minimum and the maximum values or within some subgroups defined in the population.
+For instance, similar to the mean-based approach, in the case of continuous focal predictors, choosing the representative values of the focal predictors can be very challenging especially if the cases are not evenly distributed around the minimum and the maximum values or within some subgroups defined in the population.
 In addition, population-based approach can be very computationally intensive for large datasets.
@@
 
-We now illustrate the construction of predictor effects with isolated confidence intervals and also demonstrate that the mean-anchored and population approaches produce different results in models with nonlinear link functions and/or additional source of potential bias.
+We now illustrate the construction of predictor effects with isolated confidence intervals and also demonstrate that the mean-based and population approaches produce different results in models with nonlinear link functions and/or additional source of potential bias.
 In addition, we demonstrate that population-based approach can be used to correct bias in the predictions.
@@
 
-Mean-anchored and population-based approaches can produce very different results in models with nonlinear link functions with additional sources of potential bias such as additional non-focal predictor(s), random effect terms or complex interactions.
+Mean-based and population-based approaches can produce very different results in models with nonlinear link functions with additional sources of potential bias such as additional non-focal predictor(s), random effect terms or complex interactions.
 To demonstrate this, we considered a two predictor binary outcome simulation described in \nameref{S2_Appendix}, such that the age effect is way smaller than that of the wealth index, and compared their predictor effect on the predicted probability of improved water quality as shown in Fig~\ref{fig:pred_bin_plots}.
@@
 \includegraphics[width=0.8\textwidth]{pred_bin_plots-figure5.pdf}
-\caption{{\bf Mean-anchored and population-based predictor effects.} The two approaches produce relatively different predictor effects for each of the predictors but the difference is more pronounced when the effect of the particular focal predictor has no strong effect -- \emph{on the left}.
+\caption{{\bf Mean-based and population-based predictor effects.} The two approaches produce relatively different predictor effects for each of the predictors but the difference is more pronounced when the effect of the particular focal predictor has no strong effect -- \emph{on the left}.
 In both cases, age and wealth index effect, the average predicted probability of improved water based on population-based approach is very close to the observed proportion in the simulated data.
-The horizontal blue, black and the red dotted lines are the average mean-anchored, population-based predicted probability and observed proportion of improved water quality, respectively.
+The horizontal blue, black and the red dotted lines are the average mean-based, population-based predicted probability and observed proportion of improved water quality, respectively.
 The vertical dotted black line represents the mean of the focal predictor, and the point at which it crosses the red dotted line represents the expected ``perfect'' prediction at the model center.
@@
 We then fitted two models -- \emph{non-mediated} which models \emph{z} as a function of \emph{x} and the \emph{mediated} which models \emph{z} as a function of both \emph{x} and \emph{y}.
-In both cases, we compared mean-anchored and the population-based predictions, as shown in Fig~\ref{fig:pred_mediated_plots}.
+In both cases, we compared mean-based and the population-based predictions, as shown in Fig~\ref{fig:pred_mediated_plots}.
 
@@
 \includegraphics[width=0.8\textwidth]{pred_mediated_plots-figure6.pdf}
-\caption{{\bf Mean-anchored and population-based predictor effects for mediated effect.} These figures compare the mean-anchored and population-based predictor effect of $x$ from a mediated, $z \sim x + y$, and non-mediated, $z \sim x$, models.
-A: In the absence of mediator variable, both mean-anchored and population-based approaches predictors are identical since there is no any other additional (non-focal predictors) sources of bias.
+\caption{{\bf Mean-based and population-based predictor effects for mediated effect.} These figures compare the mean-based and population-based predictor effect of $x$ from a mediated, $z \sim x + y$, and non-mediated, $z \sim x$, models.
+A: In the absence of mediator variable, both mean-based and population-based approaches predictors are identical since there is no any other additional (non-focal predictors) sources of bias.
 Consequently, in both approaches, the average predicted probability is very close to the observed proportion.
@@
 However, the population-based approach still closely approximates the average proportion in the simulated data.
-The horizontal blue, black and the red dotted lines are the average mean-anchored, population-based predicted probability and observed proportion, respectively.
+The horizontal blue, black and the red dotted lines are the average mean-based, population-based predicted probability and observed proportion, respectively.
 The vertical dotted black line represents the mean of the focal predictor, and the point at which it crosses the red dotted line represents the expected ``perfect'' prediction at the model center.
@@
 
-Although commonly used \proglang{R} software packages, by default, implements mean-anchored approach, our simulation results demonstrated that the population-based approach has a potential of yielding results which are more consistent with the observed data.
-We would, therefore, argue that the use of mean-anchored approach should have some theoretical justification, especially in complex models.
+Although commonly used \proglang{R} software packages, by default, implements mean-based approach, our simulation results demonstrated that the population-based approach has a potential of yielding results which are more consistent with the observed data.
+We would, therefore, argue that the use of mean-based approach should have some theoretical justification, especially in complex models.
 

### 2022-02-08 manuscript/glm_two_predictor_preds.R

@@ -13,7 +13,7 @@ startGraphics()
 pred_age_none <- varpred(glm_mod
 	, "age"
 	, bias.adjust="none"
-	, modelname="Mean-anchored"
+	, modelname="Mean-based"
 )
 pred_age_none_mean <- getmeans(pred_age_none, what="estimate")
 
@@ -42,10 +42,10 @@ pred_age_plots <- (comparevarpred(vlist=vlist
 	+ geom_hline(data=pred_age_pop_mean, aes(yintercept=fit, colour=model, lty=model))
 	+ geom_hline(data=true_prop_df, aes(yintercept=status, colour=model,lty=model))
 	+ geom_point(data=binned_df, aes(x=age, y=status), colour="grey")
-	+ scale_colour_manual(breaks = c("Observed mean", "Mean-anchored", "Population-based")
-		, values=c("Observed mean"="red", "Mean-anchored"="blue", "Population-based"="black")
+	+ scale_colour_manual(breaks = c("Observed mean", "Mean-based", "Population-based")
+		, values=c("Observed mean"="red", "Mean-based"="blue", "Population-based"="black")
 	)
-	+ scale_linetype_manual(values=c("Observed mean"=2, "Mean-anchored"=1, "Population-based"=1))
+	+ scale_linetype_manual(values=c("Observed mean"=2, "Mean-based"=1, "Population-based"=1))
 	+ labs(y="Probability of improved water quality", colour="Method", linetype="Method")
 	+ theme(legend.position="bottom")
 )
@@ -56,7 +56,7 @@ pred_age_plots <- (comparevarpred(vlist=vlist
 pred_wealthindex_none <- varpred(glm_mod
 	, "wealthindex"
 	, bias.adjust="none"
-	, modelname="Mean-anchored"
+	, modelname="Mean-based"
 )
 pred_wealthindex_none_mean <- getmeans(pred_wealthindex_none, what="estimate")
 
@@ -85,10 +85,10 @@ pred_wealthindex_plots <- (comparevarpred(vlist=vlist
 	+ geom_hline(data=pred_wealthindex_pop_mean, aes(yintercept=fit, colour=model, lty=model))
 	+ geom_hline(data=true_prop_df, aes(yintercept=status, colour=model,lty=model))
 	+ geom_point(data=binned_df, aes(x=wealthindex, y=status), colour="grey")
-	+ scale_colour_manual(breaks = c("Observed mean", "Mean-anchored", "Population-based")
[…]

### 2022-02-08 manuscript/justify.R

@@ -9,7 +9,7 @@ set.seed(9991)
 
 ############################################################################################
 # Global variables
-nHH_obs <- 100
+nHH_obs <- 30
 perHH <- 1
 
 ############################################################################################

### 2022-02-08 manuscript/mediate_preds.R

@@ -13,7 +13,7 @@ startGraphics(height = 5.3)
 pred_notmediated_none <- varpred(mod_notmediated
 	, "x"
 	, bias.adjust="none"
-	, modelname="Mean-anchored"
+	, modelname="Mean-based"
 )
 pred_notmediated_none_mean <- getmeans(pred_notmediated_none, what="estimate")
 
@@ -47,10 +47,10 @@ pred_notmediated_plots <- (comparevarpred(vlist=vlist
 	+ geom_hline(data=observed_df_med, aes(yintercept=z, colour=model, lty=model))
 	+ geom_vline(data=observed_df_med, aes(xintercept=x), lty=2, col="black")
 	+ geom_point(data=binned_df, aes(x=x, y=z), colour="grey")
-	+ scale_colour_manual(breaks = c("Observed mean", "Mean-anchored", "Population-based")
-		, values=c("Observed mean"="red", "Mean-anchored"="blue", "Population-based"="black")
+	+ scale_colour_manual(breaks = c("Observed mean", "Mean-based", "Population-based")
+		, values=c("Observed mean"="red", "Mean-based"="blue", "Population-based"="black")
 	)
-	+ scale_linetype_manual(values=c("Observed mean"=2, "Mean-anchored"=1, "Population-based"=1))
+	+ scale_linetype_manual(values=c("Observed mean"=2, "Mean-based"=1, "Population-based"=1))
 	+ labs(colour="Method", linetype="Method", title="A) Non-mediated", y="Predicted probability")
 	+ theme(legend.position="bottom")
 )
@@ -61,7 +61,7 @@ pred_notmediated_plots <- (comparevarpred(vlist=vlist
 pred_mediated_none <- varpred(mod_mediated
 	, "x"
 	, bias.adjust="none"
-	, modelname="Mean-anchored"
+	, modelname="Mean-based"
 )
 pred_mediated_none_mean <- getmeans(pred_mediated_none, what="estimate")
 
@@ -90,10 +90,10 @@ pred_mediated_plots <- (comparevarpred(vlist=vlist
 	+ geom_hline(data=observed_df_med, aes(yintercept=z, colour=model, lty=model))
 	+ geom_vline(data=observed_df_med, aes(xintercept=x), lty=2, col="black")
 	+ geom_point(data=binned_df, aes(x=x, y=z), colour="grey")
-	+ scale_colour_manual(breaks = c("Observed mean", "Mean-anchored", "Population-based")
[…]

### 2022-02-10 manuscript/vareffects_plosone.tex
messing around

@@
 % Version 3.5 March 2018
-%
+
 % % % % % % % % % % % % % % % % % % % % % %
@@
 % % % % % % % % % % % % % % % % % % % % % % %
-%
+
 % Once your paper is accepted for publication, 
@@
 We also present an alternative way, isolated confidence intervals, to describe uncertainties associated with these predictions.
+
+\section*{Glossary}
+
+\begin{itemize}
+\item \textbf{Input variables}: These are measured variables (and may be transformed)
+\item \textbf{Predictors}: These are statistical variables i.e., those terms which are entered into the model. They include the main effects and also polynomials of input variables
+\item \textbf{Anchor}: The value of the focal variable (input or predictor) used to calculate CIs. Common choice is the average of the focal predictor
+\item \textbf{Reference point}: Values chosen for the non-focal predictors, and can include population averages, quantiles or population values
+\end{itemize}
 

### 2022-02-24 manuscript/jd_notes.Rmd
research check

---
title: JD notes on effects
---

## Language

It might be better to say focal variable, or focal input variable, otherwise we are conflating predictor with something we use to make a prediction. I'm also thinking it might be good to just say "variable" for input variable, but in this case we need yet another name for the statistical variables.

I'm worried right now about "predictor" for "statistical variable", since we then say "focal predictor" to refer to an input variable instead. How about "variable" (sometimes "input variable") for the science and "vector" (sometimes "input vector") for the stats – an input variable may correspond to more than one input vector

## Predictions vs. effects

I'd like to start by talking about predictions vs. effects. Predictions are the standard curves; they tell us what we expect to see if we know the value of the focal predictor.

The lines inside (effects) are a little different: they are a good way to visualize the effect of the focal predictor. Unlike predictions, where the uncertainty captures the uncertainty in the whole model, the effects plot focuses on uncertainty in the effect of the focal predictor only -- thus, for example, if an effect is barely significant (P exactly 0.05), one end of the CI will be exactly a straight line for the effects plot, but not for the prediction plot. Also, the uncertainty curves for the effect, but not for the prediction, depend on the "anchor".

It is interesting, I think, that all of this comes down to choosing a reference, point, an anchor, and then whether or not to include uncertainty in the intercept (so that all of the uncertainty from all of the other variables in the model is somehow channeled through the intercept).

### 2022-03-09 manuscript/vareffects_plosone.tex
Adding glossary
Working on definitions.

--- 2022-02-08 manuscript/vareffects_plosone.tex
@@
 % Header and Footer with logo
-\usepackage{lastpage,fancyhdr,graphicx}
-\usepackage{epstopdf}
+\usepackage{lastpage,fancyhdr,graphicx,grffile}
 %\pagestyle{myheadings}
@@
 
+\section*{Definitions}
+
+\begin{itemize}
+\item \textbf{Input variables}: Scientific variables underlying an inference or exploration.
+\item \textbf{Model variables}: Variables that represent columns in the model matrix.
+Each input variable will correspond to one or more model variables.
+In particular, variables with more than two categories, or variables modeled with a spline or polynomial response, will correspond to more than one input variable.
+\item\textbf{Center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more model variables).
+The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable.
+\item \textbf{Reference point}: Value or values chosen for \emph{non-focal predictors}, when estimating effects.
+Typically the center point, but can instead be a population of quantiles or observations.
+\item \textbf{Anchor}: The value chosen for the \emph{focal predictor}. Typically chosen as the center point.
+\end{itemize}
+
 % Please keep the Author Summary between 150 and 200 words
@@
 \centering
-\includegraphics[width=0.8\textwidth]{justify_plots-figure1.pdf}
+\includegraphics[width=0.9\textwidth]{justify_preds.inter.pdf}
 \caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} predictor effects for $x_1$ on $y$ for models with and without interactions.}
@@
 \centering
-\includegraphics[width=0.8\textwidth]{justify_ci_plots-figure2.pdf}
+\includegraphics[width=0.9\textwidth]{justify_preds.isolate.pdf}
 \caption{{\bf A comparison of traditional and isolated $95\%$ confidence bands.} The description of horizontal, vertical and trend lines remain the same as above.

### 2022-03-09 manuscript/vareffects_plosone.tex
Working with Bicko during an alarm

@@
 \begin{itemize}
-\item \textbf{Input variables}: Scientific variables underlying an inference or exploration.
+\item \textbf{Input variables}: Scientific variables underlying an inference or exploration. The focal predictor we use for effects or predictions is an input variable.
 \item \textbf{Model variables}: Variables that represent columns in the model matrix.
@@
 Typically the center point, but can instead be a population of quantiles or observations.
-\item \textbf{Anchor}: The value chosen for the \emph{focal predictor}. Typically chosen as the center point.
+\item \textbf{Anchor}: The value chosen for the \emph{focal predictor} when estimating effect confidence intervals (anchor choice does not affect the estimate).
+Typically chosen as the center point of the model variables corresponding to the focal input variable.
 \end{itemize}

### 2022-03-12 manuscript/justify_preds.R
Autocommit

@@ -13,7 +13,8 @@ startGraphics()
 quants <- seq(0,1,length.out=100)
 x1_focal <- quantile(justify_sim_df$x1, quants, names=FALSE)
 binned_df <- binfun(justify_mod, focal="x1", bins=20, groups=NULL)
-x1_truepred_df <- truepredfun(justify_mod, "x1", true_betas=justify_sim_betas, modelname="observed")
+## Preds based on true parameter values
+x1_truepred_df <- varpred(justify_mod, "x1", true.bata=justify_sim_betas, bias.adjust="population", modelname="observed")$preds
 
 justify_plots <- (combinepreds(justify_mod
 		, c("varpred", "emmeans", "Effect")
@@ -40,8 +41,9 @@ justify_plots <- (combinepreds(justify_mod
 )
 
 ## Add CIs
+### Preds based on true parameter values
+x2_truepred_df <- varpred(justify_mod, "x2", true.beta=justify_sim_betas, bias.adjust="population", modelname="observed")$preds
 x2_focal <- quantile(justify_sim_df$x2, quants, names=FALSE)
-x2_truepred_df <- truepredfun(justify_mod, "x2", true_betas=justify_sim_betas, modelname="observed")
 justify_ci_plots <- (combinepreds(justify_mod
 		, c("varpred", "emmeans", "Effect")
 		, focal="x2"
@@ -74,7 +76,7 @@ dev.off()
 quants <- seq(0,1,length.out=100)
 x1_focal <- quantile(justify_inter_sim_df$x1, quants, names=FALSE)
 binned_df <- binfun(justify_inter_mod, focal="x1", bins=20, groups=NULL)
-x1_truepred_df <- truepredfun(justify_inter_mod, "x1", true_betas=justify_inter_sim_betas, modelname="observed")
+x1_truepred_df <- varpred(justify_inter_mod, "x1", true.beta=justify_inter_sim_betas, bias.adjust="population", modelname="observed")$preds
 
 justify_inter_plots <- (combinepreds(justify_inter_mod
 		, c("varpred", "emmeans", "Effect")

### 2022-03-12 manuscript/mediate.R

@@ -11,14 +11,20 @@ set.seed(991)
 
 N <- 1e4
 
+rho <- 0.8
 b0 <- 5
-bxz <- 0
-bxy <- 1
+bxz <- 0.2
 byz <- 1.5
 
 df <- data.frame(x=rnorm(N))
+#sim_df_mediate <- (df
+#	%>% mutate(y = rnorm(N) + bxy*x
+#		, z = b0 + bxz*x + byz*y
+#		, z = rbinom(N, 1, plogis(z))
+#	)
+#)
 sim_df_mediate <- (df
-	%>% mutate(y = rnorm(N) + bxy*x
+	%>% mutate(y = rho*x + sqrt(1-rho^2)*rnorm(N)
 		, z = b0 + bxz*x + byz*y
 		, z = rbinom(N, 1, plogis(z))
 	)

### 2022-03-12 manuscript/mediate_preds.R

@@ -98,13 +98,13 @@ pred_mediated_plots <- (comparevarpred(vlist=vlist
 	+ theme(legend.position="bottom")
 )
 
-pred_mediate_plots <- ggarrange(pred_notmediated_plots
+pred_mediated_plots <- ggarrange(pred_notmediated_plots
 	, pred_mediated_plots + rremove("ylab")
 	, common.legend=TRUE
 	, legend="bottom"
 	, ncol=2
 )
-print(pred_mediate_plots)
+print(pred_mediated_plots)
 
 ## Figure 6
 pdf("pred_mediated_plots-figure6.pdf", height = 5.3)

### 2022-03-13 manuscript/jd_notes.Rmd

@@
 The lines inside (effects) are a little different: they are a good way to visualize the effect of the focal predictor.
-Unlike predictions, where the uncertainty captures the uncertainty in the whole model, the effects plot focuses on uncertainty in the effect of the focal predictor only -- thus, for example, if an effect is barely significant (P exactly 0.05), one end of the CI will be exactly a straight line for the effects plot, but not for the prediction plot.
+Unlike predictions, where the uncertainty captures the uncertainty in the whole model, the effects plot focuses on uncertainty in the effect of the focal predictor only -- thus, for example, if an effect is barely significant (P exactly 0.05), one end of the CI will be exactly a horizontal line for the effects plot, but not for the prediction plot.
 Also, the uncertainty curves for the effect, but not for the prediction, depend on the "anchor".

### 2022-03-23 manuscript/vareffects_plosone.tex
Autocommit
population-based > sample-based

@@
 In the presence of sources of bias such as additional non-focal predictors, nonlinear link functions, random effect terms, etc., mean-based approach generate predictor effects that are biased and may not be consistent with the observed quantities.
-An alternative is the population-based approach which estimates the average effect in the population.
+An alternative is the whole-sample-based approach which estimates the average effect in the population.
 Moreover, isolated confidence intervals provide an alternative and a more clear way to describe uncertainty associated with the focal predictor of interest.
-In addition to theoretical and methodical comparison, using simulation, we illustrate the two approaches and show that they can produce substantially different results and that population-based approach can not only produce estimates consistent with the observed values, but also appropriate for bias correction.
+In addition to theoretical and methodical comparison, using simulation, we illustrate the two approaches and show that they can produce substantially different results and that whole-sample-based approach can not only produce estimates consistent with the observed values, but also appropriate for bias correction.
 We also present an alternative way, isolated confidence intervals, to describe uncertainties associated with these predictions.
@@
 In particular, bias in expected mean prediction induced by nonlinear transformation of the response variable can lead inaccurate predictions.
-An alternative to mean-based reference point, i.e., averaging the non-focal predictors, is the \emph{population-based} approach, discussed in detail later, which involves computing the prediction over the population of the non-focal predictors and then averaging across the values of the focal predictor \cite{hanmer2013behind}.
+An alternative to mean-based reference point, i.e., averaging the non-focal predictors, is the \emph{whole-sample-based} approach, discussed in detail later, which involves computing the prediction over the population of the non-focal predictors and then averaging across the values of the focal predictor \cite{hanmer2013behind}.
 
@@
 Another potential source of bias comes from additional non-focal predictors, especially in models with nonlinear link functions.
-Here, we describe and implement a different approach, population-based approach for bias correction.
+Here, we describe and implement a different approach, whole-sample-based approach for bias correction.
 
@@
 
-\section*{Mean-based vs population-based}
+\section*{Mean-based vs whole-sample-based}
 
-As mentioned above, the two common choices to generate predictor effects are: 1) setting the non-focal predictors to their mean -- mean-based; and 2) using the entire population of the non-focal predictors -- population-based.
+As mentioned above, the two common choices to generate predictor effects are: 1) setting the non-focal predictors to their mean -- mean-based; and 2) using the entire population of the non-focal predictors -- whole-sample-based.
 If the link function is nonlinear and/or the model has complex higher order interactions, the average of the predictions evaluated at the mean of the non-focal predictors and the average of predictions evaluated at the population level (and then averaged) of the non-focal predictor are not equivalent, i.e.,
@@
 This may limit the inferences we can make about the entire population.
-In general, the mean-based approach provides the predictor effect of an average case, whereas, the population-based approach, summarizes the predictor effect over the entire population -- and in some applications, the effect of an average case might not be generalizable to the entire population, especially, if the average does not represent the population.
-The population-based focuses on specific observations since the prediction is first obtained for each observation and then averaged across the levels of the focal predictor.
+In general, the mean-based approach provides the predictor effect of an average case, whereas, the whole-sample-based approach, summarizes the predictor effect over the entire population -- and in some applications, the effect of an average case might not be generalizable to the entire population, especially, if the average does not represent the population.
+The whole-sample-based focuses on specific observations since the prediction is first obtained for each observation and then averaged across the levels of the focal predictor.
 
@@
 For instance, similar to the mean-based approach, in the case of continuous focal predictors, choosing the representative values of the focal predictors can be very challenging especially if the cases are not evenly distributed around the minimum and the maximum values or within some subgroups defined in the population.
-In addition, population-based approach can be very computationally intensive for large datasets.
+In addition, whole-sample-based approach can be very computationally intensive for large datasets.
 
@@
 We now illustrate the construction of predictor effects with isolated confidence intervals and also demonstrate that the mean-based and population approaches produce different results in models with nonlinear link functions and/or additional source of potential bias.
-In addition, we demonstrate that population-based approach can be used to correct bias in the predictions.
+In addition, we demonstrate that whole-sample-based approach can be used to correct bias in the predictions.
 
@@
 
-Mean-based and population-based approaches can produce very different results in models with nonlinear link functions with additional sources of potential bias such as additional non-focal predictor(s), random effect terms or complex interactions.
+Mean-based and whole-sample-based approaches can produce very different results in models with nonlinear link functions with additional sources of potential bias such as additional non-focal predictor(s), random effect terms or complex interactions.
 To demonstrate this, we considered a two predictor binary outcome simulation described in \nameref{S2_Appendix}, such that the age effect is way smaller than that of the wealth index, and compared their predictor effect on the predicted probability of improved water quality as shown in Fig~\ref{fig:pred_bin_plots}.
@@
 \includegraphics[width=0.8\textwidth]{pred_bin_plots-figure5.pdf}
-\caption{{\bf Mean-based and population-based predictor effects.} The two approaches produce relatively different predictor effects for each of the predictors but the difference is more pronounced when the effect of the particular focal predictor has no strong effect -- \emph{on the left}.
-In both cases, age and wealth index effect, the average predicted probability of improved water based on population-based approach is very close to the observed proportion in the simulated data.
-The horizontal blue, black and the red dotted lines are the average mean-based, population-based predicted probability and observed proportion of improved water quality, respectively.
+\caption{{\bf Mean-based and whole-sample-based predictor effects.} The two approaches produce relatively different predictor effects for each of the predictors but the difference is more pronounced when the effect of the particular focal predictor has no strong effect -- \emph{on the left}.
+In both cases, age and wealth index effect, the average predicted probability of improved water based on whole-sample-based approach is very close to the observed proportion in the simulated data.
+The horizontal blue, black and the red dotted lines are the average mean-based, whole-sample-based predicted probability and observed proportion of improved water quality, respectively.
 The vertical dotted black line represents the mean of the focal predictor, and the point at which it crosses the red dotted line represents the expected ``perfect'' prediction at the model center.
 The grey points are binned observations -- observed proportions of improved water quality in each bin.
-We would expect all the curves, horizontal, and vertical lines to cross at the model center but due the nonlinear averaging (at least in the bias corrected population-based approach) of the predictions on the response scale, this may not always be the case.}
+We would expect all the curves, horizontal, and vertical lines to cross at the model center but due the nonlinear averaging (at least in the bias corrected whole-sample-based approach) of the predictions on the response scale, this may not always be the case.}
 \label{fig:pred_bin_plots}
@@
 We then fitted two models -- \emph{non-mediated} which models \emph{z} as a function of \emph{x} and the \emph{mediated} which models \emph{z} as a function of both \emph{x} and \emph{y}.
-In both cases, we compared mean-based and the population-based predictions, as shown in Fig~\ref{fig:pred_mediated_plots}.
+In both cases, we compared mean-based and the whole-sample-based predictions, as shown in Fig~\ref{fig:pred_mediated_plots}.
 
@@
 \includegraphics[width=0.8\textwidth]{pred_mediated_plots-figure6.pdf}
-\caption{{\bf Mean-based and population-based predictor effects for mediated effect.} These figures compare the mean-based and population-based predictor effect of $x$ from a mediated, $z \sim x + y$, and non-mediated, $z \sim x$, models.
-A: In the absence of mediator variable, both mean-based and population-based approaches predictors are identical since there is no any other additional (non-focal predictors) sources of bias.
+\caption{{\bf Mean-based and whole-sample-based predictor effects for mediated effect.} These figures compare the mean-based and whole-sample-based predictor effect of $x$ from a mediated, $z \sim x + y$, and non-mediated, $z \sim x$, models.
+A: In the absence of mediator variable, both mean-based and whole-sample-based approaches predictors are identical since there is no any other additional (non-focal predictors) sources of bias.
 Consequently, in both approaches, the average predicted probability is very close to the observed proportion.
 B: When the mediator variable is included, there is no direct effect of $x$ on $z$ and as a result the predictor effect curve does not align with the observations.
-However, the population-based approach still closely approximates the average proportion in the simulated data.
-The horizontal blue, black and the red dotted lines are the average mean-based, population-based predicted probability and observed proportion, respectively.
+However, the whole-sample-based approach still closely approximates the average proportion in the simulated data.
+The horizontal blue, black and the red dotted lines are the average mean-based, whole-sample-based predicted probability and observed proportion, respectively.
 The vertical dotted black line represents the mean of the focal predictor, and the point at which it crosses the red dotted line represents the expected ``perfect'' prediction at the model center.
@@
 
-Although commonly used \proglang{R} software packages, by default, implements mean-based approach, our simulation results demonstrated that the population-based approach has a potential of yielding results which are more consistent with the observed data.
+Although commonly used \proglang{R} software packages, by default, implements mean-based approach, our simulation results demonstrated that the whole-sample-based approach has a potential of yielding results which are more consistent with the observed data.
 We would, therefore, argue that the use of mean-based approach should have some theoretical justification, especially in complex models.
@@
 
-From our theoretical, methodological and simulation results, researchers using these kind of models should, in the absence of theoretical justification, report predictions based on population-based approach or at least attempt to do a comparison of the two approaches before settling on the most appropriate in answering their research question.
+From our theoretical, methodological and simulation results, researchers using these kind of models should, in the absence of theoretical justification, report predictions based on whole-sample-based approach or at least attempt to do a comparison of the two approaches before settling on the most appropriate in answering their research question.
 Moreover, we provide \proglang{R} package, \pkg{vareffects}, which implements these methods and is available on github (\href{https://github.com/mac-theobio/effects}{https://github.com/mac-theobio/effects}).
@@
 \mathrm{logit(P}(z_i = 1)) &= \eta_i \nonumber\\
-\eta_i &= \beta_0 + \beta_{yz} y_i \nonumber\\
-y_i &= \beta_{xy} x_i + \epsilon_y \nonumber\\
+\eta_i &= \beta_0 + \beta_{xz} x_i + \beta_{yz} y_i \nonumber\\
+y_i &= \rho x_i + \sqrt{1-\rho^2} y_y \nonumber\\
 x_i &\sim \mathrm{Normal(0, 1)} \nonumber\\
-\epsilon_y &\sim \mathrm{Normal(0, 1)} \nonumber\\
+y_y &\sim \mathrm{Normal(0, 1)} \nonumber\\
+\rho &= 0.8 \nonumber\\
 \beta_0 &= 5 \nonumber\\
-\beta_{xy} &= 1 \nonumber\\
+\beta_{xz} &= 0.2 \nonumber\\
 \beta_{yz} &= 1.5 \nonumber\\

### 2022-03-23 manuscript/vareffects_plosone.tex
Spell-check

@@
 This generates -- \emph{predictor effects}\cite{fox2009effect}, \emph{marginal predictions} \cite{leeper2017package} or \emph{estimated marginal means} \cite{lenth2018package}.
-In this article, we refer to this quantity as predictor effects since it should, for example, tell us what we would expect the presponse to be at a particular value or level of the predictor, for an ``average case''.
-Formerly, predictor effects computes the expected outcome by meaningfully holding the non-focal predictors constant (or averaged in some meaningful way) while varying the focal predictor, with the goal that the outcome expected prediction represents how the model responds to the changes in the focal predictor.
+In this article, we refer to this quantity as predictor effects since it should, for example, tell us what we would expect the response to be at a particular value or level of the predictor, for an ``average case''.
+More specifically, predictor effects computes the expected outcome by meaningfully holding the non-focal predictors constant (or averaged in some meaningful way) while varying the focal predictor, with the goal that the outcome expected prediction represents how the model responds to the changes in the focal predictor.
 
 The commonly used \proglang{R} software packages (\pkg{emmeans} and \pkg{effects}) for generating predictor effects, by default, use the average of the non-focal predictors as the reference point.
-However, there are a number of choices one can make when considering this approach -- for example, in the presence of interaction, averaging the interactions (averaging product of interacting predictors) versus product of the averages of the interacting predictors (default for \pkg{emmeans} and \pkg{effects}).
+However, there are a number of choices one can make when considering this approach -- for example, in the presence of interaction, averaging the interactions (averaging product of interacting model variables) versus product of the averages of the interacting model variables (default for \pkg{emmeans} and \pkg{effects}).
 We claim that neither of these two approaches is the most appropriate but we show that averaging the interaction closely matches the observed values.
@@
 
-We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$ and $\beta_{23} = 5$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.
+We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$ and $\beta_{23} = 5$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.
 
@@
 The grey points are the binned simulated $y$.
-The trend lines represents the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other predictors at their average.
+The trend lines represents the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other predictors at their average, based on observed values (red) and the three methods.
 A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the simulated in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$.
-B: Even with the simple interaction between the non-focal predictors, we start seeing deviation of the predicted mean, $\bar{\hat{y}}$, from the simulated, $\bar{y}$, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not, the proposed \pkg{varpred}.}
+Similarly, the predictions based on the three approaches closely matches the ``true'' predictions.
+B: Even with the simple interaction between the non-focal predictors, we start seeing deviation in both predicted mean, $\bar{\hat{y}}$, and overall predictions from the simulated, $\bar{y}$, and ``true'' predictions, respectively, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not, the proposed \pkg{varpred}.}
 \label{fig:justify_plots}
@@
 For \pkg{emmeans} and \pkg{effects}, the confidence bands are much wider because they include uncertainties associated with the non-focal predictors, but narrower and crosses at the mean of the focal predictor, i.e., model center, in \pkg{varpred}.
-In other words, with \pkg{varpred}, we are able to generate isolated confidence bands indicating zero uncertainty at the value of the focal predictor we are more certain about, i.e., mean of the focal predictor.
+In other words, with \pkg{varpred}, we are able to generate isolated confidence bands indicating zero uncertainty at the value of the focal predictor we are more certain about, i.e., mean of the model variable corresponding to the focal predictor.
 For simple models, the point where the confidence bands crosses is the model center and it corresponds to anchor we choose, in this case, the mean of the focal predictor.
@@
 
-\subsection*{Population-based approach for bias correction}
+\subsection*{Whole-sample-based approach for bias correction}
 
@@
 
-Population-based approach is not entirely foolproof.
+Whole-sample-based approach is not entirely foolproof.
 For instance, similar to the mean-based approach, in the case of continuous focal predictors, choosing the representative values of the focal predictors can be very challenging especially if the cases are not evenly distributed around the minimum and the maximum values or within some subgroups defined in the population.

### 2022-03-23 manuscript/mediate_preds.R

@@ -21,7 +21,7 @@ pred_notmediated_none_mean <- getmeans(pred_notmediated_none, what="estimate")
 pred_notmediated_pop <- varpred(mod_notmediated
 	, "x"
 	, bias.adjust="population"
-	, modelname="Population-based"
+	, modelname="Whole-sample-based"
 )
 pred_notmediated_pop_mean <- getmeans(pred_notmediated_pop, what="estimate")
 
@@ -47,10 +47,10 @@ pred_notmediated_plots <- (comparevarpred(vlist=vlist
 	+ geom_hline(data=observed_df_med, aes(yintercept=z, colour=model, lty=model))
 	+ geom_vline(data=observed_df_med, aes(xintercept=x), lty=2, col="black")
 	+ geom_point(data=binned_df, aes(x=x, y=z), colour="grey")
-	+ scale_colour_manual(breaks = c("Observed mean", "Mean-based", "Population-based")
-		, values=c("Observed mean"="red", "Mean-based"="blue", "Population-based"="black")
+	+ scale_colour_manual(breaks = c("Observed mean", "Mean-based", "Whole-sample-based")
+		, values=c("Observed mean"="red", "Mean-based"="blue", "Whole-sample-based"="black")
 	)
-	+ scale_linetype_manual(values=c("Observed mean"=2, "Mean-based"=1, "Population-based"=1))
+	+ scale_linetype_manual(values=c("Observed mean"=2, "Mean-based"=1, "Whole-sample-based"=1))
 	+ labs(colour="Method", linetype="Method", title="A) Non-mediated", y="Predicted probability")
 	+ theme(legend.position="bottom")
 )
@@ -69,7 +69,7 @@ pred_mediated_none_mean <- getmeans(pred_mediated_none, what="estimate")
 pred_mediated_pop <- varpred(mod_mediated
 	, "x"
 	, bias.adjust="population"
-	, modelname="Population-based"
+	, modelname="Whole-sample-based"
 )
 pred_mediated_pop_mean <- getmeans(pred_mediated_pop, what="estimate")
 
@@ -90,10 +90,10 @@ pred_mediated_plots <- (comparevarpred(vlist=vlist
 	+ geom_hline(data=observed_df_med, aes(yintercept=z, colour=model, lty=model))
 	+ geom_vline(data=observed_df_med, aes(xintercept=x), lty=2, col="black")
 	+ geom_point(data=binned_df, aes(x=x, y=z), colour="grey")
[…]

### 2022-03-23 manuscript/vareffects_plosone.tex
Autocommit

@@
 The first one is the coefficient estimates.
-Others are \emph{input variables},  \emph{predictors}, \emph{model matrix}, \emph{predicted values} and \emph{marginal effects}.
-Input variables are the variables that were measured (may be transformed), while predictors are the terms that are entered in the model.
-Hence, predictors encompass the main effects, but also polynomials of input variables and interaction terms (for example in polynomials and splines) \cite{schielzeth2010simple}.
-Model matrix refers to the design matrix whose rows include all combination of variables appearing in the interaction terms, along with the typical values of the focal and non-focal predictors.
+Others are \emph{input and model variables}, \emph{model matrix}, \emph{model center}, \emph{reference point}, \emph{anchor}, \emph{predictors values} and \emph{marginal effects}.
+In particular: input variables are scientific variables underlying an inference or exploration.
+The focal predictor we use for effects or predictions is an input variable \cite{schielzeth2010simple}; model matrix refers to the design matrix whose rows include all combination of variables appearing in the interaction terms, along with the typical values of the focal and non-focal predictors; model variables are variables that represent columns in the model matrix.
+Each input variable will correspond to one or more model variables.
+In particular, variables with more than two categories, or variables modeled with a spline or polynomial response, will correspond to more than one input variable; model center is a point corresponding to a column-wise mean of the model matrix (the mean of one or more model variables).
+The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable; reference point is a (or are) value (or values) chosen for non-focal predictors, when estimating effects.
+Typically the center point, but can instead be a population of quantiles or observations; and lastly, an anchor is the value chosen for the focal predictor when estimating effect confidence intervals (anchor choice does not affect the estimate).
+Typically chosen as the center point of the model variables corresponding to the focal input variable.
 

### 2022-03-24 manuscript/vareffects_plosone.tex
Autocommit

@@
 However, for the interacting predictors, we first choose representative values for the focal predictor and some particular values of the interacting predictor similar to the way we choose values for the focal predictor or predetermined.
-The compute the possible combination of all the chosen levels of the focal predictor with chose or predetermined levels of the interacting predictor.
-In our example, suppose we pick $k$ unique values of the focal predictor, $x_2$, then
+In our example, suppose we pick $i$ and $j$ unique values of the focal predictor, $x_2$ and interacting predictor $x_3$, respectively, then the prediction on linear scale is given by
 
 \begin{align*}
-\eta^\star(x_{2[1:k,j]}, \nset{{\bar{x}^\star}}) = \beta_0 + \beta_1 \bar{x}_1 + \beta_2x_{2[1:k,j]} + \beta_3x_{3[j,1:k]} + \beta_{23}x_{2[1:k,j]}x_{3[1:k,j]}
+\eta^\star(x_{2i}, x_{3j}, \nset{{\bar{x}^\star}}) = \beta_0 + \beta_1 \bar{x}_1 + \beta_2x_{2i} + \beta_3x_{3j} + \beta_{23}x_{2i}x_{3j}.
 \end{align*}
-where $j = 1, 2, \cdots$ are the indices of the unique values of the interacting predictor, $x_3$, $x_{2[1:k,j]}$, for example, are the values of the focal predictor corresponding to the $j$th interacting predictor.
 

### 2022-03-29 manuscript/algebra.Rmd
Autocommit

---
title: Is anchoring better?
---

\newcommand{\yH}{\hat{y}_0}
\newcommand{\yHStarl}{\hat{y}_{0l}^\star}
\newcommand{\yHStarr}{\hat{y}_{0r}^\star}
\newcommand{\xHS}{\bar{x}^\star}
\newcommand{\bH}{\hat{\beta}_1}
\newcommand{\bh}{\hat{\beta}_0}
\newcommand{\sxxb}{\sum{(x_i - \bar{x})^2}}
\newcommand{\sxxbsl}{\sum{(x_i - \xHS_l)^2}}
\newcommand{\sxxbsr}{\sum{(x_i - \xHS_r)^2}}
\newcommand{\var}[1]{\mathrm{var(#1)}}
\newcommand{\ss}{\sigma^2}
\newcommand{\varbH}{\frac{\ss}{\sxxb}}
\newcommand{\varbh}{\ss\left(\frac{1}{n} + \frac{\bar{x}^2}{\sxxb} \right)}

Maybe we can start with simple linear model:

\[
y_i = \beta_0 + \beta_1 x_i + \epsilon_i
\]

so that

\[
\hat{\beta_0} = \bar{y} - \bH\bar{x} 
\]
with a variance of 
\[
\varbh
\]

and

\[
\bH = \frac{\sum{(x_i - \bar{x})}(y_i - \bar{y})}{\sxxb}
\]
with a variance of 
\[
\varbH
\]
where $\ss$ is the residual variance.

## Prediction

We want to describe the uncertainties associated with the new data point, $x_0$, based on the fitted model, i.e.,

\begin{align}
\var{\yH} &= \var{\bh + \bH x_0}\\
&= \varbh + \varbH - 2\ss\frac{x_0\bar{x}}{\sxxb}\\
&= \ss \left(\frac{1}{n} + \frac{(x_0 - \bar{x})^2}{\sxxb}\right).
\end{align}

The default anchor for prediction is the $\bar{x}$. However, we want to investigate what happens to the $\var{\yH}$ for various choices of anchor, $\xHS_l < \bar{x}$ and $\xHS_r > \bar{x}$. In particular, we want to compare

\begin{align}
\var{\yHStarl} = \ss \left(\frac{1}{n} + \frac{(x_0 - \xHS_l)^2}{\sxxbsl}\right)
\end{align}
and 
\begin{align}
\var{\yHStarr} = \ss \left(\frac{1}{n} + \frac{(x_0 - \xHS_r)^2}{\sxxbsr}\right)
\end{align}
with $\var{\yH}$.

### 2022-03-30 manuscript/jd_notes.Rmd
working on notes

@@
 
+Also talk about the fact that what _we_ call prediction plots can have either confidence intervals (for the predicted mean) or ``prediction intervals'' for points (we should not show the latter).
+
 It is interesting, I think, that all of this comes down to choosing a reference, point, an anchor, and then whether or not to include uncertainty in the intercept (so that all of the uncertainty from all of the other variables in the model is somehow channeled through the intercept).
+
+FIGURE showing a splash plot.
+
+## Model center
+
+A linear model naturally centers around the model center.
+This is the average point of the columns of the model matrix.
+The prediction at the model center is always the average value of the response variable.
+
+Note that the model center does not always correspond to logically possible values of the input variables, since there can be more vectors than variables.
+
+## Anchor
+
+If we are trying to focus on effects, the anchor from which we calculate CIs becomes important.
+This is not true for the more conventional prediction plots: we can choose any anchor, and the linear algebra sorts out to the same answer.
+The model center seems like a a natural, stable choice.
+
+FIGURE showing a center-anchored and a zero-anchored isolated effects plot.
+
+If we have more than one input vector for our focal variable, however, the model center does not correspond to any given value of the variable, and we don't get the nice crossing we get in simpler cases.
+
+FIGURE showing a multi-parameter focal variable, and a failure to cross. Compare with using the variable center?
+
+## Reference point ##
+
+In simple models, the reference point affects only the intercept. We can always get this exactly right by using the model center.
+
+FIGURE showing a model-centered reference point and a variable-centered reference point (and showing clearly how the former works better).
+We should make the latter ourselves, but also mention that it is the default for competing packages.
+
+If we have focal interactions, then the reference point also affects our slope. _I feel like we should know about how this part actually works!_ What do the competitors do about focal interactions??
+
+## Confounding
+
+If we are looking to disentangle direct and indirect effects, effects plots seem to make more sense than prediction plots.
+How well can we explain this?
+In the case of a linear model, we expect a univariate fit (or a fit involving covariates orthogonal to the focal predictor) to visually match the data.
+This is not true for the general confounding case, which will help guide us when we get to generalized models.
+
+FIGURE showing an unconfounded fit going through the data and a confounded fit not going through the data.
+The “good” confounded fit still goes through the center, not sure if we need to talk abou tthat.

### 2022-04-20 manuscript/jd_notes.Rmd
meeting prep

@@
 ---
+
+We should talk about estimates before we get to confidence intervals.
+The main thing about estimates is the reference point.
+We should talk specifically about how taking the average as a reference point makes clear sense in a linear model.
+
+We should talk about effect-style plots and prediction-style plots.
+A prediction-style plot is about predicting observations (or the ensemble mean of observations) for a given value of the focal predictor.
+For this, we want to use the classic curved confidence intervals (or prediction intervals).
+A prediction-style plot will generally focus on _total_ effects, and so will often be based on a univariate fit.
+
+An effect-style plot is an attempt to visualize the effect of a focal predictor.
+In this case, we suggest the narrower confidence intervals.
+Here, it makes equal sense to think about the total effect of our focal predictor, from a univariate model, or the direct effects – after controlling for whatever – from a multivariate model.
 

### 2022-04-27 manuscript/algebra.Rmd
Autocommit

@@
 
-\newcommand{\yH}{\hat{y}_0}
-\newcommand{\yHStarl}{\hat{y}_{0l}^\star}
-\newcommand{\yHStarr}{\hat{y}_{0r}^\star}
-\newcommand{\xHS}{\bar{x}^\star}
-\newcommand{\bH}{\hat{\beta}_1}
-\newcommand{\bh}{\hat{\beta}_0}
-\newcommand{\sxxb}{\sum{(x_i - \bar{x})^2}}
-\newcommand{\sxxbsl}{\sum{(x_i - \xHS_l)^2}}
-\newcommand{\sxxbsr}{\sum{(x_i - \xHS_r)^2}}
-\newcommand{\var}[1]{\mathrm{var(#1)}}
-\newcommand{\ss}{\sigma^2}
-\newcommand{\varbH}{\frac{\ss}{\sxxb}}
-\newcommand{\varbh}{\ss\left(\frac{1}{n} + \frac{\bar{x}^2}{\sxxb} \right)}
+\newcommand{\yH}{\hat{y}}
+\newcommand{\x}{\mathbf{X}}
+\newcommand{\xt}{\x^\top}
+\newcommand{\xtx}{\xt\x}
+\newcommand{\xtxI}{(\xtx)^{-1}}
+\newcommand{\b}{\beta}
+\newcommand{\bt}{\b^\top}
+\newcommand{\xb}{\mathbf{\x\b}}
+\newcommand{\ep}{\epsilon}
+\newcommand{\bH}{\hat{\b}}
+\newcommand{\bHt}{\bH^\top}
+\newcommand{\xbH}{\x\bH}
+\newcommand{\xtxBH}{(\xtx)\bH}
+\newcommand{\xty}{\xt y}
+\newcommand{\E}{\mathrm{E}}
+\newcommand{\s}{\sigma}
+\newcommand{\ss}{\s^2}
+\newcommand{\ssH}{\hat{\ss}}
+\newcommand{\xo}{\x_0}
+\newcommand{\xof}{\x_{0,f}}
+\newcommand{\xot}{\xo^\top}
+\newcommand{\v}{\mathrm{var}}
+\newcommand{\a}{\mathbf{A}}
+\newcommand{\xoma}{(\xo - \a)}
+\newcommand{\xomat}{\xoma^\top}
 
-Maybe we can start with simple linear model:
+## OLS estimator
+
+Consider linear model
 
 \[
-y_i = \beta_0 + \beta_1 x_i + \epsilon_i
+y = \xb + \ep
 \]
 
-so that
+Let $\bH$ be the estimate of the population parameter $\b$, we want to minimize the sum of squared residuals $\sum \ep_i^2$, with
+
+\begin{align*}
+\ep &= y - \xbH \\
+\ep^\top\ep &= (y - \xbH) (y - \xbH) \\
+&=y^\top y - 2\bHt\xt y + \bHt \xtx \bH
+\end{align*}
+
+Minimizing above, i.e, taking derivative w.r.t $\bH$ and equating to $0$, we have
 
 \[
-\hat{\beta_0} = \bar{y} - \bH\bar{x} 
+\xtxBH = \xty
 \]
-with a variance of 
-\[
-\varbh
-\]
+
+Provided $\xtxI$ exists:
+
+\begin{align*}
+\xtxI \xtxBH &= \xtxI \xty\\
+\implies \bH &= \xtxI \xty
+\end{align*}
+
+From the result above
+
+\begin{align*}
+\bH &= \xtxI \xty\\
+&= \xtxI \xt (\xb + \ep) \\
+&= \b + \xtxI \xt \ep
+\end{align*}
+
+## Variance-covariance matrix
+
+\begin{align*}
+\E[(\bH - \b)(\bH - \b)^\top] &= \E[(\xtxI \xt \ep) (\xtxI \xt \ep)^T] \\
+&= \E[\xtxI \xt \ep\ep^\top \x \xtxI] \\
+&= \xtxI \xt \E[\ep\ep^\top] \x \xtxI \\
+&= \ss \xtxI
+\end{align*}
 
@@
 \[
-\bH = \frac{\sum{(x_i - \bar{x})}(y_i - \bar{y})}{\sxxb}
+\ssH = \frac{\ep^\top\ep}{n - k}
 \]
-with a variance of 
-\[
-\varbH
-\]
-where $\ss$ is the residual variance.
 
-## Prediction
+## Predicting new data points
 
-We want to describe the uncertainties associated with the new data point, $x_0$, based on the fitted model, i.e.,
+Let $\xo$ be the design matrix containing values of the predictors, from which we want to make predictions, then
 
-\begin{align}
-\var{\yH} &= \var{\bh + \bH x_0}\\
-&= \varbh + \varbH - 2\ss\frac{x_0\bar{x}}{\sxxb}\\
-&= \ss \left(\frac{1}{n} + \frac{(x_0 - \bar{x})^2}{\sxxb}\right).
-\end{align}
+\begin{align*}
+\hat{y} &= \xo\bH\\
+&= \xo\xtxI \xty
+\end{align*}
 
-The default anchor for prediction is the $\bar{x}$.
-However, we want to investigate what happens to the $\var{\yH}$ for various choices of anchor, $\xHS_l < \bar{x}$ and $\xHS_r > \bar{x}$.
-In particular, we want to compare
+and
 
-\begin{align}
-\var{\yHStarl} = \ss \left(\frac{1}{n} + \frac{(x_0 - \xHS_l)^2}{\sxxbsl}\right)
-\end{align}
-and 
-\begin{align}
-\var{\yHStarr} = \ss \left(\frac{1}{n} + \frac{(x_0 - \xHS_r)^2}{\sxxbsr}\right)
-\end{align}
-with $\var{\yH}$.
+\begin{align*}
+\v{(\hat{y})} &= \v[\xo\xtxI \xty]\\
+&= \xo\xtxI \xt \v(y) [\xo\xtxI \xt]^\top \\
+&= \ss \xo\xtxI \xt \x \xtxI \xot\\
+&= \ss \xo \xtxI \xot
+\end{align*}
+
+## Anchoring and effects
+
+Let $\xo$ be a centered model matrix containing appropriately chosen values of focal predictors and non-focal predictors as the average of the corresponding columns in the model matrix, and let $\a$ be the anchor matrix, with same dimensions as $\xo$ and values in each columns repeated $n$ times.
+Also, let $\a_f$ be the anchor column corresponding to the focal predictor, the rest of the columns corresponds the non-focal predictors with same values as in $\xo$.
+Any appropriate value (within the range of focal values) can be chosen for $\a_f$ but by default center point is chosen.
+Then
+
+\begin{align*}
+\v{(\hat{y})} &= \ss \xoma \xtxI \xomat
+\end{align*}
+
+We can see that, for any values $\a_f=\xof$, $\v(\hat{y}) = \boldsymbol{0}$. If $\a_f = \overline{\xof}$, then the anchor is the center point at this observation will still hold.
+
+We need to figure out how to quantify and compare all the variances resulting from choosing other anchors, i.e., $\a_f < \overline{\xof}$ and $\a_f > \overline{\xof}$.

### 2022-05-04 manuscript/vareffects_plosone.tex
Autocommit

--- 2022-03-23 manuscript/vareffects_plosone.tex
@@
 
+\newcommand{\bxo}{\bX^{\star}}
+\newcommand{\bxof}{\bX_f^{\star}}
+\newcommand{\ba}{{\mathbf A}}
+\newcommand{\bao}{{\mathbf A}_0}
+\newcommand{\baf}{{\mathbf A_f}}
+\newcommand{\bafc}{{\mathbf A_{\bar{f}}}}
+
 %% END MACROS SECTION
@@
 Each input variable will correspond to one or more model variables.
-In particular, variables with more than two categories, or variables modeled with a spline or polynomial response, will correspond to more than one input variable.
-\item\textbf{Center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more model variables).
+In particular, variables with more than two categories, or variables modeled with a spline or polynomial response, will correspond to more than one model variables.
+\item\textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more model variables).
 The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable.
@@
 Also, coefficient estimates of models involving nonlinear link functions or interactions lose their direct interpretation \cite{leeper2017interpreting}, meaning that interpretation of derived quantities from these estimates requires some understanding of the specified model.
-An alternative is to explore the \emph{effects} of predictors on the predictions or probabilities of the outcome at various level of the predictor of interest -- \emph{predictor effects} \cite{fox2009effect, leeper2017package, lenth2018package}.
+An alternative is to explore the \emph{effects} of predictors on the predictions or probabilities of the outcome at various level of the predictor of interest -- \emph{prediction and effects} \cite{fox2009effect, leeper2017package, lenth2018package}.
 For example, as a way to plan, public health officials may want to know the effects of household income, wealth index, etc, on the predicted probability of having improved water services among slum dwellers.
 
-When visually presented, predictor effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind of transformations on the predictors whose estimates are usually, but not always, a subject to less clarity of interpretation.
-Further, generating predictor effects together with the associated confidence intervals for regression models has a number of challenges.
+When visually presented, prediction and effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind of transformations on the predictors whose estimates are usually, but not always, a subject to less clarity of interpretation.
+Further, generating predictor and effect plots together with the associated confidence intervals for regression models has a number of challenges.
 In particular:
 \begin{enumerate}
-\item choice of representative values of \emph{focal} predictor and the \emph{reference point} for the non-focal predictors especially in multivariate models
+\item choice of representative values of \emph{focal} predictor and the \emph{reference point} for non-focal predictors especially in multivariate models
 \item  propagation of uncertainty -- appropriate choice of \emph{anchor} for computing confidence intervals; and how to incorporate the uncertainty due to non-focal predictors 
@@
 The most common way of dealing with the first challenge is taking unique levels of the focal (predictor of interest) predictor if discrete or taking appropriately sized quantiles (or bins) if continuous, and then calculating the predictions while holding non-focal (other predictors other than the focal predictor) predictors at their reference point (e.g., means) \cite{hanmer2013behind}.
-This generates -- \emph{predictor effects}\cite{fox2009effect}, \emph{marginal predictions} \cite{leeper2017package} or \emph{estimated marginal means} \cite{lenth2018package}.
-In this article, we refer to this quantity as predictor effects since it should, for example, tell us what we would expect the response to be at a particular value or level of the predictor, for an ``average case''.
-More specifically, predictor effects computes the expected outcome by meaningfully holding the non-focal predictors constant (or averaged in some meaningful way) while varying the focal predictor, with the goal that the outcome expected prediction represents how the model responds to the changes in the focal predictor.
+This generates -- \emph{predictions}, \emph{predictor effects}\cite{fox2009effect}, \emph{marginal predictions} \cite{leeper2017package} or \emph{estimated marginal means} \cite{lenth2018package}.
+In this article, we refer to this quantity as predictions since it should, for example, tell us what we would expect the response to be at a particular value or level of the predictor, for an ``average case''.
+More specifically, we compute the expected outcome by meaningfully holding non-focal predictors constant (or averaged in some meaningful way) while varying the focal predictor, with the goal that the outcome expected prediction represents how the model responds to the changes in the focal predictor.
+On the other hand, effects differs with predictions in the sense that the former is about how we describe the uncertainty around the latter and depends on the anchor.
 
-The commonly used \proglang{R} software packages (\pkg{emmeans} and \pkg{effects}) for generating predictor effects, by default, use the average of the non-focal predictors as the reference point.
+The commonly used \proglang{R} software packages (\pkg{emmeans} and \pkg{effects}) for generating predictions, by default, use the average of non-focal predictors as the reference point.
 However, there are a number of choices one can make when considering this approach -- for example, in the presence of interaction, averaging the interactions (averaging product of interacting model variables) versus product of the averages of the interacting model variables (default for \pkg{emmeans} and \pkg{effects}).
@@
 \includegraphics[width=0.9\textwidth]{justify_preds.inter.pdf}
-\caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} predictor effects for $x_1$ on $y$ for models with and without interactions.}
+\caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} predictions for $x_1$ on $y$ for models with and without interactions.}
 The horizontal blue, green and black lines are the mean predictions, i.e., $\bar{\hat{y}}$, the red ones are the means of the simulated $y$, i.e., $\bar{y}$, while the vertical dotted grey lines are the means of the corresponding focal predictors.
 The grey points are the binned simulated $y$.
-The trend lines represents the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other predictors at their average, based on observed values (red) and the three methods.
+The trend lines represents the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other predictors at their average for the three mthods and the ``true'' predictions based on the simulation parameters.
 A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the simulated in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$.
 Similarly, the predictions based on the three approaches closely matches the ``true'' predictions.
-B: Even with the simple interaction between the non-focal predictors, we start seeing deviation in both predicted mean, $\bar{\hat{y}}$, and overall predictions from the simulated, $\bar{y}$, and ``true'' predictions, respectively, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not, the proposed \pkg{varpred}.}
+B: Even with the simple interaction between non-focal predictors, we start seeing deviation in both predicted mean, $\bar{\hat{y}}$, and overall predictions from the simulated, $\bar{y}$, and ``true'' predictions, respectively, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not, the proposed \pkg{varpred}.}
 \label{fig:justify_plots}
@@
 In particular, \pkg{emmeans} and \pkg{effects} give similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is very close to the simulated average ($\bar{y}$), Fig~\ref{fig:justify_plots}B.
-To generate Fig~\ref{fig:justify_plots}A, all the three packages averages the non-focal predictors $x_2$ and $x_3$.
+To generate Fig~\ref{fig:justify_plots}A, all the three packages averages non-focal predictors $x_2$ and $x_3$.
 On other hand, to generate Fig~\ref{fig:justify_plots}B, the difference in the estimates lies on how each of the packages average the interaction term ($x_2x_3$).
-In particular \pkg{emmeans} and \pkg{effects} computes $\bar{x_2}\bar{x_3}$ while \pkg{varpred} computes $\over{x_2x_3}$.
+In particular \pkg{emmeans} and \pkg{effects} uses \emph{input variables} to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses \emph{model variables} to compute $\over{x_2x_3}$.
 
-Sometimes, one may be interested in the uncertainties associated with the focal predictor of interest only, excluding other uncertainties due to other non-focal predictors -- \emph{isolated} confidence intervals.
+Sometimes, one may be interested in the uncertainties associated with the focal predictor of interest only, excluding other uncertainties due to other non-focal predictors -- effects.
 However, currently, the two packages do not provide straightforward way do achieve this.
-To illustrate this, we generate predictor effect of $x_2$ from model~\ref{eq:simple_inter_higher_no_interaction} together with the associated $95\%$ confidence bands, as shown in Fig~\ref{fig:justify_ci_plots}.
+To illustrate this, we generate predictions of $x_2$ from model~\ref{eq:simple_inter_higher_no_interaction} together with the associated $95\%$ confidence bands, as shown in Fig~\ref{fig:justify_ci_plots}.
 
@@
 \includegraphics[width=0.9\textwidth]{justify_preds.isolate.pdf}
-\caption{{\bf A comparison of traditional and isolated $95\%$ confidence bands.} The description of horizontal, vertical and trend lines remain the same as above.
-The wider dotted blue curves overlaying the green curves are the traditional confidence bands from \pkg{emmeans} and \pkg{effects}, while the narrower black curves crossing at the mean of the focal predictor, i.e., the model center, are the isolated confidence bands from \pkg{varpred}.
-For simple models, the isolated confidence bands crosses at the model center.}
+\caption{{\bf A comparison of conventional $95\%$ confidence bands and effects.} The description of horizontal, vertical and trend lines remain the same as above.
+The wider dotted blue curves overlaying the green curves corresponds to the conventional confidence bands around the predcitions from \pkg{emmeans} and \pkg{effects}, while the narrower black curves crossing at the mean of the focal predictor, i.e., the model center, corresponds to the effects of the focal predictor from \pkg{varpred}.
+For simple models, the isolated confidence bands crosses at the model center.
+The wider curves incorporates not only the uncertainties due to intercept term but also other non-focal predictors.
+On the other hand, effects (narrower crossing lines) only take into account the main effect uncertainty of the focal predictor.}
 \label{fig:justify_ci_plots}
@@
 
-For \pkg{emmeans} and \pkg{effects}, the confidence bands are much wider because they include uncertainties associated with the non-focal predictors, but narrower and crosses at the mean of the focal predictor, i.e., model center, in \pkg{varpred}.
-In other words, with \pkg{varpred}, we are able to generate isolated confidence bands indicating zero uncertainty at the value of the focal predictor we are more certain about, i.e., mean of the model variable corresponding to the focal predictor.
+For \pkg{emmeans} and \pkg{effects}, the confidence bands are much wider because they include uncertainties associated with the intercept and non-focal predictors, but narrower and crosses at the mean of the focal predictor, i.e., \emph{center point or mean-based}, in \pkg{varpred}.
+In other words, with \pkg{varpred}, we are able to generate effects indicating zero uncertainty at the value of the focal predictor we are more certain about, i.e., mean of the model variable corresponding to the focal predictor (center point or mean-based).
 For simple models, the point where the confidence bands crosses is the model center and it corresponds to anchor we choose, in this case, the mean of the focal predictor.
@@
 In particular, bias in expected mean prediction induced by nonlinear transformation of the response variable can lead inaccurate predictions.
-An alternative to mean-based reference point, i.e., averaging the non-focal predictors, is the \emph{whole-sample-based} approach, discussed in detail later, which involves computing the prediction over the population of the non-focal predictors and then averaging across the values of the focal predictor \cite{hanmer2013behind}.
+An alternative to mean-based reference point, i.e., averaging non-focal predictors, is the \emph{whole-sample-based} approach, discussed in detail later, which involves computing the prediction over the population of non-focal predictors and then averaging across the values of the focal predictor \cite{hanmer2013behind}.
 
-The main purpose of this article is to discuss and implement various approaches for computing predictor effects and provide an alternative method for computing the associated confidence intervals.
+The main purpose of this article is to discuss and implement various approaches for computing predictions and provide an alternative method for computing the associated confidence intervals, i.e., effects.
 We further explore and demonstrate, using simulated data, approaches for correcting bias in predictions for GL(M)Ms involving nonlinear link functions.
@@
 The first one is the coefficient estimates.
-Others are \emph{input variables},  \emph{predictors}, \emph{model matrix}, \emph{predicted values} and \emph{marginal effects}.
-Input variables are the variables that were measured (may be transformed), while predictors are the terms that are entered in the model.
-Hence, predictors encompass the main effects, but also polynomials of input variables and interaction terms (for example in polynomials and splines) \cite{schielzeth2010simple}.
+Others are \emph{input and model variables}, \emph{model matrix}, \emph{model center}, \emph{reference point} and \emph{anchor}.
+
+\subsection*{Input variables}
+
+These are scientific variables underlying an inference or exploration. The focal and non-focal predictors we use for effects or predictions are an input variables.
+
+\subsection*{Model matrix}
+
 Model matrix refers to the design matrix whose rows include all combination of variables appearing in the interaction terms, along with the typical values of the focal and non-focal predictors.
 
-In simple linear models with no interaction terms, the default output for the coefficient estimates are simple and directly interpretable as the expected change in outcome for a unit change in focal predictor.
-This is the \emph{unconditional marginal effect} \cite{leeper2017interpreting} and it is constant across all the observations and levels of all other predictors.
-Consider models ~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher}.
-The marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher_no_interaction} is $\frac{\partial y}{\partial x_2} = \beta_2$.
-On the other hand, the marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher} is given by $\frac{\partial y}{\partial x_2} = \beta_2 + \beta_{23}x_3$.
-In other words, if there are no interactions, the marginal effect of $x_2$ on $y$ is constant, while, if there are interactions in the model, the marginal effect of a change in $x_2$ on $y$ depends on the value of the other \emph{conditioning} predictor, $x_3$.
+\subsection*{Model variables}
 
-% In general, for a continuous variable, $x_k$, the marginal effect is $\frac{\partial y}{\partial x_k} = g(x\beta)\beta_k$, where $g()$ is the probability density function.
-This implies that, in addition to coefficient estimates, the marginal effect depends on the nature of $g()$ and the values of the other non-focal predictors.
-For nonlinear link functions, $g()$, the effect is not constant and instead depends on the function and the values over which the curve is evaluated.
+These are variables that represent columns in the model matrix.
+Each input variable may correspond to one or more model variables.
+In particular, variables with more than two categories, or variables modeled with a spline or polynomial response, will correspond to more than one model variables.
 
-To distinguish between predictor effects and marginal effects, consider result from a hypothetical  simulated example -- regression of household size as function of household wealth index and age of household head, as shown in Fig~\ref{fig:qoi_age_pred_plot}.
-Since the model has no interactions, the relationship between the predicted household size and age is linear, hence the marginal effect of age is the slope ($\frac{\Delta \mathrm{hh size}}{\Delta \mathrm{age}}$) of the predictor effect line and can be calculated irrespective of the values of wealth index.
+\subsection*{Model center}
+
+A point corresponding to a column-wise mean of the model matrix (the mean of one or more model variables).
+We also refer to this as mean-based approach.
+The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable -- the case in Fig~\ref{fig:justify_plots}.
+In simple linear models (without) interaction, the input variable can be used as a proxy.
+
+\subsection*{Reference point}
+
+Corresponds to the value or values chosen for non-focal predictors, when estimating the predictions.
+Typically the center point (mean-based), but can instead be a whole-sample of quantiles or observations.
+Appropriately choosing the reference point is very crucial in generating ``correct'' predictions especially when dealing with non-linear link functions.
+
+\subsection*{Anchor}
+
+The value chosen for the focal predictor when estimating effect confidence intervals (anchor choice does not affect the main prediction estimates).
+Typically chosen as the center point of the model variables corresponding to the focal input variable but other sensible values can be chosen too.
+
+If we are trying to focus on effects, the anchor from which we calculate CIs becomes important.
+This is not true for the more conventional prediction plots: we can choose any anchor, e.g., zero-anchored, but model center seems like a natural, stable choice, i.e., center-anchored.
+In Fig~\ref{fig:justify_anchors}, we compare effects plots anchored at the minimum and center (mean) of the focal predictor.
 
@@
 \centering
-\includegraphics[width=0.8\textwidth]{qoi_age_pred_plot-figure3.pdf}
-\caption{{\bf A comparison of predictor effect and unconditional marginal effect of age on household size.} The black line is the predictor effect trend line, while the dotted lines indicate the changes along the \code{age} and \code{household size} axis.
-For a linear model with no interaction, the marginal effect is the slope of the predictor effect line.}
-\label{fig:qoi_age_pred_plot}
+\includegraphics[width=0.7\textwidth]{justify_anchors.combined.pdf}
+\caption{{\bf A comparison of min-based and center- (mean-) based anchors.} The center-based effects are based on the mean of the focal predictor while the min-anchored effects are based on the minimum of the focal predictor.
+As we mentioned earlier, the choice of the anchor do not affect the main estimates (central trend lines).}
+\label{fig:justify_anchors}
 \end{figure}
 
-Predictor effect, on the other hand, is the expected household size for a particular age, holding wealth index at its mean.
-In particular, the purpose and goal of a predictor effect seems fairly straightforward; for specified values of a focal predictor, we want to give a point estimate and confidence intervals for the prediction of the model for a ``reference point'' individual with those values of the predictors.
+If we have more than one model variables for our focal predictor, however, the model center does not correspond to any given value of the variable, and we don't get the nice crossing we get in simpler cases.
+See Fig~\ref{fig:pred_cubic_plots}.
+
+\subsection*{Prediction- vs. effect- styled plots}
+
+A prediction-style plot is about predicting observations (or the ensemble mean of observations) for a given value of the focal predictor.
+For this, we want to use the classic curved confidence intervals (or prediction intervals).
+A prediction-style plot will generally focus on \emph{total} effects, and so will often be suited for a univariate fit.
+An effect-style plot, however, is an attempt to visualize the effect of a focal predictor.
+In this case, we suggest the narrower confidence intervals.
+Here, it makes equal sense to think about the total effect of our focal predictor, from a univariate model, or the direct effects – after controlling for whatever – from a multivariate model.
+See Fig~\ref{fig:justify_ci_plots}.
+
+\subsection*{Predictions vs. effects}
+
+We would like to summarize by talking about predictions vs. effects. Predictions are the standard curves; they tell us what we expect to see if we know the value of the focal predictor.
+
+The lines inside (effects) are a little different: they are a good way to visualize the effect of the focal predictor.
+Unlike predictions, where the uncertainty captures the uncertainty in the whole model, the effects plot focuses on uncertainty in the effect of the focal predictor only -- thus, for example, if an effect is barely significant (P exactly 0.05), one end of the CI will be exactly a horizontal line for the effects plot, but not for the prediction plot.
+Also, the uncertainty curves for the effect, but not for the prediction, depend on the ``anchor''.
 
@@
 
-To get an intuition of how conditioning on the mean values of the non-focal predictors work, suppose we are interested in predictor effects of a particular predictor, i.e., focal, $x_f$, from the set of predictors.
-To keep it simple, assume that the model has no interaction terms.
+To get an intuition of how conditioning on the mean values of non-focal predictors work, suppose we are interested in predictions at the levels of a particular predictor, i.e., focal, $x_f$, from the set of predictors.
+As a starting point, assume that the model has no interaction terms.
 The idea is to choose a reference point for the values of non-focal predictors.
 For example, fixing the values of non-focal predictor(s) at some typical values -- typically determined by averaging (for now) in some meaningful way, for example, arithmetic mean  for continuous and average over the levels of the factors for categorical non-focal predictors.
-We refer to this as \emph{mean-based} approach.
+We refer to this as centered or mean-based approach.
 The most convenient way to achieve this is by constructing $\bX^\star$ by averaging the columns of non-focal predictors in model matrix $\bX$, and together with appropriately chosen values of focal predictor.
@@
 Consider a simple linear model with linear predictor $\eta = \bX\bbeta$ and let $g(\boldmu) = \boldeta$ be an identity link function (in the case of simple linear model), where $\boldmu$ is the expected value of response variable $y$.
-Let $\hat{\bbeta}$ be the estimate of $\bbeta$, together with the estimated covariance matrix $\Sigma = V(\hat{\bbeta})$ of $\hat{\bbeta}$.
+Let $\hat{\bbeta}$ be the estimate of $\bbeta$, together with the estimated covariance matrix $\boldsymbol{\Sigma} = V(\hat{\bbeta})$ of $\hat{\bbeta}$.
 Let $\mathbf{X^*}$ be the model matrix, inheriting most of its key properties, for example transformations on predictors and interactions from the model matrix, $\mathbf{X}$.
-Then the prediction $\hat{\boldeta}^\star = \bX^\star\hat{\bbeta}$ is the predictor effect for the focal predictor in question \cite{fox2009effect}.
+Then the quantity $\hat{\boldeta}^\star = \bX^\star\hat{\bbeta}$ is the prediction with respect to the focal predictor in question \cite{fox2009effect}.
 
-An alternative formulation of predictor effect involves, expressing the linear predictor as the sum of the focal and non-focal predictors' linear predictor. In particular,
+An alternative formulation of $\hat{\boldeta}^\star$ involves expressing the linear predictor as the sum of the focal and non-focal predictors' linear predictor. In particular,
 
@@
 Higher order terms such as interactions, splines, polynomials, etc., can be within the focal predictor, within (between) non-focal predictor(s), between the focal and non-focal predictor(s).
-To distinguish the three, suppose the model which describes the hypothetical household size simulation described above is
+To distinguish the three, suppose the model which describes the hypothetical simulated household size simulation based on a number of socio-demographic factors is
 
@@
 
-In the first case, with $Age$ as the focal predictor, the interaction is within its polynomial terms.
-In this case, each of the polynomial terms of the focal predictor are evaluated independently across the chosen levels of the focal predictor, $Age_i$.
-Specifically, the higher order terms are treated as additional columns of the model matrix evaluated with the same values of the focal predictor, while the non-focal predictors are fixed at their reference point as discussed in the previous section.
-In this case, the predictor effect of $Age$ on the linear scale becomes
+In the first case, with $Age$ as the focal predictor, the interaction is within its polynomial terms (and 3 model variables associated with it).
+In this case, each of the polynomial terms (model variables) of the focal predictor are evaluated independently across the chosen levels of the focal predictor, $Age_i$.
+Specifically, the higher order terms are treated as additional columns of the model matrix evaluated with the same values of the focal predictor, while non-focal predictors are fixed at their reference point as discussed in the previous section.
+In this case, the predictions associated with $Age$ on the linear predictor scale becomes
 
@@
 
-In the second case, with $Wealthindex$ as the focal predictor, the higher order interactions in the non-focal predictor, $Age$, are treated simply as additional columns in the variable space (of model matrix) and appropriate choice of reference point applies just like in the models without interactions.
-For instance, in mean-based approach, we simply average all the non-focal interaction terms.
-In this setup, the predictor effect of the previous model, on link scale, becomes
+In the second case, with $Wealthindex$ as the focal predictor, the higher order interactions in non-focal predictor, $Age$, are treated simply as additional columns in the variable space (of model matrix) and appropriate choice of reference point applies just like in the models without interactions.
+For instance, in mean-based approach, we simply average all non-focal interaction terms.
+Thus
 
@@
 
-Lastly, consider model~\ref{eq:simple_inter_higher}, with the interactions between the focal predictor, $x_2$, and the non-focal predictor, $x_3$.
+Lastly, consider model~\ref{eq:simple_inter_higher}, with the interactions between the focal predictor, $x_2$, and non-focal predictor, $x_3$.
 In this case, the non-interacting predictors are treated as non-focal predictors, with appropriate reference points.
 However, for the interacting predictors, we first choose representative values for the focal predictor and some particular values of the interacting predictor similar to the way we choose values for the focal predictor or predetermined.
-The compute the possible combination of all the chosen levels of the focal predictor with chose or predetermined levels of the interacting predictor.
-In our example, suppose we pick $k$ unique values of the focal predictor, $x_2$, then
+In our example, suppose we pick $i$ and $j$ unique values of the focal predictor, $x_2$ and interacting predictor $x_3$, respectively, then the prediction on linear predictor scale is given by
 
 \begin{align*}
-\eta^\star(x_{2[1:k,j]}, \nset{{\bar{x}^\star}}) = \beta_0 + \beta_1 \bar{x}_1 + \beta_2x_{2[1:k,j]} + \beta_3x_{3[j,1:k]} + \beta_{23}x_{2[1:k,j]}x_{3[1:k,j]}
+\eta^\star(x_{2i}, x_{3j}, \nset{{\bar{x}^\star}}) = \beta_0 + \beta_1 \bar{x}_1 + \beta_2x_{2i} + \beta_3x_{3j} + \beta_{23}x_{2i}x_{3j}.
 \end{align*}
-where $j = 1, 2, \cdots$ are the indices of the unique values of the interacting predictor, $x_3$, $x_{2[1:k,j]}$, for example, are the values of the focal predictor corresponding to the $j$th interacting predictor.
 
@@
 
-\section*{Uncertainty propagation}
+\section*{Uncertainty estimation}
 
@@
 In principle, every value of focal predictor has a different CI.
-The traditional way to compute variances for predictions is $\sigma^2 = \textrm{Diag}(\bX^\star \Sigma \bX^{\star\top})$ \cite{lenth2018package, fox2009effect}, so that the confidence intervals are $\eta \pm q\sigma$, where $q$ is an appropriate quantile of Normal or t distribution.
-This approach incorporates all the uncertainties -- including the uncertainty due to non-focal predictors.
+The conventional way to compute variances for predictions is $\sigma^2 = \textrm{Diag}(\bX^\star \boldsymbol{\Sigma} \bX^{\star\top})$ \cite{lenth2018package, fox2009effect}, so that the confidence intervals are $\eta \pm q\sigma$, where $q$ is an appropriate quantile of Normal or t distribution.
+This generates predictions-styled CIs and incorporates all the uncertainties -- including the uncertainty due to the intercept and non-focal predictors.
 But what if we are only interested in the uncertainty as a result of the focal predictor, so that the confidence intervals are $\eta \pm q \sigma_f$?
-We call this isolated confidence intervals.
+We call this anchored CIs and generates effects-styled CIs.
 
-Currently, commonly used \proglang{R} packages for constructing predictions, by default, do not exclude the uncertainties resulting from the non-focal predictors when computing the CIs.
+Currently, commonly used \proglang{R} packages for constructing predictions, by default, do not exclude the uncertainties resulting from non-focal predictors when computing the CIs.
 A non-trivial way to exclude uncertainties associated with non-focal predictors in some of these packages is to provide a user defined variance-covariance matrix with the covariances of non-focal terms set to $0$ -- \emph{zeroing-out} variance-covariance matrix.
 This only works when the input predictors are \emph{centered} prior to model fitting, in case of numerical predictors, and even much complicated when the predictors are categorical.
-We first describe the variance-covariance based approach and then discuss our proposed method which is based on \emph{centering model matrix} and does not require input predictors to be scaled prior to model fitting.
+We first describe the variance-covariance based approach and then discuss our proposed method which is based on the model center and does not require input predictors to be scaled prior to model fitting.
 
@@
 The computation of $\hat{\eta}^\star$ remains the same as described above.
-However, to compute $\sigma$, $\Sigma$ is modified by \emph{zeroing-out} (the variance-covariance of all non-focal predictors are set to zero) variances of non-focal terms.
+However, to compute $\sigma$, $\boldsymbol{\Sigma}$ is modified by \emph{zeroing-out} (the variance-covariance of all non-focal predictors are set to zero) variances of non-focal terms.
 Although this is the simplest approach, it requires centering of continuous, i.e., $x_c = x - \bar{x}$, predictors prior to model fitting and proper way to average categorical predictors.
 
-\subsection*{Centered model matrix}
+\subsection*{Model center based}
 
-Consider centered model matrix $\bX^{\star}_{c} = \{\bX_f^\star, \nset{{\bX}^\star} - \nset{{\bar{\bX}}^\star}\}$.
-It follows that the non-focal terms in $\bX^{\star}_{c}$ are all zero in simple models without interactions but are isolated (narrower) around the model center in models involving complex interactions.
-Consequently the uncertainty due to non-focal predictors are isolated in the computation of $\sigma^2 = \textrm{Diag}(\bX^\star_c \Sigma \bX^{\star\top}_c)$.
-In addition, the computation of $\bX^{\star}_c$ impacts only on the intercepts and the non-focal terms, i.e., the slopes and variance of the focal predictors are not affected.
-This means that we can still generate isolated CIs without necessarily centering the predictors prior to model fitting.
+Let $\bxo$ be a centered model matrix containing appropriately chosen values of focal predictors and centered non-focal predictors, and let $\ba$ be the anchor matrix, with same dimensions as $\bxo$ and values in each columns repeated $n$ times.
+Also, let $\baf$ be the column of $\ba$ corresponding to the focal predictor, the rest of the columns corresponds non-focal predictors with same values as in $\bxo$.
+Any appropriate values (within the range of focal values) can be chosen for $\baf$ but for model center (center-anchored), we have $\bafc$ .
+Thus $\sigma_f^2 = \textrm{Diag}((\bxo - \bafc) \boldsymbol{\Sigma} (\bxo - \bafc)^\top)$.
+We can see that $\forall ~\bxof=\baf$, $\sigma_f^2 = \boldsymbol{0}$.
+This is true for the center-anchored provided $\bxof=\bafc$.
+In other words non-focal terms (model variables) in $\bX^{\star}_{c} = (\bxo - \bafc)$ are all zero in simple models without interactions but are isolated (narrower) around the model center in models involving complex interactions.
+In addition, the computation of $\bX^{\star}_c$ impacts only on the intercepts and non-focal model variables, i.e., the slopes and variance of the focal predictors are not affected.
+This means that we can still generate effects without necessarily centering the predictors prior to model fitting.
+
+%% \subsection*{Is center-anchored better?}
+%% 
+%% We want to compare the amount of uncertainty when there is/no anchor.
+Let $\bao = \boldsymbol{0}$ be the anchoring matrix with all entries in $0$s, and let $\bafc$ be the center-anchored matrix defined above.
+Then entries in $(\bxo - \bao) \geq (\bxo - \bafc)$.
+%%
 
@@
 In many applications, it usually important to report the estimates that reflect the expected values of the untransformed response.
-However, when dealing with nonlinear link functions, it is even harder to generate correct predictions that reflect the untransformed response due to the bias in the expected mean induced by the nonlinear transformation of the response variable.
+This is not a major issue in simple linear models.
+However, when dealing with nonlinear link functions, it is even harder to generate correct predictions that reflect this due to the bias in the expected mean induced by the nonlinear transformation of the response variable.
 In such cases, bias correction is needed when back-transforming the predictions to the original scales.
@@
 The most precise (although not necessarily accurate!) way to predict is to condition on values of the focal predictor and make predictions for all observations (members of the population) \cite{hanmer2013behind}.
-A key point is that the nonlinear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population to a one-dimensional distribution of $\eta^\star(x_f, \nset{x})$, which is a function of focal predictor and the observed values of the non-focal predictors, as opposed to the definition in Equation~\ref{eq:eta_mean}.
+A key point is that the nonlinear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population to a one-dimensional distribution of $\eta^\star(x_f, \nset{x})$, which is a function of focal predictor and the observed values of non-focal predictors, as opposed to the definition in Equation~\ref{eq:eta_mean}.
 More specifically:
@@
 \begin{itemize}
-\item compute linear predictor associated with the non-focal predictors, $\nset{\eta} = \sum \nset{\beta} \nset{x}$
+\item compute linear predictor associated with non-focal predictors, $\nset{\eta} = \sum \nset{\beta} \nset{x}$
 \item compute linear predictor associated with the focal predictors, $\eta_{jf} = \sum{\beta_f x_{jf}}$
@@
 \begin{align}
-\sigma_{jf}^2 = \textrm{Diag}(\bX^\star_{jc} \Sigma \bX^{\star\top}_{jc})
+\sigma_{jf}^2 = \textrm{Diag}(\bX^\star_{jc} \boldsymbol{\Sigma} \bX^{\star\top}_{jc})
 \end{align}
@@
 
-\section*{Mean-based vs whole-sample-based}
+\section*{Mean-based vs. whole-sample-based}
 
-As mentioned above, the two common choices to generate predictor effects are: 1) setting the non-focal predictors to their mean -- mean-based; and 2) using the entire population of the non-focal predictors -- whole-sample-based.
-If the link function is nonlinear and/or the model has complex higher order interactions, the average of the predictions evaluated at the mean of the non-focal predictors and the average of predictions evaluated at the population level (and then averaged) of the non-focal predictor are not equivalent, i.e.,
+As mentioned above, the two common choices to generate predictions are: 1) setting non-focal predictors to their mean -- centered- (mean-) based; and 2) using the entire population of non-focal predictors -- whole-sample-based.
+If the link function is nonlinear and/or the model has complex higher order interactions, the average of the predictions evaluated at the mean of non-focal predictors and the average of predictions evaluated at the population level (and then averaged) of non-focal predictor are not equivalent, i.e.,
 
@@
 However, picking a single value, e.g., mean of the predictor, on which to draw conclusions about the effect can be problematic, unrealistic or not contained in or representative of the population.
-In addition, the mean-based approach fails to use every values of the non-focal predictors hence not utilizing the full potential of the information contained in the data.
+In addition, the mean-based approach fails to use every values of non-focal predictors hence not utilizing the full potential of the information contained in the data.
 This may limit the inferences we can make about the entire population.
-In general, the mean-based approach provides the predictor effect of an average case, whereas, the whole-sample-based approach, summarizes the predictor effect over the entire population -- and in some applications, the effect of an average case might not be generalizable to the entire population, especially, if the average does not represent the population.
+In general, the mean-based approach provides the predictions of an average case, whereas, the whole-sample-based approach, summarizes the predictions over the entire population -- and in some applications, the effect of an average case might not be generalizable to the entire population, especially, if the average does not represent the population.
 The whole-sample-based focuses on specific observations since the prediction is first obtained for each observation and then averaged across the levels of the focal predictor.
@@
 Another potential concern with the mean-based approach arises in situations where direct naive use leads to rare or meaningless basis for generalization.
-For example, setting categorical dummy categorical variables in the model matrix to their means, which, by default, sets them to their sample means or observed proportions.
+For example, setting dummy categorical variables in the model matrix to their means, which, by default, sets them to their sample means or observed proportions.
 For example, a dummy variable for christian household heads may be set to $0.2$, translating to prediction for an household head who is $20\%$ christian.
@@
 
-We now illustrate the construction of predictor effects with isolated confidence intervals and also demonstrate that the mean-based and population approaches produce different results in models with nonlinear link functions and/or additional source of potential bias.
-In addition, we demonstrate that whole-sample-based approach can be used to correct bias in the predictions.
+We now illustrate the construction of prediction-styled and effect-styled plots and also demonstrate that the mean-based and whole-sample-based approaches produce different results in models with nonlinear link functions and/or additional source(s) of potential bias.
+In addition, we demonstrate that whole-sample-based approach can be used for bias correction.
 
-\subsection*{Isolated confidence intervals}
+\subsection*{Predictions-styled and effects-styled plots}
 
 As mentioned above, in some applications, one may be interested in uncertainties associated with the focal predictor only.
-In that case, it is important to exclude all the uncertainties as a result of the non-focal predictors and only show confidence intervals describing the predictor of interest.
+In that case, it is important to exclude all the uncertainties as a result of non-focal predictors and only show confidence intervals describing the predictor of interest.
 To illustrate this, we consider the simulation model described in \nameref{S1_Appendix}.
 
-We start by fitting two models from the simulated data, i.e., consider a cubic polynomial predictor, age, as a focal predictor and also, consider the non-polynomial predictor, wealth index as the focal predictor.
-After fitting the model, we constructed predictor effects and then compared isolated vs non-isolated confidence intervals associated with these predictions (see Fig~\ref{fig:pred_cubic_plots}).
+We start by generating two predictions from the fitted model \nameref{S1_Appendix}: 1) one with cubic polynomial of \code{age} as a focal predictor; and 2) one with a simple focal predictor (non-polynomial) of \code{Wealthindex}.
+Then we compared non-anchored (more of predictions) and mean-anchored (effects) confidence intervals around the predicted values (see Fig~\ref{fig:pred_cubic_plots}).
 
@@
 \includegraphics[width=0.8\textwidth]{pred_cubic_plots-figure4.pdf}
-\caption{{\bf Isolated and non-isolated predictor effects $95\%$ confidence bands.} These figures compare the predictor effects together with their corresponding $95\%$ confidence bands.
-A: The focal predictor is a cubic polynomial.
-B: The focal predictor is not a polynomial function, implying that the complex interaction is in the non-focal predictor.
-In both cases, the confidence bands are narrower in the isolated case, and crosses at the mean of the focal predictor (model center) in the case of simple (non-polynomial) focal predictor -- B.
-In the case of cubic polynomial focal predictor, the model center is not a point in the predictor space, i.e., mean of the focal predictor, but a combination of all the terms in the cubic polynomial, consequently, the isolated bands do not cross at a point -- A.
-The horizontal black and red dotted lines are the observed and predicted average household size, i.e., $\over{hh~size}$ and $\over{\widehat{hh~size}}$, respectively; the inner red dotted curves and crossing line are the isolated confidence bands for A and B, respectively, while the outer blue curves are the non-isolated confidence bands in both cases.
-The central curve and trend line, in A and B, respectively, are the predictor effect of age and wealth index on the predicted household size, respectively.}
+\caption{{\bf Prediction-styled and effect-styled plots for cubic polynomial and simple focal predictors.} A: The focal predictor is a cubic polynomial.
+B: The focal predictor is not a polynomial function, implying that the complex interaction is in non-focal predictor.
+In both cases, the mean-anchored effects are narrower than the non-anchored (predictions), but crosses at the mean of the focal predictor (model center) only in the case of simple (non-polynomial) focal predictor -- B.
+In the case of cubic polynomial focal predictor, the model center is not a point in the predictor space, i.e., mean of the focal predictor, but a combination of all the model variables in the cubic polynomial, consequently, mean-anchored curves do not cross at a point -- A.
+The horizontal black and red dotted lines are the observed and predicted average household size, i.e., $\over{hh~size}$ and $\over{\widehat{hh~size}}$, respectively; the inner red dotted curves and crossing lines are the anchored confidence bands for A and B, respectively, while the outer blue curves are the non-anchored confidence bands in both cases.
+The central curve and trend line, in A and B, respectively, are the predicted household size based on age and wealth index, respectively.}
 \label{fig:pred_cubic_plots}
@@
 
-In the absence of complex higher order interaction terms or transformations on the focal predictor in simple linear models, we would expect, at the model center, the observed average, $\over{hh~size}$, and the average of the predictions, $\over{\widehat{hh~size}}$, to be identical and cross, in the case of isolated confidence bands, at the model center (see Fig~\ref{fig:pred_cubic_plots}B).
-On the other hand, if the focal predictor is characterized by interactions or any other form of transformation, the model center is not necessarily a point in the predictor space.
-In this case, the isolated confidence bands will be narrower than the non-isolated but not necessarily crossing at the model center (see Fig~\ref{fig:pred_cubic_plots}A).
+In the absence of complex higher order interaction terms or transformations on the focal predictor in simple linear models, we would expect, at the model center, the observed average, $\over{hh~size}$, and the average of the predictions, $\over{\widehat{hh~size}}$, to be identical and cross with the mean-anchored confidence bands (effects), at the model center (see Fig~\ref{fig:pred_cubic_plots}B).
+On the other hand, if the focal predictor is characterized by complex interactions or any other form of transformation, the model center is not necessarily a point in the predictor space.
+In this case, the mean-anchored confidence bands will be narrower than the non-anchored ones but not necessarily crossing at the model center (see Fig~\ref{fig:pred_cubic_plots}A).
 
@@
 Mean-based and whole-sample-based approaches can produce very different results in models with nonlinear link functions with additional sources of potential bias such as additional non-focal predictor(s), random effect terms or complex interactions.
-To demonstrate this, we considered a two predictor binary outcome simulation described in \nameref{S2_Appendix}, such that the age effect is way smaller than that of the wealth index, and compared their predictor effect on the predicted probability of improved water quality as shown in Fig~\ref{fig:pred_bin_plots}.
+To demonstrate this, we considered a two predictor binary outcome simulation described in \nameref{S2_Appendix}, such that the age effect is way smaller than that of the wealth index, and compared their effects on the predicted probability of improved water quality as shown in Fig~\ref{fig:pred_bin_plots}.
 
@@
 \includegraphics[width=0.8\textwidth]{pred_bin_plots-figure5.pdf}
-\caption{{\bf Mean-based and whole-sample-based predictor effects.} The two approaches produce relatively different predictor effects for each of the predictors but the difference is more pronounced when the effect of the particular focal predictor has no strong effect -- \emph{on the left}.
-In both cases, age and wealth index effect, the average predicted probability of improved water based on whole-sample-based approach is very close to the observed proportion in the simulated data.
+\caption{{\bf Mean-based and whole-sample-based predictions.} The two approaches produce relatively different estimates in each of the predictors but the difference is more pronounced when the effect of the particular focal predictor has no strong effect -- \emph{on the left}.
+In both cases, \emph{right} and \emph{left}, the average predicted probability of improved water based on whole-sample-based approach is very close to the observed proportion in the simulated data.
 The horizontal blue, black and the red dotted lines are the average mean-based, whole-sample-based predicted probability and observed proportion of improved water quality, respectively.
@@
 
-Consider the simple indirect mediation simulation described in \nameref{S3_Appendix}.
-We then fitted two models -- \emph{non-mediated} which models \emph{z} as a function of \emph{x} and the \emph{mediated} which models \emph{z} as a function of both \emph{x} and \emph{y}.
+Using simple indirect mediation simulation model described in \nameref{S3_Appendix}, fitted two models -- \emph{non-mediated} which models \emph{z} as a function of \emph{x} (univariate model) and the \emph{mediated} which models \emph{z} as a function of both \emph{x} and \emph{y} (multivariate model).
 In both cases, we compared mean-based and the whole-sample-based predictions, as shown in Fig~\ref{fig:pred_mediated_plots}.
@@
 The argument and results we present in this paper supports a greater need for a shift on focus on what and how to present predictions from generalized models.
-For example, we believe that isolated confidence intervals could provide more clarity concerning the uncertainty due to the predictor of interests as opposed to the traditional way of incorporating everything.
+For example, we believe that isolated confidence intervals could provide more clarity concerning the uncertainty due to the predictor of interests as opposed to the conventional way of incorporating everything.
 
@@
 \begin{align}\label{sim:glm_two_pred}
-\mathrm{status}_i &\sim \mathrm{Binomial}(1, \mathrm{P(status}_i = 1)) \nonumber\\
-\mathrm{logit(P(status}_i = 1)) &= \eta_i \nonumber\\
+\mathrm{status}_i &\sim \mathrm{Bern}(\mathrm{P_i}) \nonumber\\
+\mathrm{logit}(\mathrm{P_i}) &= \eta_i \nonumber\\
 \mathrm{\eta}_i &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i \nonumber\\
@@
 \begin{align}\label{sim:simple_mediate}
-z_i &\sim \mathrm{Binomial(1, P(z_i = 1))} \nonumber\\
-\mathrm{logit(P}(z_i = 1)) &= \eta_i \nonumber\\
+z_i &\sim \mathrm{Bern}(\mathrm{P_i}) \nonumber\\
+\mathrm{logit}(\mathrm{P_i}) &= \eta_i \nonumber\\
 \eta_i &= \beta_0 + \beta_{xz} x_i + \beta_{yz} y_i \nonumber\\

### 2022-05-05 manuscript/vareffects_plosone.tex
Autocommit

@@
 {\Large
-\textbf\newline{Describing the curves: uncertainty propagation and bias correction for predictor effects in simple and generalized linear (mixed) models} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+\textbf\newline{Describing the curves: uncertainty estimation and bias correction for predictor effects in simple and generalized linear (mixed) models} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
 }
@@
 For models that involve complex multiplicative interactions, additional non-focal predictors or nonlinear link functions, the estimated coefficient are not readily interpretable.
-A general way to summarize these kind of models is through predictor effects, which are not only dependent on choice of representative values of focal predictor but also sensitive to which values of non-focal predictors are chosen.
-The most common approach is generating predictor effects at a ``reference point'', usually the mean, of non-focal predictor, i.e., mean-based which could be considered as the effect of an ``average case'' in the population.
+A general way to summarize these kind of models is through predictions, which are not only dependent on choice of representative values of focal predictor but also sensitive to which values of non-focal predictors are chosen.
+The most common approach is generating predictions at a ``reference point'', usually the mean, of non-focal predictor, i.e., mean-based (centered) which could be considered as the effect of an ``average case'' in the population.
 In the presence of sources of bias such as additional non-focal predictors, nonlinear link functions, random effect terms, etc., mean-based approach generate predictor effects that are biased and may not be consistent with the observed quantities.
 An alternative is the whole-sample-based approach which estimates the average effect in the population.
-Moreover, isolated confidence intervals provide an alternative and a more clear way to describe uncertainty associated with the focal predictor of interest.
+Moreover, anchored effects provide an alternative and a more clear way to describe uncertainty associated with the focal predictor of interest.
 In addition to theoretical and methodical comparison, using simulation, we illustrate the two approaches and show that they can produce substantially different results and that whole-sample-based approach can not only produce estimates consistent with the observed values, but also appropriate for bias correction.
-We also present an alternative way, isolated confidence intervals, to describe uncertainties associated with these predictions.
+We also present an alternative way, anchored confidence intervals, to describe uncertainties associated with these predictions.
 
@@
 \includegraphics[width=0.8\textwidth]{pred_mediated_plots-figure6.pdf}
-\caption{{\bf Mean-based and whole-sample-based predictor effects for mediated effect.} These figures compare the mean-based and whole-sample-based predictor effect of $x$ from a mediated, $z \sim x + y$, and non-mediated, $z \sim x$, models.
+\caption{{\bf Mean-based and whole-sample-based predictions for mediated effects.} These figures compare the mean-based and whole-sample-based predictions for the mediated (multivariate) and non-mediated (univariate) models.
 A: In the absence of mediator variable, both mean-based and whole-sample-based approaches predictors are identical since there is no any other additional (non-focal predictors) sources of bias.
-Consequently, in both approaches, the average predicted probability is very close to the observed proportion.
-B: When the mediator variable is included, there is no direct effect of $x$ on $z$ and as a result the predictor effect curve does not align with the observations.
+Consequently, for both approaches, the average predicted probability is very close to the observed proportion.
+B: When the mediator variable is included, there is no direct effect of $x$ on $z$ and as a result the predictions do not necessarily align with the observations.
 However, the whole-sample-based approach still closely approximates the average proportion in the simulated data.
@@
 The argument and results we present in this paper supports a greater need for a shift on focus on what and how to present predictions from generalized models.
-For example, we believe that isolated confidence intervals could provide more clarity concerning the uncertainty due to the predictor of interests as opposed to the conventional way of incorporating everything.
+For example, we believe that effect-styled confidence intervals could provide more clarity concerning the uncertainty due to the predictor of interests as opposed to the conventional way of incorporating everything.
 

### 2022-05-13 manuscript/abstract.tex
Refactoring manuscript.tex

--- 2022-05-05 manuscript/vareffects_plosone.tex
@@
-% Template for PLoS
-% Version 3.5 March 2018
-
-% % % % % % % % % % % % % % % % % % % % % %
-%
-% -- IMPORTANT NOTE
-%
-% This template contains comments intended 
-% to minimize problems and delays during our production 
-% process. Please follow the template instructions
-% whenever possible.
-%
-% % % % % % % % % % % % % % % % % % % % % % %
-
-% Once your paper is accepted for publication, 
-% PLEASE REMOVE ALL TRACKED CHANGES in this file 
-% and leave only the final text of your manuscript. 
-% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
-% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
-%
-%
-% There are no restrictions on package use within the LaTeX files except that 
-% no packages listed in the template may be deleted.
-%
-% Please do not include colors or graphics in the text.
-%
-% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
-%
-% % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- FIGURES AND TABLES
-%
-% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
-%
-% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
-% - Figures should be uploaded separately from your manuscript file. 
-% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
-% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
-% For figure citations, please use "Fig" instead of "Figure".
-% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
-%
-% Tables should be cell-based and may not contain:
-% - spacing/line breaks within cells to alter layout or alignment
-% - do not nest tabular environments (no tabular environments within tabular environments)
-% - no graphics or colored text (cell background color/shading OK)
-% See http://journals.plos.org/plosone/s/tables for table guidelines.
-%
-% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
-%
-% IMPORTANT
-% Below are a few tips to help format your equations and other special characters according to our specifications.
-For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
-%
-% For inline equations, please be sure to include all portions of an equation in the math environment.
-For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
-%
-% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
-%
-% Please add line breaks to long display equations when possible in order to fit size of the column. 
-%
-% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
-%
-% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
-%
-% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
-%
-% % % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Please contact latex@plos.org with any questions.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-
-\documentclass[10pt,letterpaper]{article}
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
-
-% amsmath and amssymb packages, useful for mathematical formulas and symbols
-\usepackage{amsmath,amssymb}
-
-% Use adjustwidth environment to exceed column width (see example table in text)
-\usepackage{changepage}
-
-% Use Unicode characters when possible
-\usepackage[utf8x]{inputenc}
-
-% textcomp package and marvosym package for additional characters
-\usepackage{textcomp,marvosym}
-
-% cite package, to clean up citations in the main text. Do not remove.
-\usepackage{cite}
-
-% Use nameref to cite supporting information files (see Supporting Information section for more info)
-\usepackage{nameref,hyperref}
-
-% line numbers
-\usepackage[right]{lineno}
-
-% ligatures disabled
-\usepackage{microtype}
-\DisableLigatures[f]{encoding = *, family = * }
-
-% color can be used to apply background shading to table cells only
-\usepackage[table]{xcolor}
-
-% array package and thick rules for tables
-\usepackage{array}
-
-% create "+" rule type for thick vertical lines
-\newcolumntype{+}{!{\vrule width 2pt}}
-
-% create \thickcline for thick horizontal lines of variable length
-\newlength\savedwidth
-\newcommand\thickcline[1]{%
-  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-  \cline{#1}%
-  \noalign{\vskip\arrayrulewidth}%
-  \noalign{\global\arrayrulewidth\savedwidth}%
-}
-
-% \thickhline command for thick horizontal lines that span the table
-\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-\hline
-\noalign{\global\arrayrulewidth\savedwidth}}
-
-% Remove comment for double spacing
-%\usepackage{setspace} 
-%\doublespacing
-
-% Text layout
-\raggedright
-\setlength{\parindent}{0.5cm}
-\textwidth 5.25in 
-\textheight 8.75in
-
-% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
-% Captions will be left justified
-\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-\renewcommand{\figurename}{Fig}
-
-% Use the PLoS provided BiBTeX style
-\bibliographystyle{plos2015}
-
-% Remove brackets from numbering in List of References
-\makeatletter
-\renewcommand{\@biblabel}[1]{\quad#1.}
-\makeatother
-
-% Header and Footer with logo
-\usepackage{lastpage,fancyhdr,graphicx,grffile}
-%\pagestyle{myheadings}
-\pagestyle{fancy}
-\fancyhf{}
-%\setlength{\headheight}{27.023pt}
-%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
-\rfoot{\thepage/\pageref{LastPage}}
-\renewcommand{\headrulewidth}{0pt}
-\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
-\fancyheadoffset[L]{2.25in}
-\fancyfootoffset[L]{2.25in}
-\lfoot{\today}
-
-%% Include all macros below
-
-\newcommand{\fix}{{\bf \textcolor{red}{FIXME}}}
-\newcommand{\JD}[1]{{\color{blue} \emph{#1}}}
-\newcommand{\bmb}[1]{{\color{RubineRed!70!} \emph{#1}}}
-\newcommand{\pkg}[1]{\textbf{#1}}
-
-\def\code#1{\texttt{#1}}
-\let\proglang=\textsf
-
-\newcommand{\bX}{{\mathbf X}}
-\newcommand{\bZ}{{\mathbf Z}}
-\newcommand{\bbeta}{{\boldsymbol \beta}}
-\newcommand{\boldeta}{{\boldsymbol \eta}}
-\newcommand{\boldmu}{{\boldsymbol \mu}}
-
-\newcommand{\nset}[1]{#1_{\{n\}}}
-\newcommand{\yref}{y_{\textrm{ref}}}
-\newcommand{\cdist}{{D(\nset{x}|x_f)}}
-\newcommand{\cdistprime}{{D(\nset{x}|x_{f'})}}
-\newcommand{\xfprime}{x_{f'}}
-
-\newcommand{\yE}{{\mathrm E}}
-\let\over=\overline
-
-\newcommand{\bxo}{\bX^{\star}}
-\newcommand{\bxof}{\bX_f^{\star}}
-\newcommand{\ba}{{\mathbf A}}
-\newcommand{\bao}{{\mathbf A}_0}
-\newcommand{\baf}{{\mathbf A_f}}
-\newcommand{\bafc}{{\mathbf A_{\bar{f}}}}
-
-%% END MACROS SECTION
-
-\begin{document}
-\vspace*{0.2in}
-
 % Title must be 250 characters or less.
@@
 \section*{Author summary}
-
-\fix
-
-\linenumbers
-
-% Use "Eq" instead of "Equation" for equation citations.
-\section*{Introduction}
-
-In many applications which use simple or generalized linear (mixed) models, outcome predictions or probabilities, are often of interest.
-Predictions provide a great way to summarize what the regression model is telling us and very useful for interpreting and visualizing model estimates.
-For example, in logistic regression models, the coefficient estimates are usually not easy to interpret and less informative.
-Since logistic models are nonlinear (due to the nonlinear link function), in multivariate models with interactions, the magnitude of the effect of change in the outcome depends on the values of the predictor of interest and other predictors.
-Hence, the conclusions one can make about the estimated effects greatly depends on how well we choose the values of the other predictors, i.e., \emph{non-focal} predictors.
-
-Both simple and generalized linear (mixed) models (GL(M)Ms) can examine very complex relationships, including nonlinear relationships between response and predictors, interactions between predictors (and via splines for example), and nonlinear transformations via link functions due to their flexibility.
-This flexibility comes at a cost, for example, complex multivariate models may risk misinterpretation, and miscalculation of quantities of interest.
-Also, coefficient estimates of models involving nonlinear link functions or interactions lose their direct interpretation \cite{leeper2017interpreting}, meaning that interpretation of derived quantities from these estimates requires some understanding of the specified model.
-An alternative is to explore the \emph{effects} of predictors on the predictions or probabilities of the outcome at various level of the predictor of interest -- \emph{prediction and effects} \cite{fox2009effect, leeper2017package, lenth2018package}.
-For example, as a way to plan, public health officials may want to know the effects of household income, wealth index, etc, on the predicted probability of having improved water services among slum dwellers.
-
-When visually presented, prediction and effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind of transformations on the predictors whose estimates are usually, but not always, a subject to less clarity of interpretation.
-Further, generating predictor and effect plots together with the associated confidence intervals for regression models has a number of challenges.
-In particular:
-\begin{enumerate}
-\item choice of representative values of \emph{focal} predictor and the \emph{reference point} for non-focal predictors especially in multivariate models
-\item  propagation of uncertainty -- appropriate choice of \emph{anchor} for computing confidence intervals; and how to incorporate the uncertainty due to non-focal predictors 
-\item bias in the expected mean prediction induced by the nonlinear transformation of the response variable (especially in GL(M)Ms)
-\end{enumerate}
-
-The most common way of dealing with the first challenge is taking unique levels of the focal (predictor of interest) predictor if discrete or taking appropriately sized quantiles (or bins) if continuous, and then calculating the predictions while holding non-focal (other predictors other than the focal predictor) predictors at their reference point (e.g., means) \cite{hanmer2013behind}.
-This generates -- \emph{predictions}, \emph{predictor effects}\cite{fox2009effect}, \emph{marginal predictions} \cite{leeper2017package} or \emph{estimated marginal means} \cite{lenth2018package}.
-In this article, we refer to this quantity as predictions since it should, for example, tell us what we would expect the response to be at a particular value or level of the predictor, for an ``average case''.
-More specifically, we compute the expected outcome by meaningfully holding non-focal predictors constant (or averaged in some meaningful way) while varying the focal predictor, with the goal that the outcome expected prediction represents how the model responds to the changes in the focal predictor.
-On the other hand, effects differs with predictions in the sense that the former is about how we describe the uncertainty around the latter and depends on the anchor.
-
-The commonly used \proglang{R} software packages (\pkg{emmeans} and \pkg{effects}) for generating predictions, by default, use the average of non-focal predictors as the reference point.
-However, there are a number of choices one can make when considering this approach -- for example, in the presence of interaction, averaging the interactions (averaging product of interacting model variables) versus product of the averages of the interacting model variables (default for \pkg{emmeans} and \pkg{effects}).
-We claim that neither of these two approaches is the most appropriate but we show that averaging the interaction closely matches the observed values.
-To illustrate this, consider models~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher} below, with $x_1$ as the focal predictor:
-
-%
-\begin{align}
-y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \epsilon \label{eq:simple_inter_higher_no_interaction}\\
-y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \beta_{23}x_2x_3 + \epsilon \label{eq:simple_inter_higher}
-\end{align}
-%
-
-We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$ and $\beta_{23} = 5$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=0.9\textwidth]{justify_preds.inter.pdf}
-\caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} predictions for $x_1$ on $y$ for models with and without interactions.}
-The horizontal blue, green and black lines are the mean predictions, i.e., $\bar{\hat{y}}$, the red ones are the means of the simulated $y$, i.e., $\bar{y}$, while the vertical dotted grey lines are the means of the corresponding focal predictors.
-The grey points are the binned simulated $y$.
-The trend lines represents the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other predictors at their average for the three mthods and the ``true'' predictions based on the simulation parameters.
-A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the simulated in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$.
-Similarly, the predictions based on the three approaches closely matches the ``true'' predictions.
-B: Even with the simple interaction between non-focal predictors, we start seeing deviation in both predicted mean, $\bar{\hat{y}}$, and overall predictions from the simulated, $\bar{y}$, and ``true'' predictions, respectively, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not, the proposed \pkg{varpred}.}
-\label{fig:justify_plots}
-\end{figure}
-
-In the absence of interaction (model~\ref{eq:simple_inter_higher_no_interaction}), the three approaches produce similar estimates, which match the simulated values, Fig~\ref{fig:justify_plots}A.
-However, in the presence of interaction, even as simple as the one in model~\ref{eq:simple_inter_higher}, the estimates starts to differ.
-In particular, \pkg{emmeans} and \pkg{effects} give similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is very close to the simulated average ($\bar{y}$), Fig~\ref{fig:justify_plots}B.
-To generate Fig~\ref{fig:justify_plots}A, all the three packages averages non-focal predictors $x_2$ and $x_3$.
-On other hand, to generate Fig~\ref{fig:justify_plots}B, the difference in the estimates lies on how each of the packages average the interaction term ($x_2x_3$).
-In particular \pkg{emmeans} and \pkg{effects} uses \emph{input variables} to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses \emph{model variables} to compute $\over{x_2x_3}$.
-
-Sometimes, one may be interested in the uncertainties associated with the focal predictor of interest only, excluding other uncertainties due to other non-focal predictors -- effects.
-However, currently, the two packages do not provide straightforward way do achieve this.
-To illustrate this, we generate predictions of $x_2$ from model~\ref{eq:simple_inter_higher_no_interaction} together with the associated $95\%$ confidence bands, as shown in Fig~\ref{fig:justify_ci_plots}.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=0.9\textwidth]{justify_preds.isolate.pdf}
-\caption{{\bf A comparison of conventional $95\%$ confidence bands and effects.} The description of horizontal, vertical and trend lines remain the same as above.
-The wider dotted blue curves overlaying the green curves corresponds to the conventional confidence bands around the predcitions from \pkg{emmeans} and \pkg{effects}, while the narrower black curves crossing at the mean of the focal predictor, i.e., the model center, corresponds to the effects of the focal predictor from \pkg{varpred}.
-For simple models, the isolated confidence bands crosses at the model center.
-The wider curves incorporates not only the uncertainties due to intercept term but also other non-focal predictors.
-On the other hand, effects (narrower crossing lines) only take into account the main effect uncertainty of the focal predictor.}
-\label{fig:justify_ci_plots}
-\end{figure}
-
-For \pkg{emmeans} and \pkg{effects}, the confidence bands are much wider because they include uncertainties associated with the intercept and non-focal predictors, but narrower and crosses at the mean of the focal predictor, i.e., \emph{center point or mean-based}, in \pkg{varpred}.
-In other words, with \pkg{varpred}, we are able to generate effects indicating zero uncertainty at the value of the focal predictor we are more certain about, i.e., mean of the model variable corresponding to the focal predictor (center point or mean-based).
-For simple models, the point where the confidence bands crosses is the model center and it corresponds to anchor we choose, in this case, the mean of the focal predictor.
-
-When dealing with nonlinear link functions or models with multiple non-focal predictors, the correct predictions, for example, are even much harder to estimate.
-One approach is to make predictions on the transformed scale (linear predictor scale), and then back-transform to the original scale.
-However, the back-transformation may either result in biased predictions or requires some approximation.
-In particular, bias in expected mean prediction induced by nonlinear transformation of the response variable can lead inaccurate predictions.
-An alternative to mean-based reference point, i.e., averaging non-focal predictors, is the \emph{whole-sample-based} approach, discussed in detail later, which involves computing the prediction over the population of non-focal predictors and then averaging across the values of the focal predictor \cite{hanmer2013behind}.
-
-The main purpose of this article is to discuss and implement various approaches for computing predictions and provide an alternative method for computing the associated confidence intervals, i.e., effects.
-We further explore and demonstrate, using simulated data, approaches for correcting bias in predictions for GL(M)Ms involving nonlinear link functions.
-
-\section*{Quantities of interest}
-
-Several quantities of interest may be derived from regression models.
-The first one is the coefficient estimates.
-Others are \emph{input and model variables}, \emph{model matrix}, \emph{model center}, \emph{reference point} and \emph{anchor}.
-
-\subsection*{Input variables}
-
-These are scientific variables underlying an inference or exploration. The focal and non-focal predictors we use for effects or predictions are an input variables.
-
-\subsection*{Model matrix}
-
-Model matrix refers to the design matrix whose rows include all combination of variables appearing in the interaction terms, along with the typical values of the focal and non-focal predictors.
-
-\subsection*{Model variables}
-
-These are variables that represent columns in the model matrix.
-Each input variable may correspond to one or more model variables.
-In particular, variables with more than two categories, or variables modeled with a spline or polynomial response, will correspond to more than one model variables.
-
-\subsection*{Model center}
-
-A point corresponding to a column-wise mean of the model matrix (the mean of one or more model variables).
-We also refer to this as mean-based approach.
-The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable -- the case in Fig~\ref{fig:justify_plots}.
-In simple linear models (without) interaction, the input variable can be used as a proxy.
-
-\subsection*{Reference point}
-
-Corresponds to the value or values chosen for non-focal predictors, when estimating the predictions.
-Typically the center point (mean-based), but can instead be a whole-sample of quantiles or observations.
-Appropriately choosing the reference point is very crucial in generating ``correct'' predictions especially when dealing with non-linear link functions.
-
-\subsection*{Anchor}
-
-The value chosen for the focal predictor when estimating effect confidence intervals (anchor choice does not affect the main prediction estimates).
-Typically chosen as the center point of the model variables corresponding to the focal input variable but other sensible values can be chosen too.
-
-If we are trying to focus on effects, the anchor from which we calculate CIs becomes important.
-This is not true for the more conventional prediction plots: we can choose any anchor, e.g., zero-anchored, but model center seems like a natural, stable choice, i.e., center-anchored.
-In Fig~\ref{fig:justify_anchors}, we compare effects plots anchored at the minimum and center (mean) of the focal predictor.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=0.7\textwidth]{justify_anchors.combined.pdf}
-\caption{{\bf A comparison of min-based and center- (mean-) based anchors.} The center-based effects are based on the mean of the focal predictor while the min-anchored effects are based on the minimum of the focal predictor.
-As we mentioned earlier, the choice of the anchor do not affect the main estimates (central trend lines).}
-\label{fig:justify_anchors}
-\end{figure}
-
-If we have more than one model variables for our focal predictor, however, the model center does not correspond to any given value of the variable, and we don't get the nice crossing we get in simpler cases.
-See Fig~\ref{fig:pred_cubic_plots}.
-
-\subsection*{Prediction- vs. effect- styled plots}
-
-A prediction-style plot is about predicting observations (or the ensemble mean of observations) for a given value of the focal predictor.
-For this, we want to use the classic curved confidence intervals (or prediction intervals).
-A prediction-style plot will generally focus on \emph{total} effects, and so will often be suited for a univariate fit.
-An effect-style plot, however, is an attempt to visualize the effect of a focal predictor.
-In this case, we suggest the narrower confidence intervals.
-Here, it makes equal sense to think about the total effect of our focal predictor, from a univariate model, or the direct effects – after controlling for whatever – from a multivariate model.
-See Fig~\ref{fig:justify_ci_plots}.
-
-\subsection*{Predictions vs. effects}
-
-We would like to summarize by talking about predictions vs. effects. Predictions are the standard curves; they tell us what we expect to see if we know the value of the focal predictor.
-
-The lines inside (effects) are a little different: they are a good way to visualize the effect of the focal predictor.
-Unlike predictions, where the uncertainty captures the uncertainty in the whole model, the effects plot focuses on uncertainty in the effect of the focal predictor only -- thus, for example, if an effect is barely significant (P exactly 0.05), one end of the CI will be exactly a horizontal line for the effects plot, but not for the prediction plot.
-Also, the uncertainty curves for the effect, but not for the prediction, depend on the ``anchor''.
-
-\section*{Statistical background}
-
-To get an intuition of how conditioning on the mean values of non-focal predictors work, suppose we are interested in predictions at the levels of a particular predictor, i.e., focal, $x_f$, from the set of predictors.
-As a starting point, assume that the model has no interaction terms.
-The idea is to choose a reference point for the values of non-focal predictors.
-For example, fixing the values of non-focal predictor(s) at some typical values -- typically determined by averaging (for now) in some meaningful way, for example, arithmetic mean  for continuous and average over the levels of the factors for categorical non-focal predictors.
-We refer to this as centered or mean-based approach.
-The most convenient way to achieve this is by constructing $\bX^\star$ by averaging the columns of non-focal predictors in model matrix $\bX$, and together with appropriately chosen values of focal predictor.
-
-Consider a simple linear model with linear predictor $\eta = \bX\bbeta$ and let $g(\boldmu) = \boldeta$ be an identity link function (in the case of simple linear model), where $\boldmu$ is the expected value of response variable $y$.
-Let $\hat{\bbeta}$ be the estimate of $\bbeta$, together with the estimated covariance matrix $\boldsymbol{\Sigma} = V(\hat{\bbeta})$ of $\hat{\bbeta}$.
-Let $\mathbf{X^*}$ be the model matrix, inheriting most of its key properties, for example transformations on predictors and interactions from the model matrix, $\mathbf{X}$.
-Then the quantity $\hat{\boldeta}^\star = \bX^\star\hat{\bbeta}$ is the prediction with respect to the focal predictor in question \cite{fox2009effect}.
-
-An alternative formulation of $\hat{\boldeta}^\star$ involves expressing the linear predictor as the sum of the focal and non-focal predictors' linear predictor. In particular,
-
-\begin{align}\label{eq:eta_mean}
-\eta^\star(x_f, \nset{{\bar{x}^\star}}) &= \beta_f x_f + \sum \nset{\beta} \nset{{\bar{x}^\star}} \\
-\hat{y}_f  &= g^{-1} \left(\eta^\star(x_f, \nset{{\bar{x}^\star}})\right)
-\end{align}
-where $\nset{{\bar{x}^\star}}$ are the appropriately averaged entries of non-focal predictors and $x_f$ is a vector of values of the focal predictors for a particular observation.
-
-\subsection*{Dealing with higher order interactions}
-
-Higher order terms such as interactions, splines, polynomials, etc., can be within the focal predictor, within (between) non-focal predictor(s), between the focal and non-focal predictor(s).
-To distinguish the three, suppose the model which describes the hypothetical simulated household size simulation based on a number of socio-demographic factors is
-
-\begin{align}\label{sim:lm_cubic}
-\mathrm{hh~size}_i &= \beta_0 + \beta_{\mathrm{A_1}}\mathrm{Age}_i + \beta_{\mathrm{A_2}}\mathrm{Age}^2_i + \beta_{\mathrm{A_3}}\mathrm{Age}^3_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \epsilon_i \nonumber\\
-\end{align}
-
-In the first case, with $Age$ as the focal predictor, the interaction is within its polynomial terms (and 3 model variables associated with it).
-In this case, each of the polynomial terms (model variables) of the focal predictor are evaluated independently across the chosen levels of the focal predictor, $Age_i$.
-Specifically, the higher order terms are treated as additional columns of the model matrix evaluated with the same values of the focal predictor, while non-focal predictors are fixed at their reference point as discussed in the previous section.
-In this case, the predictions associated with $Age$ on the linear predictor scale becomes
-
-\begin{align*}
-\eta^\star(\mathrm{Age}_i, \nset{{\over{\mathrm{Wealthindex}}}}) &= \beta_0 + \beta_{\mathrm{A_1}}\mathrm{Age}_i + \beta_{\mathrm{A_2}}\mathrm{Age}^2_i + \beta_{\mathrm{A_3}}\mathrm{Age}^3_i\\
-	& + \beta_{\mathrm{W}}\over{\mathrm{Wealthindex}}
-\end{align*}
-
-In the second case, with $Wealthindex$ as the focal predictor, the higher order interactions in non-focal predictor, $Age$, are treated simply as additional columns in the variable space (of model matrix) and appropriate choice of reference point applies just like in the models without interactions.
-For instance, in mean-based approach, we simply average all non-focal interaction terms.
-Thus
-
-\begin{align*}
-\eta^\star(\mathrm{Wealthindex}_i, \nset{{\{\over{\mathrm{Age}}, \over{\mathrm{Age}^2}, \over{\mathrm{Age}^2}\}}}) &= \beta_0 + \beta_{\mathrm{A_1}}\over{\mathrm{Age}} + \beta_{\mathrm{A_2}}\over{\mathrm{Age}^2} + \beta_{\mathrm{A_3}}\over{\mathrm{Age}^3}\\
-	& + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i
-\end{align*}
-
-Lastly, consider model~\ref{eq:simple_inter_higher}, with the interactions between the focal predictor, $x_2$, and non-focal predictor, $x_3$.
-In this case, the non-interacting predictors are treated as non-focal predictors, with appropriate reference points.
-However, for the interacting predictors, we first choose representative values for the focal predictor and some particular values of the interacting predictor similar to the way we choose values for the focal predictor or predetermined.
-In our example, suppose we pick $i$ and $j$ unique values of the focal predictor, $x_2$ and interacting predictor $x_3$, respectively, then the prediction on linear predictor scale is given by
-
-\begin{align*}
-\eta^\star(x_{2i}, x_{3j}, \nset{{\bar{x}^\star}}) = \beta_0 + \beta_1 \bar{x}_1 + \beta_2x_{2i} + \beta_3x_{3j} + \beta_{23}x_{2i}x_{3j}.
-\end{align*}
-
-In general, our formulation, even for more complicated interactions, follow these three basic principles -- interaction within focal predictor, interaction between non-focal predictors and interaction between focal and non-focal predictors.
-
-\section*{Uncertainty estimation}
-
-What about the confidence intervals (CI)?
-The limits of the confidence intervals are points, not mean values.
-In principle, every value of focal predictor has a different CI.
-The conventional way to compute variances for predictions is $\sigma^2 = \textrm{Diag}(\bX^\star \boldsymbol{\Sigma} \bX^{\star\top})$ \cite{lenth2018package, fox2009effect}, so that the confidence intervals are $\eta \pm q\sigma$, where $q$ is an appropriate quantile of Normal or t distribution.
-This generates predictions-styled CIs and incorporates all the uncertainties -- including the uncertainty due to the intercept and non-focal predictors.
-But what if we are only interested in the uncertainty as a result of the focal predictor, so that the confidence intervals are $\eta \pm q \sigma_f$?
-We call this anchored CIs and generates effects-styled CIs.
-
-Currently, commonly used \proglang{R} packages for constructing predictions, by default, do not exclude the uncertainties resulting from non-focal predictors when computing the CIs.
-A non-trivial way to exclude uncertainties associated with non-focal predictors in some of these packages is to provide a user defined variance-covariance matrix with the covariances of non-focal terms set to $0$ -- \emph{zeroing-out} variance-covariance matrix.
-This only works when the input predictors are \emph{centered} prior to model fitting, in case of numerical predictors, and even much complicated when the predictors are categorical.
-We first describe the variance-covariance based approach and then discuss our proposed method which is based on the model center and does not require input predictors to be scaled prior to model fitting.
-
-\subsection*{Variance-covariance}
-
-The computation of $\hat{\eta}^\star$ remains the same as described above.
-However, to compute $\sigma$, $\boldsymbol{\Sigma}$ is modified by \emph{zeroing-out} (the variance-covariance of all non-focal predictors are set to zero) variances of non-focal terms.
-Although this is the simplest approach, it requires centering of continuous, i.e., $x_c = x - \bar{x}$, predictors prior to model fitting and proper way to average categorical predictors.
-
-\subsection*{Model center based}
-
-Let $\bxo$ be a centered model matrix containing appropriately chosen values of focal predictors and centered non-focal predictors, and let $\ba$ be the anchor matrix, with same dimensions as $\bxo$ and values in each columns repeated $n$ times.
-Also, let $\baf$ be the column of $\ba$ corresponding to the focal predictor, the rest of the columns corresponds non-focal predictors with same values as in $\bxo$.
-Any appropriate values (within the range of focal values) can be chosen for $\baf$ but for model center (center-anchored), we have $\bafc$ .
-Thus $\sigma_f^2 = \textrm{Diag}((\bxo - \bafc) \boldsymbol{\Sigma} (\bxo - \bafc)^\top)$.
-We can see that $\forall ~\bxof=\baf$, $\sigma_f^2 = \boldsymbol{0}$.
-This is true for the center-anchored provided $\bxof=\bafc$.
-In other words non-focal terms (model variables) in $\bX^{\star}_{c} = (\bxo - \bafc)$ are all zero in simple models without interactions but are isolated (narrower) around the model center in models involving complex interactions.
-In addition, the computation of $\bX^{\star}_c$ impacts only on the intercepts and non-focal model variables, i.e., the slopes and variance of the focal predictors are not affected.
-This means that we can still generate effects without necessarily centering the predictors prior to model fitting.
-
-%% \subsection*{Is center-anchored better?}
-%% 
-%% We want to compare the amount of uncertainty when there is/no anchor.
-Let $\bao = \boldsymbol{0}$ be the anchoring matrix with all entries in $0$s, and let $\bafc$ be the center-anchored matrix defined above.
-Then entries in $(\bxo - \bao) \geq (\bxo - \bafc)$.
-%%
-
-\section*{Bias correction}
-
-In many applications, it usually important to report the estimates that reflect the expected values of the untransformed response.
-This is not a major issue in simple linear models.
-However, when dealing with nonlinear link functions, it is even harder to generate correct predictions that reflect this due to the bias in the expected mean induced by the nonlinear transformation of the response variable.
-In such cases, bias correction is needed when back-transforming the predictions to the original scales.
-Most common approach for bias-adjustment is second-order Taylor approximation \cite{lenth2018package, duursma2003bias}.
-Another potential source of bias comes from additional non-focal predictors, especially in models with nonlinear link functions.
-Here, we describe and implement a different approach, whole-sample-based approach for bias correction.
-
-\subsection*{Whole-sample-based approach for bias correction}
-
-The most precise (although not necessarily accurate!) way to predict is to condition on values of the focal predictor and make predictions for all observations (members of the population) \cite{hanmer2013behind}.
-A key point is that the nonlinear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population to a one-dimensional distribution of $\eta^\star(x_f, \nset{x})$, which is a function of focal predictor and the observed values of non-focal predictors, as opposed to the definition in Equation~\ref{eq:eta_mean}.
-More specifically:
-
-\begin{itemize}
-\item compute linear predictor associated with non-focal predictors, $\nset{\eta} = \sum \nset{\beta} \nset{x}$
-\item compute linear predictor associated with the focal predictors, $\eta_{jf} = \sum{\beta_f x_{jf}}$
-\item for every value of the focal predictor, $\eta_{jf}$:
-
-\begin{align}\label{eq:pop_eta} 
-\eta_j^\star(\eta_{jf}, \nset{\eta})  &= \eta_{jf} + \nset{\eta} \nonumber \\
-&= \eta_j^\star(x_f, \nset{x})
-\end{align}
-\end{itemize}
-
-Once Equation~\ref{eq:pop_eta} is computed, one can back-transform the estimates to the original scale and the average over the levels of the focal predictors, $j$:
-
-\begin{align}\label{eq:pop_response} 
-\hat{y}_f  &= \textrm{mean} ~ g^{-1} \left(\eta_j^\star(x_f, \nset{x})\right)
-\end{align}
-
-We make similar adjustments to compute the variances of the predictions at every level of the focal predictor:
-
-\begin{align}
-\sigma_{jf}^2 = \textrm{Diag}(\bX^\star_{jc} \boldsymbol{\Sigma} \bX^{\star\top}_{jc})
-\end{align}
-where $\bX^{\star}_{jc} = \{\bX_{jf}^\star, \nset{{\bX}^\star} - \nset{{\bar{\bX}}^\star}\}$ and
-
-\begin{align}
-\mathrm{CI}_f = \mathrm{mean} ~ g^{-1} \left(\eta_j^\star(x_f, \nset{x}) \pm q\sigma_{jf}\right)
-\end{align}
-
-For models with random effects components, we make further adjustment to correct for bias induced by the random effects terms.
-In the population approach, we treat the random effects terms as additional non-focal predictors and simply make adjustment to Equation~\ref{eq:pop_eta}.
-In particular
-
-\begin{align}\label{eq:pop_eta_re} 
-\tau &= \bZ b \nonumber \\
-\eta_j^\star(x_f, \nset{x}, \tau)  &= \eta_j^\star(x_f, \nset{x}) + \tau
-\end{align}
-where $\bZ$ and $b$ are the design matrix and a vector of random effects, respectively.
-
-\section*{Mean-based vs. whole-sample-based}
-
-As mentioned above, the two common choices to generate predictions are: 1) setting non-focal predictors to their mean -- centered- (mean-) based; and 2) using the entire population of non-focal predictors -- whole-sample-based.
-If the link function is nonlinear and/or the model has complex higher order interactions, the average of the predictions evaluated at the mean of non-focal predictors and the average of predictions evaluated at the population level (and then averaged) of non-focal predictor are not equivalent, i.e.,
-
-\begin{align}\label{eq:compare_anchored_pop_based}
-g^{-1} \left(\eta_j^\star(\bar{x}_f, \nset{\bar{x}})\right) \neq \frac{1}{n} \sum_{i=1}^n{ g^{-1} \left(\eta_j^\star(\bar{x}_f, \nset{x})\right)}.
-\end{align}
-
-If $g()$ is strictly concave or strictly convex, one can use Jensen's inequality to determine which side of Equation~\ref{eq:compare_anchored_pop_based} is smaller or larger.
-However, for non-fully convex or concave link function, one can use Taylor series expansion to approximate the range over which the two are smaller or smaller than the other \cite{hanmer2013behind}.
-
-As stated above, in simple linear models without interactions, the effect is constant, so both approaches yield similar results.
-However, picking a single value, e.g., mean of the predictor, on which to draw conclusions about the effect can be problematic, unrealistic or not contained in or representative of the population.
-In addition, the mean-based approach fails to use every values of non-focal predictors hence not utilizing the full potential of the information contained in the data.
-This may limit the inferences we can make about the entire population.
-In general, the mean-based approach provides the predictions of an average case, whereas, the whole-sample-based approach, summarizes the predictions over the entire population -- and in some applications, the effect of an average case might not be generalizable to the entire population, especially, if the average does not represent the population.
-The whole-sample-based focuses on specific observations since the prediction is first obtained for each observation and then averaged across the levels of the focal predictor.
-
-Another potential concern with the mean-based approach arises in situations where direct naive use leads to rare or meaningless basis for generalization.
-For example, setting dummy categorical variables in the model matrix to their means, which, by default, sets them to their sample means or observed proportions.
-For example, a dummy variable for christian household heads may be set to $0.2$, translating to prediction for an household head who is $20\%$ christian.
-This problem may extend to models with other complex interaction terms \cite{hanmer2013behind}.
-
-Whole-sample-based approach is not entirely foolproof.
-For instance, similar to the mean-based approach, in the case of continuous focal predictors, choosing the representative values of the focal predictors can be very challenging especially if the cases are not evenly distributed around the minimum and the maximum values or within some subgroups defined in the population.
-In addition, whole-sample-based approach can be very computationally intensive for large datasets.
-
-\section*{Simulation examples}
-
-We now illustrate the construction of prediction-styled and effect-styled plots and also demonstrate that the mean-based and whole-sample-based approaches produce different results in models with nonlinear link functions and/or additional source(s) of potential bias.
-In addition, we demonstrate that whole-sample-based approach can be used for bias correction.
-
-\subsection*{Predictions-styled and effects-styled plots}
-
-As mentioned above, in some applications, one may be interested in uncertainties associated with the focal predictor only.
-In that case, it is important to exclude all the uncertainties as a result of non-focal predictors and only show confidence intervals describing the predictor of interest.
-To illustrate this, we consider the simulation model described in \nameref{S1_Appendix}.
-
-We start by generating two predictions from the fitted model \nameref{S1_Appendix}: 1) one with cubic polynomial of \code{age} as a focal predictor; and 2) one with a simple focal predictor (non-polynomial) of \code{Wealthindex}.
-Then we compared non-anchored (more of predictions) and mean-anchored (effects) confidence intervals around the predicted values (see Fig~\ref{fig:pred_cubic_plots}).
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=0.8\textwidth]{pred_cubic_plots-figure4.pdf}
-\caption{{\bf Prediction-styled and effect-styled plots for cubic polynomial and simple focal predictors.} A: The focal predictor is a cubic polynomial.
-B: The focal predictor is not a polynomial function, implying that the complex interaction is in non-focal predictor.
-In both cases, the mean-anchored effects are narrower than the non-anchored (predictions), but crosses at the mean of the focal predictor (model center) only in the case of simple (non-polynomial) focal predictor -- B.
-In the case of cubic polynomial focal predictor, the model center is not a point in the predictor space, i.e., mean of the focal predictor, but a combination of all the model variables in the cubic polynomial, consequently, mean-anchored curves do not cross at a point -- A.
-The horizontal black and red dotted lines are the observed and predicted average household size, i.e., $\over{hh~size}$ and $\over{\widehat{hh~size}}$, respectively; the inner red dotted curves and crossing lines are the anchored confidence bands for A and B, respectively, while the outer blue curves are the non-anchored confidence bands in both cases.
-The central curve and trend line, in A and B, respectively, are the predicted household size based on age and wealth index, respectively.}
-\label{fig:pred_cubic_plots}
-\end{figure}
-
-In the absence of complex higher order interaction terms or transformations on the focal predictor in simple linear models, we would expect, at the model center, the observed average, $\over{hh~size}$, and the average of the predictions, $\over{\widehat{hh~size}}$, to be identical and cross with the mean-anchored confidence bands (effects), at the model center (see Fig~\ref{fig:pred_cubic_plots}B).
-On the other hand, if the focal predictor is characterized by complex interactions or any other form of transformation, the model center is not necessarily a point in the predictor space.
-In this case, the mean-anchored confidence bands will be narrower than the non-anchored ones but not necessarily crossing at the model center (see Fig~\ref{fig:pred_cubic_plots}A).
-
-\subsection*{Bias correction}
-
-Mean-based and whole-sample-based approaches can produce very different results in models with nonlinear link functions with additional sources of potential bias such as additional non-focal predictor(s), random effect terms or complex interactions.
-To demonstrate this, we considered a two predictor binary outcome simulation described in \nameref{S2_Appendix}, such that the age effect is way smaller than that of the wealth index, and compared their effects on the predicted probability of improved water quality as shown in Fig~\ref{fig:pred_bin_plots}.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=0.8\textwidth]{pred_bin_plots-figure5.pdf}
-\caption{{\bf Mean-based and whole-sample-based predictions.} The two approaches produce relatively different estimates in each of the predictors but the difference is more pronounced when the effect of the particular focal predictor has no strong effect -- \emph{on the left}.
-In both cases, \emph{right} and \emph{left}, the average predicted probability of improved water based on whole-sample-based approach is very close to the observed proportion in the simulated data.
-The horizontal blue, black and the red dotted lines are the average mean-based, whole-sample-based predicted probability and observed proportion of improved water quality, respectively.
-The vertical dotted black line represents the mean of the focal predictor, and the point at which it crosses the red dotted line represents the expected ``perfect'' prediction at the model center.
-The grey points are binned observations -- observed proportions of improved water quality in each bin.
-We would expect all the curves, horizontal, and vertical lines to cross at the model center but due the nonlinear averaging (at least in the bias corrected whole-sample-based approach) of the predictions on the response scale, this may not always be the case.}
-\label{fig:pred_bin_plots}
-\end{figure}
-
-If there was no effect of nonlinear averaging, then we would expect the observed proportion and the average of the predicted probabilities to cross at the model center as we see in Fig~\ref{fig:pred_cubic_plots}B.
-However, the differences we see in Fig~\ref{fig:pred_bin_plots} are due to nonlinear averaging since both observed status and predicted probabilities are averaged on the response scale as opposed to link scale.
-In particular, if the range of values are bigger than $0.5$ (seemingly the case here), then we would expected the averages to be slightly higher than what we would expect at the model center, and vice versa.
-
-\subsection*{Mediated effect}
-
-Using simple indirect mediation simulation model described in \nameref{S3_Appendix}, fitted two models -- \emph{non-mediated} which models \emph{z} as a function of \emph{x} (univariate model) and the \emph{mediated} which models \emph{z} as a function of both \emph{x} and \emph{y} (multivariate model).
-In both cases, we compared mean-based and the whole-sample-based predictions, as shown in Fig~\ref{fig:pred_mediated_plots}.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=0.8\textwidth]{pred_mediated_plots-figure6.pdf}
-\caption{{\bf Mean-based and whole-sample-based predictions for mediated effects.} These figures compare the mean-based and whole-sample-based predictions for the mediated (multivariate) and non-mediated (univariate) models.
-A: In the absence of mediator variable, both mean-based and whole-sample-based approaches predictors are identical since there is no any other additional (non-focal predictors) sources of bias.
-Consequently, for both approaches, the average predicted probability is very close to the observed proportion.
-B: When the mediator variable is included, there is no direct effect of $x$ on $z$ and as a result the predictions do not necessarily align with the observations.
-However, the whole-sample-based approach still closely approximates the average proportion in the simulated data.
-The horizontal blue, black and the red dotted lines are the average mean-based, whole-sample-based predicted probability and observed proportion, respectively.
-The vertical dotted black line represents the mean of the focal predictor, and the point at which it crosses the red dotted line represents the expected ``perfect'' prediction at the model center.
-The grey points are the binned observations.}
-\label{fig:pred_mediated_plots}
-\end{figure}
-
-From Fig~\ref{fig:pred_mediated_plots}A, we see what we would expect in the absence of additional sources of bias, even though in the simulation, the effect of $x$ on $z$ is mediated through $y$.
-By ignoring $y$ in the model, we are still able to capture the effect of $x$ and closely match the observed values using both approaches.
-However, if there were additional non-focal predictors, we would expect to see the differences similar to those in Fig~\ref{fig:pred_bin_plots}.
-Including both $y$ and $x$ ``dilutes'' the direct effect of $x$ on $y$ and as a result, our prediction do not necessarily match the observed binned observations (see Fig~\ref{fig:pred_mediated_plots}B).
-
-\section*{Discussion}
-
-Our simulations examples majorly focused on simple linear and logistic models due their wide range of usage and application.
-In addition, these models act as a starting point for building other complex models, including mixed effect models and models with categorical predictors.
-However, the logic for extension of approaches to more complex models, including other forms of nonlinear link functions is very straightforward.
-In addition, our \proglang{R} package implementation already extends to and supports most of the nonlinear link functions and mixed model framework, including multivariate binary outcome models.
-
-Although commonly used \proglang{R} software packages, by default, implements mean-based approach, our simulation results demonstrated that the whole-sample-based approach has a potential of yielding results which are more consistent with the observed data.
-We would, therefore, argue that the use of mean-based approach should have some theoretical justification, especially in complex models.
-
-\section*{Conclusion}
-
-Generating outcome predictions or predicted probabilities from simple and generalized linear (mixed) models is not only important but also, generating quantities which are consistent with the observed values should be of interest.
-However, many studies still report coefficient estimates from generalized models like probit, logistic, etc.,\cite{hanmer2013behind}, which are subject of less clarity due to lack of direct link to the outcome of interest.
-
-The argument and results we present in this paper supports a greater need for a shift on focus on what and how to present predictions from generalized models.
-For example, we believe that effect-styled confidence intervals could provide more clarity concerning the uncertainty due to the predictor of interests as opposed to the conventional way of incorporating everything.
-
-From our theoretical, methodological and simulation results, researchers using these kind of models should, in the absence of theoretical justification, report predictions based on whole-sample-based approach or at least attempt to do a comparison of the two approaches before settling on the most appropriate in answering their research question.
-Moreover, we provide \proglang{R} package, \pkg{vareffects}, which implements these methods and is available on github (\href{https://github.com/mac-theobio/effects}{https://github.com/mac-theobio/effects}).
-
-\section*{Supporting information}
-
-% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
-\paragraph*{S1 Appendix.}
-\label{S1_Appendix}
-{\bf Cubic polynomial interaction simulation.} Consider an hypothetical simulation which simulates household size as a function of household wealth index and cubic function of the age of the household head, specified as follows:
-
-\begin{align}\label{sim:lm_cubic}
-\mathrm{hh~size}_i &= \beta_0 + \beta_{\mathrm{A_1}}\mathrm{Age}_i + \beta_{\mathrm{A_2}}\mathrm{Age}^2_i + \beta_{\mathrm{A_3}}\mathrm{Age}^3_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \epsilon_i \nonumber\\
-\mathrm{Age}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
-\mathrm{Wealthindex}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
-\epsilon_i &\sim \mathrm{Normal}(0, 10) \nonumber\\
-\beta_0 &= 20 \nonumber\\
-\beta_{\mathrm{A}_1} &= 0.1 \nonumber\\
-\beta_{\mathrm{A}_2} &= 0.8 \nonumber\\
-\beta_{\mathrm{A}_3} &= 0.3 \nonumber\\
-\beta_{\mathrm{W}} &= -0.5 \nonumber\\
-i &= 1,\cdots, 100
-\end{align}
-
-\paragraph*{S2 Appendix.}
-\label{S2_Appendix}
-{\bf Binary outcome simulation.} Consider a simple simulation for improved water quality in Nairobi slums, such that the status is $1$ for improved and $0$ for unimproved water quality.
-In additional to the focal predictor, age of the household head, we add wealth index.
-In particular:
-
-\begin{align}\label{sim:glm_two_pred}
-\mathrm{status}_i &\sim \mathrm{Bern}(\mathrm{P_i}) \nonumber\\
-\mathrm{logit}(\mathrm{P_i}) &= \eta_i \nonumber\\
-\mathrm{\eta}_i &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i \nonumber\\
-\mathrm{Age}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
-\mathrm{Wealthindex}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
-\beta_0 &= 5 \nonumber\\
-\beta_{\mathrm{A}} &= 0.5 \nonumber\\
-\beta_{\mathrm{W}} &= 1.5 \nonumber\\
-i &= 1,\cdots, 10000
-\end{align}
-
-\paragraph*{S3 Appendix.}
-\label{S3_Appendix}
-{\bf Mediated effect simulation.} Next, we consider a simple indirect mediation simulation such that $x$ has direct effect on $y$ which in turn has effect on $z$ but $x$ has no direct effect on $y$, i.e., $x \rightarrow y \rightarrow z$.
-In particular:
-
-\begin{align}\label{sim:simple_mediate}
-z_i &\sim \mathrm{Bern}(\mathrm{P_i}) \nonumber\\
-\mathrm{logit}(\mathrm{P_i}) &= \eta_i \nonumber\\
-\eta_i &= \beta_0 + \beta_{xz} x_i + \beta_{yz} y_i \nonumber\\
-y_i &= \rho x_i + \sqrt{1-\rho^2} y_y \nonumber\\
-x_i &\sim \mathrm{Normal(0, 1)} \nonumber\\
-y_y &\sim \mathrm{Normal(0, 1)} \nonumber\\
-\rho &= 0.8 \nonumber\\
-\beta_0 &= 5 \nonumber\\
-\beta_{xz} &= 0.2 \nonumber\\
-\beta_{yz} &= 1.5 \nonumber\\
-i &= 1,\cdots, 10000
-\end{align}
-
-\section*{Acknowledgments}
-
-This work was supported by a grant to Jonathan Dushoff from the Natural Sciences and Engineering Research Council of Canada (NSERC) Discovery.
-
-\section*{Author Contributions}
-
-\textbf{Conceptualization:} Jonathan Dushoff
-
-\noindent\textbf{Software:} Steve Cygu, Benjamin M. Bolker
-
-\noindent\textbf{Supervision:} Jonathan Dushoff, Benjamin M. Bolker
-
-\noindent\textbf{Writing – original draft:} Steve Cygu
-
-\nolinenumbers
-
-% Either type in your references using
-% \begin{thebibliography}{}
-% \bibitem{}
-% Text
-% \end{thebibliography}
-%
-% or
-%
-% Compile your BiBTeX database using our plos2015.bst
-% style file and paste the contents of your .bbl file
-% here. See http://journals.plos.org/plosone/s/latex for 
-% step-by-step instructions.
-%
-
-\bibliography{vareffects_plosone}
-
-\end{document}

### 2022-05-13 manuscript/body.tex
Refactoring manuscript.tex

--- 2022-05-05 manuscript/vareffects_plosone.tex
@@
-% Template for PLoS
-% Version 3.5 March 2018
-
-% % % % % % % % % % % % % % % % % % % % % %
-%
-% -- IMPORTANT NOTE
-%
-% This template contains comments intended 
-% to minimize problems and delays during our production 
-% process. Please follow the template instructions
-% whenever possible.
-%
-% % % % % % % % % % % % % % % % % % % % % % %
-
-% Once your paper is accepted for publication, 
-% PLEASE REMOVE ALL TRACKED CHANGES in this file 
-% and leave only the final text of your manuscript. 
-% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
-% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
-%
-%
-% There are no restrictions on package use within the LaTeX files except that 
-% no packages listed in the template may be deleted.
-%
-% Please do not include colors or graphics in the text.
-%
-% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
-%
-% % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- FIGURES AND TABLES
-%
-% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
-%
-% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
-% - Figures should be uploaded separately from your manuscript file. 
-% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
-% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
-% For figure citations, please use "Fig" instead of "Figure".
-% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
-%
-% Tables should be cell-based and may not contain:
-% - spacing/line breaks within cells to alter layout or alignment
-% - do not nest tabular environments (no tabular environments within tabular environments)
-% - no graphics or colored text (cell background color/shading OK)
-% See http://journals.plos.org/plosone/s/tables for table guidelines.
-%
-% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
-%
-% IMPORTANT
-% Below are a few tips to help format your equations and other special characters according to our specifications.
-For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
-%
-% For inline equations, please be sure to include all portions of an equation in the math environment.
-For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
-%
-% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
-%
-% Please add line breaks to long display equations when possible in order to fit size of the column. 
-%
-% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
-%
-% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
-%
-% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
-%
-% % % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Please contact latex@plos.org with any questions.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-
-\documentclass[10pt,letterpaper]{article}
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
-
-% amsmath and amssymb packages, useful for mathematical formulas and symbols
-\usepackage{amsmath,amssymb}
-
-% Use adjustwidth environment to exceed column width (see example table in text)
-\usepackage{changepage}
-
-% Use Unicode characters when possible
-\usepackage[utf8x]{inputenc}
-
-% textcomp package and marvosym package for additional characters
-\usepackage{textcomp,marvosym}
-
-% cite package, to clean up citations in the main text. Do not remove.
-\usepackage{cite}
-
-% Use nameref to cite supporting information files (see Supporting Information section for more info)
-\usepackage{nameref,hyperref}
-
-% line numbers
-\usepackage[right]{lineno}
-
-% ligatures disabled
-\usepackage{microtype}
-\DisableLigatures[f]{encoding = *, family = * }
-
-% color can be used to apply background shading to table cells only
-\usepackage[table]{xcolor}
-
-% array package and thick rules for tables
-\usepackage{array}
-
-% create "+" rule type for thick vertical lines
-\newcolumntype{+}{!{\vrule width 2pt}}
-
-% create \thickcline for thick horizontal lines of variable length
-\newlength\savedwidth
-\newcommand\thickcline[1]{%
-  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-  \cline{#1}%
-  \noalign{\vskip\arrayrulewidth}%
-  \noalign{\global\arrayrulewidth\savedwidth}%
-}
-
-% \thickhline command for thick horizontal lines that span the table
-\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-\hline
-\noalign{\global\arrayrulewidth\savedwidth}}
-
-% Remove comment for double spacing
-%\usepackage{setspace} 
-%\doublespacing
-
-% Text layout
-\raggedright
-\setlength{\parindent}{0.5cm}
-\textwidth 5.25in 
-\textheight 8.75in
-
-% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
-% Captions will be left justified
-\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-\renewcommand{\figurename}{Fig}
-
-% Use the PLoS provided BiBTeX style
-\bibliographystyle{plos2015}
-
-% Remove brackets from numbering in List of References
-\makeatletter
-\renewcommand{\@biblabel}[1]{\quad#1.}
-\makeatother
-
-% Header and Footer with logo
-\usepackage{lastpage,fancyhdr,graphicx,grffile}
-%\pagestyle{myheadings}
-\pagestyle{fancy}
-\fancyhf{}
-%\setlength{\headheight}{27.023pt}
-%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
-\rfoot{\thepage/\pageref{LastPage}}
-\renewcommand{\headrulewidth}{0pt}
-\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
-\fancyheadoffset[L]{2.25in}
-\fancyfootoffset[L]{2.25in}
-\lfoot{\today}
-
-%% Include all macros below
-
-\newcommand{\fix}{{\bf \textcolor{red}{FIXME}}}
-\newcommand{\JD}[1]{{\color{blue} \emph{#1}}}
-\newcommand{\bmb}[1]{{\color{RubineRed!70!} \emph{#1}}}
-\newcommand{\pkg}[1]{\textbf{#1}}
-
-\def\code#1{\texttt{#1}}
-\let\proglang=\textsf
-
-\newcommand{\bX}{{\mathbf X}}
-\newcommand{\bZ}{{\mathbf Z}}
-\newcommand{\bbeta}{{\boldsymbol \beta}}
-\newcommand{\boldeta}{{\boldsymbol \eta}}
-\newcommand{\boldmu}{{\boldsymbol \mu}}
-
-\newcommand{\nset}[1]{#1_{\{n\}}}
-\newcommand{\yref}{y_{\textrm{ref}}}
-\newcommand{\cdist}{{D(\nset{x}|x_f)}}
-\newcommand{\cdistprime}{{D(\nset{x}|x_{f'})}}
-\newcommand{\xfprime}{x_{f'}}
-
-\newcommand{\yE}{{\mathrm E}}
-\let\over=\overline
-
-\newcommand{\bxo}{\bX^{\star}}
-\newcommand{\bxof}{\bX_f^{\star}}
-\newcommand{\ba}{{\mathbf A}}
-\newcommand{\bao}{{\mathbf A}_0}
-\newcommand{\baf}{{\mathbf A_f}}
-\newcommand{\bafc}{{\mathbf A_{\bar{f}}}}
-
-%% END MACROS SECTION
-
-\begin{document}
-\vspace*{0.2in}
-
-% Title must be 250 characters or less.
-\begin{flushleft}
-{\Large
-\textbf\newline{Describing the curves: uncertainty estimation and bias correction for predictor effects in simple and generalized linear (mixed) models} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
-}
-\newline
-% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
-\\
-Steve Cygu\textsuperscript{1},
-Benjamin M. Bolker\textsuperscript{1,2},
-Jonathan Dushoff\textsuperscript{1,2}
-\\
-\bigskip
-\textbf{1} School of Computational Science and Engineering, McMaster University, Hamilton, Ontario, Canada
-\\
-\textbf{2} Department of Biology, McMaster University, Hamilton, Ontario, Canada
-\\
-\bigskip
-
-% Use the asterisk to denote corresponding authorship and provide email address in note below.
-* cygu@aims.ac.za
-
-\end{flushleft}
-% Please keep the abstract below 300 words
-\section*{Abstract}
-
-In many applications which use generalized linear (mixed) models, outcome predictions or predicted probabilities are often of interest.
-For models that involve complex multiplicative interactions, additional non-focal predictors or nonlinear link functions, the estimated coefficient are not readily interpretable.
-A general way to summarize these kind of models is through predictions, which are not only dependent on choice of representative values of focal predictor but also sensitive to which values of non-focal predictors are chosen.
-The most common approach is generating predictions at a ``reference point'', usually the mean, of non-focal predictor, i.e., mean-based (centered) which could be considered as the effect of an ``average case'' in the population.
-In the presence of sources of bias such as additional non-focal predictors, nonlinear link functions, random effect terms, etc., mean-based approach generate predictor effects that are biased and may not be consistent with the observed quantities.
-An alternative is the whole-sample-based approach which estimates the average effect in the population.
-Moreover, anchored effects provide an alternative and a more clear way to describe uncertainty associated with the focal predictor of interest.
-In addition to theoretical and methodical comparison, using simulation, we illustrate the two approaches and show that they can produce substantially different results and that whole-sample-based approach can not only produce estimates consistent with the observed values, but also appropriate for bias correction.
-We also present an alternative way, anchored confidence intervals, to describe uncertainties associated with these predictions.
-
-\section*{Definitions}
-
-\begin{itemize}
-\item \textbf{Input variables}: Scientific variables underlying an inference or exploration. The focal predictor we use for effects or predictions is an input variable.
-\item \textbf{Model variables}: Variables that represent columns in the model matrix.
-Each input variable will correspond to one or more model variables.
-In particular, variables with more than two categories, or variables modeled with a spline or polynomial response, will correspond to more than one model variables.
-\item\textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more model variables).
-The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable.
-\item \textbf{Reference point}: Value or values chosen for \emph{non-focal predictors}, when estimating effects.
-Typically the center point, but can instead be a population of quantiles or observations.
-\item \textbf{Anchor}: The value chosen for the \emph{focal predictor} when estimating effect confidence intervals (anchor choice does not affect the estimate).
-Typically chosen as the center point of the model variables corresponding to the focal input variable.
-\end{itemize}
-
-% Please keep the Author Summary between 150 and 200 words
-% Use first person. PLOS ONE authors please skip this step. 
-% Author Summary not valid for PLOS ONE submissions.   
-\section*{Author summary}
-
-\fix
-
 \linenumbers
@@
 \nolinenumbers
-
-% Either type in your references using
-% \begin{thebibliography}{}
-% \bibitem{}
-% Text
-% \end{thebibliography}
-%
-% or
-%
-% Compile your BiBTeX database using our plos2015.bst
-% style file and paste the contents of your .bbl file
-% here. See http://journals.plos.org/plosone/s/latex for 
-% step-by-step instructions.
-%
-
-\bibliography{vareffects_plosone}
-
-\end{document}

### 2022-05-13 manuscript/body.tex
Autocommit

@@
 \centering
-\includegraphics[width=0.8\textwidth]{pred_cubic_plots-figure4.pdf}
+\includegraphics[width=0.8\textwidth]{cubic_predictors_preds.ggp.pdf}
 \caption{{\bf Prediction-styled and effect-styled plots for cubic polynomial and simple focal predictors.} A: The focal predictor is a cubic polynomial.
@@
 \centering
-\includegraphics[width=0.8\textwidth]{pred_bin_plots-figure5.pdf}
+\includegraphics[width=0.8\textwidth]{glm_two_predictor_preds.ggp.pdf}
 \caption{{\bf Mean-based and whole-sample-based predictions.} The two approaches produce relatively different estimates in each of the predictors but the difference is more pronounced when the effect of the particular focal predictor has no strong effect -- \emph{on the left}.
@@
 \centering
-\includegraphics[width=0.8\textwidth]{pred_mediated_plots-figure6.pdf}
+\includegraphics[width=0.8\textwidth]{mediate_preds.ggp.pdf}
 \caption{{\bf Mean-based and whole-sample-based predictions for mediated effects.} These figures compare the mean-based and whole-sample-based predictions for the mediated (multivariate) and non-mediated (univariate) models.

### 2022-05-14 manuscript/body.tex
Autocommit

@@
 This flexibility comes at a cost, for example, complex multivariate models may risk misinterpretation, and miscalculation of quantities of interest.
-Also, coefficient estimates of models involving nonlinear link functions or interactions lose their direct interpretation \cite{leeper2017interpreting}, meaning that interpretation of derived quantities from these estimates requires some understanding of the specified model.
-An alternative is to explore the \emph{effects} of predictors on the predictions or probabilities of the outcome at various level of the predictor of interest -- \emph{prediction and effects} \cite{fox2009effect, leeper2017package, lenth2018package}.
+Also, coefficient estimates of models involving nonlinear link functions or interactions lose their direct interpretation \citep{leeper2017interpreting}, meaning that interpretation of derived quantities from these estimates requires some understanding of the specified model.
+An alternative is to explore the \emph{effects} of predictors on the predictions or probabilities of the outcome at various level of the predictor of interest -- \emph{prediction and effects} \citep{fox2009effect, leeper2017package, lenth2018package}.
 For example, as a way to plan, public health officials may want to know the effects of household income, wealth index, etc, on the predicted probability of having improved water services among slum dwellers.
@@
 
-The most common way of dealing with the first challenge is taking unique levels of the focal (predictor of interest) predictor if discrete or taking appropriately sized quantiles (or bins) if continuous, and then calculating the predictions while holding non-focal (other predictors other than the focal predictor) predictors at their reference point (e.g., means) \cite{hanmer2013behind}.
-This generates -- \emph{predictions}, \emph{predictor effects}\cite{fox2009effect}, \emph{marginal predictions} \cite{leeper2017package} or \emph{estimated marginal means} \cite{lenth2018package}.
+The most common way of dealing with the first challenge is taking unique levels of the focal (predictor of interest) predictor if discrete or taking appropriately sized quantiles (or bins) if continuous, and then calculating the predictions while holding non-focal (other predictors other than the focal predictor) predictors at their reference point (e.g., means) \citep{hanmer2013behind}.
+This generates -- \emph{predictions}, \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}.
 In this article, we refer to this quantity as predictions since it should, for example, tell us what we would expect the response to be at a particular value or level of the predictor, for an ``average case''.
@@
 In particular, bias in expected mean prediction induced by nonlinear transformation of the response variable can lead inaccurate predictions.
-An alternative to mean-based reference point, i.e., averaging non-focal predictors, is the \emph{whole-sample-based} approach, discussed in detail later, which involves computing the prediction over the population of non-focal predictors and then averaging across the values of the focal predictor \cite{hanmer2013behind}.
+An alternative to mean-based reference point, i.e., averaging non-focal predictors, is the \emph{whole-sample-based} approach, discussed in detail later, which involves computing the prediction over the population of non-focal predictors and then averaging across the values of the focal predictor \citep{hanmer2013behind}.
 
@@
 Let $\mathbf{X^*}$ be the model matrix, inheriting most of its key properties, for example transformations on predictors and interactions from the model matrix, $\mathbf{X}$.
-Then the quantity $\hat{\boldeta}^\star = \bX^\star\hat{\bbeta}$ is the prediction with respect to the focal predictor in question \cite{fox2009effect}.
+Then the quantity $\hat{\boldeta}^\star = \bX^\star\hat{\bbeta}$ is the prediction with respect to the focal predictor in question \citep{fox2009effect}.
 
@@
 In principle, every value of focal predictor has a different CI.
-The conventional way to compute variances for predictions is $\sigma^2 = \textrm{Diag}(\bX^\star \boldsymbol{\Sigma} \bX^{\star\top})$ \cite{lenth2018package, fox2009effect}, so that the confidence intervals are $\eta \pm q\sigma$, where $q$ is an appropriate quantile of Normal or t distribution.
+The conventional way to compute variances for predictions is $\sigma^2 = \textrm{Diag}(\bX^\star \boldsymbol{\Sigma} \bX^{\star\top})$ \citep{lenth2018package, fox2009effect}, so that the confidence intervals are $\eta \pm q\sigma$, where $q$ is an appropriate quantile of Normal or t distribution.
 This generates predictions-styled CIs and incorporates all the uncertainties -- including the uncertainty due to the intercept and non-focal predictors.
@@
 In such cases, bias correction is needed when back-transforming the predictions to the original scales.
-Most common approach for bias-adjustment is second-order Taylor approximation \cite{lenth2018package, duursma2003bias}.
+Most common approach for bias-adjustment is second-order Taylor approximation \citep{lenth2018package, duursma2003bias}.
 Another potential source of bias comes from additional non-focal predictors, especially in models with nonlinear link functions.
@@
 
-The most precise (although not necessarily accurate!) way to predict is to condition on values of the focal predictor and make predictions for all observations (members of the population) \cite{hanmer2013behind}.
+The most precise (although not necessarily accurate!) way to predict is to condition on values of the focal predictor and make predictions for all observations (members of the population) \citep{hanmer2013behind}.
 A key point is that the nonlinear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population to a one-dimensional distribution of $\eta^\star(x_f, \nset{x})$, which is a function of focal predictor and the observed values of non-focal predictors, as opposed to the definition in Equation~\ref{eq:eta_mean}.
@@
 If $g()$ is strictly concave or strictly convex, one can use Jensen's inequality to determine which side of Equation~\ref{eq:compare_anchored_pop_based} is smaller or larger.
-However, for non-fully convex or concave link function, one can use Taylor series expansion to approximate the range over which the two are smaller or smaller than the other \cite{hanmer2013behind}.
+However, for non-fully convex or concave link function, one can use Taylor series expansion to approximate the range over which the two are smaller or smaller than the other \citep{hanmer2013behind}.
 
@@
 For example, a dummy variable for christian household heads may be set to $0.2$, translating to prediction for an household head who is $20\%$ christian.
-This problem may extend to models with other complex interaction terms \cite{hanmer2013behind}.
+This problem may extend to models with other complex interaction terms \citep{hanmer2013behind}.
 
@@
 Generating outcome predictions or predicted probabilities from simple and generalized linear (mixed) models is not only important but also, generating quantities which are consistent with the observed values should be of interest.
-However, many studies still report coefficient estimates from generalized models like probit, logistic, etc.,\cite{hanmer2013behind}, which are subject of less clarity due to lack of direct link to the outcome of interest.
+However, many studies still report coefficient estimates from generalized models like probit, logistic, etc., \citep{hanmer2013behind}, which are subject of less clarity due to lack of direct link to the outcome of interest.
 

### 2022-05-25 manuscript/abstract.tex
Autocommit
Working on BMB's feedback

@@
 
-In many applications which use generalized linear (mixed) models, outcome predictions or predicted probabilities are often of interest.
-For models that involve complex multiplicative interactions, additional non-focal predictors or nonlinear link functions, the estimated coefficient are not readily interpretable.
-A general way to summarize these kind of models is through predictions, which are not only dependent on choice of representative values of focal predictor but also sensitive to which values of non-focal predictors are chosen.
-The most common approach is generating predictions at a ``reference point'', usually the mean, of non-focal predictor, i.e., mean-based (centered) which could be considered as the effect of an ``average case'' in the population.
-In the presence of sources of bias such as additional non-focal predictors, nonlinear link functions, random effect terms, etc., mean-based approach generate predictor effects that are biased and may not be consistent with the observed quantities.
+In applications that use generalized linear (mixed) models, outcome predictions are often of interest.
+For models that involve complex multiplicative interactions, additional non-focal predictors or nonlinear link functions, the estimated coefficients are not readily interpretable.
+A general way to summarize these kind of models is through predictions, that depend not only on choice of representative values of focal predictor but are also sensitive to which values of non-focal predictors are chosen.
+The most common approach is generating predictions at a reference point, usually the mean, of non-focal predictor, i.e., mean-based (centered) which could be considered as the effect of an average case in the population.
+In the presence of additional non-focal predictors, nonlinear link functions, random effect terms, etc., mean-based approach generate predictions that are biased and may not be consistent with the observed quantities.
 An alternative is the whole-sample-based approach which estimates the average effect in the population.
@@
 \section*{Author summary}
+
+\fix

### 2022-05-25 manuscript/body.tex
Autocommit
Working on BMB's feedback

@@
+%% Main tex
+
 \linenumbers
@@
 
-In many applications which use simple or generalized linear (mixed) models, outcome predictions or probabilities, are often of interest.
+In applications that use regression models, outcome predictions are often of interest.
 Predictions provide a great way to summarize what the regression model is telling us and very useful for interpreting and visualizing model estimates.
-For example, in logistic regression models, the coefficient estimates are usually not easy to interpret and less informative.
-Since logistic models are nonlinear (due to the nonlinear link function), in multivariate models with interactions, the magnitude of the effect of change in the outcome depends on the values of the predictor of interest and other predictors.
+For example, in logistic regression models, the coefficient estimates, when reported as log odds, are not easy to interpret and less informative.
+Since logistic models are nonlinear (due to the nonlinear link function), in multivariate models with\bmb{?only}\BC{clarify with BMB} interactions, the magnitude of the effect of change in the outcome depends on the values of the predictor of interest and other predictors.
 Hence, the conclusions one can make about the estimated effects greatly depends on how well we choose the values of the other predictors, i.e., \emph{non-focal} predictors.
 
-Both simple and generalized linear (mixed) models (GL(M)Ms) can examine very complex relationships, including nonlinear relationships between response and predictors, interactions between predictors (and via splines for example), and nonlinear transformations via link functions due to their flexibility.
-This flexibility comes at a cost, for example, complex multivariate models may risk misinterpretation, and miscalculation of quantities of interest.
-Also, coefficient estimates of models involving nonlinear link functions or interactions lose their direct interpretation \citep{leeper2017interpreting}, meaning that interpretation of derived quantities from these estimates requires some understanding of the specified model.
-An alternative is to explore the \emph{effects} of predictors on the predictions or probabilities of the outcome at various level of the predictor of interest -- \emph{prediction and effects} \citep{fox2009effect, leeper2017package, lenth2018package}.
-For example, as a way to plan, public health officials may want to know the effects of household income, wealth index, etc, on the predicted probability of having improved water services among slum dwellers.
+Ordinary least squares (OLS) regression provides an easy way to express multivariate relationships as an additive linear combination of many predictors with the outcome.
+The coefficient estimates from OLS regression models are generally easy to interpret as the expected change in the outcome with a unit change in the continuous predictor or as the expected outcome difference between the levels of the categorical predictor with the baseline category.
+However, with additional complexities such interaction between predictors, the resulting coefficients risk misinterpretation or interpretation complexity \citep{brambor_understanding_2006, berry_improving_2012}.
 
-When visually presented, prediction and effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind of transformations on the predictors whose estimates are usually, but not always, a subject to less clarity of interpretation.
-Further, generating predictor and effect plots together with the associated confidence intervals for regression models has a number of challenges.
+Simple OLS regression models are flexible and can easily be generalized, as generalized linear (mixed) models (GL(M)Ms), to examine more complex relationships, including \bmb{nonlinear relationships -- need to clarify}\BC{clarity with BMB} between response and predictors e.g., log transformed response or predictors, interactions between predictors (and via splines for example), and nonlinear transformations via link functions \bmb{due to their flexibility}\BC{clarify why BMB circled it}.
+This flexibility comes with additional complications, for example, complex multivariate models may risk misinterpretation, and miscalculation of quantities of interest \citep{berry_improving_2012, leeper2017interpreting}.
+Further, coefficient estimates of GLM models involving nonlinear link functions lack direct interpretation\citep{leeper2017interpreting}, meaning that interpretation of derived quantities from these estimates requires some understanding of the specified model.
+An alternative is to explore the \emph{effects} of predictors on the predictions of the outcome at various levels of the predictor of interest -- \emph{predictions and effects} \citep{fox2009effect, leeper2017package, lenth2018package}.
+For example, as a way to plan, public health officials may want to know the effects of household income, wealthindex, etc, on the predicted probability that slum dwellers will have improved water services.
+
+When visually presented, prediction and effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind of transformations on the predictors whose estimates are usually, but not always, harder to interpret.
+Generating predictions and effects plots together with the associated confidence intervals for regression models has a number of challenges.
 In particular:
@@
 
-The most common way of dealing with the first challenge is taking unique levels of the focal (predictor of interest) predictor if discrete or taking appropriately sized quantiles (or bins) if continuous, and then calculating the predictions while holding non-focal (other predictors other than the focal predictor) predictors at their reference point (e.g., means) \citep{hanmer2013behind}.
-This generates -- \emph{predictions}, \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}.
-In this article, we refer to this quantity as predictions since it should, for example, tell us what we would expect the response to be at a particular value or level of the predictor, for an ``average case''.
-More specifically, we compute the expected outcome by meaningfully holding non-focal predictors constant (or averaged in some meaningful way) while varying the focal predictor, with the goal that the outcome expected prediction represents how the model responds to the changes in the focal predictor.
-On the other hand, effects differs with predictions in the sense that the former is about how we describe the uncertainty around the latter and depends on the anchor.
+The most common way of choosing the representative values is taking unique levels of the focal predictor if discrete or taking appropriately sized quantiles if continuous, and then calculating the predictions while holding non-focal (other predictors other than the focal predictor) predictors at their reference point (e.g., means) \citep{hanmer2013behind}.
+This procedure generates -- \emph{predictions}, \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}.
+In this article, we refer to these quantities as predictions since they should, for example, tell us what we would expect the response to be at a particular value or level of the predictor, for an ``average case''.
+\bmb{More specifically, we compute the expected outcome by holding non-focal predictors constant (or averaged in some meaningful way) while varying the focal predictor, with the goal that the predictions represents how the model responds to the changes in the focal predictor.
+On the other hand, effects differ with predictions in the sense that the former is about how we describe the uncertainty around the latter and depends on the anchor.
+-- see next/unclear}\BC{clarify with BMB}
 
-The commonly used \proglang{R} software packages (\pkg{emmeans} and \pkg{effects}) for generating predictions, by default, use the average of non-focal predictors as the reference point.
-However, there are a number of choices one can make when considering this approach -- for example, in the presence of interaction, averaging the interactions (averaging product of interacting model variables) versus product of the averages of the interacting model variables (default for \pkg{emmeans} and \pkg{effects}).
-We claim that neither of these two approaches is the most appropriate but we show that averaging the interaction closely matches the observed values.
+The most \bmb{commonly used \proglang{R} software packages (\pkg{emmeans} and \pkg{effects})}\BC{not sure the underline} for generating predictions, by default, use the average of non-focal predictors as the reference point.
+However, there are a number of choices one can make when considering this approach -- for example, in the presence of interaction, averaging the interactions (averaging the product of interacting model variables) versus taking the product of the averages of the interacting model variables (the default for \pkg{emmeans} and \pkg{effects}).
+For simple models (with no interactions), the two approaches give the same predictions.
+However, for complex models, we claim that neither of these two approaches is the most appropriate but we show that averaging the interaction closely matches the observed values.
 To illustrate this, consider models~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher} below, with $x_1$ as the focal predictor:
-
+\bmb{This whole example .... example of nonlinear averaging}\BC{Yes and handling interaction terms?}
 %
@@
 
-We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$ and $\beta_{23} = 5$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.
+We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$ $\beta_{23} = 5$, $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and $\sigma^2 = 1$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.
 
 \begin{figure}[!h]
-\centering
-\includegraphics[width=0.9\textwidth]{justify_preds.inter.pdf}
+\begin{center}
+\includegraphics{justify_preds.inter.pdf}
+\end{center}
 \caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} predictions for $x_1$ on $y$ for models with and without interactions.}
@@
 The grey points are the binned simulated $y$.
-The trend lines represents the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other predictors at their average for the three mthods and the ``true'' predictions based on the simulation parameters.
-A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the simulated in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$.
-Similarly, the predictions based on the three approaches closely matches the ``true'' predictions.
-B: Even with the simple interaction between non-focal predictors, we start seeing deviation in both predicted mean, $\bar{\hat{y}}$, and overall predictions from the simulated, $\bar{y}$, and ``true'' predictions, respectively, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not, the proposed \pkg{varpred}.}
+The trend lines represent the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other predictors at their average for the three methods and the true predictions based on the simulation parameters.
+A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the simulated value in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$.
+Similarly, the predictions based on the three approaches closely match the true predictions.
+B: Even with the simple interaction between non-focal predictors, we start seeing deviation in both predicted mean, $\bar{\hat{y}}$, and overall predictions from the simulated, $\bar{y}$, and true predictions, respectively; in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not the proposed \pkg{varpred}.}
 \label{fig:justify_plots}
@@
 
-Sometimes, one may be interested in the uncertainties associated with the focal predictor of interest only, excluding other uncertainties due to other non-focal predictors -- effects.
+One may be interested in the uncertainties associated with the focal predictor of interest only, excluding other uncertainties due to other non-focal predictors -- effects.
 However, currently, the two packages do not provide straightforward way do achieve this.
@@
 \begin{figure}[!h]
-\centering
-\includegraphics[width=0.9\textwidth]{justify_preds.isolate.pdf}
+\begin{center}
+\includegraphics{justify_preds.isolate.pdf}
+\end{center}
 \caption{{\bf A comparison of conventional $95\%$ confidence bands and effects.} The description of horizontal, vertical and trend lines remain the same as above.
@@
 In other words, with \pkg{varpred}, we are able to generate effects indicating zero uncertainty at the value of the focal predictor we are more certain about, i.e., mean of the model variable corresponding to the focal predictor (center point or mean-based).
-For simple models, the point where the confidence bands crosses is the model center and it corresponds to anchor we choose, in this case, the mean of the focal predictor.
+For simple models, the point where the confidence bands crosses is the model center; it corresponds to anchor we choose, in this case, the mean of the focal predictor.
 
-When dealing with nonlinear link functions or models with multiple non-focal predictors, the correct predictions, for example, are even much harder to estimate.
+When dealing with nonlinear link functions or models with multiple non-focal predictors, the correct predictions are much harder to estimate.
 One approach is to make predictions on the transformed scale (linear predictor scale), and then back-transform to the original scale.
@@
 \begin{figure}[!h]
-\centering
-\includegraphics[width=0.7\textwidth]{justify_anchors.combined.pdf}
+\begin{center}
+\includegraphics{justify_anchors.combined.pdf}
+\end{center}
 \caption{{\bf A comparison of min-based and center- (mean-) based anchors.} The center-based effects are based on the mean of the focal predictor while the min-anchored effects are based on the minimum of the focal predictor.
@@
 
-We would like to summarize by talking about predictions vs. effects. Predictions are the standard curves; they tell us what we expect to see if we know the value of the focal predictor.
-
+We would like to summarize by talking about predictions vs.
+effects.
+Predictions are the standard curves; they tell us what we expect to see if we know the value of the focal predictor.
 The lines inside (effects) are a little different: they are a good way to visualize the effect of the focal predictor.
-Unlike predictions, where the uncertainty captures the uncertainty in the whole model, the effects plot focuses on uncertainty in the effect of the focal predictor only -- thus, for example, if an effect is barely significant (P exactly 0.05), one end of the CI will be exactly a horizontal line for the effects plot, but not for the prediction plot.
-Also, the uncertainty curves for the effect, but not for the prediction, depend on the ``anchor''.
+Unlike predictions, where the uncertainty captures the uncertainty in the whole model, the effects plot focuses on uncertainty in the effect of the focal predictor only and depend on the anchor.
 
@@
 We refer to this as centered or mean-based approach.
-The most convenient way to achieve this is by constructing $\bX^\star$ by averaging the columns of non-focal predictors in model matrix $\bX$, and together with appropriately chosen values of focal predictor.
+The most convenient way to achieve this is by constructing centered model matrix, $\bX^\star$, by averaging the columns of non-focal predictors in model matrix $\bX$, and together with appropriately chosen values of focal predictor.
 
-Consider a simple linear model with linear predictor $\eta = \bX\bbeta$ and let $g(\boldmu) = \boldeta$ be an identity link function (in the case of simple linear model), where $\boldmu$ is the expected value of response variable $y$.
-Let $\hat{\bbeta}$ be the estimate of $\bbeta$, together with the estimated covariance matrix $\boldsymbol{\Sigma} = V(\hat{\bbeta})$ of $\hat{\bbeta}$.
-Let $\mathbf{X^*}$ be the model matrix, inheriting most of its key properties, for example transformations on predictors and interactions from the model matrix, $\mathbf{X}$.
-Then the quantity $\hat{\boldeta}^\star = \bX^\star\hat{\bbeta}$ is the prediction with respect to the focal predictor in question \citep{fox2009effect}.
+Consider a simple linear model with linear predictor $\eta = \bX\bbeta$ and let $g(\boldmu) = \boldeta$ be an identity link function (in the case of simple linear model), where $\boldmuh$ is the expected value of response variable $\hat{y}$.
+Let $\bbetah$ be the estimate of $\bbeta$, together with the estimated covariance matrix $\Sigma = V(\bbetah)$ of $\bbetah$.
+Then the quantity $\hat{\boldeta}^\star = \bX^\star\bbetah$ is the prediction with respect to the focal predictor in question \citep{fox2009effect}.
 
-An alternative formulation of $\hat{\boldeta}^\star$ involves expressing the linear predictor as the sum of the focal and non-focal predictors' linear predictor. In particular,
+An alternative formulation of $\boldetah^\star$ involves expressing the linear predictor as the sum of the focal and non-focal predictors' linear predictors. In particular,
 
 \begin{align}\label{eq:eta_mean}
-\eta^\star(x_f, \nset{{\bar{x}^\star}}) &= \beta_f x_f + \sum \nset{\beta} \nset{{\bar{x}^\star}} \\
+\eta^\star(x_f, \nset{{\bar{x}^\star}}) &= \hat{\beta}_f x_f + \sum \nset{\hat{\beta}} \nset{{\bar{x}^\star}} \\
 \hat{y}_f  &= g^{-1} \left(\eta^\star(x_f, \nset{{\bar{x}^\star}})\right)
 \end{align}
-where $\nset{{\bar{x}^\star}}$ are the appropriately averaged entries of non-focal predictors and $x_f$ is a vector of values of the focal predictors for a particular observation.
+where $\nset{{\bar{x}^\star}}$ and $x_f$ are columns of $\bX^\star$ corresponding to the non-focal predictors and the focal predictor, respectively.
 
@@
 
-Higher order terms such as interactions, splines, polynomials, etc., can be within the focal predictor, within (between) non-focal predictor(s), between the focal and non-focal predictor(s).
-To distinguish the three, suppose the model which describes the hypothetical simulated household size simulation based on a number of socio-demographic factors is
+Higher order terms such as interactions, splines, polynomials, etc., can be within the focal predictor, within (between) non-focal predictor(s) or between the focal and non-focal predictor(s).
+To distinguish the three, suppose the model which describes the hypothetical simulated household size simulation based on a number of socio-demographic factors such as age and wealthindex is
 
@@
 
-In the first case, with $Age$ as the focal predictor, the interaction is within its polynomial terms (and 3 model variables associated with it).
+In the first case, with \code{Age} as the focal predictor, with cubic polynomial interaction terms (and 3 model variables associated with it).
 In this case, each of the polynomial terms (model variables) of the focal predictor are evaluated independently across the chosen levels of the focal predictor, $Age_i$.
 Specifically, the higher order terms are treated as additional columns of the model matrix evaluated with the same values of the focal predictor, while non-focal predictors are fixed at their reference point as discussed in the previous section.
-In this case, the predictions associated with $Age$ on the linear predictor scale becomes
+In this case, the predictions associated with \code{Age} on the linear predictor scale become
 
@@
 
-In the second case, with $Wealthindex$ as the focal predictor, the higher order interactions in non-focal predictor, $Age$, are treated simply as additional columns in the variable space (of model matrix) and appropriate choice of reference point applies just like in the models without interactions.
-For instance, in mean-based approach, we simply average all non-focal interaction terms.
+In the second case, with $Wealthindex$ as the focal predictor, the higher order terms in non-focal predictor, \code{Age}, are treated simply as additional columns in the variable space (model variables) and an appropriate choice of reference point applies just like in the models without higher order interactions.
+For instance, in mean-based approach, we average all non-focal model variables.
 Thus
@@
 \begin{align*}
-\eta^\star(\mathrm{Wealthindex}_i, \nset{{\{\over{\mathrm{Age}}, \over{\mathrm{Age}^2}, \over{\mathrm{Age}^2}\}}}) &= \beta_0 + \beta_{\mathrm{A_1}}\over{\mathrm{Age}} + \beta_{\mathrm{A_2}}\over{\mathrm{Age}^2} + \beta_{\mathrm{A_3}}\over{\mathrm{Age}^3}\\
+\eta^\star(\mathrm{Wealthindex}_i, \nset{{\{\over{\mathrm{Age\mathop{\vphantom{^2}}}}, \over{\mathrm{Age}^2}, \over{\mathrm{Age}^2}\}}}) &= \beta_0 + \beta_{\mathrm{A_1}}\over{\mathrm{Age\mathop{\vphantom{^2}}}} + \beta_{\mathrm{A_2}}\over{\mathrm{Age}^2} + \beta_{\mathrm{A_3}}\over{\mathrm{Age}^3}\\
 	& + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i
@@
 
-Lastly, consider model~\ref{eq:simple_inter_higher}, with the interactions between the focal predictor, $x_2$, and non-focal predictor, $x_3$.
+Lastly, consider model shown in Equation~\ref{eq:simple_inter_higher}, previously used to demonstrate how to handle interactions in non-focal predictors.
+Here, we use the same model to demonstrate how to handle interactions between the focal predictor, $x_2$, and non-focal predictor, $x_3$.
 In this case, the non-interacting predictors are treated as non-focal predictors, with appropriate reference points.
 However, for the interacting predictors, we first choose representative values for the focal predictor and some particular values of the interacting predictor similar to the way we choose values for the focal predictor or predetermined.
-In our example, suppose we pick $i$ and $j$ unique values of the focal predictor, $x_2$ and interacting predictor $x_3$, respectively, then the prediction on linear predictor scale is given by
+In our example, suppose we pick $i$ and $j$ unique values of the focal predictor, $x_2$ and interacting predictor $x_3$, respectively.
+The prediction on linear predictor scale is given by
 
@@
 
-What about the confidence intervals (CI)?
-The limits of the confidence intervals are points, not mean values.
+We describe the uncertainty around the estimates using confidence intervals (CI).
 In principle, every value of focal predictor has a different CI.
-The conventional way to compute variances for predictions is $\sigma^2 = \textrm{Diag}(\bX^\star \boldsymbol{\Sigma} \bX^{\star\top})$ \citep{lenth2018package, fox2009effect}, so that the confidence intervals are $\eta \pm q\sigma$, where $q$ is an appropriate quantile of Normal or t distribution.
-This generates predictions-styled CIs and incorporates all the uncertainties -- including the uncertainty due to the intercept and non-focal predictors.
-But what if we are only interested in the uncertainty as a result of the focal predictor, so that the confidence intervals are $\eta \pm q \sigma_f$?
-We call this anchored CIs and generates effects-styled CIs.
+The conventional way to compute variances for predictions is $\boldsymbol\sigma^2 = \textrm{Diag}(\bX^\star \boldsymbol{\Sigma} \bX^{\star\top})$ \citep{lenth2018package, fox2009effect}, so that the confidence intervals are $\boldeta \pm q\boldsymbol\sigma$, where $q$ is an appropriate quantile of Normal or t distribution.
+This generates prediction CIs and incorporates all the uncertainties -- including the uncertainty due to the intercept and non-focal predictors.
+But what if we are only interested in the uncertainty as a result of the focal predictor, so that the confidence intervals are $\boldeta \pm q \boldsymbol \sigma_f$, i.e., effects?
 
-Currently, commonly used \proglang{R} packages for constructing predictions, by default, do not exclude the uncertainties resulting from non-focal predictors when computing the CIs.
-A non-trivial way to exclude uncertainties associated with non-focal predictors in some of these packages is to provide a user defined variance-covariance matrix with the covariances of non-focal terms set to $0$ -- \emph{zeroing-out} variance-covariance matrix.
-This only works when the input predictors are \emph{centered} prior to model fitting, in case of numerical predictors, and even much complicated when the predictors are categorical.
-We first describe the variance-covariance based approach and then discuss our proposed method which is based on the model center and does not require input predictors to be scaled prior to model fitting.
-
-\subsection*{Variance-covariance}
-
-The computation of $\hat{\eta}^\star$ remains the same as described above.
-However, to compute $\sigma$, $\boldsymbol{\Sigma}$ is modified by \emph{zeroing-out} (the variance-covariance of all non-focal predictors are set to zero) variances of non-focal terms.
-Although this is the simplest approach, it requires centering of continuous, i.e., $x_c = x - \bar{x}$, predictors prior to model fitting and proper way to average categorical predictors.
+Commonly used \proglang{R} packages for constructing predictions do not exclude the uncertainties resulting from non focal predictors when computing the CIs.
+Currently a non-trivial way to exclude uncertainties associated with non-focal predictors in some of these packages is to provide a user defined variance-covariance matrix with the covariances of non-focal terms set to $0$.
+We refer to this procedure as \emph{zeroing-out} variance-covariance matrix; and only works when the input predictors are \emph{centered} prior to model fitting, in case of numerical predictors, and more complicated when the predictors are categorical.
+We describe our proposed method which is based on the model center and does not require input predictors to be centered prior to model fitting.
 
@@
 
-Let $\bxo$ be a centered model matrix containing appropriately chosen values of focal predictors and centered non-focal predictors, and let $\ba$ be the anchor matrix, with same dimensions as $\bxo$ and values in each columns repeated $n$ times.
-Also, let $\baf$ be the column of $\ba$ corresponding to the focal predictor, the rest of the columns corresponds non-focal predictors with same values as in $\bxo$.
-Any appropriate values (within the range of focal values) can be chosen for $\baf$ but for model center (center-anchored), we have $\bafc$ .
-Thus $\sigma_f^2 = \textrm{Diag}((\bxo - \bafc) \boldsymbol{\Sigma} (\bxo - \bafc)^\top)$.
-We can see that $\forall ~\bxof=\baf$, $\sigma_f^2 = \boldsymbol{0}$.
-This is true for the center-anchored provided $\bxof=\bafc$.
-In other words non-focal terms (model variables) in $\bX^{\star}_{c} = (\bxo - \bafc)$ are all zero in simple models without interactions but are isolated (narrower) around the model center in models involving complex interactions.
-In addition, the computation of $\bX^{\star}_c$ impacts only on the intercepts and non-focal model variables, i.e., the slopes and variance of the focal predictors are not affected.
+Let $\bxo$ be a model matrix containing appropriately chosen values of focal predictor and centered non-focal model variables, and let $\ba$ be the anchor matrix, with same dimension and entries in all non-focal model variables as $\bxo$.
+Let $\baf$ be the column of $\ba$ corresponding to focal predictor model variable(s) defined in $\bxo$.
+Any appropriate values can be chosen for $\baf$ but for model center (center-anchored), we use $\bafc$ which is the mean of the focal predictor model variable(s).
+Thus the variances for the predictions we defined in the previous section is
+
+%
+\begin{align}\label{eq:centered_variance}
+\boldsymbol\sigma_f^2 = \textrm{Diag}((\bxo - \bafc) \boldsymbol{\Sigma} (\bxo - \bafc)^\top)
+\end{align}
+%
+We can see that $\forall ~\bxof=\baf$, $\bxo - \baf = \boldsymbol 0$, hence $\boldsymbol\sigma_f^2 = \boldsymbol{0}$.
+Similarly, for all values of $\bxof$ close to anchor point $\bafc$ the term $(\bxo - \bafc)$ and $\boldsymbol\sigma_f^2$ goes to $\boldsymbol 0$ and $\boldsymbol\sigma_f^2 = \boldsymbol 0$ if $\bxof=\bafc$.
+This means that $\boldsymbol\sigma_f^2$ close to the anchor point are smaller than those away from the anchor point; and results to confidence intervals which are narrower (isolated) around the anchor or crosses at the anchor point for simple models.
+
+The computation of $\bxo - \bafc$ impacts only on the intercepts and non-focal model variables, i.e., the slopes and variance corresponding to the focal predictors are not affected.
 This means that we can still generate effects without necessarily centering the predictors prior to model fitting.
@@
 
-In many applications, it usually important to report the estimates that reflect the expected values of the untransformed response.
+It is usually important to report the estimates that reflect the expected values of the untransformed response.
 This is not a major issue in simple linear models.
-However, when dealing with nonlinear link functions, it is even harder to generate correct predictions that reflect this due to the bias in the expected mean induced by the nonlinear transformation of the response variable.
+However, when dealing with nonlinear link functions,  generated predictions may not reflect the observed response due to the bias in the expected mean induced by the nonlinear transformation of the response variable.
 In such cases, bias correction is needed when back-transforming the predictions to the original scales.
-Most common approach for bias-adjustment is second-order Taylor approximation \citep{lenth2018package, duursma2003bias}.
+The common approach for bias-adjustment is second-order Taylor approximation \citep{lenth2018package, duursma2003bias}.
 Another potential source of bias comes from additional non-focal predictors, especially in models with nonlinear link functions.
-Here, we describe and implement a different approach, whole-sample-based approach for bias correction.
+Here, we describe and implement a different approach; whole-sample-based approach for bias correction.
 
@@
 
-The most precise (although not necessarily accurate!) way to predict is to condition on values of the focal predictor and make predictions for all observations (members of the population) \citep{hanmer2013behind}.
-A key point is that the nonlinear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population to a one-dimensional distribution of $\eta^\star(x_f, \nset{x})$, which is a function of focal predictor and the observed values of non-focal predictors, as opposed to the definition in Equation~\ref{eq:eta_mean}.
+One way to predict is to condition on values of the focal predictor and make predictions for all observations (members of the population) \citep{hanmer2013behind}.
+The nonlinear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population to a one-dimensional distribution of $\eta^\star(x_f, \nset{x})$, which is a function of focal predictor and the observed values of non-focal predictors, as opposed to the definition in Equation~\ref{eq:eta_mean}.
 More specifically:
@@
 
+\bmb{does averaging the CI ...}\BC{Theoretical proof or just a statement?}
+
 For models with random effects components, we make further adjustment to correct for bias induced by the random effects terms.
@@
 
-As mentioned above, the two common choices to generate predictions are: 1) setting non-focal predictors to their mean -- centered- (mean-) based; and 2) using the entire population of non-focal predictors -- whole-sample-based.
+As mentioned above, the two common choices to generate predictions are: 1) setting non-focal predictors to their mean -- mean-based; and 2) using the entire population of non-focal predictors -- whole-sample-based.
 If the link function is nonlinear and/or the model has complex higher order interactions, the average of the predictions evaluated at the mean of non-focal predictors and the average of predictions evaluated at the population level (and then averaged) of non-focal predictor are not equivalent, i.e.,
@@
 
-If $g()$ is strictly concave or strictly convex, one can use Jensen's inequality to determine which side of Equation~\ref{eq:compare_anchored_pop_based} is smaller or larger.
-However, for non-fully convex or concave link function, one can use Taylor series expansion to approximate the range over which the two are smaller or smaller than the other \citep{hanmer2013behind}.
+If $g()$ is strictly concave or strictly convex, one can use Jensen's inequality to determine which side of Equation~\ref{eq:compare_anchored_pop_based} is decelerating or accelerating.
+For accelerating functions, the Jensen’s inequality describes how the changes in input values elevate the predictions and otherwise describes how these changes depress the predictions.
+For a non-fully convex or concave link function, one can use Taylor series expansion to approximate the range over which the two are smaller or smaller than the other \citep{hanmer2013behind}.
+\bmb{Just knowing is not enough and discuss much earlier}\BC{Clarify with BMB the current version and appropriate section to move to}
 
@@
 This may limit the inferences we can make about the entire population.
-In general, the mean-based approach provides the predictions of an average case, whereas, the whole-sample-based approach, summarizes the predictions over the entire population -- and in some applications, the effect of an average case might not be generalizable to the entire population, especially, if the average does not represent the population.
-The whole-sample-based focuses on specific observations since the prediction is first obtained for each observation and then averaged across the levels of the focal predictor.
+In general, the mean-based approach provides the predictions of an average case, whereas the whole-sample-based approach, summarizes the predictions over the entire population -- and in some applications, the effect of an average case might not be generalizable to the entire population, especially, if the average does not represent the population.
+In particular the whole-sample-based approach focuses on specific observations since the prediction is first obtained for each observation and then averaged across the levels of the focal predictor.
 
@@
 For example, a dummy variable for christian household heads may be set to $0.2$, translating to prediction for an household head who is $20\%$ christian.
-This problem may extend to models with other complex interaction terms \citep{hanmer2013behind}.
+This may be problematic in models with complex interaction terms in which the values may not be converge to population mean\citep{hanmer2013behind}.
 
@@
 For instance, similar to the mean-based approach, in the case of continuous focal predictors, choosing the representative values of the focal predictors can be very challenging especially if the cases are not evenly distributed around the minimum and the maximum values or within some subgroups defined in the population.
-In addition, whole-sample-based approach can be very computationally intensive for large datasets.
+In addition, whole-sample-based approach can be computationally intensive for large datasets.
 
@@
 \begin{figure}[!h]
-\centering
-\includegraphics[width=0.8\textwidth]{cubic_predictors_preds.ggp.pdf}
+\begin{center}
+\includegraphics{cubic_predictors_preds.ggp.pdf}
+\end{center}
 \caption{{\bf Prediction-styled and effect-styled plots for cubic polynomial and simple focal predictors.} A: The focal predictor is a cubic polynomial.
@@
 In the case of cubic polynomial focal predictor, the model center is not a point in the predictor space, i.e., mean of the focal predictor, but a combination of all the model variables in the cubic polynomial, consequently, mean-anchored curves do not cross at a point -- A.
-The horizontal black and red dotted lines are the observed and predicted average household size, i.e., $\over{hh~size}$ and $\over{\widehat{hh~size}}$, respectively; the inner red dotted curves and crossing lines are the anchored confidence bands for A and B, respectively, while the outer blue curves are the non-anchored confidence bands in both cases.
-The central curve and trend line, in A and B, respectively, are the predicted household size based on age and wealth index, respectively.}
+The horizontal black and red dotted lines are the observed and predicted average household size, i.e., $\over{hh~size}$ and $\over{\widehat{hh~size}}$, respectively; the inner red dotted curves and crossing lines are the anchored confidence bands for A and B, respectively; while the outer blue curves are the non-anchored confidence bands in both cases.
+The central curve and trend line, in A and B, respectively; are the predicted household size based on age and wealth index, respectively.}
 \label{fig:pred_cubic_plots}
@@
 \begin{figure}[!h]
-\centering
-\includegraphics[width=0.8\textwidth]{glm_two_predictor_preds.ggp.pdf}
+\begin{center}
+\includegraphics{glm_two_predictor_preds.ggp.pdf}
+\end{center}
 \caption{{\bf Mean-based and whole-sample-based predictions.} The two approaches produce relatively different estimates in each of the predictors but the difference is more pronounced when the effect of the particular focal predictor has no strong effect -- \emph{on the left}.
@@
 \begin{figure}[!h]
-\centering
-\includegraphics[width=0.8\textwidth]{mediate_preds.ggp.pdf}
+\begin{center}
+\includegraphics{mediate_preds.ggp.pdf}
+\end{center}
 \caption{{\bf Mean-based and whole-sample-based predictions for mediated effects.} These figures compare the mean-based and whole-sample-based predictions for the mediated (multivariate) and non-mediated (univariate) models.

### 2022-05-25 manuscript/body.tex
Some refactoring and some MS comments (and tiny changes)

--- 2022-05-14 manuscript/body.tex
@@
+%% Main tex
+
 \linenumbers
+
+\jd{How did all the figures wind up in back?}
 
@@
 
+\jd{Can we stake out early what we mean by predictions vs.~effects?
+It still seems muddled together.
+I guess it's complicated because in many cases the estimate is the same (only CIs differ).
+But I think we should do it anyway.}
 In many applications which use simple or generalized linear (mixed) models, outcome predictions or probabilities, are often of interest.
@@
 An alternative is to explore the \emph{effects} of predictors on the predictions or probabilities of the outcome at various level of the predictor of interest -- \emph{prediction and effects} \citep{fox2009effect, leeper2017package, lenth2018package}.
-For example, as a way to plan, public health officials may want to know the effects of household income, wealth index, etc, on the predicted probability of having improved water services among slum dwellers.
+\jd{Are we still saying these (prediction and effects) are the same?
+Or are we now saying they're different, with the same citations?} For example, as a way to plan, public health officials may want to know the effects of household income, wealth index, etc, on the predicted probability of having improved water services among slum dwellers.
 
@@
 
-The most common way of dealing with the first challenge is taking unique levels of the focal (predictor of interest) predictor if discrete or taking appropriately sized quantiles (or bins) if continuous, and then calculating the predictions while holding non-focal (other predictors other than the focal predictor) predictors at their reference point (e.g., means) \citep{hanmer2013behind}.
-This generates -- \emph{predictions}, \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}.
+The most common way of dealing with the first challenge is taking unique levels of the focal (predictor of interest) predictor if discrete or taking appropriately sized quantiles (or bins) \jd{Don't understand what the bins are doing here (when we are holding the reference point constant); it's not what we do, is it?} if continuous, and then calculating the predictions while holding non-focal (other predictors other than the focal predictor) predictors at their reference point (e.g., means) \citep{hanmer2013behind}.
+This generates \jd{Are predictions and predictor effects really used for the same thing?} -- \emph{predictions}, \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}.
 In this article, we refer to this quantity as predictions since it should, for example, tell us what we would expect the response to be at a particular value or level of the predictor, for an ``average case''.
-More specifically, we compute the expected outcome by meaningfully holding non-focal predictors constant (or averaged in some meaningful way) while varying the focal predictor, with the goal that the outcome expected prediction represents how the model responds to the changes in the focal predictor.
+More specifically, we compute the expected outcome by holding non-focal predictors constant (or averaged in some meaningful way) while varying the focal predictor, with the goal that the outcome expected prediction represents how the model responds to the changes in the focal predictor.
 On the other hand, effects differs with predictions in the sense that the former is about how we describe the uncertainty around the latter and depends on the anchor.
+\jd{I don't understand this: certainly both predictions and effects have uncertainty.}
 
 The commonly used \proglang{R} software packages (\pkg{emmeans} and \pkg{effects}) for generating predictions, by default, use the average of non-focal predictors as the reference point. 
+\jd{This is ambiguous; we need to be talking about input variables and model variables by now. Also: what do the packages do about categorical non-focal predictors? Do we know?}
 However, there are a number of choices one can make when considering this approach -- for example, in the presence of interaction, averaging the interactions (averaging product of interacting model variables) versus product of the averages of the interacting model variables (default for \pkg{emmeans} and \pkg{effects}).
@@
 We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$ and $\beta_{23} = 5$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.
+\jd{What are true predictions? Why does it look like effects matches varpred, but it sounds like we're saying it matches emmeans?}
 
 \begin{figure}[!h]
-\centering
-\includegraphics[width=0.9\textwidth]{justify_preds.inter.pdf}
+\begin{center}
+\includegraphics{justify_preds.inter.pdf}
+\end{center}
 \caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} predictions for $x_1$ on $y$ for models with and without interactions.}
@@
 
+\jd{Why do we say ``similar"? Are they not the same?}
 In the absence of interaction (model~\ref{eq:simple_inter_higher_no_interaction}), the three approaches produce similar estimates, which match the simulated values, Fig~\ref{fig:justify_plots}A.
@@
 
+\jd{Hard to parse.
+Why ``for example''?
+The key here is that the lm case works out well because it's all about linear averaging (the mean prediction is the prediction from the mean point), whereas the glm case is non-linear averaging.
+There should be a link also to why we use the ``model center'' (we're using the correct center point for the linear averaging to work out.}
 When dealing with nonlinear link functions or models with multiple non-focal predictors, the correct predictions, for example, are even much harder to estimate.
@@
 \section*{Quantities of interest}
+
+\jd{Is this borrowed from me? I'm not really following the structure of the MS right now. Maybe I'll be clearer in the morning.}
 

### 2022-05-25 manuscript/jd_notes.Rmd
Autocommit

@@
 ---
+
+**Bicko**:
+
+- It seems distinguishing between prediction and effects is a bit hard. How about prediction and prediction effects.
+	- Predictions: the central line
+	- Prediction (predictor?) effects: the curves which describe our uncertainty around the predictions and can be non-isolated (non-anchored) or isolated (anchored)?
 

### 2022-05-25 manuscript/body.tex
Autocommit

--- 2022-05-25 manuscript/body.tex
@@
 
+\jd{How did all the figures wind up in back?}
+
 % Use "Eq" instead of "Equation" for equation citations.
 \section*{Introduction}
+
+\jd{Can we stake out early what we mean by predictions vs.~effects?
+It still seems muddled together.
+I guess it's complicated because in many cases the estimate is the same (only CIs differ).
+But I think we should do it anyway.}
 
@@
 Further, coefficient estimates of GLM models involving nonlinear link functions lack direct interpretation\citep{leeper2017interpreting}, meaning that interpretation of derived quantities from these estimates requires some understanding of the specified model.
-An alternative is to explore the \emph{effects} of predictors on the predictions of the outcome at various levels of the predictor of interest -- \emph{predictions and effects} \citep{fox2009effect, leeper2017package, lenth2018package}.
+An alternative is to explore the \emph{effects} of predictors on the predictions of the outcome at various levels of the predictor of interest -- \emph{predictions and effects}\jd{Are we still saying these (prediction and effects) are the same?
+Or are we now saying they're different, with the same citations?} \citep{fox2009effect, leeper2017package, lenth2018package}.
 For example, as a way to plan, public health officials may want to know the effects of household income, wealthindex, etc, on the predicted probability that slum dwellers will have improved water services.
@@
 \begin{enumerate}
-\item choice of representative values of \emph{focal} predictor and the \emph{reference point} for non-focal predictors especially in multivariate models
+\item choice of representative values of \emph{focal} predictor and the \emph{reference point} for non-focal predictors, especially in multivariate models
 \item  propagation of uncertainty -- appropriate choice of \emph{anchor} for computing confidence intervals; and how to incorporate the uncertainty due to non-focal predictors 
@@
 
-The most common way of choosing the representative values is taking unique levels of the focal predictor if discrete or taking appropriately sized quantiles if continuous, and then calculating the predictions while holding non-focal (other predictors other than the focal predictor) predictors at their reference point (e.g., means) \citep{hanmer2013behind}.
-This procedure generates -- \emph{predictions}, \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}.
+The most common way of choosing the representative values is taking unique levels of the focal predictor if discrete or taking appropriately sized quantiles \jd{Don't understand what the bins are doing here (when we are holding the reference point constant); it's not what we do, is it?} if continuous, and then calculating the predictions while holding non-focal (other predictors other than the focal predictor) predictors at their reference point (e.g., means) \citep{hanmer2013behind}.
+This procedure generates \jd{Are predictions and predictor effects really used for the same thing?} -- \emph{predictions}, \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}.
 In this article, we refer to these quantities as predictions since they should, for example, tell us what we would expect the response to be at a particular value or level of the predictor, for an ``average case''.
@@
 The most \bmb{commonly used \proglang{R} software packages (\pkg{emmeans} and \pkg{effects})}\BC{not sure the underline} for generating predictions, by default, use the average of non-focal predictors as the reference point.
+\jd{This is ambiguous; we need to be talking about input variables and model variables by now. Also: what do the packages do about categorical non-focal predictors? Do we know?}
 However, there are a number of choices one can make when considering this approach -- for example, in the presence of interaction, averaging the interactions (averaging the product of interacting model variables) versus taking the product of the averages of the interacting model variables (the default for \pkg{emmeans} and \pkg{effects}).
@@
 We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$ $\beta_{23} = 5$, $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and $\sigma^2 = 1$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.
+\jd{What are true predictions? Why does it look like effects matches varpred, but it sounds like we're saying it matches emmeans?}
 
-\begin{figure}[!h]
+\begin{figure}[h]
 \begin{center}
@@
 
-\begin{figure}[!h]
+\begin{figure}[h]
 \begin{center}
@@
 
+\jd{Hard to parse.
+Why ``for example''?
+The key here is that the lm case works out well because it's all about linear averaging (the mean prediction is the prediction from the mean point), whereas the glm case is non-linear averaging.
+There should be a link also to why we use the ``model center'' (we're using the correct center point for the linear averaging to work out.}
+
 When dealing with nonlinear link functions or models with multiple non-focal predictors, the correct predictions are much harder to estimate.

### 2022-05-25 manuscript/body.tex
Predictions and predictor effects?

@@
 Or are we now saying they're different, with the same citations?} \citep{fox2009effect, leeper2017package, lenth2018package}.
-For example, as a way to plan, public health officials may want to know the effects of household income, wealthindex, etc, on the predicted probability that slum dwellers will have improved water services.
+For example, as a way to plan, public health officials may want to know the effects of household \code{income}, \code{wealthindex}, etc, on the predicted probability that slum dwellers will have improved water services.
 
@@
 We start by generating two predictions from the fitted model \nameref{S1_Appendix}: 1) one with cubic polynomial of \code{age} as a focal predictor; and 2) one with a simple focal predictor (non-polynomial) of \code{Wealthindex}.
-Then we compared non-anchored (more of predictions) and mean-anchored (effects) confidence intervals around the predicted values (see Fig~\ref{fig:pred_cubic_plots}).
+Then we compared non-anchored and mean-anchored (effects) confidence intervals around the predicted values (see Fig~\ref{fig:pred_cubic_plots}).
 
@@
 \caption{{\bf Prediction-styled and effect-styled plots for cubic polynomial and simple focal predictors.} A: The focal predictor is a cubic polynomial.
-B: The focal predictor is not a polynomial function, implying that the complex interaction is in non-focal predictor.
-In both cases, the mean-anchored effects are narrower than the non-anchored (predictions), but crosses at the mean of the focal predictor (model center) only in the case of simple (non-polynomial) focal predictor -- B.
-In the case of cubic polynomial focal predictor, the model center is not a point in the predictor space, i.e., mean of the focal predictor, but a combination of all the model variables in the cubic polynomial, consequently, mean-anchored curves do not cross at a point -- A.
-The horizontal black and red dotted lines are the observed and predicted average household size, i.e., $\over{hh~size}$ and $\over{\widehat{hh~size}}$, respectively; the inner red dotted curves and crossing lines are the anchored confidence bands for A and B, respectively; while the outer blue curves are the non-anchored confidence bands in both cases.
-The central curve and trend line, in A and B, respectively; are the predicted household size based on age and wealth index, respectively.}
+B: The focal predictor is a linear function, with a cubic polynomial non-focal predictor.
+In both cases, the mean-anchored effects are narrower than the non-anchored (predictions), but intersects at the mean of the focal predictor (model center) only in the case of linear focal predictor -- B.
+In the case of cubic polynomial focal predictor, the model center is not a point in the predictor space, i.e., mean of the focal predictor, but a combination of all the model variables in the cubic polynomial, consequently, mean-anchored curves will not necessarily cross at a point -- A.
+The horizontal black and red dashed lines are the observed and predicted average household size, i.e., $\over{hh~size}$ and $\over{\widehat{hh~size}}$, respectively; the inner red dotted curves and crossing lines are the effects for A and B, respectively; while the outer blue curves are the \BC{predictions (place holder for now)} in both cases.
+The central curve and trend line, in A and B, respectively; are the predicted household size based on age and wealth index, respectively.
+\bmb{Use different line types ...}\BC{Confirm with BMB}}
 \label{fig:pred_cubic_plots}
@@
 
-In the absence of complex higher order interaction terms or transformations on the focal predictor in simple linear models, we would expect, at the model center, the observed average, $\over{hh~size}$, and the average of the predictions, $\over{\widehat{hh~size}}$, to be identical and cross with the mean-anchored confidence bands (effects), at the model center (see Fig~\ref{fig:pred_cubic_plots}B).
+In the absence of complex higher order interaction terms or transformations on the focal predictor in simple linear models, we would expect, at the model center, the observed average, $\over{hh~size}$, and the average of the predictions, $\over{\widehat{hh~size}}$, to be identical and all intersect at the mean-anchored confidence bands (effects), at the model center (Fig~\ref{fig:pred_cubic_plots}B).
 On the other hand, if the focal predictor is characterized by complex interactions or any other form of transformation, the model center is not necessarily a point in the predictor space.
-In this case, the mean-anchored confidence bands will be narrower than the non-anchored ones but not necessarily crossing at the model center (see Fig~\ref{fig:pred_cubic_plots}A).
+In this case, the mean-anchored confidence bands will be narrower than the non-anchored ones but will not necessarily cross at the model center (Fig~\ref{fig:pred_cubic_plots}A).
 

### 2022-05-26 manuscript/body.tex
Autocommit

@@
 Further, coefficient estimates of GLM models involving nonlinear link functions lack direct interpretation\citep{leeper2017interpreting}, meaning that interpretation of derived quantities from these estimates requires some understanding of the specified model.
-An alternative is to explore the \emph{effects} of predictors on the predictions of the outcome at various levels of the predictor of interest -- \emph{predictions and effects}\jd{Are we still saying these (prediction and effects) are the same?
+An alternative is to explore the \emph{effects} of predictors on the predictions of the outcome at various levels of the predictor of interest \jd{Are we still saying these (prediction and effects) are the same?
 Or are we now saying they're different, with the same citations?} \citep{fox2009effect, leeper2017package, lenth2018package}.
@@
 
-When visually presented, prediction and effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind of transformations on the predictors whose estimates are usually, but not always, harder to interpret.
-Generating predictions and effects plots together with the associated confidence intervals for regression models has a number of challenges.
+When visually presented, predictions provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind of transformations on the predictors whose estimates are usually, but not always, harder to interpret.
+Generating predictions together with the associated confidence intervals for regression models has a number of challenges.
 In particular:

### 2022-06-13 manuscript/body.tex
All chapters draft
Draft sync

@@
 % Use "Eq" instead of "Equation" for equation citations.
-\section*{Introduction}
+\section{Introduction}
 
@@
 Predictions provide a great way to summarize what the regression model is telling us and very useful for interpreting and visualizing model estimates.
-For example, in logistic regression models, the coefficient estimates, when reported as log odds, are not easy to interpret and less informative.
-Since logistic models are nonlinear (due to the nonlinear link function), in multivariate models with\bmb{?only}\BC{clarify with BMB} interactions, the magnitude of the effect of change in the outcome depends on the values of the predictor of interest and other predictors.
+For example, in logistic regression models, the coefficient estimates, when reported as log odds, are not easy to interpret and are less informative.
+Since logistic models are nonlinear (due to the nonlinear link function), in multivariate models interactions or other nonlinear transformations, the magnitude of the effect of change in the outcome depends on the values of the predictor of interest and other predictors.
 Hence, the conclusions one can make about the estimated effects greatly depends on how well we choose the values of the other predictors, i.e., \emph{non-focal} predictors.
@@
 The coefficient estimates from OLS regression models are generally easy to interpret as the expected change in the outcome with a unit change in the continuous predictor or as the expected outcome difference between the levels of the categorical predictor with the baseline category.
-However, with additional complexities such interaction between predictors, the resulting coefficients risk misinterpretation or interpretation complexity \citep{brambor_understanding_2006, berry_improving_2012}.
+However, with additional complexities such as interactions between predictors, the resulting coefficients risk misinterpretation or interpretation complexity \citep{brambor_understanding_2006, berry_improving_2012}.
 
-Simple OLS regression models are flexible and can easily be generalized, as generalized linear (mixed) models (GL(M)Ms), to examine more complex relationships, including \bmb{nonlinear relationships -- need to clarify}\BC{clarity with BMB} between response and predictors e.g., log transformed response or predictors, interactions between predictors (and via splines for example), and nonlinear transformations via link functions \bmb{due to their flexibility}\BC{clarify why BMB circled it}.
+Simple OLS regression models are flexible and can easily be generalized, as generalized linear (mixed) models (GL(M)Ms), to examine more complex relationships, including nonlinear relationships between response and predictors e.g., nonlinear transformation on the predictors, interactions between predictors (and via splines for example), and nonlinear transformations via link functions.
 This flexibility comes with additional complications, for example, complex multivariate models may risk misinterpretation, and miscalculation of quantities of interest \citep{berry_improving_2012, leeper2017interpreting}.
 Further, coefficient estimates of GLM models involving nonlinear link functions lack direct interpretation\citep{leeper2017interpreting}, meaning that interpretation of derived quantities from these estimates requires some understanding of the specified model.
-An alternative is to explore the \emph{effects} of predictors on the predictions of the outcome at various levels of the predictor of interest \jd{Are we still saying these (prediction and effects) are the same?
-Or are we now saying they're different, with the same citations?} \citep{fox2009effect, leeper2017package, lenth2018package}.
+An alternative is to explore \emph{predictions} and \emph{effects} \citep{fox2009effect, leeper2017package, lenth2018package}.
 For example, as a way to plan, public health officials may want to know the effects of household \code{income}, \code{wealthindex}, etc, on the predicted probability that slum dwellers will have improved water services.
 
-When visually presented, predictions provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind of transformations on the predictors whose estimates are usually, but not always, harder to interpret.
-Generating predictions together with the associated confidence intervals for regression models has a number of challenges.
+Both predictions and effects tell us what we expect the response to be for a given value of \emph{focal} predictor (predictor of interest).
+The major difference between the two lies in the construction of the associated confidence intervals.
+In particular, for predictions, we want to use the classic curved confidence intervals (or prediction intervals) which capture the uncertainty in the whole model and generally focus on the \emph{total} effects, and so will often be suited for a univariate model.
+On the other hand, effects, attempt to show the effect of a focal predictor.
+In this case, we suggest the narrower confidence intervals which focus on uncertainty in the effect of the focal predictor only, and so will often be suited for the direct effects – after controlling for other predictors – from a multivariate model.
+
+When visually presented, predictions and effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind of transformations on the predictors whose estimates are usually, but not always, harder to interpret.
+Generating these quantities together with the associated confidence intervals for regression models has a number of challenges.
 In particular:
 \begin{enumerate}
-\item choice of representative values of \emph{focal} predictor and the \emph{reference point} for non-focal predictors, especially in multivariate models
-\item  propagation of uncertainty -- appropriate choice of \emph{anchor} for computing confidence intervals; and how to incorporate the uncertainty due to non-focal predictors 
+\item choice of representative values of focal predictor and the \emph{reference point} for non-focal predictors, especially in multivariate models
+\item  uncertainty estimation -- appropriate choice of \emph{anchor} for computing confidence intervals; and how to incorporate the uncertainty due to non-focal predictors 
 \item bias in the expected mean prediction induced by the nonlinear transformation of the response variable (especially in GL(M)Ms)
@@
 
-The most common way of choosing the representative values is taking unique levels of the focal predictor if discrete or taking appropriately sized quantiles \jd{Don't understand what the bins are doing here (when we are holding the reference point constant); it's not what we do, is it?} if continuous, and then calculating the predictions while holding non-focal (other predictors other than the focal predictor) predictors at their reference point (e.g., means) \citep{hanmer2013behind}.
-This procedure generates \jd{Are predictions and predictor effects really used for the same thing?} -- \emph{predictions}, \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}.
-In this article, we refer to these quantities as predictions since they should, for example, tell us what we would expect the response to be at a particular value or level of the predictor, for an ``average case''.
-\bmb{More specifically, we compute the expected outcome by holding non-focal predictors constant (or averaged in some meaningful way) while varying the focal predictor, with the goal that the predictions represents how the model responds to the changes in the focal predictor.
-On the other hand, effects differ with predictions in the sense that the former is about how we describe the uncertainty around the latter and depends on the anchor.
--- see next/unclear}\BC{clarify with BMB}
+The most common way of choosing the representative values is taking unique levels of the focal predictor if discrete or taking appropriately sized quantiles if continuous, and then calculating the predictions while holding non-focal (other predictors other than the focal predictor) predictors at their reference point (e.g., means) \citep{fox2009effect, hanmer2013behind}.
+This procedure generates -- predictions (at least in our case), \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}.
+In this article, we refer to these quantities as predictions and effects since they should, for example, tell us what we would expect the response to be at a particular value or level of the predictor, for an ``average case''.
+More specifically, the basic idea is to compute the expected outcome by holding the non-focal predictors constant, for example at their mean (centered), while varying the focal predictor, with the goal that estimates represent how the model responds to the changes in the focal predictor.
+As mentioned above, in our case, the key difference between predictions and effects is in the computation of the confidence intervals.
 
-The most \bmb{commonly used \proglang{R} software packages (\pkg{emmeans} and \pkg{effects})}\BC{not sure the underline} for generating predictions, by default, use the average of non-focal predictors as the reference point.
-\jd{This is ambiguous; we need to be talking about input variables and model variables by now. Also: what do the packages do about categorical non-focal predictors? Do we know?}
+The most commonly used \proglang{R} software packages for generating predictions (\pkg{emmeans} and \pkg{effects}), by default, use the average of input and model variables as the reference point for continuous and categorical predictors, respectively.
 However, there are a number of choices one can make when considering this approach -- for example, in the presence of interaction, averaging the interactions (averaging the product of interacting model variables) versus taking the product of the averages of the interacting model variables (the default for \pkg{emmeans} and \pkg{effects}).
 For simple models (with no interactions), the two approaches give the same predictions.
-However, for complex models, we claim that neither of these two approaches is the most appropriate but we show that averaging the interaction closely matches the observed values.
+However, for complex models, we demonstrate that the two approaches can produce substantially different results and show that averaging the interactions (model variables) closely matches the observed values.
 To illustrate this, consider models~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher} below, with $x_1$ as the focal predictor:
-\bmb{This whole example .... example of nonlinear averaging}\BC{Yes and handling interaction terms?}
 %
@@
 %
-
-We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$ $\beta_{23} = 5$, $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and $\sigma^2 = 1$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.
-\jd{What are true predictions? Why does it look like effects matches varpred, but it sounds like we're saying it matches emmeans?}
+We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$ $\beta_{23} = 5$, $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and $\sigma^2 = 1$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions (predictions based on the true simulation parameters) and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.
 
@@
 
-\section*{Quantities of interest}
+\section{Quantities of interest}
 
@@
 
-\subsection*{Input variables}
+\subsection{Input variables}
 
@@
 
-\subsection*{Model matrix}
+\subsection{Model matrix}
 
@@
 
-\subsection*{Model variables}
+\subsection{Model variables}
 
@@
 
-\subsection*{Model center}
+\subsection{Model center}
 
@@
 
-\subsection*{Reference point}
+\subsection{Reference point}
 
@@
 
-\subsection*{Anchor}
+\subsection{Anchor}
 
@@
 
-\subsection*{Prediction- vs. effect- styled plots}
+\subsection{Prediction- vs. effect- styled plots}
 
@@
 
-\subsection*{Predictions vs. effects}
+\subsection{Predictions vs. effects}
 
@@
 
-\section*{Statistical background}
+\section{Statistical background}
 
@@
 
-\subsection*{Dealing with higher order interactions}
+\subsection{Dealing with higher order interactions}
 
@@
 
-\section*{Uncertainty estimation}
+\section{Uncertainty estimation}
 
@@
 
-\subsection*{Model center based}
+\subsection{Model center based}
 
@@
 
-%% \subsection*{Is center-anchored better?}
+%% \subsection{Is center-anchored better?}
 %% 
@@
 
-\section*{Bias correction}
+\section{Bias correction}
 
@@
 
-\subsection*{Whole-sample-based approach for bias correction}
+\subsection{Whole-sample-based approach for bias correction}
 
@@
 
-\section*{Mean-based vs. whole-sample-based}
+\section{Mean-based vs. whole-sample-based}
 
@@
 
-\section*{Simulation examples}
+\section{Simulation examples}
 
@@
 
-\subsection*{Predictions-styled and effects-styled plots}
+\subsection{Predictions-styled and effects-styled plots}
 
@@
 
-\subsection*{Bias correction}
+\subsection{Bias correction}
 
@@
 
-\subsection*{Mediated effect}
+\subsection{Mediated effect}
 
@@
 
-\section*{Discussion}
+\section{Discussion}
 
@@
 
-\section*{Conclusion}
+\section{Conclusion}
 
@@
 
-\section*{Supporting information}
+\section{Supporting information}
 

### 2022-06-13 manuscript/body.tex
Autocommit

@@
 \section{Introduction}
-
-\jd{Can we stake out early what we mean by predictions vs.~effects?
-It still seems muddled together.
-I guess it's complicated because in many cases the estimate is the same (only CIs differ).
-But I think we should do it anyway.}
 
@@
 
-The most commonly used \proglang{R} software packages for generating predictions (\pkg{emmeans} and \pkg{effects}), by default, use the average of input and model variables as the reference point for continuous and categorical predictors, respectively.
+The most commonly used \proglang{R} software packages for generating predictions (\pkg{emmeans} and \pkg{effects}), by default, use the average of \emph{input} and \emph{model variables} as the reference point for continuous and categorical predictors, respectively.
 However, there are a number of choices one can make when considering this approach -- for example, in the presence of interaction, averaging the interactions (averaging the product of interacting model variables) versus taking the product of the averages of the interacting model variables (the default for \pkg{emmeans} and \pkg{effects}).
@@
 We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$ $\beta_{23} = 5$, $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and $\sigma^2 = 1$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions (predictions based on the true simulation parameters) and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.
-
+%
 \begin{figure}[h]
@@
 \end{figure}
-
-\jd{Why do we say ``similar"? Are they not the same?}
-In the absence of interaction (model~\ref{eq:simple_inter_higher_no_interaction}), the three approaches produce similar estimates, which match the simulated values, Fig~\ref{fig:justify_plots}A.
+%
+In the absence of interaction (model~\ref{eq:simple_inter_higher_no_interaction}), the three approaches produce the same estimates, which match the simulated values, Fig~\ref{fig:justify_plots}A.
 However, in the presence of interaction, even as simple as the one in model~\ref{eq:simple_inter_higher}, the estimates starts to differ.
 In particular, \pkg{emmeans} and \pkg{effects} give similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is very close to the simulated average ($\bar{y}$), Fig~\ref{fig:justify_plots}B.
-To generate Fig~\ref{fig:justify_plots}A, all the three packages averages non-focal predictors $x_2$ and $x_3$.
+To generate Fig~\ref{fig:justify_plots}A, all the three packages average $x_2$ and $x_3$.
 On other hand, to generate Fig~\ref{fig:justify_plots}B, the difference in the estimates lies on how each of the packages average the interaction term ($x_2x_3$).
-In particular \pkg{emmeans} and \pkg{effects} uses \emph{input variables} to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses \emph{model variables} to compute $\over{x_2x_3}$.
+In particular \pkg{emmeans} and \pkg{effects} use \emph{input variables} to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses \emph{model variables} to compute $\over{x_2x_3}$.
+In the case of Fig~\ref{fig:justify_plots}A, averaging the input variables is equivalent to averaging the model variables.
 
@@
 
-\jd{Hard to parse.
-Why ``for example''?
-The key here is that the lm case works out well because it's all about linear averaging (the mean prediction is the prediction from the mean point), whereas the glm case is non-linear averaging.
-There should be a link also to why we use the ``model center'' (we're using the correct center point for the linear averaging to work out.}
-
-When dealing with nonlinear link functions or models with multiple non-focal predictors, the correct predictions are much harder to estimate.
-One approach is to make predictions on the transformed scale (linear predictor scale), and then back-transform to the original scale.
-However, the back-transformation may either result in biased predictions or requires some approximation.
+For linear models (including models with complex interactions), the averaging is done on the linear scale, i.e., linear averaging.
+As a result the bias in the expected (mean) predictions can easily be corrected by performing predictions from the \emph{model center} (average of the model variables) or the center point.
+The mean prediction is the prediction at the model center.
+On the other hand, when dealing with GL(M)Ms with nonlinear link functions, the averaging is done on the nonlinear scale, i.e., nonlinear averaging which makes estimating correct predictions much harder.
 In particular, bias in expected mean prediction induced by nonlinear transformation of the response variable can lead inaccurate predictions.
+One way to deal with this is to make predictions on the transformed scale (linear predictor scale), and then back-transform to the original scale.
+However, the back-transformation may either result in biased predictions or requires some approximation, for example, second order Taylor approximation implemented in \pkg{emmeans} \citep{lenth2018package}.
+In linear models, mean-based (alias center point or model center) gives estimates consistent with the observed values, however, in models with nonlinear link functions, this is not always the case.
 An alternative to mean-based reference point, i.e., averaging non-focal predictors, is the \emph{whole-sample-based} approach, discussed in detail later, which involves computing the prediction over the population of non-focal predictors and then averaging across the values of the focal predictor \citep{hanmer2013behind}.

### 2022-06-14 manuscript/abstract.tex
Adopting th notion of predictors as opposed to model variables

@@
 {\Large
-\textbf\newline{Describing the curves: uncertainty estimation and bias correction for predictor effects in simple and generalized linear (mixed) models} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+\textbf\newline{Describing the curves: uncertainty estimation and bias correction for predictions in simple and generalized linear (mixed) models} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
 }
@@
 In applications that use generalized linear (mixed) models, outcome predictions are often of interest.
-For models that involve complex multiplicative interactions, additional non-focal predictors or nonlinear link functions, the estimated coefficients are not readily interpretable.
-A general way to summarize these kind of models is through predictions, that depend not only on choice of representative values of focal predictor but are also sensitive to which values of non-focal predictors are chosen.
+For models that involve complex multiplicative interactions, additional non-focal input variables or nonlinear link functions, the estimated coefficients are not readily interpretable.
+A general way to summarize these kind of models is through predictions, that depend not only on choice of representative values of focal input variables but are also sensitive to which values of non-focal input variables are chosen.
 The most common approach is generating predictions at a reference point, usually the mean, of non-focal predictor, i.e., mean-based (centered) which could be considered as the effect of an average case in the population.
-In the presence of additional non-focal predictors, nonlinear link functions, random effect terms, etc., mean-based approach generate predictions that are biased and may not be consistent with the observed quantities.
+In the presence of additional non-focal input variables, nonlinear link functions, random effect terms, etc., mean-based approach generate predictions that are biased and may not be consistent with the observed quantities.
 An alternative is the whole-sample-based approach which estimates the average effect in the population.
-Moreover, anchored effects provide an alternative and a more clear way to describe uncertainty associated with the focal predictor of interest.
+Moreover, anchored effects provide an alternative and a more clear way to describe uncertainty associated with the focal input variable.
 In addition to theoretical and methodical comparison, using simulation, we illustrate the two approaches and show that they can produce substantially different results and that whole-sample-based approach can not only produce estimates consistent with the observed values, but also appropriate for bias correction.
 We also present an alternative way, anchored confidence intervals, to describe uncertainties associated with these predictions.
-
-\section*{Definitions}
-
-\begin{itemize}
-\item \textbf{Input variables}: Scientific variables underlying an inference or exploration. The focal predictor we use for effects or predictions is an input variable.
-\item \textbf{Model variables}: Variables that represent columns in the model matrix.
-Each input variable will correspond to one or more model variables.
-In particular, variables with more than two categories, or variables modeled with a spline or polynomial response, will correspond to more than one model variables.
-\item\textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more model variables).
-The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable.
-\item \textbf{Reference point}: Value or values chosen for \emph{non-focal predictors}, when estimating effects.
-Typically the center point, but can instead be a population of quantiles or observations.
-\item \textbf{Anchor}: The value chosen for the \emph{focal predictor} when estimating effect confidence intervals (anchor choice does not affect the estimate).
-Typically chosen as the center point of the model variables corresponding to the focal input variable.
-\end{itemize}
 

### 2022-06-14 manuscript/body.tex
Adopting th notion of predictors as opposed to model variables

@@
 \linenumbers
-
-\jd{How did all the figures wind up in back?}
 
@@
 For example, in logistic regression models, the coefficient estimates, when reported as log odds, are not easy to interpret and are less informative.
-Since logistic models are nonlinear (due to the nonlinear link function), in multivariate models interactions or other nonlinear transformations, the magnitude of the effect of change in the outcome depends on the values of the predictor of interest and other predictors.
-Hence, the conclusions one can make about the estimated effects greatly depends on how well we choose the values of the other predictors, i.e., \emph{non-focal} predictors.
+Since logistic models are nonlinear (due to the nonlinear link function), in multivariate models interactions or other nonlinear transformations, the magnitude of the effect of change in the outcome depends on the values of the input variable of interest and the other remaining input variables.
+Hence, the conclusions one can make about the estimated predictions greatly depends on how well we choose the values of the other input variables, i.e., \emph{non-focal} input variables.
 
@@
 
-\section{Quantities of interest}
+\section{Definitions}
 
-\jd{Is this borrowed from me? I'm not really following the structure of the MS right now. Maybe I'll be clearer in the morning.}
+In order to discuss more general statistical background and mathematical formulation, we need to understand and formally define a number of terms, some of which have been introduced in the previous section.
 
-Several quantities of interest may be derived from regression models.
-The first one is the coefficient estimates.
-Others are \emph{input and model variables}, \emph{model matrix}, \emph{model center}, \emph{reference point} and \emph{anchor}.
-
-\subsection{Input variables}
-
-These are scientific variables underlying an inference or exploration. The focal and non-focal predictors we use for effects or predictions are an input variables.
-
-\subsection{Model matrix}
-
-Model matrix refers to the design matrix whose rows include all combination of variables appearing in the interaction terms, along with the typical values of the focal and non-focal predictors.
-
-\subsection{Model variables}
-
-These are variables that represent columns in the model matrix.
+\begin{itemize}
+\item \textbf{Predictors}: Refer to the X-variables which go into the regression model.
+For example, consider the model the includes the interaction in Equation~\ref{eq:simple_inter_higher}.
+This regression model has $5$ predictors -- $x_1, x_2, x_3, x_2x_3$ and the intercept term.
+\item \textbf{Focal and non-focal predictors}: Focal predictor refers to the predictor we are interested in generating its effects on the predictions.
+For multivariate models, all other remaining predictors are non-focal predictors.
+For example, consider model described by Equation~\ref{eq:simple_inter_higher_no_interaction}.
+Suppose we are interested in the effect of $x_1$ on the predicted $y$, then $x_1$ is the focal predictor while $x_2$ and $x_3$ are non-focal predictors.
+\item \textbf{Input variables}: Refers to the observed (or scientific) variables underlying an inference or exploration, and forms the X-variables.
+For example, the regression model described by Equation~\ref{eq:simple_inter_higher} has only $4$ input variables -- $x_1, x_2, x_3$.
+\item \textbf{Model matrix}: Refers to the design matrix whose rows include all combination of variables appearing in the interaction terms.
+Consider an example for three hypothetical households -- the first household head is a Christian with an income of $\$ 50$, the second household head is Muslim with an income of $\$ 100$ while the third household head is a Jew with an income of $\$ 77$.
+Suppose we want to model the household size (\code{hhsize}) as function of these household characteristics, i.e., $$\mathrm{hhsize} = \beta_0 + \beta_1\times\mathrm{income} + \beta_{2[r]}\times\mathrm{religion} + \mathrm{error}.$$ The model matrix corresponding to this model is given by
+$$\begin{bmatrix}{}
+ Intercept & income & religionJew & religionMuslim \\
+ 1 & 50 & 0 & 0 \\
+  1 & 100 & 0 & 1 \\
+  1 & 77 & 1 & 0 \\
+\end{bmatrix}.$$ The first column represents the constant term in our model, $\beta_0$.
+For continuous variables, the representation in the model matrix is the same as the correspond input variables (for example second column representing income).
+For categorical variables, however, by default, the model matrix  creates additional dummy variables using the reference cell parameterization.
+This means that, if an input variable has $L$ factor levels, then there will be $L-1$ dummy columns representing all but the first level created in the model matrix.
+In our example, the column \code{religion} is missing and instead we have \code{religionJew} and \code{religionMuslim}; the missing category \code{religionChristian} is treated as the reference category.
+\item \textbf{Model variables:} It is easy to interchangeably refer to predictors as input variables, and vice versa.
+To avoid this issue, we adopt the term model variables which are variables that represent columns in the model matrix.
 Each input variable may correspond to one or more model variables.
-In particular, variables with more than two categories, or variables modeled with a spline or polynomial response, will correspond to more than one model variables.
-
-\subsection{Model center}
-
-A point corresponding to a column-wise mean of the model matrix (the mean of one or more model variables).
-We also refer to this as mean-based approach.
+In particular, variables with more than two categories (for instance religion in our previous example), or variables modeled with a spline or polynomial term, will correspond to more than one model variables.
+\item \textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more model variables).
+Also referred to as center point.
+We refer to the prediction approach which uses model center as mean-based.
 The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable -- the case in Fig~\ref{fig:justify_plots}.
-In simple linear models (without) interaction, the input variable can be used as a proxy.
-
-\subsection{Reference point}
-
-Corresponds to the value or values chosen for non-focal predictors, when estimating the predictions.
+In simple linear models (without) interaction, the mean of the input variable can be used as a proxy for center point.
+\item \textbf{Reference point:} Corresponds to the value or values chosen for non-focal predictors, when estimating the predictions.
 Typically the center point (mean-based), but can instead be a whole-sample of quantiles or observations.
-Appropriately choosing the reference point is very crucial in generating ``correct'' predictions especially when dealing with non-linear link functions.
-
-\subsection{Anchor}
-
-The value chosen for the focal predictor when estimating effect confidence intervals (anchor choice does not affect the main prediction estimates).
+Appropriately choosing the reference point is very crucial in generating correct predictions especially when dealing with non-linear link functions.
+\item \textbf{Anchor:} The value chosen for the focal predictor when estimating effect confidence intervals (anchor choice does not affect the main prediction estimates).
 Typically chosen as the center point of the model variables corresponding to the focal input variable but other sensible values can be chosen too.
-
 If we are trying to focus on effects, the anchor from which we calculate CIs becomes important.
@@
 In Fig~\ref{fig:justify_anchors}, we compare effects plots anchored at the minimum and center (mean) of the focal predictor.
+\end{itemize}
 

### 2022-06-14 manuscript/body.tex
Intro changed

@@
 
-Ordinary least squares (OLS) regression provides an easy way to express multivariate relationships as an additive linear combination of many predictors with the outcome.
-The coefficient estimates from OLS regression models are generally easy to interpret as the expected change in the outcome with a unit change in the continuous predictor or as the expected outcome difference between the levels of the categorical predictor with the baseline category.
-However, with additional complexities such as interactions between predictors, the resulting coefficients risk misinterpretation or interpretation complexity \citep{brambor_understanding_2006, berry_improving_2012}.
+Ordinary least squares (OLS) regression provides an easy way to express multivariate relationships as an additive linear combination of many input variables with the outcome.
+The coefficient estimates from OLS regression models are generally easy to interpret as the expected change in the outcome with a unit change in the continuous variable or as the expected outcome difference between the levels of the categorical variable with the baseline category.
+However, with additional complexities such as interactions between input variables, the resulting coefficients risk misinterpretation or interpretation complexity \citep{brambor_understanding_2006, berry_improving_2012}.
 
-Simple OLS regression models are flexible and can easily be generalized, as generalized linear (mixed) models (GL(M)Ms), to examine more complex relationships, including nonlinear relationships between response and predictors e.g., nonlinear transformation on the predictors, interactions between predictors (and via splines for example), and nonlinear transformations via link functions.
+Simple OLS regression models are flexible and can easily be generalized, as generalized linear (mixed) models (GL(M)Ms), to examine more complex relationships, including nonlinear relationships between response and input variables e.g., interactions between and nonlinear transformation on the input variables, (and via splines for example), and nonlinear transformations via link functions.
 This flexibility comes with additional complications, for example, complex multivariate models may risk misinterpretation, and miscalculation of quantities of interest \citep{berry_improving_2012, leeper2017interpreting}.
@@
 
-Both predictions and effects tell us what we expect the response to be for a given value of \emph{focal} predictor (predictor of interest).
+Both predictions and effects tell us what we expect the response to be for a given value of \emph{focal} input variable (scientific variable of interest).
 The major difference between the two lies in the construction of the associated confidence intervals.
 In particular, for predictions, we want to use the classic curved confidence intervals (or prediction intervals) which capture the uncertainty in the whole model and generally focus on the \emph{total} effects, and so will often be suited for a univariate model.
-On the other hand, effects, attempt to show the effect of a focal predictor.
-In this case, we suggest the narrower confidence intervals which focus on uncertainty in the effect of the focal predictor only, and so will often be suited for the direct effects – after controlling for other predictors – from a multivariate model.
+On the other hand, effects, attempt to show the effect of a focal input variable.
+In this case, we suggest the narrower confidence intervals which focus on uncertainty in the effect of the focal input variable only, and so will often be suited for the direct effects – after controlling for other non-focal input variables – from a multivariate model.
 
-When visually presented, predictions and effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind of transformations on the predictors whose estimates are usually, but not always, harder to interpret.
+When visually presented, predictions and effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind of transformations on the input variables whose estimates are usually, but not always, harder to interpret.
 Generating these quantities together with the associated confidence intervals for regression models has a number of challenges.
@@
 \begin{enumerate}
-\item choice of representative values of focal predictor and the \emph{reference point} for non-focal predictors, especially in multivariate models
-\item  uncertainty estimation -- appropriate choice of \emph{anchor} for computing confidence intervals; and how to incorporate the uncertainty due to non-focal predictors 
+\item choice of representative values of focal input variable and the \emph{reference point} for non-focal input variables, especially in multivariate models
+\item  uncertainty estimation -- appropriate choice of \emph{anchor} for computing confidence intervals; and how to incorporate the uncertainty due to non-focal input variables 
 \item bias in the expected mean prediction induced by the nonlinear transformation of the response variable (especially in GL(M)Ms)
@@
 
-The most common way of choosing the representative values is taking unique levels of the focal predictor if discrete or taking appropriately sized quantiles if continuous, and then calculating the predictions while holding non-focal (other predictors other than the focal predictor) predictors at their reference point (e.g., means) \citep{fox2009effect, hanmer2013behind}.
+The most common way of choosing the representative values is taking unique levels of the focal input variable if discrete or taking appropriately sized quantiles if continuous, and then calculating the predictions while holding non-focal (other input variables other than the focal input variable) input variables at their reference point (values we chose for the non-focal input variables e.g., means) \citep{fox2009effect, hanmer2013behind}.
 This procedure generates -- predictions (at least in our case), \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}.
-In this article, we refer to these quantities as predictions and effects since they should, for example, tell us what we would expect the response to be at a particular value or level of the predictor, for an ``average case''.
-More specifically, the basic idea is to compute the expected outcome by holding the non-focal predictors constant, for example at their mean (centered), while varying the focal predictor, with the goal that estimates represent how the model responds to the changes in the focal predictor.
+In this article, we refer to these quantities as predictions and effects since they should, for example, tell us what we would expect the response to be at a particular value or level of the input variable, for an ``average case''.
+More specifically, the basic idea is to compute the expected outcome by holding the non-focal input variables constant, for example at their mean, while varying the focal input variable, with the goal that estimates represent how the model responds to the changes in the focal input variable.
 As mentioned above, in our case, the key difference between predictions and effects is in the computation of the confidence intervals.
 
-The most commonly used \proglang{R} software packages for generating predictions (\pkg{emmeans} and \pkg{effects}), by default, use the average of \emph{input} and \emph{model variables} as the reference point for continuous and categorical predictors, respectively.
-However, there are a number of choices one can make when considering this approach -- for example, in the presence of interaction, averaging the interactions (averaging the product of interacting model variables) versus taking the product of the averages of the interacting model variables (the default for \pkg{emmeans} and \pkg{effects}).
+The most commonly used \proglang{R} software packages for generating predictions (\pkg{emmeans} and \pkg{effects}), by default, use the average of \emph{input} and \emph{predictor variables} as the reference point for continuous and categorical input variables, respectively.
+However, there are a number of choices one can make when considering this approach -- for example, in the presence of interaction, averaging the interactions (averaging the product of interacting predictor variables) versus taking the product of the averages of the interacting predictor variables (the default for \pkg{emmeans} and \pkg{effects}).
 For simple models (with no interactions), the two approaches give the same predictions.
-However, for complex models, we demonstrate that the two approaches can produce substantially different results and show that averaging the interactions (model variables) closely matches the observed values.
-To illustrate this, consider models~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher} below, with $x_1$ as the focal predictor:
+However, for complex models, we demonstrate that the two approaches can produce substantially different results and show that averaging the interactions closely matches the observed values.
+To illustrate this, consider models~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher} below, with $x_1$ as the focal input variable:
 %
@@
 \caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} predictions for $x_1$ on $y$ for models with and without interactions.}
-The horizontal blue, green and black lines are the mean predictions, i.e., $\bar{\hat{y}}$, the red ones are the means of the simulated $y$, i.e., $\bar{y}$, while the vertical dotted grey lines are the means of the corresponding focal predictors.
+The horizontal blue, green and black lines are the mean predictions, i.e., $\bar{\hat{y}}$, the red ones are the means of the simulated $y$, i.e., $\bar{y}$, while the vertical dotted grey lines are the means of the corresponding focal input variable.
 The grey points are the binned simulated $y$.
-The trend lines represent the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other predictors at their average for the three methods and the true predictions based on the simulation parameters.
+The trend lines represent the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other input variables at their average for the three methods and the true predictions based on the simulation parameters.
 A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the simulated value in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$.
 Similarly, the predictions based on the three approaches closely match the true predictions.
-B: Even with the simple interaction between non-focal predictors, we start seeing deviation in both predicted mean, $\bar{\hat{y}}$, and overall predictions from the simulated, $\bar{y}$, and true predictions, respectively; in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not the proposed \pkg{varpred}.}
+B: Even with the simple interaction between non-focal input variables, we start seeing deviation in both predicted mean, $\bar{\hat{y}}$, and overall predictions from the simulated, $\bar{y}$, and true predictions, respectively; in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not the proposed \pkg{varpred}.}
 \label{fig:justify_plots}
@@
 In particular, \pkg{emmeans} and \pkg{effects} give similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is very close to the simulated average ($\bar{y}$), Fig~\ref{fig:justify_plots}B.
-To generate Fig~\ref{fig:justify_plots}A, all the three packages average $x_2$ and $x_3$.
+To generate Fig~\ref{fig:justify_plots}A, all the three packages average $x_2$ and $x_3$, i.e., the input variables.
 On other hand, to generate Fig~\ref{fig:justify_plots}B, the difference in the estimates lies on how each of the packages average the interaction term ($x_2x_3$).
-In particular \pkg{emmeans} and \pkg{effects} use \emph{input variables} to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses \emph{model variables} to compute $\over{x_2x_3}$.
-In the case of Fig~\ref{fig:justify_plots}A, averaging the input variables is equivalent to averaging the model variables.
+In particular \pkg{emmeans} and \pkg{effects} use input variables to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses predictor variables to compute $\over{x_2x_3}$.
+In the case of Fig~\ref{fig:justify_plots}A, averaging the input variables is equivalent to averaging the predictor variables.
 
-One may be interested in the uncertainties associated with the focal predictor of interest only, excluding other uncertainties due to other non-focal predictors -- effects.
+One may be interested in the uncertainties associated with the focal input variable of interest only, excluding other uncertainties due to other non-focal input variables -- effects.
 However, currently, the two packages do not provide straightforward way do achieve this.
@@
 \caption{{\bf A comparison of conventional $95\%$ confidence bands and effects.} The description of horizontal, vertical and trend lines remain the same as above.
-The wider dotted blue curves overlaying the green curves corresponds to the conventional confidence bands around the predcitions from \pkg{emmeans} and \pkg{effects}, while the narrower black curves crossing at the mean of the focal predictor, i.e., the model center, corresponds to the effects of the focal predictor from \pkg{varpred}.
+The wider dotted blue curves overlaying the green curves corresponds to the conventional confidence bands around the predcitions from \pkg{emmeans} and \pkg{effects}, while the narrower black curves crossing at the mean of the focal input variable, i.e., the model center, corresponds to the effects of the focal input variable from \pkg{varpred}.
 For simple models, the isolated confidence bands crosses at the model center.
-The wider curves incorporates not only the uncertainties due to intercept term but also other non-focal predictors.
-On the other hand, effects (narrower crossing lines) only take into account the main effect uncertainty of the focal predictor.}
+The wider curves incorporates not only the uncertainties due to intercept term but also other non-focal input variables.
+On the other hand, effects (narrower crossing lines) only take into account the main effect uncertainty of the focal input variable.}
 \label{fig:justify_ci_plots}
@@
 
-For \pkg{emmeans} and \pkg{effects}, the confidence bands are much wider because they include uncertainties associated with the intercept and non-focal predictors, but narrower and crosses at the mean of the focal predictor, i.e., \emph{center point or mean-based}, in \pkg{varpred}.
-In other words, with \pkg{varpred}, we are able to generate effects indicating zero uncertainty at the value of the focal predictor we are more certain about, i.e., mean of the model variable corresponding to the focal predictor (center point or mean-based).
-For simple models, the point where the confidence bands crosses is the model center; it corresponds to anchor we choose, in this case, the mean of the focal predictor.
+For \pkg{emmeans} and \pkg{effects}, the confidence bands are much wider because they include uncertainties associated with the intercept and non-focal input variables, but narrower and crosses at the mean of the focal input variable in \pkg{varpred}.
+We refer to this point as the \emph{center point} and the approach as \emph{mean-based}.
+In other words, with \pkg{varpred}, we are able to generate effects indicating zero uncertainty at the value of the focal input variable we are more certain about, i.e., mean of the predictor variable corresponding to the focal input variable.
+For simple models, the point where the confidence bands crosses is the model center; it corresponds to anchor we choose, in this case, the mean of the focal input variable, but we generally choose the mean of the focal predictor (predictors corresponding to the focal input variable).
 
 For linear models (including models with complex interactions), the averaging is done on the linear scale, i.e., linear averaging.
-As a result the bias in the expected (mean) predictions can easily be corrected by performing predictions from the \emph{model center} (average of the model variables) or the center point.
+As a result the bias in the expected (mean) predictions can easily be corrected by performing predictions from the \emph{model center} (average of the predictor variables).
 The mean prediction is the prediction at the model center.
 On the other hand, when dealing with GL(M)Ms with nonlinear link functions, the averaging is done on the nonlinear scale, i.e., nonlinear averaging which makes estimating correct predictions much harder.
-In particular, bias in expected mean prediction induced by nonlinear transformation of the response variable can lead inaccurate predictions.
+In particular, bias in expected mean prediction induced by nonlinear transformation of the response variable can lead to inaccurate predictions.
 One way to deal with this is to make predictions on the transformed scale (linear predictor scale), and then back-transform to the original scale.
 However, the back-transformation may either result in biased predictions or requires some approximation, for example, second order Taylor approximation implemented in \pkg{emmeans} \citep{lenth2018package}.
-In linear models, mean-based (alias center point or model center) gives estimates consistent with the observed values, however, in models with nonlinear link functions, this is not always the case.
-An alternative to mean-based reference point, i.e., averaging non-focal predictors, is the \emph{whole-sample-based} approach, discussed in detail later, which involves computing the prediction over the population of non-focal predictors and then averaging across the values of the focal predictor \citep{hanmer2013behind}.
+In linear models, mean-based gives estimates consistent with the observed values, however, in models with nonlinear link functions, this is not always the case.
+An alternative to mean-based reference point, i.e., averaging non-focal predictor variables, is the \emph{whole-sample-based} approach, discussed in detail later, which involves computing the prediction over the population of non-focal predictor variables and then averaging across the values of the focal predictor variable \citep{hanmer2013behind}.
+
+Predictions are the standard curves; they tell us what we expect to see if we know the value of the focal predictor.
+The lines inside (effects) are a little different: they are a good way to visualize the effect of the focal predictor.
+Unlike predictions, where the uncertainty captures the uncertainty in the whole model, the effects plot focuses on uncertainty in the effect of the focal predictor only and depend on the anchor.
 
@@
 \begin{itemize}
-\item \textbf{Predictors}: Refer to the X-variables which go into the regression model.
-For example, consider the model the includes the interaction in Equation~\ref{eq:simple_inter_higher}.
-This regression model has $5$ predictors -- $x_1, x_2, x_3, x_2x_3$ and the intercept term.
-\item \textbf{Focal and non-focal predictors}: Focal predictor refers to the predictor we are interested in generating its effects on the predictions.
-For multivariate models, all other remaining predictors are non-focal predictors.
+\item \textbf{Input variables}: Refers to the observed (or scientific) variables underlying an inference or exploration.
+For example, the regression models described by Equation~\ref{eq:simple_inter_higher_no_interaction} and Equation~\ref{eq:simple_inter_higher} both have $3$ input variables -- $x_1, x_2, x_3$.
+\item \textbf{Focal and non-focal input variables}: Focal input variable refers to the input variable we are interested in.
+For multivariate models, all other remaining input variables are non-focal input variables.
 For example, consider model described by Equation~\ref{eq:simple_inter_higher_no_interaction}.
-Suppose we are interested in the effect of $x_1$ on the predicted $y$, then $x_1$ is the focal predictor while $x_2$ and $x_3$ are non-focal predictors.
-\item \textbf{Input variables}: Refers to the observed (or scientific) variables underlying an inference or exploration, and forms the X-variables.
-For example, the regression model described by Equation~\ref{eq:simple_inter_higher} has only $4$ input variables -- $x_1, x_2, x_3$.
-\item \textbf{Model matrix}: Refers to the design matrix whose rows include all combination of variables appearing in the interaction terms.
+Suppose we are interested in the effect of $x_1$ on the predicted $y$, then $x_1$ is the focal input variable while $x_2$ and $x_3$ are non-focal non-focal variables.
+\item \textbf{Model matrix}: Refers to the design matrix whose rows include all combination of input variables appearing in the interaction terms.
 Consider an example for three hypothetical households -- the first household head is a Christian with an income of $\$ 50$, the second household head is Muslim with an income of $\$ 100$ while the third household head is a Jew with an income of $\$ 77$.
@@
 \end{bmatrix}.$$ The first column represents the constant term in our model, $\beta_0$.
-For continuous variables, the representation in the model matrix is the same as the correspond input variables (for example second column representing income).
+For continuous input variables, the representation in the model matrix is the same as the corresponding input variables (for example second column representing income).
 For categorical variables, however, by default, the model matrix  creates additional dummy variables using the reference cell parameterization.
@@
 In our example, the column \code{religion} is missing and instead we have \code{religionJew} and \code{religionMuslim}; the missing category \code{religionChristian} is treated as the reference category.
-\item \textbf{Model variables:} It is easy to interchangeably refer to predictors as input variables, and vice versa.
-To avoid this issue, we adopt the term model variables which are variables that represent columns in the model matrix.
-Each input variable may correspond to one or more model variables.
-In particular, variables with more than two categories (for instance religion in our previous example), or variables modeled with a spline or polynomial term, will correspond to more than one model variables.
-\item \textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more model variables).
+\item \textbf{Predictor variables (or simply predictors)}: Refer to the X-variables which go into the regression model; and represent columns in the model matrix.
+Each input variable may correspond to one or more predictor variables.
+In particular, variables with more than two categories (for instance religion in our previous example), or input variables modeled with interactions, a spline or polynomial interactions, will correspond to more than one predictor variables.
+For example, the regression models described by Equation~\ref{eq:simple_inter_higher_no_interaction} and Equation~\ref{eq:simple_inter_higher} have $4$ predictors -- $x_1, x_2, x_3$ plus intercept term and $5$ predictors -- $x_1, x_2, x_3, x_2x_3$ plus the intercept term, respectively.
+\item \textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more predictor variables).
 Also referred to as center point.
-We refer to the prediction approach which uses model center as mean-based.
-The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable -- the case in Fig~\ref{fig:justify_plots}.
+We refer to the prediction approach that uses model center as mean-based.
+The center point for a set of predictor variables corresponding to an input variable may not represent a possible value of the input variable -- the case in Fig~\ref{fig:justify_plots} (with interaction).
 In simple linear models (without) interaction, the mean of the input variable can be used as a proxy for center point.
-\item \textbf{Reference point:} Corresponds to the value or values chosen for non-focal predictors, when estimating the predictions.
+If we have more than one predictor variables associated with the focal input variable, however, the model center does not correspond to any given value of the variable, and we don't get the nice crossing we get in simpler cases.
+See Fig~\ref{fig:pred_cubic_plots}.
+\item \textbf{Reference point:} Corresponds to the value or values chosen for non-focal input variables, when estimating the predictions.
 Typically the center point (mean-based), but can instead be a whole-sample of quantiles or observations.
 Appropriately choosing the reference point is very crucial in generating correct predictions especially when dealing with non-linear link functions.
-\item \textbf{Anchor:} The value chosen for the focal predictor when estimating effect confidence intervals (anchor choice does not affect the main prediction estimates).
-Typically chosen as the center point of the model variables corresponding to the focal input variable but other sensible values can be chosen too.
+\item \textbf{Anchor:} The value chosen for the focal input variable when estimating effect confidence intervals.
+The anchor choice does not affect the main prediction estimates.
+Typically chosen as the center point of the predictor variables corresponding to the focal input variable but other sensible values can be chosen too.
 If we are trying to focus on effects, the anchor from which we calculate CIs becomes important.
 This is not true for the more conventional prediction plots: we can choose any anchor, e.g., zero-anchored, but model center seems like a natural, stable choice, i.e., center-anchored.
-In Fig~\ref{fig:justify_anchors}, we compare effects plots anchored at the minimum and center (mean) of the focal predictor.
+In Fig~\ref{fig:justify_anchors}, we compare effects plots anchored at the minimum and center (mean) of the focal input variable.
 \end{itemize}
@@
 \end{figure}
-
-If we have more than one model variables for our focal predictor, however, the model center does not correspond to any given value of the variable, and we don't get the nice crossing we get in simpler cases.
-See Fig~\ref{fig:pred_cubic_plots}.
-
-\subsection{Prediction- vs. effect- styled plots}
-
-A prediction-style plot is about predicting observations (or the ensemble mean of observations) for a given value of the focal predictor.
-For this, we want to use the classic curved confidence intervals (or prediction intervals).
-A prediction-style plot will generally focus on \emph{total} effects, and so will often be suited for a univariate fit.
-An effect-style plot, however, is an attempt to visualize the effect of a focal predictor.
-In this case, we suggest the narrower confidence intervals.
-Here, it makes equal sense to think about the total effect of our focal predictor, from a univariate model, or the direct effects – after controlling for whatever – from a multivariate model.
-See Fig~\ref{fig:justify_ci_plots}.
-
-\subsection{Predictions vs. effects}
-
-We would like to summarize by talking about predictions vs.
-effects.
-Predictions are the standard curves; they tell us what we expect to see if we know the value of the focal predictor.
-The lines inside (effects) are a little different: they are a good way to visualize the effect of the focal predictor.
-Unlike predictions, where the uncertainty captures the uncertainty in the whole model, the effects plot focuses on uncertainty in the effect of the focal predictor only and depend on the anchor.
 

### 2022-06-15 manuscript/body.tex
Autocommit

@@
 For example, the regression models described by Equation~\ref{eq:simple_inter_higher_no_interaction} and Equation~\ref{eq:simple_inter_higher} have $4$ predictors -- $x_1, x_2, x_3$ plus intercept term and $5$ predictors -- $x_1, x_2, x_3, x_2x_3$ plus the intercept term, respectively.
+\item \textbf{Focal predictors:} All columns of the model matrix associated with the focal input variable.
+For example, all the terms of a cubic polynomial focal variable or columns corresponding to a categorical focal variable with more than two levels.
+\item \textbf{Non-focal predictors:} All columns of the model matrix associated with the non-focal input variable(s).
+For example, in the previous household size example, all the columns associated with religion are non-focal predictors.
 \item \textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more predictor variables).
@@
 If we are trying to focus on effects, the anchor from which we calculate CIs becomes important.
-This is not true for the more conventional prediction plots: we can choose any anchor, e.g., zero-anchored, but model center seems like a natural, stable choice, i.e., center-anchored.
+This is not true for the more conventional prediction CIs.
+We can choose any anchor, e.g., zero-anchored (the value of the anchor is $0$), but center point seems like a natural, stable choice, we call this center-anchored.
 In Fig~\ref{fig:justify_anchors}, we compare effects plots anchored at the minimum and center (mean) of the focal input variable.
@@
 
-To get an intuition of how conditioning on the mean values of non-focal predictors work, suppose we are interested in predictions at the levels of a particular predictor, i.e., focal, $x_f$, from the set of predictors.
+To illustrate mean-based approach, suppose we are interested in predictions at the levels of a particular focal input variable, $x_f$, from a multivariate model.
 As a starting point, assume that the model has no interaction terms.
-The idea is to choose a reference point for the values of non-focal predictors.
-For example, fixing the values of non-focal predictor(s) at some typical values -- typically determined by averaging (for now) in some meaningful way, for example, arithmetic mean  for continuous and average over the levels of the factors for categorical non-focal predictors.
+The idea is to choose a reference point for the values of non-focal input variables.
+For example, fixing the values of non-focal predictor(s) at some typical values -- typically determined by averaging in some meaningful way, for example, arithmetic mean  for continuous and average over the levels of the factors for categorical non-focal predictors.
 We refer to this as centered or mean-based approach.
@@
 \begin{align}\label{eq:eta_mean}
-\eta^\star(x_f, \nset{{\bar{x}^\star}}) &= \hat{\beta}_f x_f + \sum \nset{\hat{\beta}} \nset{{\bar{x}^\star}} \\
-\hat{y}_f  &= g^{-1} \left(\eta^\star(x_f, \nset{{\bar{x}^\star}})\right)
+\eta^\star(x^\star_f, \nset{{\bar{x}^\star}}) &= \hat{\beta}_f x^\star_f + \sum \nset{\hat{\beta}} \nset{{\bar{x}^\star}} \\
+\hat{y}_f  &= g^{-1} \left(\eta^\star(x^\star_f, \nset{{\bar{x}^\star}})\right)
 \end{align}
-where $\nset{{\bar{x}^\star}}$ and $x_f$ are columns of $\bX^\star$ corresponding to the non-focal predictors and the focal predictor, respectively.
+where $\nset{{\bar{x}^\star}}$ and $x^\star_f$ are columns of $\bX^\star$ corresponding to the non-focal predictors and the focal predictor, respectively.
 
@@
 
-Higher order terms such as interactions, splines, polynomials, etc., can be within the focal predictor, within (between) non-focal predictor(s) or between the focal and non-focal predictor(s).
+Higher order terms such as interactions, splines, polynomials, etc., can be within the focal input variable, within (between) non-focal input variable(s) or between the focal and non-focal input variable(s).
 To distinguish the three, suppose the model which describes the hypothetical simulated household size simulation based on a number of socio-demographic factors such as age and wealthindex is
@@
 
-In the first case, with \code{Age} as the focal predictor, with cubic polynomial interaction terms (and 3 model variables associated with it).
-In this case, each of the polynomial terms (model variables) of the focal predictor are evaluated independently across the chosen levels of the focal predictor, $Age_i$.
-Specifically, the higher order terms are treated as additional columns of the model matrix evaluated with the same values of the focal predictor, while non-focal predictors are fixed at their reference point as discussed in the previous section.
+In the first case, with \code{Age} as the focal input variable, with cubic polynomial interaction ($3$ focal predictors associated with it).
+In this case, each of the focal predictors ($\mathrm{Age}_i, \mathrm{Age}^2_i$ and $\mathrm{Age}^3_i$) are evaluated independently across the chosen levels of the focal input variable, $Age_i$.
+Specifically, the higher order terms are treated as additional columns of the model matrix evaluated with the same values of the focal input variable, while non-focal predictors are fixed at their reference point as discussed in the previous section.
 In this case, the predictions associated with \code{Age} on the linear predictor scale become
@@
 
-In the second case, with $Wealthindex$ as the focal predictor, the higher order terms in non-focal predictor, \code{Age}, are treated simply as additional columns in the variable space (model variables) and an appropriate choice of reference point applies just like in the models without higher order interactions.
-For instance, in mean-based approach, we average all non-focal model variables.
+In the second case, with \code{Wealthindex} as the focal input variable, the non-focal variable, \code{Age}, is a cubic polynomial.
+In this case, the non-focal predictors ($\mathrm{Age}_i, \mathrm{Age}^2_i$ and $\mathrm{Age}^3_i$) are simply treated as additional columns in the predictor space and an appropriate choice of reference point applies just like in the models without higher non-focal variable interactions.
+For instance, in mean-based approach, we average all non-focal predictors.
 Thus
@@
 
-Lastly, consider model shown in Equation~\ref{eq:simple_inter_higher}, previously used to demonstrate how to handle interactions in non-focal predictors.
-Here, we use the same model to demonstrate how to handle interactions between the focal predictor, $x_2$, and non-focal predictor, $x_3$.
+Lastly, consider model described by Equation~\ref{eq:simple_inter_higher}, previously used to demonstrate how to handle interactions in non-focal input variables.
+Here, we use the same model to demonstrate how to handle interactions between the focal input variable, $x_2$, and non-focal input variable, $x_3$.
 In this case, the non-interacting predictors are treated as non-focal predictors, with appropriate reference points.
-However, for the interacting predictors, we first choose representative values for the focal predictor and some particular values of the interacting predictor similar to the way we choose values for the focal predictor or predetermined.
-In our example, suppose we pick $i$ and $j$ unique values of the focal predictor, $x_2$ and interacting predictor $x_3$, respectively.
+However, for the interacting predictors, we first choose representative values for the focal input variable and some particular values of the interacting non-focal input variable in a similar way we choose values for the focal input variable or predetermined.
+In our example, suppose we pick $i$ and $j$ unique values of the focal input variable, $x_2$ and interacting input variable $x_3$, respectively.
 The prediction on linear predictor scale is given by
@@
 \begin{align*}
-\eta^\star(x_{2i}, x_{3j}, \nset{{\bar{x}^\star}}) = \beta_0 + \beta_1 \bar{x}_1 + \beta_2x_{2i} + \beta_3x_{3j} + \beta_{23}x_{2i}x_{3j}.
+\eta^\star(x_{2i}, x_{3j}, {\bar{x}^\star_1}) = \beta_0 + \beta_1 \bar{x}^\star_1 + \beta_2x_{2i} + \beta_3x_{3j} + \beta_{23}x_{2i}x_{3j}.
 \end{align*}
 
-In general, our formulation, even for more complicated interactions, follow these three basic principles -- interaction within focal predictor, interaction between non-focal predictors and interaction between focal and non-focal predictors.
+In general, our formulation, even for more complicated interactions, follow these three basic principles -- interaction within focal input variables, interaction between non-focal input variables and interaction between focal and non-focal input variables.
 
@@
 
-We describe the uncertainty around the estimates using confidence intervals (CI).
-In principle, every value of focal predictor has a different CI.
-The conventional way to compute variances for predictions is $\boldsymbol\sigma^2 = \textrm{Diag}(\bX^\star \boldsymbol{\Sigma} \bX^{\star\top})$ \citep{lenth2018package, fox2009effect}, so that the confidence intervals are $\boldeta \pm q\boldsymbol\sigma$, where $q$ is an appropriate quantile of Normal or t distribution.
-This generates prediction CIs and incorporates all the uncertainties -- including the uncertainty due to the intercept and non-focal predictors.
-But what if we are only interested in the uncertainty as a result of the focal predictor, so that the confidence intervals are $\boldeta \pm q \boldsymbol \sigma_f$, i.e., effects?
+We describe the uncertainty around the estimates using confidence intervals (CIs).
+In principle, every value of focal input variable has a different CI.
+The conventional way to compute variances for predictions is
+%
+\begin{align}\label{eq:conventional_variance}
+\boldsymbol\sigma^2 = \textrm{Diag}(\bX^\star \boldsymbol{\Sigma} \bX^{\star\top}), 
+\end{align}
+so that the confidence intervals are $\boldeta \pm q\boldsymbol\sigma$, where $q$ is an appropriate quantile of Normal or t distribution \citep{lenth2018package, fox2009effect}.
+This generates conventional CIs which incorporate all the uncertainties -- including the uncertainties due to the intercept and non-focal input variables.
+But what if we are interested in the uncertainty as a result of the focal input variable only, so that the CIs are $\boldeta \pm q \boldsymbol \sigma_f$, i.e., effects?
 
-Commonly used \proglang{R} packages for constructing predictions do not exclude the uncertainties resulting from non focal predictors when computing the CIs.
-Currently a non-trivial way to exclude uncertainties associated with non-focal predictors in some of these packages is to provide a user defined variance-covariance matrix with the covariances of non-focal terms set to $0$.
-We refer to this procedure as \emph{zeroing-out} variance-covariance matrix; and only works when the input predictors are \emph{centered} prior to model fitting, in case of numerical predictors, and more complicated when the predictors are categorical.
-We describe our proposed method which is based on the model center and does not require input predictors to be centered prior to model fitting.
+Commonly used \proglang{R} packages for constructing predictions do not exclude the uncertainties resulting from non focal input variables when computing the CIs.
+Currently obscure way to exclude uncertainties associated with non-focal input variables in some of these packages is to provide a user defined variance-covariance matrix with the covariances of non-focal predictors set to $0$.
+We refer to this procedure as \emph{zeroing-out} variance-covariance matrix; and only works when the input variables are \emph{centered} prior to model fitting, in case of numerical variables, and more complicated when the input variables are categorical.
+An alternative is our proposed approach which uses the model center and does not require the input variables to be centered prior to model fitting.
 
-\subsection{Model center based}
-
-Let $\bxo$ be a model matrix containing appropriately chosen values of focal predictor and centered non-focal model variables, and let $\ba$ be the anchor matrix, with same dimension and entries in all non-focal model variables as $\bxo$.
-Let $\baf$ be the column of $\ba$ corresponding to focal predictor model variable(s) defined in $\bxo$.
-Any appropriate values can be chosen for $\baf$ but for model center (center-anchored), we use $\bafc$ which is the mean of the focal predictor model variable(s).
-Thus the variances for the predictions we defined in the previous section is
-
+Let $\bxo$ be a centered model matrix previously defined, and let $\ba$ be an anchor matrix, with the same dimensions and entries in all non-focal predictors as $\bxo$.
+Let $\baf$ be the column of $\ba$ corresponding to focal predictor(s) defined in $\bxo$.
+Any appropriate values can be chosen for $\baf$ but for model center (center-anchored), we use $\bafc$ which is the mean of the focal predictor(s).
+Thus
 %
 \begin{align}\label{eq:centered_variance}
-\boldsymbol\sigma_f^2 = \textrm{Diag}((\bxo - \bafc) \boldsymbol{\Sigma} (\bxo - \bafc)^\top)
+\boldsymbol\sigma_f^2 = \textrm{Diag}((\bxo - \bafc) \boldsymbol{\Sigma} (\bxo - \bafc)^\top).
 \end{align}
@@
 We can see that $\forall ~\bxof=\baf$, $\bxo - \baf = \boldsymbol 0$, hence $\boldsymbol\sigma_f^2 = \boldsymbol{0}$.
-Similarly, for all values of $\bxof$ close to anchor point $\bafc$ the term $(\bxo - \bafc)$ and $\boldsymbol\sigma_f^2$ goes to $\boldsymbol 0$ and $\boldsymbol\sigma_f^2 = \boldsymbol 0$ if $\bxof=\bafc$.
-This means that $\boldsymbol\sigma_f^2$ close to the anchor point are smaller than those away from the anchor point; and results to confidence intervals which are narrower (isolated) around the anchor or crosses at the anchor point for simple models.
+Similarly, for all values of $\bxof$ close to anchor point $\bafc$, the term $(\bxo - \bafc)$ and $\boldsymbol\sigma_f^2$ goes to $\boldsymbol 0$ and $\boldsymbol\sigma_f^2 = \boldsymbol 0$ if $\bxof=\bafc$.
+This means that $\boldsymbol\sigma_f^2$ close to the anchor point are smaller than those away from the anchor point; and results to confidence intervals which are narrower around the anchor or crosses at the anchor point for simple models.
+This is the effect of the particular focal input variable.
+By setting $\baf = \boldsymbol 0$, we get the variances for the conventional CIs, Equation~\ref{eq:conventional_variance}.
 
-The computation of $\bxo - \bafc$ impacts only on the intercepts and non-focal model variables, i.e., the slopes and variance corresponding to the focal predictors are not affected.
+The computation of $\bxo - \bafc$ impacts only on the intercepts and non-focal predictors, i.e., the slopes and variance corresponding to the focal predictors are not affected.
 This means that we can still generate effects without necessarily centering the predictors prior to model fitting.

### 2022-06-15 manuscript/body.tex
Autocommit

@@
 One way to deal with this is to make predictions on the transformed scale (linear predictor scale), and then back-transform to the original scale.
-However, the back-transformation may either result in biased predictions or requires some approximation, for example, second order Taylor approximation implemented in \pkg{emmeans} \citep{lenth2018package}.
+However, the back-transformation may either result in biased predictions or requires some approximation, for example, second-order Taylor approximation implemented in \pkg{emmeans} \citep{lenth2018package}.
 In linear models, mean-based gives estimates consistent with the observed values, however, in models with nonlinear link functions, this is not always the case.
@@
 
-\begin{align}\label{sim:lm_cubic}
+\begin{align}\label{eq:lm_cubic}
 \mathrm{hh~size}_i &= \beta_0 + \beta_{\mathrm{A_1}}\mathrm{Age}_i + \beta_{\mathrm{A_2}}\mathrm{Age}^2_i + \beta_{\mathrm{A_3}}\mathrm{Age}^3_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \epsilon_i \nonumber\\
@@
 The common approach for bias-adjustment is second-order Taylor approximation \citep{lenth2018package, duursma2003bias}.
-Another potential source of bias comes from additional non-focal predictors, especially in models with nonlinear link functions.
+Another potential source of bias comes from the non-focal input variables in a multivariate model, especially in models with nonlinear link functions.
 Here, we describe and implement a different approach; whole-sample-based approach for bias correction.
@@
 
-One way to predict is to condition on values of the focal predictor and make predictions for all observations (members of the population) \citep{hanmer2013behind}.
-The nonlinear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population to a one-dimensional distribution of $\eta^\star(x_f, \nset{x})$, which is a function of focal predictor and the observed values of non-focal predictors, as opposed to the definition in Equation~\ref{eq:eta_mean}.
+An alternative approach to choosing a reference point is to make predictions over all observations of the non-focal input variables (members of the population) \citep{hanmer2013behind}.
+The nonlinear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population to a one-dimensional distribution of $\eta^\star(x_f, \nset{x})$, which is a function of the chosen values of the focal predictor and the whole sample of non-focal predictors, as opposed to the definition in Equation~\ref{eq:eta_mean}.
 More specifically:
-
 \begin{itemize}
-\item compute linear predictor associated with non-focal predictors, $\nset{\eta} = \sum \nset{\beta} \nset{x}$
-\item compute linear predictor associated with the focal predictors, $\eta_{jf} = \sum{\beta_f x_{jf}}$
-\item for every value of the focal predictor, $\eta_{jf}$:
+\item compute linear predictor associated with whole sample of the non-focal predictors, $\nset{\eta} = \sum \nset{\beta} \nset{x}$
+\item compute linear predictor associated with the chosen values focal input variable, $\eta_{jf} = \sum{\beta_f x_{jf}}$
+\item for every value of the focal linear predictor, $\eta_{jf}$, compute
 
@@
 
-Once Equation~\ref{eq:pop_eta} is computed, one can back-transform the estimates to the original scale and the average over the levels of the focal predictors, $j$:
+Once Equation~\ref{eq:pop_eta} is computed, we back-transform the estimates to the original scale and average over the levels of the focal predictors, $j$:
 
@@
 
-We make similar adjustments to compute the variances of the predictions at every level of the focal predictor:
+We make similar adjustments to compute the variances of the predictions at every level of the focal input variable:
 
@@
 
-\bmb{does averaging the CI ...}\BC{Theoretical proof or just a statement?}
-
 For models with random effects components, we make further adjustment to correct for bias induced by the random effects terms.
-In the population approach, we treat the random effects terms as additional non-focal predictors and simply make adjustment to Equation~\ref{eq:pop_eta}.
+In whole-sample approach, we treat the random effects terms as additional non-focal predictors and simply make adjustment to Equation~\ref{eq:pop_eta}.
 In particular
@@
 
-As mentioned above, the two common choices to generate predictions are: 1) setting non-focal predictors to their mean -- mean-based; and 2) using the entire population of non-focal predictors -- whole-sample-based.
+As mentioned above, the two common choices to generate predictions are: 1) using mean of the non-focal predictors as the reference point -- mean-based; and 2) using the entire observed values of non-focal predictors -- whole-sample-based.
 If the link function is nonlinear and/or the model has complex higher order interactions, the average of the predictions evaluated at the mean of non-focal predictors and the average of predictions evaluated at the population level (and then averaged) of non-focal predictor are not equivalent, i.e.,
@@
 For accelerating functions, the Jensen’s inequality describes how the changes in input values elevate the predictions and otherwise describes how these changes depress the predictions.
-For a non-fully convex or concave link function, one can use Taylor series expansion to approximate the range over which the two are smaller or smaller than the other \citep{hanmer2013behind}.
-\bmb{Just knowing is not enough and discuss much earlier}\BC{Clarify with BMB the current version and appropriate section to move to}
+For a non-fully convex or concave link function, one can use Taylor series expansion to approximate the range over which the two are smaller or smaller than the other and then apply second-order approximation \citep{hanmer2013behind}.
+
+In whole-sample-based approach, the ensemble of predictions and CIs are back-transformed before averaging, see Equation~\ref{eq:pop_response}, so we do not need to worry about the nonlinear averaging.
+In other words, the averaging is no longer on the link scale and are likely to be bounded by the original data scale.
 
@@
 This may limit the inferences we can make about the entire population.
-In general, the mean-based approach provides the predictions of an average case, whereas the whole-sample-based approach, summarizes the predictions over the entire population -- and in some applications, the effect of an average case might not be generalizable to the entire population, especially, if the average does not represent the population.
-In particular the whole-sample-based approach focuses on specific observations since the prediction is first obtained for each observation and then averaged across the levels of the focal predictor.
+In general, the mean-based approach provides the predictions of an average case, whereas the whole-sample-based approach, summarizes the predictions over the entire population.
+In some applications, the effect of an average case might not be generalizable to the entire population, especially, if the average does not represent the population.
+This is might not be a problem in whole-sample-based approach since it focuses on specific observations because the prediction is first obtained for each observation and then averaged across the levels of the focal input variable.
 
@@
 For example, setting dummy categorical variables in the model matrix to their means, which, by default, sets them to their sample means or observed proportions.
-For example, a dummy variable for christian household heads may be set to $0.2$, translating to prediction for an household head who is $20\%$ christian.
-This may be problematic in models with complex interaction terms in which the values may not be converge to population mean\citep{hanmer2013behind}.
+For example, a dummy variable for christian household heads may be set to $0.2$, translating to prediction for an household head who is $20\%$ christian \citep{hanmer2013behind}.
 
@@
 
-We now illustrate the construction of prediction-styled and effect-styled plots and also demonstrate that the mean-based and whole-sample-based approaches produce different results in models with nonlinear link functions and/or additional source(s) of potential bias.
+We now illustrate the construction of prediction and effect plots and also demonstrate that the mean-based and whole-sample-based approaches produce different results in models with nonlinear link functions and/or additional source(s) of potential bias.
 In addition, we demonstrate that whole-sample-based approach can be used for bias correction.

### 2022-06-16 manuscript/body.tex
Autocommit

@@
 %
-\begin{figure}[h]
+\begin{figure}
 \begin{center}
@@
 
-\begin{figure}[h]
+\begin{figure}
 \begin{center}
@@
 
-\subsection{Predictions-styled and effects-styled plots}
+\subsection{Prediction and effect styled plots}
 
-As mentioned above, in some applications, one may be interested in uncertainties associated with the focal predictor only.
-In that case, it is important to exclude all the uncertainties as a result of non-focal predictors and only show confidence intervals describing the predictor of interest.
-To illustrate this, we consider the simulation model described in \nameref{S1_Appendix}.
+In multivariate models, we may want to visualize the effect of a particular focal input variable.
+In this case, effects styled plots are preferable over the conventional prediction plots which generally focus on the \emph{total} effects.
+To illustrate this, we simulate a multivariate model described in \nameref{S1_Appendix}.
+We also use the same model to illustrate how predictions and effects differ in the presence of higher order interactions in focal or non-focal input variables.
 
-We start by generating two predictions from the fitted model \nameref{S1_Appendix}: 1) one with cubic polynomial of \code{age} as a focal predictor; and 2) one with a simple focal predictor (non-polynomial) of \code{Wealthindex}.
-Then we compared non-anchored and mean-anchored (effects) confidence intervals around the predicted values (see Fig~\ref{fig:pred_cubic_plots}).
+We generated two sets of predictions and effects: 1) one with \code{age} as the focal input variable (a cubic polynomial polynomial focal predictor); and 2) one with \code{Wealthindex} as the focal in put variable (linear focal input variable).
+Then compared the predictions and effects as shown in Fig~\ref{fig:pred_cubic_plots}.
 
-\begin{figure}[h]
+\begin{figure}
 \begin{center}

### 2022-06-16 manuscript/cubic_predictors_preds.R

@@ -2,6 +2,7 @@ library(shellpipes)
 library(vareffects); varefftheme()
 library(ggpubr)
 library(ggplot2)
+library(ggthemes)
 library(dplyr)
 library(emmeans)
 
@@ -10,56 +11,63 @@ startGraphics()
 
 ## Focal polynomial
 
-### Non-isolated CIs 
+### Predictions CIs 
 pred_age_non_cubic <- varpred(mod_cubic
 	, "age"
+	, steps=500
 	, isolate=FALSE
-	, modelname="Non-isolated"
+	, modelname="Predictions"
 )
 
-### Isolated CIs
+### Effects CIs
 pred_age_isolated_cubic <- varpred(mod_cubic
 	, "age"
+	, steps=500
 	, isolate=TRUE
-	, modelname="Isolated"
+	, modelname="Effects"
 )
 pred_fit_mean <- getmeans(pred_age_isolated_cubic, what="estimate", modelname="Predicted mean")
 print(pred_fit_mean)
 
+## Colour limits
+col_limits <- c("Observed mean", "Predicted mean", "Effects", "Predictions")
+
 ### Combine
 vlist <- list(pred_age_non_cubic, pred_age_isolated_cubic)
 
-pred_age_cubic_plots <- (comparevarpred(vlist=vlist
+pred_age_cubic_df <- (comparevarpred(vlist=vlist
 		, lnames=NULL
-		, plotit=TRUE
+		, plotit=FALSE
 		, addmarginals=FALSE
 	)
-#	+ geom_hline(data=pred_fit_mean, aes(yintercept=fit, colour=model), lty=1)
-	+ geom_hline(data=true_prop_df, aes(yintercept=fit, colour=model), lty=2)
-	+ scale_colour_manual(breaks = c("Observed mean", "Predicted mean", "Isolated", "Non-isolated")
-		, values=c("Observed mean"="black", "Predicted mean"="red", "Isolated"="red", "Non-isolated"="blue")
-	)
-	+ scale_linetype_manual(values=c("Observed mean"=2, "Predicted mean"=1, "Isolated"=1, "Non-isolated"=3))
-	+ labs(title="A) Focal polynomial", y="Predicted household size", linetype="Method", colour="Method")
-	+ theme(legend.position="bottom")
+)
+pred_age_cubic_plots <- (ggplot(pred_age_cubic_df$preds, aes(x=age, y=fit))
[…]

### 2022-06-17 manuscript/body.tex
Autocommit

@@
 
-In multivariate models, we may want to visualize the effect of a particular focal input variable.
-In this case, effects styled plots are preferable over the conventional prediction plots which generally focus on the \emph{total} effects.
-To illustrate this, we simulate a multivariate model described in \nameref{S1_Appendix}.
+To illustrate the distinction between the two approaches, we simulate a multivariate model described in \nameref{S1_Appendix}.
 We also use the same model to illustrate how predictions and effects differ in the presence of higher order interactions in focal or non-focal input variables.
-
-We generated two sets of predictions and effects: 1) one with \code{age} as the focal input variable (a cubic polynomial polynomial focal predictor); and 2) one with \code{Wealthindex} as the focal in put variable (linear focal input variable).
+We then generated two sets of mean-based predictions and effects: 1) one with \code{age} as the focal input variable (a cubic polynomial focal predictor); and 2) one with \code{Wealthindex} as the focal in put variable (linear focal input variable).
+In both cases the, the center point was used as the anchor for the effects.
 Then compared the predictions and effects as shown in Fig~\ref{fig:pred_cubic_plots}.
@@
 \end{center}
-\caption{{\bf Prediction-styled and effect-styled plots for cubic polynomial and simple focal predictors.} A: The focal predictor is a cubic polynomial.
-B: The focal predictor is a linear function, with a cubic polynomial non-focal predictor.
-In both cases, the mean-anchored effects are narrower than the non-anchored (predictions), but intersects at the mean of the focal predictor (model center) only in the case of linear focal predictor -- B.
-In the case of cubic polynomial focal predictor, the model center is not a point in the predictor space, i.e., mean of the focal predictor, but a combination of all the model variables in the cubic polynomial, consequently, mean-anchored curves will not necessarily cross at a point -- A.
-The horizontal black and red dashed lines are the observed and predicted average household size, i.e., $\over{hh~size}$ and $\over{\widehat{hh~size}}$, respectively; the inner red dotted curves and crossing lines are the effects for A and B, respectively; while the outer blue curves are the \BC{predictions (place holder for now)} in both cases.
-The central curve and trend line, in A and B, respectively; are the predicted household size based on age and wealth index, respectively.
-\bmb{Use different line types ...}\BC{Confirm with BMB}}
+\caption{{\bf Prediction and effect plots for cubic polynomial and simple focal predictors.} A: The focal input variable is a cubic polynomial.
+B: The focal input variable is linear, with polynomial non-focal input variable.
+The central curve and trend lines are the same for predictions and effects plots in the respective plots.
+However, the confidence intervals for the effects are narrower than those of the predictions in both cases.
+The default anchor for generating effects is the center point.
+For simple models (or simple focal input variables), the center point is a point within the range of focal values; and therefore the confidence interval lines intersect at the center point -- B.
+In the case of higher order interactions such as cubic polynomials, the center point is not a point but a combination of all the predictor variables associated with the cubic polynomial, consequently, curves will not necessarily cross at a point -- A.
+The horizontal black and red dashed lines are the observed and predicted average household size, i.e., $\over{hh~size}$ and $\over{\widehat{hh~size}}$, respectively.}
 \label{fig:pred_cubic_plots}
@@
 
-In the absence of complex higher order interaction terms or transformations on the focal predictor in simple linear models, we would expect, at the model center, the observed average, $\over{hh~size}$, and the average of the predictions, $\over{\widehat{hh~size}}$, to be identical and all intersect at the mean-anchored confidence bands (effects), at the model center (Fig~\ref{fig:pred_cubic_plots}B).
-On the other hand, if the focal predictor is characterized by complex interactions or any other form of transformation, the model center is not necessarily a point in the predictor space.
-In this case, the mean-anchored confidence bands will be narrower than the non-anchored ones but will not necessarily cross at the model center (Fig~\ref{fig:pred_cubic_plots}A).
+Predictions are the standard curves (wider); they tell us what we expect to see if we know the value of the focal input variable.
+In multivariate models, at times, we may want to focus on the effect of the focal input variable only.
+Unlike prediction plot, which captures the uncertainty in the entire model, effect plot focuses on the uncertainty of the focal input variable only.
+In this case, effects styled plots are preferable over the conventional prediction plots which generally focus on the \emph{total} effects, and are usually narrower around the anchor.
+
+In the absence of complex higher order interactions or transformations on the focal input variable, in simple OLS models we would expect the observed average, $\over{hh~size}$, and the average of the predictions, $\over{\widehat{hh~size}}$, to be identical and interactions at the center point, for the effects plot.
+On the other hand, if the focal input variable is characterized by complex interactions or any other form of transformations, the center point is not necessarily a point in the range of focal input values.
+In this case, the effects will be narrower than the predictions plots but will not necessarily intersect.
 
@@
 
-Mean-based and whole-sample-based approaches can produce very different results in models with nonlinear link functions with additional sources of potential bias such as additional non-focal predictor(s), random effect terms or complex interactions.
-To demonstrate this, we considered a two predictor binary outcome simulation described in \nameref{S2_Appendix}, such that the age effect is way smaller than that of the wealth index, and compared their effects on the predicted probability of improved water quality as shown in Fig~\ref{fig:pred_bin_plots}.
+Mean-based and whole-sample-based approaches can produce very different results in models with nonlinear link functions with additional sources of potential bias such as additional non-focal input variable(s), random effects or complex interactions.
+To demonstrate this, we considered a two predictor binary outcome simulation described in \nameref{S2_Appendix}, such that the age a very small effect size in comparison to wealth index, and compared their effects on the predicted probability of improved water quality as shown in Fig~\ref{fig:pred_bin_plots}.
 
@@
 \end{center}
-\caption{{\bf Mean-based and whole-sample-based predictions.} The two approaches produce relatively different estimates in each of the predictors but the difference is more pronounced when the effect of the particular focal predictor has no strong effect -- \emph{on the left}.
-In both cases, \emph{right} and \emph{left}, the average predicted probability of improved water based on whole-sample-based approach is very close to the observed proportion in the simulated data.
-The horizontal blue, black and the red dotted lines are the average mean-based, whole-sample-based predicted probability and observed proportion of improved water quality, respectively.
-The vertical dotted black line represents the mean of the focal predictor, and the point at which it crosses the red dotted line represents the expected ``perfect'' prediction at the model center.
-The grey points are binned observations -- observed proportions of improved water quality in each bin.
-We would expect all the curves, horizontal, and vertical lines to cross at the model center but due the nonlinear averaging (at least in the bias corrected whole-sample-based approach) of the predictions on the response scale, this may not always be the case.}
+\caption{{\bf Mean-based and whole-sample-based predictions.} The \emph{truth} is the prediction based on the true simulation parameters.
+If the focal input variable has small effect size, mean-based and whole-sample-based approaches produce different predictions -- A.
+In the case of strong effect size, however, the two approaches produce similar predictions -- B.
+In both cases, the whole-sample-based predictions are close to the truth.
+The difference in mean-based and whole-sample-based is due to the bias induced by the non-focal variables and the nonlinear link function in the logistic model.
+Mean-based approach suffers from nonlinear averaging; and if focal input variable has small effect size, the bias is even pronounced.
+However, this is not the case if the non-focal input variables have small effect since the bias they induce is equally likely to be small.
+Since whole-sample-based approach averages over the whole population of the non-focal predictor variables, it is less likely to be affected by the nonlinear averaging for the reasons already mentioned.
+The horizontal dashed lines correspond to the respective averages.
+The vertical dashed lines represent the center point, and the point at which they intersect the horizontal lines represents the expected ``perfect'' prediction at the model center.
+The grey points are binned observations -- observed proportions of improved water quality in each bin.}
 \label{fig:pred_bin_plots}
@@
 
-If there was no effect of nonlinear averaging, then we would expect the observed proportion and the average of the predicted probabilities to cross at the model center as we see in Fig~\ref{fig:pred_cubic_plots}B.
-However, the differences we see in Fig~\ref{fig:pred_bin_plots} are due to nonlinear averaging since both observed status and predicted probabilities are averaged on the response scale as opposed to link scale.
-In particular, if the range of values are bigger than $0.5$ (seemingly the case here), then we would expected the averages to be slightly higher than what we would expect at the model center, and vice versa.
+If there was no effect of nonlinear averaging, then we would expect the average observed proportion and the average predictions to intersect at the center point as we see in Fig~\ref{fig:pred_cubic_plots}B.
+One of possible reason for the variations we see in Fig~\ref{fig:pred_bin_plots} is the nonlinear averaging; since both observed status and predicted probabilities are averaged on the response scale as opposed to link scale.
+For example, if the range of values are bigger than $0.5$ (seemingly the case here), then we would expected the averages to be slightly higher than what we would expect at the center point.
 

### 2022-06-17 manuscript/glm_two_predictor.R

@@ -8,7 +8,7 @@ set.seed(991)
 
 ############################################################################################
 # Global variables
-nHH_obs <- 1e4
+nHH_obs <- 1e3
 perHH <- 1
 
 ############################################################################################
@@ -18,17 +18,20 @@ perHH <- 1
 ### No interaction model
 
 #### Simulation
-glm_sim_df <- linearsim(nHH=nHH_obs
+glm_sim <- linearsim(nHH=nHH_obs
 	, perHH=perHH
 	, addnoiseGLM=FALSE
 	, noiseSD=1 # ignored if addnoiseGLM=FALSE
 	, form=~1+x1+x2
-	, betas = c(5, 0.5, 1.5)
+	, betas = c(2, 0.5, 1.5)
 	, pgausian=list(p=2,fun=rnorm, mean=0, sd=1)
 	, pcat=list(p=0)
 	, link_scale=FALSE
 	, vnames=c("age", "wealthindex", "status")
-)$data
+)
+glm_sim_betas <- glm_sim$betas
+print(glm_sim_betas)
+glm_sim_df <- glm_sim$data
 head(glm_sim_df)
 true_prop_df <- (glm_sim_df
 	%>% summarize_all(mean)

### 2022-06-18 manuscript/abstract.tex
Heading to school

@@
 An alternative is the whole-sample-based approach which estimates the average effect in the population.
-Moreover, anchored effects provide an alternative and a more clear way to describe uncertainty associated with the focal input variable.
+Moreover, effects confidence intervals provide an alternative and a more clear way to describe uncertainty associated with the focal input variable.
 In addition to theoretical and methodical comparison, using simulation, we illustrate the two approaches and show that they can produce substantially different results and that whole-sample-based approach can not only produce estimates consistent with the observed values, but also appropriate for bias correction.
-We also present an alternative way, anchored confidence intervals, to describe uncertainties associated with these predictions.
+We also present an alternative way, effects confidence intervals, to describe uncertainties associated with these predictions.
 

### 2022-06-18 manuscript/body.tex
Heading to school

@@
 For example, in logistic regression models, the coefficient estimates, when reported as log odds, are not easy to interpret and are less informative.
-Since logistic models are nonlinear (due to the nonlinear link function), in multivariate models interactions or other nonlinear transformations, the magnitude of the effect of change in the outcome depends on the values of the input variable of interest and the other remaining input variables.
-Hence, the conclusions one can make about the estimated predictions greatly depends on how well we choose the values of the other input variables, i.e., \emph{non-focal} input variables.
+Since logistic models are nonlinear (due to the nonlinear link function), in multivariate models with complex interactions, the magnitude of the effect of change in the outcome depends on the values of the input variable of interest and the other input variables, i.e., \emph{non-focal} input variables.
+In other words, the conclusions we make about the predictions greatly depends on how well we choose the values of the non-focal input variables.
 
@@
 
-Simple OLS regression models are flexible and can easily be generalized, as generalized linear (mixed) models (GL(M)Ms), to examine more complex relationships, including nonlinear relationships between response and input variables e.g., interactions between and nonlinear transformation on the input variables, (and via splines for example), and nonlinear transformations via link functions.
+Simple OLS regression models are flexible and can easily be generalized, as generalized linear (mixed) models (GL(M)Ms), to examine more complex relationships, including nonlinear relationships between response and input variables e.g., higher order interactions via polynomials and splines, nonlinear transformation on the input variables, and nonlinear transformations via link functions.
 This flexibility comes with additional complications, for example, complex multivariate models may risk misinterpretation, and miscalculation of quantities of interest \citep{berry_improving_2012, leeper2017interpreting}.
-Further, coefficient estimates of GLM models involving nonlinear link functions lack direct interpretation\citep{leeper2017interpreting}, meaning that interpretation of derived quantities from these estimates requires some understanding of the specified model.
+Further, coefficient estimates of GLMs involving nonlinear link functions lack direct interpretation\citep{leeper2017interpreting}, meaning that interpretation of derived quantities from these estimates requires some understanding of the specified model.
 An alternative is to explore \emph{predictions} and \emph{effects} \citep{fox2009effect, leeper2017package, lenth2018package}.
@@
 
-When visually presented, predictions and effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind of transformations on the input variables whose estimates are usually, but not always, harder to interpret.
+When visually presented, predictions and effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction or some kind of transformations on the input variables whose estimates are usually, but not always, harder to interpret.
 Generating these quantities together with the associated confidence intervals for regression models has a number of challenges.
@@
 
-The most common way of choosing the representative values is taking unique levels of the focal input variable if discrete or taking appropriately sized quantiles if continuous, and then calculating the predictions while holding non-focal (other input variables other than the focal input variable) input variables at their reference point (values we chose for the non-focal input variables e.g., means) \citep{fox2009effect, hanmer2013behind}.
-This procedure generates -- predictions (at least in our case), \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}.
-In this article, we refer to these quantities as predictions and effects since they should, for example, tell us what we would expect the response to be at a particular value or level of the input variable, for an ``average case''.
-More specifically, the basic idea is to compute the expected outcome by holding the non-focal input variables constant, for example at their mean, while varying the focal input variable, with the goal that estimates represent how the model responds to the changes in the focal input variable.
+The most common way of choosing the representative values is taking unique levels of the focal input variable if discrete or taking appropriately sized quantiles if continuous.
+The predictions are then calculated by holding the non-focal input variables at their reference point (values chosen for the non-focal input variables, e.g., means), while varying the focal input variable, with the goal that the estimates represent how the model responds to the changes in the focal input variable \citep{fox2009effect, hanmer2013behind}.
+This procedure generates -- \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}.
+In this article, we refer to these quantities as predictions and effects since they should, for example, tell us what we would expect the response to be at a particular value or level of the focal input variable, for an ``average case''.
+.
 As mentioned above, in our case, the key difference between predictions and effects is in the computation of the confidence intervals.
@@
 The most commonly used \proglang{R} software packages for generating predictions (\pkg{emmeans} and \pkg{effects}), by default, use the average of \emph{input} and \emph{predictor variables} as the reference point for continuous and categorical input variables, respectively.
-However, there are a number of choices one can make when considering this approach -- for example, in the presence of interaction, averaging the interactions (averaging the product of interacting predictor variables) versus taking the product of the averages of the interacting predictor variables (the default for \pkg{emmeans} and \pkg{effects}).
+However, there are a number of choices one can make when generating predictions -- for example, in the presence of interactions, averaging the interactions (averaging the product of interacting predictor variables) versus taking the product of the averages of the interacting predictor variables (the default for \pkg{emmeans} and \pkg{effects}).
 For simple models (with no interactions), the two approaches give the same predictions.
@@
 %
-We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$ $\beta_{23} = 5$, $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and $\sigma^2 = 1$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions (predictions based on the true simulation parameters) and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.
+We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$, $\beta_{23} = 5$, $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and $\sigma^2 = 1$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions (predictions based on the true simulation parameters) and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.
 %
@@
 %
-In the absence of interaction (model~\ref{eq:simple_inter_higher_no_interaction}), the three approaches produce the same estimates, which match the simulated values, Fig~\ref{fig:justify_plots}A.
-However, in the presence of interaction, even as simple as the one in model~\ref{eq:simple_inter_higher}, the estimates starts to differ.
+In the absence of interaction (Equation~\ref{eq:simple_inter_higher_no_interaction}), the three approaches produce the same estimates, which match the simulated values, Fig~\ref{fig:justify_plots}A.
+However, in the presence of interaction, even as simple as the one in Equation~\ref{eq:simple_inter_higher}, the estimates starts to differ.
 In particular, \pkg{emmeans} and \pkg{effects} give similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is very close to the simulated average ($\bar{y}$), Fig~\ref{fig:justify_plots}B.
-To generate Fig~\ref{fig:justify_plots}A, all the three packages average $x_2$ and $x_3$, i.e., the input variables.
+To generate Fig~\ref{fig:justify_plots}A,  \pkg{emmeans} and \pkg{effects} average $x_2$ and $x_3$, i.e., the input variables while \pkg{varpred} averages the predictor variables corresponds to $x_2$ and $x_3$.
 On other hand, to generate Fig~\ref{fig:justify_plots}B, the difference in the estimates lies on how each of the packages average the interaction term ($x_2x_3$).
@@
 
-One may be interested in the uncertainties associated with the focal input variable of interest only, excluding other uncertainties due to other non-focal input variables -- effects.
-However, currently, the two packages do not provide straightforward way do achieve this.
-To illustrate this, we generate predictions of $x_2$ from model~\ref{eq:simple_inter_higher_no_interaction} together with the associated $95\%$ confidence bands, as shown in Fig~\ref{fig:justify_ci_plots}.
+One may be interested in the uncertainties associated with the focal input variable only, excluding other uncertainties due to other non-focal input variables -- effects.
+However, currently, the two packages do not provide a straightforward way do achieve this.
+To illustrate this, we generate predictions of $x_2$ from Equation~\ref{eq:simple_inter_higher_no_interaction} together with the associated $95\%$ confidence intervals, as shown in Fig~\ref{fig:justify_ci_plots}.
 
@@
 \caption{{\bf A comparison of conventional $95\%$ confidence bands and effects.} The description of horizontal, vertical and trend lines remain the same as above.
-The wider dotted blue curves overlaying the green curves corresponds to the conventional confidence bands around the predcitions from \pkg{emmeans} and \pkg{effects}, while the narrower black curves crossing at the mean of the focal input variable, i.e., the model center, corresponds to the effects of the focal input variable from \pkg{varpred}.
+The wider dotted blue curves overlaying the green curves corresponds to the conventional confidence bands around the predictions from \pkg{emmeans} and \pkg{effects}, while the narrower black curves crossing at the mean of the focal input variable, i.e., the model center, corresponds to the effects of the focal input variable from \pkg{varpred}.
 For simple models, the isolated confidence bands crosses at the model center.
@@
 
-Using simple indirect mediation simulation model described in \nameref{S3_Appendix}, fitted two models -- \emph{non-mediated} which models \emph{z} as a function of \emph{x} (univariate model) and the \emph{mediated} which models \emph{z} as a function of both \emph{x} and \emph{y} (multivariate model).
+A \emph{mediated} effect is a situation in which only some of the input variables have a \emph{causal} impact on the outcome, even though all the variables are (somehow) associated with the outcome.
+For example, the age at marriage may be strongly associated with divorce rate but only married people divorce.
+So does marriage \emph{cause} divorce?
+
+For simplicity, we consider an hypothetical simulation for three variables: predictors $x$ and $y$; and outcome variable $z$; such that
+%
+\begin{center}
+\begin{tikzpicture}
+    \node (1) at (0,0) {x};
+    \node (2) [right = of 1] {y};
+    \node (3) [below = of 2] {z};
+    \path (1) edge  (2);
+    \path (1) edge (2);
+    \path (2) edge (3);
+    \path (1) edge (3);
+\end{tikzpicture}
+\end{center}
+%
+The arrows above show the direction of influence; and the implication of $x\rightarrow y$ and $y \rightarrow z$ is that $x$ affects $z$ in two ways.
+First, it has (some) direct effect; and second, it can have indirect effect by influencing $y$, regulated by parameter $\rho$ as shown in \nameref{S3_Appendix}.
+
+To make inferences about this mediated effect, we need more than one model.
+First, the regression of $z$ on $x$, i.e., the \emph{non-mediated} (or univariate) model which tells on the \emph{total} effect of $x$ on $z$; and second, the regression of $z$ on $x$ and $y$, i.e., the \emph{mediated} (or multivariate or full) model.
+We simulated the data as shown in \nameref{S3_Appendix} and fitted the two models -- non-mediated and mediated.
 In both cases, we compared mean-based and the whole-sample-based predictions, as shown in Fig~\ref{fig:pred_mediated_plots}.
@@
 \caption{{\bf Mean-based and whole-sample-based predictions for mediated effects.} These figures compare the mean-based and whole-sample-based predictions for the mediated (multivariate) and non-mediated (univariate) models.
-A: In the absence of mediator variable, both mean-based and whole-sample-based approaches predictors are identical since there is no any other additional (non-focal predictors) sources of bias.
+A: In the absence of mediator variable, both mean-based and whole-sample-based approaches predictors are identical since there is no any other additional (non-focal input variables) sources of bias.
 Consequently, for both approaches, the average predicted probability is very close to the observed proportion.
-B: When the mediator variable is included, there is no direct effect of $x$ on $z$ and as a result the predictions do not necessarily align with the observations.
+B: When the mediator variable is included,  $x$ has no direct effect on $z$ and as a result the predictions do not align with the observations.
 However, the whole-sample-based approach still closely approximates the average proportion in the simulated data.
-The horizontal blue, black and the red dotted lines are the average mean-based, whole-sample-based predicted probability and observed proportion, respectively.
-The vertical dotted black line represents the mean of the focal predictor, and the point at which it crosses the red dotted line represents the expected ``perfect'' prediction at the model center.
+The horizontal dashed lines are the respective average predictions and observed proportions.
+The vertical dashed black represents the center point, and the point at which it crosses the horizontal lines represents the expected ``perfect'' prediction at the model center.
 The grey points are the binned observations.}
@@
 From Fig~\ref{fig:pred_mediated_plots}A, we see what we would expect in the absence of additional sources of bias, even though in the simulation, the effect of $x$ on $z$ is mediated through $y$.
+This is the total effect of $x$, which only tells us the influence of $x$ on $z$.
 By ignoring $y$ in the model, we are still able to capture the effect of $x$ and closely match the observed values using both approaches.
 However, if there were additional non-focal predictors, we would expect to see the differences similar to those in Fig~\ref{fig:pred_bin_plots}.
-Including both $y$ and $x$ ``dilutes'' the direct effect of $x$ on $y$ and as a result, our prediction do not necessarily match the observed binned observations (see Fig~\ref{fig:pred_mediated_plots}B).
+Including both $y$ and $x$ ``dilutes'' the direct effect of $x$ on $z$ and as a result, our predictions do not necessarily match the observed binned observations (see Fig~\ref{fig:pred_mediated_plots}B).
 
@@
 
-Our simulations examples majorly focused on simple linear and logistic models due their wide range of usage and application.
-In addition, these models act as a starting point for building other complex models, including mixed effect models and models with categorical predictors.
-However, the logic for extension of approaches to more complex models, including other forms of nonlinear link functions is very straightforward.
+Our simulations examples focused on simple linear and logistic models due their wide range of usage and application.
+These models also act as a starting point for building other complex models, including mixed effect models and models with categorical predictors.
+The logic for extending to more complex models, including other forms of nonlinear link functions is very straightforward.
+In fact the components needed for extension are the correct linear predictor and the inverse link function; everything else generalizes.
 In addition, our \proglang{R} package implementation already extends to and supports most of the nonlinear link functions and mixed model framework, including multivariate binary outcome models.
@@
 Generating outcome predictions or predicted probabilities from simple and generalized linear (mixed) models is not only important but also, generating quantities which are consistent with the observed values should be of interest.
-However, many studies still report coefficient estimates from generalized models like probit, logistic, etc., \citep{hanmer2013behind}, which are subject of less clarity due to lack of direct link to the outcome of interest.
+However, many studies still report coefficient estimates from generalized models like probit, logistic, etc., \citep{hanmer2013behind}, which are subject of less clarity and complexities in interpretation.
 
 The argument and results we present in this paper supports a greater need for a shift on focus on what and how to present predictions from generalized models.
-For example, we believe that effect-styled confidence intervals could provide more clarity concerning the uncertainty due to the predictor of interests as opposed to the conventional way of incorporating everything.
+For example, the effect-styled confidence intervals could provide more clarity concerning the uncertainty due to the input variable of interests as opposed to the conventional way of incorporating everything.
 
@@
 
-\noindent\textbf{Supervision:} Jonathan Dushoff, Benjamin M. Bolker
-
 \noindent\textbf{Writing – original draft:} Steve Cygu

### 2022-06-18 manuscript/mediate_preds.R

@@ -1,6 +1,7 @@
 library(shellpipes)
 library(vareffects); varefftheme()
 library(ggpubr)
+library(ggthemes)
 library(ggplot2)
 library(dplyr)
 
@@ -35,6 +36,7 @@ summary(binned_df)
 
 ### Combine all predictions
 vlist <- list(pred_notmediated_none, pred_notmediated_pop)
+col_limits <- c("Observed mean", "Mean-based", "Whole-sample-based")
 
 pred_notmediated_plots <- (comparevarpred(vlist=vlist
 		, lnames=NULL
@@ -47,10 +49,8 @@ pred_notmediated_plots <- (comparevarpred(vlist=vlist
 	+ geom_hline(data=observed_df_med, aes(yintercept=z, colour=model, lty=model))
 	+ geom_vline(data=observed_df_med, aes(xintercept=x), lty=2, col="black")
 	+ geom_point(data=binned_df, aes(x=x, y=z), colour="grey")
-	+ scale_colour_manual(breaks = c("Observed mean", "Mean-based", "Whole-sample-based")
-		, values=c("Observed mean"="red", "Mean-based"="blue", "Whole-sample-based"="black")
-	)
-	+ scale_linetype_manual(values=c("Observed mean"=2, "Mean-based"=1, "Whole-sample-based"=1))
+	+ scale_color_colorblind(limits=col_limits)
+ 	+ scale_linetype_discrete(limits=col_limits)
 	+ labs(colour="Method", linetype="Method", title="A) Non-mediated", y="Predicted probability")
 	+ theme(legend.position="bottom")
 )
@@ -90,10 +90,8 @@ pred_mediated_plots <- (comparevarpred(vlist=vlist
 	+ geom_hline(data=observed_df_med, aes(yintercept=z, colour=model, lty=model))
 	+ geom_vline(data=observed_df_med, aes(xintercept=x), lty=2, col="black")
 	+ geom_point(data=binned_df, aes(x=x, y=z), colour="grey")
-	+ scale_colour_manual(breaks = c("Observed mean", "Mean-based", "Whole-sample-based")
-		, values=c("Observed mean"="red", "Mean-based"="blue", "Whole-sample-based"="black")
-	)
-	+ scale_linetype_manual(values=c("Observed mean"=2, "Mean-based"=1, "Whole-sample-based"=1))
+	+ scale_color_colorblind(limits=col_limits)
+ 	+ scale_linetype_discrete(limits=col_limits)
[…]

### 2022-06-18 manuscript/body.tex
Gone through all BB's comments?

@@
 The major difference between the two lies in the construction of the associated confidence intervals.
-In particular, for predictions, we want to use the classic curved confidence intervals (or prediction intervals) which capture the uncertainty in the whole model and generally focus on the \emph{total} effects, and so will often be suited for a univariate model.
+In particular, for predictions, we want to use the conventional curved confidence intervals (or prediction intervals) which capture the uncertainty in the whole model and generally focus on the \emph{total} effects, and so will often be suited for a univariate model.
 On the other hand, effects, attempt to show the effect of a focal input variable.
@@
 As mentioned above, in our case, the key difference between predictions and effects is in the computation of the confidence intervals.
+The concept of effects has been presented in the literature -- Shi et al.
+\citep{shi_evidence_2017} used effect-styled plots to compare the difference in sexual risk behaviors between circumcised and uncircumcised men.
 
@@
 \end{center}
-\caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} predictions for $x_1$ on $y$ for models with and without interactions.}
-The horizontal blue, green and black lines are the mean predictions, i.e., $\bar{\hat{y}}$, the red ones are the means of the simulated $y$, i.e., $\bar{y}$, while the vertical dotted grey lines are the means of the corresponding focal input variable.
+\caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} predictions for regression of $x_1$ on $y$.} The horizontal dashed lines are the mean of the predicted and simulated $y$, $\bar{\hat{y}}$ and $\bar{y}$, respectively.
 The grey points are the binned simulated $y$.
-The trend lines represent the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other input variables at their average for the three methods and the true predictions based on the simulation parameters.
+The trend lines represent the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other input variables at their mean and the true predictions based on the simulation parameters.
 A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the simulated value in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$.
-Similarly, the predictions based on the three approaches closely match the true predictions.
+Similarly, the predictions based on the three approaches match the true predictions.
 B: Even with the simple interaction between non-focal input variables, we start seeing deviation in both predicted mean, $\bar{\hat{y}}$, and overall predictions from the simulated, $\bar{y}$, and true predictions, respectively; in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not the proposed \pkg{varpred}.}
@@
 %
-In the absence of interaction (Equation~\ref{eq:simple_inter_higher_no_interaction}), the three approaches produce the same estimates, which match the simulated values, Fig~\ref{fig:justify_plots}A.
-However, in the presence of interaction, even as simple as the one in Equation~\ref{eq:simple_inter_higher}, the estimates starts to differ.
+In the absence of interactions (Equation~\ref{eq:simple_inter_higher_no_interaction}), the three approaches produce the same estimates, which match the simulated values, Fig~\ref{fig:justify_plots}A.
+However, in the presence of interactions, even as simple as the one in Equation~\ref{eq:simple_inter_higher}, the estimates start to differ.
 In particular, \pkg{emmeans} and \pkg{effects} give similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is very close to the simulated average ($\bar{y}$), Fig~\ref{fig:justify_plots}B.
@@
 \end{center}
-\caption{{\bf A comparison of conventional $95\%$ confidence bands and effects.} The description of horizontal, vertical and trend lines remain the same as above.
-The wider dotted blue curves overlaying the green curves corresponds to the conventional confidence bands around the predictions from \pkg{emmeans} and \pkg{effects}, while the narrower black curves crossing at the mean of the focal input variable, i.e., the model center, corresponds to the effects of the focal input variable from \pkg{varpred}.
-For simple models, the isolated confidence bands crosses at the model center.
-The wider curves incorporates not only the uncertainties due to intercept term but also other non-focal input variables.
-On the other hand, effects (narrower crossing lines) only take into account the main effect uncertainty of the focal input variable.}
+\caption{{\bf The prediction-styled and effect-styled $95\%$ confidence intervals.} The description of horizontal, vertical and trend lines remain the same as above.
+The wider dotted blue curves overlaying the green curves corresponds to the conventional confidence intervals around the predictions from \pkg{emmeans} and \pkg{effects}, while the narrower black curves crossing at the mean of the focal input variable, i.e., the center point, corresponds to the effects of the focal input variable from \pkg{varpred}.
+For simple models, the effect-styled confidence bands crosses at the center point (an anchor).
+The prediction-styled curves incorporates not only the uncertainties due to intercept term but also other non-focal input variables.
+On the other hand, effect-styled curves only take into account the main effect uncertainty of the focal input variable.}
 \label{fig:justify_ci_plots}
@@
 
-For \pkg{emmeans} and \pkg{effects}, the confidence bands are much wider because they include uncertainties associated with the intercept and non-focal input variables, but narrower and crosses at the mean of the focal input variable in \pkg{varpred}.
+For \pkg{emmeans} and \pkg{effects}, the confidence intervals are much wider because they include uncertainties associated with the intercept and non-focal input variables, but narrower and crosses at the mean of the focal input variable in \pkg{varpred}.
 We refer to this point as the \emph{center point} and the approach as \emph{mean-based}.
-In other words, with \pkg{varpred}, we are able to generate effects indicating zero uncertainty at the value of the focal input variable we are more certain about, i.e., mean of the predictor variable corresponding to the focal input variable.
-For simple models, the point where the confidence bands crosses is the model center; it corresponds to anchor we choose, in this case, the mean of the focal input variable, but we generally choose the mean of the focal predictor (predictors corresponding to the focal input variable).
+In other words, with \pkg{varpred}, we are able to generate effects indicating zero uncertainty at the value of the focal input variable we are more certain about, i.e., the center point.
+For simple models, the point where the confidence intervals curves cross is the model center or simply center point; it corresponds to anchor we choose, in this case, the center point (which is generally an appropriate choice).
+Predictions are the standard curves; they tell us what we expect to see if we know the value of the focal predictor.
+The lines inside (effects) are a little different: they are a good way to visualize the effect of the focal predictor.
+Unlike predictions, where the uncertainty captures the uncertainty in the whole model, the effects plot focuses on uncertainty in the effect of the focal predictor only and depend on the anchor.
 
@@
 
-Predictions are the standard curves; they tell us what we expect to see if we know the value of the focal predictor.
-The lines inside (effects) are a little different: they are a good way to visualize the effect of the focal predictor.
-Unlike predictions, where the uncertainty captures the uncertainty in the whole model, the effects plot focuses on uncertainty in the effect of the focal predictor only and depend on the anchor.
-
 The main purpose of this article is to discuss and implement various approaches for computing predictions and provide an alternative method for computing the associated confidence intervals, i.e., effects.
 We further explore and demonstrate, using simulated data, approaches for correcting bias in predictions for GL(M)Ms involving nonlinear link functions.
+The proposed method and \proglang{R} software package will complement the existing ones by providing: 1) a straightforward way to generate effects, and 2) an alternative and a more robust way to correct for predictions bias in GL(M)Ms.
 
@@
 \item \textbf{Focal and non-focal input variables}: Focal input variable refers to the input variable we are interested in.
-For multivariate models, all other remaining input variables are non-focal input variables.
+For multivariate models, all other variables other than focal input variable are non-focal input variables.
 For example, consider model described by Equation~\ref{eq:simple_inter_higher_no_interaction}.
-Suppose we are interested in the effect of $x_1$ on the predicted $y$, then $x_1$ is the focal input variable while $x_2$ and $x_3$ are non-focal non-focal variables.
+Suppose we are interested in the effect of $x_1$ on the predicted $y$, then $x_1$ is the focal input variable while $x_2$ and $x_3$ are non-focal input variables.
 \item \textbf{Model matrix}: Refers to the design matrix whose rows include all combination of input variables appearing in the interaction terms.
@@
 \item \textbf{Non-focal predictors:} All columns of the model matrix associated with the non-focal input variable(s).
-For example, in the previous household size example, all the columns associated with religion are non-focal predictors.
+For example, in the previous household size example, all the columns associated with religion are non-focal predictors if income is the focal input variable.
 \item \textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more predictor variables).
@@
 We refer to the prediction approach that uses model center as mean-based.
-The center point for a set of predictor variables corresponding to an input variable may not represent a possible value of the input variable -- the case in Fig~\ref{fig:justify_plots} (with interaction).
-In simple linear models (without) interaction, the mean of the input variable can be used as a proxy for center point.
-If we have more than one predictor variables associated with the focal input variable, however, the model center does not correspond to any given value of the variable, and we don't get the nice crossing we get in simpler cases.
-See Fig~\ref{fig:pred_cubic_plots}.
+The center point for a set of predictor variables corresponding to an input variable may not represent a possible value of the input variable -- the case in Fig~\ref{fig:justify_plots}B.
+In simple linear models (without) interaction, the mean of the input variable can be used as a proxy for the center point.
+If we have more than one predictor variables associated with the focal input variable, however, the model center does not correspond to any given value of the variable, and we do not necessarily get the ``nice'' crossing we get in simpler cases, see Fig~\ref{fig:pred_cubic_plots}.
 \item \textbf{Reference point:} Corresponds to the value or values chosen for non-focal input variables, when estimating the predictions.
 Typically the center point (mean-based), but can instead be a whole-sample of quantiles or observations.
-Appropriately choosing the reference point is very crucial in generating correct predictions especially when dealing with non-linear link functions.
+Appropriately choosing the reference point is very crucial in generating correct predictions (for bias correction) especially when dealing with non-linear link functions.
 \item \textbf{Anchor:} The value chosen for the focal input variable when estimating effect confidence intervals.
@@
 Typically chosen as the center point of the predictor variables corresponding to the focal input variable but other sensible values can be chosen too.
-If we are trying to focus on effects, the anchor from which we calculate CIs becomes important.
-This is not true for the more conventional prediction CIs.
+If we are trying to focus on effects, the anchor from which we calculate confidence intervals becomes important.
+This is not true for the more conventional prediction confidence intervals.
 We can choose any anchor, e.g., zero-anchored (the value of the anchor is $0$), but center point seems like a natural, stable choice, we call this center-anchored.
@@
 \end{center}
-\caption{{\bf A comparison of min-based and center- (mean-) based anchors.} The center-based effects are based on the mean of the focal predictor while the min-anchored effects are based on the minimum of the focal predictor.
+\caption{{\bf A comparison of min- and center- anchored effects.} The center-anchored effects are based on the mean of the focal predictor while the min-anchored effects are based on the minimum of the focal predictor.
 As we mentioned earlier, the choice of the anchor do not affect the main estimates (central trend lines).}
@@
 
-Consider a simple linear model with linear predictor $\eta = \bX\bbeta$ and let $g(\boldmu) = \boldeta$ be an identity link function (in the case of simple linear model), where $\boldmuh$ is the expected value of response variable $\hat{y}$.
+Consider a simple linear model with linear predictor $\eta = \bX\bbeta$ and let $g(\boldmu) = \boldeta$ be an identity link function, where $\boldmuh$ is the expected value of response variable $\hat{y}$.
 Let $\bbetah$ be the estimate of $\bbeta$, together with the estimated covariance matrix $\Sigma = V(\bbetah)$ of $\bbetah$.
-Then the quantity $\hat{\boldeta}^\star = \bX^\star\bbetah$ is the prediction with respect to the focal predictor in question \citep{fox2009effect}.
+Then the quantity $\hat{\boldeta}^\star = \bX^\star\bbetah$ is the prediction with respect to the focal input variable in question \citep{fox2009effect}.
 
@@
 Higher order terms such as interactions, splines, polynomials, etc., can be within the focal input variable, within (between) non-focal input variable(s) or between the focal and non-focal input variable(s).
-To distinguish the three, suppose the model which describes the hypothetical simulated household size simulation based on a number of socio-demographic factors such as age and wealthindex is
-
+To distinguish the three, suppose the model which describes the hypothetical simulated household size simulation based on a number of socio-demographic factors such as age and wealth index is
+%
 \begin{align}\label{eq:lm_cubic}
-\mathrm{hh~size}_i &= \beta_0 + \beta_{\mathrm{A_1}}\mathrm{Age}_i + \beta_{\mathrm{A_2}}\mathrm{Age}^2_i + \beta_{\mathrm{A_3}}\mathrm{Age}^3_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \epsilon_i \nonumber\\
+\mathrm{hh~size}_i &= \beta_0 + \beta_{\mathrm{A_1}}\mathrm{Age}_i + \beta_{\mathrm{A_2}}\mathrm{Age}^2_i + \beta_{\mathrm{A_3}}\mathrm{Age}^3_i\nonumber \\
+&+ \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \epsilon_i.
 \end{align}
-
+%
 In the first case, with \code{Age} as the focal input variable, with cubic polynomial interaction ($3$ focal predictors associated with it).
 In this case, each of the focal predictors ($\mathrm{Age}_i, \mathrm{Age}^2_i$ and $\mathrm{Age}^3_i$) are evaluated independently across the chosen levels of the focal input variable, $Age_i$.
-Specifically, the higher order terms are treated as additional columns of the model matrix evaluated with the same values of the focal input variable, while non-focal predictors are fixed at their reference point as discussed in the previous section.
+Specifically, the higher order terms are treated as additional columns of the model matrix evaluated with the same values chosen for the focal input variable, while non-focal predictors are fixed at their reference point as discussed in the previous section.
 In this case, the predictions associated with \code{Age} on the linear predictor scale become
-
-\begin{align*}
-\eta^\star(\mathrm{Age}_i, \nset{{\over{\mathrm{Wealthindex}}}}) &= \beta_0 + \beta_{\mathrm{A_1}}\mathrm{Age}_i + \beta_{\mathrm{A_2}}\mathrm{Age}^2_i + \beta_{\mathrm{A_3}}\mathrm{Age}^3_i\\
-	& + \beta_{\mathrm{W}}\over{\mathrm{Wealthindex}}
-\end{align*}
-
+%
+\begin{align}
+\eta^\star(\mathrm{Age}_i, \nset{{\over{\mathrm{Wealthindex}}}}) &= \beta_0 + \beta_{\mathrm{A_1}}\mathrm{Age}_i + \beta_{\mathrm{A_2}}\mathrm{Age}^2_i + \beta_{\mathrm{A_3}}\mathrm{Age}^3_i \nonumber\\
+	& + \beta_{\mathrm{W}}\over{\mathrm{Wealthindex}}.
+\end{align}
+%
 In the second case, with \code{Wealthindex} as the focal input variable, the non-focal variable, \code{Age}, is a cubic polynomial.
@@
 Thus
-
-\begin{align*}
-\eta^\star(\mathrm{Wealthindex}_i, \nset{{\{\over{\mathrm{Age\mathop{\vphantom{^2}}}}, \over{\mathrm{Age}^2}, \over{\mathrm{Age}^2}\}}}) &= \beta_0 + \beta_{\mathrm{A_1}}\over{\mathrm{Age\mathop{\vphantom{^2}}}} + \beta_{\mathrm{A_2}}\over{\mathrm{Age}^2} + \beta_{\mathrm{A_3}}\over{\mathrm{Age}^3}\\
-	& + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i
-\end{align*}
-
+%
+\begin{align}
+\eta^\star(\mathrm{Wealthindex}_i, \nset{{\{\over{\mathrm{Age\mathop{\vphantom{^2}}}}, \over{\mathrm{Age}^2}, \over{\mathrm{Age}^2}\}}}) &= \beta_0 + \beta_{\mathrm{A_1}}\over{\mathrm{Age\mathop{\vphantom{^2}}}} + \beta_{\mathrm{A_2}}\over{\mathrm{Age}^2} + \beta_{\mathrm{A_3}}\over{\mathrm{Age}^3}\nonumber\\
+	& + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i.
+\end{align}
+%
 Lastly, consider model described by Equation~\ref{eq:simple_inter_higher}, previously used to demonstrate how to handle interactions in non-focal input variables.
@@
 In this case, the non-interacting predictors are treated as non-focal predictors, with appropriate reference points.
-However, for the interacting predictors, we first choose representative values for the focal input variable and some particular values of the interacting non-focal input variable in a similar way we choose values for the focal input variable or predetermined.
+However, for the interacting predictors, we first choose representative values for the focal input variable and then some particular values of the interacting non-focal input variable in a similar way we choose values for the focal input variable or predetermined.
 In our example, suppose we pick $i$ and $j$ unique values of the focal input variable, $x_2$ and interacting input variable $x_3$, respectively.
 The prediction on linear predictor scale is given by
-
+%
 \begin{align*}
@@
 \end{align*}
+%
 
@@
 This is the effect of the particular focal input variable.
-By setting $\baf = \boldsymbol 0$, we get the variances for the conventional CIs, Equation~\ref{eq:conventional_variance}.
+By setting $\baf = \boldsymbol 0$, we get the variances for the conventional CIs in Equation~\ref{eq:conventional_variance}.
 
@@
 \item for every value of the focal linear predictor, $\eta_{jf}$, compute
-
+%
 \begin{align}\label{eq:pop_eta} 
 \eta_j^\star(\eta_{jf}, \nset{\eta})  &= \eta_{jf} + \nset{\eta} \nonumber \\
-&= \eta_j^\star(x_f, \nset{x})
+&= \eta_j^\star(x_f, \nset{x}).
 \end{align}
 \end{itemize}
+%
 
 Once Equation~\ref{eq:pop_eta} is computed, we back-transform the estimates to the original scale and average over the levels of the focal predictors, $j$:
-
+%
 \begin{align}\label{eq:pop_response} 
-\hat{y}_f  &= \textrm{mean} ~ g^{-1} \left(\eta_j^\star(x_f, \nset{x})\right)
+\hat{y}_f  &= \textrm{mean} ~ g^{-1} \left(\eta_j^\star(x_f, \nset{x})\right).
 \end{align}
+%
 
 We make similar adjustments to compute the variances of the predictions at every level of the focal input variable:
-
+%
 \begin{align}
@@
 where $\bX^{\star}_{jc} = \{\bX_{jf}^\star, \nset{{\bX}^\star} - \nset{{\bar{\bX}}^\star}\}$ and 
-
+%
 \begin{align}
-\mathrm{CI}_f = \mathrm{mean} ~ g^{-1} \left(\eta_j^\star(x_f, \nset{x}) \pm q\sigma_{jf}\right)
+\mathrm{CI}_f = \mathrm{mean} ~ g^{-1} \left(\eta_j^\star(x_f, \nset{x}) \pm q\sigma_{jf}\right).
 \end{align}
+%
 
@@
 In particular
-
+%
 \begin{align}\label{eq:pop_eta_re} 
@@
 We also use the same model to illustrate how predictions and effects differ in the presence of higher order interactions in focal or non-focal input variables.
-We then generated two sets of mean-based predictions and effects: 1) one with \code{age} as the focal input variable (a cubic polynomial focal predictor); and 2) one with \code{Wealthindex} as the focal in put variable (linear focal input variable).
+We then generated two sets of mean-based predictions and effects: 1) one with \code{age} as the focal input variable (a cubic polynomial focal input variable); and 2) one with \code{Wealthindex} as the focal in put variable (linear focal input variable).
 In both cases the, the center point was used as the anchor for the effects.
@@
 B: The focal input variable is linear, with polynomial non-focal input variable.
-The central curve and trend lines are the same for predictions and effects plots in the respective plots.
+The central curve and trend lines are the same for predictions and effects plots in each case.
 However, the confidence intervals for the effects are narrower than those of the predictions in both cases.
 The default anchor for generating effects is the center point.
-For simple models (or simple focal input variables), the center point is a point within the range of focal values; and therefore the confidence interval lines intersect at the center point -- B.
+For simple models (or simple focal input variables), the center point is a point within the range of focal values; and therefore the confidence interval lines intersect at that that point -- B.
 In the case of higher order interactions such as cubic polynomials, the center point is not a point but a combination of all the predictor variables associated with the cubic polynomial, consequently, curves will not necessarily cross at a point -- A.
@@
 Unlike prediction plot, which captures the uncertainty in the entire model, effect plot focuses on the uncertainty of the focal input variable only.
-In this case, effects styled plots are preferable over the conventional prediction plots which generally focus on the \emph{total} effects, and are usually narrower around the anchor.
+In this case, effects styled plots which are usually narrower around the anchor are preferable over the conventional prediction plots which generally focus on the \emph{total} effects.
 
-In the absence of complex higher order interactions or transformations on the focal input variable, in simple OLS models we would expect the observed average, $\over{hh~size}$, and the average of the predictions, $\over{\widehat{hh~size}}$, to be identical and interactions at the center point, for the effects plot.
+In the absence of complex higher order interactions or transformations on the focal input variable, in simple OLS models we would expect the observed average, $\over{hh~size}$, and the average of the predictions, $\over{\widehat{hh~size}}$, to be identical and intersect with the trend lines at the center point, for the effects plot.
 On the other hand, if the focal input variable is characterized by complex interactions or any other form of transformations, the center point is not necessarily a point in the range of focal input values.
@@
 Mean-based and whole-sample-based approaches can produce very different results in models with nonlinear link functions with additional sources of potential bias such as additional non-focal input variable(s), random effects or complex interactions.
-To demonstrate this, we considered a two predictor binary outcome simulation described in \nameref{S2_Appendix}, such that the age a very small effect size in comparison to wealth index, and compared their effects on the predicted probability of improved water quality as shown in Fig~\ref{fig:pred_bin_plots}.
+To demonstrate this, we considered a two predictor binary outcome simulation described in \nameref{S2_Appendix}, such that age has a very small effect size in comparison to wealth index, and compared their effects on the predicted probability of improved water quality as shown in Fig~\ref{fig:pred_bin_plots}.
 
@@
 The difference in mean-based and whole-sample-based is due to the bias induced by the non-focal variables and the nonlinear link function in the logistic model.
-Mean-based approach suffers from nonlinear averaging; and if focal input variable has small effect size, the bias is even pronounced.
+Mean-based approach suffers from nonlinear averaging; and if focal input variable has small effect size, the bias is even pronounced since the effect of nonlinear averaging is driven by the non-focal input variable with the strong effect.
 However, this is not the case if the non-focal input variables have small effect since the bias they induce is equally likely to be small.
@@
 If there was no effect of nonlinear averaging, then we would expect the average observed proportion and the average predictions to intersect at the center point as we see in Fig~\ref{fig:pred_cubic_plots}B.
-One of possible reason for the variations we see in Fig~\ref{fig:pred_bin_plots} is the nonlinear averaging; since both observed status and predicted probabilities are averaged on the response scale as opposed to link scale.
+One possible reason for the variations we see in Fig~\ref{fig:pred_bin_plots} is the nonlinear averaging; since both observed status and predicted probabilities are averaged on the response scale as opposed to link scale.
 For example, if the range of values are bigger than $0.5$ (seemingly the case here), then we would expected the averages to be slightly higher than what we would expect at the center point.
@@
 
-For simplicity, we consider an hypothetical simulation for three variables: predictors $x$ and $y$; and outcome variable $z$; such that
+For simplicity, we consider an hypothetical simulation for three variables: input variables $x$ and $y$; and outcome variable $z$; such that
 %
@@
 \end{center}
-\caption{{\bf Mean-based and whole-sample-based predictions for mediated effects.} These figures compare the mean-based and whole-sample-based predictions for the mediated (multivariate) and non-mediated (univariate) models.
-A: In the absence of mediator variable, both mean-based and whole-sample-based approaches predictors are identical since there is no any other additional (non-focal input variables) sources of bias.
+\caption{{\bf Mean-based and whole-sample-based predictions for mediated effects for the mediated (multivariate) and non-mediated (univariate) models.} A: In the absence of mediator variable, both mean-based and whole-sample-based approaches predictions are identical since there is no any other additional (non-focal input variables) sources of bias.
 Consequently, for both approaches, the average predicted probability is very close to the observed proportion.
@@
 By ignoring $y$ in the model, we are still able to capture the effect of $x$ and closely match the observed values using both approaches.
-However, if there were additional non-focal predictors, we would expect to see the differences similar to those in Fig~\ref{fig:pred_bin_plots}.
+However, if there were additional non-focal predictors, we would expect to see the differences similar to those in Fig~\ref{fig:pred_bin_plots}A.
 Including both $y$ and $x$ ``dilutes'' the direct effect of $x$ on $z$ and as a result, our predictions do not necessarily match the observed binned observations (see Fig~\ref{fig:pred_mediated_plots}B).
@@
 \label{S3_Appendix}
-{\bf Mediated effect simulation.} Next, we consider a simple indirect mediation simulation such that $x$ has direct effect on $y$ which in turn has effect on $z$ but $x$ has no direct effect on $y$, i.e., $x \rightarrow y \rightarrow z$.
-In particular:
+{\bf Mediated effect simulation.} Next, we consider a simple indirect mediation previously described and simulate a binary outcome model such that:
 

### 2022-06-19 manuscript/abstract.tex
Gone through all BB's comments?

@@
 A general way to summarize these kind of models is through predictions, that depend not only on choice of representative values of focal input variables but are also sensitive to which values of non-focal input variables are chosen.
-The most common approach is generating predictions at a reference point, usually the mean, of non-focal predictor, i.e., mean-based (centered) which could be considered as the effect of an average case in the population.
-In the presence of additional non-focal input variables, nonlinear link functions, random effect terms, etc., mean-based approach generate predictions that are biased and may not be consistent with the observed quantities.
+The most common approach is generating predictions at a reference point, usually the mean of non-focal predictor.
+We call this mean-based approach and it estimates effect of an average case in the population.
+In the presence of additional non-focal input variables, nonlinear link functions, random effect terms, etc., mean-based approach generates predictions that are biased and may not be consistent with the observed quantities.
 An alternative is the whole-sample-based approach which estimates the average effect in the population.
@@
 % Author Summary not valid for PLOS ONE submissions.   
-\section*{Author summary}
-
-\fix
+% \section*{Author summary}

### 2022-06-19 manuscript/body.tex
Gone through all BB's comments?

@@
 For example, in logistic regression models, the coefficient estimates, when reported as log odds, are not easy to interpret and are less informative.
-Since logistic models are nonlinear (due to the nonlinear link function), in multivariate models with complex interactions, the magnitude of the effect of change in the outcome depends on the values of the input variable of interest and the other input variables, i.e., \emph{non-focal} input variables.
+Since logistic models are nonlinear (due to the nonlinear link function), in multivariate models with complex interactions, the magnitude of change in the outcome depends on the values of the input variable of interest and the other input variables, i.e., \emph{non-focal} input variables.
 In other words, the conclusions we make about the predictions greatly depends on how well we choose the values of the non-focal input variables.
@@
 \item  uncertainty estimation -- appropriate choice of \emph{anchor} for computing confidence intervals; and how to incorporate the uncertainty due to non-focal input variables 
-\item bias in the expected mean prediction induced by the nonlinear transformation of the response variable (especially in GL(M)Ms)
+\item bias in the expected mean prediction induced by the nonlinear transformation of the response variable (especially in GL(M)Ms).
 \end{enumerate}
@@
 In this article, we refer to these quantities as predictions and effects since they should, for example, tell us what we would expect the response to be at a particular value or level of the focal input variable, for an ``average case''.
-.
 As mentioned above, in our case, the key difference between predictions and effects is in the computation of the confidence intervals.
@@
 However, there are a number of choices one can make when generating predictions -- for example, in the presence of interactions, averaging the interactions (averaging the product of interacting predictor variables) versus taking the product of the averages of the interacting predictor variables (the default for \pkg{emmeans} and \pkg{effects}).
-For simple models (with no interactions), the two approaches give the same predictions.
+For simple OLS models (with no interactions), the two approaches give the same predictions.
 However, for complex models, we demonstrate that the two approaches can produce substantially different results and show that averaging the interactions closely matches the observed values.
@@
 y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \epsilon \label{eq:simple_inter_higher_no_interaction}\\
-y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \beta_{23}x_2x_3 + \epsilon \label{eq:simple_inter_higher}
+y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \beta_{23}x_2x_3 + \epsilon \label{eq:simple_inter_higher}.
 \end{align}
 %
+
 We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$, $\beta_{23} = 5$, $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and $\sigma^2 = 1$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions (predictions based on the true simulation parameters) and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.
@@
 Similarly, the predictions based on the three approaches match the true predictions.
-B: Even with the simple interaction between non-focal input variables, we start seeing deviation in both predicted mean, $\bar{\hat{y}}$, and overall predictions from the simulated, $\bar{y}$, and true predictions, respectively; in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not the proposed \pkg{varpred}.}
+B: Even with the simple interaction between non-focal input variables, we start seeing deviation in both predicted mean, $\bar{\hat{y}}$, and overall predictions from the simulated, $\bar{y}$, and true predictions, respectively, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not the proposed \pkg{varpred}.}
 \label{fig:justify_plots}
@@
 On other hand, to generate Fig~\ref{fig:justify_plots}B, the difference in the estimates lies on how each of the packages average the interaction term ($x_2x_3$).
-In particular \pkg{emmeans} and \pkg{effects} use input variables to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses predictor variables to compute $\over{x_2x_3}$.
+In particular, \pkg{emmeans} and \pkg{effects} use input variables to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses predictor variables to compute $\over{x_2x_3}$.
 In the case of Fig~\ref{fig:justify_plots}A, averaging the input variables is equivalent to averaging the predictor variables.
@@
 \end{center}
-\caption{{\bf The prediction-styled and effect-styled $95\%$ confidence intervals.} The description of horizontal, vertical and trend lines remain the same as above.
-The wider dotted blue curves overlaying the green curves corresponds to the conventional confidence intervals around the predictions from \pkg{emmeans} and \pkg{effects}, while the narrower black curves crossing at the mean of the focal input variable, i.e., the center point, corresponds to the effects of the focal input variable from \pkg{varpred}.
-For simple models, the effect-styled confidence bands crosses at the center point (an anchor).
+\caption{{\bf The prediction and effect styled plots at $95\%$ confidence intervals.} The description of horizontal, vertical and trend lines remain the same as above.
+The wider dashed curves corresponds to the conventional confidence intervals around the predictions from \pkg{emmeans} and \pkg{effects}, while the narrower curves crossing at the mean of the focal input variable, i.e., the center point, correspond to the effect of the focal input variable from \pkg{varpred}.
+For simple OLS models, the effect-styled confidence bands crosses at the center point (an anchor).
 The prediction-styled curves incorporates not only the uncertainties due to intercept term but also other non-focal input variables.
@@
 For \pkg{emmeans} and \pkg{effects}, the confidence intervals are much wider because they include uncertainties associated with the intercept and non-focal input variables, but narrower and crosses at the mean of the focal input variable in \pkg{varpred}.
-We refer to this point as the \emph{center point} and the approach as \emph{mean-based}.
 In other words, with \pkg{varpred}, we are able to generate effects indicating zero uncertainty at the value of the focal input variable we are more certain about, i.e., the center point.
-For simple models, the point where the confidence intervals curves cross is the model center or simply center point; it corresponds to anchor we choose, in this case, the center point (which is generally an appropriate choice).
+For simple OLS models, the point where the confidence intervals curves cross is the model center or simply center point; it corresponds to anchor we choose, in this case, the center point (which is generally an appropriate choice).
 Predictions are the standard curves; they tell us what we expect to see if we know the value of the focal predictor.
-The lines inside (effects) are a little different: they are a good way to visualize the effect of the focal predictor.
+The lines inside (effects) are a little different; they are a good way to visualize the effect of the focal predictor.
 Unlike predictions, where the uncertainty captures the uncertainty in the whole model, the effects plot focuses on uncertainty in the effect of the focal predictor only and depend on the anchor.
@@
 The mean prediction is the prediction at the model center.
-On the other hand, when dealing with GL(M)Ms with nonlinear link functions, the averaging is done on the nonlinear scale, i.e., nonlinear averaging which makes estimating correct predictions much harder.
-In particular, bias in expected mean prediction induced by nonlinear transformation of the response variable can lead to inaccurate predictions.
+On the other hand, when dealing with GL(M)Ms with nonlinear link functions, the averaging is done on the nonlinear scale, i.e., nonlinear averaging, making estimating correct predictions much harder.
+This is the \emph{bias} in expected mean prediction induced by nonlinear transformation of the response variable and it lead to inaccurate predictions.
 One way to deal with this is to make predictions on the transformed scale (linear predictor scale), and then back-transform to the original scale.
@@
 We further explore and demonstrate, using simulated data, approaches for correcting bias in predictions for GL(M)Ms involving nonlinear link functions.
-The proposed method and \proglang{R} software package will complement the existing ones by providing: 1) a straightforward way to generate effects, and 2) an alternative and a more robust way to correct for predictions bias in GL(M)Ms.
+The proposed method and \proglang{R} software package will complement the existing ones by providing: 1) a straightforward way to generate effects, and 2) an alternative and a more robust way to correct for prediction bias in GL(M)Ms.
 
@@
 
-In order to discuss more general statistical background and mathematical formulation, we need to understand and formally define a number of terms, some of which have been introduced in the previous section.
+In order to discuss the statistical background and mathematical formulation of the proposed approaches, we need to understand and formally define a number of terms, some of which have been introduced in the previous section:
 
@@
 Suppose we are interested in the effect of $x_1$ on the predicted $y$, then $x_1$ is the focal input variable while $x_2$ and $x_3$ are non-focal input variables.
-\item \textbf{Model matrix}: Refers to the design matrix whose rows include all combination of input variables appearing in the interaction terms.
+\item \textbf{Model matrix}: Refers to the design matrix whose rows include all combination of input variables.
 Consider an example for three hypothetical households -- the first household head is a Christian with an income of $\$ 50$, the second household head is Muslim with an income of $\$ 100$ while the third household head is a Jew with an income of $\$ 77$.
@@
 In simple linear models (without) interaction, the mean of the input variable can be used as a proxy for the center point.
-If we have more than one predictor variables associated with the focal input variable, however, the model center does not correspond to any given value of the variable, and we do not necessarily get the ``nice'' crossing we get in simpler cases, see Fig~\ref{fig:pred_cubic_plots}.
-\item \textbf{Reference point:} Corresponds to the value or values chosen for non-focal input variables, when estimating the predictions.
+If there are more than one predictor variables associated with the focal input variable, however, the model center does not correspond to any given value of the variable, and we do not necessarily get the ``nice'' crossing we get in simpler cases, see Fig~\ref{fig:pred_cubic_plots}.
+\item \textbf{Reference point:} Corresponds to the value or values chosen for non-focal input variables, when estimating the predictions and effects.
 Typically the center point (mean-based), but can instead be a whole-sample of quantiles or observations.
@@
 In Fig~\ref{fig:justify_anchors}, we compare effects plots anchored at the minimum and center (mean) of the focal input variable.
-\end{itemize}
-
+%
 \begin{figure}
@@
 \end{figure}
+%
+\item \textbf{Marginal effects:} In simple linear models with no interaction terms, the coefficient estimates are simple and directly interpretable as the expected change in outcome for a unit change in focal input variable.
+This is the \emph{unconditional marginal effect} and it is constant across all the observations and levels of all other input variables.
+Consider models described by Equation~\ref{eq:simple_inter_higher_no_interaction} and Equation~\ref{eq:simple_inter_higher}.
+The marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher_no_interaction} is $\frac{\partial y}{\partial x_2} = \beta_2$.
+On the other hand, the marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher} is given by $\frac{\partial y}{\partial x_2} = \beta_2 + \beta_{23}x_3$.
+In other words, if there are no interactions, the marginal effect of $x_2$ on $y$ is constant, while, if there are interactions in the model, the marginal effect of a change in $x_2$ on $y$ depends on the value of the other \emph{conditioning} predictor, $x_3$.
+In essence, computation of marginal effects requires the use of partial derivatives and can be considered as the slope multi-dimensional surface with response to the surface of interest \cite{leeper2017interpreting}.
+The \proglang{R} software package \pkg{margins} \cite{lenth2018package} generates marginal effects.
+Consider result from a hypothetical  simulated example -- regression of household size as function of household wealth index and age of household head, as shown in Fig~\ref{fig:qoi_age_pred_plot}.
+Since the model has no interactions, the relationship between the predicted household size and age is linear, hence the marginal effect of age is the slope ($\frac{\Delta \mathrm{hh size}}{\Delta \mathrm{age}}$) of the prediction (equivalently effect) line and can be calculated irrespective of the values of wealth index.
+%
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.8\textwidth]{quantity_of_interest.ggp.pdf}
+\caption{{\bf A comparison of prediction, effect and marginal effect.} A: The black solid line is the prediction (same as effect), while the dashed lines indicate the changes along the \code{age} and \code{household size} axis.
+For a linear model with no interaction, the marginal effect is the slope of the prediction line.
+B: The marginal effect (ME) at representative values of the focal input variable \code{age} together with the $95\%$ confidence intervals.
+For simple OLS models, the ME is constant across all the values of \code{age} chosen.}
+\label{fig:qoi_age_pred_plot}
+\end{figure}
+%
+\end{itemize}
 
@@
 
-Higher order terms such as interactions, splines, polynomials, etc., can be within the focal input variable, within (between) non-focal input variable(s) or between the focal and non-focal input variable(s).
-To distinguish the three, suppose the model which describes the hypothetical simulated household size simulation based on a number of socio-demographic factors such as age and wealth index is
+Higher order input variables such as interactions, splines, polynomials, etc., can be within the focal input variable, within (between) non-focal input variable(s) or between the focal and non-focal input variable(s).
+To distinguish the three, suppose the model which describes the hypothetical simulation of household size based on a number of socio-demographic factors such as age and wealth index is
 %
@@
 In this case, the non-interacting predictors are treated as non-focal predictors, with appropriate reference points.
-However, for the interacting predictors, we first choose representative values for the focal input variable and then some particular values of the interacting non-focal input variable in a similar way we choose values for the focal input variable or predetermined.
+However, for the interacting predictors, we first choose representative values for the focal input variable and then some particular values of the interacting non-focal input variable in a similar way we chose the values for the focal input variable or use predetermined values.
 In our example, suppose we pick $i$ and $j$ unique values of the focal input variable, $x_2$ and interacting input variable $x_3$, respectively.
@@
 The default anchor for generating effects is the center point.
-For simple models (or simple focal input variables), the center point is a point within the range of focal values; and therefore the confidence interval lines intersect at that that point -- B.
+For simple OLS models (or simple focal input variables), the center point is a point within the range of focal values; and therefore the confidence interval lines intersect at that that point -- B.
 In the case of higher order interactions such as cubic polynomials, the center point is not a point but a combination of all the predictor variables associated with the cubic polynomial, consequently, curves will not necessarily cross at a point -- A.
@@
 For example, if the range of values are bigger than $0.5$ (seemingly the case here), then we would expected the averages to be slightly higher than what we would expect at the center point.
+
+In Fig~\ref{fig:pred_bin_prediction_effects_plots} below, we zoom into Fig~\ref{fig:pred_bin_plots}{A} and compare the prediction and effect styled plots.
+
+\begin{figure}
+\begin{center}
+\includegraphics{glm_two_predictor_preds_effects.ggp.pdf}
+\end{center}
+\caption{{\bf Prediction and effect styled plots and bias correction.} The central curves for predictions and effects are the same in the respective case.
+These patterns are  consistent with the ones in Fig~\ref{fig:pred_bin_plots}A with respect to the \emph{truth}.
+Further, the prediction and effect CIs exhibit patterns similar to the ones we previously described and generalizes in a similar way to the ones in simple linear models.
+}
+\label{fig:pred_bin_prediction_effects_plots}
+\end{figure}
 

### 2022-06-19 manuscript/justify_preds.R

@@ -83,7 +83,7 @@ teeGG(simple_plot, desc="inter")
 ### Preds based on true parameter values
 x2_truepred_df <- varpred(justify_mod, "x2", true.beta=justify_sim_betas, bias.adjust="population", modelname="observed")$preds
 x2_focal <- quantile(justify_sim_df$x2, quants, names=FALSE)
-justify_ci_plots <- (combinepreds(justify_mod
+justify_ci_df <- (combinepreds(justify_mod
 		, c("varpred", "emmeans", "Effect")
 		, focal="x2"
 		, x.var="x2"
@@ -91,17 +91,31 @@ justify_ci_plots <- (combinepreds(justify_mod
 		, type="response"
 		, nesting=NULL
 		, ci=TRUE
+		, plotit=FALSE
 	)
-	+ geom_hline(data=true_prop_df, aes(yintercept=y, colour="observed"), lty=4)
-	+ geom_vline(xintercept=mean(x2_focal), lty=2, col="grey")
-	+ scale_color_colorblind(limits=col_limits 
-		, labels=col_labels
-	)
-#	+ geom_line(data=x2_truepred_df, aes(x=x2, y=fit), lty=4, colour="red")
-	+ guides(linetype="none")
-	+ labs(colour="Method", linetype="Method")
-	+ theme(legend.position="bottom")
 )
+pred_means_df <- (justify_ci_df
+	%>% group_by(model)
+	%>% mutate(fit=mean(fit))
+)
+
+justify_ci_plots <- (ggplot(justify_ci_df, aes(x=xvar, colour=model, linetype=model))
+	+ geom_line(aes(y=fit))
+	+ geom_line(aes(y=lwr))
+	+ geom_line(aes(y=upr))
+ 	+ geom_hline(data=true_prop_df, aes(yintercept=y, linetype="observed", colour="observed"))
+ 	+ geom_hline(data=pred_means_df, aes(yintercept=fit, linetype=model, colour=model))
+	+ geom_vline(xintercept=mean(x2_focal), lty=2, col="grey")
+ 	+ scale_linetype_discrete(limits=col_limits#[4:2]
+ 		, labels=col_labels#[4:2]
+ 	)
+ 	+ scale_color_colorblind(limits=col_limits#[4:2] 
+ 		, labels=col_labels#[4:2]
+ 	)
+	+ labs(colour="Method", linetype="Method")
+)
+
+
 teeGG(justify_ci_plots, desc="isolate")
 
 

### 2022-06-20 manuscript/body.tex
Added a plot for effects in the case of Bias

@@
 Another potential source of bias comes from the non-focal input variables in a multivariate model, especially in models with nonlinear link functions.
-Here, we describe and implement a different approach; whole-sample-based approach for bias correction.
+Here, we describe and implement a different approach -- whole-sample-based approach for bias correction.
 
@@
 We then generated two sets of mean-based predictions and effects: 1) one with \code{age} as the focal input variable (a cubic polynomial focal input variable); and 2) one with \code{Wealthindex} as the focal in put variable (linear focal input variable).
-In both cases the, the center point was used as the anchor for the effects.
-Then compared the predictions and effects as shown in Fig~\ref{fig:pred_cubic_plots}.
+In both cases the, the center point was used as the anchor for the effects, as shown in Fig~\ref{fig:pred_cubic_plots}.
 

### 2022-06-20 manuscript/abstract.tex
How is the title?

@@
 {\Large
-\textbf\newline{Describing the curves: uncertainty estimation and bias correction for predictions in simple and generalized linear (mixed) models} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+\textbf\newline{Describing the curves: uncertainty estimation and bias correction for predictions and effects in simple and generalized linear (mixed) models} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
 }

### 2022-06-21 manuscript/body.tex
Float for marginal effects
Fixing conclusion

@@
 %
-\begin{figure}[!h]
+\begin{figure}
 \centering
@@
 
-The argument and results we present in this paper supports a greater need for a shift on focus on what and how to present predictions from generalized models.
+The argument and results we present in this paper support a greater need for a shift on focus on how to present predictions from generalized models.
 For example, the effect-styled confidence intervals could provide more clarity concerning the uncertainty due to the input variable of interests as opposed to the conventional way of incorporating everything.

### 2022-06-26 manuscript/body.tex
working on plane

@@
 % Use "Eq" instead of "Equation" for equation citations.
+% Best to use a command that combines the name and the reference
 \section{Introduction}
+
+Plots of predicted values of an outcome against predictors (often called effects plots, or prediction plots) are often a useful way to summarize the results of a regression model.
+Then can be used to illustrate model uncertainty, or to give a more explicit quantitative sense of how the outcome is expected to change.
+In generalized models with a non-linear link function, or models with a spline or polynomial response to a predictor variable.
 
@@
 %
-In the absence of interactions (Equation~\ref{eq:simple_inter_higher_no_interaction}), the three approaches produce the same estimates, which match the simulated values, Fig~\ref{fig:justify_plots}A.
+In the absence of interactions (\eref{simple_inter_higher_no_interaction}), the three approaches produce the same estimates, which match the simulated values, Fig~\ref{fig:justify_plots}A.
 However, in the presence of interactions, even as simple as the one in Equation~\ref{eq:simple_inter_higher}, the estimates start to differ.

### 2022-06-26 manuscript/body.tex
hacking on airplane

@@
 Plots of predicted values of an outcome against predictors (often called effects plots, or prediction plots) are often a useful way to summarize the results of a regression model.
-Then can be used to illustrate model uncertainty, or to give a more explicit quantitative sense of how the outcome is expected to change.
-In generalized models with a non-linear link function, or models with a spline or polynomial response to a predictor variable.
+These can be used to illustrate model uncertainty, or to give a more explicit quantitative sense of how the outcome is expected to change.
+In generalized models with a non-linear link function, or models with a spline or polynomial response to a predictor variable, they can also aid in undertanding difficult-to-interpret coefficient estimates \citep{brambor_understanding_2006, berry_improving_2012}.
+\jd{Do these citations make sense where I put them?}
 
-In applications that use regression models, outcome predictions are often of interest.
-Predictions provide a great way to summarize what the regression model is telling us and very useful for interpreting and visualizing model estimates.
-For example, in logistic regression models, the coefficient estimates, when reported as log odds, are not easy to interpret and are less informative.
-Since logistic models are nonlinear (due to the nonlinear link function), in multivariate models with complex interactions, the magnitude of change in the outcome depends on the values of the input variable of interest and the other input variables, i.e., \emph{non-focal} input variables.
-In other words, the conclusions we make about the predictions greatly depends on how well we choose the values of the non-focal input variables.
+To make an outcome plot, we use a \emph{focal} predictor on the x-axis.
+The resulting plot will also depend on choices we make about other (non-focal) predictor variables.
+In ordinary (Gaussian) linear regression, the non-focal choices may have a simple additive effect on the outcome estimate.
+However, when the focal predictor has interactions, or when there is a non-linear link function, non-focal choices can also affect the slope of the response.
 
-Ordinary least squares (OLS) regression provides an easy way to express multivariate relationships as an additive linear combination of many input variables with the outcome.
-The coefficient estimates from OLS regression models are generally easy to interpret as the expected change in the outcome with a unit change in the continuous variable or as the expected outcome difference between the levels of the categorical variable with the baseline category.
-However, with additional complexities such as interactions between input variables, the resulting coefficients risk misinterpretation or interpretation complexity \citep{brambor_understanding_2006, berry_improving_2012}.
-
+\jd{How much do we need of this \P? To what extent can it be merged into the two above.}
 Simple OLS regression models are flexible and can easily be generalized, as generalized linear (mixed) models (GL(M)Ms), to examine more complex relationships, including nonlinear relationships between response and input variables e.g., higher order interactions via polynomials and splines, nonlinear transformation on the input variables, and nonlinear transformations via link functions.
@@
 Further, coefficient estimates of GLMs involving nonlinear link functions lack direct interpretation\citep{leeper2017interpreting}, meaning that interpretation of derived quantities from these estimates requires some understanding of the specified model.
+
+\jd{We need to stake out what we need by predictions and effects first.
+We could do this by first talking about the CI choices, and about univariate vs.
+multivariate?
+Or by talking about what people might want from predictions and effects first?
+We can't imply that this stuff is already known.}
+
 An alternative is to explore \emph{predictions} and \emph{effects} \citep{fox2009effect, leeper2017package, lenth2018package}.
 For example, as a way to plan, public health officials may want to know the effects of household \code{income}, \code{wealthindex}, etc, on the predicted probability that slum dwellers will have improved water services.
+
+\jd{We should discuss before continuing this part.}
+
+The outcome plots described above are often loosely called either prediction plots or effects plots.
+We endeavor here to make a conceptual distinction.
+If our goal is to \emph{predict} what we learn about the outcome variable by measuring the focal predictor,
 
@@
 
-When visually presented, predictions and effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction or some kind of transformations on the input variables whose estimates are usually, but not always, harder to interpret.
+When visually presented, predictions and effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction or some kind of transformations on the input variables whose estimates are usually, but not always, harder to interpret \jd{Keep working on reducing repetition if you can.}.
 Generating these quantities together with the associated confidence intervals for regression models has a number of challenges.
@@
 
+\jd{Not sure I agree with this sentence.
+Also, picking representative focal values seems not interesting?
+The reason I'm not sure I agree is that our method of spanning the range with a fixed number of points is not exactly a quantile method; also, what does "sized" mean here?}
 The most common way of choosing the representative values is taking unique levels of the focal input variable if discrete or taking appropriately sized quantiles if continuous.
@@
 
+\jd{This seems like too much about the other packages.
+Maybe we could: 1) stake things out a bit better by closing off our overview with the "splash plot" (make that the first figure); 2) write the other-package stuff specifically as a comparison, and make it a bit shorter.}
+
 \section{Definitions}
@@
 
+\jd{I should edit if we can finally consense on nomenclature. I feel it's my fault.}
 \begin{itemize}
@@
 For example, the regression models described by Equation~\ref{eq:simple_inter_higher_no_interaction} and Equation~\ref{eq:simple_inter_higher} both have $3$ input variables -- $x_1, x_2, x_3$.
-\item \textbf{Focal and non-focal input variables}: Focal input variable refers to the input variable we are interested in.
+\item \textbf{Focal and non-focal input variables}: The focal input variable refers to the input variable we are interested in.
 For multivariate models, all other variables other than focal input variable are non-focal input variables.
@@
 \caption{{\bf A comparison of min- and center- anchored effects.} The center-anchored effects are based on the mean of the focal predictor while the min-anchored effects are based on the minimum of the focal predictor.
-As we mentioned earlier, the choice of the anchor do not affect the main estimates (central trend lines).}
+As we mentioned earlier, the choice of the anchor do not affect the main estimates (central trend lines).
+\jd{Can we use zero anchor instead of min?}}
 \label{fig:justify_anchors}
@@
 B: The marginal effect (ME) at representative values of the focal input variable \code{age} together with the $95\%$ confidence intervals.
-For simple OLS models, the ME is constant across all the values of \code{age} chosen.}
+For simple OLS models, the ME is constant across all the values of \code{age} chosen.
+\jd{The definitions should be shorter and it's definitely weird to have a figure in the definitions.
+Maybe you should consider making a box?
+Also, if we're going to have this figure, we should try to illustrate a more difficult example for comparison.}}
 \label{fig:qoi_age_pred_plot}
@@
 
+\jd{Are we kind of going backward now from the comparison section? Should all of these definitions come first?}
+
+\jd{This \P\ seems unnecessary. Why not just say “To illustrate \ldots", and then jump to the next \P (“, we consider")}
 To illustrate mean-based approach, suppose we are interested in predictions at the levels of a particular focal input variable, $x_f$, from a multivariate model.
@@
 
-\subsection{Dealing with higher order interactions}
+\subsection{Dealing with higher-order terms}
+
+\jd{I'm having trouble staking out higher-order terms as a thing.
+Yes, many times some or all of the terms in a multi-parameter predictor can be described as higher order, but not always (e.g., categorical predictors).
+Also, why are we interested in things that don't involve the focal predictor (and why only interactions?).
+I know that we have to average properly, but we should do that once.
+Does it make sense to talk separately about interactions, and multi-parameter focal predictors?}
 
 Higher order input variables such as interactions, splines, polynomials, etc., can be within the focal input variable, within (between) non-focal input variable(s) or between the focal and non-focal input variable(s).
-To distinguish the three, suppose the model which describes the hypothetical simulation of household size based on a number of socio-demographic factors such as age and wealth index is
+To distinguish the three \jd{Right?
+Why are there three?
+There should be two or four.}, suppose the model which describes the hypothetical simulation of household size based on a number of socio-demographic factors such as age and wealth index is
 %
@@
 In general, our formulation, even for more complicated interactions, follow these three basic principles -- interaction within focal input variables, interaction between non-focal input variables and interaction between focal and non-focal input variables.
+\jd{We should talk about what this s. means. What are the principles?}
 
@@
 Commonly used \proglang{R} packages for constructing predictions do not exclude the uncertainties resulting from non focal input variables when computing the CIs.
-Currently obscure way to exclude uncertainties associated with non-focal input variables in some of these packages is to provide a user defined variance-covariance matrix with the covariances of non-focal predictors set to $0$.
+\jd{We should first explain what we want to do, and then do our comparisons with existing packages.} Currently obscure way to exclude uncertainties associated with non-focal input variables in some of these packages is to provide a user defined variance-covariance matrix with the covariances of non-focal predictors set to $0$.
 We refer to this procedure as \emph{zeroing-out} variance-covariance matrix; and only works when the input variables are \emph{centered} prior to model fitting, in case of numerical variables, and more complicated when the input variables are categorical.
@@
 
+\jd{We need to explain why we think this is a problem, not just state that it is “usually important".}
 It is usually important to report the estimates that reflect the expected values of the untransformed response.
@@
 Here, we describe and implement a different approach -- whole-sample-based approach for bias correction.
+\jd{I don't understand how this \P\ describes two separate problems.
+What source of bias is being described in the first case?
+Are you sure it exists?
+What does Lenth say about the bias they correct for?}
 
@@
 
-As mentioned above, the two common choices to generate predictions are: 1) using mean of the non-focal predictors as the reference point -- mean-based; and 2) using the entire observed values of non-focal predictors -- whole-sample-based.
-If the link function is nonlinear and/or the model has complex higher order interactions, the average of the predictions evaluated at the mean of non-focal predictors and the average of predictions evaluated at the population level (and then averaged) of non-focal predictor are not equivalent, i.e.,
+\jd{Why is whole-sample common? Does anyone do it at all except us?}
+
+We can directly compare the outcome estimate at the model center (mean-based estimate) with the average of predictions evaluated across the population values of non-focal predictors (sample-based estimate):
 
 \begin{align}\label{eq:compare_anchored_pop_based}
-g^{-1} \left(\eta_j^\star(\bar{x}_f, \nset{\bar{x}})\right) \neq \frac{1}{n} \sum_{i=1}^n{ g^{-1} \left(\eta_j^\star(\bar{x}_f, \nset{x})\right)}.
+g^{-1} \left(\eta_j^\star(\bar{x}_f, \nset{\bar{x}})\right) : \frac{1}{n} \sum_{i=1}^n{ g^{-1} \left(\eta_j^\star(\bar{x}_f, \nset{x})\right)}.
 \end{align}
 
-If $g()$ is strictly concave or strictly convex, one can use Jensen's inequality to determine which side of Equation~\ref{eq:compare_anchored_pop_based} is decelerating or accelerating.
+In an ordinary linear model, the link function $g$ is the identity function, so the two means are the same. For non-trivial links, we expect them to be different in general. 
+\jd{What are the key ideas if any from the rest of this \P?}.
+I would like to clearly describe what happens with the log link, and also what happens near each end in a logit link.
+Why is this a place to talk about the explicit Taylor?}
+strictly concave or strictly convex, one can use Jensen's inequality to determine which side of Equation~\ref{eq:compare_anchored_pop_based} is decelerating or accelerating.
 For accelerating functions, the Jensen’s inequality describes how the changes in input values elevate the predictions and otherwise describes how these changes depress the predictions.
@@
 
-As stated above, in simple linear models without interactions, the effect is constant, so both approaches yield similar results.
+\jd{What does this mean?
+I was going to say that $g$ collapses out of \eqref{compare_anchored_pop_based} and the results are identical.
+Is that wrong?} As stated above, in simple linear models without interactions, the effect is constant, so both approaches yield similar results.
 However, picking a single value, e.g., mean of the predictor, on which to draw conclusions about the effect can be problematic, unrealistic or not contained in or representative of the population.
@@
 
+\jd{I meant to say earlier that we need to talk about the common “sum-to-zero-like" method of averaging over dsitnct values of a categorical input variable vs.~the true model-center method.
+This might be another place where we can show we do better than some existing packages!
+See notes.}
 Another potential concern with the mean-based approach arises in situations where direct naive use leads to rare or meaningless basis for generalization.

### 2022-06-27 manuscript/body.tex
Big RSA arrival merge

@@
 In an ordinary linear model, the link function $g$ is the identity function, so the two means are the same. For non-trivial links, we expect them to be different in general. 
-\jd{What are the key ideas if any from the rest of this \P?}.
+\jd{What are the key ideas if any from the rest of this \P?.
 I would like to clearly describe what happens with the log link, and also what happens near each end in a logit link.
@@
 \jd{What does this mean?
-I was going to say that $g$ collapses out of \eqref{compare_anchored_pop_based} and the results are identical.
+I was going to say that $g$ collapses out of \eref{compare_anchored_pop_based} and the results are identical.
 Is that wrong?} As stated above, in simple linear models without interactions, the effect is constant, so both approaches yield similar results.

### 2022-06-28 manuscript/body.tex
Working on JD's comments

@@
 These can be used to illustrate model uncertainty, or to give a more explicit quantitative sense of how the outcome is expected to change.
-In generalized models with a non-linear link function, or models with a spline or polynomial response to a predictor variable, they can also aid in undertanding difficult-to-interpret coefficient estimates \citep{brambor_understanding_2006, berry_improving_2012}.
+In generalized models with a non-linear link function, or models with a spline or polynomial response to a predictor variable, they can also aid in undertanding difficult-to-interpret coefficient estimates \citep{brambor_understanding_2006, berry_improving_2012, leeper2017interpreting}.
 \jd{Do these citations make sense where I put them?}
@@
 However, when the focal predictor has interactions, or when there is a non-linear link function, non-focal choices can also affect the slope of the response.
+Additional challenges arise when dealing with generalized linear (mixed) models (GL(M)Ms) which extend ordinary linear models to incorporate random effects.
+How we treat (include or exclude) the random effects also affect the slope and nature of the response.
 
 \jd{How much do we need of this \P? To what extent can it be merged into the two above.}
-Simple OLS regression models are flexible and can easily be generalized, as generalized linear (mixed) models (GL(M)Ms), to examine more complex relationships, including nonlinear relationships between response and input variables e.g., higher order interactions via polynomials and splines, nonlinear transformation on the input variables, and nonlinear transformations via link functions.
+
+\BC{Added a sentence in the last \P. I think we can delete the \P}
+
+\dya{Simple OLS regression models are flexible and can easily be generalized, as generalized linear (mixed) models (GL(M)Ms), to examine more complex relationships, including nonlinear relationships between response and input variables e.g., higher order interactions via polynomials and splines, nonlinear transformation on the input variables, and nonlinear transformations via link functions.
 This flexibility comes with additional complications, for example, complex multivariate models may risk misinterpretation, and miscalculation of quantities of interest \citep{berry_improving_2012, leeper2017interpreting}.
-Further, coefficient estimates of GLMs involving nonlinear link functions lack direct interpretation\citep{leeper2017interpreting}, meaning that interpretation of derived quantities from these estimates requires some understanding of the specified model.
+Further, coefficient estimates of GLMs involving nonlinear link functions lack direct interpretation\citep{leeper2017interpreting}, meaning that interpretation of derived quantities from these estimates requires some understanding of the specified model.}
 
@@
 
-An alternative is to explore \emph{predictions} and \emph{effects} \citep{fox2009effect, leeper2017package, lenth2018package}.
-For example, as a way to plan, public health officials may want to know the effects of household \code{income}, \code{wealthindex}, etc, on the predicted probability that slum dwellers will have improved water services.
+\dya{An alternative is to explore \emph{predictions} and \emph{effects} \citep{fox2009effect, leeper2017package, lenth2018package}.
+For example, as a way to plan, public health officials may want to know the effects of household \code{income}, \code{wealthindex}, etc, on the predicted probability that slum dwellers will have improved water services.}
 
 \jd{We should discuss before continuing this part.}
-
+\BC{Okay. Some thoughts below your \P}
 The outcome plots described above are often loosely called either prediction plots or effects plots.
 We endeavor here to make a conceptual distinction.
-If our goal is to \emph{predict} what we learn about the outcome variable by measuring the focal predictor,
+If our goal is to \emph{predict} what we learn about the outcome variable by measuring the focal predictor, \textcolor{red}{then we would focus on the central (outcome) estimate which is the same for both prediction and effects plots.
+It simply tells us what we expect the outcome to be for any given value of the focal input variable.
+We may also want to provide uncertainty associated with the central estimate.
+How we choose to describe the uncertainty can inform the choice we make between prediction and effects plots.
+For instance, we may want to capture the uncertainty in the whole model (these include uncertainty due to intercept, focal and non-focal predictors, and random effects) or we may be interested in showing the uncertainty associated with the focal predictor only.
+In the former case, we use the prediction plots which generally focus on the \emph{total} effect, and so will often be suited for a univariate model.
+On the other hand, in the latter case, we introduce effects plots which focus on the direct effect of the focal predictor -- after controlling for other non-focal input variables -- from a multivariate model, and so will be preferable for multivariate models.
+Since effects plots focus on focal-specific uncertainty, they are usually narrower than prediction plots.} \BC{Not sure whether to say that uncertainties are described by the CIs?}
 
-Both predictions and effects tell us what we expect the response to be for a given value of \emph{focal} input variable (scientific variable of interest).
+\dya{Both predictions and effects tell us what we expect the response to be for a given value of \emph{focal} input variable (scientific variable of interest).
 The major difference between the two lies in the construction of the associated confidence intervals.
@@
 On the other hand, effects, attempt to show the effect of a focal input variable.
-In this case, we suggest the narrower confidence intervals which focus on uncertainty in the effect of the focal input variable only, and so will often be suited for the direct effects – after controlling for other non-focal input variables – from a multivariate model.
+In this case, we suggest the narrower confidence intervals which focus on uncertainty in the effect of the focal input variable only, and so will often be suited for the direct effects – after controlling for other non-focal input variables – from a multivariate model.}
 
-When visually presented, predictions and effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction or some kind of transformations on the input variables whose estimates are usually, but not always, harder to interpret \jd{Keep working on reducing repetition if you can.}.
-Generating these quantities together with the associated confidence intervals for regression models has a number of challenges.
-In particular:
+\dya{When visually presented, predictions and effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction or some kind of transformations on the input variables whose estimates are usually, but not always, harder to interpret \jd{Keep working on reducing repetition if you can.}.}
+
+Generating quantities, i.e., central estimate together with the corresponding confidence intervals, associated  prediction and effects plots has a number of challenges. In particular:
 \begin{enumerate}
@@
 The reason I'm not sure I agree is that our method of spanning the range with a fixed number of points is not exactly a quantile method; also, what does "sized" mean here?}
-The most common way of choosing the representative values is taking unique levels of the focal input variable if discrete or taking appropriately sized quantiles if continuous.
+\BC{I somehow agree it is not very interesting but I think we need to mention what we do.
+We do both -- spanning a range of values and quantiles of the observed focal values (our default).
+"sized" is supposed to the number of sampled data points.
+Maybe I should just say we take quantiles of the focal predictor?.}
+The most common way of choosing the representative values is taking unique levels or quantiles of the focal input variable if discrete or continuous, respectively.
 The predictions are then calculated by holding the non-focal input variables at their reference point (values chosen for the non-focal input variables, e.g., means), while varying the focal input variable, with the goal that the estimates represent how the model responds to the changes in the focal input variable \citep{fox2009effect, hanmer2013behind}.

### 2022-06-28 manuscript/body.tex
Still working on JD's feedback

@@
 We endeavor here to make a conceptual distinction.
-If our goal is to \emph{predict} what we learn about the outcome variable by measuring the focal predictor, \textcolor{red}{then we would focus on the central (outcome) estimate which is the same for both prediction and effects plots.
+If our goal is to \emph{predict} what we learn about the outcome variable by measuring the focal predictor, \NEW{then we would focus on the central (outcome) estimate which is the same for both prediction and effects plots.
 It simply tells us what we expect the outcome to be for any given value of the focal input variable.
@@
 Maybe I should just say we take quantiles of the focal predictor?.}
+
 The most common way of choosing the representative values is taking unique levels or quantiles of the focal input variable if discrete or continuous, respectively.
@@
 This procedure generates -- \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}.
-In this article, we refer to these quantities as predictions and effects since they should, for example, tell us what we would expect the response to be at a particular value or level of the focal input variable, for an ``average case''.
+\BC{We need to decide on how to proceed from here. I think we can refer to this as the central estimate. Re-writing and we can delete older version if that's okay.}
+\NEW{In this article, we refer to these quantities as the central (outcome) estimates for an ``average case'' in a population; and the associated confidence intervals will either be a prerequisite for prediction or effects plots.
+In the case of effects plots, we can also choose an anchor (value chosen for the focal input variable).
+The concept of effects plots has been presented in the literature -- Shi et al.
+\citep{shi_evidence_2017} used effects plots to compare the difference in sexual risk behaviors between circumcised and uncircumcised men.}
+\dya{In this article, we refer to these quantities as predictions and effects since they should, for example, tell us what we would expect the response to be at a particular value or level of the focal input variable, for an ``average case''.
 As mentioned above, in our case, the key difference between predictions and effects is in the computation of the confidence intervals.
 The concept of effects has been presented in the literature -- Shi et al.
-\citep{shi_evidence_2017} used effect-styled plots to compare the difference in sexual risk behaviors between circumcised and uncircumcised men.
+\citep{shi_evidence_2017} used effect-styled plots to compare the difference in sexual risk behaviors between circumcised and uncircumcised men.}
 
-The most commonly used \proglang{R} software packages for generating predictions (\pkg{emmeans} and \pkg{effects}), by default, use the average of \emph{input} and \emph{predictor variables} as the reference point for continuous and categorical input variables, respectively.
+\BC{I think we should move all with "MOVE" tag to somewhere after the statistical background but Ben suggested this format. I think moving everthing downstream will streamline the flow.}
+\MOVE{The most commonly used \proglang{R} software packages for generating predictions (\pkg{emmeans} and \pkg{effects}), by default, use the average of \emph{input} and \emph{predictor variables} as the reference point for continuous and categorical input variables, respectively.
 However, there are a number of choices one can make when generating predictions -- for example, in the presence of interactions, averaging the interactions (averaging the product of interacting predictor variables) versus taking the product of the averages of the interacting predictor variables (the default for \pkg{emmeans} and \pkg{effects}).
@@
 %
+}
 
-We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$, $\beta_{23} = 5$, $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and $\sigma^2 = 1$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions (predictions based on the true simulation parameters) and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.
+\MOVE{We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$, $\beta_{23} = 5$, $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and $\sigma^2 = 1$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions (predictions based on the true simulation parameters) and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.
 %
@@
 In the case of Fig~\ref{fig:justify_plots}A, averaging the input variables is equivalent to averaging the predictor variables.
+}
 
-One may be interested in the uncertainties associated with the focal input variable only, excluding other uncertainties due to other non-focal input variables -- effects.
+\MOVE{One may be interested in the uncertainties associated with the focal input variable only, excluding other uncertainties due to other non-focal input variables -- effects.
 However, currently, the two packages do not provide a straightforward way do achieve this.
 To illustrate this, we generate predictions of $x_2$ from Equation~\ref{eq:simple_inter_higher_no_interaction} together with the associated $95\%$ confidence intervals, as shown in Fig~\ref{fig:justify_ci_plots}.
-
+%
 \begin{figure}
@@
 \end{figure}
-
+%
 For \pkg{emmeans} and \pkg{effects}, the confidence intervals are much wider because they include uncertainties associated with the intercept and non-focal input variables, but narrower and crosses at the mean of the focal input variable in \pkg{varpred}.
@@
 Unlike predictions, where the uncertainty captures the uncertainty in the whole model, the effects plot focuses on uncertainty in the effect of the focal predictor only and depend on the anchor.
+}
 

### 2022-06-29 manuscript/body.tex
Will JD be available?

@@
 Maybe we could: 1) stake things out a bit better by closing off our overview with the "splash plot" (make that the first figure); 2) write the other-package stuff specifically as a comparison, and make it a bit shorter.}
+\BC{By ``splash plot'' you mean Fig 1? I am suggesting we move all the package comparison stuff to/after statistical background section or maybe to simulation example section ...}
 
@@
 \end{center}
-\caption{{\bf A comparison of min- and center- anchored effects.} The center-anchored effects are based on the mean of the focal predictor while the min-anchored effects are based on the minimum of the focal predictor.
-As we mentioned earlier, the choice of the anchor do not affect the main estimates (central trend lines).
-\jd{Can we use zero anchor instead of min?}}
+\caption{{\bf A comparison of zero- and center- anchored effects.} The center-anchored effects are based on the mean of the focal predictor while the zero-anchored effects are based on the minimum of the focal predictor.
+As we mentioned earlier, the choice of the anchor do not affect the main estimates (central trend lines).}
 \label{fig:justify_anchors}

### 2022-06-29 manuscript/body.tex
Pausing a little, to meet JD?

@@
 \caption{{\bf A comparison of zero- and center- anchored effects.} The center-anchored effects are based on the mean of the focal predictor while the zero-anchored effects are based on the minimum of the focal predictor.
-As we mentioned earlier, the choice of the anchor do not affect the main estimates (central trend lines).}
+As we mentioned earlier, the choice of the anchor do not affect the main estimates (central trend lines).
+\MOVE{Should me move this downstream}}
 \label{fig:justify_anchors}
@@
 The \proglang{R} software package \pkg{margins} \cite{lenth2018package} generates marginal effects.
-Consider result from a hypothetical  simulated example -- regression of household size as function of household wealth index and age of household head, as shown in Fig~\ref{fig:qoi_age_pred_plot}.
-Since the model has no interactions, the relationship between the predicted household size and age is linear, hence the marginal effect of age is the slope ($\frac{\Delta \mathrm{hh size}}{\Delta \mathrm{age}}$) of the prediction (equivalently effect) line and can be calculated irrespective of the values of wealth index.
+To illustrate the distinction between prediction and effects plots, and marginal effects, we use the simulated data described in \nameref{S1_Appendix}.
 %
@@
 \centering
-\includegraphics[width=0.8\textwidth]{quantity_of_interest.ggp.pdf}
-\caption{{\bf A comparison of prediction, effect and marginal effect.} A: The black solid line is the prediction (same as effect), while the dashed lines indicate the changes along the \code{age} and \code{household size} axis.
-For a linear model with no interaction, the marginal effect is the slope of the prediction line.
+\includegraphics{cubic_varpred_margins_pred.ggp.pdf}
+\caption{{\bf A comparison of effects and marginal effect plots.} A:  The black solid line is the central (outcome) estimate, while the dashed lines corresponds to the effects styled $95\%$ confidence interval.
 B: The marginal effect (ME) at representative values of the focal input variable \code{age} together with the $95\%$ confidence intervals.
-For simple OLS models, the ME is constant across all the values of \code{age} chosen.
+For simple linear models with no interactions, the ME should be constant across all the values of the focal input variable; and it would correspond to the slope of the central estimate of the prediction or effects.
+However, in this example, the focal input variable is a cubic polynomial, and so, the effect is not constant.
 \jd{The definitions should be shorter and it's definitely weird to have a figure in the definitions.
 Maybe you should consider making a box?
-Also, if we're going to have this figure, we should try to illustrate a more difficult example for comparison.}}
+Also, if we're going to have this figure, we should try to illustrate a more difficult example for comparison.}\BC{Is this better example?} \MOVE{I think this should also be moved downstream.} \BC{What should we call the central and/or y-axis line of the effect/prediction plot?}}
 \label{fig:qoi_age_pred_plot}
@@
 \jd{Are we kind of going backward now from the comparison section? Should all of these definitions come first?}
+\BC{I agree. We can move all the package comparisons to simulation section. \MOVE{}}
 
 \jd{This \P\ seems unnecessary. Why not just say “To illustrate \ldots", and then jump to the next \P (“, we consider")}
-To illustrate mean-based approach, suppose we are interested in predictions at the levels of a particular focal input variable, $x_f$, from a multivariate model.
-As a starting point, assume that the model has no interaction terms.
-The idea is to choose a reference point for the values of non-focal input variables.
-For example, fixing the values of non-focal predictor(s) at some typical values -- typically determined by averaging in some meaningful way, for example, arithmetic mean  for continuous and average over the levels of the factors for categorical non-focal predictors.
-We refer to this as centered or mean-based approach.
-The most convenient way to achieve this is by constructing centered model matrix, $\bX^\star$, by averaging the columns of non-focal predictors in model matrix $\bX$, and together with appropriately chosen values of focal predictor.
-
-Consider a simple linear model with linear predictor $\eta = \bX\bbeta$ and let $g(\boldmu) = \boldeta$ be an identity link function, where $\boldmuh$ is the expected value of response variable $\hat{y}$.
+\BC{How is this?}
+To illustrate mean-based approach, consider a simple linear model with linear predictor $\eta = \bX\bbeta$ and let $g(\boldmu) = \boldeta$ be an identity link function, where $\boldmuh$ is the expected value of response variable $\hat{y}$.
 Let $\bbetah$ be the estimate of $\bbeta$, together with the estimated covariance matrix $\Sigma = V(\bbetah)$ of $\bbetah$.
-Then the quantity $\hat{\boldeta}^\star = \bX^\star\bbetah$ is the prediction with respect to the focal input variable in question \citep{fox2009effect}.
+Then the quantity $\hat{\boldeta}^\star = \bX^\star\bbetah$ is the \fix{central (outcome) estimate} with respect to the focal input variable in question \citep{fox2009effect}; where $\bX^\star$ is the centered model matrix constructed by averaging the columns of non-focal predictors in model matrix $\bX$, and together with appropriately chosen values of focal predictor.
 
@@
 
+\BC{Should we just be more specific and say "Dealing with polynomial input variables"?}
+\BC{JD to explain second part ...}
+
 Higher order input variables such as interactions, splines, polynomials, etc., can be within the focal input variable, within (between) non-focal input variable(s) or between the focal and non-focal input variable(s).
@@
 Why are there three?
-There should be two or four.}, suppose the model which describes the hypothetical simulation of household size based on a number of socio-demographic factors such as age and wealth index is
+There should be two or four.} \BC{Three: 1) on the focal, 2) on the non-focal, and 3) between focal and non-focal.
+Which one is redundant or missing?}, suppose the model which describes the hypothetical simulation of household size based on a number of socio-demographic factors such as age and wealth index is
 %
@@
 \jd{We should talk about what this s. means. What are the principles?}
+\BC{DISCUSS with JD}
 
@@
 
-Commonly used \proglang{R} packages for constructing predictions do not exclude the uncertainties resulting from non focal input variables when computing the CIs.
-\jd{We should first explain what we want to do, and then do our comparisons with existing packages.} Currently obscure way to exclude uncertainties associated with non-focal input variables in some of these packages is to provide a user defined variance-covariance matrix with the covariances of non-focal predictors set to $0$.
+We may be interested in the uncertainties associated with the focal input variable only, excluding other uncertainties due to other non-focal input variables -- effects.
+However, commonly used \proglang{R} packages for constructing predictions do not exclude the uncertainties resulting from non focal input variables when computing the CIs.
+\jd{We should first explain what we want to do, and then do our comparisons with existing packages.} \BC{How about the new version?} Currently obscure way to exclude uncertainties associated with non-focal input variables in some of these packages is to provide a user defined variance-covariance matrix with the covariances of non-focal predictors set to $0$.
 We refer to this procedure as \emph{zeroing-out} variance-covariance matrix; and only works when the input variables are \emph{centered} prior to model fitting, in case of numerical variables, and more complicated when the input variables are categorical.

### 2022-06-29 manuscript/cubic_predictors_model.R

@@ -20,7 +20,11 @@ true_prop_df <- (sim_df_cubic
 print(true_prop_df)
 
 ## Models
-mod_cubic <- lm(hhsize ~ poly(age, degree=3, raw=TRUE)+wealthindex
+# mod_cubic <- lm(hhsize ~ poly(age, degree=3, raw=TRUE)+wealthindex
+# 	, data=sim_df_cubic
+# )
+## Using I() because margins doesn't work with poly()
+mod_cubic <- lm(hhsize ~ I(age)+I(age^2)+I(age^3)+wealthindex
 	, data=sim_df_cubic
 )
 summary(mod_cubic)

### 2022-06-29 manuscript/cubic_varpred_margins_pred.R

@@ -2,31 +2,34 @@ library(shellpipes)
 library(vareffects); varefftheme()
 library(ggpubr)
 library(ggplot2)
-library(ggthemes)
-library(dplyr)
 library(margins)
-library(emmeans)
-library(effects)
 
 loadEnvironments()
 startGraphics()
 
-
 ## marginal effect
-m <- lm(hhsize ~I(age)+I(age^2)+I(age^3)+wealthindex, data=sim_df_cubic)
-d1 <- cplot(m, "age", what="effect")
-d2 <- data.frame(d1)
-d2
-ggplot(d2, aes(x=xvals, y=yvals)) +  geom_line() + geom_point(data=sim_df_cubic, aes(x=age, y=hhsize))
-p1 <- varpred(m, "age", isolate=FALSE, at=list(age=d2$xvals), bias.adjust="none", returnall=TRUE)
-plot(p1) + geom_point(data=sim_df_cubic, aes(x=age, y=hhsize))
-m1 <- emmeans(m, specs=~age, at=list(age=d2$xvals), nesting=NULL)
-as.data.frame(m1)
-ef <- Effect("age", m, xlevels=list(age=d2$xvals))
-p1
-as.data.frame(ef)
-head(p1$raw$model.matrix)
-head(ef$model.matrix)
-quit()
-meffect_df <- cplot(mod_cubic, "age", what = "effect", draw=FALSE)
-meffect_df
+meffect_df <- cplot(mod_cubic, "age", what="effect", draw=FALSE)
+meffect_df <- data.frame(meffect_df)
+meffect_df$model <- "B) Marginal effect"
+meffect_plot <- (ggplot(meffect_df, aes(x=xvals))
+	+ geom_line(aes(y=yvals))
+	+ geom_line(aes(y=lower), lty=2)
+	+ geom_line(aes(y=upper), lty=2)
+	+ labs(x="age", y="ME of age", title="B) Marginal effect")
+)
+
+## varpred
+varpred_df <- varpred(mod_cubic, "age", steps=500, modelname="A) Prediction and effects")
+varpred_plot <- (plot(varpred_df)
+	+ labs(y="Predicted household size", title="A) Prediction and effects")
+)
+
+## Combine all prediction for faceting
+pred_plots <- ggarrange(varpred_plot
+	, meffect_plot
+	, common.legend=TRUE
+	, legend="bottom"
+	, ncol=2
+)
+
+teeGG(pred_plots)

### 2022-06-30 manuscript/body.tex
To meet with JD...?

@@
 \includegraphics{cubic_varpred_margins_pred.ggp.pdf}
-\caption{{\bf A comparison of effects and marginal effect plots.} A:  The black solid line is the central (outcome) estimate, while the dashed lines corresponds to the effects styled $95\%$ confidence interval.
+\caption{{\bf A comparison of effects and marginal effect plots.} \NEW{A:  The black solid line is the central (outcome) estimate, while the dashed lines corresponds to the effects styled $95\%$ confidence interval.
 B: The marginal effect (ME) at representative values of the focal input variable \code{age} together with the $95\%$ confidence intervals.
 For simple linear models with no interactions, the ME should be constant across all the values of the focal input variable; and it would correspond to the slope of the central estimate of the prediction or effects.
-However, in this example, the focal input variable is a cubic polynomial, and so, the effect is not constant.
-\jd{The definitions should be shorter and it's definitely weird to have a figure in the definitions.
+However, in this example, the focal input variable is a cubic polynomial, and so, the effect is not constant.} \jd{The definitions should be shorter and it's definitely weird to have a figure in the definitions.
 Maybe you should consider making a box?
-Also, if we're going to have this figure, we should try to illustrate a more difficult example for comparison.}\BC{Is this better example?} \MOVE{I think this should also be moved downstream.} \BC{What should we call the central and/or y-axis line of the effect/prediction plot?}}
+Also, if we're going to have this figure, we should try to illustrate a more difficult example for comparison.} \MOVE{I think this should also be moved downstream.} \BC{What should we call the central and/or y-axis line of the effect/prediction plot?}}
 \label{fig:qoi_age_pred_plot}
@@
 \jd{This \P\ seems unnecessary. Why not just say “To illustrate \ldots", and then jump to the next \P (“, we consider")}
-\BC{How is this?}
-To illustrate mean-based approach, consider a simple linear model with linear predictor $\eta = \bX\bbeta$ and let $g(\boldmu) = \boldeta$ be an identity link function, where $\boldmuh$ is the expected value of response variable $\hat{y}$.
+
+\NEW{To illustrate mean-based approach, consider a simple linear model with linear predictor $\eta = \bX\bbeta$ and let $g(\boldmu) = \boldeta$ be an identity link function, where $\boldmuh$ is the expected value of response variable $\hat{y}$.
 Let $\bbetah$ be the estimate of $\bbeta$, together with the estimated covariance matrix $\Sigma = V(\bbetah)$ of $\bbetah$.
 Then the quantity $\hat{\boldeta}^\star = \bX^\star\bbetah$ is the \fix{central (outcome) estimate} with respect to the focal input variable in question \citep{fox2009effect}; where $\bX^\star$ is the centered model matrix constructed by averaging the columns of non-focal predictors in model matrix $\bX$, and together with appropriately chosen values of focal predictor.
+}
 
@@
 
-We may be interested in the uncertainties associated with the focal input variable only, excluding other uncertainties due to other non-focal input variables -- effects.
-However, commonly used \proglang{R} packages for constructing predictions do not exclude the uncertainties resulting from non focal input variables when computing the CIs.
-\jd{We should first explain what we want to do, and then do our comparisons with existing packages.} \BC{How about the new version?} Currently obscure way to exclude uncertainties associated with non-focal input variables in some of these packages is to provide a user defined variance-covariance matrix with the covariances of non-focal predictors set to $0$.
+\NEW{We may be interested in the uncertainties associated with the focal input variable only, excluding other uncertainties due to other non-focal input variables -- effects.
+However, commonly used \proglang{R} packages for constructing predictions do not exclude the uncertainties resulting from non focal input variables when computing the CIs.} \jd{We should first explain what we want to do, and then do our comparisons with existing packages.} Currently obscure way to exclude uncertainties associated with non-focal input variables in some of these packages is to provide a user defined variance-covariance matrix with the covariances of non-focal predictors set to $0$.
 We refer to this procedure as \emph{zeroing-out} variance-covariance matrix; and only works when the input variables are \emph{centered} prior to model fitting, in case of numerical variables, and more complicated when the input variables are categorical.

### 2022-07-01 manuscript/body.tex
eref conflicts eref in the thesis, changed to EREF

@@
 %
-In the absence of interactions (\eref{simple_inter_higher_no_interaction}), the three approaches produce the same estimates, which match the simulated values, Fig~\ref{fig:justify_plots}A.
+In the absence of interactions (\EREF{simple_inter_higher_no_interaction}), the three approaches produce the same estimates, which match the simulated values, Fig~\ref{fig:justify_plots}A.
 However, in the presence of interactions, even as simple as the one in Equation~\ref{eq:simple_inter_higher}, the estimates start to differ.
@@
 \jd{What does this mean?
-I was going to say that $g$ collapses out of \eref{compare_anchored_pop_based} and the results are identical.
+I was going to say that $g$ collapses out of \EREF{compare_anchored_pop_based} and the results are identical.
 Is that wrong?} As stated above, in simple linear models without interactions, the effect is constant, so both approaches yield similar results.

### 2022-07-01 manuscript/body.tex
A little rest.

@@
 
-\jd{We need to explain why we think this is a problem, not just state that it is “usually important".}
-It is usually important to report the estimates that reflect the expected values of the untransformed response.
+\jd{Why is whole-sample common? Does anyone do it at all except us?}
+
+We can directly compare the outcome estimate at the model center (mean-based estimate) with the average of predictions evaluated across the population values of non-focal predictors (sample-based estimate):
+
+\begin{align}\label{eq:compare_anchored_pop_based}
+g^{-1} \left(\eta_j^\star(\bar{x}_f, \nset{\bar{x}})\right) : \frac{1}{n} \sum_{i=1}^n{ g^{-1} \left(\eta_j^\star(\bar{x}_f, \nset{x})\right)}.
+\end{align}
+
+In an ordinary linear model, the link function $g$ is the identity function, so the two means are the same.
+For non-trivial link functions, we expect them to be different in general -- \NEW{a consequent of Jensen inequality \citep{duursma_bias_2003, denny_performance_2019, ye_generalized_2021}.
+In GL(M)Ms the assumed link function, predetermines the curvature of the response surface and also the sign of the Jensen effect.
+For example, logistic model has a positive curvature if the predicted probabilities are low and negative curvature if the predicted probabilities are high.
+In other words, logistic model may be concave or convex depending on the predicted probabilities, hence the sign of the Jensen effect is not readily available.
+However, in log link functions, the inverse of the link (exponential) function is strictly convex, i.e., an accelerating function.
+For accelerating functions, the Jensen’s inequality describes how the changes in input values elevate the predictions and otherwise describes how these changes depress the predictions.
+Further, our models may include input variables and/or random effects terms other than the focal one.
+In non-trivial link functions, the Jensen effect may depend on the additional input variables \citep{ye_generalized_2021}}.
+
+\jd{We need to explain why we think this is a problem, not just state that it is “usually important".} \BC{See previous \P}
+
+\NEW{It is usually important to report the estimates that reflect the expected values of the untransformed response.
 This is not a major issue in simple linear models.
-However, when dealing with nonlinear link functions,  generated predictions may not reflect the observed response due to the bias in the expected mean induced by the nonlinear transformation of the response variable.
-In such cases, bias correction is needed when back-transforming the predictions to the original scales.
-The common approach for bias-adjustment is second-order Taylor approximation \citep{lenth2018package, duursma2003bias}.
-Another potential source of bias comes from the non-focal input variables in a multivariate model, especially in models with nonlinear link functions.
-Here, we describe and implement a different approach -- whole-sample-based approach for bias correction.
-\jd{I don't understand how this \P\ describes two separate problems.
-What source of bias is being described in the first case?
-Are you sure it exists?
-What does Lenth say about the bias they correct for?}
+However, when dealing with nonlinear link functions,  generated predictions may not reflect the observed response due to the bias in the expected mean induced by the nonlinear transformation of the response variable -- a consequent of Jensen effect mentioned above.
+In such cases, bias correction is needed to back-transform the predictions to the original scales.
+The common approach for bias-adjustment is second-order Taylor approximation \citep{duursma2003bias, hanmer2013behind}; already implemented in \pkg{emmeans} \citep{lenth2018package}.
+Here, we describe and implement a different approach -- whole-sample-based approach for bias correction.}
 
@@
 \section{Mean-based vs. whole-sample-based}
-
-\jd{Why is whole-sample common? Does anyone do it at all except us?}
-
-We can directly compare the outcome estimate at the model center (mean-based estimate) with the average of predictions evaluated across the population values of non-focal predictors (sample-based estimate):
-
-\begin{align}\label{eq:compare_anchored_pop_based}
-g^{-1} \left(\eta_j^\star(\bar{x}_f, \nset{\bar{x}})\right) : \frac{1}{n} \sum_{i=1}^n{ g^{-1} \left(\eta_j^\star(\bar{x}_f, \nset{x})\right)}.
-\end{align}
-
-In an ordinary linear model, the link function $g$ is the identity function, so the two means are the same. For non-trivial links, we expect them to be different in general. 
-\jd{What are the key ideas if any from the rest of this \P?.
-I would like to clearly describe what happens with the log link, and also what happens near each end in a logit link.
-Why is this a place to talk about the explicit Taylor?}
-strictly concave or strictly convex, one can use Jensen's inequality to determine which side of Equation~\ref{eq:compare_anchored_pop_based} is decelerating or accelerating.
-For accelerating functions, the Jensen’s inequality describes how the changes in input values elevate the predictions and otherwise describes how these changes depress the predictions.
-For a non-fully convex or concave link function, one can use Taylor series expansion to approximate the range over which the two are smaller or smaller than the other and then apply second-order approximation \citep{hanmer2013behind}.
 

### 2022-07-01 manuscript/body.tex
Moved Jensen's stuff up

@@
 The common approach for bias-adjustment is second-order Taylor approximation \citep{duursma2003bias, hanmer2013behind}; already implemented in \pkg{emmeans} \citep{lenth2018package}.
-Here, we describe and implement a different approach -- whole-sample-based approach for bias correction.}
+Here, we describe and implement a different approach -- whole-sample-based approach for bias correction.
+Hammer and Kalkan \citep{hanmer2013behind} introduced mean- and whole-sample- based approaches but limited their discussion to binary response models only.}
 

### 2022-07-03 manuscript/body.tex
trying very hard ☺

@@
 Additional challenges arise when dealing with generalized linear (mixed) models (GL(M)Ms) which extend ordinary linear models to incorporate random effects.
-How we treat (include or exclude) the random effects also affect the slope and nature of the response.
+How we treat random effects can also the estimated response.
 
@@
 \BC{Added a sentence in the last \P. I think we can delete the \P}
-
-\dya{Simple OLS regression models are flexible and can easily be generalized, as generalized linear (mixed) models (GL(M)Ms), to examine more complex relationships, including nonlinear relationships between response and input variables e.g., higher order interactions via polynomials and splines, nonlinear transformation on the input variables, and nonlinear transformations via link functions.
-This flexibility comes with additional complications, for example, complex multivariate models may risk misinterpretation, and miscalculation of quantities of interest \citep{berry_improving_2012, leeper2017interpreting}.
-Further, coefficient estimates of GLMs involving nonlinear link functions lack direct interpretation\citep{leeper2017interpreting}, meaning that interpretation of derived quantities from these estimates requires some understanding of the specified model.}
 
@@
 
-\dya{An alternative is to explore \emph{predictions} and \emph{effects} \citep{fox2009effect, leeper2017package, lenth2018package}.
-For example, as a way to plan, public health officials may want to know the effects of household \code{income}, \code{wealthindex}, etc, on the predicted probability that slum dwellers will have improved water services.}
-
 \jd{We should discuss before continuing this part.}
@@
 If our goal is to \emph{predict} what we learn about the outcome variable by measuring the focal predictor, \NEW{then we would focus on the central (outcome) estimate which is the same for both prediction and effects plots.
+\jd{This \P\ should be about distinguishing -- what do they mean?
+I feel we can make good progress after a terminology meeting.}.
 It simply tells us what we expect the outcome to be for any given value of the focal input variable.
@@
 
-The most common way of choosing the representative values is taking unique levels or quantiles of the focal input variable if discrete or continuous, respectively.
+The most common way of choosing the representative values is taking unique levels or quantiles \jd{I still don't understand where quantiles come in?
+What's an example of the use of quantiles?} of the focal input variable if discrete or continuous, respectively.
 The predictions are then calculated by holding the non-focal input variables at their reference point (values chosen for the non-focal input variables, e.g., means), while varying the focal input variable, with the goal that the estimates represent how the model responds to the changes in the focal input variable \citep{fox2009effect, hanmer2013behind}.
@@
 \BC{We need to decide on how to proceed from here. I think we can refer to this as the central estimate. Re-writing and we can delete older version if that's okay.}
-\NEW{In this article, we refer to these quantities as the central (outcome) estimates for an ``average case'' in a population; and the associated confidence intervals will either be a prerequisite for prediction or effects plots.
+\jd{I had a suggestion in the notes.}
+In this article, we refer to these quantities as the central (outcome) estimates for an ``average case'' in a population; and the associated confidence intervals will either be \jd{WHAT??
+a prerequisite for} prediction or effects plots.
 In the case of effects plots, we can also choose an anchor (value chosen for the focal input variable).
 The concept of effects plots has been presented in the literature -- Shi et al.
-\citep{shi_evidence_2017} used effects plots to compare the difference in sexual risk behaviors between circumcised and uncircumcised men.}
-\dya{In this article, we refer to these quantities as predictions and effects since they should, for example, tell us what we would expect the response to be at a particular value or level of the focal input variable, for an ``average case''.
-As mentioned above, in our case, the key difference between predictions and effects is in the computation of the confidence intervals.
-The concept of effects has been presented in the literature -- Shi et al.
-\citep{shi_evidence_2017} used effect-styled plots to compare the difference in sexual risk behaviors between circumcised and uncircumcised men.}
+\citep{shi_evidence_2017} used effects plots to compare the difference in sexual risk behaviors between circumcised and uncircumcised men.
+\jd{This is useful text, but also feels repetitive; why are we coming back to the “concept” of effects plots?}
 
@@
 This is not true for the more conventional prediction confidence intervals.
-We can choose any anchor, e.g., zero-anchored (the value of the anchor is $0$), but center point seems like a natural, stable choice, we call this center-anchored.
-In Fig~\ref{fig:justify_anchors}, we compare effects plots anchored at the minimum and center (mean) of the focal input variable.
+We can choose any anchor, but the center point is a natural, stable choice, we call this.
+\jd{Why do we need this in the definitions section?:} In Fig~\ref{fig:justify_anchors}, we compare effects plots anchored at the center (mean) of the focal input variable with \ldots.
 %
@@
 \end{center}
-\caption{{\bf A comparison of zero- and center- anchored effects.} The center-anchored effects are based on the mean of the focal predictor while the zero-anchored effects are based on the minimum of the focal predictor.
+\caption{{\bf A comparison of zero- and center-anchored effects.} \jd{This explanation is not currently correct, but do we really need to explain that the zero anchor means that the anchor is at zero?
+Also, the real difference between anchors is for multi-parameter predictors?
+Do we have an example of that?} The center-anchored effects are based on the mean of the focal predictor while the zero-anchored effects are based on the minimum of the focal predictor.
 As we mentioned earlier, the choice of the anchor do not affect the main estimates (central trend lines).

### 2022-07-04 manuscript/body.tex
No major progress, we might need to discuss first.

@@
 In generalized models with a non-linear link function, or models with a spline or polynomial response to a predictor variable, they can also aid in undertanding difficult-to-interpret coefficient estimates \citep{brambor_understanding_2006, berry_improving_2012, leeper2017interpreting}.
-\jd{Do these citations make sense where I put them?}
 
@@
 How we treat random effects can also the estimated response.
-
-\jd{How much do we need of this \P? To what extent can it be merged into the two above.}
-
-\BC{Added a sentence in the last \P. I think we can delete the \P}
-
-\jd{We need to stake out what we need by predictions and effects first.
-We could do this by first talking about the CI choices, and about univariate vs.
-multivariate?
-Or by talking about what people might want from predictions and effects first?
-We can't imply that this stuff is already known.}
 
@@
 Since effects plots focus on focal-specific uncertainty, they are usually narrower than prediction plots.} \BC{Not sure whether to say that uncertainties are described by the CIs?}
-
-\dya{Both predictions and effects tell us what we expect the response to be for a given value of \emph{focal} input variable (scientific variable of interest).
-The major difference between the two lies in the construction of the associated confidence intervals.
-In particular, for predictions, we want to use the conventional curved confidence intervals (or prediction intervals) which capture the uncertainty in the whole model and generally focus on the \emph{total} effects, and so will often be suited for a univariate model.
-On the other hand, effects, attempt to show the effect of a focal input variable.
-In this case, we suggest the narrower confidence intervals which focus on uncertainty in the effect of the focal input variable only, and so will often be suited for the direct effects – after controlling for other non-focal input variables – from a multivariate model.}
-
-\dya{When visually presented, predictions and effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction or some kind of transformations on the input variables whose estimates are usually, but not always, harder to interpret \jd{Keep working on reducing repetition if you can.}.}
 
@@
 \BC{I think we should move all with "MOVE" tag to somewhere after the statistical background but Ben suggested this format. I think moving everthing downstream will streamline the flow.}
-\MOVE{The most commonly used \proglang{R} software packages for generating predictions (\pkg{emmeans} and \pkg{effects}), by default, use the average of \emph{input} and \emph{predictor variables} as the reference point for continuous and categorical input variables, respectively.
-However, there are a number of choices one can make when generating predictions -- for example, in the presence of interactions, averaging the interactions (averaging the product of interacting predictor variables) versus taking the product of the averages of the interacting predictor variables (the default for \pkg{emmeans} and \pkg{effects}).
-For simple OLS models (with no interactions), the two approaches give the same predictions.
-However, for complex models, we demonstrate that the two approaches can produce substantially different results and show that averaging the interactions closely matches the observed values.
-To illustrate this, consider models~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher} below, with $x_1$ as the focal input variable:
-%
-\begin{align}
-y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \epsilon \label{eq:simple_inter_higher_no_interaction}\\
-y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \beta_{23}x_2x_3 + \epsilon \label{eq:simple_inter_higher}.
-\end{align}
-%
-}
-
-\MOVE{We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$, $\beta_{23} = 5$, $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and $\sigma^2 = 1$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions (predictions based on the true simulation parameters) and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.
-%
-\begin{figure}
-\begin{center}
-\includegraphics{justify_preds.inter.pdf}
-\end{center}
-\caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} predictions for regression of $x_1$ on $y$.} The horizontal dashed lines are the mean of the predicted and simulated $y$, $\bar{\hat{y}}$ and $\bar{y}$, respectively.
-The grey points are the binned simulated $y$.
-The trend lines represent the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other input variables at their mean and the true predictions based on the simulation parameters.
-A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the simulated value in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$.
-Similarly, the predictions based on the three approaches match the true predictions.
-B: Even with the simple interaction between non-focal input variables, we start seeing deviation in both predicted mean, $\bar{\hat{y}}$, and overall predictions from the simulated, $\bar{y}$, and true predictions, respectively, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not the proposed \pkg{varpred}.}
-\label{fig:justify_plots}
-\end{figure}
-%
-In the absence of interactions (\EREF{simple_inter_higher_no_interaction}), the three approaches produce the same estimates, which match the simulated values, Fig~\ref{fig:justify_plots}A.
-However, in the presence of interactions, even as simple as the one in Equation~\ref{eq:simple_inter_higher}, the estimates start to differ.
-In particular, \pkg{emmeans} and \pkg{effects} give similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is very close to the simulated average ($\bar{y}$), Fig~\ref{fig:justify_plots}B.
-To generate Fig~\ref{fig:justify_plots}A,  \pkg{emmeans} and \pkg{effects} average $x_2$ and $x_3$, i.e., the input variables while \pkg{varpred} averages the predictor variables corresponds to $x_2$ and $x_3$.
-On other hand, to generate Fig~\ref{fig:justify_plots}B, the difference in the estimates lies on how each of the packages average the interaction term ($x_2x_3$).
-In particular, \pkg{emmeans} and \pkg{effects} use input variables to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses predictor variables to compute $\over{x_2x_3}$.
-In the case of Fig~\ref{fig:justify_plots}A, averaging the input variables is equivalent to averaging the predictor variables.
-}
-
-\MOVE{One may be interested in the uncertainties associated with the focal input variable only, excluding other uncertainties due to other non-focal input variables -- effects.
-However, currently, the two packages do not provide a straightforward way do achieve this.
-To illustrate this, we generate predictions of $x_2$ from Equation~\ref{eq:simple_inter_higher_no_interaction} together with the associated $95\%$ confidence intervals, as shown in Fig~\ref{fig:justify_ci_plots}.
-%
-\begin{figure}
-\begin{center}
-\includegraphics{justify_preds.isolate.pdf}
-\end{center}
-\caption{{\bf The prediction and effect styled plots at $95\%$ confidence intervals.} The description of horizontal, vertical and trend lines remain the same as above.
-The wider dashed curves corresponds to the conventional confidence intervals around the predictions from \pkg{emmeans} and \pkg{effects}, while the narrower curves crossing at the mean of the focal input variable, i.e., the center point, correspond to the effect of the focal input variable from \pkg{varpred}.
-For simple OLS models, the effect-styled confidence bands crosses at the center point (an anchor).
-The prediction-styled curves incorporates not only the uncertainties due to intercept term but also other non-focal input variables.
-On the other hand, effect-styled curves only take into account the main effect uncertainty of the focal input variable.}
-\label{fig:justify_ci_plots}
-\end{figure}
-%
-For \pkg{emmeans} and \pkg{effects}, the confidence intervals are much wider because they include uncertainties associated with the intercept and non-focal input variables, but narrower and crosses at the mean of the focal input variable in \pkg{varpred}.
-In other words, with \pkg{varpred}, we are able to generate effects indicating zero uncertainty at the value of the focal input variable we are more certain about, i.e., the center point.
-For simple OLS models, the point where the confidence intervals curves cross is the model center or simply center point; it corresponds to anchor we choose, in this case, the center point (which is generally an appropriate choice).
-Predictions are the standard curves; they tell us what we expect to see if we know the value of the focal predictor.
-The lines inside (effects) are a little different; they are a good way to visualize the effect of the focal predictor.
-Unlike predictions, where the uncertainty captures the uncertainty in the whole model, the effects plot focuses on uncertainty in the effect of the focal predictor only and depend on the anchor.
-}
 
@@
 We can choose any anchor, but the center point is a natural, stable choice, we call this.
-\jd{Why do we need this in the definitions section?:} In Fig~\ref{fig:justify_anchors}, we compare effects plots anchored at the center (mean) of the focal input variable with \ldots.
-%
-\begin{figure}
-\begin{center}
-\includegraphics{justify_anchors.combined.pdf}
-\end{center}
-\caption{{\bf A comparison of zero- and center-anchored effects.} \jd{This explanation is not currently correct, but do we really need to explain that the zero anchor means that the anchor is at zero?
-Also, the real difference between anchors is for multi-parameter predictors?
-Do we have an example of that?} The center-anchored effects are based on the mean of the focal predictor while the zero-anchored effects are based on the minimum of the focal predictor.
-As we mentioned earlier, the choice of the anchor do not affect the main estimates (central trend lines).
-\MOVE{Should me move this downstream}}
-\label{fig:justify_anchors}
-\end{figure}
-%
 \item \textbf{Marginal effects:} In simple linear models with no interaction terms, the coefficient estimates are simple and directly interpretable as the expected change in outcome for a unit change in focal input variable.
@@
 In essence, computation of marginal effects requires the use of partial derivatives and can be considered as the slope multi-dimensional surface with response to the surface of interest \cite{leeper2017interpreting}.
-The \proglang{R} software package \pkg{margins} \cite{lenth2018package} generates marginal effects.
-To illustrate the distinction between prediction and effects plots, and marginal effects, we use the simulated data described in \nameref{S1_Appendix}.
-%
-\begin{figure}
-\centering
-\includegraphics{cubic_varpred_margins_pred.ggp.pdf}
-\caption{{\bf A comparison of effects and marginal effect plots.} \NEW{A:  The black solid line is the central (outcome) estimate, while the dashed lines corresponds to the effects styled $95\%$ confidence interval.
-B: The marginal effect (ME) at representative values of the focal input variable \code{age} together with the $95\%$ confidence intervals.
-For simple linear models with no interactions, the ME should be constant across all the values of the focal input variable; and it would correspond to the slope of the central estimate of the prediction or effects.
-However, in this example, the focal input variable is a cubic polynomial, and so, the effect is not constant.} \jd{The definitions should be shorter and it's definitely weird to have a figure in the definitions.
-Maybe you should consider making a box?
-Also, if we're going to have this figure, we should try to illustrate a more difficult example for comparison.} \MOVE{I think this should also be moved downstream.} \BC{What should we call the central and/or y-axis line of the effect/prediction plot?}}
-\label{fig:qoi_age_pred_plot}
-\end{figure}
-%
 \end{itemize}
@@
 
-\jd{Are we kind of going backward now from the comparison section? Should all of these definitions come first?}
-\BC{I agree. We can move all the package comparisons to simulation section. \MOVE{}}
-
-\jd{This \P\ seems unnecessary. Why not just say “To illustrate \ldots", and then jump to the next \P (“, we consider")}
-
-\NEW{To illustrate mean-based approach, consider a simple linear model with linear predictor $\eta = \bX\bbeta$ and let $g(\boldmu) = \boldeta$ be an identity link function, where $\boldmuh$ is the expected value of response variable $\hat{y}$.
+To illustrate mean-based approach, consider a simple linear model with linear predictor $\eta = \bX\bbeta$ and let $g(\boldmu) = \boldeta$ be an identity link function, where $\boldmuh$ is the expected value of response variable $\hat{y}$.
 Let $\bbetah$ be the estimate of $\bbeta$, together with the estimated covariance matrix $\Sigma = V(\bbetah)$ of $\bbetah$.
 Then the quantity $\hat{\boldeta}^\star = \bX^\star\bbetah$ is the \fix{central (outcome) estimate} with respect to the focal input variable in question \citep{fox2009effect}; where $\bX^\star$ is the centered model matrix constructed by averaging the columns of non-focal predictors in model matrix $\bX$, and together with appropriately chosen values of focal predictor.
-}
 
@@
 
+\subsection{Comparison with other implementations}
+
+The most commonly used \proglang{R} software packages for generating predictions (\pkg{emmeans} and \pkg{effects}), by default, use the average of \emph{input} and \emph{predictor variables} as the reference point for continuous and categorical input variables, respectively.
+However, there are a number of choices one can make when generating predictions -- for example, in the presence of interactions, averaging the interactions (averaging the product of interacting predictor variables) versus taking the product of the averages of the interacting predictor variables (the default for \pkg{emmeans} and \pkg{effects}).
+For simple OLS models (with no interactions), the two approaches give the same predictions.
+However, for complex models, we demonstrate that the two approaches can produce substantially different results and show that averaging the interactions closely matches the observed values.
+To illustrate this, consider models~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher} below, with $x_1$ as the focal input variable:
+%
+\begin{align}
+y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \epsilon \label{eq:simple_inter_higher_no_interaction}\\
+y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \beta_{23}x_2x_3 + \epsilon \label{eq:simple_inter_higher}.
+\end{align}
+%
+
+We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$, $\beta_{23} = 5$, $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and $\sigma^2 = 1$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions (predictions based on the true simulation parameters) and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.
+%
+\begin{figure}
+\begin{center}
+\includegraphics{justify_preds.inter.pdf}
+\end{center}
+\caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} predictions for regression of $x_1$ on $y$.} The horizontal dashed lines are the mean of the predicted and simulated $y$, $\bar{\hat{y}}$ and $\bar{y}$, respectively.
+The grey points are the binned simulated $y$.
+The trend lines represent the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other input variables at their mean and the true predictions based on the simulation parameters.
+A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the simulated value in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$.
+Similarly, the predictions based on the three approaches match the true predictions.
+B: Even with the simple interaction between non-focal input variables, we start seeing deviation in both predicted mean, $\bar{\hat{y}}$, and overall predictions from the simulated, $\bar{y}$, and true predictions, respectively, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not the proposed \pkg{varpred}.}
+\label{fig:justify_plots}
+\end{figure}
+%
+In the absence of interactions (\EREF{simple_inter_higher_no_interaction}), the three approaches produce the same estimates, which match the simulated values, Fig~\ref{fig:justify_plots}A.
+However, in the presence of interactions, even as simple as the one in Equation~\ref{eq:simple_inter_higher}, the estimates start to differ.
+In particular, \pkg{emmeans} and \pkg{effects} give similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is very close to the simulated average ($\bar{y}$), Fig~\ref{fig:justify_plots}B.
+To generate Fig~\ref{fig:justify_plots}A,  \pkg{emmeans} and \pkg{effects} average $x_2$ and $x_3$, i.e., the input variables while \pkg{varpred} averages the predictor variables corresponds to $x_2$ and $x_3$.
+On other hand, to generate Fig~\ref{fig:justify_plots}B, the difference in the estimates lies on how each of the packages average the interaction term ($x_2x_3$).
+In particular, \pkg{emmeans} and \pkg{effects} use input variables to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses predictor variables to compute $\over{x_2x_3}$.
+In the case of Fig~\ref{fig:justify_plots}A, averaging the input variables is equivalent to averaging the predictor variables.
+
+One may be interested in the uncertainties associated with the focal input variable only, excluding other uncertainties due to other non-focal input variables -- effects.
+However, currently, the two packages do not provide a straightforward way do achieve this.
+To illustrate this, we generate predictions of $x_2$ from Equation~\ref{eq:simple_inter_higher_no_interaction} together with the associated $95\%$ confidence intervals, as shown in Fig~\ref{fig:justify_ci_plots}.
+%
+\begin{figure}
+\begin{center}
+\includegraphics{justify_preds.isolate.pdf}
+\end{center}
+\caption{{\bf The prediction and effect styled plots at $95\%$ confidence intervals.} The description of horizontal, vertical and trend lines remain the same as above.
+The wider dashed curves corresponds to the conventional confidence intervals around the predictions from \pkg{emmeans} and \pkg{effects}, while the narrower curves crossing at the mean of the focal input variable, i.e., the center point, correspond to the effect of the focal input variable from \pkg{varpred}.
+For simple OLS models, the effect-styled confidence bands crosses at the center point (an anchor).
+The prediction-styled curves incorporates not only the uncertainties due to intercept term but also other non-focal input variables.
+On the other hand, effect-styled curves only take into account the main effect uncertainty of the focal input variable.}
+\label{fig:justify_ci_plots}
+\end{figure}
+%
+For \pkg{emmeans} and \pkg{effects}, the confidence intervals are much wider because they include uncertainties associated with the intercept and non-focal input variables, but narrower and crosses at the mean of the focal input variable in \pkg{varpred}.
+In other words, with \pkg{varpred}, we are able to generate effects indicating zero uncertainty at the value of the focal input variable we are more certain about, i.e., the center point.
+For simple OLS models, the point where the confidence intervals curves cross is the model center or simply center point; it corresponds to anchor we choose, in this case, the center point (which is generally an appropriate choice).
+Predictions are the standard curves; they tell us what we expect to see if we know the value of the focal predictor.
+The lines inside (effects) are a little different; they are a good way to visualize the effect of the focal predictor.
+Unlike predictions, where the uncertainty captures the uncertainty in the whole model, the effects plot focuses on uncertainty in the effect of the focal predictor only and depend on the anchor.
+
+\jd{Why do we need this in the definitions section?:} In Fig~\ref{fig:justify_anchors}, we compare effects plots anchored at the center (mean) of the focal input variable with \ldots.
+%
+\begin{figure}
+\begin{center}
+\includegraphics{justify_anchors.combined.pdf}
+\end{center}
+\caption{{\bf A comparison of zero- and center-anchored effects.} \jd{This explanation is not currently correct, but do we really need to explain that the zero anchor means that the anchor is at zero?
+Also, the real difference between anchors is for multi-parameter predictors?
+Do we have an example of that?} The center-anchored effects are based on the mean of the focal predictor while the zero-anchored effects are based on the minimum of the focal predictor.
+As we mentioned earlier, the choice of the anchor do not affect the main estimates (central trend lines).}
+\label{fig:justify_anchors}
+\end{figure}
+%
+
+The \proglang{R} software package \pkg{margins} \cite{lenth2018package} generates marginal effects.
+To illustrate the distinction between prediction and effects plots, and marginal effects, we use the simulated data described in \nameref{S1_Appendix}.
+%
+\begin{figure}
+\centering
+\includegraphics{cubic_varpred_margins_pred.ggp.pdf}
+\caption{{\bf A comparison of effects and marginal effect plots.} \NEW{A:  The black solid line is the central (outcome) estimate, while the dashed lines corresponds to the effects styled $95\%$ confidence interval.
+B: The marginal effect (ME) at representative values of the focal input variable \code{age} together with the $95\%$ confidence intervals.
+For simple linear models with no interactions, the ME should be constant across all the values of the focal input variable; and it would correspond to the slope of the central estimate of the prediction or effects.
+However, in this example, the focal input variable is a cubic polynomial, and so, the effect is not constant.} \jd{The definitions should be shorter and it's definitely weird to have a figure in the definitions.
+Maybe you should consider making a box?
+Also, if we're going to have this figure, we should try to illustrate a more difficult example for comparison.} \BC{What should we call the central and/or y-axis line of the effect/prediction plot?}}
+\label{fig:qoi_age_pred_plot}
+\end{figure}
+%
+
 \subsection{Prediction and effect styled plots}

### 2022-07-05 manuscript/body.tex
After JD, BB and BC meeting. Going for tea

@@
 
-Plots of predicted values of an outcome against predictors (often called effects plots, or prediction plots) are often a useful way to summarize the results of a regression model.
-These can be used to illustrate model uncertainty, or to give a more explicit quantitative sense of how the outcome is expected to change.
-In generalized models with a non-linear link function, or models with a spline or polynomial response to a predictor variable, they can also aid in undertanding difficult-to-interpret coefficient estimates \citep{brambor_understanding_2006, berry_improving_2012, leeper2017interpreting}.
+Plots of predicted values of an outcome against predictors (often called effect plots or prediction plots) are often a useful way to summarize the results of a regression model.
+These can be used to illustrate model uncertainty or to give a more explicit quantitative sense of how the outcome is expected to change.
+In generalized models with a non-linear link function, or models with a spline or polynomial response to a predictor variable, they can also aid in understanding difficult-to-interpret coefficient estimates \citep{brambor_understanding_2006, berry_improving_2012, leeper2017interpreting}.
 
 To make an outcome plot, we use a \emph{focal} predictor on the x-axis.
-The resulting plot will also depend on choices we make about other (non-focal) predictor variables.
+The resulting plot will depend on choices we make about other (non-focal) predictor(s).
 In ordinary (Gaussian) linear regression, the non-focal choices may have a simple additive effect on the outcome estimate.
-However, when the focal predictor has interactions, or when there is a non-linear link function, non-focal choices can also affect the slope of the response.
-Additional challenges arise when dealing with generalized linear (mixed) models (GL(M)Ms) which extend ordinary linear models to incorporate random effects.
-How we treat random effects can also the estimated response.
+However, when the focal predictor has interactions or a non-linear link function, non-focal choices can also affect the slope of the response.
+Additional challenges arise when dealing with generalized linear (mixed) models (GL(M)Ms), which extend ordinary linear models to incorporate random effects.
+How we treat random effects can also have an effect.
 
-\jd{We should discuss before continuing this part.}
-\BC{Okay. Some thoughts below your \P}
-The outcome plots described above are often loosely called either prediction plots or effects plots.
+The outcome plots described above are often called prediction plots or effects plots.
 We endeavor here to make a conceptual distinction.
-If our goal is to \emph{predict} what we learn about the outcome variable by measuring the focal predictor, \NEW{then we would focus on the central (outcome) estimate which is the same for both prediction and effects plots.
-\jd{This \P\ should be about distinguishing -- what do they mean?
-I feel we can make good progress after a terminology meeting.}.
-It simply tells us what we expect the outcome to be for any given value of the focal input variable.
-We may also want to provide uncertainty associated with the central estimate.
-How we choose to describe the uncertainty can inform the choice we make between prediction and effects plots.
-For instance, we may want to capture the uncertainty in the whole model (these include uncertainty due to intercept, focal and non-focal predictors, and random effects) or we may be interested in showing the uncertainty associated with the focal predictor only.
-In the former case, we use the prediction plots which generally focus on the \emph{total} effect, and so will often be suited for a univariate model.
-On the other hand, in the latter case, we introduce effects plots which focus on the direct effect of the focal predictor -- after controlling for other non-focal input variables -- from a multivariate model, and so will be preferable for multivariate models.
-Since effects plots focus on focal-specific uncertainty, they are usually narrower than prediction plots.} \BC{Not sure whether to say that uncertainties are described by the CIs?}
+If our goal is to \emph{predict} what we learn about the outcome variable by measuring the focal predictor, \NEW{then we would focus on the central estimate, which is the same for both prediction and effects plots.
+However, the distinction between the two lies in how we describe the uncertainties around the central estimate.
+In particular, we may want to capture the uncertainty in the whole model (these include uncertainty due to intercept, focal and non-focal predictors, and random effects), or we may be interested in showing the uncertainty associated with the focal predictor only.
+In the former case, we use the prediction plots, which generally focus on the \emph{total} effect and will often be suited for a univariate model.
+In the latter case, we use effect plots that focus on the direct effect of the focal predictor -- after controlling for other non-focal predictors -- from a multivariate model, which will be preferable for multivariate models.
+Since effects plots focus on focal-specific uncertainty, they are usually narrower than prediction plots.
+Shi et al.
+\citep{shi_evidence_2017} used effects plots to compare the difference in sexual risk behaviors between circumcised and uncircumcised men.}
 
-Generating quantities, i.e., central estimate together with the corresponding confidence intervals, associated  prediction and effects plots has a number of challenges. In particular:
+Generating quantities, i.e., central estimate, together with the corresponding confidence intervals, i.e.,  prediction and effects plots, has some challenges. In particular:
 \begin{enumerate}
-\item choice of representative values of focal input variable and the \emph{reference point} for non-focal input variables, especially in multivariate models
-\item  uncertainty estimation -- appropriate choice of \emph{anchor} for computing confidence intervals; and how to incorporate the uncertainty due to non-focal input variables 
-\item bias in the expected mean prediction induced by the nonlinear transformation of the response variable (especially in GL(M)Ms).
+\item choice of representative values of focal predictor and the \emph{reference point} for non-focal predictors, especially in multivariate models
+\item uncertainty estimation -- appropriate choice of \emph{anchor} for computing confidence intervals; and how to incorporate the uncertainty due to non-focal predictors 
+\item bias in the expected mean prediction induced by the non-linear transformation of the response variable (especially in GL(M)Ms).
 \end{enumerate}
 
-\jd{Not sure I agree with this sentence.
-Also, picking representative focal values seems not interesting?
-The reason I'm not sure I agree is that our method of spanning the range with a fixed number of points is not exactly a quantile method; also, what does "sized" mean here?}
-\BC{I somehow agree it is not very interesting but I think we need to mention what we do.
-We do both -- spanning a range of values and quantiles of the observed focal values (our default).
-"sized" is supposed to the number of sampled data points.
-Maybe I should just say we take quantiles of the focal predictor?.}
-
-The most common way of choosing the representative values is taking unique levels or quantiles \jd{I still don't understand where quantiles come in?
-What's an example of the use of quantiles?} of the focal input variable if discrete or continuous, respectively.
-The predictions are then calculated by holding the non-focal input variables at their reference point (values chosen for the non-focal input variables, e.g., means), while varying the focal input variable, with the goal that the estimates represent how the model responds to the changes in the focal input variable \citep{fox2009effect, hanmer2013behind}.
+The most common way of choosing the representative values is taking unique levels or quantiles of the focal predictor if discrete or continuous, respectively.
+The central estimates are then calculated by holding the non-focal predictors at their reference point (values chosen for the non-focal predictors, e.g., means) while varying the focal predictor, with the goal that the estimates represent how the model responds to the changes in the focal predictor \citep{fox2009effect, hanmer2013behind}.
 This procedure generates -- \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}.
-\BC{We need to decide on how to proceed from here. I think we can refer to this as the central estimate. Re-writing and we can delete older version if that's okay.}
-\jd{I had a suggestion in the notes.}
-In this article, we refer to these quantities as the central (outcome) estimates for an ``average case'' in a population; and the associated confidence intervals will either be \jd{WHAT??
-a prerequisite for} prediction or effects plots.
-In the case of effects plots, we can also choose an anchor (value chosen for the focal input variable).
-The concept of effects plots has been presented in the literature -- Shi et al.
-\citep{shi_evidence_2017} used effects plots to compare the difference in sexual risk behaviors between circumcised and uncircumcised men.
-\jd{This is useful text, but also feels repetitive; why are we coming back to the “concept” of effects plots?}
-
-\BC{I think we should move all with "MOVE" tag to somewhere after the statistical background but Ben suggested this format. I think moving everthing downstream will streamline the flow.}
+In this article, we refer to these quantities as the \emph{central estimates} for an ``average case'' in a population; the associated confidence intervals will either describe a prediction or effect plot.
+In the effect plots, we can also choose an anchor (value chosen for the focal predictor).
 
 For linear models (including models with complex interactions), the averaging is done on the linear scale, i.e., linear averaging.
-As a result the bias in the expected (mean) predictions can easily be corrected by performing predictions from the \emph{model center} (average of the predictor variables).
-The mean prediction is the prediction at the model center.
-On the other hand, when dealing with GL(M)Ms with nonlinear link functions, the averaging is done on the nonlinear scale, i.e., nonlinear averaging, making estimating correct predictions much harder.
-This is the \emph{bias} in expected mean prediction induced by nonlinear transformation of the response variable and it lead to inaccurate predictions.
-One way to deal with this is to make predictions on the transformed scale (linear predictor scale), and then back-transform to the original scale.
+As a result, the \emph{bias} in the expected (mean) predictions can easily be corrected using \emph{mean-based reference point}, i.e., averaging of the non-focal model variables -- currently not implemented in commonly used \proglang{R} software packages.
+The mean of the central estimates is the prediction at the \emph{model center}.
+In this case, the estimates are usually consistent with the observed values.
+However, in models with non-linear link functions, this is not always the case.
+In particular, when dealing with GL(M)Ms with non-linear link functions, the averaging is done on the non-linear scale, i.e., non-linear averaging, making generating correct predictions much harder.
+This is the bias in the expected mean prediction induced by the non-linear transformation of the response variable.
+One way to address this is to make predictions on the linear predictor scale and back-transform to the original scale.
 However, the back-transformation may either result in biased predictions or requires some approximation, for example, second-order Taylor approximation implemented in \pkg{emmeans} \citep{lenth2018package}.
-In linear models, mean-based gives estimates consistent with the observed values, however, in models with nonlinear link functions, this is not always the case.
-An alternative to mean-based reference point, i.e., averaging non-focal predictor variables, is the \emph{whole-sample-based} approach, discussed in detail later, which involves computing the prediction over the population of non-focal predictor variables and then averaging across the values of the focal predictor variable \citep{hanmer2013behind}.
+An alternative to the mean-based reference point is the \emph{whole-sample-based} approach, discussed in detail later, which involves computing the prediction over the population of non-focal predictors and then averaging across the values of the focal predictor \citep{hanmer2013behind}.
 
-The main purpose of this article is to discuss and implement various approaches for computing predictions and provide an alternative method for computing the associated confidence intervals, i.e., effects.
-We further explore and demonstrate, using simulated data, approaches for correcting bias in predictions for GL(M)Ms involving nonlinear link functions.
+This article aims to discuss and implement various approaches for computing predictions and effects, together with the associated plots.
+We further explore and demonstrate, using simulated data, approaches for correcting bias in central estimates for GL(M)Ms involving non-linear link functions.
 The proposed method and \proglang{R} software package will complement the existing ones by providing: 1) a straightforward way to generate effects, and 2) an alternative and a more robust way to correct for prediction bias in GL(M)Ms.
-
-\jd{This seems like too much about the other packages.
-Maybe we could: 1) stake things out a bit better by closing off our overview with the "splash plot" (make that the first figure); 2) write the other-package stuff specifically as a comparison, and make it a bit shorter.}
-\BC{By ``splash plot'' you mean Fig 1? I am suggesting we move all the package comparison stuff to/after statistical background section or maybe to simulation example section ...}
 
@@
 For example, the regression models described by Equation~\ref{eq:simple_inter_higher_no_interaction} and Equation~\ref{eq:simple_inter_higher} both have $3$ input variables -- $x_1, x_2, x_3$.
-\item \textbf{Focal and non-focal input variables}: The focal input variable refers to the input variable we are interested in.
-For multivariate models, all other variables other than focal input variable are non-focal input variables.
+\item \textbf{Focal and non-focal predictors}: The focal predictor refers to the input variable we are interested in.
+For multivariate models, all other variables other than focal predictors are non-focal predictors.
 For example, consider model described by Equation~\ref{eq:simple_inter_higher_no_interaction}.
-Suppose we are interested in the effect of $x_1$ on the predicted $y$, then $x_1$ is the focal input variable while $x_2$ and $x_3$ are non-focal input variables.
+Suppose we are interested in the effect of $x_1$ on the predicted $y$, then $x_1$ is the focal predictor while $x_2$ and $x_3$ are non-focal predictors.
 \item \textbf{Model matrix}: Refers to the design matrix whose rows include all combination of input variables.
@@
 In our example, the column \code{religion} is missing and instead we have \code{religionJew} and \code{religionMuslim}; the missing category \code{religionChristian} is treated as the reference category.
-\item \textbf{Predictor variables (or simply predictors)}: Refer to the X-variables which go into the regression model; and represent columns in the model matrix.
-Each input variable may correspond to one or more predictor variables.
-In particular, variables with more than two categories (for instance religion in our previous example), or input variables modeled with interactions, a spline or polynomial interactions, will correspond to more than one predictor variables.
-For example, the regression models described by Equation~\ref{eq:simple_inter_higher_no_interaction} and Equation~\ref{eq:simple_inter_higher} have $4$ predictors -- $x_1, x_2, x_3$ plus intercept term and $5$ predictors -- $x_1, x_2, x_3, x_2x_3$ plus the intercept term, respectively.
-\item \textbf{Focal predictors:} All columns of the model matrix associated with the focal input variable.
-For example, all the terms of a cubic polynomial focal variable or columns corresponding to a categorical focal variable with more than two levels.
-\item \textbf{Non-focal predictors:} All columns of the model matrix associated with the non-focal input variable(s).
-For example, in the previous household size example, all the columns associated with religion are non-focal predictors if income is the focal input variable.
-\item \textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more predictor variables).
+\item \textbf{Model variables}: Refer to the X-variables which go into the regression model; and represent columns in the model matrix.
+Each input variable may correspond to one or more model variable.
+In particular, variables with more than two categories (for instance religion in our previous example), or input variables modeled with interactions, a spline or polynomial interactions, will correspond to more than one model variable.
+For example, the regression models described by Equation~\ref{eq:simple_inter_higher_no_interaction} and Equation~\ref{eq:simple_inter_higher} have $4$ model variables -- $x_1, x_2, x_3$ plus intercept term and $5$ model variables -- $x_1, x_2, x_3, x_2x_3$ plus the intercept term, respectively.
+\item \textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more model variables).
 Also referred to as center point.
 We refer to the prediction approach that uses model center as mean-based.
-The center point for a set of predictor variables corresponding to an input variable may not represent a possible value of the input variable -- the case in Fig~\ref{fig:justify_plots}B.
+The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable -- the case in Fig~\ref{fig:justify_plots}B.
 In simple linear models (without) interaction, the mean of the input variable can be used as a proxy for the center point.
-If there are more than one predictor variables associated with the focal input variable, however, the model center does not correspond to any given value of the variable, and we do not necessarily get the ``nice'' crossing we get in simpler cases, see Fig~\ref{fig:pred_cubic_plots}.
+If there are more than one model variables associated with the focal input variable, however, the model center does not correspond to any given value of the variable, and we do not necessarily get the ``nice'' crossing we get in simpler cases, see Fig~\ref{fig:pred_cubic_plots}.
 \item \textbf{Reference point:} Corresponds to the value or values chosen for non-focal input variables, when estimating the predictions and effects.
@@
 
-In whole-sample-based approach, the ensemble of predictions and CIs are back-transformed before averaging, see Equation~\ref{eq:pop_response}, so we do not need to worry about the nonlinear averaging.
-In other words, the averaging is no longer on the link scale and are likely to be bounded by the original data scale.
+In the whole-sample-based approach, the ensemble of predictions and CIs are back-transformed before averaging, see Equation~\ref{eq:pop_response}, so we do not need to worry about the nonlinear averaging.
+In other words, the averaging is no longer on the link scale and is likely to be bounded by the original data scale.
+In simple linear models without interactions, averaging on the link scale is identical to averaging on the response, so both approaches yield similar results.
+However, picking a single value, e.g., the mean of the predictor, on which to draw conclusions about the effect can be problematic, unrealistic, or not contained in or representative of the population.
+In addition, the mean-based approach fails to use every value of non-focal predictors hence not utilizing the full potential of the information contained in the data.
+This may limit the inferences we can make about the entire population.
+In general, the mean-based approach provides the predictions of an average case, whereas the whole-sample-based approach summarizes the predictions over the entire population.
+In some applications, the effect of an average case might not be generalizable to the entire population, especially, if the average does not represent the population.
+This might not be a problem in the whole-sample-based approach since it focuses on specific observations -- the prediction is first obtained for each observation and then averaged across the levels of the focal input variable.
 
-\jd{What does this mean?
-I was going to say that $g$ collapses out of \EREF{compare_anchored_pop_based} and the results are identical.
-Is that wrong?} As stated above, in simple linear models without interactions, the effect is constant, so both approaches yield similar results.
-However, picking a single value, e.g., mean of the predictor, on which to draw conclusions about the effect can be problematic, unrealistic or not contained in or representative of the population.
-In addition, the mean-based approach fails to use every values of non-focal predictors hence not utilizing the full potential of the information contained in the data.
-This may limit the inferences we can make about the entire population.
-In general, the mean-based approach provides the predictions of an average case, whereas the whole-sample-based approach, summarizes the predictions over the entire population.
-In some applications, the effect of an average case might not be generalizable to the entire population, especially, if the average does not represent the population.
-This is might not be a problem in whole-sample-based approach since it focuses on specific observations because the prediction is first obtained for each observation and then averaged across the levels of the focal input variable.
+Another potential concern with the mean-based approach arises when direct naive use leads to a rare or meaningless basis for generalization.
+For example, if our sample has 20\% Jews, 30\% Muslims, and 50\% Christians.
+One approach (default in common packages) is assigning equal category weight, i.e., 1/3 Jews, 1/3 Muslims, and 1/3 Christians (the ``sum-to-zero'' approach).
+The second approach (our default) is setting dummy categorical variables in the model matrix to their means, which, by default, set them to their sample means or observed proportions.
+The first or second approach translates to prediction for a household head who is 1/3 or 50\% Christian, respectively \citep{hanmer2013behind}.
+The second approach seems more realistic and will converge to the population mean in many cases.
 
-\jd{I meant to say earlier that we need to talk about the common “sum-to-zero-like" method of averaging over dsitnct values of a categorical input variable vs.~the true model-center method.
-This might be another place where we can show we do better than some existing packages!
-See notes.}
-Another potential concern with the mean-based approach arises in situations where direct naive use leads to rare or meaningless basis for generalization.
-For example, setting dummy categorical variables in the model matrix to their means, which, by default, sets them to their sample means or observed proportions.
-For example, a dummy variable for christian household heads may be set to $0.2$, translating to prediction for an household head who is $20\%$ christian \citep{hanmer2013behind}.
-
-Whole-sample-based approach is not entirely foolproof.
-For instance, similar to the mean-based approach, in the case of continuous focal predictors, choosing the representative values of the focal predictors can be very challenging especially if the cases are not evenly distributed around the minimum and the maximum values or within some subgroups defined in the population.
-In addition, whole-sample-based approach can be computationally intensive for large datasets.
+The whole-sample-based approach is not entirely foolproof.
+For instance, similar to the mean-based approach, in the case of continuous focal predictors, choosing the representative values of the focal predictors can be very challenging, especially if the cases are not evenly distributed around the minimum and the maximum values or within some subgroups defined in the population.
+In addition, the whole-sample-based approach can be computationally intensive for large datasets.
 

### 2022-07-05 manuscript/abstract.tex
trying very hard ☺
No major progress, we might need to discuss first.
Handing over to JD after meeting

--- 2022-06-19 manuscript/abstract.tex
@@
 {\Large
-\textbf\newline{Describing the curves: uncertainty estimation and bias correction for predictions in simple and generalized linear (mixed) models} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+\textbf\newline{Outcome plots: uncertainty estimation and bias correction for predictions and effects in simple and generalized linear (mixed) models} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
 }
@@
 
+\jd{This talks about a lot of terms that we don't define. Remember to rewrite!}
+\fix{BC to rewrite}
+
 In applications that use generalized linear (mixed) models, outcome predictions are often of interest.

### 2022-07-05 manuscript/body.tex
Handing over to JD after meeting

@@
 In simple linear models (without) interaction, the mean of the input variable can be used as a proxy for the center point.
-If there are more than one model variables associated with the focal input variable, however, the model center does not correspond to any given value of the variable, and we do not necessarily get the ``nice'' crossing we get in simpler cases, see Fig~\ref{fig:pred_cubic_plots}.
-\item \textbf{Reference point:} Corresponds to the value or values chosen for non-focal input variables, when estimating the predictions and effects.
+If there are more than one model variables associated with the focal input variable, however, the model center does not correspond to any given value of the focal predictor, and we do not necessarily get the ``nice'' crossing we get in simpler cases, see Fig~\ref{fig:pred_cubic_plots}.
+\item \textbf{Reference point:} Corresponds to the value or values chosen for non-focal predictors, when estimating the predictions and effects.
 Typically the center point (mean-based), but can instead be a whole-sample of quantiles or observations.
 Appropriately choosing the reference point is very crucial in generating correct predictions (for bias correction) especially when dealing with non-linear link functions.
-\item \textbf{Anchor:} The value chosen for the focal input variable when estimating effect confidence intervals.
-The anchor choice does not affect the main prediction estimates.
-Typically chosen as the center point of the predictor variables corresponding to the focal input variable but other sensible values can be chosen too.
+\item \textbf{Anchor:} The value chosen for the focal predictor when estimating effect confidence intervals.
+The anchor choice does not affect the central estimates.
+Typically chosen as the center point of the model variables corresponding to the focal predictor but other sensible values can be chosen too.
 If we are trying to focus on effects, the anchor from which we calculate confidence intervals becomes important.
-This is not true for the more conventional prediction confidence intervals.
-We can choose any anchor, but the center point is a natural, stable choice, we call this.
+This is not true for the predictions.
+We can choose any anchor, but the center point is a natural and stable choice.
 \item \textbf{Marginal effects:} In simple linear models with no interaction terms, the coefficient estimates are simple and directly interpretable as the expected change in outcome for a unit change in focal input variable.
@@
 Let $\bbetah$ be the estimate of $\bbeta$, together with the estimated covariance matrix $\Sigma = V(\bbetah)$ of $\bbetah$.
-Then the quantity $\hat{\boldeta}^\star = \bX^\star\bbetah$ is the \fix{central (outcome) estimate} with respect to the focal input variable in question \citep{fox2009effect}; where $\bX^\star$ is the centered model matrix constructed by averaging the columns of non-focal predictors in model matrix $\bX$, and together with appropriately chosen values of focal predictor.
+Then the quantity $\hat{\boldeta}^\star = \bX^\star\bbetah$ is the central estimate with respect to the focal predictor in question \citep{fox2009effect}; where $\bX^\star$ is the centered model matrix constructed by averaging the columns of non-focal predictors in model matrix $\bX$, and together with appropriately chosen values of focal predictor.
 
@@
 \section{Bias correction}
-
-\jd{Why is whole-sample common? Does anyone do it at all except us?}
 

### 2022-07-06 manuscript/body.tex
Meeting JD?

@@
 
-\subsection{Dealing with higher-order terms}
+\subsection{Dealing with multi-parameter predictors}
 
-\jd{I'm having trouble staking out higher-order terms as a thing.
-Yes, many times some or all of the terms in a multi-parameter predictor can be described as higher order, but not always (e.g., categorical predictors).
-Also, why are we interested in things that don't involve the focal predictor (and why only interactions?).
-I know that we have to average properly, but we should do that once.
-Does it make sense to talk separately about interactions, and multi-parameter focal predictors?}
-
-\BC{Should we just be more specific and say "Dealing with polynomial input variables"?}
-\BC{JD to explain second part ...}
-
-Higher order input variables such as interactions, splines, polynomials, etc., can be within the focal input variable, within (between) non-focal input variable(s) or between the focal and non-focal input variable(s).
-To distinguish the three \jd{Right?
-Why are there three?
-There should be two or four.} \BC{Three: 1) on the focal, 2) on the non-focal, and 3) between focal and non-focal.
-Which one is redundant or missing?}, suppose the model which describes the hypothetical simulation of household size based on a number of socio-demographic factors such as age and wealth index is
+Muli-parameter variables (MPVs) such as splines, polynomials, etc., can be within the focal predictor, or within (between) non-focal predictor(s).
+To distinguish the two, suppose the model which describes the hypothetical simulation of household size based on a number of socio-demographic factors such as age and wealth index is
 %
@@
 %
-In the first case, with \code{Age} as the focal input variable, with cubic polynomial interaction ($3$ focal predictors associated with it).
-In this case, each of the focal predictors ($\mathrm{Age}_i, \mathrm{Age}^2_i$ and $\mathrm{Age}^3_i$) are evaluated independently across the chosen levels of the focal input variable, $Age_i$.
+\BC{Got to this point \dots}
+In the first case, with \code{Age} as the focal input variable, is a cubic polynomial (with $3$ model variables).
+In this case, each model variables ($\mathrm{Age}_i, \mathrm{Age}^2_i$ and $\mathrm{Age}^3_i$) is evaluated separately across the chosen levels of the focal input variable, $Age_i$.
 Specifically, the higher order terms are treated as additional columns of the model matrix evaluated with the same values chosen for the focal input variable, while non-focal predictors are fixed at their reference point as discussed in the previous section.

### 2022-07-06 manuscript/body.tex
Past MPVs

@@
 
-Generating quantities, i.e., central estimate, together with the corresponding confidence intervals, i.e.,  prediction and effects plots, has some challenges. In particular:
+Generating quantities, i.e., central estimate together with the corresponding confidence intervals has some challenges. In particular:
 \begin{enumerate}
@@
 
-\subsection{Dealing with multi-parameter predictors}
+\subsection{Dealing with multi-parameter variables}
 
@@
 %
-\BC{Got to this point \dots}
-In the first case, with \code{Age} as the focal input variable, is a cubic polynomial (with $3$ model variables).
-In this case, each model variables ($\mathrm{Age}_i, \mathrm{Age}^2_i$ and $\mathrm{Age}^3_i$) is evaluated separately across the chosen levels of the focal input variable, $Age_i$.
-Specifically, the higher order terms are treated as additional columns of the model matrix evaluated with the same values chosen for the focal input variable, while non-focal predictors are fixed at their reference point as discussed in the previous section.
-In this case, the predictions associated with \code{Age} on the linear predictor scale become
+In the first case, with \code{Age} as the focal input variable, is a cubic polynomial with three model variables ($\mathrm{Age}_i, \mathrm{Age}^2_i$ and $\mathrm{Age}^3_i$).
+In this case, each model variable is evaluated separately across the chosen levels of the focal input variable, $Age_i$.
+Specifically, the model variables are treated as additional columns of the model matrix evaluated with the same values chosen for the focal input variable, while non-focal predictors are fixed at their reference point as discussed in the previous section.
+The central estimates associated with \code{Age} on the linear predictor scale become
 %
@@
 %
-In the second case, with \code{Wealthindex} as the focal input variable, the non-focal variable, \code{Age}, is a cubic polynomial.
-In this case, the non-focal predictors ($\mathrm{Age}_i, \mathrm{Age}^2_i$ and $\mathrm{Age}^3_i$) are simply treated as additional columns in the predictor space and an appropriate choice of reference point applies just like in the models without higher non-focal variable interactions.
-For instance, in mean-based approach, we average all non-focal predictors.
+In the second case, with \code{Wealthindex} as the focal predictor, the non-focal predictor, \code{Age}, is a cubic polynomial.
+In this case, the non-focal model variables ($\mathrm{Age}_i, \mathrm{Age}^2_i$ and $\mathrm{Age}^3_i$) are all treated as separate non-focal model variables and an appropriate choice of reference point applies just like in the models without MPVs.
+For instance, in mean-based approach, we average all non-focal MPVs.
 Thus
@@
 %
-Lastly, consider model described by Equation~\ref{eq:simple_inter_higher}, previously used to demonstrate how to handle interactions in non-focal input variables.
-Here, we use the same model to demonstrate how to handle interactions between the focal input variable, $x_2$, and non-focal input variable, $x_3$.
-In this case, the non-interacting predictors are treated as non-focal predictors, with appropriate reference points.
-However, for the interacting predictors, we first choose representative values for the focal input variable and then some particular values of the interacting non-focal input variable in a similar way we chose the values for the focal input variable or use predetermined values.
-In our example, suppose we pick $i$ and $j$ unique values of the focal input variable, $x_2$ and interacting input variable $x_3$, respectively.
-The prediction on linear predictor scale is given by
+
+We could also have interactions between focal and non-focal predictors.
+To illustrate, we consider simple input variables but the logic extends to more complex ones like MPVs.
+Consider model described by Equation~\ref{eq:simple_inter_higher}, used to demonstrate how to handle interactions in non-focal predictors.
+Here, we use it to demonstrate how to handle interactions between the focal, $x_2$, and non-focal, $x_3$, predictors.
+In this case, the non-interacting predictors are treated as non-focal model variables, with appropriate choice of reference points.
+However, for the interacting model variables, we first choose representative values for the focal predictor and then some particular values of the interacting non-focal predictors in a similar way we choose the values of the focal predictor or use predetermined values.
+In our example, suppose we pick $i$ and $j$ unique values of the focal predictor, $x_2$ and interacting non-focal predictor, $x_3$, respectively.
+The central estimate on linear predictor scale is given by
 %
@@
 %
-
-In general, our formulation, even for more complicated interactions, follow these three basic principles -- interaction within focal input variables, interaction between non-focal input variables and interaction between focal and non-focal input variables.
-\jd{We should talk about what this s. means. What are the principles?}
-\BC{DISCUSS with JD}
 
@@
 \NEW{We may be interested in the uncertainties associated with the focal input variable only, excluding other uncertainties due to other non-focal input variables -- effects.
-However, commonly used \proglang{R} packages for constructing predictions do not exclude the uncertainties resulting from non focal input variables when computing the CIs.} \jd{We should first explain what we want to do, and then do our comparisons with existing packages.} Currently obscure way to exclude uncertainties associated with non-focal input variables in some of these packages is to provide a user defined variance-covariance matrix with the covariances of non-focal predictors set to $0$.
+However, commonly used \proglang{R} packages for constructing predictions do not exclude the uncertainties resulting from non focal input variables when computing the CIs.} Currently obscure way to exclude uncertainties associated with non-focal input variables in some of these packages is to provide a user defined variance-covariance matrix with the covariances of non-focal predictors set to $0$.
 We refer to this procedure as \emph{zeroing-out} variance-covariance matrix; and only works when the input variables are \emph{centered} prior to model fitting, in case of numerical variables, and more complicated when the input variables are categorical.
@@
 In non-trivial link functions, the Jensen effect may depend on the additional input variables \citep{ye_generalized_2021}}.
-
-\jd{We need to explain why we think this is a problem, not just state that it is “usually important".} \BC{See previous \P}
 

### 2022-07-07 manuscript/body.tex
More progress

@@
 % Use "Eq" instead of "Equation" for equation citations. 
+%% BC: use \EREF{} for equations and \FREF{} for figures
 % Best to use a command that combines the name and the reference
@@
 \item \textbf{Input variables}: Refers to the observed (or scientific) variables underlying an inference or exploration.
-For example, the regression models described by Equation~\ref{eq:simple_inter_higher_no_interaction} and Equation~\ref{eq:simple_inter_higher} both have $3$ input variables -- $x_1, x_2, x_3$.
+For example, the regression models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher} both have $3$ input variables -- $x_1, x_2, x_3$.
 \item \textbf{Focal and non-focal predictors}: The focal predictor refers to the input variable we are interested in.
 For multivariate models, all other variables other than focal predictors are non-focal predictors.
-For example, consider model described by Equation~\ref{eq:simple_inter_higher_no_interaction}.
+For example, consider model described by \EREF{simple_inter_higher_no_interaction}.
 Suppose we are interested in the effect of $x_1$ on the predicted $y$, then $x_1$ is the focal predictor while $x_2$ and $x_3$ are non-focal predictors.
@@
 In particular, variables with more than two categories (for instance religion in our previous example), or input variables modeled with interactions, a spline or polynomial interactions, will correspond to more than one model variable.
-For example, the regression models described by Equation~\ref{eq:simple_inter_higher_no_interaction} and Equation~\ref{eq:simple_inter_higher} have $4$ model variables -- $x_1, x_2, x_3$ plus intercept term and $5$ model variables -- $x_1, x_2, x_3, x_2x_3$ plus the intercept term, respectively.
+For example, the regression models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher} have $4$ model variables -- $x_1, x_2, x_3$ plus intercept term and $5$ model variables -- $x_1, x_2, x_3, x_2x_3$ plus the intercept term, respectively.
 \item \textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more model variables).
@@
 We refer to the prediction approach that uses model center as mean-based.
-The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable -- the case in Fig~\ref{fig:justify_plots}B.
+The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable -- the case in \FREF{justify_plots}B.
 In simple linear models (without) interaction, the mean of the input variable can be used as a proxy for the center point.
-If there are more than one model variables associated with the focal input variable, however, the model center does not correspond to any given value of the focal predictor, and we do not necessarily get the ``nice'' crossing we get in simpler cases, see Fig~\ref{fig:pred_cubic_plots}.
+If there are more than one model variables associated with the focal input variable, however, the model center does not correspond to any given value of the focal predictor, and we do not necessarily get the ``nice'' crossing we get in simpler cases, see \FREF{pred_cubic_plots}.
 \item \textbf{Reference point:} Corresponds to the value or values chosen for non-focal predictors, when estimating the predictions and effects.
@@
 This is the \emph{unconditional marginal effect} and it is constant across all the observations and levels of all other input variables.
-Consider models described by Equation~\ref{eq:simple_inter_higher_no_interaction} and Equation~\ref{eq:simple_inter_higher}.
+Consider models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher}.
 The marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher_no_interaction} is $\frac{\partial y}{\partial x_2} = \beta_2$.
@@
 To illustrate, we consider simple input variables but the logic extends to more complex ones like MPVs.
-Consider model described by Equation~\ref{eq:simple_inter_higher}, used to demonstrate how to handle interactions in non-focal predictors.
+Consider model described by \EREF{simple_inter_higher}, used to demonstrate how to handle interactions in non-focal predictors.
 Here, we use it to demonstrate how to handle interactions between the focal, $x_2$, and non-focal, $x_3$, predictors.
@@
 This is the effect of the particular focal input variable.
-By setting $\baf = \boldsymbol 0$, we get the variances for the conventional CIs in Equation~\ref{eq:conventional_variance}.
+By setting $\baf = \boldsymbol 0$, we get the variances for the conventional CIs in \EREF{conventional_variance}.
 
@@
 An alternative approach to choosing a reference point is to make predictions over all observations of the non-focal input variables (members of the population) \citep{hanmer2013behind}.
-The nonlinear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population to a one-dimensional distribution of $\eta^\star(x_f, \nset{x})$, which is a function of the chosen values of the focal predictor and the whole sample of non-focal predictors, as opposed to the definition in Equation~\ref{eq:eta_mean}.
+The nonlinear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population to a one-dimensional distribution of $\eta^\star(x_f, \nset{x})$, which is a function of the chosen values of the focal predictor and the whole sample of non-focal predictors, as opposed to the definition in \EREF{eta_mean}.
 More specifically:
@@
 
-Once Equation~\ref{eq:pop_eta} is computed, we back-transform the estimates to the original scale and average over the levels of the focal predictors, $j$:
+Once \EREF{pop_eta} is computed, we back-transform the estimates to the original scale and average over the levels of the focal predictors, $j$:
 %
@@
 For models with random effects components, we make further adjustment to correct for bias induced by the random effects terms.
-In whole-sample approach, we treat the random effects terms as additional non-focal predictors and simply make adjustment to Equation~\ref{eq:pop_eta}.
+In whole-sample approach, we treat the random effects terms as additional non-focal predictors and simply make adjustment to \EREF{pop_eta}.
 In particular
@@
 
-In the whole-sample-based approach, the ensemble of predictions and CIs are back-transformed before averaging, see Equation~\ref{eq:pop_response}, so we do not need to worry about the nonlinear averaging.
+In the whole-sample-based approach, the ensemble of predictions and CIs are back-transformed before averaging, see \EREF{pop_response}, so we do not need to worry about the nonlinear averaging.
 In other words, the averaging is no longer on the link scale and is likely to be bounded by the original data scale.
@@
 
-We now illustrate the construction of prediction and effect plots and also demonstrate that the mean-based and whole-sample-based approaches produce different results in models with nonlinear link functions and/or additional source(s) of potential bias.
+We now illustrate the construction of outcome plots, compare with other packages and also demonstrate that the mean-based and whole-sample-based approaches produce different results in models with nonlinear link functions and/or additional source(s) of potential bias.
 In addition, we demonstrate that whole-sample-based approach can be used for bias correction.
@@
 
-The most commonly used \proglang{R} software packages for generating predictions (\pkg{emmeans} and \pkg{effects}), by default, use the average of \emph{input} and \emph{predictor variables} as the reference point for continuous and categorical input variables, respectively.
-However, there are a number of choices one can make when generating predictions -- for example, in the presence of interactions, averaging the interactions (averaging the product of interacting predictor variables) versus taking the product of the averages of the interacting predictor variables (the default for \pkg{emmeans} and \pkg{effects}).
-For simple OLS models (with no interactions), the two approaches give the same predictions.
+The most commonly used \proglang{R} software packages for outcome plots (\pkg{emmeans} and \pkg{effects}), by default, use the average of input variables and ``sum-to-zero'' as the reference point approach for continuous and categorical input variables, respectively.
+However, there are a number of choices one can make when constructing outcome plots -- for example, in the presence of interactions, averaging the interactions (averaging the product of model variables) versus taking the product of the averages of the input variables (the default for \pkg{emmeans} and \pkg{effects}).
+For simple OLS models (with no interactions), the two approaches give the estimates.
 However, for complex models, we demonstrate that the two approaches can produce substantially different results and show that averaging the interactions closely matches the observed values.
@@
 %
-
-We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$, $\beta_{23} = 5$, $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and $\sigma^2 = 1$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions (predictions based on the true simulation parameters) and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.
+We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$, $\beta_{23} = 5$, $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and $\sigma^2 = 1$, and then compare estimates from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' central estimates (predictions based on the true simulation parameters) and observed average i.e., $\bar{y}$, as shown in \FREF{justify_plots}.
 %
@@
 \end{center}
-\caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} predictions for regression of $x_1$ on $y$.} The horizontal dashed lines are the mean of the predicted and simulated $y$, $\bar{\hat{y}}$ and $\bar{y}$, respectively.
+\caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} central estimates for regression of $x_1$ on $y$ (outcome plots).} The horizontal dashed lines are the mean of the central estimates and simulated $y$, $\bar{\hat{y}}$ and $\bar{y}$, respectively.
 The grey points are the binned simulated $y$.
-The trend lines represent the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other input variables at their mean and the true predictions based on the simulation parameters.
-A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the simulated value in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$.
-Similarly, the predictions based on the three approaches match the true predictions.
-B: Even with the simple interaction between non-focal input variables, we start seeing deviation in both predicted mean, $\bar{\hat{y}}$, and overall predictions from the simulated, $\bar{y}$, and true predictions, respectively, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not the proposed \pkg{varpred}.}
+The trend lines represent the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other input variables at their mean and the true predictions based on the simulation parameters -- central estimate.
+A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the simulated one in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$.
+Similarly, the central estimate based on the three approaches match the truth.
+B: Even with the simple interaction between non-focal predictors, we see differences between predicted mean, $\bar{\hat{y}}$ and simulated mean , $\bar{y}$.
+Similarly, there are differences between central estimates and the truth, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not the proposed \pkg{varpred}.}
 \label{fig:justify_plots}
@@
 %
-In the absence of interactions (\EREF{simple_inter_higher_no_interaction}), the three approaches produce the same estimates, which match the simulated values, Fig~\ref{fig:justify_plots}A.
-However, in the presence of interactions, even as simple as the one in Equation~\ref{eq:simple_inter_higher}, the estimates start to differ.
-In particular, \pkg{emmeans} and \pkg{effects} give similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is very close to the simulated average ($\bar{y}$), Fig~\ref{fig:justify_plots}B.
-To generate Fig~\ref{fig:justify_plots}A,  \pkg{emmeans} and \pkg{effects} average $x_2$ and $x_3$, i.e., the input variables while \pkg{varpred} averages the predictor variables corresponds to $x_2$ and $x_3$.
-On other hand, to generate Fig~\ref{fig:justify_plots}B, the difference in the estimates lies on how each of the packages average the interaction term ($x_2x_3$).
-In particular, \pkg{emmeans} and \pkg{effects} use input variables to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses predictor variables to compute $\over{x_2x_3}$.
-In the case of Fig~\ref{fig:justify_plots}A, averaging the input variables is equivalent to averaging the predictor variables.
+In the absence of interactions (\EREF{simple_inter_higher_no_interaction}), the three approaches produce the same estimates, which match the simulated values, \FREF{justify_plots}A.
+However, in the presence of interactions, even as simple as the one in \EREF{simple_inter_higher}, the estimates start to differ.
+In particular, \pkg{emmeans} and \pkg{effects} have similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is very close to the simulated average ($\bar{y}$), \FREF{justify_plots}B.
+To generate \FREF{justify_plots}A,  \pkg{emmeans} and \pkg{effects} average $x_2$ and $x_3$, i.e., the input variables while \pkg{varpred} averages the model variables corresponding to $x_2$ and $x_3$.
+On other hand, to generate \FREF{justify_plots}B, the difference in the estimates lies on how each of the packages average the interaction model variable ($x_2x_3$).
+In particular, \pkg{emmeans} and \pkg{effects} use input variables to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses model variables to compute $\over{x_2x_3}$.
+In the case of \FREF{justify_plots}A, averaging the input variables is equivalent to averaging the model variables.
 
-One may be interested in the uncertainties associated with the focal input variable only, excluding other uncertainties due to other non-focal input variables -- effects.
-However, currently, the two packages do not provide a straightforward way do achieve this.
-To illustrate this, we generate predictions of $x_2$ from Equation~\ref{eq:simple_inter_higher_no_interaction} together with the associated $95\%$ confidence intervals, as shown in Fig~\ref{fig:justify_ci_plots}.
+\BC{Got to this point \dots}
+
+One may be interested in the uncertainties associated with the focal input variable only, excluding other uncertainties due to other non-focal input variables -- effect plots.
+However, as previously mentioned, the two packages do not provide a straightforward way do achieve this.
+To illustrate this, we consider \EREF{simple_inter_higher_no_interaction} and use $x_2$ as the focal predictor, and generate the outcome plot shown in \FREF{justify_ci_plots}.
 %
@@
 
-\jd{Why do we need this in the definitions section?:} In Fig~\ref{fig:justify_anchors}, we compare effects plots anchored at the center (mean) of the focal input variable with \ldots.
+\jd{Why do we need this in the definitions section?:} In \FREF{justify_anchors}, we compare effects plots anchored at the center (mean) of the focal input variable with \ldots.
 %
@@
 We then generated two sets of mean-based predictions and effects: 1) one with \code{age} as the focal input variable (a cubic polynomial focal input variable); and 2) one with \code{Wealthindex} as the focal in put variable (linear focal input variable).
-In both cases the, the center point was used as the anchor for the effects, as shown in Fig~\ref{fig:pred_cubic_plots}.
+In both cases the, the center point was used as the anchor for the effects, as shown in \FREF{pred_cubic_plots}.
 
@@
 Mean-based and whole-sample-based approaches can produce very different results in models with nonlinear link functions with additional sources of potential bias such as additional non-focal input variable(s), random effects or complex interactions.
-To demonstrate this, we considered a two predictor binary outcome simulation described in \nameref{S2_Appendix}, such that age has a very small effect size in comparison to wealth index, and compared their effects on the predicted probability of improved water quality as shown in Fig~\ref{fig:pred_bin_plots}.
+To demonstrate this, we considered a two predictor binary outcome simulation described in \nameref{S2_Appendix}, such that age has a very small effect size in comparison to wealth index, and compared their effects on the predicted probability of improved water quality as shown in \FREF{pred_bin_plots}.
 
@@
 
-If there was no effect of nonlinear averaging, then we would expect the average observed proportion and the average predictions to intersect at the center point as we see in Fig~\ref{fig:pred_cubic_plots}B.
-One possible reason for the variations we see in Fig~\ref{fig:pred_bin_plots} is the nonlinear averaging; since both observed status and predicted probabilities are averaged on the response scale as opposed to link scale.
+If there was no effect of nonlinear averaging, then we would expect the average observed proportion and the average predictions to intersect at the center point as we see in \FREF{pred_cubic_plots}B.
+One possible reason for the variations we see in \FREF{pred_bin_plots} is the nonlinear averaging; since both observed status and predicted probabilities are averaged on the response scale as opposed to link scale.
 For example, if the range of values are bigger than $0.5$ (seemingly the case here), then we would expected the averages to be slightly higher than what we would expect at the center point.
 
-In Fig~\ref{fig:pred_bin_prediction_effects_plots} below, we zoom into Fig~\ref{fig:pred_bin_plots}{A} and compare the prediction and effect styled plots.
+In \FREF{pred_bin_prediction_effects_plots} below, we zoom into \FREF{pred_bin_plots}A and compare the prediction and effect styled plots.
 
@@
 \caption{{\bf Prediction and effect styled plots and bias correction.} The central curves for predictions and effects are the same in the respective case.
-These patterns are  consistent with the ones in Fig~\ref{fig:pred_bin_plots}A with respect to the \emph{truth}.
+These patterns are  consistent with the ones in \FREF{pred_bin_plots}A with respect to the \emph{truth}.
 Further, the prediction and effect CIs exhibit patterns similar to the ones we previously described and generalizes in a similar way to the ones in simple linear models.
@@
 We simulated the data as shown in \nameref{S3_Appendix} and fitted the two models -- non-mediated and mediated.
-In both cases, we compared mean-based and the whole-sample-based predictions, as shown in Fig~\ref{fig:pred_mediated_plots}.
+In both cases, we compared mean-based and the whole-sample-based predictions, as shown in \FREF{pred_mediated_plots}.
 
@@
 
-From Fig~\ref{fig:pred_mediated_plots}A, we see what we would expect in the absence of additional sources of bias, even though in the simulation, the effect of $x$ on $z$ is mediated through $y$.
+From \FREF{pred_mediated_plots}A, we see what we would expect in the absence of additional sources of bias, even though in the simulation, the effect of $x$ on $z$ is mediated through $y$.
 This is the total effect of $x$, which only tells us the influence of $x$ on $z$.
 By ignoring $y$ in the model, we are still able to capture the effect of $x$ and closely match the observed values using both approaches.
-However, if there were additional non-focal predictors, we would expect to see the differences similar to those in Fig~\ref{fig:pred_bin_plots}A.
-Including both $y$ and $x$ ``dilutes'' the direct effect of $x$ on $z$ and as a result, our predictions do not necessarily match the observed binned observations (see Fig~\ref{fig:pred_mediated_plots}B).
+However, if there were additional non-focal predictors, we would expect to see the differences similar to those in \FREF{pred_bin_plots}A.
+Including both $y$ and $x$ ``dilutes'' the direct effect of $x$ on $z$ and as a result, our predictions do not necessarily match the observed binned observations (see \FREF{pred_mediated_plots}B).
 

### 2022-07-07 manuscript/body.tex
Few minutes break

@@
 However, there are a number of choices one can make when constructing outcome plots -- for example, in the presence of interactions, averaging the interactions (averaging the product of model variables) versus taking the product of the averages of the input variables (the default for \pkg{emmeans} and \pkg{effects}).
-For simple OLS models (with no interactions), the two approaches give the estimates.
+For simple OLS models (with no interactions), the two approaches give the same estimates.
 However, for complex models, we demonstrate that the two approaches can produce substantially different results and show that averaging the interactions closely matches the observed values.
@@
 %
-We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$, $\beta_{23} = 5$, $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and $\sigma^2 = 1$, and then compare estimates from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' central estimates (predictions based on the true simulation parameters) and observed average i.e., $\bar{y}$, as shown in \FREF{justify_plots}.
+We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$, $\beta_{23} = 5$, $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and $\sigma^2 = 1$, and then compare estimates from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' central estimates (based on the true simulation parameters) and observed average, i.e., $\bar{y}$, as shown in \FREF{justify_plots}.
 %
@@
 
-\BC{Got to this point \dots}
-
-One may be interested in the uncertainties associated with the focal input variable only, excluding other uncertainties due to other non-focal input variables -- effect plots.
+One may be interested in the uncertainties associated with the focal predictor only, excluding other uncertainties due to other non-focal predictors -- effect plots.
 However, as previously mentioned, the two packages do not provide a straightforward way do achieve this.
@@
 \begin{center}
-\includegraphics{justify_preds.isolate.pdf}
+\includegraphics{justify_anchors_all.ggp.pdf}
 \end{center}
-\caption{{\bf The prediction and effect styled plots at $95\%$ confidence intervals.} The description of horizontal, vertical and trend lines remain the same as above.
-The wider dashed curves corresponds to the conventional confidence intervals around the predictions from \pkg{emmeans} and \pkg{effects}, while the narrower curves crossing at the mean of the focal input variable, i.e., the center point, correspond to the effect of the focal input variable from \pkg{varpred}.
-For simple OLS models, the effect-styled confidence bands crosses at the center point (an anchor).
-The prediction-styled curves incorporates not only the uncertainties due to intercept term but also other non-focal input variables.
-On the other hand, effect-styled curves only take into account the main effect uncertainty of the focal input variable.}
+\caption{{\bf Prediction and effect plots.} The description of horizontal, vertical and trend lines remain the same as above.
+A: The wider dashed curves corresponds to the conventional prediction curves from \pkg{emmeans} and \pkg{effects} packages, while the narrower curves crossing at the mean of the focal predictor, also referred to as the center point, represent the effect from \pkg{varpred}.
+For simple OLS models, the effect-styled curves cross at the center point (which is the default anchor and also referred to as center-anchored).
+The prediction-styled curves incorporates not only the uncertainties due to intercept term but also other non-focal predictors.
+On the other hand, effect-styled curves only take into account the main effect uncertainty of the focal predictor.
+B: center- and zero- anchored effects.
+The center-anchored effects (as described above) while the zero-anchored means that the anchor is at zero value of the focal predictor.
+The choice of the anchor do not affect the central estimates, hence they are identical in both choices of the anchor.}
 \label{fig:justify_ci_plots}
@@
 %
-For \pkg{emmeans} and \pkg{effects}, the confidence intervals are much wider because they include uncertainties associated with the intercept and non-focal input variables, but narrower and crosses at the mean of the focal input variable in \pkg{varpred}.
-In other words, with \pkg{varpred}, we are able to generate effects indicating zero uncertainty at the value of the focal input variable we are more certain about, i.e., the center point.
-For simple OLS models, the point where the confidence intervals curves cross is the model center or simply center point; it corresponds to anchor we choose, in this case, the center point (which is generally an appropriate choice).
+For \pkg{emmeans} and \pkg{effects}, the confidence intervals are much wider because they include uncertainties associated with the intercept and non-focal predictors, but narrower and crosses at the mean of the focal predictor in \pkg{varpred}.
+In other words, with \pkg{varpred}, we are able to generate effects indicating zero uncertainty at the value of the focal predictor we are more certain about, i.e., the anchor.
+For simple OLS models, the point where the curves cross is the model center or simply center point; it corresponds to anchor we choose, in this case, the center point (which is generally an appropriate and stable choice).
 Predictions are the standard curves; they tell us what we expect to see if we know the value of the focal predictor.
@@
 Unlike predictions, where the uncertainty captures the uncertainty in the whole model, the effects plot focuses on uncertainty in the effect of the focal predictor only and depend on the anchor.
-
-\jd{Why do we need this in the definitions section?:} In \FREF{justify_anchors}, we compare effects plots anchored at the center (mean) of the focal input variable with \ldots.
-%
-\begin{figure}
-\begin{center}
-\includegraphics{justify_anchors.combined.pdf}
-\end{center}
-\caption{{\bf A comparison of zero- and center-anchored effects.} \jd{This explanation is not currently correct, but do we really need to explain that the zero anchor means that the anchor is at zero?
-Also, the real difference between anchors is for multi-parameter predictors?
-Do we have an example of that?} The center-anchored effects are based on the mean of the focal predictor while the zero-anchored effects are based on the minimum of the focal predictor.
-As we mentioned earlier, the choice of the anchor do not affect the main estimates (central trend lines).}
-\label{fig:justify_anchors}
-\end{figure}
-%
 
@@
 For simple linear models with no interactions, the ME should be constant across all the values of the focal input variable; and it would correspond to the slope of the central estimate of the prediction or effects.
-However, in this example, the focal input variable is a cubic polynomial, and so, the effect is not constant.} \jd{The definitions should be shorter and it's definitely weird to have a figure in the definitions.
-Maybe you should consider making a box?
-Also, if we're going to have this figure, we should try to illustrate a more difficult example for comparison.} \BC{What should we call the central and/or y-axis line of the effect/prediction plot?}}
+However, in this example, the focal input variable is a cubic polynomial, and so, the effect is not constant.}}
 \label{fig:qoi_age_pred_plot}
@@
 
-\subsection{Prediction and effect styled plots}
+\BC{Got to this point \dots}
+
+\subsection{Prediction and effect styled plots for MPVs}
 

### 2022-07-07 manuscript/cubic_varpred_margins_pred.R

@@ -21,7 +21,7 @@ meffect_plot <- (ggplot(meffect_df, aes(x=xvals))
 ## varpred
 varpred_df <- varpred(mod_cubic, "age", steps=500, modelname="A) Prediction and effects")
 varpred_plot <- (plot(varpred_df)
-	+ labs(y="Predicted household size", title="A) Prediction and effects")
+	+ labs(y="Household size", title="A) Prediction and effects")
 )
 
 ## Combine all prediction for faceting

### 2022-07-07 manuscript/body.tex
Autocommit

@@
 In other words, if there are no interactions, the marginal effect of $x_2$ on $y$ is constant, while, if there are interactions in the model, the marginal effect of a change in $x_2$ on $y$ depends on the value of the other \emph{conditioning} predictor, $x_3$.
-In essence, computation of marginal effects requires the use of partial derivatives and can be considered as the slope multi-dimensional surface with response to the surface of interest \cite{leeper2017interpreting}.
+In essence, computation of marginal effects requires the use of partial derivatives and can be considered as the slope of a multi-dimensional surface with respect to the surface of interest \cite{leeper2017interpreting}.
 \end{itemize}
@@
 %
-In the first case, with \code{Age} as the focal input variable, is a cubic polynomial with three model variables ($\mathrm{Age}_i, \mathrm{Age}^2_i$ and $\mathrm{Age}^3_i$).
-In this case, each model variable is evaluated separately across the chosen levels of the focal input variable, $Age_i$.
-Specifically, the model variables are treated as additional columns of the model matrix evaluated with the same values chosen for the focal input variable, while non-focal predictors are fixed at their reference point as discussed in the previous section.
+In the first case, with \code{Age} as the focal predictor, is a cubic polynomial with three model variables ($\mathrm{Age}_i, \mathrm{Age}^2_i$ and $\mathrm{Age}^3_i$).
+In this case, each model variable is evaluated separately across the chosen levels of the focal predictor, $Age_i$.
+Specifically, the model variables are treated as additional columns of the model matrix evaluated with the same values chosen for the focal predictor, while non-focal predictors are fixed at their reference point as discussed in the previous section.
 The central estimates associated with \code{Age} on the linear predictor scale become
@@
 We describe the uncertainty around the estimates using confidence intervals (CIs).
-In principle, every value of focal input variable has a different CI.
+In principle, every value of focal predictor has a different CI.
 The conventional way to compute variances for predictions is
@@
 so that the confidence intervals are $\boldeta \pm q\boldsymbol\sigma$, where $q$ is an appropriate quantile of Normal or t distribution \citep{lenth2018package, fox2009effect}.
-This generates conventional CIs which incorporate all the uncertainties -- including the uncertainties due to the intercept and non-focal input variables.
-But what if we are interested in the uncertainty as a result of the focal input variable only, so that the CIs are $\boldeta \pm q \boldsymbol \sigma_f$, i.e., effects?
+This generates conventional CIs which incorporate all the uncertainties -- including the uncertainties due to the intercept and non-focal predictor.
+But what if we are interested in the uncertainty as a result of the focal predictor only, so that the CIs are $\boldeta \pm q \boldsymbol \sigma_f$, i.e., effects?
 
-\NEW{We may be interested in the uncertainties associated with the focal input variable only, excluding other uncertainties due to other non-focal input variables -- effects.
-However, commonly used \proglang{R} packages for constructing predictions do not exclude the uncertainties resulting from non focal input variables when computing the CIs.} Currently obscure way to exclude uncertainties associated with non-focal input variables in some of these packages is to provide a user defined variance-covariance matrix with the covariances of non-focal predictors set to $0$.
+\NEW{We may be interested in the uncertainties associated with the focal predictor only, excluding other uncertainties due to other non-focal predictors -- effects.
+However, commonly used \proglang{R} packages for constructing predictions do not exclude the uncertainties resulting from non focal predictors when computing the CIs.} Currently obscure way to exclude uncertainties associated with non-focal predictors in some of these packages is to provide a user defined variance-covariance matrix with the covariances of non-focal predictors set to $0$.
 We refer to this procedure as \emph{zeroing-out} variance-covariance matrix; and only works when the input variables are \emph{centered} prior to model fitting, in case of numerical variables, and more complicated when the input variables are categorical.
@@
 This means that $\boldsymbol\sigma_f^2$ close to the anchor point are smaller than those away from the anchor point; and results to confidence intervals which are narrower around the anchor or crosses at the anchor point for simple models.
-This is the effect of the particular focal input variable.
+This is the effect of the particular focal predictor.
 By setting $\baf = \boldsymbol 0$, we get the variances for the conventional CIs in \EREF{conventional_variance}.
@@
 
-An alternative approach to choosing a reference point is to make predictions over all observations of the non-focal input variables (members of the population) \citep{hanmer2013behind}.
+An alternative approach to choosing a reference point is to make predictions over all observations of the non-focal predictors (members of the population) \citep{hanmer2013behind}.
 The nonlinear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population to a one-dimensional distribution of $\eta^\star(x_f, \nset{x})$, which is a function of the chosen values of the focal predictor and the whole sample of non-focal predictors, as opposed to the definition in \EREF{eta_mean}.
@@
 \item compute linear predictor associated with whole sample of the non-focal predictors, $\nset{\eta} = \sum \nset{\beta} \nset{x}$
-\item compute linear predictor associated with the chosen values focal input variable, $\eta_{jf} = \sum{\beta_f x_{jf}}$
+\item compute linear predictor associated with the chosen values focal predictor, $\eta_{jf} = \sum{\beta_f x_{jf}}$
 \item for every value of the focal linear predictor, $\eta_{jf}$, compute
@@
 
-We make similar adjustments to compute the variances of the predictions at every level of the focal input variable:
+We make similar adjustments to compute the variances of the predictions at every level of the focal predictor:
 %
@@
 In some applications, the effect of an average case might not be generalizable to the entire population, especially, if the average does not represent the population.
-This might not be a problem in the whole-sample-based approach since it focuses on specific observations -- the prediction is first obtained for each observation and then averaged across the levels of the focal input variable.
+This might not be a problem in the whole-sample-based approach since it focuses on specific observations -- the prediction is first obtained for each observation and then averaged across the levels of the focal predictor.
 
@@
 However, for complex models, we demonstrate that the two approaches can produce substantially different results and show that averaging the interactions closely matches the observed values.
-To illustrate this, consider models~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher} below, with $x_1$ as the focal input variable:
+To illustrate this, consider models~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher} below, with $x_1$ as the focal predictor:
 %
@@
 The \proglang{R} software package \pkg{margins} \cite{lenth2018package} generates marginal effects.
-To illustrate the distinction between prediction and effects plots, and marginal effects, we use the simulated data described in \nameref{S1_Appendix}.
+To illustrate the distinction between effects and marginal effects, we use the simulated data described in \nameref{S1_Appendix}.
 %
@@
 \includegraphics{cubic_varpred_margins_pred.ggp.pdf}
-\caption{{\bf A comparison of effects and marginal effect plots.} \NEW{A:  The black solid line is the central (outcome) estimate, while the dashed lines corresponds to the effects styled $95\%$ confidence interval.
-B: The marginal effect (ME) at representative values of the focal input variable \code{age} together with the $95\%$ confidence intervals.
-For simple linear models with no interactions, the ME should be constant across all the values of the focal input variable; and it would correspond to the slope of the central estimate of the prediction or effects.
-However, in this example, the focal input variable is a cubic polynomial, and so, the effect is not constant.}}
+\caption{{\bf Effect and marginal effect plots.} \NEW{A:  The effect plot.
+B: The marginal effect (ME) at representative values of the focal predictor \code{age} together with the $95\%$ confidence intervals.
+For simple linear models with no interactions, the ME should be constant across all the values of the focal predictor; and it would correspond to the slope of the central estimate of central estimate.
+However, in this example, the focal predictor is a cubic polynomial, and so, the effect is not constant but a function of the partial derivatives of the corresponding model variables.}}
 \label{fig:qoi_age_pred_plot}
@@
 %
-
-\BC{Got to this point \dots}
 

### 2022-07-07 manuscript/body.tex
Cape Town airport

--- 2022-07-07 manuscript/body.tex
@@
 However, when the focal predictor has interactions or a non-linear link function, non-focal choices can also affect the slope of the response.
-Additional challenges arise when dealing with generalized linear (mixed) models (GL(M)Ms), which extend ordinary linear models to incorporate random effects.
-How we treat random effects can also have an effect.
+Additional challenges arise when dealing with “mixed” models, which incorporate random effects.
 
-The outcome plots described above are often called prediction plots or effects plots.
-We endeavor here to make a conceptual distinction.
-If our goal is to \emph{predict} what we learn about the outcome variable by measuring the focal predictor, \NEW{then we would focus on the central estimate, which is the same for both prediction and effects plots.
-However, the distinction between the two lies in how we describe the uncertainties around the central estimate.
-In particular, we may want to capture the uncertainty in the whole model (these include uncertainty due to intercept, focal and non-focal predictors, and random effects), or we may be interested in showing the uncertainty associated with the focal predictor only.
-In the former case, we use the prediction plots, which generally focus on the \emph{total} effect and will often be suited for a univariate model.
-In the latter case, we use effect plots that focus on the direct effect of the focal predictor -- after controlling for other non-focal predictors -- from a multivariate model, which will be preferable for multivariate models.
-Since effects plots focus on focal-specific uncertainty, they are usually narrower than prediction plots.
+As noted, the outcome plots described above are often called prediction plots or effects plots. We endeavor here to make a conceptual distinction. 
+The primary distinction between the two lies in how we describe the uncertainties around the central estimate. 
+If our goal is to \emph{predict} what we learn about the outcome variable by measuring the focal predictor, then we may want to capture a variety of sources of uncertainty, including that due to the intercept, focal and non-focal predictors, and random effects.
+Conversely, if we wish to focus on the \emph{effect} of a focal predictor only, we might want to isolate uncertainty due to coefficients associated with that predictor.
+If we follow this convention, we expect effects plots to have narrower confidence intervals (CIs) than prediction plots.
+
+The other distinction relates not to the method of calculating CIs, but to the model chosen for the plot. 
+In general, if we want to predict based on a focal parameter, we are likely to want to fit a univariate model, that only contains terms related to that predictor.
+If we want to know the effects of a predictor, we may want to control for covariates with a multivariate model, in order to estimate “direct” effects, or leave covariates out, in order to estimate “total” effects (direct plus indirect), or take an intermediate strategy.
 Shi et al.
-\citep{shi_evidence_2017} used effects plots to compare the difference in sexual risk behaviors between circumcised and uncircumcised men.}
+\citep{shi_evidence_2017} used multivariate effects plots to visualize estimated direct effects of different predictors in a paper that compared the difference in sexual risk behaviors between circumcised and uncircumcised men.
 
@@
 \begin{enumerate}
-\item choice of representative values of focal predictor and the \emph{reference point} for non-focal predictors, especially in multivariate models
-\item uncertainty estimation -- appropriate choice of \emph{anchor} for computing confidence intervals; and how to incorporate the uncertainty due to non-focal predictors 
-\item bias in the expected mean prediction induced by the non-linear transformation of the response variable (especially in GL(M)Ms).
+\item choosing the \emph{reference point} for non-focal predictors in multivariate models
+\item uncertainty estimation -- appropriate choice of \emph{anchor} for computing confidence intervals in effects plots
+\item \jd{When is this a challenge? Doesn't it just work like magic with the lm machinery? and how to incorporate the uncertainty due to non-focal predictors}
+\item biases induced by non-linear transformations of the response variable in generalized models (especially generalized mixed models).
+\item choice of representative values of the focal predictor \jd{I still don't see this as much of a challenge.}
 \end{enumerate}
 
-The most common way of choosing the representative values is taking unique levels or quantiles of the focal predictor if discrete or continuous, respectively.
+Representative values for a focal predictor are generally chosen using quantiles, or equally spaced values, for a continuous predictor, or an exhaustive set of levels for a categorical predictor.
 The central estimates are then calculated by holding the non-focal predictors at their reference point (values chosen for the non-focal predictors, e.g., means) while varying the focal predictor, with the goal that the estimates represent how the model responds to the changes in the focal predictor \citep{fox2009effect, hanmer2013behind}.
@@
 
-Muli-parameter variables (MPVs) such as splines, polynomials, etc., can be within the focal predictor, or within (between) non-focal predictor(s).
+Multi-parameter variables (MPVs) such as splines, polynomials, etc., can be within the focal predictor, or within non-focal predictor(s).
 To distinguish the two, suppose the model which describes the hypothetical simulation of household size based on a number of socio-demographic factors such as age and wealth index is

### 2022-07-08 manuscript/body.tex
GEtting on airplane

@@
 %% BC: use \EREF{} for equations and \FREF{} for figures
+%% JD: Why do you capitalize these? It's scary.
 % Best to use a command that combines the name and the reference
@@
 
-Representative values for a focal predictor are generally chosen using quantiles, or equally spaced values, for a continuous predictor, or an exhaustive set of levels for a categorical predictor.
-The central estimates are then calculated by holding the non-focal predictors at their reference point (values chosen for the non-focal predictors, e.g., means) while varying the focal predictor, with the goal that the estimates represent how the model responds to the changes in the focal predictor \citep{fox2009effect, hanmer2013behind}.
-This procedure generates -- \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}.
-In this article, we refer to these quantities as the \emph{central estimates} for an ``average case'' in a population; the associated confidence intervals will either describe a prediction or effect plot.
-In the effect plots, we can also choose an anchor (value chosen for the focal predictor).
+Representative values for a focal predictor are generally chosen using quantiles, or equally spaced values, for a continuous predictor; or an exhaustive set of levels for a categorical predictor.
+The central estimates are then calculated by holding the non-focal predictors at a reference point while varying the focal predictor, with the goal that the estimates represent how the model responds to the changes in the focal predictor \citep{fox2009effect, hanmer2013behind}.
+These values have been called: \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package} \jd{Is this also the same as least-squares means?}.
+In this article, we refer to these quantities as the \emph{central estimates} of the outcome.
 
+\jd{It looks like you're trying to introduce column averaging and non-linear averaging in the same \P. How can we simplify?}
 For linear models (including models with complex interactions), the averaging is done on the linear scale, i.e., linear averaging.
@@
 
-\jd{I should edit if we can finally consense on nomenclature. I feel it's my fault.}
 \begin{itemize}
@@
 For example, the regression models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher} both have $3$ input variables -- $x_1, x_2, x_3$.
-\item \textbf{Focal and non-focal predictors}: The focal predictor refers to the input variable we are interested in.
-For multivariate models, all other variables other than focal predictors are non-focal predictors.
-For example, consider model described by \EREF{simple_inter_higher_no_interaction}.
-Suppose we are interested in the effect of $x_1$ on the predicted $y$, then $x_1$ is the focal predictor while $x_2$ and $x_3$ are non-focal predictors.
+\item \textbf{Focal predictors}: We call the input variable on the x-axis of an outcome plot the focal predictor. 
+Any other predictor variables are “non-focal” predictors. 
+For example, consider the model described by \EREF{simple_inter_higher_no_interaction}. 
+If we are focusing on the effect of $x_1$ on the response $y$, then $x_1$ is the focal predictor while $x_2$ and $x_3$ are non-focal predictors.
+\jd{Is this trivial example necessary?
+Is any example necessary here?}
+
 \item \textbf{Model matrix}: Refers to the design matrix whose rows include all combination of input variables.
@@
 In our example, the column \code{religion} is missing and instead we have \code{religionJew} and \code{religionMuslim}; the missing category \code{religionChristian} is treated as the reference category.
+
 \item \textbf{Model variables}: Refer to the X-variables which go into the regression model; and represent columns in the model matrix.
@@
 For example, the regression models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher} have $4$ model variables -- $x_1, x_2, x_3$ plus intercept term and $5$ model variables -- $x_1, x_2, x_3, x_2x_3$ plus the intercept term, respectively.
+
 \item \textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more model variables).
@@
 The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable -- the case in \FREF{justify_plots}B.
-In simple linear models (without) interaction, the mean of the input variable can be used as a proxy for the center point.
-If there are more than one model variables associated with the focal input variable, however, the model center does not correspond to any given value of the focal predictor, and we do not necessarily get the ``nice'' crossing we get in simpler cases, see \FREF{pred_cubic_plots}.
+In simple linear models with no interactions, the mean of the input variable can be used as a proxy for the center point \jd{Is it a “proxy” or is it the thing??}.
+If there are more than one model variables associated with the focal input variable, however, the model center does not correspond to any given value of the focal predictor \jd{This seems like bad repetition}, and we do not necessarily get the ``nice'' crossing we get in simpler cases, see \FREF{pred_cubic_plots}.
 \item \textbf{Reference point:} Corresponds to the value or values chosen for non-focal predictors, when estimating the predictions and effects.

### 2022-07-08 manuscript/body.tex
Trying to work on plane

@@
 
+\jd{I guess this list should really be a desc, not an itemize.}
+
 \begin{itemize}
@@
 
-\item \textbf{Model variables}: Refer to the X-variables which go into the regression model; and represent columns in the model matrix.
+\item \textbf{Model variables}: Refer to the variables which are combined to make the linear predictor (corresponding to the columns in the model matrix).
 Each input variable may correspond to one or more model variable.
-In particular, variables with more than two categories (for instance religion in our previous example), or input variables modeled with interactions, a spline or polynomial interactions, will correspond to more than one model variable.
-For example, the regression models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher} have $4$ model variables -- $x_1, x_2, x_3$ plus intercept term and $5$ model variables -- $x_1, x_2, x_3, x_2x_3$ plus the intercept term, respectively.
+Input variables with more than two categories (for instance religion in our previous example), or input variables with non-linear response functions (e.g., spline or polynomial) will correspond to more than one model variable -- we call such variables “multi-parameter variables”.
 
@@
 Also referred to as center point.
-We refer to the prediction approach that uses model center as mean-based.
 The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable -- the case in \FREF{justify_plots}B.
@@
 If there are more than one model variables associated with the focal input variable, however, the model center does not correspond to any given value of the focal predictor \jd{This seems like bad repetition}, and we do not necessarily get the ``nice'' crossing we get in simpler cases, see \FREF{pred_cubic_plots}.
-\item \textbf{Reference point:} Corresponds to the value or values chosen for non-focal predictors, when estimating the predictions and effects.
-Typically the center point (mean-based), but can instead be a whole-sample of quantiles or observations.
-Appropriately choosing the reference point is very crucial in generating correct predictions (for bias correction) especially when dealing with non-linear link functions.
-\item \textbf{Anchor:} The value chosen for the focal predictor when estimating effect confidence intervals.
-The anchor choice does not affect the central estimates.
-Typically chosen as the center point of the model variables corresponding to the focal predictor but other sensible values can be chosen too.
-If we are trying to focus on effects, the anchor from which we calculate confidence intervals becomes important.
-This is not true for the predictions.
-We can choose any anchor, but the center point is a natural and stable choice.
-\item \textbf{Marginal effects:} In simple linear models with no interaction terms, the coefficient estimates are simple and directly interpretable as the expected change in outcome for a unit change in focal input variable.
+
+\item \textbf{Reference point:} The values (or sets of values) chosen for non-focal predictors, when estimating the predictions and effects.
+Typically the center point, but can also be chosen as a baseline value or as a mean across categories.
+We will also discuss using a set of quantiles or observations as a reference.
+
+\item \textbf{Anchor:} The value chosen for the focal predictor when estimating effect-style confidence intervals.
+The anchor choice does not affect the central estimates, nor prediction-style.
+Often chosen as the center point of the model variables corresponding to the focal predictor.
+
+\item \textbf{Marginal effects:} \jd{This isn't really one of our definitions; maybe some of this belongs in the text part.} In simple linear models with no interaction terms, the coefficient estimates are simple and directly interpretable as the expected change in outcome for a unit change in focal input variable.
 This is the \emph{unconditional marginal effect} and it is constant across all the observations and levels of all other input variables.
@@
 In essence, computation of marginal effects requires the use of partial derivatives and can be considered as the slope of a multi-dimensional surface with respect to the surface of interest \cite{leeper2017interpreting}.
+
+\item \jd{Should we consider defining effects-style and prediction-style CIs? Can we think of better words?}
+
 \end{itemize}
@@
 
+\jd{This should be more general.
+The natural way to do the comparison is to first write out the two ways of averaging, and then point out that they're the same when $g$ is the identity.
+In other words, do the general derivation before making the assumption.
+It doesn't ever seem good to write $g$ and then \emph{define} it as the identity.}
 To illustrate mean-based approach, consider a simple linear model with linear predictor $\eta = \bX\bbeta$ and let $g(\boldmu) = \boldeta$ be an identity link function, where $\boldmuh$ is the expected value of response variable $\hat{y}$.

### 2022-07-09 manuscript/body.tex
x-axis labels

--- 2022-07-08 manuscript/body.tex
@@
 
-We now illustrate the construction of outcome plots, compare with other packages and also demonstrate that the mean-based and whole-sample-based approaches produce different results in models with nonlinear link functions and/or additional source(s) of potential bias.
+We start by comparing our proposed approach with the existing implementations and then illustrate the construction of outcome plots and also demonstrate that the mean-based and whole-sample-based approaches produce different results in models with nonlinear link functions and/or additional source(s) of potential bias.
 In addition, we demonstrate that whole-sample-based approach can be used for bias correction.
@@
 
-The \proglang{R} software package \pkg{margins} \cite{lenth2018package} generates marginal effects.
+The \proglang{R} software package \pkg{margins} \citep{lenth2018package} generates marginal effects.
 To illustrate the distinction between effects and marginal effects, we use the simulated data described in \nameref{S1_Appendix}.
@@
 
-To illustrate the distinction between the two approaches, we simulate a multivariate model described in \nameref{S1_Appendix}.
-We also use the same model to illustrate how predictions and effects differ in the presence of higher order interactions in focal or non-focal input variables.
-We then generated two sets of mean-based predictions and effects: 1) one with \code{age} as the focal input variable (a cubic polynomial focal input variable); and 2) one with \code{Wealthindex} as the focal in put variable (linear focal input variable).
-In both cases the, the center point was used as the anchor for the effects, as shown in \FREF{pred_cubic_plots}.
+To illustrate the distinction between predictions and effects in MPVs, we simulated a multivariate model described in \nameref{S1_Appendix}.
+We also used the same model to illustrate how predictions and effects differ if MPVs are in the focal or non-focal predictor.
+In particular, we generated two sets of mean-based predictions and effects: 1) one with \code{age} as the focal predictor (a cubic polynomial focal input variable); and 2) one with \code{Wealthindex} as the focal predictor (linear focal input variable).
+In both cases the, we used center point as the anchor for the effects, as shown in \FREF{pred_cubic_plots}.
 
@@
 \end{center}
-\caption{{\bf Prediction and effect plots for cubic polynomial and simple focal predictors.} A: The focal input variable is a cubic polynomial.
-B: The focal input variable is linear, with polynomial non-focal input variable.
-The central curve and trend lines are the same for predictions and effects plots in each case.
-However, the confidence intervals for the effects are narrower than those of the predictions in both cases.
+\caption{{\bf Prediction and effect plots for cubic polynomial and simple focal predictors.} A: The focal predictor is a MPV cubic polynomial.
+B: The focal input variable is linear, with MPV non-focal predictor.
+The central estimates (dashed central curves or lines) are the same for predictions and effects, in each case (A and B).
+However, the effects are narrower than the predictions.
 The default anchor for generating effects is the center point.
-For simple OLS models (or simple focal input variables), the center point is a point within the range of focal values; and therefore the confidence interval lines intersect at that that point -- B.
-In the case of higher order interactions such as cubic polynomials, the center point is not a point but a combination of all the predictor variables associated with the cubic polynomial, consequently, curves will not necessarily cross at a point -- A.
+For linear (or simple) focal input variables, the center point is a point within the range of focal values; and therefore the effect curves intersect at that that point -- B.
+In the case of MPVs such as cubic polynomials, the center point is not a point but a combination of all the model variables associated with the cubic polynomial, consequently, curves will not necessarily cross at a point -- A.
 The horizontal black and red dashed lines are the observed and predicted average household size, i.e., $\over{hh~size}$ and $\over{\widehat{hh~size}}$, respectively.}
@@
 
-Predictions are the standard curves (wider); they tell us what we expect to see if we know the value of the focal input variable.
-In multivariate models, at times, we may want to focus on the effect of the focal input variable only.
-Unlike prediction plot, which captures the uncertainty in the entire model, effect plot focuses on the uncertainty of the focal input variable only.
-In this case, effects styled plots which are usually narrower around the anchor are preferable over the conventional prediction plots which generally focus on the \emph{total} effects.
-
-In the absence of complex higher order interactions or transformations on the focal input variable, in simple OLS models we would expect the observed average, $\over{hh~size}$, and the average of the predictions, $\over{\widehat{hh~size}}$, to be identical and intersect with the trend lines at the center point, for the effects plot.
-On the other hand, if the focal input variable is characterized by complex interactions or any other form of transformations, the center point is not necessarily a point in the range of focal input values.
+In the absence of, complex interactions, MPVs, or transformations on the focal input variable, in simple OLS models we would expect the observed average, $\over{hh~size}$, and the average of the predictions, $\over{\widehat{hh~size}}$, to be identical and intersect with the trend lines at the center point, for the effect plot.
+On the other hand, if the focal input variable is characterized by complex interactions, MPVs, or any other form of transformations, the center point is not necessarily a point in the range of focal input values.
 In this case, the effects will be narrower than the predictions plots but will not necessarily intersect.
@@
 Mean-based and whole-sample-based approaches can produce very different results in models with nonlinear link functions with additional sources of potential bias such as additional non-focal input variable(s), random effects or complex interactions.
-To demonstrate this, we considered a two predictor binary outcome simulation described in \nameref{S2_Appendix}, such that age has a very small effect size in comparison to wealth index, and compared their effects on the predicted probability of improved water quality as shown in \FREF{pred_bin_plots}.
+To demonstrate this, we considered a two predictor binary outcome simulation described in \nameref{S2_Appendix}, such that age has a very small effect size in comparison to wealth index, and compared their effects on the estimated probability of improved water quality as shown in \FREF{pred_bin_plots}.
 
@@
 \end{center}
-\caption{{\bf Mean-based and whole-sample-based predictions.} The \emph{truth} is the prediction based on the true simulation parameters.
-If the focal input variable has small effect size, mean-based and whole-sample-based approaches produce different predictions -- A.
-In the case of strong effect size, however, the two approaches produce similar predictions -- B.
-In both cases, the whole-sample-based predictions are close to the truth.
-The difference in mean-based and whole-sample-based is due to the bias induced by the non-focal variables and the nonlinear link function in the logistic model.
-Mean-based approach suffers from nonlinear averaging; and if focal input variable has small effect size, the bias is even pronounced since the effect of nonlinear averaging is driven by the non-focal input variable with the strong effect.
-However, this is not the case if the non-focal input variables have small effect since the bias they induce is equally likely to be small.
+\caption{{\bf Mean-based and whole-sample-based central estimates.} The \emph{truth} is based on the true simulation parameters.
+If the focal predictor has a small effect size, mean-based and whole-sample-based approaches produce different estimates -- A.
+In the case of strong effect size, however, the two approaches produce very close estimates -- B.
+In both cases, the whole-sample-based estimates are closer to the truth.
+The difference in mean-based and whole-sample-based is due to the bias induced by the non-focal predictor and the nonlinear link function in the logistic model.
+Mean-based approach suffers from nonlinear averaging; and if focal predictor has small effect size, the bias is even pronounced since the effect of nonlinear averaging is driven by the non-focal predictor(s) with the strong effect.
+However, this is not the case if the non-focal predictor(s) have small effect since the bias they induce is equally likely to be small.
 Since whole-sample-based approach averages over the whole population of the non-focal predictor variables, it is less likely to be affected by the nonlinear averaging for the reasons already mentioned.
 The horizontal dashed lines correspond to the respective averages.
-The vertical dashed lines represent the center point, and the point at which they intersect the horizontal lines represents the expected ``perfect'' prediction at the model center.
+The vertical dashed lines represent the center point, and the point at which they intersect the horizontal lines represents the expected ``perfect'' estimate at the model center.
 The grey points are binned observations -- observed proportions of improved water quality in each bin.}
@@
 
-In \FREF{pred_bin_prediction_effects_plots} below, we zoom into \FREF{pred_bin_plots}A and compare the prediction and effect styled plots.
+In \FREF{pred_bin_prediction_effects_plots} below, we use \FREF{pred_bin_plots}A and compare the prediction and effect plots associated with the central estimates.
 
@@
 \end{center}
-\caption{{\bf Prediction and effect styled plots and bias correction.} The central curves for predictions and effects are the same in the respective case.
-These patterns are  consistent with the ones in \FREF{pred_bin_plots}A with respect to the \emph{truth}.
-Further, the prediction and effect CIs exhibit patterns similar to the ones we previously described and generalizes in a similar way to the ones in simple linear models.
+\caption{{\bf Bias correction for prediction and effect plots.} In each case (A and B), the central estimates for predictions and effects are the same.
+These patterns are  consistent with the ones in \FREF{pred_bin_plots}A compared to the truth.
+Further, the predictions and effects exhibit patterns and generalize in a similar way to the ones in simple linear models.
 }
@@
 \end{center}
-\caption{{\bf Mean-based and whole-sample-based predictions for mediated effects for the mediated (multivariate) and non-mediated (univariate) models.} A: In the absence of mediator variable, both mean-based and whole-sample-based approaches predictions are identical since there is no any other additional (non-focal input variables) sources of bias.
-Consequently, for both approaches, the average predicted probability is very close to the observed proportion.
+\caption{{\bf Mean-based and whole-sample-based estimates for mediated (multivariate) and non-mediated (univariate) models.} A: In the absence of mediator variable, both mean-based and whole-sample-based approaches predictions are identical since there is no any other additional (non-focal predictor) sources of bias.
+Consequently, in both approaches, the average estimated probability is very close to the observed proportion.
 B: When the mediator variable is included,  $x$ has no direct effect on $z$ and as a result the predictions do not align with the observations.
-However, the whole-sample-based approach still closely approximates the average proportion in the simulated data.
-The horizontal dashed lines are the respective average predictions and observed proportions.
-The vertical dashed black represents the center point, and the point at which it crosses the horizontal lines represents the expected ``perfect'' prediction at the model center.
+However, the whole-sample-based approach still closely approximates the average proportion of the simulated data.
+The horizontal dashed lines are the respective average estimates and observed proportions.
+The vertical dashed black represents the center point, and the point at which it crosses the horizontal lines represents the expected ``perfect'' estimate at the model center.
 The grey points are the binned observations.}
@@
 However, if there were additional non-focal predictors, we would expect to see the differences similar to those in \FREF{pred_bin_plots}A.
-Including both $y$ and $x$ ``dilutes'' the direct effect of $x$ on $z$ and as a result, our predictions do not necessarily match the observed binned observations (see \FREF{pred_mediated_plots}B).
+Including both $y$ and $x$ ``dilutes'' the direct effect of $x$ on $z$ and as a result, our estimates do not necessarily match the observed binned observations (see \FREF{pred_mediated_plots}B).
 
@@
 The argument and results we present in this paper support a greater need for a shift on focus on how to present predictions from generalized models.
-For example, the effect-styled confidence intervals could provide more clarity concerning the uncertainty due to the input variable of interests as opposed to the conventional way of incorporating everything.
+For example, effect plots could provide more clarity concerning the uncertainty due to the input variable of interests as opposed to the conventional way of incorporating everything.
 

### 2022-07-09 manuscript/cubic_predictors_preds.R

@@ -49,7 +49,7 @@ pred_age_cubic_plots <- (ggplot(pred_age_cubic_df$preds, aes(x=age, y=fit))
  	+ scale_color_colorblind(limits=col_limits)
  	+ scale_fill_colorblind(limits=col_limits)
  	+ scale_linetype_discrete(limits=col_limits)
- 	+ labs(title="A) Polynomial", y="Predicted household size", linetype="Method", colour="Method", fill="Method")
+ 	+ labs(title="A) Polynomial", y="Household size", linetype="Method", colour="Method", fill="Method")
 )
 print(pred_age_cubic_plots)
 
@@ -90,7 +90,7 @@ pred_wealthindex_cubic_plots <- (ggplot(pred_wealthindex_cubic_df$preds, aes(x=w
  	+ scale_color_colorblind(limits=col_limits)
  	+ scale_fill_colorblind(limits=col_limits)
  	+ scale_linetype_discrete(limits=col_limits)
- 	+ labs(title="B) Simple (linear)", y="Predicted household size", linetype="Method", colour="Method", fill="Method")
+ 	+ labs(title="B) Simple (linear)", y="Household size", linetype="Method", colour="Method", fill="Method")
 )
 print(pred_wealthindex_cubic_plots)
 

### 2022-07-11 manuscript/body.tex
More progress, I guess.

@@
 For linear models (including models with complex interactions), the averaging is done on the linear scale, i.e., linear averaging.
-As a result, the \emph{bias} in the expected (mean) predictions can easily be corrected using \emph{mean-based reference point}, i.e., averaging of the non-focal model variables -- currently not implemented in commonly used \proglang{R} software packages.
+As a result, the \emph{bias} in the expected (mean) predictions can easily be corrected using \emph{mean-based reference point}, i.e., averaging of the non-focal linear predictor variables -- currently not implemented in commonly used \proglang{R} software packages.
 The mean of the central estimates is the prediction at the \emph{model center}.
@@
 The proposed method and \proglang{R} software package will complement the existing ones by providing: 1) a straightforward way to generate effects, and 2) an alternative and a more robust way to correct for prediction bias in GL(M)Ms.
+\jd{We should revisit this list.}
 
@@
 
-\item \textbf{Model variables}: Refer to the variables which are combined to make the linear predictor (corresponding to the columns in the model matrix).
+\item \textbf{Linear predictor variables}: Refer to the variables which are combined to make the linear predictor (corresponding to the columns in the model matrix).
 Each input variable may correspond to one or more model variable.
-Input variables with more than two categories (for instance religion in our previous example), or input variables with non-linear response functions (e.g., spline or polynomial) will correspond to more than one model variable -- we call such variables “multi-parameter variables”.
+Input variables with more than two categories (for instance religion in our previous example), or input variables with non-linear response functions (e.g., spline or polynomial) will correspond to more than one linear predictor variable -- we call such variables “multi-parameter variables”.
 
-\item \textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more model variables).
+\item \textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more linear predictor variables).
 Also referred to as center point.
-The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable -- the case in \FREF{justify_plots}B.
+The center point for a set of linear predictor variables corresponding to an input variable may not represent a possible value of the input variable -- the case in \FREF{justify_plots}B.
 In simple linear models with no interactions, the mean of the input variable can be used as a proxy for the center point \jd{Is it a “proxy” or is it the thing??}.
-If there are more than one model variables associated with the focal input variable, however, the model center does not correspond to any given value of the focal predictor \jd{This seems like bad repetition}, and we do not necessarily get the ``nice'' crossing we get in simpler cases, see \FREF{pred_cubic_plots}.
+If there are more than one linear predictor variables associated with the focal input variable, however, the model center does not correspond to any given value of the focal predictor \jd{This seems like bad repetition}, and we do not necessarily get the ``nice'' crossing we get in simpler cases, see \FREF{pred_cubic_plots}.
 
@@
 The anchor choice does not affect the central estimates, nor prediction-style.
-Often chosen as the center point of the model variables corresponding to the focal predictor.
+Often chosen as the center point of the linear predictor variables corresponding to the focal predictor.
 
@@
 
+\jd{I feel the first s.~is perfectly clear, and I don't know why it needs any explanation at all.}
 Multi-parameter variables (MPVs) such as splines, polynomials, etc., can be within the focal predictor, or within non-focal predictor(s).
@@
 %
-In the first case, with \code{Age} as the focal predictor, is a cubic polynomial with three model variables ($\mathrm{Age}_i, \mathrm{Age}^2_i$ and $\mathrm{Age}^3_i$).
-In this case, each model variable is evaluated separately across the chosen levels of the focal predictor, $Age_i$.
-Specifically, the model variables are treated as additional columns of the model matrix evaluated with the same values chosen for the focal predictor, while non-focal predictors are fixed at their reference point as discussed in the previous section.
+In the first case, with \code{Age} as the focal predictor, is a cubic polynomial with three linear predictor variables ($\mathrm{Age}_i, \mathrm{Age}^2_i$ and $\mathrm{Age}^3_i$).
+In this case, each linear predictor variable is evaluated separately across the chosen levels of the focal predictor, $Age_i$.
+Specifically, the linear predictor variables are treated as additional columns of the model matrix evaluated with the same values chosen for the focal predictor, while non-focal predictors are fixed at their reference point as discussed in the previous section.
 The central estimates associated with \code{Age} on the linear predictor scale become
@@
 In the second case, with \code{Wealthindex} as the focal predictor, the non-focal predictor, \code{Age}, is a cubic polynomial.
-In this case, the non-focal model variables ($\mathrm{Age}_i, \mathrm{Age}^2_i$ and $\mathrm{Age}^3_i$) are all treated as separate non-focal model variables and an appropriate choice of reference point applies just like in the models without MPVs.
+In this case, the non-focal linear predictor variables ($\mathrm{Age}_i, \mathrm{Age}^2_i$ and $\mathrm{Age}^3_i$) are all treated as separate non-focal linear predictor variables and an appropriate choice of reference point applies just like in the models without MPVs.
 For instance, in mean-based approach, we average all non-focal MPVs.
@@
 
-We could also have interactions between focal and non-focal predictors.
+\jd{Why is this under MPV as a subhead?
+Why “also”?
+Do we discuss non-focal interaction?} We could also have interactions between focal and non-focal predictors.
 To illustrate, we consider simple input variables but the logic extends to more complex ones like MPVs.
-Consider model described by \EREF{simple_inter_higher}, used to demonstrate how to handle interactions in non-focal predictors.
+Consider model described by \EREF{simple_inter_higher}, \jd{This is referring back to the definition list, which is a strange place to put examples} used to demonstrate how to handle interactions in non-focal predictors.
 Here, we use it to demonstrate how to handle interactions between the focal, $x_2$, and non-focal, $x_3$, predictors.
-In this case, the non-interacting predictors are treated as non-focal model variables, with appropriate choice of reference points.
-However, for the interacting model variables, we first choose representative values for the focal predictor and then some particular values of the interacting non-focal predictors in a similar way we choose the values of the focal predictor or use predetermined values.
+In this case, the non-interacting predictors are treated as non-focal linear predictor variables, with appropriate choice of reference points \jd{What is the point of saying this again?
+AFAICT this s.~says if we have a non-interacting predictor we treat it as a non-interacting predictor as already described Can we delete?}.
+However, for the interacting linear predictor variables, we first choose representative values for the focal predictor and then some particular values of the interacting non-focal predictors in a similar way we choose the values of the focal predictor or use predetermined values.
 In our example, suppose we pick $i$ and $j$ unique values of the focal predictor, $x_2$ and interacting non-focal predictor, $x_3$, respectively.
@@
 %
+\jd{I don't understand how this part fits in at all.
+Obviously in theory we could pick values of any non-focal predictor, but what's actually going on here?
+What's the actual difference between how we pick reference values for interacting vs.~non-interacting non-focal predictors?}
 
@@
 
-We describe the uncertainty around the estimates using confidence intervals (CIs).
-In principle, every value of focal predictor has a different CI.
-The conventional way to compute variances for predictions is
+We describe the uncertainty around the estimates using confidence intervals (CIs). In principle, every prediction has a different CI. The conventional way to compute variances for predictions is 
 %
 \begin{align}\label{eq:conventional_variance}
-\boldsymbol\sigma^2 = \textrm{Diag}(\bX^\star \boldsymbol{\Sigma} \bX^{\star\top}), 
+\boldsymbol\sigma^2 = \mathrm{Diag}(\bX^\star \boldsymbol{\Sigma} \bX^{\star\top}), 
 \end{align}
+\jd{I don't understand how a scalar ($\sigma^2$) is equal to a vector (the $\mathrm{Diag}$). Is there a citation for this formula?}
 so that the confidence intervals are $\boldeta \pm q\boldsymbol\sigma$, where $q$ is an appropriate quantile of Normal or t distribution \citep{lenth2018package, fox2009effect}.
@@
 
+\jd{Haven't we already discussed this idea?
+The new sentence reads like we didn't.
+We should start by saying what we want, then continue by explaining that there is an equivalent calculation (is it always equivalent?) that can be done in the packages.}
 \NEW{We may be interested in the uncertainties associated with the focal predictor only, excluding other uncertainties due to other non-focal predictors -- effects.
@@
 
+\jd{You explain it like the 0 variance at the center is the point.
+The point is to look at the effect of changing the focal parameter from the anchor value; the 0 variance at the anchor value is a corollary.
+You should start by explaining why something like \EREF{centered_variance} is what you want, and then explain the zero variance.}
+\jd{The explanation of \bxo is not clear.
+I guess you mean that the values for everything except for the focal columns are centered?
+What is the point of defining \bxo\ and then defining \bafc\ so that it exactly cancels?
+Is it better to just define a matrix with zeroes in all the right places?}
 Let $\bxo$ be a centered model matrix previously defined, and let $\ba$ be an anchor matrix, with the same dimensions and entries in all non-focal predictors as $\bxo$.
@@
 This means that we can still generate effects without necessarily centering the predictors prior to model fitting.
+\jd{What is the point of this s.?
+What does it mean “impacts only on”?}
 
@@
 
+\jd{Why at the model center? Aren't we interested in comparisons at the reference point for any value of the focal variable?}
 We can directly compare the outcome estimate at the model center (mean-based estimate) with the average of predictions evaluated across the population values of non-focal predictors (sample-based estimate):
@@
 In an ordinary linear model, the link function $g$ is the identity function, so the two means are the same.
-For non-trivial link functions, we expect them to be different in general -- \NEW{a consequent of Jensen inequality \citep{duursma_bias_2003, denny_performance_2019, ye_generalized_2021}.
+For non-trivial link functions, we expect them to be different in general -- \NEW{a consequent of Jensen inequality \citep{duursma_bias_2003, denny_performance_2019, ye_generalized_2021} \jd{Drop this claim; it's obvious we expect them to be different, Jensen is just about the sign}.
 In GL(M)Ms the assumed link function, predetermines the curvature of the response surface and also the sign of the Jensen effect.
@@
 In non-trivial link functions, the Jensen effect may depend on the additional input variables \citep{ye_generalized_2021}}.
+\jd{I think I asked you to describe how Jensen is working in these cases.
+In each of the cases it would be both easier and clearer to say whether we expect the left side or the right side to be larger than to spin all of those words about “describes how the changes in input values elevate the predictions and otherwise describes how these changes depress the predictions”.}
 
-\NEW{It is usually important to report the estimates that reflect the expected values of the untransformed response.
-This is not a major issue in simple linear models.
+\NEW{It is usually important to report the estimates that reflect the expected values of the untransformed response \jd{Hard to understand this s.}.
+This is not a major issue in simple linear models \jd{What do you mean by major issue?
+In what sense is it an issue?
+In what sense does it exist?}.
 However, when dealing with nonlinear link functions,  generated predictions may not reflect the observed response due to the bias in the expected mean induced by the nonlinear transformation of the response variable -- a consequent of Jensen effect mentioned above.
@@
 Here, we describe and implement a different approach -- whole-sample-based approach for bias correction.
-Hammer and Kalkan \citep{hanmer2013behind} introduced mean- and whole-sample- based approaches but limited their discussion to binary response models only.}
+Hanmer and Kalkan \citep{hanmer2013behind} introduced mean- and whole-sample- based approaches but limited their discussion to binary response models only.}
+\jd{We mostly talk about binary-response models as well. Also, they did not introduce (there are much older citations to it, and they use a different name, which we should mention (or maybe adopt).}
 
@@
 The most commonly used \proglang{R} software packages for outcome plots (\pkg{emmeans} and \pkg{effects}), by default, use the average of input variables and ``sum-to-zero'' as the reference point approach for continuous and categorical input variables, respectively.
-However, there are a number of choices one can make when constructing outcome plots -- for example, in the presence of interactions, averaging the interactions (averaging the product of model variables) versus taking the product of the averages of the input variables (the default for \pkg{emmeans} and \pkg{effects}).
+However, there are a number of choices one can make when constructing outcome plots -- for example, in the presence of interactions, averaging the interactions (averaging the product of linear predictor variables) versus taking the product of the averages of the input variables (the default for \pkg{emmeans} and \pkg{effects}).
 For simple OLS models (with no interactions), the two approaches give the same estimates.
@@
 In particular, \pkg{emmeans} and \pkg{effects} have similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is very close to the simulated average ($\bar{y}$), \FREF{justify_plots}B.
-To generate \FREF{justify_plots}A,  \pkg{emmeans} and \pkg{effects} average $x_2$ and $x_3$, i.e., the input variables while \pkg{varpred} averages the model variables corresponding to $x_2$ and $x_3$.
-On other hand, to generate \FREF{justify_plots}B, the difference in the estimates lies on how each of the packages average the interaction model variable ($x_2x_3$).
-In particular, \pkg{emmeans} and \pkg{effects} use input variables to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses model variables to compute $\over{x_2x_3}$.
-In the case of \FREF{justify_plots}A, averaging the input variables is equivalent to averaging the model variables.
+To generate \FREF{justify_plots}A,  \pkg{emmeans} and \pkg{effects} average $x_2$ and $x_3$, i.e., the input variables while \pkg{varpred} averages the linear predictor variables corresponding to $x_2$ and $x_3$.
+On other hand, to generate \FREF{justify_plots}B, the difference in the estimates lies on how each of the packages average the interaction linear predictor variable ($x_2x_3$).
+In particular, \pkg{emmeans} and \pkg{effects} use input variables to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses linear predictor variables to compute $\over{x_2x_3}$.
+In the case of \FREF{justify_plots}A, averaging the input variables is equivalent to averaging the linear predictor variables.
 
@@
 For simple linear models with no interactions, the ME should be constant across all the values of the focal predictor; and it would correspond to the slope of the central estimate of central estimate.
-However, in this example, the focal predictor is a cubic polynomial, and so, the effect is not constant but a function of the partial derivatives of the corresponding model variables.}}
+However, in this example, the focal predictor is a cubic polynomial, and so, the effect is not constant but a function of the partial derivatives of the corresponding linear predictor variables.}}
 \label{fig:qoi_age_pred_plot}
@@
 For linear (or simple) focal input variables, the center point is a point within the range of focal values; and therefore the effect curves intersect at that that point -- B.
-In the case of MPVs such as cubic polynomials, the center point is not a point but a combination of all the model variables associated with the cubic polynomial, consequently, curves will not necessarily cross at a point -- A.
+In the case of MPVs such as cubic polynomials, the center point is not a point but a combination of all the linear predictor variables associated with the cubic polynomial, consequently, curves will not necessarily cross at a point -- A.
 The horizontal black and red dashed lines are the observed and predicted average household size, i.e., $\over{hh~size}$ and $\over{\widehat{hh~size}}$, respectively.}

### 2022-07-11 manuscript/body.tex
Little beak :( after intro

--- 2022-07-10 manuscript/body.tex
@@
 
-To make an outcome plot, we use a \emph{focal} predictor on the x-axis.
+To make an outcome plot, we use a \emph{focal} predictor on the x-axis and central estimate (predicted values) on the y-axis.
 The resulting plot will depend on choices we make about other (non-focal) predictor(s).
-In ordinary (Gaussian) linear regression, the non-focal choices may have a simple additive effect on the outcome estimate.
-However, when the focal predictor has interactions or a non-linear link function, non-focal choices can also affect the slope of the response.
+In ordinary (Gaussian) linear regression, the non-focal choices may have a simple additive effect on the central estimate.
+However, when the focal predictor has interactions or a non-linear link function, non-focal choices can also affect the slope of the estimate.
 Additional challenges arise when dealing with “mixed” models, which incorporate random effects.
@@
 \item uncertainty estimation -- appropriate choice of \emph{anchor} for computing confidence intervals in effects plots
-\item \jd{When is this a challenge? Doesn't it just work like magic with the lm machinery? and how to incorporate the uncertainty due to non-focal predictors}
+\item \jd{When is this a challenge?
+Doesn't it just work like magic with the lm machinery?
+and how to incorporate the uncertainty due to non-focal predictors} \BC{Clarify with JD.
+Another challenge or a continuation  of the above?}
 \item biases induced by non-linear transformations of the response variable in generalized models (especially generalized mixed models).
-\item choice of representative values of the focal predictor \jd{I still don't see this as much of a challenge.}
+\item choice of representative values of the focal predictor \jd{I still don't see this as much of a challenge.} \dya{We needed to mention somewhere that we also have to choose focal values.
+I think \P~ below is suffecient.}
 \end{enumerate}
@@
 Representative values for a focal predictor are generally chosen using quantiles, or equally spaced values, for a continuous predictor; or an exhaustive set of levels for a categorical predictor.
-The central estimates are then calculated by holding the non-focal predictors at a reference point while varying the focal predictor, with the goal that the estimates represent how the model responds to the changes in the focal predictor \citep{fox2009effect, hanmer2013behind}.
-These values have been called: \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package} \jd{Is this also the same as least-squares means?}.
+The central estimates are then calculated by holding the non-focal predictors at a reference point (a value chosen for a non-focal predictor) while varying the focal predictor, with the goal that the estimates represent how the model responds to the changes in the focal predictor \citep{fox2009effect, hanmer2013behind}.
+These values have been called: \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package} \jd{Is this also the same as least-squares means?} \BC{Yes.
+But was later renamed to emmeans}.
 In this article, we refer to these quantities as the \emph{central estimates} of the outcome.
 
-\jd{It looks like you're trying to introduce column averaging and non-linear averaging in the same \P. How can we simplify?}
-For linear models (including models with complex interactions), the averaging is done on the linear scale, i.e., linear averaging.
-As a result, the \emph{bias} in the expected (mean) predictions can easily be corrected using \emph{mean-based reference point}, i.e., averaging of the non-focal model variables -- currently not implemented in commonly used \proglang{R} software packages.
+In models with non-focal predictors such as multivariate models, reference point can be chosen as the average of the non-focal \emph{model variables} -- we call this approach \emph{mean-based} reference point and is currently not implemented in commonly used \proglang{R} software packages.
 The mean of the central estimates is the prediction at the \emph{model center}.
-In this case, the estimates are usually consistent with the observed values.
-However, in models with non-linear link functions, this is not always the case.
-In particular, when dealing with GL(M)Ms with non-linear link functions, the averaging is done on the non-linear scale, i.e., non-linear averaging, making generating correct predictions much harder.
+We introduce an alternative choice for the reference point below.
+
+For linear models, the averaging is done on the linear scale, i.e., linear averaging.
+As a result, the \emph{bias} in model center estimates can easily be corrected using mean-based approach and the estimates are usually consistent with the observed values.
+However, in models with non-linear link functions, this is not always the true.
+In particular, when dealing with generalized models with non-linear link functions, the averaging is done on the non-linear scale, i.e., non-linear averaging, making generating correct predictions much harder.
 This is the bias in the expected mean prediction induced by the non-linear transformation of the response variable.
@@
 This article aims to discuss and implement various approaches for computing predictions and effects, together with the associated plots.
-We further explore and demonstrate, using simulated data, approaches for correcting bias in central estimates for GL(M)Ms involving non-linear link functions.
-The proposed method and \proglang{R} software package will complement the existing ones by providing: 1) a straightforward way to generate effects, and 2) an alternative and a more robust way to correct for prediction bias in GL(M)Ms.
+We further explore and demonstrate, using simulated data, approaches for correcting bias in central estimates for generalized (mixed) models involving non-linear link functions.
+The proposed method and \proglang{R} software package will complement the existing ones by providing: 1) a straightforward way to generate effects, and 2) an alternative and a more robust way to correct for prediction bias in generalized (mixed) models.
 
@@
 
-\jd{I guess this list should really be a desc, not an itemize.}
+\jd{I guess this list should really be a desc, not an itemize.}\BC{Clarify}
 
 \begin{itemize}
-\item \textbf{Input variables}: Refers to the observed (or scientific) variables underlying an inference or exploration.
+\item \textbf{Input variables:} Refers to the observed (or scientific) variables underlying an inference or exploration.
 For example, the regression models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher} both have $3$ input variables -- $x_1, x_2, x_3$.
-\item \textbf{Focal predictors}: We call the input variable on the x-axis of an outcome plot the focal predictor. 
-Any other predictor variables are “non-focal” predictors. 
-For example, consider the model described by \EREF{simple_inter_higher_no_interaction}. 
-If we are focusing on the effect of $x_1$ on the response $y$, then $x_1$ is the focal predictor while $x_2$ and $x_3$ are non-focal predictors.
-\jd{Is this trivial example necessary?
-Is any example necessary here?}
+\item \textbf{Focal predictors}: We call the input variable on the x-axis of an outcome plot the focal predictor.  Any other predictor variables are ``non-focal'' predictors.
 
-\item \textbf{Model matrix}: Refers to the design matrix whose rows include all combination of input variables.
+\item \textbf{Model matrix:} Refers to the design matrix whose rows include all combination of input variables.
 Consider an example for three hypothetical households -- the first household head is a Christian with an income of $\$ 50$, the second household head is Muslim with an income of $\$ 100$ while the third household head is a Jew with an income of $\$ 77$.
@@
 
-\item \textbf{Model variables}: Refer to the variables which are combined to make the linear predictor (corresponding to the columns in the model matrix).
+\item \textbf{Model variables:} Refer to the variables which are combined to make the linear predictor (corresponding to the columns in the model matrix).
 Each input variable may correspond to one or more model variable.
@@
 The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable -- the case in \FREF{justify_plots}B.
-In simple linear models with no interactions, the mean of the input variable can be used as a proxy for the center point \jd{Is it a “proxy” or is it the thing??}.
-If there are more than one model variables associated with the focal input variable, however, the model center does not correspond to any given value of the focal predictor \jd{This seems like bad repetition}, and we do not necessarily get the ``nice'' crossing we get in simpler cases, see \FREF{pred_cubic_plots}.
+In simple linear models with no interactions, the mean of the input variable can be used is a center point.
+If there are more than one model variables associated with the focal input variable, however, the model center does not correspond to a single value of the focal predictor, see \FREF{pred_cubic_plots}.
 
@@
 
-\item \textbf{Marginal effects:} \jd{This isn't really one of our definitions; maybe some of this belongs in the text part.} In simple linear models with no interaction terms, the coefficient estimates are simple and directly interpretable as the expected change in outcome for a unit change in focal input variable.
-This is the \emph{unconditional marginal effect} and it is constant across all the observations and levels of all other input variables.
-Consider models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher}.
-The marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher_no_interaction} is $\frac{\partial y}{\partial x_2} = \beta_2$.
-On the other hand, the marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher} is given by $\frac{\partial y}{\partial x_2} = \beta_2 + \beta_{23}x_3$.
-In other words, if there are no interactions, the marginal effect of $x_2$ on $y$ is constant, while, if there are interactions in the model, the marginal effect of a change in $x_2$ on $y$ depends on the value of the other \emph{conditioning} predictor, $x_3$.
-In essence, computation of marginal effects requires the use of partial derivatives and can be considered as the slope of a multi-dimensional surface with respect to the surface of interest \cite{leeper2017interpreting}.
-
-\item \jd{Should we consider defining effects-style and prediction-style CIs? Can we think of better words?}
+\item \textbf{Prediction-style and effect-style plots:} A prediction-style plot is about predicting observations for a given value of the focal predictor.
+For this, we want to use the classic curved confidence intervals.
+An effect-style plot attempts to visualize the effect of a focal predictor and are characterized with narrower confidence intervals.
+Unlike a prediction-style plot, where the confidence intervals capture the uncertainty associated all predictors in the model, the effect-style plot focuses on uncertainty associated with the focal predictor only and depend on the anchor.
 
@@
 For simple OLS models, the point where the curves cross is the model center or simply center point; it corresponds to anchor we choose, in this case, the center point (which is generally an appropriate and stable choice).
-Predictions are the standard curves; they tell us what we expect to see if we know the value of the focal predictor.
-The lines inside (effects) are a little different; they are a good way to visualize the effect of the focal predictor.
-Unlike predictions, where the uncertainty captures the uncertainty in the whole model, the effects plot focuses on uncertainty in the effect of the focal predictor only and depend on the anchor.
+
+\BC{To clean%
+\textbf{Marginal effects:} \jd{This isn't really one of our definitions; maybe some of this belongs in the text part.} In simple linear models with no interaction terms, the coefficient estimates are simple and directly interpretable as the expected change in outcome for a unit change in focal input variable.
+This is the \emph{unconditional marginal effect} and it is constant across all the observations and levels of all other input variables.
+Consider models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher}.
+The marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher_no_interaction} is $\frac{\partial y}{\partial x_2} = \beta_2$.
+On the other hand, the marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher} is given by $\frac{\partial y}{\partial x_2} = \beta_2 + \beta_{23}x_3$.
+In other words, if there are no interactions, the marginal effect of $x_2$ on $y$ is constant, while, if there are interactions in the model, the marginal effect of a change in $x_2$ on $y$ depends on the value of the other \emph{conditioning} predictor, $x_3$.
+In essence, computation of marginal effects requires the use of partial derivatives and can be considered as the slope of a multi-dimensional surface with respect to the surface of interest \cite{leeper2017interpreting}.%
+}
 

### 2022-07-13 manuscript/body.tex
To start on e^x

@@
 
+\jd{We should revisit this list.}
+
 \section{Definitions}
@@
 
-\item \textbf{Model variables:} Refer to the variables which are combined to make the linear predictor (corresponding to the columns in the model matrix).
+\item \textbf{Linear predictor variables}: Refer to the variables which are combined to make the linear predictor (corresponding to the columns in the model matrix).
 Each input variable may correspond to one or more model variable.
-Input variables with more than two categories (for instance religion in our previous example), or input variables with non-linear response functions (e.g., spline or polynomial) will correspond to more than one model variable -- we call such variables “multi-parameter variables”.
+Input variables with more than two categories (for instance religion in our previous example), or input variables with non-linear response functions (e.g., spline or polynomial) will correspond to more than one linear predictor variable -- we call such variables “multi-parameter variables”.
 
@@
 
-\section{Statistical background}
+\section{Statistical formulation}
 
@@
 It doesn't ever seem good to write $g$ and then \emph{define} it as the identity.}
-To illustrate mean-based approach, consider a simple linear model with linear predictor $\eta = \bX\bbeta$ and let $g(\boldmu) = \boldeta$ be an identity link function, where $\boldmuh$ is the expected value of response variable $\hat{y}$.
-Let $\bbetah$ be the estimate of $\bbeta$, together with the estimated covariance matrix $\Sigma = V(\bbetah)$ of $\bbetah$.
-Then the quantity $\hat{\boldeta}^\star = \bX^\star\bbetah$ is the central estimate with respect to the focal predictor in question \citep{fox2009effect}; where $\bX^\star$ is the centered model matrix constructed by averaging the columns of non-focal predictors in model matrix $\bX$, and together with appropriately chosen values of focal predictor.
 
-An alternative formulation of $\boldetah^\star$ involves expressing the linear predictor as the sum of the focal and non-focal predictors' linear predictors. In particular,
+\BC{Trying to adopt new notation style, is this better?}
 
+To estimate the central estimate of an outcome plot on the response scale, we need a link function $g$ and appropriate values of the focal predictor $X^{f}$ and reference point for the non-focal predictors $\uset{{\bX}}$.
+As previously mentioned, we can either use mean-based or whole-sample-based reference point.
+
+Using mean-based reference point, the central estimate is
+%
+\begin{align}\label{eq:eta_linear_mean}
+\etahfi &= \bbetah \bX^{f}_c \nonumber\\
+\yfi &= g^{-1}(\etahfi),
+\end{align}
+%
+where $\bX^{f}_c = \{X^{f}, \uset{{\bX}}_c\}$ is a centered model matrix with appropriately chosen values of the focal predictor $X^{f}$ and centered non-focal linear variables $\uset{{\bX}}_c$ constructed by replacing the entries in the corresponding columns with their averages.
+
+On the other hand, using whole-sample-based reference point, \EREF{eta_linear_mean}
+becomes
+%
+\begin{align}\label{eq:eta_linear_sample}
+\etahfj &= \bbetah \bX^{f}_j \nonumber\\
+\yfi  &= \underset{f}{\textrm{mean}} ~ g^{-1} (\hat{\eta}^\star_j),
+\end{align}
+%
+where $\bX^{f}_j = \{X^{f}_j, \uset{{\bX}}\}$ is the model matrix of the $j$th
+observation and $\uset{{\bX}}$ is the entire population of the non-focal linear
+predictors.
+Here, we generate a vector of size $J \times N$ and then average over the values of the focal predictor; $J$ and $N$ represents the number of focal values and population of non-focal predictors, respectively.
+
+For identity link function, e.g., in a simple linear model, \EREF{eta_linear_mean} and \EREF{eta_linear_sample} are equivalent.
+In the subsequent sections, we will first discuss general formulation of mean-based approach and how to generate associated confidence intervals.
+Thereafter, we will discuss the second approach, who-sample-based, and its application to bias correction.
+
+\subsection{Mean-based approach}
+
+An alternative formulation of \EREF{eta_linear_mean} involves expressing the linear predictor as the sum of the focal and non-focal predictors' linear predictors. In particular, 
+%
 \begin{align}\label{eq:eta_mean}
-\eta^\star(x^\star_f, \nset{{\bar{x}^\star}}) &= \hat{\beta}_f x^\star_f + \sum \nset{\hat{\beta}} \nset{{\bar{x}^\star}} \\
-\hat{y}_f  &= g^{-1} \left(\eta^\star(x^\star_f, \nset{{\bar{x}^\star}})\right)
+\etahfi(x^f, \uset{{\bar{x}}}) &= \hat{\beta}^f x^f + \sum \uset{\hat{\beta}} \uset{{\bar{x}}} \\
+\yfi  &= g^{-1} \left(\eta_i(x^f, \uset{{\bar{x}}})\right)
 \end{align}
-where $\nset{{\bar{x}^\star}}$ and $x^\star_f$ are columns of $\bX^\star$ corresponding to the non-focal predictors and the focal predictor, respectively.
+where $x^f$ and $\uset{{\bar{x}}}$ are columns of $\bX^{f}_c$ focal predictor and non-focal predictors, respectively.
 
-\subsection{Dealing with multi-parameter variables}
+In a model with a multi-parameter variable or input variables with complex interactions, construction of $\etahfi(x^f, \uset{{\bar{x}}})$ is not usually straightforward since we want $\uset{{\bar{x}}}$ to be or represent the ``true'' model center.
+We therefore illustrate some of these cases.
+
+\subsubsection{Dealing with multi-parameter variables}
+
+\jd{I feel the first s.~is perfectly clear, and I don't know why it needs any explanation at all.}
+\BC{I suggest we keep the illustrations.}
 
@@
 \begin{align}
-\eta^\star(\mathrm{Age}_i, \nset{{\over{\mathrm{Wealthindex}}}}) &= \beta_0 + \beta_{\mathrm{A_1}}\mathrm{Age}_i + \beta_{\mathrm{A_2}}\mathrm{Age}^2_i + \beta_{\mathrm{A_3}}\mathrm{Age}^3_i \nonumber\\
+\etahfi(\mathrm{Age}_i, \nset{{\over{\mathrm{Wealthindex}}}}) &= \beta_0 + \beta_{\mathrm{A_1}}\mathrm{Age}_i + \beta_{\mathrm{A_2}}\mathrm{Age}^2_i + \beta_{\mathrm{A_3}}\mathrm{Age}^3_i \nonumber\\
 	& + \beta_{\mathrm{W}}\over{\mathrm{Wealthindex}}.
@@
 %
-\begin{align}
-\eta^\star(\mathrm{Wealthindex}_i, \nset{{\{\over{\mathrm{Age\mathop{\vphantom{^2}}}}, \over{\mathrm{Age}^2}, \over{\mathrm{Age}^2}\}}}) &= \beta_0 + \beta_{\mathrm{A_1}}\over{\mathrm{Age\mathop{\vphantom{^2}}}} + \beta_{\mathrm{A_2}}\over{\mathrm{Age}^2} + \beta_{\mathrm{A_3}}\over{\mathrm{Age}^3}\nonumber\\
+\begin{align}\label{eq:mpv_second}
+\etahfi(\mathrm{Wealthindex}_i, \nset{{\{\over{\mathrm{Age\mathop{\vphantom{^2}}}}, \over{\mathrm{Age}^2}, \over{\mathrm{Age}^2}\}}}) &= \beta_0 + \beta_{\mathrm{A_1}}\over{\mathrm{Age\mathop{\vphantom{^2}}}} + \beta_{\mathrm{A_2}}\over{\mathrm{Age}^2} + \beta_{\mathrm{A_3}}\over{\mathrm{Age}^3}\nonumber\\
 	& + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i.
@@
 
-We could also have interactions between focal and non-focal predictors.
-To illustrate, we consider simple input variables but the logic extends to more complex ones like MPVs.
-Consider model described by \EREF{simple_inter_higher}, used to demonstrate how to handle interactions in non-focal predictors.
-Here, we use it to demonstrate how to handle interactions between the focal, $x_2$, and non-focal, $x_3$, predictors.
-In this case, the non-interacting predictors are treated as non-focal model variables, with appropriate choice of reference points.
-However, for the interacting model variables, we first choose representative values for the focal predictor and then some particular values of the interacting non-focal predictors in a similar way we choose the values of the focal predictor or use predetermined values.
-In our example, suppose we pick $i$ and $j$ unique values of the focal predictor, $x_2$ and interacting non-focal predictor, $x_3$, respectively.
+\subsubsection{Dealing with interactions in input variables}
+
+Interactions can be between non-focal predictor or between focal and non-focal
+predictors. Handling former case is similar to that of the second case
+in MPVs (\EREF{mpv_second}).
+In the latter case, consider model described by \EREF{simple_inter_higher}.
+The interaction is between the focal, $x_2$, and non-focal, $x_3$, predictors.
+In this case, we chose the values of the focal predictors as previously described and the reference point for the interacting non-focal predictor is predetermined or appropriately chosen set of values.
+In our example, suppose we pick $i$ and $j$ unique values of the focal predictor, $x_2$, and interacting non-focal predictor, $x_3$, respectively.
 The central estimate on linear predictor scale is given by
@@
 \begin{align*}
-\eta^\star(x_{2i}, x_{3j}, {\bar{x}^\star_1}) = \beta_0 + \beta_1 \bar{x}^\star_1 + \beta_2x_{2i} + \beta_3x_{3j} + \beta_{23}x_{2i}x_{3j}.
+\etahfi(x_{2i}, x_{3j}, {\uset{\bar{x}}_1}) = \beta_0 + \beta_1 \uset{\bar{x}}_1 + \beta_2x_{2i} + \beta_3x_{3j} + \beta_{23}x_{2i}x_{3j}.
 \end{align*}
 %
+The main point is that, in the case of non-interacting non-focal predictors, the reference point is a center point while in the case of interacting non-focal predictors, the choice of the reference point is not necessarily a center point but can be any appropriate value or set of values.
 
-\section{Uncertainty estimation}
+\subsection{Uncertainty estimation}
 
-We describe the uncertainty around the estimates using confidence intervals (CIs).
-In principle, every value of focal predictor has a different CI.
-The conventional way to compute variances for predictions is
+We describe the uncertainty around the estimates using confidence intervals (CIs). In principle, every prediction has a different CI. The conventional way to compute variances for predictions is 
 %
 \begin{align}\label{eq:conventional_variance}
-\boldsymbol\sigma^2 = \textrm{Diag}(\bX^\star \boldsymbol{\Sigma} \bX^{\star\top}), 
+\sigma^2_i = \mathrm{Diag}(\bXstar \boldsymbol{\Sigma} {\bXstar}^\top), 
 \end{align}
-so that the confidence intervals are $\boldeta \pm q\boldsymbol\sigma$, where $q$ is an appropriate quantile of Normal or t distribution \citep{lenth2018package, fox2009effect}.
+%
+so that the confidence intervals are $\etahfi \pm q\sigma_i$, where $\boldsymbol\Sigma$ is the variance matrix of $\bbetah$ and $q$ is an appropriate quantile of Normal or t distribution \citep{fox2009effect}.
 This generates conventional CIs which incorporate all the uncertainties -- including the uncertainties due to the intercept and non-focal predictor.
-But what if we are interested in the uncertainty as a result of the focal predictor only, so that the CIs are $\boldeta \pm q \boldsymbol \sigma_f$, i.e., effects?
+We call this prediction-style CIs.
 
-\NEW{We may be interested in the uncertainties associated with the focal predictor only, excluding other uncertainties due to other non-focal predictors -- effects.
-However, commonly used \proglang{R} packages for constructing predictions do not exclude the uncertainties resulting from non focal predictors when computing the CIs.} Currently obscure way to exclude uncertainties associated with non-focal predictors in some of these packages is to provide a user defined variance-covariance matrix with the covariances of non-focal predictors set to $0$.
-We refer to this procedure as \emph{zeroing-out} variance-covariance matrix; and only works when the input variables are \emph{centered} prior to model fitting, in case of numerical variables, and more complicated when the input variables are categorical.
-An alternative is our proposed approach which uses the model center and does not require the input variables to be centered prior to model fitting.
+\subsubsection{Effect-style CIs}
+
+But what if we are interested in the uncertainty as a result of the focal predictor only (effect-style CIs), so that the CIs are $\etahfi \pm q\sigma^f_i$, i.e., effects?
+For effect-style CIs, we need an anchor and centered model matrix with all non-focal predictor variables set to zero.
 
@@
 \begin{align}\label{eq:centered_variance}
-\boldsymbol\sigma_f^2 = \textrm{Diag}((\bxo - \bafc) \boldsymbol{\Sigma} (\bxo - \bafc)^\top).
+\boldsymbol\sigma^2_i = \textrm{Diag}((\bxo - \bafc) \boldsymbol{\Sigma} (\bxo - \bafc)^\top).
 \end{align}
 %
-We can see that $\forall ~\bxof=\baf$, $\bxo - \baf = \boldsymbol 0$, hence $\boldsymbol\sigma_f^2 = \boldsymbol{0}$.
-Similarly, for all values of $\bxof$ close to anchor point $\bafc$, the term $(\bxo - \bafc)$ and $\boldsymbol\sigma_f^2$ goes to $\boldsymbol 0$ and $\boldsymbol\sigma_f^2 = \boldsymbol 0$ if $\bxof=\bafc$.
-This means that $\boldsymbol\sigma_f^2$ close to the anchor point are smaller than those away from the anchor point; and results to confidence intervals which are narrower around the anchor or crosses at the anchor point for simple models.
-This is the effect of the particular focal predictor.
-By setting $\baf = \boldsymbol 0$, we get the variances for the conventional CIs in \EREF{conventional_variance}.
+We can see that $\forall ~\bxof=\baf$, $\bxo - \baf = \boldsymbol 0$, hence $\boldsymbol\sigma_i^2 = \boldsymbol{0}$.
+Similarly, for all values of $\bxof$ close to anchor point $\bafc$, the term $(\bxo - \bafc)$ and $\boldsymbol\sigma_i^2$ goes to $\boldsymbol 0$ and $\boldsymbol\sigma_i^2 = \boldsymbol 0$ if $\bxof=\bafc$.
+This means that $\boldsymbol\sigma_i^2$ close to the anchor point are smaller than those away from the anchor point; and results to confidence intervals which are narrower around the anchor or crosses at the anchor point for simple models.
+In other words, this shows the effect of changing the focal value from the anchor value.
+By setting $\baf = \boldsymbol 0$, we get the variances for the prediction-style CIs in \EREF{conventional_variance}.
 
-The computation of $\bxo - \bafc$ impacts only on the intercepts and non-focal predictors, i.e., the slopes and variance corresponding to the focal predictors are not affected.
-This means that we can still generate effects without necessarily centering the predictors prior to model fitting.
+An alternative way to compute $\boldsymbol\sigma^2_i$ in \EREF{centered_variance} is
+by \emph{zeroing-out} the covariance matrix, which involves setting all the non-focal
+linear predictor variables in $\Sigma$ to $0$. This procedure can be implemented 
+in commonly used \proglang{R} packages for effect plots and prediction plots, but only
+works when the input variables are \emph{centered} prior to model fitting, in case of
+numerical variables, and more complicated when the input variables are categorical. 
+The anchor approach, however, does not require the input variables to be centered
+prior to model fitting since the computation of $\bxo - \bafc$ affects only the 
+intercepts and non-focal predictor linear variables -- the slopes and variance
+corresponding to the focal predictor linear variables are not affected.
 
@@
 
-We can directly compare the outcome estimate at the model center (mean-based estimate) with the average of predictions evaluated across the population values of non-focal predictors (sample-based estimate):
+We can directly compare the outcome estimate at the model center for non-focal parameters (mean-based estimate) with the average of predictions evaluated across the population values of non-focal predictors (sample-based estimate, the colon below indicates that we are comparing two quantities):
 
@@
 
-\BC{To clean%
-\textbf{Marginal effects:} \jd{This isn't really one of our definitions; maybe some of this belongs in the text part.} In simple linear models with no interaction terms, the coefficient estimates are simple and directly interpretable as the expected change in outcome for a unit change in focal input variable.
-This is the \emph{unconditional marginal effect} and it is constant across all the observations and levels of all other input variables.
-Consider models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher}.
-The marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher_no_interaction} is $\frac{\partial y}{\partial x_2} = \beta_2$.
-On the other hand, the marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher} is given by $\frac{\partial y}{\partial x_2} = \beta_2 + \beta_{23}x_3$.
-In other words, if there are no interactions, the marginal effect of $x_2$ on $y$ is constant, while, if there are interactions in the model, the marginal effect of a change in $x_2$ on $y$ depends on the value of the other \emph{conditioning} predictor, $x_3$.
-In essence, computation of marginal effects requires the use of partial derivatives and can be considered as the slope of a multi-dimensional surface with respect to the surface of interest \cite{leeper2017interpreting}.%
-}
-
-The \proglang{R} software package \pkg{margins} \citep{lenth2018package} generates marginal effects.
-To illustrate the distinction between effects and marginal effects, we use the simulated data described in \nameref{S1_Appendix}.
+Lastly, we compare effect plots and the marginal effects \citep{leeper2017interpreting} plots implemented in package \pkg{margins} \citep{lenth2018package}.
+To illustrate this, we use the simulated data described in \nameref{S1_Appendix}.
+The results are shown in \FREF{qoi_age_pred_plot}.
 %

### 2022-07-13 manuscript/body.tex
A walk...

@@
 However, the back-transformation may either result in biased predictions or requires some approximation, for example, second-order Taylor approximation implemented in \pkg{emmeans} \citep{lenth2018package}.
-An alternative to the mean-based reference point is the \emph{whole-sample-based} approach, discussed in detail later, which involves computing the prediction over the population of non-focal predictors and then averaging across the values of the focal predictor \citep{hanmer2013behind}.
+An alternative to the mean-based reference point is the \emph{observed-value-based} approach, discussed in detail later, which involves computing the prediction over the population of non-focal predictors and then averaging across the values of the focal predictor \citep{hanmer2013behind}.
 
@@
 To estimate the central estimate of an outcome plot on the response scale, we need a link function $g$ and appropriate values of the focal predictor $X^{f}$ and reference point for the non-focal predictors $\uset{{\bX}}$.
-As previously mentioned, we can either use mean-based or whole-sample-based reference point.
+As previously mentioned, we can either use mean-based or observed-value-based reference point.
 
@@
 
-On the other hand, using whole-sample-based reference point, \EREF{eta_linear_mean}
+On the other hand, using observed-value-based reference point, \EREF{eta_linear_mean}
 becomes
@@
 \etahfj &= \bbetah \bX^{f}_j \nonumber\\
-\yfi  &= \underset{f}{\textrm{mean}} ~ g^{-1} (\hat{\eta}^\star_j),
+\yfi  &= \underset{j}{\textrm{mean}} ~ g^{-1} (\etahfj),
 \end{align}
@@
 \begin{align}
-\etahfi(\mathrm{Age}_i, \nset{{\over{\mathrm{Wealthindex}}}}) &= \beta_0 + \beta_{\mathrm{A_1}}\mathrm{Age}_i + \beta_{\mathrm{A_2}}\mathrm{Age}^2_i + \beta_{\mathrm{A_3}}\mathrm{Age}^3_i \nonumber\\
-	& + \beta_{\mathrm{W}}\over{\mathrm{Wealthindex}}.
+\etahfi(\mathrm{Age}_i, \nset{{\over{\mathrm{Wealthindex}}}}) &= \betah_0 + \betah_{\mathrm{A_1}}\mathrm{Age}_i + \betah_{\mathrm{A_2}}\mathrm{Age}^2_i + \betah_{\mathrm{A_3}}\mathrm{Age}^3_i \nonumber\\
+	& + \betah_{\mathrm{W}}\over{\mathrm{Wealthindex}}.
 \end{align}
@@
 \begin{align}\label{eq:mpv_second}
-\etahfi(\mathrm{Wealthindex}_i, \nset{{\{\over{\mathrm{Age\mathop{\vphantom{^2}}}}, \over{\mathrm{Age}^2}, \over{\mathrm{Age}^2}\}}}) &= \beta_0 + \beta_{\mathrm{A_1}}\over{\mathrm{Age\mathop{\vphantom{^2}}}} + \beta_{\mathrm{A_2}}\over{\mathrm{Age}^2} + \beta_{\mathrm{A_3}}\over{\mathrm{Age}^3}\nonumber\\
-	& + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i.
+\etahfi(\mathrm{Wealthindex}_i, \nset{{\{\over{\mathrm{Age\mathop{\vphantom{^2}}}}, \over{\mathrm{Age}^2}, \over{\mathrm{Age}^2}\}}}) &= \betah_0 + \betah_{\mathrm{A_1}}\over{\mathrm{Age\mathop{\vphantom{^2}}}} + \betah_{\mathrm{A_2}}\over{\mathrm{Age}^2} + \betah_{\mathrm{A_3}}\over{\mathrm{Age}^3}\nonumber\\
+	& + \betah_{\mathrm{W}}\mathrm{Wealthindex}_i.
 \end{align}
@@
 \begin{align*}
-\etahfi(x_{2i}, x_{3j}, {\uset{\bar{x}}_1}) = \beta_0 + \beta_1 \uset{\bar{x}}_1 + \beta_2x_{2i} + \beta_3x_{3j} + \beta_{23}x_{2i}x_{3j}.
+\etahfi(x_{2i}, x_{3j}, {\uset{\bar{x}}_1}) = \betah_0 + \betah_1 \uset{\bar{x}}_1 + \betah_2x_{2i} + \betah_3x_{3j} + \betah_{23}x_{2i}x_{3j}.
 \end{align*}
@@
 We can directly compare the outcome estimate at the model center for non-focal parameters (mean-based estimate) with the average of predictions evaluated across the population values of non-focal predictors (sample-based estimate, the colon below indicates that we are comparing two quantities):
-
+%
 \begin{align}\label{eq:compare_anchored_pop_based}
@@
 \end{align}
+%
+In an ordinary linear model, the link function $g$ is the identity function, so the two means are the same. For non-trivial link functions, we expect them to be different in general.
 
-In an ordinary linear model, the link function $g$ is the identity function, so the two means are the same.
-For non-trivial link functions, we expect them to be different in general -- \NEW{a consequent of Jensen inequality \citep{duursma_bias_2003, denny_performance_2019, ye_generalized_2021} \jd{Drop this claim; it's obvious we expect them to be different, Jensen is just about the sign}.
-In GL(M)Ms the assumed link function, predetermines the curvature of the response surface and also the sign of the Jensen effect.
-For example, logistic model has a positive curvature if the predicted probabilities are low and negative curvature if the predicted probabilities are high.
-In other words, logistic model may be concave or convex depending on the predicted probabilities, hence the sign of the Jensen effect is not readily available.
-However, in log link functions, the inverse of the link (exponential) function is strictly convex, i.e., an accelerating function.
-For accelerating functions, the Jensen’s inequality describes how the changes in input values elevate the predictions and otherwise describes how these changes depress the predictions.
-Further, our models may include input variables and/or random effects terms other than the focal one.
-In non-trivial link functions, the Jensen effect may depend on the additional input variables \citep{ye_generalized_2021}}.
-\jd{I think I asked you to describe how Jensen is working in these cases.
-In each of the cases it would be both easier and clearer to say whether we expect the left side or the right side to be larger than to spin all of those words about “describes how the changes in input values elevate the predictions and otherwise describes how these changes depress the predictions”.}
+From Jensen's inequality, the exponential link function, for example, is concave up; hence we expect the right-hand side of \EREF{compare_anchored_pop_based} to be greater than the left-hand side, i.e., the average of the predictions is greater than the prediction at the average of the focal and non-focal parameters.
+On the other hand, the logistic function is concave up and concave down at low and high probabilities, respectively.
+We expect a pattern similar to that of exponential function when the logistic function is concave up and the opposite when the logistic function is concave down.
+In other words, the mean-based approach can under-estimate the prediction in exponential and low probability logistic function and over-estimate in high probability logistic function.
 
-\NEW{It is usually important to report the estimates that reflect the expected values of the untransformed response \jd{Hard to understand this s.}.
-This is not a major issue in simple linear models \jd{What do you mean by major issue?
-In what sense is it an issue?
-In what sense does it exist?}.
-However, when dealing with nonlinear link functions,  generated predictions may not reflect the observed response due to the bias in the expected mean induced by the nonlinear transformation of the response variable -- a consequent of Jensen effect mentioned above.
-In such cases, bias correction is needed to back-transform the predictions to the original scales.
+When dealing with nonlinear link functions,  generated estimates may not reflect the observed response due to the bias in the expected mean induced by the nonlinear transformation of the response variable -- as illustrated by the two examples above.
+In such cases, bias correction is needed when back-transforming the estimates to the original scales.
 The common approach for bias-adjustment is second-order Taylor approximation \citep{duursma2003bias, hanmer2013behind}; already implemented in \pkg{emmeans} \citep{lenth2018package}.
-Here, we describe and implement a different approach -- whole-sample-based approach for bias correction.
-Hanmer and Kalkan \citep{hanmer2013behind} introduced mean- and whole-sample- based approaches but limited their discussion to binary response models only.}
-\jd{We mostly talk about binary-response models as well. Also, they did not introduce (there are much older citations to it, and they use a different name, which we should mention (or maybe adopt).}
+Here, we describe and implement a different approach -- observed-value-based approach for bias correction.
+Hanmer and Kalkan \citep{hanmer2013behind} discuss and implement mean- and observed-value- based approaches in binary response models only.
+Our formulation is more general and can easily be extended to other link functions other than logistic.
 
-\subsection{Whole-sample-based approach for bias correction}
+\subsection{Observed-value-based approach for bias correction}
 
 An alternative approach to choosing a reference point is to make predictions over all observations of the non-focal predictors (members of the population) \citep{hanmer2013behind}.
-The nonlinear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population to a one-dimensional distribution of $\eta^\star(x_f, \nset{x})$, which is a function of the chosen values of the focal predictor and the whole sample of non-focal predictors, as opposed to the definition in \EREF{eta_mean}.
+The nonlinear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population to a one-dimensional distribution of $\etahfi(x_f, \nset{x})$, which is a function of the chosen values of the focal predictor and the whole sample of non-focal predictors, as opposed to the definition in \EREF{eta_mean}.
 More specifically:
 \begin{itemize}
-\item compute linear predictor associated with whole sample of the non-focal predictors, $\nset{\eta} = \sum \nset{\beta} \nset{x}$
-\item compute linear predictor associated with the chosen values focal predictor, $\eta_{jf} = \sum{\beta_f x_{jf}}$
-\item for every value of the focal linear predictor, $\eta_{jf}$, compute
+\item compute linear predictor associated with whole sample of the non-focal predictors, $\etanfj = \sum \uset{\hat{\beta}} \uset{x}$
+\item compute linear predictor associated with the chosen values focal predictor, $\etahfi = \hat{\beta}^f x^f$
+\item for every value of the focal linear predictor, $\etahfi$, compute
 %
 \begin{align}\label{eq:pop_eta} 
-\eta_j^\star(\eta_{jf}, \nset{\eta})  &= \eta_{jf} + \nset{\eta} \nonumber \\
-&= \eta_j^\star(x_f, \nset{x}).
+\etahfj(\etahfi, \etanfj)  &= \etahfi + \etanfj \nonumber \\
+&= \etahfj(x^f, \uset{x}).
 \end{align}
@@
 %
+HERE
 
@@
 \begin{align}\label{eq:pop_response} 
-\hat{y}_f  &= \textrm{mean} ~ g^{-1} \left(\eta_j^\star(x_f, \nset{x})\right).
+\yfi  &= \underset{j}{\textrm{mean}} ~ g^{-1} \left(\etahfj(x^f, \uset{x})\right).
 \end{align}
@@
 \begin{align}
-\sigma_{jf}^2 = \textrm{Diag}(\bX^\star_{jc} \boldsymbol{\Sigma} \bX^{\star\top}_{jc})
+\sigma_j^2 = \textrm{Diag}(\bX^{f}_j \boldsymbol{\Sigma} {\bX^{f}_j}^\top)
 \end{align}
-where $\bX^{\star}_{jc} = \{\bX_{jf}^\star, \nset{{\bX}^\star} - \nset{{\bar{\bX}}^\star}\}$ and 
+%
+and
 %
 \begin{align}
-\mathrm{CI}_f = \mathrm{mean} ~ g^{-1} \left(\eta_j^\star(x_f, \nset{x}) \pm q\sigma_{jf}\right).
+\mathrm{CI}_i = \underset{j}{\textrm{mean}} ~ g^{-1} \left(\etahfj(x^f, \uset{x}) \pm q\sigma_{j}\right).
 \end{align}
@@
 For models with random effects components, we make further adjustment to correct for bias induced by the random effects terms.
-In whole-sample approach, we treat the random effects terms as additional non-focal predictors and simply make adjustment to \EREF{pop_eta}.
+In observed-value approach, we treat the random effects terms as additional non-focal predictors and simply make adjustment to \EREF{pop_eta}.
 In particular
@@
 \begin{align}\label{eq:pop_eta_re} 
-\tau &= \bZ b \nonumber \\
-\eta_j^\star(x_f, \nset{x}, \tau)  &= \eta_j^\star(x_f, \nset{x}) + \tau
+\uset{\tau} &= \bZ b \nonumber \\
+\etahfj(x^f, \uset{x}, \uset{\tau})  &= \etahfj(x^f, \uset{x}) + \uset{\tau}
 \end{align}
@@
 
-\section{Mean-based vs. whole-sample-based}
+\section{Mean-based vs. observed-value-based}
 
-In the whole-sample-based approach, the ensemble of predictions and CIs are back-transformed before averaging, see \EREF{pop_response}, so we do not need to worry about the nonlinear averaging.
+In the observed-value-based approach, the ensemble of predictions and CIs are back-transformed before averaging, see \EREF{pop_response}, so we do not need to worry about the nonlinear averaging.
 In other words, the averaging is no longer on the link scale and is likely to be bounded by the original data scale.
@@
 This may limit the inferences we can make about the entire population.
-In general, the mean-based approach provides the predictions of an average case, whereas the whole-sample-based approach summarizes the predictions over the entire population.
+In general, the mean-based approach provides the predictions of an average case, whereas the observed-value-based approach summarizes the predictions over the entire population.
 In some applications, the effect of an average case might not be generalizable to the entire population, especially, if the average does not represent the population.
-This might not be a problem in the whole-sample-based approach since it focuses on specific observations -- the prediction is first obtained for each observation and then averaged across the levels of the focal predictor.
+This might not be a problem in the observed-value-based approach since it focuses on specific observations -- the prediction is first obtained for each observation and then averaged across the levels of the focal predictor.
 
@@
 
-The whole-sample-based approach is not entirely foolproof.
+The observed-value-based approach is not entirely foolproof.
 For instance, similar to the mean-based approach, in the case of continuous focal predictors, choosing the representative values of the focal predictors can be very challenging, especially if the cases are not evenly distributed around the minimum and the maximum values or within some subgroups defined in the population.
-In addition, the whole-sample-based approach can be computationally intensive for large datasets.
+In addition, the observed-value-based approach can be computationally intensive for large datasets.
 
@@
 
-We start by comparing our proposed approach with the existing implementations and then illustrate the construction of outcome plots and also demonstrate that the mean-based and whole-sample-based approaches produce different results in models with nonlinear link functions and/or additional source(s) of potential bias.
-In addition, we demonstrate that whole-sample-based approach can be used for bias correction.
+We start by comparing our proposed approach with the existing implementations and then illustrate the construction of outcome plots and also demonstrate that the mean-based and observed-value-based approaches produce different results in models with nonlinear link functions and/or additional source(s) of potential bias.
+In addition, we demonstrate that observed-value-based approach can be used for bias correction.
 
@@
 
-Mean-based and whole-sample-based approaches can produce very different results in models with nonlinear link functions with additional sources of potential bias such as additional non-focal input variable(s), random effects or complex interactions.
+Mean-based and observed-value-based approaches can produce very different results in models with nonlinear link functions with additional sources of potential bias such as additional non-focal input variable(s), random effects or complex interactions.
 To demonstrate this, we considered a two predictor binary outcome simulation described in \nameref{S2_Appendix}, such that age has a very small effect size in comparison to wealth index, and compared their effects on the estimated probability of improved water quality as shown in \FREF{pred_bin_plots}.
@@
 \end{center}
-\caption{{\bf Mean-based and whole-sample-based central estimates.} The \emph{truth} is based on the true simulation parameters.
-If the focal predictor has a small effect size, mean-based and whole-sample-based approaches produce different estimates -- A.
+\caption{{\bf Mean-based and observed-value-based central estimates.} The \emph{truth} is based on the true simulation parameters.
+If the focal predictor has a small effect size, mean-based and observed-value-based approaches produce different estimates -- A.
 In the case of strong effect size, however, the two approaches produce very close estimates -- B.
-In both cases, the whole-sample-based estimates are closer to the truth.
-The difference in mean-based and whole-sample-based is due to the bias induced by the non-focal predictor and the nonlinear link function in the logistic model.
+In both cases, the observed-value-based estimates are closer to the truth.
+The difference in mean-based and observed-value-based is due to the bias induced by the non-focal predictor and the nonlinear link function in the logistic model.
 Mean-based approach suffers from nonlinear averaging; and if focal predictor has small effect size, the bias is even pronounced since the effect of nonlinear averaging is driven by the non-focal predictor(s) with the strong effect.
 However, this is not the case if the non-focal predictor(s) have small effect since the bias they induce is equally likely to be small.
-Since whole-sample-based approach averages over the whole population of the non-focal predictor variables, it is less likely to be affected by the nonlinear averaging for the reasons already mentioned.
+Since observed-value-based approach averages over the whole population of the non-focal predictor variables, it is less likely to be affected by the nonlinear averaging for the reasons already mentioned.
 The horizontal dashed lines correspond to the respective averages.
@@
 We simulated the data as shown in \nameref{S3_Appendix} and fitted the two models -- non-mediated and mediated.
-In both cases, we compared mean-based and the whole-sample-based predictions, as shown in \FREF{pred_mediated_plots}.
+In both cases, we compared mean-based and the observed-value-based predictions, as shown in \FREF{pred_mediated_plots}.
 
@@
 \end{center}
-\caption{{\bf Mean-based and whole-sample-based estimates for mediated (multivariate) and non-mediated (univariate) models.} A: In the absence of mediator variable, both mean-based and whole-sample-based approaches predictions are identical since there is no any other additional (non-focal predictor) sources of bias.
+\caption{{\bf Mean-based and observed-value-based estimates for mediated (multivariate) and non-mediated (univariate) models.} A: In the absence of mediator variable, both mean-based and observed-value-based approaches predictions are identical since there is no any other additional (non-focal predictor) sources of bias.
 Consequently, in both approaches, the average estimated probability is very close to the observed proportion.
 B: When the mediator variable is included,  $x$ has no direct effect on $z$ and as a result the predictions do not align with the observations.
-However, the whole-sample-based approach still closely approximates the average proportion of the simulated data.
+However, the observed-value-based approach still closely approximates the average proportion of the simulated data.
 The horizontal dashed lines are the respective average estimates and observed proportions.
@@
 
-Although commonly used \proglang{R} software packages, by default, implements mean-based approach, our simulation results demonstrated that the whole-sample-based approach has a potential of yielding results which are more consistent with the observed data.
+Although commonly used \proglang{R} software packages, by default, implements mean-based approach, our simulation results demonstrated that the observed-value-based approach has a potential of yielding results which are more consistent with the observed data.
 We would, therefore, argue that the use of mean-based approach should have some theoretical justification, especially in complex models.
@@
 
-From our theoretical, methodological and simulation results, researchers using these kind of models should, in the absence of theoretical justification, report predictions based on whole-sample-based approach or at least attempt to do a comparison of the two approaches before settling on the most appropriate in answering their research question.
+From our theoretical, methodological and simulation results, researchers using these kind of models should, in the absence of theoretical justification, report predictions based on observed-value-based approach or at least attempt to do a comparison of the two approaches before settling on the most appropriate in answering their research question.
 Moreover, we provide \proglang{R} package, \pkg{vareffects}, which implements these methods and is available on github (\href{https://github.com/mac-theobio/effects}{https://github.com/mac-theobio/effects}).

### 2022-07-13 manuscript/body.tex
Sending back to JD.

@@
 
-To make an outcome plot, we use a \emph{focal} predictor on the x-axis and central estimate (predicted values) on the y-axis.
+To make an outcome plot, we use a \emph{focal} predictor on the x-axis and \emph{central estimate} (predicted values) on the y-axis.
 The resulting plot will depend on choices we make about other (non-focal) predictor(s).
 In ordinary (Gaussian) linear regression, the non-focal choices may have a simple additive effect on the central estimate.
-However, when the focal predictor has interactions or a non-linear link function, non-focal choices can also affect the slope of the estimate.
+However, when the focal predictor has interactions or non-linear response functions (e.g., spline or polynomial), non-focal choices can also affect the slope of the estimate.
 Additional challenges arise when dealing with “mixed” models, which incorporate random effects.
@@
 \item biases induced by non-linear transformations of the response variable in generalized models (especially generalized mixed models).
-\item choice of representative values of the focal predictor \jd{I still don't see this as much of a challenge.} \dya{We needed to mention somewhere that we also have to choose focal values.
-I think \P~ below is suffecient.}
 \end{enumerate}
@@
 
-In models with non-focal predictors such as multivariate models, reference point can be chosen as the average of the non-focal \emph{model variables} -- we call this approach \emph{mean-based} reference point and is currently not implemented in commonly used \proglang{R} software packages.
+In a model with non-focal predictors such as multivariate models, reference points can be chosen as the average of the non-focal \emph{linear predictor variables} -- we call this approach \emph{mean-based} reference point and is currently not implemented in commonly used \proglang{R} software packages.
 The mean of the central estimates is the prediction at the \emph{model center}.
@@
 
-For linear models, the averaging is done on the linear scale, i.e., linear averaging.
+For a linear model, the averaging is done on the linear scale, i.e., linear averaging.
 As a result, the \emph{bias} in model center estimates can easily be corrected using mean-based approach and the estimates are usually consistent with the observed values.
-However, in models with non-linear link functions, this is not always the true.
-In particular, when dealing with generalized models with non-linear link functions, the averaging is done on the non-linear scale, i.e., non-linear averaging, making generating correct predictions much harder.
+However, in a model with non-linear link function, this is not usually true.
+In particular, when dealing with a generalized model with non-linear link function, the averaging is done on the non-linear scale, i.e., non-linear averaging, making generating correct predictions much harder.
 This is the bias in the expected mean prediction induced by the non-linear transformation of the response variable.
-One way to address this is to make predictions on the linear predictor scale and back-transform to the original scale.
-However, the back-transformation may either result in biased predictions or requires some approximation, for example, second-order Taylor approximation implemented in \pkg{emmeans} \citep{lenth2018package}.
 An alternative to the mean-based reference point is the \emph{observed-value-based} approach, discussed in detail later, which involves computing the prediction over the population of non-focal predictors and then averaging across the values of the focal predictor \citep{hanmer2013behind}.
@@
 \item \textbf{Linear predictor variables}: Refer to the variables which are combined to make the linear predictor (corresponding to the columns in the model matrix).
-Each input variable may correspond to one or more model variable.
+Each input variable may correspond to one or more linear predictor variables.
 Input variables with more than two categories (for instance religion in our previous example), or input variables with non-linear response functions (e.g., spline or polynomial) will correspond to more than one linear predictor variable -- we call such variables “multi-parameter variables”.
 
-\item \textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more model variables).
+\item \textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more linear predictor variables).
 Also referred to as center point.
-The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable -- the case in \FREF{justify_plots}B.
-In simple linear models with no interactions, the mean of the input variable can be used is a center point.
-If there are more than one model variables associated with the focal input variable, however, the model center does not correspond to a single value of the focal predictor, see \FREF{pred_cubic_plots}.
+The center point for a set of linear predictor variables corresponding to an input variable may not represent a possible value of the input variable -- the case in \FREF{justify_plots}B.
+In a simple linear model with no interactions, the mean of the input variable can be used as a center point.
+If there are more than one linear predictor variables associated with the focal input variable, however, the model center does not correspond to a single value of the focal predictor, see \FREF{pred_cubic_plots}.
 
@@
 
-\jd{This should be more general.
-The natural way to do the comparison is to first write out the two ways of averaging, and then point out that they're the same when $g$ is the identity.
-In other words, do the general derivation before making the assumption.
-It doesn't ever seem good to write $g$ and then \emph{define} it as the identity.}
-
-\BC{Trying to adopt new notation style, is this better?}
-
-To estimate the central estimate of an outcome plot on the response scale, we need a link function $g$ and appropriate values of the focal predictor $X^{f}$ and reference point for the non-focal predictors $\uset{{\bX}}$.
+To estimate the central estimate of an outcome plot on the response scale, we need a link function $g$ and appropriate values of the focal predictor $\bX^{f}$ and reference point for the non-focal predictors $\uset{{\bX}}$.
 As previously mentioned, we can either use mean-based or observed-value-based reference point.
@@
 %
-where $\bX^{f}_c = \{X^{f}, \uset{{\bX}}_c\}$ is a centered model matrix with appropriately chosen values of the focal predictor $X^{f}$ and centered non-focal linear variables $\uset{{\bX}}_c$ constructed by replacing the entries in the corresponding columns with their averages.
+where $\bX^{f}_c = \{\bX^{f}, \uset{{\bX}}_c\}$ is a centered model matrix with appropriately chosen values of the focal predictor $\bX^{f}$ and centered non-focal linear predictor variables $\uset{{\bX}}_c$ constructed by replacing the corresponding values in the model matrix with their averages.
 
@@
 observation and $\uset{{\bX}}$ is the entire population of the non-focal linear
-predictors.
+predictor variables.
 Here, we generate a vector of size $J \times N$ and then average over the values of the focal predictor; $J$ and $N$ represents the number of focal values and population of non-focal predictors, respectively.
@@
 In the subsequent sections, we will first discuss general formulation of mean-based approach and how to generate associated confidence intervals.
-Thereafter, we will discuss the second approach, who-sample-based, and its application to bias correction.
+Thereafter, we will discuss the second approach, observed-value-based, and its application to bias correction.
 
@@
 
-Interactions can be between non-focal predictor or between focal and non-focal
+Interactions can be between non-focal predictors or between focal and non-focal
 predictors. Handling former case is similar to that of the second case
@@
 The interaction is between the focal, $x_2$, and non-focal, $x_3$, predictors.
-In this case, we chose the values of the focal predictors as previously described and the reference point for the interacting non-focal predictor is predetermined or appropriately chosen set of values.
+In this case, the values of the focal predictors are chosen as previously described and the reference point for the interacting non-focal predictor is predetermined or appropriately chosen set of values.
 In our example, suppose we pick $i$ and $j$ unique values of the focal predictor, $x_2$, and interacting non-focal predictor, $x_3$, respectively.
-The central estimate on linear predictor scale is given by
+The central estimate on linear predictor becomes
 %
@@
 
-Let $\bxo$ be a centered model matrix previously defined, and let $\ba$ be an anchor matrix, with the same dimensions and entries in all non-focal predictors as $\bxo$.
-Let $\baf$ be the column of $\ba$ corresponding to focal predictor(s) defined in $\bxo$.
-Any appropriate values can be chosen for $\baf$ but for model center (center-anchored), we use $\bafc$ which is the mean of the focal predictor(s).
+Let $\bxo$ be a centered model matrix previously defined, and let $\ba$ be an anchor matrix, with the same dimensions and entries of all non-focal linear predictor variables as $\bxo$.
+Let $\baf$ be the column of $\ba$ corresponding to focal linear predictor variable defined in $\bxo$.
+Any appropriate values can be chosen for $\baf$ but for model center (center-anchored), we use $\bafc$ which is the mean of the focal predictor.
 Thus
@@
 \begin{align}\label{eq:centered_variance}
-\boldsymbol\sigma^2_i = \textrm{Diag}((\bxo - \bafc) \boldsymbol{\Sigma} (\bxo - \bafc)^\top).
+\boldsymbol\sigma^2_i = \textrm{Diag}((\bxo - \bafc) \boldsymbol{\Sigma} {(\bxo - \bafc)}^\top).
 \end{align}
@@
 
-We can directly compare the outcome estimate at the model center for non-focal parameters (mean-based estimate) with the average of predictions evaluated across the population values of non-focal predictors (sample-based estimate, the colon below indicates that we are comparing two quantities):
+We can directly compare the outcome estimate at the model center for non-focal parameters (mean-based estimate) with the average of predictions evaluated across the population values of non-focal predictors (observed-value-based estimate, the colon below indicates that we are comparing two quantities):
 %
@@
 We expect a pattern similar to that of exponential function when the logistic function is concave up and the opposite when the logistic function is concave down.
-In other words, the mean-based approach can under-estimate the prediction in exponential and low probability logistic function and over-estimate in high probability logistic function.
+In other words, the mean-based approach can under-estimate the prediction in exponential and low probability logistic functions and over-estimate in high probability logistic function.
+
+\fix{Thinking about RE}
 
@@
 Here, we describe and implement a different approach -- observed-value-based approach for bias correction.
-Hanmer and Kalkan \citep{hanmer2013behind} discuss and implement mean- and observed-value- based approaches in binary response models only.
+Hanmer and Kalkan \citep{hanmer2013behind} discussed and implemented mean- and observed-value- based approaches in binary response models only.
 Our formulation is more general and can easily be extended to other link functions other than logistic.
@@
 %
-HERE
 
@@
 \begin{align}
-\mathrm{CI}_i = \underset{j}{\textrm{mean}} ~ g^{-1} \left(\etahfj(x^f, \uset{x}) \pm q\sigma_{j}\right).
+\mathrm{CI}_i = \underset{j}{\textrm{mean}} ~ g^{-1} \left(\etahfj(x^f, \uset{x}) \pm q\sigma^f_{j}\right).
 \end{align}

### 2022-07-16 manuscript/abstract.tex
Polishing

@@
 \\
-Steve Cygu\textsuperscript{1},
+Steve Cygu\textsuperscript{1, *},
 Benjamin M. Bolker\textsuperscript{1,2},
@@
 \bigskip
-\textbf{1} School of Computational Science and Engineering, McMaster University, Hamilton, Ontario, Canada
+\textsuperscript{1}School of Computational Science and Engineering, McMaster University, Hamilton, Ontario, Canada
 \\
-\textbf{2} Department of Biology, McMaster University, Hamilton, Ontario, Canada
+\textsuperscript{2}Department of Biology, McMaster University, Hamilton, Ontario, Canada
 \\

### 2022-07-17 manuscript/abstract.tex
Fixing abstract

@@
 
-\jd{This talks about a lot of terms that we don't define. Remember to rewrite!}
-\fix{BC to rewrite}
-
-In applications that use generalized linear (mixed) models, outcome predictions are often of interest.
-For models that involve complex multiplicative interactions, additional non-focal input variables or nonlinear link functions, the estimated coefficients are not readily interpretable.
-A general way to summarize these kind of models is through predictions, that depend not only on choice of representative values of focal input variables but are also sensitive to which values of non-focal input variables are chosen.
-The most common approach is generating predictions at a reference point, usually the mean of non-focal predictor.
+In generalized (mixed) models that involve complex multiplicative interactions, multi-parameter variables, additional non-focal predictors or a nonlinear link function, outcome plots (prediction and effect plots) can aid in understanding difficult-to-interpret coefficient estimates.
+Outcome plots depend on not only values of the focal predictors but are also on the choices we make about the non-focal predictors.
+The most common approach is to generate estimates (central estimates, predictions and effects) at a reference point, usually the mean of non-focal predictor.
 We call this mean-based approach and it estimates effect of an average case in the population.
-In the presence of additional non-focal input variables, nonlinear link functions, random effect terms, etc., mean-based approach generates predictions that are biased and may not be consistent with the observed quantities.
-An alternative is the whole-sample-based approach which estimates the average effect in the population.
-Moreover, effects confidence intervals provide an alternative and a more clear way to describe uncertainty associated with the focal input variable.
-In addition to theoretical and methodical comparison, using simulation, we illustrate the two approaches and show that they can produce substantially different results and that whole-sample-based approach can not only produce estimates consistent with the observed values, but also appropriate for bias correction.
-We also present an alternative way, effects confidence intervals, to describe uncertainties associated with these predictions.
+In the presence of additional non-focal input variables, nonlinear link functions, random effect terms, etc., mean-based approach generates estimates that are biased and may not be consistent with the observed quantities.
+An alternative is the observed-value-based approach which estimates the average effect in the population.
+Moreover, effect-styled confidence intervals provides an alternative and a more clear way to describe uncertainty associated with the focal input variable.
+In addition to theoretical and methodical comparison, using simulation, we illustrate the two approaches and show that they can produce substantially different results and that observed-value-based approach can not only produce estimates consistent with the observed values, but also appropriate for bias correction.
+We also present an alternative way, effect-styled confidence intervals, to describe uncertainties associated with these predictions.
 

### 2022-07-18 manuscript/abstract.tex
Upto simulations

@@
 
-In generalized (mixed) models that involve complex multiplicative interactions, multi-parameter variables, additional non-focal predictors or a nonlinear link function, outcome plots (prediction and effect plots) can aid in understanding difficult-to-interpret coefficient estimates.
-Outcome plots depend on not only values of the focal predictors but are also on the choices we make about the non-focal predictors.
+In generalized linear (mixed) models that involve complex multiplicative interactions, multi-parameter variables, additional non-focal predictors or a nonlinear link function, outcome plots (prediction and effect plots) can aid in understanding difficult-to-interpret coefficient estimates.
+Outcome plots depend on the choices we make about the non-focal predictors.
 The most common approach is to generate estimates (central estimates, predictions and effects) at a reference point, usually the mean of non-focal predictor.
 We call this mean-based approach and it estimates effect of an average case in the population.
-In the presence of additional non-focal input variables, nonlinear link functions, random effect terms, etc., mean-based approach generates estimates that are biased and may not be consistent with the observed quantities.
+In the presence of additional non-focal predictors, nonlinear link functions, random effect terms, etc., mean-based approach generates estimates that are biased and may not be consistent with the observed quantities.
 An alternative is the observed-value-based approach which estimates the average effect in the population.
-Moreover, effect-styled confidence intervals provides an alternative and a more clear way to describe uncertainty associated with the focal input variable.
+Moreover, effect-styled confidence intervals provides an alternative and a more clear way to describe uncertainty associated with the focal predictor.
 In addition to theoretical and methodical comparison, using simulation, we illustrate the two approaches and show that they can produce substantially different results and that observed-value-based approach can not only produce estimates consistent with the observed values, but also appropriate for bias correction.
-We also present an alternative way, effect-styled confidence intervals, to describe uncertainties associated with these predictions.
+We also present an alternative way, effect-styled confidence intervals, to describe uncertainties associated with the central estimates.
 

### 2022-07-18 manuscript/body.tex
Upto simulations

@@
 Another challenge or a continuation  of the above?}
-\item biases induced by non-linear transformations of the response variable in generalized models (especially generalized mixed models).
+\item biases induced by non-linear transformations of the response variable in generalized linear models (especially generalized mixed models).
 \end{enumerate}
@@
 In this article, we refer to these quantities as the \emph{central estimates} of the outcome.
+The mean of the central estimates is the prediction at the \emph{model center}.
 
 In a model with non-focal predictors such as multivariate models, reference points can be chosen as the average of the non-focal \emph{linear predictor variables} -- we call this approach \emph{mean-based} reference point and is currently not implemented in commonly used \proglang{R} software packages.
-The mean of the central estimates is the prediction at the \emph{model center}.
-We introduce an alternative choice for the reference point below.
+We introduce an alternative choice for the reference point.
 
@@
 
-When dealing with nonlinear link functions,  generated estimates may not reflect the observed response due to the bias in the expected mean induced by the nonlinear transformation of the response variable -- as illustrated by the two examples above.
+When dealing with nonlinear link functions and additional non-focal predictors,  generated estimates may not reflect the observed response due to the bias in the expected mean induced by the nonlinear transformation of the response variable -- as illustrated by the two examples above.
 In such cases, bias correction is needed when back-transforming the estimates to the original scales.
@@
 Here, we describe and implement a different approach -- observed-value-based approach for bias correction.
-Hanmer and Kalkan \citep{hanmer2013behind} discussed and implemented mean- and observed-value- based approaches in binary response models only.
+Hanmer and Kalkan \citep{hanmer2013behind} discussed and implemented observed-value-based and input-variable-based mean-based approaches in binary response models only.
 Our formulation is more general and can easily be extended to other link functions other than logistic.
@@
 However, for complex models, we demonstrate that the two approaches can produce substantially different results and show that averaging the interactions closely matches the observed values.
-To illustrate this, consider models~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher} below, with $x_1$ as the focal predictor:
+To illustrate this, consider models description by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher} below, with $x_1$ as the focal predictor:
 %

### 2022-07-18 manuscript/body.tex
Upto MPVs

@@
 \end{center}
-\caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} central estimates for regression of $x_1$ on $y$ (outcome plots).} The horizontal dashed lines are the mean of the central estimates and simulated $y$, $\bar{\hat{y}}$ and $\bar{y}$, respectively.
+\caption{{\bf A comparison of central estimates for \pkg{emmeans}, \pkg{effects} and \pkg{varpred}.} The horizontal dashed lines are the mean of the central estimates and simulated $y$, $\bar{\hat{y}}$ and $\bar{y}$, respectively.
 The grey points are the binned simulated $y$.
-The trend lines represent the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other input variables at their mean and the true predictions based on the simulation parameters -- central estimate.
+The trend lines represent the $\hat{y}$ -- central estimate.
 A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the simulated one in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$.
 Similarly, the central estimate based on the three approaches match the truth.
-B: Even with the simple interaction between non-focal predictors, we see differences between predicted mean, $\bar{\hat{y}}$ and simulated mean , $\bar{y}$.
+B: Even with the simple interaction between non-focal predictors, we see differences between predicted mean, $\bar{\hat{y}}$ and simulated mean, $\bar{y}$.
 Similarly, there are differences between central estimates and the truth, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not the proposed \pkg{varpred}.}
@@
 %
-In the absence of interactions (\EREF{simple_inter_higher_no_interaction}), the three approaches produce the same estimates, which match the simulated values, \FREF{justify_plots}A.
+In the absence of interactions (\EREF{simple_inter_higher_no_interaction}), the three approaches produce identical estimates, which match the simulated values, \FREF{justify_plots}A.
 However, in the presence of interactions, even as simple as the one in \EREF{simple_inter_higher}, the estimates start to differ.
@@
 \caption{{\bf Prediction and effect plots.} The description of horizontal, vertical and trend lines remain the same as above.
-A: The wider dashed curves corresponds to the conventional prediction curves from \pkg{emmeans} and \pkg{effects} packages, while the narrower curves crossing at the mean of the focal predictor, also referred to as the center point, represent the effect from \pkg{varpred}.
+A: The wider dashed curves corresponds to the conventional prediction curves from \pkg{emmeans} and \pkg{effects} packages, while the narrower curves crossing at the center point represent the effect from \pkg{varpred}.
 For simple OLS models, the effect-styled curves cross at the center point (which is the default anchor and also referred to as center-anchored).
@@
 B: center- and zero- anchored effects.
-The center-anchored effects (as described above) while the zero-anchored means that the anchor is at zero value of the focal predictor.
+The zero-anchored effects means that the anchor is at zero value of the focal predictor.
 The choice of the anchor do not affect the central estimates, hence they are identical in both choices of the anchor.}
@@
 B: The marginal effect (ME) at representative values of the focal predictor \code{age} together with the $95\%$ confidence intervals.
-For simple linear models with no interactions, the ME should be constant across all the values of the focal predictor; and it would correspond to the slope of the central estimate of central estimate.
+For simple linear models with no interactions, the ME should be constant across all the values of the focal predictor; and it would correspond to the slope of the central estimate.
 However, in this example, the focal predictor is a cubic polynomial, and so, the effect is not constant but a function of the partial derivatives of the corresponding linear predictor variables.}}

### 2022-07-19 manuscript/body.tex
JD can now have a look

@@
 In a simple linear model with no interactions, the mean of the input variable can be used as a center point.
-If there are more than one linear predictor variables associated with the focal input variable, however, the model center does not correspond to a single value of the focal predictor, see \FREF{pred_cubic_plots}.
+If there are more than one linear predictor variables associated with the focal predictor, however, the model center does not correspond to a single value of the focal predictor, see \FREF{pred_cubic_plots}.
 
@@
 We also used the same model to illustrate how predictions and effects differ if MPVs are in the focal or non-focal predictor.
-In particular, we generated two sets of mean-based predictions and effects: 1) one with \code{age} as the focal predictor (a cubic polynomial focal input variable); and 2) one with \code{Wealthindex} as the focal predictor (linear focal input variable).
+In particular, we generated two sets of mean-based predictions and effects: 1) one with \code{age} as the focal predictor (a cubic polynomial - MPV); and 2) one with \code{Wealthindex} as the focal predictor (linear).
 In both cases the, we used center point as the anchor for the effects, as shown in \FREF{pred_cubic_plots}.
@@
 \caption{{\bf Prediction and effect plots for cubic polynomial and simple focal predictors.} A: The focal predictor is a MPV cubic polynomial.
-B: The focal input variable is linear, with MPV non-focal predictor.
-The central estimates (dashed central curves or lines) are the same for predictions and effects, in each case (A and B).
+B: The focal predictor is linear, with MPV non-focal predictor.
+The central estimates (dashed central curves or lines) are the same for predictions and effects, in respective case (A and B).
 However, the effects are narrower than the predictions.
 The default anchor for generating effects is the center point.
-For linear (or simple) focal input variables, the center point is a point within the range of focal values; and therefore the effect curves intersect at that that point -- B.
+For linear (or simple) focal predictor, the center point is a point within the range of focal values; and therefore the effect curves intersect at that that point -- B.
 In the case of MPVs such as cubic polynomials, the center point is not a point but a combination of all the linear predictor variables associated with the cubic polynomial, consequently, curves will not necessarily cross at a point -- A.
-The horizontal black and red dashed lines are the observed and predicted average household size, i.e., $\over{hh~size}$ and $\over{\widehat{hh~size}}$, respectively.}
+The horizontal black and yellow dashed lines are the observed and predicted average household size, i.e., $\over{hh~size}$ and $\over{\widehat{hh~size}}$, respectively.}
 \label{fig:pred_cubic_plots}
@@
 
-In the absence of, complex interactions, MPVs, or transformations on the focal input variable, in simple OLS models we would expect the observed average, $\over{hh~size}$, and the average of the predictions, $\over{\widehat{hh~size}}$, to be identical and intersect with the trend lines at the center point, for the effect plot.
-On the other hand, if the focal input variable is characterized by complex interactions, MPVs, or any other form of transformations, the center point is not necessarily a point in the range of focal input values.
+In the absence of, complex interactions, MPVs, or transformations on the focal predictor, in simple OLS models we would expect the observed average, $\over{hh~size}$, and the average of the predictions, $\over{\widehat{hh~size}}$, to be identical and intersect with the trend lines at the center point, for the effect plot.
+On the other hand, if the focal predictor is a MPVs or has any other form of transformation, the center point is not necessarily a point in the range of focal predictor.
 In this case, the effects will be narrower than the predictions plots but will not necessarily intersect.
@@
 
-Mean-based and observed-value-based approaches can produce very different results in models with nonlinear link functions with additional sources of potential bias such as additional non-focal input variable(s), random effects or complex interactions.
-To demonstrate this, we considered a two predictor binary outcome simulation described in \nameref{S2_Appendix}, such that age has a very small effect size in comparison to wealth index, and compared their effects on the estimated probability of improved water quality as shown in \FREF{pred_bin_plots}.
+In a generalized linear model with at least one non-focal predictor, mean-based and observed-value-based approaches can produce very different.
+To demonstrate this, we considered a two input variables binary outcome simulation described in \nameref{S2_Appendix}, such that \code{Age} has a very small effect size in comparison to \code{Wealthindex}, and compared their effects on the estimated probability of improved water quality as shown in \FREF{pred_bin_plots}.
 
@@
 
-In \FREF{pred_bin_prediction_effects_plots} below, we use \FREF{pred_bin_plots}A and compare the prediction and effect plots associated with the central estimates.
+In \FREF{pred_bin_prediction_effects_plots}, we use the results in \FREF{pred_bin_plots}A and compare the prediction and effect plots associated with the central estimates.
 
@@
 We simulated the data as shown in \nameref{S3_Appendix} and fitted the two models -- non-mediated and mediated.
-In both cases, we compared mean-based and the observed-value-based predictions, as shown in \FREF{pred_mediated_plots}.
+In both cases, we compared mean-based and the observed-value-based central estimates, as shown in \FREF{pred_mediated_plots}.
 
@@
 \end{center}
-\caption{{\bf Mean-based and observed-value-based estimates for mediated (multivariate) and non-mediated (univariate) models.} A: In the absence of mediator variable, both mean-based and observed-value-based approaches predictions are identical since there is no any other additional (non-focal predictor) sources of bias.
+\caption{{\bf Mean-based and observed-value-based estimates for mediated (multivariate) and non-mediated (univariate) models.} A: In the absence of mediator variable, both mean-based and observed-value-based approaches central estimates are identical since there is no any other additional (non-focal predictor) sources of bias.
 Consequently, in both approaches, the average estimated probability is very close to the observed proportion.
-B: When the mediator variable is included,  $x$ has no direct effect on $z$ and as a result the predictions do not align with the observations.
+B: When the mediator variable is included,  $x$ has no direct effect on $z$ and as a result the central estimates do not align with the observations.
 However, the observed-value-based approach still closely approximates the average proportion of the simulated data.

### 2022-07-19 manuscript/glm_two_predictor_preds_effects.R

@@ -54,7 +54,7 @@ pred_age <- comparevarpred(vlist=vlist
 pred_age_df <- pred_age$preds
 pred_age_df$.method <- "A) Mean-based" 
 
-## Corrected (Whole-sample)
+## Corrected (Observed-value)
 ### Predictions
 pred_age_prediction_ws <- varpred(glm_mod
 	, "age"
@@ -80,9 +80,9 @@ pred_age_ws <- comparevarpred(vlist=vlist
 	, ci=TRUE
 )
 pred_age_ws_df <- pred_age_ws$preds
-pred_age_ws_df$.method <- "B) Whole-sample-based"
+pred_age_ws_df$.method <- "B) Observed-value-based"
 pred_age_pop_ws_true_df <- pred_age_pop_true_df
-pred_age_pop_ws_true_df$.method <- "B) Whole-sample-based"
+pred_age_pop_ws_true_df$.method <- "B) Observed-value-based"
 
 ## Compare the predictions
 pred_df <- do.call("rbind"

### 2022-07-20 manuscript/body.tex
comments on the beginning (sorry!)

@@
 The other distinction relates not to the method of calculating CIs, but to the model chosen for the plot. 
-In general, if we want to predict based on a focal parameter, we are likely to want to fit a univariate model, that only contains terms related to that predictor.
-If we want to know the effects of a predictor, we may want to control for covariates with a multivariate model, in order to estimate “direct” effects, or leave covariates out, in order to estimate “total” effects (direct plus indirect), or take an intermediate strategy.
+In general, if we want to predict based on a focal parameter, we are likely to want to fit a univariate model that only contains terms related to that predictor.
+If we want to know the effects of a predictor, we may want to control for covariates with a multivariate model, in order to estimate “direct” effects; leave covariates out, in order to estimate “total” effects (direct plus indirect); or take an intermediate strategy.
 Shi et al.
-\citep{shi_evidence_2017} used multivariate effects plots to visualize estimated direct effects of different predictors in a paper that compared the difference in sexual risk behaviors between circumcised and uncircumcised men.
+\citep{shi_evidence_2017}, for example, used multivariate effects plots to visualize estimated direct effects of different predictors in a paper that compared the difference in sexual risk behaviors between circumcised and uncircumcised men.
 
-Generating quantities, i.e., central estimate together with the corresponding confidence intervals has some challenges. In particular:
+Producing this sort of outcome plot has some challenges, including:
 \begin{enumerate}
@@
 The central estimates are then calculated by holding the non-focal predictors at a reference point (a value chosen for a non-focal predictor) while varying the focal predictor, with the goal that the estimates represent how the model responds to the changes in the focal predictor \citep{fox2009effect, hanmer2013behind}.
-These values have been called: \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package} \jd{Is this also the same as least-squares means?} \BC{Yes.
-But was later renamed to emmeans}.
-In this article, we refer to these quantities as the \emph{central estimates} of the outcome.
-The mean of the central estimates is the prediction at the \emph{model center}.
+These values have been called: \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package} In this article, we refer to these quantities as the \emph{central estimates} of the outcome.
+
+\jd{The model center is the average of predictor rows of the model matrix.
+It is a calculation that this is equal to the mean of estimates for a linear model (identity link function).
+Maybe discuss later.} The mean of the central estimates is the prediction at the \emph{model center}.
 
@@
 We introduce an alternative choice for the reference point.
+\jd{This is where I would introduce model center. Maybe we should drop it? Or maybe your definition is better?}
 
@@
 As a result, the \emph{bias} in model center estimates can easily be corrected using mean-based approach and the estimates are usually consistent with the observed values.
+\jd{Why is there a bias to be corrected?
+Is this stuff we're going to explain later?}
 However, in a model with non-linear link function, this is not usually true.
-In particular, when dealing with a generalized model with non-linear link function, the averaging is done on the non-linear scale, i.e., non-linear averaging, making generating correct predictions much harder.
-This is the bias in the expected mean prediction induced by the non-linear transformation of the response variable.
-An alternative to the mean-based reference point is the \emph{observed-value-based} approach, discussed in detail later, which involves computing the prediction over the population of non-focal predictors and then averaging across the values of the focal predictor \citep{hanmer2013behind}.
+When averaging is done on a separate link scale, the mean of the estimates is not the same as the estimate at the mean point.
+This leads to bias: in this case a systematic difference between the values seen on average for a given value of the focal predictor and the value predicted by the mean-based approach. 
+An alternative to the mean-based reference point is the \emph{observed-value-based} approach, discussed later, which involves computing the prediction over the population of non-focal predictors and then averaging across the values of the focal predictor \citep{hanmer2013behind}.
 
-This article aims to discuss and implement various approaches for computing predictions and effects, together with the associated plots.
-We further explore and demonstrate, using simulated data, approaches for correcting bias in central estimates for generalized (mixed) models involving non-linear link functions.
-The proposed method and \proglang{R} software package will complement the existing ones by providing: 1) a straightforward way to generate effects, and 2) an alternative and a more robust way to correct for prediction bias in generalized (mixed) models.
-
-\jd{We should revisit this list.}
+This article will discuss and implement various approaches for computing predictions and effects plots.
+We further explore and demonstrate, using simulated data, approaches for correcting bias in central estimates for generalized models involving non-linear link functions, including models with random effects.
+The proposed method and \proglang{R} software package will complement the existing ones by providing: 1) a straightforward way to generate effects plots (in our sense), and 2) a robust way to correct for non-linear averaging bias in generalized (mixed) models.
 
@@
 
-\jd{I guess this list should really be a desc, not an itemize.}\BC{Clarify}
+\begin{description}
+\item [Input variables] Refers to the observed (or scientific) variables underlying an inference or exploration.
+For example, the regression models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher} both have $3$ input variables -- $x_1, x_2, x_3$.
+\item [Focal predictors] We call the input variable on the x-axis of an outcome plot the focal predictor.  Any other predictor variables are ``non-focal'' predictors.
 
-\begin{itemize}
-\item \textbf{Input variables:} Refers to the observed (or scientific) variables underlying an inference or exploration.
-For example, the regression models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher} both have $3$ input variables -- $x_1, x_2, x_3$.
-\item \textbf{Focal predictors}: We call the input variable on the x-axis of an outcome plot the focal predictor.  Any other predictor variables are ``non-focal'' predictors.
-
-\item \textbf{Model matrix:} Refers to the design matrix whose rows include all combination of input variables.
+\item [Model matrix] Refers to the design matrix whose rows include all combination of input variables.
 Consider an example for three hypothetical households -- the first household head is a Christian with an income of $\$ 50$, the second household head is Muslim with an income of $\$ 100$ while the third household head is a Jew with an income of $\$ 77$.
@@
 
-\item \textbf{Linear predictor variables}: Refer to the variables which are combined to make the linear predictor (corresponding to the columns in the model matrix).
+\item [Linear predictor variables] Refer to the variables which are combined to make the linear predictor (corresponding to the columns in the model matrix).
 Each input variable may correspond to one or more linear predictor variables.
@@
 
-\item \textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more linear predictor variables).
+\item [Model center] A point corresponding to a column-wise mean of the model matrix (the mean of one or more linear predictor variables).
+In a simple linear model with no interactions, the center point is just the point corresponding to the mean of the input variables.
 Also referred to as center point.
 The center point for a set of linear predictor variables corresponding to an input variable may not represent a possible value of the input variable -- the case in \FREF{justify_plots}B.
-In a simple linear model with no interactions, the mean of the input variable can be used as a center point.
 If there are more than one linear predictor variables associated with the focal predictor, however, the model center does not correspond to a single value of the focal predictor, see \FREF{pred_cubic_plots}.
+\jd{Here we're using my original definition of model center.}
 
-\item \textbf{Reference point:} The values (or sets of values) chosen for non-focal predictors, when estimating the predictions and effects.
+\item [Reference point] The values (or sets of values) chosen for non-focal predictors, when estimating the predictions and effects.
 Typically the center point, but can also be chosen as a baseline value or as a mean across categories.
@@
 
-\item \textbf{Anchor:} The value chosen for the focal predictor when estimating effect-style confidence intervals.
+\item [Anchor] The value chosen for the focal predictor when estimating effect-style confidence intervals.
 The anchor choice does not affect the central estimates, nor prediction-style.
@@
 
-\item \textbf{Prediction-style and effect-style plots:} A prediction-style plot is about predicting observations for a given value of the focal predictor.
+\item [Prediction-style and effect-style plots] A prediction-style plot is about predicting observations for a given value of the focal predictor.
 For this, we want to use the classic curved confidence intervals.
@@
 
-\end{itemize}
+\end{description}
 

### 2022-07-20 manuscript/body.tex
share with Bicko

@@
 
-We start by comparing our proposed approach with the existing implementations and then illustrate the construction of outcome plots and also demonstrate that the mean-based and observed-value-based approaches produce different results in models with nonlinear link functions and/or additional source(s) of potential bias.
+We start by comparing our proposed approach with the existing implementations and then illustrate the construction of outcome plots and also demonstrate that the mean-based and observed-value-based approaches produce different results in models with nonlinear link functions \jd{What‽: and/or additional source(s) of potential bias}.
 In addition, we demonstrate that observed-value-based approach can be used for bias correction.
@@
 The most commonly used \proglang{R} software packages for outcome plots (\pkg{emmeans} and \pkg{effects}), by default, use the average of input variables and ``sum-to-zero'' as the reference point approach for continuous and categorical input variables, respectively.
-However, there are a number of choices one can make when constructing outcome plots -- for example, in the presence of interactions, averaging the interactions (averaging the product of linear predictor variables) versus taking the product of the averages of the input variables (the default for \pkg{emmeans} and \pkg{effects}).
-For simple OLS models (with no interactions), the two approaches give the same estimates.
-However, for complex models, we demonstrate that the two approaches can produce substantially different results and show that averaging the interactions closely matches the observed values.
-To illustrate this, consider models description by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher} below, with $x_1$ as the focal predictor:
+However, there are a number of choices one can make when constructing outcome plots -- for example, in the presence of interactions, the default for \pkg{emmeans} and \pkg{effects} is to average the predictor variables and use these averaged values for the interaction, as opposed to our preferred model-center approach (averaging each linear predictor variable separately).
+The packages give the same results as our method for models without interactions, but when interactions are present, the two approaches can produce substantially different results, with the model-center approach more closely matching the observed values.
+
+To illustrate this, we simulate data from the models below:
 %
@@
 %
-We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$, $\beta_{23} = 5$, $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and $\sigma^2 = 1$, and then compare estimates from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' central estimates (based on the true simulation parameters) and observed average, i.e., $\bar{y}$, as shown in \FREF{justify_plots}.
+We simulate using $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$, $\beta_{23} = 5$, $\{x_{1,2,3}, \epsilon\} \sim \mathrm{Normal}(0, 1)$, and then compare estimates from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' central estimates (calculated using the simulation parameters) and observed average, i.e., $\bar{y}$, as shown in \FREF{justify_plots}.
 %
@@
 \caption{{\bf A comparison of central estimates for \pkg{emmeans}, \pkg{effects} and \pkg{varpred}.} The horizontal dashed lines are the mean of the central estimates and simulated $y$, $\bar{\hat{y}}$ and $\bar{y}$, respectively.
-The grey points are the binned simulated $y$.
+The grey points are averages of the simulated points binned according to value of $x_1$.
 The trend lines represent the $\hat{y}$ -- central estimate.
-A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the simulated one in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$.
-Similarly, the central estimate based on the three approaches match the truth.
-B: Even with the simple interaction between non-focal predictors, we see differences between predicted mean, $\bar{\hat{y}}$ and simulated mean, $\bar{y}$.
+A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, is the same in all three approaches, and closely matches the true mean; the central estimate likewise matches the truth (horizontal lines).
+B: With a simple interaction between non-focal predictors, we see differences between predicted mean, $\bar{\hat{y}}$ and simulated mean, $\bar{y}$.
 Similarly, there are differences between central estimates and the truth, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not the proposed \pkg{varpred}.}

### 2022-07-20 manuscript/justify.R

@@ -9,7 +9,7 @@ set.seed(9991)
 
 ############################################################################################
 # Global variables
-nHH_obs <- 30
+nHH_obs <- 50
 perHH <- 1
 
 ############################################################################################

### 2022-07-20 manuscript/body.tex
some progress

@@
 
-We start by comparing our proposed approach with the existing implementations and then illustrate the construction of outcome plots and also demonstrate that the mean-based and observed-value-based approaches produce different results in models with nonlinear link functions \jd{What‽: and/or additional source(s) of potential bias}.
-In addition, we demonstrate that observed-value-based approach can be used for bias correction.
+We start by comparing our proposed approach with the existing implementations and then illustrate the construction of outcome plots and also demonstrate that the mean-based approach works well for interactions and multi-parameter variables, and that and observed-value approach can correct for bias induced by non-linear averaging.
 

### 2022-07-21 manuscript/body.tex
working before dentist

@@
 \subsubsection{Dealing with multi-parameter variables}
-
-\jd{I feel the first s.~is perfectly clear, and I don't know why it needs any explanation at all.}
-\BC{I suggest we keep the illustrations.}
 
@@
 A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, is the same in all three approaches, and closely matches the true mean; the central estimate likewise matches the truth (horizontal lines).
-B: With a simple interaction between non-focal predictors, we see differences between predicted mean, $\bar{\hat{y}}$ and simulated mean, $\bar{y}$.
-Similarly, there are differences between central estimates and the truth, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not the proposed \pkg{varpred}.}
+B: With a simple interaction between non-focal predictors, results from \pkg{emmeans} and \pkg{effects}, but not from the proposed \pkg{varpred}, are biased}
 \label{fig:justify_plots}
@@
 However, in the presence of interactions, even as simple as the one in \EREF{simple_inter_higher}, the estimates start to differ.
-In particular, \pkg{emmeans} and \pkg{effects} have similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is very close to the simulated average ($\bar{y}$), \FREF{justify_plots}B.
-To generate \FREF{justify_plots}A,  \pkg{emmeans} and \pkg{effects} average $x_2$ and $x_3$, i.e., the input variables while \pkg{varpred} averages the linear predictor variables corresponding to $x_2$ and $x_3$.
-On other hand, to generate \FREF{justify_plots}B, the difference in the estimates lies on how each of the packages average the interaction linear predictor variable ($x_2x_3$).
-In particular, \pkg{emmeans} and \pkg{effects} use input variables to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses linear predictor variables to compute $\over{x_2x_3}$.
-In the case of \FREF{justify_plots}A, averaging the input variables is equivalent to averaging the linear predictor variables.
+In particular, estimates from \pkg{emmeans} and \pkg{effects} but differ from \pkg{varpred}'s, which is very close to the simulated average ($\bar{y}$), \FREF{justify_plots}B.
+In the simple model \FREF{justify_plots}A the input variables are the same as the linear predictor variables, so all the two methods produce identical results.
+In the interaction model \FREF{justify_plots}B, there is an additional linear predictor variable ($x_2x_3$).
+\pkg{emmeans} and \pkg{effects} first average the input variables to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} first calculates the corresponding vector of linear predictor values and then averages.
 
+\jd{I still feel you're too interested in the comparison. Can't we first explain want we're trying to do? And then explain the ways of doing it.}
 One may be interested in the uncertainties associated with the focal predictor only, excluding other uncertainties due to other non-focal predictors -- effect plots.

### 2022-07-21 manuscript/body.tex
confused now

@@
 A: The wider dashed curves corresponds to the conventional prediction curves from \pkg{emmeans} and \pkg{effects} packages, while the narrower curves crossing at the center point represent the effect from \pkg{varpred}.
-For simple OLS models, the effect-styled curves cross at the center point (which is the default anchor and also referred to as center-anchored).
-The prediction-styled curves incorporates not only the uncertainties due to intercept term but also other non-focal predictors.
-On the other hand, effect-styled curves only take into account the main effect uncertainty of the focal predictor.
+For simple OLS models, the effect-style curves cross at the center point (which is the default anchor and also referred to as center-anchored).
+The prediction-style curves incorporates not only the uncertainties due to intercept term but also other non-focal predictors.
+On the other hand, effect-style curves only take into account the main effect uncertainty of the focal predictor.
 B: center- and zero- anchored effects.
@@
 For simple OLS models, the point where the curves cross is the model center or simply center point; it corresponds to anchor we choose, in this case, the center point (which is generally an appropriate and stable choice).
+\jd{I agree with my point above even more now.
+The main point is that varpred can make \emph{both} effect-style and prediction-style plots, and to show those.
+As far as I'm concerned, we can just briefly say that it's hard to make effect-style plots in the other packages?
+If it even is, I guess?}
 
+\jd{Why is this here? How does panel A compare to what we would get from either effect- or prediction-style plots?}
 Lastly, we compare effect plots and the marginal effects \citep{leeper2017interpreting} plots implemented in package \pkg{margins} \citep{lenth2018package}.
@@
 
-\subsection{Prediction and effect styled plots for MPVs}
+\subsection{Prediction- and effect-style plots for MPVs}
 
@@
 In both cases the, we used center point as the anchor for the effects, as shown in \FREF{pred_cubic_plots}.
-
+\jd{Deleted some stuff about transformations: I don't think transformations affect the intersecting part. }
+In a model with a simple focal predictor, we expect the CIs for effect-style plots to cross at the anchor point.
+On the other hand, if the focal predictor is a MPV, the center point is not expected to correspond to a single value of the focal predictor \jd{It's nothing to do with range; it's an \emph{impossible} value (e.g., $x=1; x^2=5$)}.
+In this case, the effects will be narrower than the predictions plots but will not necessarily intersect.
+For linear models, we expect the observed average, the average of the predictions and the prediction at the model to all be identical. 
 \begin{figure}
@@
 B: The focal predictor is linear, with MPV non-focal predictor.
-The central estimates (dashed central curves or lines) are the same for predictions and effects, in respective case (A and B).
-However, the effects are narrower than the predictions.
-The default anchor for generating effects is the center point.
-For linear (or simple) focal predictor, the center point is a point within the range of focal values; and therefore the effect curves intersect at that that point -- B.
-In the case of MPVs such as cubic polynomials, the center point is not a point but a combination of all the linear predictor variables associated with the cubic polynomial, consequently, curves will not necessarily cross at a point -- A.
+The central estimates (dashed central curves or lines) are the same for predictions and effects in both cases.
+We use the model center as the anchor for effect-style CIs.
+For a simple predictor (panel B), this corresponds simply to the mean of the input variable, thus the effect curves intersect at that that point.
+For an MPV predictor (panel A), the center point does not correspond to a single value of the predictor, and the curves do not cross.
 The horizontal black and yellow dashed lines are the observed and predicted average household size, i.e., $\over{hh~size}$ and $\over{\widehat{hh~size}}$, respectively.}
@@
 
-In the absence of, complex interactions, MPVs, or transformations on the focal predictor, in simple OLS models we would expect the observed average, $\over{hh~size}$, and the average of the predictions, $\over{\widehat{hh~size}}$, to be identical and intersect with the trend lines at the center point, for the effect plot.
-On the other hand, if the focal predictor is a MPVs or has any other form of transformation, the center point is not necessarily a point in the range of focal predictor.
-In this case, the effects will be narrower than the predictions plots but will not necessarily intersect.
-
 \subsection{Bias correction}
 
-In a generalized linear model with at least one non-focal predictor, mean-based and observed-value-based approaches can produce very different.
-To demonstrate this, we considered a two input variables binary outcome simulation described in \nameref{S2_Appendix}, such that \code{Age} has a very small effect size in comparison to \code{Wealthindex}, and compared their effects on the estimated probability of improved water quality as shown in \FREF{pred_bin_plots}.
+\jd{Why is this your lead in?
+I don't understand the focus on us being “different”.
+Did the \pkg{pcoxtime} reviews scare you?
+We did all of this because our uncorrected predictions looked \emph{wrong}.
+Tie this back to your discussion of the Jensen stuff.}
+
+In a generalized linear model with at least one non-focal predictor, mean-based and observed-value-based approaches can produce very different \jd{different what?}.
+To demonstrate this, we considered a binary-outcome simulation with two input variables binary outcome simulation (described in \nameref{S2_Appendix}), such that \code{Age} has a very small effect size in comparison to \code{Wealthindex}, and compared their effects on the estimated probability of improved water quality as shown in \FREF{pred_bin_plots}.
+\jd{Is it easy to explain why we did it this way?}
 

### 2022-07-21 manuscript/body.tex
modified:   body.tex

@@
 \section{Bias correction}
+
+\jd{After conversation: please re-write this whole section to: describe the “bias” -- talk about Jensen, the problem is that the curve is not actually wrong, but is above or below the points, which is confusing; re-do the plots without any “truth”, the point is to show that bias correction is a sensible way to do it that has the curve closer to the points.}
 

### 2022-07-21 manuscript/body.tex
handing back!

@@
 So does marriage \emph{cause} divorce?
+\jd{This is a very weird example.
+Of course marriage causes divorce in the statistical sense (if you reduce marriage, you reduce divorce).
+Please see if you can come up with a good, clear example of mediation.}
 
@@
 
+\jd{We don't \emph{need} more than one model; better to say there is more than one question we could ask. In this case, we should make a distinction between direct and total effects.}
 To make inferences about this mediated effect, we need more than one model.
@@
 In both cases, we compared mean-based and the observed-value-based central estimates, as shown in \FREF{pred_mediated_plots}.
+
+\jd{“no” direct effect is a weird and confusing choice.
+The absence of direct effect is not what's causing the bias: it's the presence of the indirect effect.
+Please re-do with a relatively small (but non-zero) direct effect.}
 
@@
 
+\jd{I'd like to see a better explanation here as well.
+In my opinion, the reason for the mismatch is non-linear averaging over the non-focal variable; I don't think dilution is a good metaphor (a linear model could be equally diluted, but would not have this effect, I think?).}
 From \FREF{pred_mediated_plots}A, we see what we would expect in the absence of additional sources of bias, even though in the simulation, the effect of $x$ on $z$ is mediated through $y$.
@@
 
+\jd{This doesn't seem like a good starting point. I guess the point is that you're worried about taking out random effects? Don't worry, and let's just take them all the way out!}
 Our simulations examples focused on simple linear and logistic models due their wide range of usage and application.
@@
 
+\jd{and now we're left with only one, boring Discussion \P. Maybe merge Discussion and Conclusion.}
 Although commonly used \proglang{R} software packages, by default, implements mean-based approach, our simulation results demonstrated that the observed-value-based approach has a potential of yielding results which are more consistent with the observed data.
@@
 Moreover, we provide \proglang{R} package, \pkg{vareffects}, which implements these methods and is available on github (\href{https://github.com/mac-theobio/effects}{https://github.com/mac-theobio/effects}).
+
+\jd{Let's do a better job of summarizing what we did.
+We talk about different purposes of outcome plots, and try to distinguish the goals of predicition and effect.
+We talk about the model center and why it's a good reference point (maybe we should talk more about why we think it's a good anchor point as well … We explain a fairly weird kind of “bias” (an accurate kind of curve that does not match well with the data) and re-discover and implement a cool way of dealing with it.}
 
@@
 
-\textbf{Conceptualization:} Jonathan Dushoff
+\textbf{Conceptualization:} Jonathan Dushoff, Steve Cygu
 

### 2022-07-22 manuscript/body.tex
Upto mediation

@@
 
-\jd{The model center is the average of predictor rows of the model matrix.
-It is a calculation that this is equal to the mean of estimates for a linear model (identity link function).
-Maybe discuss later.} The mean of the central estimates is the prediction at the \emph{model center}.
-
 In a model with non-focal predictors such as multivariate models, reference points can be chosen as the average of the non-focal \emph{linear predictor variables} -- we call this approach \emph{mean-based} reference point and is currently not implemented in commonly used \proglang{R} software packages.
 We introduce an alternative choice for the reference point.
-\jd{This is where I would introduce model center. Maybe we should drop it? Or maybe your definition is better?}
 
@@
 \jd{Why is there a bias to be corrected?
-Is this stuff we're going to explain later?}
+Is this stuff we're going to explain later?} \BC{We talk about this in Bias correction.}
 However, in a model with non-linear link function, this is not usually true.
@@
 
-\item [Model center] A point corresponding to a column-wise mean of the model matrix (the mean of one or more linear predictor variables).
-In a simple linear model with no interactions, the center point is just the point corresponding to the mean of the input variables.
-Also referred to as center point.
-The center point for a set of linear predictor variables corresponding to an input variable may not represent a possible value of the input variable -- the case in \FREF{justify_plots}B.
-If there are more than one linear predictor variables associated with the focal predictor, however, the model center does not correspond to a single value of the focal predictor, see \FREF{pred_cubic_plots}.
-\jd{Here we're using my original definition of model center.}
+\item [Focal model center] The average of the linear predictor variables corresponding to the focal predictor.
+The average of the linear predictor variables corresponding to the non-focal predictor(s) is \emph{non-focal model center}.
+For simple continuous input variables, averaging linear predictor variables is the same as averaging the input variables (see \FREF{justify_plots}).
+Focal and non-focal model center values is also referred to as \emph{center point}.
+
+\item [Model center] Is the value of the central estimate calculated using focal and non-focal model center values (center point).
+It is a calculation that is equal to the average of central estimates for a linear model (identity link function) without complex interactions.
 
@@
 
-\jd{I still feel you're too interested in the comparison. Can't we first explain want we're trying to do? And then explain the ways of doing it.}
-One may be interested in the uncertainties associated with the focal predictor only, excluding other uncertainties due to other non-focal predictors -- effect plots.
-However, as previously mentioned, the two packages do not provide a straightforward way do achieve this.
-To illustrate this, we consider \EREF{simple_inter_higher_no_interaction} and use $x_2$ as the focal predictor, and generate the outcome plot shown in \FREF{justify_ci_plots}.
+Our implementation, \pkg{varpred}, can generate both prediction-style and effect-style plots.
+However, as previously mentioned, it is hard to generate an effect-style plot in \pkg{emmeans} and \pkg{effects}.
+We consider \EREF{simple_inter_higher_no_interaction} and use $x_2$ as the focal predictor, and compare prediction and effect plots using \pkg{varpred}.
+We also compare different anchors for effect plot.
+See \FREF{justify_ci_plots}.
 %
@@
 \caption{{\bf Prediction and effect plots.} The description of horizontal, vertical and trend lines remain the same as above.
-A: The wider dashed curves corresponds to the conventional prediction curves from \pkg{emmeans} and \pkg{effects} packages, while the narrower curves crossing at the center point represent the effect from \pkg{varpred}.
+A: The wider dashed curves corresponds to the conventional prediction curves, while the narrower curves crossing at the center point represent the effect from \pkg{varpred}.
 For simple OLS models, the effect-style curves cross at the center point (which is the default anchor and also referred to as center-anchored).
@@
 %
-For \pkg{emmeans} and \pkg{effects}, the confidence intervals are much wider because they include uncertainties associated with the intercept and non-focal predictors, but narrower and crosses at the mean of the focal predictor in \pkg{varpred}.
+For a prediction-style plot, the confidence intervals are much wider because they include uncertainties associated with the intercept and non-focal predictors, but narrower and crosses at the mean of the focal predictor.
 In other words, with \pkg{varpred}, we are able to generate effects indicating zero uncertainty at the value of the focal predictor we are more certain about, i.e., the anchor.
 For simple OLS models, the point where the curves cross is the model center or simply center point; it corresponds to anchor we choose, in this case, the center point (which is generally an appropriate and stable choice).
-\jd{I agree with my point above even more now.
-The main point is that varpred can make \emph{both} effect-style and prediction-style plots, and to show those.
-As far as I'm concerned, we can just briefly say that it's hard to make effect-style plots in the other packages?
-If it even is, I guess?}
-
-\jd{Why is this here? How does panel A compare to what we would get from either effect- or prediction-style plots?}
-Lastly, we compare effect plots and the marginal effects \citep{leeper2017interpreting} plots implemented in package \pkg{margins} \citep{lenth2018package}.
-To illustrate this, we use the simulated data described in \nameref{S1_Appendix}.
-The results are shown in \FREF{qoi_age_pred_plot}.
-%
-\begin{figure}
-\centering
-\includegraphics{cubic_varpred_margins_pred.ggp.pdf}
-\caption{{\bf Effect and marginal effect plots.} \NEW{A:  The effect plot.
-B: The marginal effect (ME) at representative values of the focal predictor \code{age} together with the $95\%$ confidence intervals.
-For simple linear models with no interactions, the ME should be constant across all the values of the focal predictor; and it would correspond to the slope of the central estimate.
-However, in this example, the focal predictor is a cubic polynomial, and so, the effect is not constant but a function of the partial derivatives of the corresponding linear predictor variables.}}
-\label{fig:qoi_age_pred_plot}
-\end{figure}
-%
 
@@
 In both cases the, we used center point as the anchor for the effects, as shown in \FREF{pred_cubic_plots}.
-\jd{Deleted some stuff about transformations: I don't think transformations affect the intersecting part. }
-In a model with a simple focal predictor, we expect the CIs for effect-style plots to cross at the anchor point.
-On the other hand, if the focal predictor is a MPV, the center point is not expected to correspond to a single value of the focal predictor \jd{It's nothing to do with range; it's an \emph{impossible} value (e.g., $x=1; x^2=5$)}.
-In this case, the effects will be narrower than the predictions plots but will not necessarily intersect.
-For linear models, we expect the observed average, the average of the predictions and the prediction at the model to all be identical. 
+%
 \begin{figure}
@@
 \end{figure}
+%
+In a model with a simple focal predictor, we expect the CIs for effect-style plots to cross at the anchor point.
+On the other hand, if the focal predictor is a MPV, the center point is not expected to correspond to a single value of the focal predictor.
+In this case, the effects will be narrower than the predictions plots but will not necessarily intersect.
+For linear models, we expect the observed average, the average of the central estimates and the prediction at the model to all be identical.
 
@@
 
-\jd{Why is this your lead in?
-I don't understand the focus on us being “different”.
-Did the \pkg{pcoxtime} reviews scare you?
-We did all of this because our uncorrected predictions looked \emph{wrong}.
-Tie this back to your discussion of the Jensen stuff.}
-
-In a generalized linear model with at least one non-focal predictor, mean-based and observed-value-based approaches can produce very different \jd{different what?}.
-To demonstrate this, we considered a binary-outcome simulation with two input variables binary outcome simulation (described in \nameref{S2_Appendix}), such that \code{Age} has a very small effect size in comparison to \code{Wealthindex}, and compared their effects on the estimated probability of improved water quality as shown in \FREF{pred_bin_plots}.
-\jd{Is it easy to explain why we did it this way?}
-
+We simulated data motivated by the water, sanitation, and hygiene (WaSH) study in which we were interested in investing the contribution of demographic and socio-economic factors to improved WaSH indicators among slum dwellers in Nairobi, Kenya.
+We used the mean-based approach to generate the predicted probabilities.
+However, we noticed that the predictions consistently over- or under- estimated the observed proportions; and did not align well with the observed data points.
+To demonstrate this, we consider a binary-outcome simulation with two input variables (described in \nameref{S2_Appendix}), such that \code{Age} has a very small effect size in comparison to \code{Wealthindex}, and compared their effects on the estimated probability of improved water quality as shown in \FREF{pred_bin_plots}.
+%
 \begin{figure}
@@
 A \emph{mediated} effect is a situation in which only some of the input variables have a \emph{causal} impact on the outcome, even though all the variables are (somehow) associated with the outcome.
-For example, the age at marriage may be strongly associated with divorce rate but only married people divorce.
-So does marriage \emph{cause} divorce?
-\jd{This is a very weird example.
-Of course marriage causes divorce in the statistical sense (if you reduce marriage, you reduce divorce).
-Please see if you can come up with a good, clear example of mediation.}
+For example, psychological factors such as water taste in the relationship between household socio-economic status and improved water quality status.
 
-For simplicity, we consider an hypothetical simulation for three variables: input variables $x$ and $y$; and outcome variable $z$; such that
+For simplicity, we consider a hypothetical simulation for three variables: input variables $x$ and $y$; and outcome variable $z$; such that
 %

### 2022-07-24 manuscript/body.tex
A little rest!

@@
 
-When dealing with nonlinear link functions and additional non-focal predictors,  generated estimates may not reflect the observed response due to the bias in the expected mean induced by the nonlinear transformation of the response variable -- as illustrated by the two examples above.
+When dealing with non-linear link functions and additional non-focal predictors,  generated estimates may not reflect the observed response due to the bias in the expected mean induced by the non-linear transformation of the response variable -- as illustrated by the two examples above.
 In such cases, bias correction is needed when back-transforming the estimates to the original scales.
@@
 An alternative approach to choosing a reference point is to make predictions over all observations of the non-focal predictors (members of the population) \citep{hanmer2013behind}.
-The nonlinear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population to a one-dimensional distribution of $\etahfi(x_f, \nset{x})$, which is a function of the chosen values of the focal predictor and the whole sample of non-focal predictors, as opposed to the definition in \EREF{eta_mean}.
+The non-linear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population to a one-dimensional distribution of $\etahfi(x_f, \nset{x})$, which is a function of the chosen values of the focal predictor and the whole sample of non-focal predictors, as opposed to the definition in \EREF{eta_mean}.
 More specifically:
@@
 
-In the observed-value-based approach, the ensemble of predictions and CIs are back-transformed before averaging, see \EREF{pop_response}, so we do not need to worry about the nonlinear averaging.
+In the observed-value-based approach, the ensemble of predictions and CIs are back-transformed before averaging, see \EREF{pop_response}, so we do not need to worry about the non-linear averaging.
 In other words, the averaging is no longer on the link scale and is likely to be bounded by the original data scale.
@@
 In both cases, the observed-value-based estimates are closer to the truth.
-The difference in mean-based and observed-value-based is due to the bias induced by the non-focal predictor and the nonlinear link function in the logistic model.
-Mean-based approach suffers from nonlinear averaging; and if focal predictor has small effect size, the bias is even pronounced since the effect of nonlinear averaging is driven by the non-focal predictor(s) with the strong effect.
+The difference in mean-based and observed-value-based is due to the bias induced by the non-focal predictor and the non-linear link function in the logistic model.
+Mean-based approach suffers from non-linear averaging; and if focal predictor has small effect size, the bias is even pronounced since the effect of non-linear averaging is driven by the non-focal predictor(s) with the strong effect.
 However, this is not the case if the non-focal predictor(s) have small effect since the bias they induce is equally likely to be small.
-Since observed-value-based approach averages over the whole population of the non-focal predictor variables, it is less likely to be affected by the nonlinear averaging for the reasons already mentioned.
+Since observed-value-based approach averages over the whole population of the non-focal predictor variables, it is less likely to be affected by the non-linear averaging for the reasons already mentioned.
 The horizontal dashed lines correspond to the respective averages.
@@
 
-If there was no effect of nonlinear averaging, then we would expect the average observed proportion and the average predictions to intersect at the center point as we see in \FREF{pred_cubic_plots}B.
-One possible reason for the variations we see in \FREF{pred_bin_plots} is the nonlinear averaging; since both observed status and predicted probabilities are averaged on the response scale as opposed to link scale.
+If there was no effect of non-linear averaging, then we would expect the average observed proportion and the average predictions to intersect at the center point as we see in \FREF{pred_cubic_plots}B.
+One possible reason for the variations we see in \FREF{pred_bin_plots} is the non-linear averaging; since both observed status and predicted probabilities are averaged on the response scale as opposed to link scale.
 For example, if the range of values are bigger than $0.5$ (seemingly the case here), then we would expected the averages to be slightly higher than what we would expect at the center point.
@@
 
-\jd{We don't \emph{need} more than one model; better to say there is more than one question we could ask. In this case, we should make a distinction between direct and total effects.}
-To make inferences about this mediated effect, we need more than one model.
-First, the regression of $z$ on $x$, i.e., the \emph{non-mediated} (or univariate) model which tells on the \emph{total} effect of $x$ on $z$; and second, the regression of $z$ on $x$ and $y$, i.e., the \emph{mediated} (or multivariate or full) model.
+To make inferences about this mediated effect, we can ask questions about the \emph{total} and \emph{direct} effects.
+In the first case, we want to fit a univariate model which regresses $z$ on $x$, excluding $y$.
+We refer to this as \emph{non-mediated} model.
+In the second case, we want to fit a multivariate model which regresses $z$ on $x$ and $y$.
+We refer to this as \emph{mediated} model.
 We simulated the data as shown in \nameref{S3_Appendix} and fitted the two models -- non-mediated and mediated.
 In both cases, we compared mean-based and the observed-value-based central estimates, as shown in \FREF{pred_mediated_plots}.
-
-\jd{“no” direct effect is a weird and confusing choice.
-The absence of direct effect is not what's causing the bias: it's the presence of the indirect effect.
-Please re-do with a relatively small (but non-zero) direct effect.}
-
+%
 \begin{figure}
@@
 \end{center}
-\caption{{\bf Mean-based and observed-value-based estimates for mediated (multivariate) and non-mediated (univariate) models.} A: In the absence of mediator variable, both mean-based and observed-value-based approaches central estimates are identical since there is no any other additional (non-focal predictor) sources of bias.
-Consequently, in both approaches, the average estimated probability is very close to the observed proportion.
-B: When the mediator variable is included,  $x$ has no direct effect on $z$ and as a result the central estimates do not align with the observations.
-However, the observed-value-based approach still closely approximates the average proportion of the simulated data.
-The horizontal dashed lines are the respective average estimates and observed proportions.
+\caption{{\bf Mean-based and observed-value-based estimates for non-mediated (total) effect models and mediated (direct).} A: In the absence of mediator variable, the central estimate aligns well with the observed data.
+B: In the presence of indirect effect (mediator variable included), the central estimates do not align with the observations.
+Since the models in A and B are both linear models with identity link function, both approaches (mean-based and observed-value) give the same estimates.
+The horizontal dashed lines are the respective average estimates and observed means ($\bar{\hat{z}} \approx \bar{z}$).
 The vertical dashed black represents the center point, and the point at which it crosses the horizontal lines represents the expected ``perfect'' estimate at the model center.
@@
 \end{figure}
+%
+From \FREF{pred_mediated_plots}A, we see what we would expect if we simulated a univariate model with no non-focal predictors, even though in the simulation, the effect of $x$ on $z$ is mediated through $y$.
+This is the total effect of $x$, which only tells us the influence of $x$ on $z$.
+By ignoring $y$ in the model, we are still able to capture the effect of $x$ and match the observed values using both approaches.
+However, if we control for the mediator variable $y$, the estimates do not necessarily match the observed values since $x$ has indirect effect on $z$.
+By controlling for the mediator variable, we able to get an indication of the direct (vary week) effect of $x$ on $z$ (see \FREF{pred_mediated_plots}B).
 
-\jd{I'd like to see a better explanation here as well.
-In my opinion, the reason for the mismatch is non-linear averaging over the non-focal variable; I don't think dilution is a good metaphor (a linear model could be equally diluted, but would not have this effect, I think?).}
-From \FREF{pred_mediated_plots}A, we see what we would expect in the absence of additional sources of bias, even though in the simulation, the effect of $x$ on $z$ is mediated through $y$.
-This is the total effect of $x$, which only tells us the influence of $x$ on $z$.
-By ignoring $y$ in the model, we are still able to capture the effect of $x$ and closely match the observed values using both approaches.
-However, if there were additional non-focal predictors, we would expect to see the differences similar to those in \FREF{pred_bin_plots}A.
-Including both $y$ and $x$ ``dilutes'' the direct effect of $x$ on $z$ and as a result, our estimates do not necessarily match the observed binned observations (see \FREF{pred_mediated_plots}B).
+\section{Discussion and conclusion}
 
-\section{Discussion}
+Generalized linear (mixed) models have a wide usage in various fields including public health.
+In a model involving difficult-to-interpret coefficient estimates, an outcome plot can aid in understanding and summarizing the results.
+In particular, if the goal is to capture every uncertainty in the model for a particular focal predictor or if we are interested on total effect, then a prediction plot would be appropriate.
+On the hand, an effect plot is preferable if we want to focus on the uncertainty associated with a focal predictor only or if we are interested in the direct effect.
 
-\jd{This doesn't seem like a good starting point. I guess the point is that you're worried about taking out random effects? Don't worry, and let's just take them all the way out!}
+Mean-based approach is widely to generate outcome plots.
+However, in a model with complex interaction, MPVs or categorical variables, it is sensitive to the choice of the reference point.
+We have demonstrated that non-focal model-center-based reference point is generally a stable choice and provide estimates which are more consistent with the observed quantities.
+
+In a model with non-linear link function such as a logistic or exponential function, the generated central estimate curve may not match well with the observed data, i.e., our description for ``bias''.
+In such such a model, observed-value-based approach provides a way to generate more consistent estimates and is preferable over conventional mean-based approach.
+
+The argument and results we present in this paper support a greater need for a shift on focus on how to summarize regression models.
+From our theoretical, methodological and simulation results, researchers using these kind of models should, in the absence of theoretical justification, report predictions based on observed-value-based approach or at least attempt to do a comparison of the two approaches before settling on the most appropriate in answering their research question.
+Moreover, we provide \proglang{R} package, \pkg{vareffects}, which implements these methods and is available on github (\href{https://github.com/mac-theobio/effects}{https://github.com/mac-theobio/effects}).
+
 Our simulations examples focused on simple linear and logistic models due their wide range of usage and application.
 These models also act as a starting point for building other complex models, including mixed effect models and models with categorical predictors.
-The logic for extending to more complex models, including other forms of nonlinear link functions is very straightforward.
+The logic for extending to more complex models, including other forms of non-linear link functions is very straightforward.
 In fact the components needed for extension are the correct linear predictor and the inverse link function; everything else generalizes.
-In addition, our \proglang{R} package implementation already extends to and supports most of the nonlinear link functions and mixed model framework, including multivariate binary outcome models.
-
-\jd{and now we're left with only one, boring Discussion \P. Maybe merge Discussion and Conclusion.}
-Although commonly used \proglang{R} software packages, by default, implements mean-based approach, our simulation results demonstrated that the observed-value-based approach has a potential of yielding results which are more consistent with the observed data.
-We would, therefore, argue that the use of mean-based approach should have some theoretical justification, especially in complex models.
-
-\section{Conclusion}
-
-Generating outcome predictions or predicted probabilities from simple and generalized linear (mixed) models is not only important but also, generating quantities which are consistent with the observed values should be of interest.
-However, many studies still report coefficient estimates from generalized models like probit, logistic, etc., \citep{hanmer2013behind}, which are subject of less clarity and complexities in interpretation.
-
-The argument and results we present in this paper support a greater need for a shift on focus on how to present predictions from generalized models.
-For example, effect plots could provide more clarity concerning the uncertainty due to the input variable of interests as opposed to the conventional way of incorporating everything.
-
-From our theoretical, methodological and simulation results, researchers using these kind of models should, in the absence of theoretical justification, report predictions based on observed-value-based approach or at least attempt to do a comparison of the two approaches before settling on the most appropriate in answering their research question.
-Moreover, we provide \proglang{R} package, \pkg{vareffects}, which implements these methods and is available on github (\href{https://github.com/mac-theobio/effects}{https://github.com/mac-theobio/effects}).
+In addition, our \proglang{R} package implementation already extends to and supports most of the non-linear link functions and mixed model framework, including multivariate binary outcome models.
 

### 2022-07-24 manuscript/abstract.tex
Sending back to JD

@@
 
-In generalized linear (mixed) models that involve complex multiplicative interactions, multi-parameter variables, additional non-focal predictors or a nonlinear link function, outcome plots (prediction and effect plots) can aid in understanding difficult-to-interpret coefficient estimates.
+In generalized linear (mixed) models that involve complex multiplicative interactions, multi-parameter variables, additional non-focal predictors or a non-linear link function, outcome plots (prediction and effect plots) can aid in understanding difficult-to-interpret coefficient estimates.
 Outcome plots depend on the choices we make about the non-focal predictors.
@@
 We call this mean-based approach and it estimates effect of an average case in the population.
-In the presence of additional non-focal predictors, nonlinear link functions, random effect terms, etc., mean-based approach generates estimates that are biased and may not be consistent with the observed quantities.
+In the presence of additional non-focal predictors, non-linear link functions, random effect terms, etc., mean-based approach generates estimates that are biased and may not be consistent with the observed quantities.
 An alternative is the observed-value-based approach which estimates the average effect in the population.

### 2022-07-24 manuscript/body.tex
Sending back to JD

@@
 Doesn't it just work like magic with the lm machinery?
-and how to incorporate the uncertainty due to non-focal predictors} \BC{Clarify with JD.
-Another challenge or a continuation  of the above?}
+and how to incorporate the uncertainty due to non-focal predictors} \BC{\fix{Clarify with JD.
+Another challenge or a continuation  of the above?}}
 \item biases induced by non-linear transformations of the response variable in generalized linear models (especially generalized mixed models).
@@
 \jd{Why is there a bias to be corrected?
-Is this stuff we're going to explain later?} \BC{We talk about this in Bias correction.}
+Is this stuff we're going to explain later?} \BC{\fix{We talk about this in Bias correction.}}
 However, in a model with non-linear link function, this is not usually true.
@@
 
-\jd{After conversation: please re-write this whole section to: describe the “bias” -- talk about Jensen, the problem is that the curve is not actually wrong, but is above or below the points, which is confusing; re-do the plots without any “truth”, the point is to show that bias correction is a sensible way to do it that has the curve closer to the points.}
+When dealing with non-linear link functions and additional non-focal predictors,  generated estimates may not reflect the observed response due to the bias, i.e., a term we use to describe a situation in which the central curve does not align well with the observed data points -- the central curve is either below or above the points.
+In such cases, bias correction is needed when back-transforming the estimates to the original scales.
+The common approach for bias-adjustment is second-order Taylor approximation \citep{duursma2003bias, hanmer2013behind}; already implemented in \pkg{emmeans} \citep{lenth2018package}.
+Here, we describe and implement a different approach -- observed-value-based approach for bias correction.
+Hanmer and Kalkan \citep{hanmer2013behind} discussed and implemented observed-value-based and input-variable-based mean-based approaches in binary response models only.
+Our formulation is more general and can easily be extended to other link functions other than logistic.
 
@@
 In other words, the mean-based approach can under-estimate the prediction in exponential and low probability logistic functions and over-estimate in high probability logistic function.
-
-\fix{Thinking about RE}
-
-When dealing with non-linear link functions and additional non-focal predictors,  generated estimates may not reflect the observed response due to the bias in the expected mean induced by the non-linear transformation of the response variable -- as illustrated by the two examples above.
-In such cases, bias correction is needed when back-transforming the estimates to the original scales.
-The common approach for bias-adjustment is second-order Taylor approximation \citep{duursma2003bias, hanmer2013behind}; already implemented in \pkg{emmeans} \citep{lenth2018package}.
-Here, we describe and implement a different approach -- observed-value-based approach for bias correction.
-Hanmer and Kalkan \citep{hanmer2013behind} discussed and implemented observed-value-based and input-variable-based mean-based approaches in binary response models only.
-Our formulation is more general and can easily be extended to other link functions other than logistic.
 
@@
 
-For models with random effects components, we make further adjustment to correct for bias induced by the random effects terms.
-In observed-value approach, we treat the random effects terms as additional non-focal predictors and simply make adjustment to \EREF{pop_eta}.
+We make further adjustments for models with random effects components to correct the bias induced by the random effects components.
+We treat the random effects components as additional non-focal predictors in the observed-value approach and modify \EREF{pop_eta}.
 In particular
@@
 In general, the mean-based approach provides the predictions of an average case, whereas the observed-value-based approach summarizes the predictions over the entire population.
-In some applications, the effect of an average case might not be generalizable to the entire population, especially, if the average does not represent the population.
+In some applications, the effect of an average case might not be generalizable to the entire population, especially if the average does not represent the population.
 This might not be a problem in the observed-value-based approach since it focuses on specific observations -- the prediction is first obtained for each observation and then averaged across the levels of the focal predictor.
@@
 
-We start by comparing our proposed approach with the existing implementations and then illustrate the construction of outcome plots and also demonstrate that the mean-based approach works well for interactions and multi-parameter variables, and that and observed-value approach can correct for bias induced by non-linear averaging.
+We start by comparing our proposed approach with the existing implementations and then illustrate the construction of outcome plots.
+We also demonstrate that the mean-based approach works well for interactions and multi-parameter variables and that the observed-value approach can correct the bias induced by non-linear averaging.
 
@@
 However, there are a number of choices one can make when constructing outcome plots -- for example, in the presence of interactions, the default for \pkg{emmeans} and \pkg{effects} is to average the predictor variables and use these averaged values for the interaction, as opposed to our preferred model-center approach (averaging each linear predictor variable separately).
-The packages give the same results as our method for models without interactions, but when interactions are present, the two approaches can produce substantially different results, with the model-center approach more closely matching the observed values.
+The packages give the same results as our method for models without interactions.
+However, when interactions are present, the two approaches can produce substantially different results, with the model-center approach more closely matching the observed values.
 
@@
 \caption{{\bf A comparison of central estimates for \pkg{emmeans}, \pkg{effects} and \pkg{varpred}.} The horizontal dashed lines are the mean of the central estimates and simulated $y$, $\bar{\hat{y}}$ and $\bar{y}$, respectively.
-The grey points are averages of the simulated points binned according to value of $x_1$.
+The grey points are averages of the simulated points binned according to the value of $x_1$.
 The trend lines represent the $\hat{y}$ -- central estimate.
-A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, is the same in all three approaches, and closely matches the true mean; the central estimate likewise matches the truth (horizontal lines).
+A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, is the same in all three approaches and closely matches the actual mean (truth); the central estimate likewise matches the truth (horizontal lines).
 B: With a simple interaction between non-focal predictors, results from \pkg{emmeans} and \pkg{effects}, but not from the proposed \pkg{varpred}, are biased}
@@
 In the absence of interactions (\EREF{simple_inter_higher_no_interaction}), the three approaches produce identical estimates, which match the simulated values, \FREF{justify_plots}A.
-However, in the presence of interactions, even as simple as the one in \EREF{simple_inter_higher}, the estimates start to differ.
+However, the estimates start to differ in the presence of interactions, even as simple as the one in \EREF{simple_inter_higher}.
 In particular, estimates from \pkg{emmeans} and \pkg{effects} but differ from \pkg{varpred}'s, which is very close to the simulated average ($\bar{y}$), \FREF{justify_plots}B.
@@
 We consider \EREF{simple_inter_higher_no_interaction} and use $x_2$ as the focal predictor, and compare prediction and effect plots using \pkg{varpred}.
-We also compare different anchors for effect plot.
+We also compare different anchors for the effect plot.
 See \FREF{justify_ci_plots}.
@@
 \end{center}
-\caption{{\bf Prediction and effect plots.} The description of horizontal, vertical and trend lines remain the same as above.
-A: The wider dashed curves corresponds to the conventional prediction curves, while the narrower curves crossing at the center point represent the effect from \pkg{varpred}.
-For simple OLS models, the effect-style curves cross at the center point (which is the default anchor and also referred to as center-anchored).
-The prediction-style curves incorporates not only the uncertainties due to intercept term but also other non-focal predictors.
-On the other hand, effect-style curves only take into account the main effect uncertainty of the focal predictor.
+\caption{{\bf Prediction and effect plots.} The description of horizontal, vertical, and trend lines remain the same as above.
+A: The wider dashed curves correspond to the conventional prediction curves, while the narrower curves crossing at the center point represent the effect from \pkg{varpred}.
+For simple OLS models, the effect-style curves cross at the center point (the default anchor and also referred to as center-anchored).
+The prediction-style curves incorporate the uncertainties due to intercept term and other non-focal predictors.
+On the other hand, effect-style curves only consider the focal predictor's main effect uncertainty.
 B: center- and zero- anchored effects.
-The zero-anchored effects means that the anchor is at zero value of the focal predictor.
-The choice of the anchor do not affect the central estimates, hence they are identical in both choices of the anchor.}
+The zero-anchored effect means that the anchor is at the zero value of the focal predictor.
+The choice of the anchor does not affect the central estimates; hence they are identical in both choices of the anchor.}
 \label{fig:justify_ci_plots}
@@
 %
-For a prediction-style plot, the confidence intervals are much wider because they include uncertainties associated with the intercept and non-focal predictors, but narrower and crosses at the mean of the focal predictor.
-In other words, with \pkg{varpred}, we are able to generate effects indicating zero uncertainty at the value of the focal predictor we are more certain about, i.e., the anchor.
-For simple OLS models, the point where the curves cross is the model center or simply center point; it corresponds to anchor we choose, in this case, the center point (which is generally an appropriate and stable choice).
+For a prediction-style plot, the confidence intervals are much wider because they include uncertainties associated with the intercept and non-focal predictors but are narrower and cross at the mean of the focal predictor.
+In other words, with \pkg{varpred}, we can generate effects indicating zero uncertainty at the value of the focal predictor we are more certain about, i.e., the anchor.
+For simple OLS models, the point where the curves cross is the model center or center point; it corresponds to the anchor we choose, in this case, the center point (which is generally an appropriate and stable choice).
 
@@
 In particular, we generated two sets of mean-based predictions and effects: 1) one with \code{age} as the focal predictor (a cubic polynomial - MPV); and 2) one with \code{Wealthindex} as the focal predictor (linear).
-In both cases the, we used center point as the anchor for the effects, as shown in \FREF{pred_cubic_plots}.
+We used the center point as the anchor for the effects in both cases, as shown in \FREF{pred_cubic_plots}.
 %
@@
 \end{center}
-\caption{{\bf Prediction and effect plots for cubic polynomial and simple focal predictors.} A: The focal predictor is a MPV cubic polynomial.
+\caption{{\bf Prediction and effect plots for cubic polynomial and simple focal predictors.} A: The focal predictor is an MPV cubic polynomial.
 B: The focal predictor is linear, with MPV non-focal predictor.
@@
 We use the model center as the anchor for effect-style CIs.
-For a simple predictor (panel B), this corresponds simply to the mean of the input variable, thus the effect curves intersect at that that point.
+For a simple predictor (panel B), this corresponds simply to the mean of the input variable; thus, the effect curves intersect at that point.
 For an MPV predictor (panel A), the center point does not correspond to a single value of the predictor, and the curves do not cross.
-The horizontal black and yellow dashed lines are the observed and predicted average household size, i.e., $\over{hh~size}$ and $\over{\widehat{hh~size}}$, respectively.}
+The horizontal black and yellow dashed lines are observed and predicted average household size, i.e., $\over{hh~size}$ and $\over{\widehat{hh~size}}$, respectively.}
 \label{fig:pred_cubic_plots}
@@
 %
-In a model with a simple focal predictor, we expect the CIs for effect-style plots to cross at the anchor point.
-On the other hand, if the focal predictor is a MPV, the center point is not expected to correspond to a single value of the focal predictor.
+We expect the CIs for effect-style plots to cross at the anchor point in a model with a simple focal predictor.
+On the other hand, if the focal predictor is an MPV, the center point is not expected to correspond to a single value of the focal predictor.
 In this case, the effects will be narrower than the predictions plots but will not necessarily intersect.
-For linear models, we expect the observed average, the average of the central estimates and the prediction at the model to all be identical.
+For linear models, we expect the observed average, the average of the central estimates and the prediction at the model center to all be identical.
 
@@
 If the focal predictor has a small effect size, mean-based and observed-value-based approaches produce different estimates -- A.
-In the case of strong effect size, however, the two approaches produce very close estimates -- B.
+However, in the case of strong effect size, the two approaches produce very close estimates -- B.
 In both cases, the observed-value-based estimates are closer to the truth.
 The difference in mean-based and observed-value-based is due to the bias induced by the non-focal predictor and the non-linear link function in the logistic model.
-Mean-based approach suffers from non-linear averaging; and if focal predictor has small effect size, the bias is even pronounced since the effect of non-linear averaging is driven by the non-focal predictor(s) with the strong effect.
-However, this is not the case if the non-focal predictor(s) have small effect since the bias they induce is equally likely to be small.
-Since observed-value-based approach averages over the whole population of the non-focal predictor variables, it is less likely to be affected by the non-linear averaging for the reasons already mentioned.
+The mean-based approach suffers from non-linear averaging.
+If the focal predictor has a small effect size, the bias is even pronounced since the effect of non-linear averaging is driven by the non-focal predictor(s) with a strong effect.
+However, this is not the case if the non-focal predictor(s) have a small effect since the bias they induce is equally likely to be small.
+Since the observed-value-based approach averages over the whole population of the non-focal predictor variables, it is less likely to be affected by the non-linear averaging for the reasons already mentioned.
 The horizontal dashed lines correspond to the respective averages.
@@
 \end{figure}
-
-If there was no effect of non-linear averaging, then we would expect the average observed proportion and the average predictions to intersect at the center point as we see in \FREF{pred_cubic_plots}B.
-One possible reason for the variations we see in \FREF{pred_bin_plots} is the non-linear averaging; since both observed status and predicted probabilities are averaged on the response scale as opposed to link scale.
-For example, if the range of values are bigger than $0.5$ (seemingly the case here), then we would expected the averages to be slightly higher than what we would expect at the center point.
+%
+If there were no effect of non-linear averaging, then we would expect the average observed proportion and the average predictions to intersect at the center point as we see in \FREF{pred_cubic_plots}B.
+One possible reason for the variations we see in \FREF{pred_bin_plots} is the non-linear averaging; since both observed status and predicted probabilities are averaged on the response scale as opposed to the link scale.
+For example, if the range of values is bigger than $0.5$ (seemingly the case here), then we would expect the averages to be slightly higher than we would expect at the center point.
 
 In \FREF{pred_bin_prediction_effects_plots}, we use the results in \FREF{pred_bin_plots}A and compare the prediction and effect plots associated with the central estimates. 
-
+%
 \begin{figure}
@@
 \caption{{\bf Bias correction for prediction and effect plots.} In each case (A and B), the central estimates for predictions and effects are the same.
-These patterns are  consistent with the ones in \FREF{pred_bin_plots}A compared to the truth.
-Further, the predictions and effects exhibit patterns and generalize in a similar way to the ones in simple linear models.
+Compared to the truth, these patterns are consistent with the ones in \FREF{pred_bin_plots}A.
+Further, the predictions and effects exhibit patterns and generalize similarly to the ones in simple linear models.
 }
@@
 %
-The arrows above show the direction of influence; and the implication of $x\rightarrow y$ and $y \rightarrow z$ is that $x$ affects $z$ in two ways.
-First, it has (some) direct effect; and second, it can have indirect effect by influencing $y$, regulated by parameter $\rho$ as shown in \nameref{S3_Appendix}.
+The arrows above show the direction of influence, and the implication of $x\rightarrow y$ and $y \rightarrow z$ is that $x$ affects $z$ in two ways.
+First, it has (some) direct effect; second, it can have an indirect effect by influencing $y$, regulated by parameter $\rho$ as shown in \nameref{S3_Appendix}.
 
@@
 We simulated the data as shown in \nameref{S3_Appendix} and fitted the two models -- non-mediated and mediated.
-In both cases, we compared mean-based and the observed-value-based central estimates, as shown in \FREF{pred_mediated_plots}.
+We compared mean-based and the observed-value-based central estimates in both cases, as shown in \FREF{pred_mediated_plots}.
 %
@@
 \end{center}
-\caption{{\bf Mean-based and observed-value-based estimates for non-mediated (total) effect models and mediated (direct).} A: In the absence of mediator variable, the central estimate aligns well with the observed data.
+\caption{{\bf Mean-based and observed-value-based estimates for non-mediated (total) and mediated (direct) effect models.} A: In the absence of a mediator variable, the central estimate aligns well with the observed data.
 B: In the presence of indirect effect (mediator variable included), the central estimates do not align with the observations.
-Since the models in A and B are both linear models with identity link function, both approaches (mean-based and observed-value) give the same estimates.
+Since the models in A and B are linear models with identity link functions, both approaches (mean-based and observed-value) give the same estimates.
 The horizontal dashed lines are the respective average estimates and observed means ($\bar{\hat{z}} \approx \bar{z}$).
@@
 This is the total effect of $x$, which only tells us the influence of $x$ on $z$.
-By ignoring $y$ in the model, we are still able to capture the effect of $x$ and match the observed values using both approaches.
-However, if we control for the mediator variable $y$, the estimates do not necessarily match the observed values since $x$ has indirect effect on $z$.
-By controlling for the mediator variable, we able to get an indication of the direct (vary week) effect of $x$ on $z$ (see \FREF{pred_mediated_plots}B).
+By ignoring $y$ in the model, we can still capture the effect of $x$ and match the observed values using both approaches.
+However, if we control for the mediator variable $y$, the estimates do not necessarily match the observed values since $x$ indirectly affects $z$.
+By controlling for the mediator variable, we can get an indication of the direct (very weak) effect of $x$ on $z$ (see \FREF{pred_mediated_plots}B).
 
@@
 
-Generalized linear (mixed) models have a wide usage in various fields including public health.
+Generalized linear (mixed) models are widely used in various fields, including public health.
 In a model involving difficult-to-interpret coefficient estimates, an outcome plot can aid in understanding and summarizing the results.
-In particular, if the goal is to capture every uncertainty in the model for a particular focal predictor or if we are interested on total effect, then a prediction plot would be appropriate.
-On the hand, an effect plot is preferable if we want to focus on the uncertainty associated with a focal predictor only or if we are interested in the direct effect.
+In particular, a prediction plot would be appropriate if the goal is to capture every uncertainty in the model for a particular focal predictor or if we are interested in total effect.
+Conversely, an effect plot is preferable if we want to focus on the uncertainty associated with a focal predictor only or if we are interested in the direct effect.
 
-Mean-based approach is widely to generate outcome plots.
+The mean-based approach is widely used to create outcome plots.
 However, in a model with complex interaction, MPVs or categorical variables, it is sensitive to the choice of the reference point.
-We have demonstrated that non-focal model-center-based reference point is generally a stable choice and provide estimates which are more consistent with the observed quantities.
+We have demonstrated that a non-focal model-center-based reference point is generally a stable choice and provides estimates more consistent with the observed quantities as compared to common input-variable mean-based approach.
 
-In a model with non-linear link function such as a logistic or exponential function, the generated central estimate curve may not match well with the observed data, i.e., our description for ``bias''.
-In such such a model, observed-value-based approach provides a way to generate more consistent estimates and is preferable over conventional mean-based approach.
+In a model with a non-linear link function such as a logistic or exponential function, the generated central estimate curve may not match well with the observed data, i.e., our description for bias.
+In such a model, the observed-value-based approach provides a way to generate more consistent estimates and is preferable to the widely used mean-based approach.
 
-The argument and results we present in this paper support a greater need for a shift on focus on how to summarize regression models.
-From our theoretical, methodological and simulation results, researchers using these kind of models should, in the absence of theoretical justification, report predictions based on observed-value-based approach or at least attempt to do a comparison of the two approaches before settling on the most appropriate in answering their research question.
-Moreover, we provide \proglang{R} package, \pkg{vareffects}, which implements these methods and is available on github (\href{https://github.com/mac-theobio/effects}{https://github.com/mac-theobio/effects}).
+The argument and results we present in this paper support a greater need for a shift in focus on how to summarize these kinds of models.
+From our theoretical, methodological and simulation results, researchers using these models should, in the absence of theoretical justification, report predictions based on the observed-value-based approach or at least attempt to compare the two approaches before settling on the most appropriate in answering their research question.
+Moreover, we provide \proglang{R} package, \pkg{vareffects}, which implements these methods and is available on GitHub (\href{https://github.com/mac-theobio/effects}{https://github.com/mac-theobio/effects}).
 
-Our simulations examples focused on simple linear and logistic models due their wide range of usage and application.
+Our simulation examples focused on simple linear and logistic models due to their wide range of usage and application.
 These models also act as a starting point for building other complex models, including mixed effect models and models with categorical predictors.
-The logic for extending to more complex models, including other forms of non-linear link functions is very straightforward.
-In fact the components needed for extension are the correct linear predictor and the inverse link function; everything else generalizes.
+The logic for extending to more complex models, including other forms of non-linear link functions, is very straightforward.
+The components needed for extension are the correct linear predictor and the inverse link function; everything else generalizes.
 In addition, our \proglang{R} package implementation already extends to and supports most of the non-linear link functions and mixed model framework, including multivariate binary outcome models.
-
-\jd{Let's do a better job of summarizing what we did.
-We talk about different purposes of outcome plots, and try to distinguish the goals of predicition and effect.
-We talk about the model center and why it's a good reference point (maybe we should talk more about why we think it's a good anchor point as well … We explain a fairly weird kind of “bias” (an accurate kind of curve that does not match well with the data) and re-discover and implement a cool way of dealing with it.}
 
@@
 \label{S1_Appendix}
-{\bf Cubic polynomial interaction simulation.} Consider an hypothetical simulation which simulates household size as a function of household wealth index and cubic function of the age of the household head, specified as follows:
-
+{\bf Cubic polynomial interaction simulation.} Consider a hypothetical simulation which simulates household size as a function of household wealth index and cubic function of the age of the household head, specified as follows:
+%
 \begin{align}\label{sim:lm_cubic}
@@
 {\bf Binary outcome simulation.} Consider a simple simulation for improved water quality in Nairobi slums, such that the status is $1$ for improved and $0$ for unimproved water quality.
-In additional to the focal predictor, age of the household head, we add wealth index.
+In addition to the focal predictor, age of the household head, we add wealth index.
 In particular:
-
+%
 \begin{align}\label{sim:glm_two_pred}
@@
 \begin{align}\label{sim:simple_mediate}
-z_i &\sim \mathrm{Bern}(\mathrm{P_i}) \nonumber\\
-\mathrm{logit}(\mathrm{P_i}) &= \eta_i \nonumber\\
-\eta_i &= \beta_0 + \beta_{xz} x_i + \beta_{yz} y_i \nonumber\\
+%% z_i &\sim \mathrm{Bern}(\mathrm{P_i}) \nonumber\\
+%% \mathrm{logit}(\mathrm{P_i}) &= \eta_i \nonumber\\
+z_i &= \beta_0 + \beta_{xz} x_i + \beta_{yz} y_i \nonumber\\
 y_i &= \rho x_i + \sqrt{1-\rho^2} y_y \nonumber\\

### 2022-07-27 manuscript/body.tex
Minor problems fixed!

@@
 
-% Use "Eq" instead of "Equation" for equation citations. 
-%% BC: use \EREF{} for equations and \FREF{} for figures
-%% JD: Why do you capitalize these? It's scary.
-% Best to use a command that combines the name and the reference
 \section{Introduction}
@@
 \item uncertainty estimation -- appropriate choice of \emph{anchor} for computing confidence intervals in effects plots
-\item \jd{When is this a challenge?
-Doesn't it just work like magic with the lm machinery?
-and how to incorporate the uncertainty due to non-focal predictors} \BC{\fix{Clarify with JD.
-Another challenge or a continuation  of the above?}}
 \item biases induced by non-linear transformations of the response variable in generalized linear models (especially generalized mixed models).
@@
 
-For a linear model, the averaging is done on the linear scale, i.e., linear averaging.
-As a result, the \emph{bias} in model center estimates can easily be corrected using mean-based approach and the estimates are usually consistent with the observed values.
-\jd{Why is there a bias to be corrected?
-Is this stuff we're going to explain later?} \BC{\fix{We talk about this in Bias correction.}}
+For a linear model, the averaging is done on the linear scale, i.e., linear averaging. As a result, the model-center estimates (made using the mean-based approach) are unbiased. 
 However, in a model with non-linear link function, this is not usually true.

### 2022-07-27 manuscript/body.tex
modified:   body.tex

@@
 Focal and non-focal model center values is also referred to as \emph{center point}.
+\jd{I don't like this as a separate item. What is the need?}
 
@@
 A: The wider dashed curves correspond to the conventional prediction curves, while the narrower curves crossing at the center point represent the effect from \pkg{varpred}.
-For simple OLS models, the effect-style curves cross at the center point (the default anchor and also referred to as center-anchored).
-The prediction-style curves incorporate the uncertainties due to intercept term and other non-focal predictors.
-On the other hand, effect-style curves only consider the focal predictor's main effect uncertainty.
+For simple OLS models, the effect-style curves cross at the center point (center-anchored, the default).
+The prediction-style curves incorporate the uncertainties due to the intercept term and other non-focal predictor, while effect-style curves only consider the focal predictor's main effect uncertainty.
 B: center- and zero- anchored effects.
 The zero-anchored effect means that the anchor is at the zero value of the focal predictor.
-The choice of the anchor does not affect the central estimates; hence they are identical in both choices of the anchor.}
+The choice of the anchor does not affect the central estimates.}
 \label{fig:justify_ci_plots}
@@
 %
-For a prediction-style plot, the confidence intervals are much wider because they include uncertainties associated with the intercept and non-focal predictors but are narrower and cross at the mean of the focal predictor.
-In other words, with \pkg{varpred}, we can generate effects indicating zero uncertainty at the value of the focal predictor we are more certain about, i.e., the anchor.
-For simple OLS models, the point where the curves cross is the model center or center point; it corresponds to the anchor we choose, in this case, the center point (which is generally an appropriate and stable choice).
+For a prediction-style plot, the confidence intervals are wider because they include uncertainties associated with the intercept and non-focal predictors but are narrower and cross at the mean of the focal predictor.
+In other words, with \pkg{varpred}, we can generate effects indicating uncertainty due only to \emph{changes} in the focal predictor (thus in the case of a simple focal predictor, there is no uncertainty at the anchor point, in this case, the model center).
 
@@
 The mean-based approach suffers from non-linear averaging.
-If the focal predictor has a small effect size, the bias is even pronounced since the effect of non-linear averaging is driven by the non-focal predictor(s) with a strong effect.
-However, this is not the case if the non-focal predictor(s) have a small effect since the bias they induce is equally likely to be small.
-Since the observed-value-based approach averages over the whole population of the non-focal predictor variables, it is less likely to be affected by the non-linear averaging for the reasons already mentioned.
+If the focal predictor has a small effect size, the bias is even pronounced since the effect of non-linear averaging is primarily driven by the non-focal predictor(s) with a strong effect.
+Since the observed-value-based approach averages over the whole population of the non-focal predictor variables, it accounts for the effects of non-linear averaging.
 The horizontal dashed lines correspond to the respective averages.
@@
 \end{center}
-\caption{{\bf Bias correction for prediction and effect plots.} In each case (A and B), the central estimates for predictions and effects are the same.
-Compared to the truth, these patterns are consistent with the ones in \FREF{pred_bin_plots}A.
-Further, the predictions and effects exhibit patterns and generalize similarly to the ones in simple linear models.
+\caption{{\bf Bias correction for prediction and effect plots.} 
+As in \FREF{pred_bin_plots}, the observed-value based estimates match the pattern of the data better than mean-based estimates. 
+\jd{Fix the caption (maybe we should say “central, effects, predictions”). I don't think you need to repeat again about central estimates being the same.}
+\jd{Add “data mean” as a horizontal line to the plots? I think data mean might be a good name, actually.
 }
@@
 
+\jd{We should maybe discuss mediated effects and other stuff. We need to be much clearer about chains of causality and this example. Let's not say “somehow”}
 A \emph{mediated} effect is a situation in which only some of the input variables have a \emph{causal} impact on the outcome, even though all the variables are (somehow) associated with the outcome.

### 2022-07-27 manuscript/body.tex
hand over

@@
 \jd{Fix the caption (maybe we should say “central, effects, predictions”). I don't think you need to repeat again about central estimates being the same.}
-\jd{Add “data mean” as a horizontal line to the plots? I think data mean might be a good name, actually.
+\jd{Add “data mean” as a horizontal line to the plots? I think data mean might be a good name, actually.}
 }
@@
 For example, psychological factors such as water taste in the relationship between household socio-economic status and improved water quality status.
+\jd{Maybe a good example for this would be using education as a mediator for effects of wealth? Can you think of a better one?}
 
@@
 We have demonstrated that a non-focal model-center-based reference point is generally a stable choice and provides estimates more consistent with the observed quantities as compared to common input-variable mean-based approach.
+\jd{I don't feel we need to say “non-focal” -- “-based” really covers it.}
 
@@
 These models also act as a starting point for building other complex models, including mixed effect models and models with categorical predictors.
-The logic for extending to more complex models, including other forms of non-linear link functions, is very straightforward.
+The logic for extending to more complex models, including other forms of non-linear link functions, is straightforward.
 The components needed for extension are the correct linear predictor and the inverse link function; everything else generalizes.

### 2022-07-28 manuscript/body.tex
Sending back to JD

@@
 These can be used to illustrate model uncertainty or to give a more explicit quantitative sense of how the outcome is expected to change.
-In generalized models with a non-linear link function, or models with a spline or polynomial response to a predictor variable, they can also aid in understanding difficult-to-interpret coefficient estimates \citep{brambor_understanding_2006, berry_improving_2012, leeper2017interpreting}.
+In generalized models with a non-linear link function, or models with a spline or polynomial response to an input variable, they can also aid in understanding difficult-to-interpret coefficient estimates \citep{brambor_understanding_2006, berry_improving_2012, leeper2017interpreting}.
 
@@
 For example, the regression models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher} both have $3$ input variables -- $x_1, x_2, x_3$.
-\item [Focal predictors] We call the input variable on the x-axis of an outcome plot the focal predictor.  Any other predictor variables are ``non-focal'' predictors.
+\item [Focal predictors] We call the input variable on the x-axis of an outcome plot the focal predictor.  Any other input variables are ``non-focal'' predictors.
 
@@
 \jd{I don't like this as a separate item. What is the need?}
+\BC{\fix{We need the terms to talk about model center \dots}}
 
@@
 But what if we are interested in the uncertainty as a result of the focal predictor only (effect-style CIs), so that the CIs are $\etahfi \pm q\sigma^f_i$, i.e., effects?
-For effect-style CIs, we need an anchor and centered model matrix with all non-focal predictor variables set to zero.
+For effect-style CIs, we need an anchor and centered model matrix with all non-focal linear predictor variables set to zero.
 
@@
 The most commonly used \proglang{R} software packages for outcome plots (\pkg{emmeans} and \pkg{effects}), by default, use the average of input variables and ``sum-to-zero'' as the reference point approach for continuous and categorical input variables, respectively.
-However, there are a number of choices one can make when constructing outcome plots -- for example, in the presence of interactions, the default for \pkg{emmeans} and \pkg{effects} is to average the predictor variables and use these averaged values for the interaction, as opposed to our preferred model-center approach (averaging each linear predictor variable separately).
+However, there are a number of choices one can make when constructing outcome plots -- for example, in the presence of interactions, the default for \pkg{emmeans} and \pkg{effects} is to average the input variables and use these averaged values for the interaction, as opposed to our preferred model-center approach (averaging each linear predictor variable separately).
 The packages give the same results as our method for models without interactions.
@@
 %
-We simulate using $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$, $\beta_{23} = 5$, $\{x_{1,2,3}, \epsilon\} \sim \mathrm{Normal}(0, 1)$, and then compare estimates from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' central estimates (calculated using the simulation parameters) and observed average, i.e., $\bar{y}$, as shown in \FREF{justify_plots}.
+We simulate using $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$, $\beta_{23} = 5$, $\{x_{1,2,3}, \epsilon\} \sim \mathrm{Normal}(0, 1)$, and then compare estimates from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' central estimates (calculated using the simulation parameters) and mean of the data, i.e., $\bar{y}$, as shown in \FREF{justify_plots}.
 %
@@
 \end{center}
-\caption{{\bf Prediction and effect plots for cubic polynomial and simple focal predictors.} A: The focal predictor is an MPV cubic polynomial.
+\caption{{\bf Central estimate, prediction and effect plots for cubic polynomial and simple focal predictors.} A: The focal predictor is an MPV cubic polynomial.
 B: The focal predictor is linear, with MPV non-focal predictor.
@@
 In this case, the effects will be narrower than the predictions plots but will not necessarily intersect.
-For linear models, we expect the observed average, the average of the central estimates and the prediction at the model center to all be identical.
+For linear models, we expect mean of the data, the average of the central estimates and the prediction at the model center to all be identical.
 
@@
 If the focal predictor has a small effect size, the bias is even pronounced since the effect of non-linear averaging is primarily driven by the non-focal predictor(s) with a strong effect.
-Since the observed-value-based approach averages over the whole population of the non-focal predictor variables, it accounts for the effects of non-linear averaging.
+Since the observed-value-based approach averages over the whole population of the non-focal linear predictor variables, it accounts for the effects of non-linear averaging.
 The horizontal dashed lines correspond to the respective averages.
@@
 \end{center}
-\caption{{\bf Bias correction for prediction and effect plots.} 
-As in \FREF{pred_bin_plots}, the observed-value based estimates match the pattern of the data better than mean-based estimates. 
-\jd{Fix the caption (maybe we should say “central, effects, predictions”). I don't think you need to repeat again about central estimates being the same.}
-\jd{Add “data mean” as a horizontal line to the plots? I think data mean might be a good name, actually.}
-}
+\caption{{\bf Bias correction for central, effect and prediction estimates.} 
+As in \FREF{pred_bin_plots}, the observed-value based estimates match the pattern of the data better than mean-based estimates.}
 \label{fig:pred_bin_prediction_effects_plots}
@@
 
+\BC{\fix{Discuss with JD}}
+
 \jd{We should maybe discuss mediated effects and other stuff. We need to be much clearer about chains of causality and this example. Let's not say “somehow”}
-A \emph{mediated} effect is a situation in which only some of the input variables have a \emph{causal} impact on the outcome, even though all the variables are (somehow) associated with the outcome.
+A \emph{mediated} effect is a situation in which only some of the input variables have a \emph{causal} impact on the outcome, even though all the variables are associated with the outcome.
 For example, psychological factors such as water taste in the relationship between household socio-economic status and improved water quality status.
 \jd{Maybe a good example for this would be using education as a mediator for effects of wealth? Can you think of a better one?}
+\BC{\fix{Discuss with JD}}
 
@@
 Since the models in A and B are linear models with identity link functions, both approaches (mean-based and observed-value) give the same estimates.
-The horizontal dashed lines are the respective average estimates and observed means ($\bar{\hat{z}} \approx \bar{z}$).
+The horizontal dashed lines are the respective average estimates and mean of the data ($\bar{\hat{z}} \approx \bar{z}$).
 The vertical dashed black represents the center point, and the point at which it crosses the horizontal lines represents the expected ``perfect'' estimate at the model center.
@@
 However, in a model with complex interaction, MPVs or categorical variables, it is sensitive to the choice of the reference point.
-We have demonstrated that a non-focal model-center-based reference point is generally a stable choice and provides estimates more consistent with the observed quantities as compared to common input-variable mean-based approach.
-\jd{I don't feel we need to say “non-focal” -- “-based” really covers it.}
+We have demonstrated that a model-center-based reference point is generally a stable choice and provides estimates more consistent with the observed quantities as compared to common input-variable mean-based approach.
 

### 2022-07-28 manuscript/cubic_predictors_preds.R

@@ -30,7 +30,7 @@ pred_fit_mean <- getmeans(pred_age_isolated_cubic, what="estimate", modelname="P
 print(pred_fit_mean)
 
 ## Colour limits
-col_limits <- c("Observed mean", "Predicted mean", "Effects", "Predictions")
+col_limits <- c("Data mean", "Predicted mean", "Effects", "Predictions")
 
 ### Combine
 vlist <- list(pred_age_non_cubic, pred_age_isolated_cubic)
@@ -44,7 +44,7 @@ pred_age_cubic_df <- (comparevarpred(vlist=vlist
 pred_age_cubic_plots <- (ggplot(pred_age_cubic_df$preds, aes(x=age, y=fit))
 	+ geom_line(aes(linetype=model, colour=model), size=1, alpha=0.5)
 	+ geom_ribbon(aes(ymin=lwr, ymax=upr, fill=model), alpha=0.5)
- 	+ geom_hline(data=true_prop_df, aes(yintercept=fit, colour=model, linetype="Observed mean"))
+ 	+ geom_hline(data=true_prop_df, aes(yintercept=fit, colour=model, linetype="Data mean"))
  	+ geom_hline(data=pred_fit_mean, aes(yintercept=fit, colour=model, linetype=model))
  	+ scale_color_colorblind(limits=col_limits)
  	+ scale_fill_colorblind(limits=col_limits)
@@ -84,7 +84,7 @@ pred_wealthindex_cubic_df <- (comparevarpred(vlist=vlist
 pred_wealthindex_cubic_plots <- (ggplot(pred_wealthindex_cubic_df$preds, aes(x=wealthindex, y=fit))
 	+ geom_line(aes(linetype=model, colour=model), size=1, alpha=0.5)
 	+ geom_ribbon(aes(ymin=lwr, ymax=upr, fill=model), alpha=0.5)
- 	+ geom_hline(data=true_prop_df, aes(yintercept=fit, colour=model, linetype="Observed mean"))
+ 	+ geom_hline(data=true_prop_df, aes(yintercept=fit, colour=model, linetype="Data mean"))
 	+ geom_vline(aes(xintercept=mean(wealthindex)), lty=2)
  	+ geom_hline(data=pred_fit_mean, aes(yintercept=fit, colour=model, linetype=model))
  	+ scale_color_colorblind(limits=col_limits)

### 2022-07-28 manuscript/glm_two_predictor_preds.R

@@ -28,7 +28,7 @@ pred_age_pop_mean <- getmeans(pred_age_pop, what="estimate")
 
 ### Binned obs
 binned_df <- binfun(glm_mod, focal="age", bins=50, groups=NULL)
-col_limits <- c("Observed mean", "Mean-based", "Observed-value-based")
+col_limits <- c("Data mean", "Mean-based", "Observed-value-based")
 
 ### Combine all predictions
 vlist <- list(pred_age_none, pred_age_pop)

### 2022-07-28 manuscript/glm_two_predictor_preds_effects.R

@@ -32,7 +32,7 @@ binned_df <- binfun(glm_mod, focal="age", bins=50, groups=NULL)
 
 ### Combine all predictions
 vlist <- list(pred_age_prediction, pred_age_effect)
-col_limits <- c("Truth", "Effects", "Predictions")
+col_limits <- c("Data mean", "Effects", "Predictions")
 
 pred_age <- comparevarpred(vlist=vlist
 	, lnames=NULL
@@ -75,6 +75,7 @@ pred_df <- do.call("rbind"
 	, list(pred_age_df, pred_age_ws_df)
 )
 pred_age_plots <- (ggplot(pred_df, aes(x=age, y=fit))
+	+ geom_hline(data=true_prop_df, aes(yintercept=status, colour=model,lty=model))
 	+ geom_line(aes(colour=model, linetype=model), size=1)
 	+ geom_ribbon(aes(ymin=lwr, ymax=upr, fill=model), alpha=0.5)
 	+ geom_point(data=binned_df, aes(x=age, y=status), colour="grey")

### 2022-07-29 manuscript/body.tex
Bias correction eq

@@
 Each input variable may correspond to one or more linear predictor variables.
-Input variables with more than two categories (for instance religion in our previous example), or input variables with non-linear response functions (e.g., spline or polynomial) will correspond to more than one linear predictor variable -- we call such variables “multi-parameter variables”.
+Input variables with more than two categories (for instance religion in our previous example), or input variables with non-linear response functions (e.g., spline or polynomial) will correspond to more than one linear predictor variable -- we call such multi-parameter variables (MPVs).
 
@@
 \end{align}
-where $x^f$ and $\uset{{\bar{x}}}$ are columns of $\bX^{f}_c$ focal predictor and non-focal predictors, respectively.
+where $x^f$ and $\uset{{\bar{x}}}$ are columns of $\bX^{f}_c$ corresponding to focal predictor and non-focal predictors, respectively.
 
-In a model with a multi-parameter variable or input variables with complex interactions, construction of $\etahfi(x^f, \uset{{\bar{x}}})$ is not usually straightforward since we want $\uset{{\bar{x}}}$ to be or represent the ``true'' model center.
+In a model with MPVs or input variables with complex interactions, construction of $\etahfi(x^f, \uset{{\bar{x}}})$ is not usually straightforward since we want $\uset{{\bar{x}}}$ to be or represent the ``true'' model center.
 We therefore illustrate some of these cases.
@@
 \begin{align}
-\etahfi(\mathrm{Age}_i, \nset{{\over{\mathrm{Wealthindex}}}}) &= \betah_0 + \betah_{\mathrm{A_1}}\mathrm{Age}_i + \betah_{\mathrm{A_2}}\mathrm{Age}^2_i + \betah_{\mathrm{A_3}}\mathrm{Age}^3_i \nonumber\\
-	& + \betah_{\mathrm{W}}\over{\mathrm{Wealthindex}}.
+\etahfi(\mathrm{Age}_i, \uset{{\over{\mathrm{Wealthindex}}}}) &= \betah_0 + \betah_{\mathrm{A_1}}\mathrm{Age}_i + \betah_{\mathrm{A_2}}\mathrm{Age}^2_i + \betah_{\mathrm{A_3}}\mathrm{Age}^3_i \nonumber\\
+	& + \betah_{\mathrm{W}}\uset{\over{\mathrm{Wealthindex}}}.
 \end{align}
@@
 \begin{align}\label{eq:mpv_second}
-\etahfi(\mathrm{Wealthindex}_i, \nset{{\{\over{\mathrm{Age\mathop{\vphantom{^2}}}}, \over{\mathrm{Age}^2}, \over{\mathrm{Age}^2}\}}}) &= \betah_0 + \betah_{\mathrm{A_1}}\over{\mathrm{Age\mathop{\vphantom{^2}}}} + \betah_{\mathrm{A_2}}\over{\mathrm{Age}^2} + \betah_{\mathrm{A_3}}\over{\mathrm{Age}^3}\nonumber\\
+\etahfi(\mathrm{Wealthindex}_i, \uset{{\{\over{\mathrm{Age\mathop{\vphantom{^2}}}}, \over{\mathrm{Age}^2}, \over{\mathrm{Age}^2}\}}}) &= \betah_0 + \betah_{\mathrm{A_1}}\over{\mathrm{Age\mathop{\vphantom{^2}}}} + \betah_{\mathrm{A_2}}\over{\mathrm{Age}^2} + \betah_{\mathrm{A_3}}\over{\mathrm{Age}^3}\nonumber\\
 	& + \betah_{\mathrm{W}}\mathrm{Wealthindex}_i.
@@
 \begin{align}\label{eq:compare_anchored_pop_based}
-g^{-1} \left(\eta_j^\star(\bar{x}_f, \nset{\bar{x}})\right) : \frac{1}{n} \sum_{i=1}^n{ g^{-1} \left(\eta_j^\star(\bar{x}_f, \nset{x})\right)}.
+g^{-1} \left(\eta_i^f(\bar{x}^f, \uset{\bar{x}})\right) : \frac{1}{n} \sum_{i=1}^n{ g^{-1} \left(\eta_i^f(\bar{x}^f, \uset{x})\right)}.
 \end{align}
@@
 
-An alternative approach to choosing a reference point is to make predictions over all observations of the non-focal predictors (members of the population) \citep{hanmer2013behind}.
+An alternative approach to choosing a reference point is to compute central estimates over all observations of the non-focal predictors (members of the population) \citep{hanmer2013behind}.
 The non-linear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population to a one-dimensional distribution of $\etahfi(x_f, \nset{x})$, which is a function of the chosen values of the focal predictor and the whole sample of non-focal predictors, as opposed to the definition in \EREF{eta_mean}.
@@
 We start by comparing our proposed approach with the existing implementations and then illustrate the construction of outcome plots.
-We also demonstrate that the mean-based approach works well for interactions and multi-parameter variables and that the observed-value approach can correct the bias induced by non-linear averaging.
+We also demonstrate that the mean-based approach works well for interactions and MPVs, and that the observed-value approach can correct the bias induced by non-linear averaging.
 
@@
 We simulated data motivated by the water, sanitation, and hygiene (WaSH) study in which we were interested in investing the contribution of demographic and socio-economic factors to improved WaSH indicators among slum dwellers in Nairobi, Kenya.
-We used the mean-based approach to generate the predicted probabilities.
+In this particular study, we used the mean-based approach to generate the predicted probabilities.
 However, we noticed that the predictions consistently over- or under- estimated the observed proportions; and did not align well with the observed data points.
@@
 \end{center}
-\caption{{\bf Mean-based and observed-value-based central estimates.} The \emph{truth} is based on the true simulation parameters.
-If the focal predictor has a small effect size, mean-based and observed-value-based approaches produce different estimates -- A.
+\caption{{\bf Mean-based and observed-value-based central estimates.} If the focal predictor has a small effect size, mean-based and observed-value-based approaches produce different estimates -- A.
 However, in the case of strong effect size, the two approaches produce very close estimates -- B.
-In both cases, the observed-value-based estimates are closer to the truth.
-The difference in mean-based and observed-value-based is due to the bias induced by the non-focal predictor and the non-linear link function in the logistic model.
-The mean-based approach suffers from non-linear averaging.
+The difference in mean-based and observed-value-based is due to the bias induced by the non-focal predictor and the non-linear averaging in the logistic model.
+The mean-based approach is affected by the non-linear averaging.
 If the focal predictor has a small effect size, the bias is even pronounced since the effect of non-linear averaging is primarily driven by the non-focal predictor(s) with a strong effect.
@@
 
-In \FREF{pred_bin_prediction_effects_plots}, we use the results in \FREF{pred_bin_plots}A and compare the prediction and effect plots associated with the central estimates. 
+In \FREF{pred_bin_prediction_effects_plots}, we use the results in \FREF{pred_bin_plots}A and compare the central estimate, prediction and effect plots associated with the central estimates. 
 %

### 2022-07-29 manuscript/body.tex
Editing discussion

--- 2022-07-28 manuscript/body.tex
@@
 The argument and results we present in this paper support a greater need for a shift in focus on how to summarize these kinds of models.
-From our theoretical, methodological and simulation results, researchers using these models should, in the absence of theoretical justification, report predictions based on the observed-value-based approach or at least attempt to compare the two approaches before settling on the most appropriate in answering their research question.
+From our theoretical, methodological and simulation results, researchers using these models should, in the absence of theoretical justification, report predictions based on the observed-value approach or at least attempt to compare the two approaches before settling on the most appropriate in answering their research question.
 Moreover, we provide \proglang{R} package, \pkg{vareffects}, which implements these methods and is available on GitHub (\href{https://github.com/mac-theobio/effects}{https://github.com/mac-theobio/effects}).

### 2022-07-30 manuscript/body.tex
Sending back to JD

@@
 
-\item [Focal model center] The average of the linear predictor variables corresponding to the focal predictor.
-The average of the linear predictor variables corresponding to the non-focal predictor(s) is \emph{non-focal model center}.
+\item [Model center] Is the value of the central estimate calculated using focal and non-focal model center values.
+It is a calculation that is equal to the average of central estimates for a linear model (identity link function) without complex interactions.
+The average of the linear predictor variables corresponding to the focal non-focal predictors are the focal and non-focal model center.
+Focal and non-focal model center values is also referred to as \emph{center point}.
 For simple continuous input variables, averaging linear predictor variables is the same as averaging the input variables (see \FREF{justify_plots}).
-Focal and non-focal model center values is also referred to as \emph{center point}.
-\jd{I don't like this as a separate item. What is the need?}
-\BC{\fix{We need the terms to talk about model center \dots}}
-
-\item [Model center] Is the value of the central estimate calculated using focal and non-focal model center values (center point).
-It is a calculation that is equal to the average of central estimates for a linear model (identity link function) without complex interactions.
 
@@
 
-\BC{\fix{Discuss with JD}}
-
-\jd{We should maybe discuss mediated effects and other stuff. We need to be much clearer about chains of causality and this example. Let's not say “somehow”}
-A \emph{mediated} effect is a situation in which only some of the input variables have a \emph{causal} impact on the outcome, even though all the variables are associated with the outcome.
-For example, psychological factors such as water taste in the relationship between household socio-economic status and improved water quality status.
-\jd{Maybe a good example for this would be using education as a mediator for effects of wealth? Can you think of a better one?}
-\BC{\fix{Discuss with JD}}
-
-For simplicity, we consider a hypothetical simulation for three variables: input variables $x$ and $y$; and outcome variable $z$; such that
+To illustrate this, consider an example in which age of the head of the household (\code{x}) has a direct on the availability of improved water service in the household (\code{z}), as well as an indirect effect through its effect on the income of the head of the household (\code{y}), as illustrated below:
 %
@@
 The arrows above show the direction of influence, and the implication of $x\rightarrow y$ and $y \rightarrow z$ is that $x$ affects $z$ in two ways.
-First, it has (some) direct effect; second, it can have an indirect effect by influencing $y$, regulated by parameter $\rho$ as shown in \nameref{S3_Appendix}.
+First, it has direct effect; second, it can have an indirect effect by influencing $y$, regulated by parameter $\rho$ as shown in \nameref{S3_Appendix}.
 

### 2022-08-03 manuscript/jonathan.R
Working on paper factoring

@@ -6,7 +6,8 @@ library(shellpipes)
 loadEnvironments()
 startGraphics()
 
-m <- glm(z ~ x+y, data=sim_df_mediate, family="binomial")
+## m <- glm(z ~ x+y, data=sim_df_mediate, family="binomial")
+m <- lm(z ~ x+y, data=sim_df_mediate)
 
 def <- varpred(m , "x", modelname="def")
 pred <- varpred(m, "x", isolate=FALSE, modelname="pred")

### 2022-08-03 manuscript/paper.tex
restarting xiangshan (glitching)

--- 2022-07-30 manuscript/body.tex
@@
-%% Main tex
-
-\linenumbers
-
 \section{Introduction}
 
-Plots of predicted values of an outcome against predictors (often called effect plots or prediction plots) are often a useful way to summarize the results of a regression model.
+Plots of predicted values of an outcome against predictors (often called effect plots or prediction plots, generalized here as “outcome plots”) are often a useful way to summarize the results of a statistical model.
 These can be used to illustrate model uncertainty or to give a more explicit quantitative sense of how the outcome is expected to change.
@@
 
-To make an outcome plot, we use a \emph{focal} predictor on the x-axis and \emph{central estimate} (predicted values) on the y-axis.
-The resulting plot will depend on choices we make about other (non-focal) predictor(s).
-In ordinary (Gaussian) linear regression, the non-focal choices may have a simple additive effect on the central estimate.
+To make an outcome plot, we use a \emph{focal} predictor on the x-axis and values of the outomcome variable on the y-axis.
+The resulting plot will depend on choices we make about other (non-focal) predictors.
+In an ordinary linear model, the non-focal choices may have a simple additive effect on the central estimate.
 However, when the focal predictor has interactions or non-linear response functions (e.g., spline or polynomial), non-focal choices can also affect the slope of the estimate.
@@
 
-As noted, the outcome plots described above are often called prediction plots or effects plots. We endeavor here to make a conceptual distinction. 
-The primary distinction between the two lies in how we describe the uncertainties around the central estimate. 
+We endeavor here to make a conceptual distinction between two functions of 
+outcome plots: prediction and effects. 
+The primary difference lies in how we describe the uncertainties around the central estimate. 
 If our goal is to \emph{predict} what we learn about the outcome variable by measuring the focal predictor, then we may want to capture a variety of sources of uncertainty, including that due to the intercept, focal and non-focal predictors, and random effects.
 Conversely, if we wish to focus on the \emph{effect} of a focal predictor only, we might want to isolate uncertainty due to coefficients associated with that predictor.
-If we follow this convention, we expect effects plots to have narrower confidence intervals (CIs) than prediction plots.
+If we follow this convention, we expect effects plots to generally have narrower confidence intervals (CIs) than prediction plots, since less uncertainty is taken into account.
 
 The other distinction relates not to the method of calculating CIs, but to the model chosen for the plot. 
-In general, if we want to predict based on a focal parameter, we are likely to want to fit a univariate model that only contains terms related to that predictor.
+In general, if we want to predict based on a focal parameter, we are likely to want to fit a univariate model that only contains terms related to that predictor, thus addressing the question: if we know (only) the focal parameter, what can we predict about the outcome?.
 If we want to know the effects of a predictor, we may want to control for covariates with a multivariate model, in order to estimate “direct” effects; leave covariates out, in order to estimate “total” effects (direct plus indirect); or take an intermediate strategy.
+These address different versions of the question: if we change the focal parameter (possibly allowing it also to affect other covariates), what is the expected effect on the outcome variable.
+
+\jd{Is there a better place to cite this?}
 Shi et al.
@@
 
+Mediated effects provide a good example of some of the choices involved in showing an effect-style plot.
 To illustrate this, consider an example in which age of the head of the household (\code{x}) has a direct on the availability of improved water service in the household (\code{z}), as well as an indirect effect through its effect on the income of the head of the household (\code{y}), as illustrated below:

### 2022-08-10 manuscript/paper.tex
merge with Talks progress?

@@
 Plots of predicted values of an outcome against predictors (often called effect plots or prediction plots, generalized here as “outcome plots”) are often a useful way to summarize the results of a statistical model.
-These can be used to illustrate model uncertainty or to give a more explicit quantitative sense of how the outcome is expected to change.
-In generalized models with a non-linear link function, or models with a spline or polynomial response to an input variable, they can also aid in understanding difficult-to-interpret coefficient estimates \citep{brambor_understanding_2006, berry_improving_2012, leeper2017interpreting}.
+These can be used to the shape of model response and associated uncertanties, and may be easier to interpret than tables of coefficient estimates \citep{brambor_understanding_2006, berry_improving_2012, leeper2017interpreting}.
 
-To make an outcome plot, we use a \emph{focal} predictor on the x-axis and values of the outomcome variable on the y-axis.
-The resulting plot will depend on choices we make about other (non-focal) predictors.
+To make an outcome plot, we use a \emph{focal} predictor on the x-axis and values of the outcome variable on the y-axis.
+The resulting plot will depend on choices about other (non-focal) predictors.
 In an ordinary linear model, the non-focal choices may have a simple additive effect on the central estimate.
@@
 The other distinction relates not to the method of calculating CIs, but to the model chosen for the plot. 
-In general, if we want to predict based on a focal parameter, we are likely to want to fit a univariate model that only contains terms related to that predictor, thus addressing the question: if we know (only) the focal parameter, what can we predict about the outcome?.
+In general, if we want to predict based on a focal parameter, we are likely to want to fit a univariate model that only contains terms related to that predictor, thus addressing the question: if we know (only) the focal parameter, what can we predict about the outcome?
 If we want to know the effects of a predictor, we may want to control for covariates with a multivariate model, in order to estimate “direct” effects; leave covariates out, in order to estimate “total” effects (direct plus indirect); or take an intermediate strategy.
-These address different versions of the question: if we change the focal parameter (possibly allowing it also to affect other covariates), what is the expected effect on the outcome variable.
+These address different versions of the question: if we change the focal parameter (possibly allowing it also to affect other covariates), what is the expected effect on the outcome variable (see, e.g., \citep{shi_evidence_2017}).
 
-\jd{Is there a better place to cite this?}
-Shi et al.
-\citep{shi_evidence_2017}, for example, used multivariate effects plots to visualize estimated direct effects of different predictors in a paper that compared the difference in sexual risk behaviors between circumcised and uncircumcised men.
-
-Producing this sort of outcome plot has some challenges, including:
-\begin{enumerate}
-\item choosing the \emph{reference point} for non-focal predictors in multivariate models
-\item uncertainty estimation -- appropriate choice of \emph{anchor} for computing confidence intervals in effects plots
-\item biases induced by non-linear transformations of the response variable in generalized linear models (especially generalized mixed models).
-\end{enumerate}
-
-Representative values for a focal predictor are generally chosen using quantiles, or equally spaced values, for a continuous predictor; or an exhaustive set of levels for a categorical predictor.
-The central estimates are then calculated by holding the non-focal predictors at a reference point (a value chosen for a non-focal predictor) while varying the focal predictor, with the goal that the estimates represent how the model responds to the changes in the focal predictor \citep{fox2009effect, hanmer2013behind}.
-These values have been called: \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package} In this article, we refer to these quantities as the \emph{central estimates} of the outcome.
-
-In a model with non-focal predictors such as multivariate models, reference points can be chosen as the average of the non-focal \emph{linear predictor variables} -- we call this approach \emph{mean-based} reference point and is currently not implemented in commonly used \proglang{R} software packages.
-We introduce an alternative choice for the reference point.
-
-For a linear model, the averaging is done on the linear scale, i.e., linear averaging. As a result, the model-center estimates (made using the mean-based approach) are unbiased. 
-However, in a model with non-linear link function, this is not usually true.
-When averaging is done on a separate link scale, the mean of the estimates is not the same as the estimate at the mean point.
-This leads to bias: in this case a systematic difference between the values seen on average for a given value of the focal predictor and the value predicted by the mean-based approach. 
-An alternative to the mean-based reference point is the \emph{observed-value-based} approach, discussed later, which involves computing the prediction over the population of non-focal predictors and then averaging across the values of the focal predictor \citep{hanmer2013behind}.
-
-This article will discuss and implement various approaches for computing predictions and effects plots.
-We further explore and demonstrate, using simulated data, approaches for correcting bias in central estimates for generalized models involving non-linear link functions, including models with random effects.
-The proposed method and \proglang{R} software package will complement the existing ones by providing: 1) a straightforward way to generate effects plots (in our sense), and 2) a robust way to correct for non-linear averaging bias in generalized (mixed) models.
+We discuss the idea of a model “center” and show how it simplifies thinking about outcome plot choices, particularly in the case of ordinary linear models.
+We also discuss biases that arise from non-linear link functions in generalized linear models, and how such biases can be addressed.
 
@@
 
-In order to discuss the statistical background and mathematical formulation of the proposed approaches, we need to understand and formally define a number of terms, some of which have been introduced in the previous section:
+\jd{We should make this a box.}
 
@@
 \item [Focal predictors] We call the input variable on the x-axis of an outcome plot the focal predictor.  Any other input variables are ``non-focal'' predictors.
-
-\item [Model matrix] Refers to the design matrix whose rows include all combination of input variables.
-Consider an example for three hypothetical households -- the first household head is a Christian with an income of $\$ 50$, the second household head is Muslim with an income of $\$ 100$ while the third household head is a Jew with an income of $\$ 77$.
-Suppose we want to model the household size (\code{hhsize}) as function of these household characteristics, i.e., $$\mathrm{hhsize} = \beta_0 + \beta_1\times\mathrm{income} + \beta_{2[r]}\times\mathrm{religion} + \mathrm{error}.$$ The model matrix corresponding to this model is given by
-$$\begin{bmatrix}{}
- Intercept & income & religionJew & religionMuslim \\
- 1 & 50 & 0 & 0 \\
-  1 & 100 & 0 & 1 \\
-  1 & 77 & 1 & 0 \\
-\end{bmatrix}.$$ The first column represents the constant term in our model, $\beta_0$.
-For continuous input variables, the representation in the model matrix is the same as the corresponding input variables (for example second column representing income).
-For categorical variables, however, by default, the model matrix  creates additional dummy variables using the reference cell parameterization.
-This means that, if an input variable has $L$ factor levels, then there will be $L-1$ dummy columns representing all but the first level created in the model matrix.
-In our example, the column \code{religion} is missing and instead we have \code{religionJew} and \code{religionMuslim}; the missing category \code{religionChristian} is treated as the reference category.
 
@@
 \item [Reference point] The values (or sets of values) chosen for non-focal predictors, when estimating the predictions and effects.
-Typically the center point, but can also be chosen as a baseline value or as a mean across categories.
+Often based on the model center, can also use chosen baseline values, or an unweighted mean across categories for a categorical variable.
 We will also discuss using a set of quantiles or observations as a reference.
 
-\item [Anchor] The value chosen for the focal predictor when estimating effect-style confidence intervals.
-The anchor choice does not affect the central estimates, nor prediction-style.
-Often chosen as the center point of the linear predictor variables corresponding to the focal predictor.
-
-\item [Prediction-style and effect-style plots] A prediction-style plot is about predicting observations for a given value of the focal predictor.
-For this, we want to use the classic curved confidence intervals.
-An effect-style plot attempts to visualize the effect of a focal predictor and are characterized with narrower confidence intervals.
-Unlike a prediction-style plot, where the confidence intervals capture the uncertainty associated all predictors in the model, the effect-style plot focuses on uncertainty associated with the focal predictor only and depend on the anchor.
+\item [Anchor] The reference value chosen for variables related the focal predictor when estimating effect-style confidence intervals.
+The anchor choice does not affect the central estimates, nor prediction-style confidence intervals.
+Often based on the model center, can also be chosen.
 

### 2022-08-19 manuscript/body.tex
good night (Te)
Autocommit

@@
 However, the estimates start to differ in the presence of interactions, even as simple as the one in \EREF{simple_inter_higher}.
-In particular, estimates from \pkg{emmeans} and \pkg{effects} but differ from \pkg{varpred}'s, which is very close to the simulated average ($\bar{y}$), \FREF{justify_plots}B.
+In particular, estimates from \pkg{emmeans} and \pkg{effects} are identical but differ from \pkg{varpred}'s, which is very close to the simulated average ($\bar{y}$), \FREF{justify_plots}B.
 In the simple model \FREF{justify_plots}A the input variables are the same as the linear predictor variables, so all the two methods produce identical results.

### 2022-08-22 manuscript/body.tex

@@
 In particular, estimates from \pkg{emmeans} and \pkg{effects} are identical but differ from \pkg{varpred}'s, which is very close to the simulated average ($\bar{y}$), \FREF{justify_plots}B.
-In the simple model \FREF{justify_plots}A the input variables are the same as the linear predictor variables, so all the two methods produce identical results.
+In the simple model \FREF{justify_plots}A the input variables are the same as the linear predictor variables, so all the three methods produce identical results.
 In the interaction model \FREF{justify_plots}B, there is an additional linear predictor variable ($x_2x_3$).

### 2022-08-25 manuscript/paper.tex
Some progress? There's a box.

@@
 Plots of predicted values of an outcome against predictors (often called effect plots or prediction plots, generalized here as “outcome plots”) are often a useful way to summarize the results of a statistical model.
-These can be used to the shape of model response and associated uncertanties, and may be easier to interpret than tables of coefficient estimates \citep{brambor_understanding_2006, berry_improving_2012, leeper2017interpreting}.
+These can be used to illustrate the shape of model response and associated uncertainties, and may be easier to interpret than tables of coefficient estimates \citep{brambor_understanding_2006, berry_improving_2012, leeper2017interpreting}.
+
+\begin{gloss}
+\subsection*{Glossary}
+\begin{description}
+\item [Input variables] Observed (scientific) variables underlying an inference or exploration.
+For example, the regression models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher} both have $3$ input variables -- $x_1, x_2, x_3$.
+
+\item [Focal predictors] The input variable on the x-axis of an outcome plot.
+
+\item [Linear predictor variables] The variables which are combined to make the linear predictor (corresponding to the columns in the model matrix).
+Each input variable corresponds to one or more linear predictor variables.
+
+\item[Multi-parameter variables (MPVs)] Input variables which correspond to more than one linear predictor variable -- for example, variables with more than two categories (religion in our first example), or variables with spline or polynomial response.
+
+\item [Model center] The point corresponding to the mean of the linear predictor variables. In an ordinary linear model, the central estimate at this point will be the mean of the response variable.
+
+\item [Reference point] The values (or sets of values) chosen for non-focal predictors when estimating the predictions and effects.
+Can be the model center, chosen baseline values, or a weighted or unweighted mean across categories.
+We will also discuss averaging over a set of quantiles or observations as a reference point.
+
+\item [Anchor] The values chosen for variables related the focal predictor when estimating effect-style confidence intervals.
+The anchor choice does not affect the central estimates, nor prediction-style confidence intervals.
+Often but not always based on the model center.
+
+\end{description}
+\end{gloss}
 
@@
 The resulting plot will depend on choices about other (non-focal) predictors.
-In an ordinary linear model, the non-focal choices may have a simple additive effect on the central estimate.
+In an \emph{ordinary} linear model, the non-focal choices may have a simple additive effect on the central estimate.
 However, when the focal predictor has interactions or non-linear response functions (e.g., spline or polynomial), non-focal choices can also affect the slope of the estimate.
+In a \emph{generalized} linear model, there is usually a non-linear link function, which complicates the picture of additivity.
 Additional challenges arise when dealing with “mixed” models, which incorporate random effects.
@@
 The other distinction relates not to the method of calculating CIs, but to the model chosen for the plot. 
-In general, if we want to predict based on a focal parameter, we are likely to want to fit a univariate model that only contains terms related to that predictor, thus addressing the question: if we know (only) the focal parameter, what can we predict about the outcome?
+In general, if we want to predict based on a focal parameter, we are likely to prefer a univariate model that only contains terms related to that predictor, thus addressing the question: if we know (only) the focal parameter, what can we predict about the outcome?
 If we want to know the effects of a predictor, we may want to control for covariates with a multivariate model, in order to estimate “direct” effects; leave covariates out, in order to estimate “total” effects (direct plus indirect); or take an intermediate strategy.
@@
 
-\section{Definitions}
-
-\jd{We should make this a box.}
-
-\begin{description}
-\item [Input variables] Refers to the observed (or scientific) variables underlying an inference or exploration.
-For example, the regression models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher} both have $3$ input variables -- $x_1, x_2, x_3$.
-\item [Focal predictors] We call the input variable on the x-axis of an outcome plot the focal predictor.  Any other input variables are ``non-focal'' predictors.
-
-\item [Linear predictor variables] Refer to the variables which are combined to make the linear predictor (corresponding to the columns in the model matrix).
-Each input variable may correspond to one or more linear predictor variables.
-Input variables with more than two categories (for instance religion in our previous example), or input variables with non-linear response functions (e.g., spline or polynomial) will correspond to more than one linear predictor variable -- we call such multi-parameter variables (MPVs).
-
-\item [Model center] Is the value of the central estimate calculated using focal and non-focal model center values.
-It is a calculation that is equal to the average of central estimates for a linear model (identity link function) without complex interactions.
-The average of the linear predictor variables corresponding to the focal non-focal predictors are the focal and non-focal model center.
-Focal and non-focal model center values is also referred to as \emph{center point}.
-For simple continuous input variables, averaging linear predictor variables is the same as averaging the input variables (see \FREF{justify_plots}).
-
-\item [Reference point] The values (or sets of values) chosen for non-focal predictors, when estimating the predictions and effects.
-Often based on the model center, can also use chosen baseline values, or an unweighted mean across categories for a categorical variable.
-We will also discuss using a set of quantiles or observations as a reference.
-
-\item [Anchor] The reference value chosen for variables related the focal predictor when estimating effect-style confidence intervals.
-The anchor choice does not affect the central estimates, nor prediction-style confidence intervals.
-Often based on the model center, can also be chosen.
-
-\end{description}
+%% \clearpage
 

### 2022-08-26 manuscript/body.tex
Additional slides

@@
 The central estimates are then calculated by holding the non-focal predictors at a reference point (a value chosen for a non-focal predictor) while varying the focal predictor, with the goal that the estimates represent how the model responds to the changes in the focal predictor \citep{fox2009effect, hanmer2013behind}.
-These values have been called: \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package} In this article, we refer to these quantities as the \emph{central estimates} of the outcome.
+These values have been called: \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}.
+In this article, we refer to these quantities as the \emph{central estimates} of the outcome.
 
@@
 It is a calculation that is equal to the average of central estimates for a linear model (identity link function) without complex interactions.
-The average of the linear predictor variables corresponding to the focal non-focal predictors are the focal and non-focal model center.
-Focal and non-focal model center values is also referred to as \emph{center point}.
+The average of the linear predictor variables corresponding to the focal and non-focal predictors are the focal and non-focal model center, respectively.
+Focal and non-focal model center values are also referred to as \emph{center point}.
 For simple continuous input variables, averaging linear predictor variables is the same as averaging the input variables (see \FREF{justify_plots}).

### 2022-08-31 manuscript/paper.tex
weekly meeting with Bicko

@@
 
-\section{Statistical formulation}
+\section{Ordinary linear models}
 
-To estimate the central estimate of an outcome plot on the response scale, we need a link function $g$ and appropriate values of the focal predictor $\bX^{f}$ and reference point for the non-focal predictors $\uset{{\bX}}$.
-As previously mentioned, we can either use mean-based or observed-value-based reference point.
+\subsection{Outcome plots}
 
-Using mean-based reference point, the central estimate is
-%
-\begin{align}\label{eq:eta_linear_mean}
-\etahfi &= \bbetah \bX^{f}_c \nonumber\\
-\yfi &= g^{-1}(\etahfi),
-\end{align}
-%
-where $\bX^{f}_c = \{\bX^{f}, \uset{{\bX}}_c\}$ is a centered model matrix with appropriately chosen values of the focal predictor $\bX^{f}$ and centered non-focal linear predictor variables $\uset{{\bX}}_c$ constructed by replacing the corresponding values in the model matrix with their averages.
+\subsection{The model center}
 
-On the other hand, using observed-value-based reference point, \EREF{eta_linear_mean}
-becomes
-%
-\begin{align}\label{eq:eta_linear_sample}
-\etahfj &= \bbetah \bX^{f}_j \nonumber\\
-\yfi  &= \underset{j}{\textrm{mean}} ~ g^{-1} (\etahfj),
-\end{align}
-%
-where $\bX^{f}_j = \{X^{f}_j, \uset{{\bX}}\}$ is the model matrix of the $j$th
-observation and $\uset{{\bX}}$ is the entire population of the non-focal linear
-predictor variables.
-Here, we generate a vector of size $J \times N$ and then average over the values of the focal predictor; $J$ and $N$ represents the number of focal values and population of non-focal predictors, respectively.
+\subsection{Multi-parameter predictors}
 
-For identity link function, e.g., in a simple linear model, \EREF{eta_linear_mean} and \EREF{eta_linear_sample} are equivalent.
-In the subsequent sections, we will first discuss general formulation of mean-based approach and how to generate associated confidence intervals.
-Thereafter, we will discuss the second approach, observed-value-based, and its application to bias correction.
+\subsection{Direct effects}
 
-\subsection{Mean-based approach}
+\section{Generalized linear models}
 
-An alternative formulation of \EREF{eta_linear_mean} involves expressing the linear predictor as the sum of the focal and non-focal predictors' linear predictors. In particular, 
-%
-\begin{align}\label{eq:eta_mean}
-\etahfi(x^f, \uset{{\bar{x}}}) &= \hat{\beta}^f x^f + \sum \uset{\hat{\beta}} \uset{{\bar{x}}} \\
-\yfi  &= g^{-1} \left(\eta_i(x^f, \uset{{\bar{x}}})\right)
-\end{align}
-where $x^f$ and $\uset{{\bar{x}}}$ are columns of $\bX^{f}_c$ corresponding to focal predictor and non-focal predictors, respectively.
-
-In a model with MPVs or input variables with complex interactions, construction of $\etahfi(x^f, \uset{{\bar{x}}})$ is not usually straightforward since we want $\uset{{\bar{x}}}$ to be or represent the ``true'' model center.
-We therefore illustrate some of these cases.
-
-\subsubsection{Dealing with multi-parameter variables}
-
-Multi-parameter variables (MPVs) such as splines, polynomials, etc., can be within the focal predictor, or within non-focal predictor(s).
-To distinguish the two, suppose the model which describes the hypothetical simulation of household size based on a number of socio-demographic factors such as age and wealth index is
-%
-\begin{align}\label{eq:lm_cubic}
-\mathrm{hh~size}_i &= \beta_0 + \beta_{\mathrm{A_1}}\mathrm{Age}_i + \beta_{\mathrm{A_2}}\mathrm{Age}^2_i + \beta_{\mathrm{A_3}}\mathrm{Age}^3_i\nonumber \\
-&+ \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \epsilon_i.
-\end{align}
-%
-In the first case, with \code{Age} as the focal predictor, is a cubic polynomial with three linear predictor variables ($\mathrm{Age}_i, \mathrm{Age}^2_i$ and $\mathrm{Age}^3_i$).
-In this case, each linear predictor variable is evaluated separately across the chosen levels of the focal predictor, $Age_i$.
-Specifically, the linear predictor variables are treated as additional columns of the model matrix evaluated with the same values chosen for the focal predictor, while non-focal predictors are fixed at their reference point as discussed in the previous section.
-The central estimates associated with \code{Age} on the linear predictor scale become
-%
-\begin{align}
-\etahfi(\mathrm{Age}_i, \uset{{\over{\mathrm{Wealthindex}}}}) &= \betah_0 + \betah_{\mathrm{A_1}}\mathrm{Age}_i + \betah_{\mathrm{A_2}}\mathrm{Age}^2_i + \betah_{\mathrm{A_3}}\mathrm{Age}^3_i \nonumber\\
-	& + \betah_{\mathrm{W}}\uset{\over{\mathrm{Wealthindex}}}.
-\end{align}
-%
-In the second case, with \code{Wealthindex} as the focal predictor, the non-focal predictor, \code{Age}, is a cubic polynomial.
-In this case, the non-focal linear predictor variables ($\mathrm{Age}_i, \mathrm{Age}^2_i$ and $\mathrm{Age}^3_i$) are all treated as separate non-focal linear predictor variables and an appropriate choice of reference point applies just like in the models without MPVs.
-For instance, in mean-based approach, we average all non-focal MPVs.
-Thus
-%
-\begin{align}\label{eq:mpv_second}
-\etahfi(\mathrm{Wealthindex}_i, \uset{{\{\over{\mathrm{Age\mathop{\vphantom{^2}}}}, \over{\mathrm{Age}^2}, \over{\mathrm{Age}^2}\}}}) &= \betah_0 + \betah_{\mathrm{A_1}}\over{\mathrm{Age\mathop{\vphantom{^2}}}} + \betah_{\mathrm{A_2}}\over{\mathrm{Age}^2} + \betah_{\mathrm{A_3}}\over{\mathrm{Age}^3}\nonumber\\
-	& + \betah_{\mathrm{W}}\mathrm{Wealthindex}_i.
-\end{align}
-%
-
-\subsubsection{Dealing with interactions in input variables}
-
-Interactions can be between non-focal predictors or between focal and non-focal
-predictors. Handling former case is similar to that of the second case
-in MPVs (\EREF{mpv_second}).
-In the latter case, consider model described by \EREF{simple_inter_higher}.
-The interaction is between the focal, $x_2$, and non-focal, $x_3$, predictors.
-In this case, the values of the focal predictors are chosen as previously described and the reference point for the interacting non-focal predictor is predetermined or appropriately chosen set of values.
-In our example, suppose we pick $i$ and $j$ unique values of the focal predictor, $x_2$, and interacting non-focal predictor, $x_3$, respectively.
-The central estimate on linear predictor becomes
-%
-\begin{align*}
-\etahfi(x_{2i}, x_{3j}, {\uset{\bar{x}}_1}) = \betah_0 + \betah_1 \uset{\bar{x}}_1 + \betah_2x_{2i} + \betah_3x_{3j} + \betah_{23}x_{2i}x_{3j}.
-\end{align*}
-%
-The main point is that, in the case of non-interacting non-focal predictors, the reference point is a center point while in the case of interacting non-focal predictors, the choice of the reference point is not necessarily a center point but can be any appropriate value or set of values.
-
-\subsection{Uncertainty estimation}
-
-We describe the uncertainty around the estimates using confidence intervals (CIs). In principle, every prediction has a different CI. The conventional way to compute variances for predictions is 
-%
-\begin{align}\label{eq:conventional_variance}
-\sigma^2_i = \mathrm{Diag}(\bXstar \boldsymbol{\Sigma} {\bXstar}^\top), 
-\end{align}
-%
-so that the confidence intervals are $\etahfi \pm q\sigma_i$, where $\boldsymbol\Sigma$ is the variance matrix of $\bbetah$ and $q$ is an appropriate quantile of Normal or t distribution \citep{fox2009effect}.
-This generates conventional CIs which incorporate all the uncertainties -- including the uncertainties due to the intercept and non-focal predictor.
-We call this prediction-style CIs.
-
-\subsubsection{Effect-style CIs}
-
-But what if we are interested in the uncertainty as a result of the focal predictor only (effect-style CIs), so that the CIs are $\etahfi \pm q\sigma^f_i$, i.e., effects?
-For effect-style CIs, we need an anchor and centered model matrix with all non-focal linear predictor variables set to zero.
-
-Let $\bxo$ be a centered model matrix previously defined, and let $\ba$ be an anchor matrix, with the same dimensions and entries of all non-focal linear predictor variables as $\bxo$.
-Let $\baf$ be the column of $\ba$ corresponding to focal linear predictor variable defined in $\bxo$.
-Any appropriate values can be chosen for $\baf$ but for model center (center-anchored), we use $\bafc$ which is the mean of the focal predictor.
-Thus
-%
-\begin{align}\label{eq:centered_variance}
-\boldsymbol\sigma^2_i = \textrm{Diag}((\bxo - \bafc) \boldsymbol{\Sigma} {(\bxo - \bafc)}^\top).
-\end{align}
-%
-We can see that $\forall ~\bxof=\baf$, $\bxo - \baf = \boldsymbol 0$, hence $\boldsymbol\sigma_i^2 = \boldsymbol{0}$.
-Similarly, for all values of $\bxof$ close to anchor point $\bafc$, the term $(\bxo - \bafc)$ and $\boldsymbol\sigma_i^2$ goes to $\boldsymbol 0$ and $\boldsymbol\sigma_i^2 = \boldsymbol 0$ if $\bxof=\bafc$.
-This means that $\boldsymbol\sigma_i^2$ close to the anchor point are smaller than those away from the anchor point; and results to confidence intervals which are narrower around the anchor or crosses at the anchor point for simple models.
-In other words, this shows the effect of changing the focal value from the anchor value.
-By setting $\baf = \boldsymbol 0$, we get the variances for the prediction-style CIs in \EREF{conventional_variance}.
-
-An alternative way to compute $\boldsymbol\sigma^2_i$ in \EREF{centered_variance} is
-by \emph{zeroing-out} the covariance matrix, which involves setting all the non-focal
-linear predictor variables in $\Sigma$ to $0$. This procedure can be implemented 
-in commonly used \proglang{R} packages for effect plots and prediction plots, but only
-works when the input variables are \emph{centered} prior to model fitting, in case of
-numerical variables, and more complicated when the input variables are categorical. 
-The anchor approach, however, does not require the input variables to be centered
-prior to model fitting since the computation of $\bxo - \bafc$ affects only the 
-intercepts and non-focal predictor linear variables -- the slopes and variance
-corresponding to the focal predictor linear variables are not affected.
-
-%% \subsection{Is center-anchored better?}
-%% 
-%% We want to compare the amount of uncertainty when there is/no anchor.
-Let $\bao = \boldsymbol{0}$ be the anchoring matrix with all entries in $0$s, and let $\bafc$ be the center-anchored matrix defined above.
-Then entries in $(\bxo - \bao) \geq (\bxo - \bafc)$.
-%%
-
-\section{Bias correction}
-
-When dealing with non-linear link functions and additional non-focal predictors,  generated estimates may not reflect the observed response due to the bias, i.e., a term we use to describe a situation in which the central curve does not align well with the observed data points -- the central curve is either below or above the points.
-In such cases, bias correction is needed when back-transforming the estimates to the original scales.
-The common approach for bias-adjustment is second-order Taylor approximation \citep{duursma2003bias, hanmer2013behind}; already implemented in \pkg{emmeans} \citep{lenth2018package}.
-Here, we describe and implement a different approach -- observed-value-based approach for bias correction.
-Hanmer and Kalkan \citep{hanmer2013behind} discussed and implemented observed-value-based and input-variable-based mean-based approaches in binary response models only.
-Our formulation is more general and can easily be extended to other link functions other than logistic.
-
-We can directly compare the outcome estimate at the model center for non-focal parameters (mean-based estimate) with the average of predictions evaluated across the population values of non-focal predictors (observed-value-based estimate, the colon below indicates that we are comparing two quantities):
-%
-\begin{align}\label{eq:compare_anchored_pop_based}
-g^{-1} \left(\eta_i^f(\bar{x}^f, \uset{\bar{x}})\right) : \frac{1}{n} \sum_{i=1}^n{ g^{-1} \left(\eta_i^f(\bar{x}^f, \uset{x})\right)}.
-\end{align}
-%
-In an ordinary linear model, the link function $g$ is the identity function, so the two means are the same. For non-trivial link functions, we expect them to be different in general.
-
-From Jensen's inequality, the exponential link function, for example, is concave up; hence we expect the right-hand side of \EREF{compare_anchored_pop_based} to be greater than the left-hand side, i.e., the average of the predictions is greater than the prediction at the average of the focal and non-focal parameters.
-On the other hand, the logistic function is concave up and concave down at low and high probabilities, respectively.
-We expect a pattern similar to that of exponential function when the logistic function is concave up and the opposite when the logistic function is concave down.
-In other words, the mean-based approach can under-estimate the prediction in exponential and low probability logistic functions and over-estimate in high probability logistic function.
-
-\subsection{Observed-value-based approach for bias correction}
-
-An alternative approach to choosing a reference point is to compute central estimates over all observations of the non-focal predictors (members of the population) \citep{hanmer2013behind}.
-The non-linear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population to a one-dimensional distribution of $\etahfi(x_f, \nset{x})$, which is a function of the chosen values of the focal predictor and the whole sample of non-focal predictors, as opposed to the definition in \EREF{eta_mean}.
-More specifically:
-\begin{itemize}
-\item compute linear predictor associated with whole sample of the non-focal predictors, $\etanfj = \sum \uset{\hat{\beta}} \uset{x}$
-\item compute linear predictor associated with the chosen values focal predictor, $\etahfi = \hat{\beta}^f x^f$
-\item for every value of the focal linear predictor, $\etahfi$, compute
-%
-\begin{align}\label{eq:pop_eta} 
-\etahfj(\etahfi, \etanfj)  &= \etahfi + \etanfj \nonumber \\
-&= \etahfj(x^f, \uset{x}).
-\end{align}
-\end{itemize}
-%
-
-Once \EREF{pop_eta} is computed, we back-transform the estimates to the original scale and average over the levels of the focal predictors, $j$:
-%
-\begin{align}\label{eq:pop_response} 
-\yfi  &= \underset{j}{\textrm{mean}} ~ g^{-1} \left(\etahfj(x^f, \uset{x})\right).
-\end{align}
-%
-
-We make similar adjustments to compute the variances of the predictions at every level of the focal predictor:
-%
-\begin{align}
-\sigma_j^2 = \textrm{Diag}(\bX^{f}_j \boldsymbol{\Sigma} {\bX^{f}_j}^\top)
-\end{align}
-%
-and
-%
-\begin{align}
-\mathrm{CI}_i = \underset{j}{\textrm{mean}} ~ g^{-1} \left(\etahfj(x^f, \uset{x}) \pm q\sigma^f_{j}\right).
-\end{align}
-%
-
-We make further adjustments for models with random effects components to correct the bias induced by the random effects components.
-We treat the random effects components as additional non-focal predictors in the observed-value approach and modify \EREF{pop_eta}.
-In particular
-%
-\begin{align}\label{eq:pop_eta_re} 
-\uset{\tau} &= \bZ b \nonumber \\
-\etahfj(x^f, \uset{x}, \uset{\tau})  &= \etahfj(x^f, \uset{x}) + \uset{\tau}
-\end{align}
-where $\bZ$ and $b$ are the design matrix and a vector of random effects, respectively.
-
-\section{Mean-based vs. observed-value-based}
-
-In the observed-value-based approach, the ensemble of predictions and CIs are back-transformed before averaging, see \EREF{pop_response}, so we do not need to worry about the non-linear averaging.
-In other words, the averaging is no longer on the link scale and is likely to be bounded by the original data scale.
-In simple linear models without interactions, averaging on the link scale is identical to averaging on the response, so both approaches yield similar results.
-However, picking a single value, e.g., the mean of the predictor, on which to draw conclusions about the effect can be problematic, unrealistic, or not contained in or representative of the population.
-In addition, the mean-based approach fails to use every value of non-focal predictors hence not utilizing the full potential of the information contained in the data.
-This may limit the inferences we can make about the entire population.
-In general, the mean-based approach provides the predictions of an average case, whereas the observed-value-based approach summarizes the predictions over the entire population.
-In some applications, the effect of an average case might not be generalizable to the entire population, especially if the average does not represent the population.
-This might not be a problem in the observed-value-based approach since it focuses on specific observations -- the prediction is first obtained for each observation and then averaged across the levels of the focal predictor.
-
-Another potential concern with the mean-based approach arises when direct naive use leads to a rare or meaningless basis for generalization.
-For example, if our sample has 20\% Jews, 30\% Muslims, and 50\% Christians.
-One approach (default in common packages) is assigning equal category weight, i.e., 1/3 Jews, 1/3 Muslims, and 1/3 Christians (the ``sum-to-zero'' approach).
-The second approach (our default) is setting dummy categorical variables in the model matrix to their means, which, by default, set them to their sample means or observed proportions.
-The first or second approach translates to prediction for a household head who is 1/3 or 50\% Christian, respectively \citep{hanmer2013behind}.
-The second approach seems more realistic and will converge to the population mean in many cases.
-
-The observed-value-based approach is not entirely foolproof.
-For instance, similar to the mean-based approach, in the case of continuous focal predictors, choosing the representative values of the focal predictors can be very challenging, especially if the cases are not evenly distributed around the minimum and the maximum values or within some subgroups defined in the population.
-In addition, the observed-value-based approach can be computationally intensive for large datasets.
-
-\section{Simulation examples}
-
-We start by comparing our proposed approach with the existing implementations and then illustrate the construction of outcome plots.
-We also demonstrate that the mean-based approach works well for interactions and MPVs, and that the observed-value approach can correct the bias induced by non-linear averaging.
-
-\subsection{Comparison with other implementations}
-
-The most commonly used \proglang{R} software packages for outcome plots (\pkg{emmeans} and \pkg{effects}), by default, use the average of input variables and ``sum-to-zero'' as the reference point approach for continuous and categorical input variables, respectively.
-However, there are a number of choices one can make when constructing outcome plots -- for example, in the presence of interactions, the default for \pkg{emmeans} and \pkg{effects} is to average the input variables and use these averaged values for the interaction, as opposed to our preferred model-center approach (averaging each linear predictor variable separately).
-The packages give the same results as our method for models without interactions.
-However, when interactions are present, the two approaches can produce substantially different results, with the model-center approach more closely matching the observed values.
-
-To illustrate this, we simulate data from the models below:
-%
-\begin{align}
-y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \epsilon \label{eq:simple_inter_higher_no_interaction}\\
-y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \beta_{23}x_2x_3 + \epsilon \label{eq:simple_inter_higher}.
-\end{align}
-%
-We simulate using $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$, $\beta_{23} = 5$, $\{x_{1,2,3}, \epsilon\} \sim \mathrm{Normal}(0, 1)$, and then compare estimates from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' central estimates (calculated using the simulation parameters) and mean of the data, i.e., $\bar{y}$, as shown in \FREF{justify_plots}.
-%
-\begin{figure}
-\begin{center}
-\includegraphics{justify_preds.inter.pdf}
-\end{center}
-\caption{{\bf A comparison of central estimates for \pkg{emmeans}, \pkg{effects} and \pkg{varpred}.} The horizontal dashed lines are the mean of the central estimates and simulated $y$, $\bar{\hat{y}}$ and $\bar{y}$, respectively.
-The grey points are averages of the simulated points binned according to the value of $x_1$.
-The trend lines represent the $\hat{y}$ -- central estimate.
-A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, is the same in all three approaches and closely matches the actual mean (truth); the central estimate likewise matches the truth (horizontal lines).
-B: With a simple interaction between non-focal predictors, results from \pkg{emmeans} and \pkg{effects}, but not from the proposed \pkg{varpred}, are biased}
-\label{fig:justify_plots}
-\end{figure}
-%
-In the absence of interactions (\EREF{simple_inter_higher_no_interaction}), the three approaches produce identical estimates, which match the simulated values, \FREF{justify_plots}A.
-However, the estimates start to differ in the presence of interactions, even as simple as the one in \EREF{simple_inter_higher}.
-In particular, estimates from \pkg{emmeans} and \pkg{effects} but differ from \pkg{varpred}'s, which is very close to the simulated average ($\bar{y}$), \FREF{justify_plots}B.
-In the simple model \FREF{justify_plots}A the input variables are the same as the linear predictor variables, so all the two methods produce identical results.
-In the interaction model \FREF{justify_plots}B, there is an additional linear predictor variable ($x_2x_3$).
-\pkg{emmeans} and \pkg{effects} first average the input variables to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} first calculates the corresponding vector of linear predictor values and then averages.
-
-Our implementation, \pkg{varpred}, can generate both prediction-style and effect-style plots.
-However, as previously mentioned, it is hard to generate an effect-style plot in \pkg{emmeans} and \pkg{effects}.
-We consider \EREF{simple_inter_higher_no_interaction} and use $x_2$ as the focal predictor, and compare prediction and effect plots using \pkg{varpred}.
-We also compare different anchors for the effect plot.
-See \FREF{justify_ci_plots}.
-%
-\begin{figure}
-\begin{center}
-\includegraphics{justify_anchors_all.ggp.pdf}
-\end{center}
-\caption{{\bf Prediction and effect plots.} The description of horizontal, vertical, and trend lines remain the same as above.
-A: The wider dashed curves correspond to the conventional prediction curves, while the narrower curves crossing at the center point represent the effect from \pkg{varpred}.
-For simple OLS models, the effect-style curves cross at the center point (center-anchored, the default).
-The prediction-style curves incorporate the uncertainties due to the intercept term and other non-focal predictor, while effect-style curves only consider the focal predictor's main effect uncertainty.
-B: center- and zero- anchored effects.
-The zero-anchored effect means that the anchor is at the zero value of the focal predictor.
-The choice of the anchor does not affect the central estimates.}
-\label{fig:justify_ci_plots}
-\end{figure}
-%
-For a prediction-style plot, the confidence intervals are wider because they include uncertainties associated with the intercept and non-focal predictors but are narrower and cross at the mean of the focal predictor.
-In other words, with \pkg{varpred}, we can generate effects indicating uncertainty due only to \emph{changes} in the focal predictor (thus in the case of a simple focal predictor, there is no uncertainty at the anchor point, in this case, the model center).
-
-\subsection{Prediction- and effect-style plots for MPVs}
-
-To illustrate the distinction between predictions and effects in MPVs, we simulated a multivariate model described in \nameref{S1_Appendix}.
-We also used the same model to illustrate how predictions and effects differ if MPVs are in the focal or non-focal predictor.
-In particular, we generated two sets of mean-based predictions and effects: 1) one with \code{age} as the focal predictor (a cubic polynomial - MPV); and 2) one with \code{Wealthindex} as the focal predictor (linear).
-We used the center point as the anchor for the effects in both cases, as shown in \FREF{pred_cubic_plots}.
-%
-\begin{figure}
-\begin{center}
-\includegraphics{cubic_predictors_preds.ggp.pdf}
-\end{center}
-\caption{{\bf Central estimate, prediction and effect plots for cubic polynomial and simple focal predictors.} A: The focal predictor is an MPV cubic polynomial.
-B: The focal predictor is linear, with MPV non-focal predictor.
-The central estimates (dashed central curves or lines) are the same for predictions and effects in both cases.
-We use the model center as the anchor for effect-style CIs.
-For a simple predictor (panel B), this corresponds simply to the mean of the input variable; thus, the effect curves intersect at that point.
-For an MPV predictor (panel A), the center point does not correspond to a single value of the predictor, and the curves do not cross.
-The horizontal black and yellow dashed lines are observed and predicted average household size, i.e., $\over{hh~size}$ and $\over{\widehat{hh~size}}$, respectively.}
-\label{fig:pred_cubic_plots}
-\end{figure}
-%
-We expect the CIs for effect-style plots to cross at the anchor point in a model with a simple focal predictor.
-On the other hand, if the focal predictor is an MPV, the center point is not expected to correspond to a single value of the focal predictor.
-In this case, the effects will be narrower than the predictions plots but will not necessarily intersect.
-For linear models, we expect mean of the data, the average of the central estimates and the prediction at the model center to all be identical.
-
-\subsection{Bias correction}
-
-We simulated data motivated by the water, sanitation, and hygiene (WaSH) study in which we were interested in investing the contribution of demographic and socio-economic factors to improved WaSH indicators among slum dwellers in Nairobi, Kenya.
-In this particular study, we used the mean-based approach to generate the predicted probabilities.
-However, we noticed that the predictions consistently over- or under- estimated the observed proportions; and did not align well with the observed data points.
-To demonstrate this, we consider a binary-outcome simulation with two input variables (described in \nameref{S2_Appendix}), such that \code{Age} has a very small effect size in comparison to \code{Wealthindex}, and compared their effects on the estimated probability of improved water quality as shown in \FREF{pred_bin_plots}.
-%
-\begin{figure}
-\begin{center}
-\includegraphics{glm_two_predictor_preds.ggp.pdf}
-\end{center}
-\caption{{\bf Mean-based and observed-value-based central estimates.} If the focal predictor has a small effect size, mean-based and observed-value-based approaches produce different estimates -- A.
-However, in the case of strong effect size, the two approaches produce very close estimates -- B.
-The difference in mean-based and observed-value-based is due to the bias induced by the non-focal predictor and the non-linear averaging in the logistic model.
-The mean-based approach is affected by the non-linear averaging.
-If the focal predictor has a small effect size, the bias is even pronounced since the effect of non-linear averaging is primarily driven by the non-focal predictor(s) with a strong effect.
-Since the observed-value-based approach averages over the whole population of the non-focal linear predictor variables, it accounts for the effects of non-linear averaging.
-The horizontal dashed lines correspond to the respective averages.
-The vertical dashed lines represent the center point, and the point at which they intersect the horizontal lines represents the expected ``perfect'' estimate at the model center.
-The grey points are binned observations -- observed proportions of improved water quality in each bin.}
-\label{fig:pred_bin_plots}
-\end{figure}
-%
-If there were no effect of non-linear averaging, then we would expect the average observed proportion and the average predictions to intersect at the center point as we see in \FREF{pred_cubic_plots}B.
-One possible reason for the variations we see in \FREF{pred_bin_plots} is the non-linear averaging; since both observed status and predicted probabilities are averaged on the response scale as opposed to the link scale.
-For example, if the range of values is bigger than $0.5$ (seemingly the case here), then we would expect the averages to be slightly higher than we would expect at the center point.
-
-In \FREF{pred_bin_prediction_effects_plots}, we use the results in \FREF{pred_bin_plots}A and compare the central estimate, prediction and effect plots associated with the central estimates. 
-%
-\begin{figure}
-\begin{center}
-\includegraphics{glm_two_predictor_preds_effects.ggp.pdf}
-\end{center}
-\caption{{\bf Bias correction for central, effect and prediction estimates.} 
-As in \FREF{pred_bin_plots}, the observed-value based estimates match the pattern of the data better than mean-based estimates.}
-\label{fig:pred_bin_prediction_effects_plots}
-\end{figure}
-
-\subsection{Mediated effect}
-
-Mediated effects provide a good example of some of the choices involved in showing an effect-style plot.
-To illustrate this, consider an example in which age of the head of the household (\code{x}) has a direct on the availability of improved water service in the household (\code{z}), as well as an indirect effect through its effect on the income of the head of the household (\code{y}), as illustrated below:
-%
-\begin{center}
-\begin{tikzpicture}
-    \node (1) at (0,0) {x};
-    \node (2) [right = of 1] {y};
-    \node (3) [below = of 2] {z};
-    \path (1) edge  (2);
-    \path (1) edge (2);
-    \path (2) edge (3);
-    \path (1) edge (3);
-\end{tikzpicture}
-\end{center}
-%
-The arrows above show the direction of influence, and the implication of $x\rightarrow y$ and $y \rightarrow z$ is that $x$ affects $z$ in two ways.
-First, it has direct effect; second, it can have an indirect effect by influencing $y$, regulated by parameter $\rho$ as shown in \nameref{S3_Appendix}.
-
-To make inferences about this mediated effect, we can ask questions about the \emph{total} and \emph{direct} effects.
-In the first case, we want to fit a univariate model which regresses $z$ on $x$, excluding $y$.
-We refer to this as \emph{non-mediated} model.
-In the second case, we want to fit a multivariate model which regresses $z$ on $x$ and $y$.
-We refer to this as \emph{mediated} model.
-We simulated the data as shown in \nameref{S3_Appendix} and fitted the two models -- non-mediated and mediated.
-We compared mean-based and the observed-value-based central estimates in both cases, as shown in \FREF{pred_mediated_plots}.
-%
-\begin{figure}
-\begin{center}
-\includegraphics{mediate_preds.ggp.pdf}
-\end{center}
-\caption{{\bf Mean-based and observed-value-based estimates for non-mediated (total) and mediated (direct) effect models.} A: In the absence of a mediator variable, the central estimate aligns well with the observed data.
-B: In the presence of indirect effect (mediator variable included), the central estimates do not align with the observations.
-Since the models in A and B are linear models with identity link functions, both approaches (mean-based and observed-value) give the same estimates.
-The horizontal dashed lines are the respective average estimates and mean of the data ($\bar{\hat{z}} \approx \bar{z}$).
-The vertical dashed black represents the center point, and the point at which it crosses the horizontal lines represents the expected ``perfect'' estimate at the model center.
-The grey points are the binned observations.}
-\label{fig:pred_mediated_plots}
-\end{figure}
-%
-From \FREF{pred_mediated_plots}A, we see what we would expect if we simulated a univariate model with no non-focal predictors, even though in the simulation, the effect of $x$ on $z$ is mediated through $y$.
-This is the total effect of $x$, which only tells us the influence of $x$ on $z$.
-By ignoring $y$ in the model, we can still capture the effect of $x$ and match the observed values using both approaches.
-However, if we control for the mediator variable $y$, the estimates do not necessarily match the observed values since $x$ indirectly affects $z$.
-By controlling for the mediator variable, we can get an indication of the direct (very weak) effect of $x$ on $z$ (see \FREF{pred_mediated_plots}B).
-
-\section{Discussion and conclusion}
+\section{Discussion}
 

### 2022-09-29 manuscript/paper.tex
slow progress

@@
 
-\subsection{Outcome plots}
+\begin{figure}
+\begin{center}
+\includegraphics{eco.splash.Rout.pdf}
+\end{center}
+\caption{Outcome plots with effect- (dashed straight lines) and prediction-style (dashed curved lines).
+The panel on the left anchors the effect-style intervals at the model center; the panel on the right has a specific anchor chosen for biological reasons.
+Anchor point does not affect prediction-style confidence intervals}
+\flab{splash}
+\end{figure}
+
+\fref{splash} shows effect- and prediction-style confidence intervals for a hypothetical data set.
+Prediction-style confidence intervals do not depend on an anchor choice, and correspond to \CIs\ for the predicted mean value of Biomass for the given value of Nitrogen (note that these differ the broader \emph{prediction intervals} which give \CIs\ for the observations rather than the predicted means).
+Effect-style confidence intervals show the range of slopes corresponding to the model's \CIs\ for the \emph{direct} effect of the focal predictor.
 
 \subsection{The model center}
+
+To make an outcome plot, we vary the value of the focal predictor, but we need to choose values for other predictors.
+Our default is to do this using the \emph{model center}.
+We define the model center as the average value of all of the “linear predictor variables” which go into the model matrix.
+This is not, in general, the same as averaging over the “input variables” which are measured and provided as data.
+
+The model center does not need to correspond to a physically possible data point.
+For example, the mean of an interaction term across the data is generally not the same as the value of the same term evaluated when each of its components is at its mean.
+The predicted outcome at the model center, however, is always the same as the mean predicted outcome -- in the case of an ordinary linear model, this means that the value is also the same as the observed mean of the outcome variable.
+
+For this reason, using the model center tends to make outcome plots match the data better than other approaches, including that of taking the mean of input variables only.
+
+\begin{figure}
+\begin{center}
+\end{center}
+\{fig:justify_plots}
+\end{figure}
+%
+\fref{center} shows an example, based on simulated data with an interaction between correlated variables.
+In the absence of interactions (\EREF{simple_inter_higher_no_interaction}), the three approaches produce identical estimates, which match the simulated values, \FREF{justify_plots}A.
+However, the estimates start to differ in the presence of interactions, even as simple as the one in \EREF{simple_inter_higher}.
+In particular, estimates from \pkg{emmeans} and \pkg{effects} are identical but differ from \pkg{varpred}'s, which is very close to the simulated average ($\bar{y}$), \FREF{justify_plots}B.
+In the simple model, \FREF{justify_plots}A, the input variables are the same as the linear predictor variables, so all the three methods produce identical results.
+In the interaction model, \FREF{justify_plots}B, there is an additional linear predictor variable ($x_2x_3$).
+\pkg{emmeans} and \pkg{effects} first average the input variables to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} first calculates the corresponding vector of linear predictor values and then averages.
 
@@
 \subsection{Direct effects}
+
+\begin{figure}
+\begin{center}
+\includegraphics{eco.mcomp.Rout.pdf}
+\end{center}
+\caption{Outcome plots with effect- (dashed straight lines) and prediction-style (dashed curved lines).
+The panel on the left shows the results of a univariate model fit; the panel on the right shows a multivariate fit (and effect-style lines only, see text.}
+\flab{mcomp}
+\end{figure}
+%%
+\jd{Alternative figure 1?}
+\fref{mcomp} shows effect- and prediction-style confidence intervals for a hypothetical data set.
+Prediction-style confidence intervals do not depend on an anchor choice, and correspond to \CIs\ for the predicted mean value of Biomass for the given value of Nitrogen (note that these differ the broader \emph{prediction intervals} which give \CIs\ for the observations rather than the predicted means).
+Effect-style confidence intervals show the range of slopes corresponding to the model's \CIs\ for the \emph{direct} effect of the focal predictor.
 

### 2022-09-29 manuscript/eco_data.R

@@ -4,16 +4,16 @@ library(tibble)
 library(shellpipes)
 
 N <- 25 
-seed <- 32
+seed <- 31
 phos_mean <- 22
 phos_spread <- 0.4
 nitro_mean <- 33
 nitro_spread <- 0.4
 mass_mean <- 100
 mass_spread <- 0.4
-beta_np <- -0.2
-beta_pm <- 0.3
-beta_nm <- 0.9
+beta_np <- 0.8
+beta_pm <- 0.8
+beta_nm <- 0.4
 
 set.seed(seed)
 

### 2022-10-05 manuscript/paper.tex
Merge with Bicko?

@@
 \fref{splash} shows effect- and prediction-style confidence intervals for a hypothetical data set.
-Prediction-style confidence intervals do not depend on an anchor choice, and correspond to \CIs\ for the predicted mean value of Biomass for the given value of Nitrogen (note that these differ the broader \emph{prediction intervals} which give \CIs\ for the observations rather than the predicted means).
+Prediction-style confidence intervals do not depend on an anchor choice, and correspond to \CIs\ for the predicted mean value of Biomass for the given value of Nitrogen (note that these differ from the broader \emph{prediction intervals} which give \CIs\ for observations rather than the predicted means, and are not discussed further here).
 Effect-style confidence intervals show the range of slopes corresponding to the model's \CIs\ for the \emph{direct} effect of the focal predictor.
+In this simple case, effect-style CIs, unlike prediction-style CIs, have a clear visual correspondence with statistical clarity: both CI lines have the same slope exactly when the P value for the effect of the predictor is below the threshold.
 
@@
 \end{center}
-\{fig:justify_plots}
+\flab{justify_plots}
 \end{figure}
@@
 %%
-\jd{Alternative figure 1?}
-\fref{mcomp} shows effect- and prediction-style confidence intervals for a hypothetical data set.
-Prediction-style confidence intervals do not depend on an anchor choice, and correspond to \CIs\ for the predicted mean value of Biomass for the given value of Nitrogen (note that these differ the broader \emph{prediction intervals} which give \CIs\ for the observations rather than the predicted means).
-Effect-style confidence intervals show the range of slopes corresponding to the model's \CIs\ for the \emph{direct} effect of the focal predictor.
+\jd{Still working on this. One panel or two?}
+\fref{mcomp} compares direct and indirect effects using univariate vs.~multivariate models
 

### 2022-10-26 manuscript/body.tex

@@
 
-When dealing with non-linear link functions and additional non-focal predictors,  generated estimates may not reflect the observed response due to the bias, i.e., a term we use to describe a situation in which the central curve does not align well with the observed data points -- the central curve is either below or above the points.
+When dealing with non-linear link functions and additional non-focal predictors,  generated estimates may not reflect the observed response due to the bias, i.e., a term we use to describe a situation in which the central curve does not align well with the observed data points -- the central curve is either below or above majority of the observed data points.
 In such cases, bias correction is needed when back-transforming the estimates to the original scales.
@@
 In particular, estimates from \pkg{emmeans} and \pkg{effects} are identical but differ from \pkg{varpred}'s, which is very close to the simulated average ($\bar{y}$), \FREF{justify_plots}B.
-In the simple model \FREF{justify_plots}A the input variables are the same as the linear predictor variables, so all the three methods produce identical results.
-In the interaction model \FREF{justify_plots}B, there is an additional linear predictor variable ($x_2x_3$).
+In the simple model, \FREF{justify_plots}A, the input variables are the same as the linear predictor variables, so all the three methods produce identical results.
+In the interaction model, \FREF{justify_plots}B, there is an additional linear predictor variable ($x_2x_3$).
 \pkg{emmeans} and \pkg{effects} first average the input variables to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} first calculates the corresponding vector of linear predictor values and then averages.
@@
 \end{center}
-\caption{{\bf Central estimate, prediction and effect plots for cubic polynomial and simple focal predictors.} A: The focal predictor is an MPV cubic polynomial.
+\caption{{\bf Central estimate, prediction and effect plots for cubic polynomial and simple focal predictors.} A: The focal predictor is a MPV cubic polynomial.
 B: The focal predictor is linear, with MPV non-focal predictor.
@@
 From our theoretical, methodological and simulation results, researchers using these models should, in the absence of theoretical justification, report predictions based on the observed-value approach or at least attempt to compare the two approaches before settling on the most appropriate in answering their research question.
-Moreover, we provide \proglang{R} package, \pkg{vareffects}, which implements these methods and is available on GitHub (\href{https://github.com/mac-theobio/effects}{https://github.com/mac-theobio/effects}).
+Moreover, we provide \proglang{R} package, \pkg{varpred}, which implements these methods and is available on GitHub (\href{https://cygubicko.github.io/varpred/}{https://cygubicko.github.io/varpred/}).
 

### 2022-11-16 paper/pabs.tex
First good refactor push

--- 2022-07-24 manuscript/abstract.tex
@@
-% Title must be 250 characters or less.
-\begin{flushleft}
-{\Large
-\textbf\newline{Outcome plots: uncertainty estimation and bias correction for predictions and effects in simple and generalized linear (mixed) models} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
-}
-\newline
-% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
-\\
-Steve Cygu\textsuperscript{1, *},
-Benjamin M. Bolker\textsuperscript{1,2},
-Jonathan Dushoff\textsuperscript{1,2}
-\\
-\bigskip
-\textsuperscript{1}School of Computational Science and Engineering, McMaster University, Hamilton, Ontario, Canada
-\\
-\textsuperscript{2}Department of Biology, McMaster University, Hamilton, Ontario, Canada
-\\
-\bigskip
-
-% Use the asterisk to denote corresponding authorship and provide email address in note below.
-* cygu@aims.ac.za
-
-\end{flushleft}
 % Please keep the abstract below 300 words

### 2022-11-16 paper/paper.tex
First good refactor push

--- 2022-10-05 manuscript/paper.tex
@@
 \item [Input variables] Observed (scientific) variables underlying an inference or exploration.
-For example, the regression models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher} both have $3$ input variables -- $x_1, x_2, x_3$.
+For example, the regression models described by \eref{simple_inter_higher_no_interaction} and \eref{simple_inter_higher} both have $3$ input variables -- $x_1, x_2, x_3$.
 
@@
 
-\begin{figure}
-\begin{center}
-\end{center}
-\flab{justify_plots}
-\end{figure}
-%
 \fref{center} shows an example, based on simulated data with an interaction between correlated variables.
-In the absence of interactions (\EREF{simple_inter_higher_no_interaction}), the three approaches produce identical estimates, which match the simulated values, \FREF{justify_plots}A.
-However, the estimates start to differ in the presence of interactions, even as simple as the one in \EREF{simple_inter_higher}.
-In particular, estimates from \pkg{emmeans} and \pkg{effects} are identical but differ from \pkg{varpred}'s, which is very close to the simulated average ($\bar{y}$), \FREF{justify_plots}B.
-In the simple model, \FREF{justify_plots}A, the input variables are the same as the linear predictor variables, so all the three methods produce identical results.
-In the interaction model, \FREF{justify_plots}B, there is an additional linear predictor variable ($x_2x_3$).
+In the absence of interactions (\eref{simple_inter_higher_no_interaction}), the three approaches produce identical estimates, which match the simulated values, \fref{justify_plots}A.
+However, the estimates start to differ in the presence of interactions, even as simple as the one in \eref{simple_inter_higher}.
+In particular, estimates from \pkg{emmeans} and \pkg{effects} are identical but differ from \pkg{varpred}'s, which is very close to the simulated average ($\bar{y}$), \fref{justify_plots}B.
+In the simple model, \fref{justify_plots}A, the input variables are the same as the linear predictor variables, so all the three methods produce identical results.
+In the interaction model, \fref{justify_plots}B, there is an additional linear predictor variable ($x_2x_3$).
 \pkg{emmeans} and \pkg{effects} first average the input variables to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} first calculates the corresponding vector of linear predictor values and then averages.

### 2022-11-16 paper/title.tex
First good refactor push

--- 2022-07-18 manuscript/abstract.tex
@@
 {\Large
-\textbf\newline{Outcome plots: uncertainty estimation and bias correction for predictions and effects in simple and generalized linear (mixed) models} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+\textbf\newline{Outcome plots: Rethinking prediction and effects plots in ordinary and generalized linear models}
 }
-\newline
-% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
-\\
+
+\bigskip
+
 Steve Cygu\textsuperscript{1, *},
@@
 Jonathan Dushoff\textsuperscript{1,2}
-\\
+
 \bigskip
+
 \textsuperscript{1}School of Computational Science and Engineering, McMaster University, Hamilton, Ontario, Canada
@@
 
-% Use the asterisk to denote corresponding authorship and provide email address in note below.
-* cygu@aims.ac.za
-
 \end{flushleft}
-% Please keep the abstract below 300 words
-\section*{Abstract}
-
-In generalized linear (mixed) models that involve complex multiplicative interactions, multi-parameter variables, additional non-focal predictors or a nonlinear link function, outcome plots (prediction and effect plots) can aid in understanding difficult-to-interpret coefficient estimates.
-Outcome plots depend on the choices we make about the non-focal predictors.
-The most common approach is to generate estimates (central estimates, predictions and effects) at a reference point, usually the mean of non-focal predictor.
-We call this mean-based approach and it estimates effect of an average case in the population.
-In the presence of additional non-focal predictors, nonlinear link functions, random effect terms, etc., mean-based approach generates estimates that are biased and may not be consistent with the observed quantities.
-An alternative is the observed-value-based approach which estimates the average effect in the population.
-Moreover, effect-styled confidence intervals provides an alternative and a more clear way to describe uncertainty associated with the focal predictor.
-In addition to theoretical and methodical comparison, using simulation, we illustrate the two approaches and show that they can produce substantially different results and that observed-value-based approach can not only produce estimates consistent with the observed values, but also appropriate for bias correction.
-We also present an alternative way, effect-styled confidence intervals, to describe uncertainties associated with the central estimates.
-
-% Please keep the Author Summary between 150 and 200 words
-% Use first person. PLOS ONE authors please skip this step. 
-% Author Summary not valid for PLOS ONE submissions.   
-% \section*{Author summary}

### 2022-11-30 paper/eco_data.R
Good night

@@ -14,6 +14,7 @@ mass_spread <- 0.4
 beta_np <- 0.8
 beta_pm <- 0.8
 beta_nm <- 0.4
+beta_pnm <- 0.1
 
 set.seed(seed)
 
@@ -22,7 +23,10 @@ dat <- (data.frame(nitro = rlnorm(N, meanlog=log(nitro_mean), sdlog=nitro_spread
 	%>% mutate(NULL
 		, phos_pred = phos_mean+(nitro-nitro_mean)*beta_np
 		, phos = rlnorm(N, meanlog=log(phos_pred), sdlog=phos_spread)
-		, mass_pred=mass_mean+(nitro-nitro_mean)*beta_nm+(phos-phos_mean)*beta_pm
+		, mass_pred=mass_mean
+			+ (nitro-nitro_mean)*beta_nm
+			+ (phos-phos_mean)*beta_pm
+			+ (nitro-nitro_mean)*(phos-phos_mean)*beta_pnm
 		, mass = rlnorm(N, meanlog=log(mass_pred), sdlog=mass_spread)
 	)
 )

### 2023-01-05 paper/paper.tex
Good night
Homing? (xiangshan)

@@
 \end{center}
-\caption{Outcome plots with effect- (dashed straight lines) and prediction-style (dashed curved lines).
+\caption{Outcome plots with effect- (dotted straight lines) and prediction-style (dashed curved lines).
 The panel on the left anchors the effect-style intervals at the model center; the panel on the right has a specific anchor chosen for biological reasons.
@@
 Effect-style confidence intervals show the range of slopes corresponding to the model's \CIs\ for the \emph{direct} effect of the focal predictor.
-In this simple case, effect-style CIs, unlike prediction-style CIs, have a clear visual correspondence with statistical clarity: both CI lines have the same slope exactly when the P value for the effect of the predictor is below the threshold.
+In this simple case, effect-style CIs, unlike prediction-style CIs, have a clear visual correspondence with statistical clarity: both CI lines slope in the same direction (both up, or both down) exactly when the P value for the effect of the predictor is below the threshold.
 
@@
 
-The model center does not need to correspond to a physically possible data point.
+The model center does not always correspond to a physically possible data point.
 For example, the mean of an interaction term across the data is generally not the same as the value of the same term evaluated when each of its components is at its mean.
-The predicted outcome at the model center, however, is always the same as the mean predicted outcome -- in the case of an ordinary linear model, this means that the value is also the same as the observed mean of the outcome variable.
+The predicted outcome at the model center, however, is always the same as the mean predicted outcome (on the link scale where the model operates) -- in the case of an ordinary linear model, this means that the value is also the same as the observed mean of the outcome variable.
 
@@
 
-\fref{center} shows an example, based on simulated data with an interaction between correlated variables.
+\fref{modelcenter} shows an example, based on simulated data with an interaction between correlated variables.
+
 In the absence of interactions (\eref{simple_inter_higher_no_interaction}), the three approaches produce identical estimates, which match the simulated values, \fref{justify_plots}A.
@@
 \end{center}
-\caption{Outcome plots with effect- (dashed straight lines) and prediction-style (dashed curved lines).
+\caption{Outcome plots with effect- (dotted straight lines) and prediction-style (dashed curved lines).
 The panel on the left shows the results of a univariate model fit; the panel on the right shows a multivariate fit (and effect-style lines only, see text.}

### 2023-02-01 paper/paper.tex
Goodnight (板橋)

@@
 In an \emph{ordinary} linear model, the non-focal choices may have a simple additive effect on the central estimate.
-However, when the focal predictor has interactions or non-linear response functions (e.g., spline or polynomial), non-focal choices can also affect the slope of the estimate.
+However, when the focal predictor has interactions or non-linear response functions (e.g., spline or polynomial), non-focal choices can also affect the shape of the estimated curve.
 In a \emph{generalized} linear model, there is usually a non-linear link function, which complicates the picture of additivity.
@@
 For example, the mean of an interaction term across the data is generally not the same as the value of the same term evaluated when each of its components is at its mean.
-The predicted outcome at the model center, however, is always the same as the mean predicted outcome (on the link scale where the model operates) -- in the case of an ordinary linear model, this means that the value is also the same as the observed mean of the outcome variable.
-
+The predicted outcome at the model center, however, is always the same as the mean predicted outcome (in a generalized linear model, this is true on the link scale where the model operates, not the original scale of the outcome variable).
 For this reason, using the model center tends to make outcome plots match the data better than other approaches, including that of taking the mean of input variables only.

### 2023-02-06 paper/paper.tex
renamed:    ncomp.R -> centers.R renamed:    mcomp.R -> uniMulti.R

@@
 \fref{splash} shows effect- and prediction-style confidence intervals for a hypothetical data set.
-Prediction-style confidence intervals do not depend on an anchor choice, and correspond to \CIs\ for the predicted mean value of Biomass for the given value of Nitrogen (note that these differ from the broader \emph{prediction intervals} which give \CIs\ for observations rather than the predicted means, and are not discussed further here).
+Prediction-style confidence intervals do not depend on an anchor choice, and correspond to \CIs\ for the predicted mean value of Biomass for the given value of Nitrogen (note that these differ from the broader \emph{prediction intervals} which give \CIs\ for observations rather than the predicted means, and which are not discussed further here).
 Effect-style confidence intervals show the range of slopes corresponding to the model's \CIs\ for the \emph{direct} effect of the focal predictor.
@@
 We define the model center as the average value of all of the “linear predictor variables” which go into the model matrix.
-This is not, in general, the same as averaging over the “input variables” which are measured and provided as data.
+This is not, in general, the same as averaging over the “input variables” which are measured and provided as data (see Glossary box).
 
@@
 
-\fref{modelcenter} shows an example, based on simulated data with an interaction between correlated variables.
+\begin{figure}
+\begin{center}
+\includegraphics{eco.uniMulti.Rout.pdf}
+\end{center}
+\caption{Outcome plots with effect- (dotted straight lines) and prediction-style (dashed curved lines).
+The panel on the left anchors the effect-style intervals at the model center; the panel on the right has a specific anchor chosen for biological reasons.
+Anchor point does not affect prediction-style confidence intervals}
+\flab{uniMulti}
+\end{figure}
+
+\fref{uniMulti} shows an example, based on simulated data with an interaction between correlated variables.
 
@@
 
+\begin{figure}
+\begin{center}
+\includegraphics{eco.uniMulti.Rout.pdf}
+\end{center}
+\caption{Outcome plots with effect- (dotted straight lines) and prediction-style (dashed curved lines).
+The panel on the left anchors the effect-style intervals at the model center; the panel on the right has a specific anchor chosen for biological reasons.
+Anchor point does not affect prediction-style confidence intervals}
+\flab{uniMulti}
+\end{figure}
+
+\fref{uniMulti} shows an example, based on simulated data with an interaction between correlated variables.
+
 \subsection{Multi-parameter predictors}
@@
 \begin{center}
-\includegraphics{eco.mcomp.Rout.pdf}
+\includegraphics{eco.uniMulti.Rout.pdf}
 \end{center}
@@
 The panel on the left shows the results of a univariate model fit; the panel on the right shows a multivariate fit (and effect-style lines only, see text.}
-\flab{mcomp}
+\flab{uniMulti}
 \end{figure}
@@
 \jd{Still working on this. One panel or two?}
-\fref{mcomp} compares direct and indirect effects using univariate vs.~multivariate models
+\fref{uniMulti} compares direct and indirect effects using univariate vs.~multivariate models
 

### 2023-02-06 paper/eco_data.R

@@ -13,11 +13,14 @@ nitro_mean <- 33
 nitro_spread <- 0.4
 mass_mean <- 100
 mass_spread <- 0.4
+
+# nitrogen, phosphorous, kpotassium
 beta_np <- 0.8
 beta_pm <- 0.8
 beta_nm <- 0.4
 beta_km <- 0.2
 beta_nk <- 0
+beta_pk <- 0.4
 beta_pkm <- 0.1
 
 set.seed(seed)
@@ -27,7 +30,9 @@ dat <- (data.frame(nitro = rlnorm(N, meanlog=log(nitro_mean), sdlog=nitro_spread
 	%>% mutate(NULL
 		, phos_pred = phos_mean+(nitro-nitro_mean)*beta_np
 		, phos = rlnorm(N, meanlog=log(phos_pred), sdlog=phos_spread)
-		, pot_pred = pot_mean+(nitro-nitro_mean)*beta_nk
+		, pot_pred = pot_mean
+			+ (nitro-nitro_mean)*beta_nk
+			+ (phos-phos_mean)*beta_pk
 		, pot = rlnorm(N, meanlog=log(pot_pred), sdlog=pot_spread)
 		, mass_pred=mass_mean
 			+ (nitro-nitro_mean)*beta_nm

### 2023-02-06 paper/uniMulti.R

@@ -7,7 +7,7 @@ startGraphics(height=4)
 
 comb <- ggarrange(
 	getEnvObj("predeff", "uniN") + ggtitle("Univariate")
-	, getEnvObj("predeff", "fullN") + ggtitle("Multivariate")
+	, getEnvObj("predeff", "multi") + ggtitle("Multivariate")
 		+ rremove("ylab")
 	, ncol=2
 )

### 2023-02-06 paper/paper.tex
take break

@@
 \item [Input variables] Observed (scientific) variables underlying an inference or exploration.
-For example, the regression models described by \eref{simple_inter_higher_no_interaction} and \eref{simple_inter_higher} both have $3$ input variables -- $x_1, x_2, x_3$.
+Input variables for our simulated ecological data set are N, P and K, plus the biomass used as a response variable.
 
@@
 \item [Linear predictor variables] The variables which are combined to make the linear predictor (corresponding to the columns in the model matrix).
-Each input variable corresponds to one or more linear predictor variables.
+Each input variable contributes to one or more linear predictor variables.
+N, P and K in our ecological data set are all linear predictor variables, but combinations like P:K or N$\mbox{}^2$ when they appear in the model.
 
@@
 \begin{center}
-\includegraphics{eco.uniMulti.Rout.pdf}
+\includegraphics{eco.multiPK.defAnchor.centers.plot.Rout.pdf}
 \end{center}
-\caption{Outcome plots with effect- (dotted straight lines) and prediction-style (dashed curved lines).
-The panel on the left anchors the effect-style intervals at the model center; the panel on the right has a specific anchor chosen for biological reasons.
-Anchor point does not affect prediction-style confidence intervals}
-\flab{uniMulti}
+\caption{Outcome plots for a model with a non-focal interaction.
+Black lines show an effect plot (mean estimate and CIs) made using the model center (average of linear predictor variables) as a reference point.
+Blue lines show the same approach made by centering based on input variables; due to the interaction this model does not go through the model center.}
+\flab{centers}
 \end{figure}
 
-\fref{uniMulti} shows an example, based on simulated data with an interaction between correlated variables.
-
-In the absence of interactions (\eref{simple_inter_higher_no_interaction}), the three approaches produce identical estimates, which match the simulated values, \fref{justify_plots}A.
-However, the estimates start to differ in the presence of interactions, even as simple as the one in \eref{simple_inter_higher}.
-In particular, estimates from \pkg{emmeans} and \pkg{effects} are identical but differ from \pkg{varpred}'s, which is very close to the simulated average ($\bar{y}$), \fref{justify_plots}B.
+\fref{centers} shows an example, based on simulated data with an interaction between correlated variables.
+The two approaches correctly show the same information about effects of the focal variable, but the one using the model center as a reference point (the \pkg{varpred} default) is centered with respect to the data, while the one that centers based on input variables alone (the most common default, including \pkg{emmeans} and \pkg{effects}:e tod	 which is very close to the simulated average ($\bar{y}$), \fref{justify_plots}B.
 In the simple model, \fref{justify_plots}A, the input variables are the same as the linear predictor variables, so all the three methods produce identical results.

### 2023-02-06 paper/eco_data.R

@@ -15,9 +15,9 @@ mass_mean <- 100
 mass_spread <- 0.4
 
 # nitrogen, phosphorous, kpotassium
-beta_np <- 0.8
+beta_np <- 0.5
 beta_pm <- 0.8
-beta_nm <- 0.4
+beta_nm <- 0.7
 beta_km <- 0.2
 beta_nk <- 0
 beta_pk <- 0.4

### 2023-02-07 paper/paper.tex
do a vdtest

@@
 \fref{centers} shows an example, based on simulated data with an interaction between correlated variables.
-The two approaches correctly show the same information about effects of the focal variable, but the one using the model center as a reference point (the \pkg{varpred} default) is centered with respect to the data, while the one that centers based on input variables alone (the most common default, including \pkg{emmeans} and \pkg{effects}:e tod	 which is very close to the simulated average ($\bar{y}$), \fref{justify_plots}B.
-In the simple model, \fref{justify_plots}A, the input variables are the same as the linear predictor variables, so all the three methods produce identical results.
-In the interaction model, \fref{justify_plots}B, there is an additional linear predictor variable ($x_2x_3$).
-\pkg{emmeans} and \pkg{effects} first average the input variables to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} first calculates the corresponding vector of linear predictor values and then averages.
+The two approaches correctly show the same information about effects of the focal variable, but the one using the model center as a reference point (the \pkg{varpred} default) is centered with respect to the data, while the one that centers based on input variables alone (the most common default, including \pkg{emmeans} and \pkg{effects} is offset vertically.
+This is due to the interaction term not being centered when the input variables are centered; these two approaches give identical results for the simpler model in \fref{splash}.
+
+\subsection{Multi-parameter predictors}
+
+A single input variable can be associated with multiple parameters either by being involved in an interaction, as above, or when a non-linear response (typically spline or polynomial) is specified.
+In this case, the anchor point can be chosen separately for each of the linear predictor variables corresponding to the non-linear response.
+Again, our recommended method is to use the model center.
+The natural alternative is to pick a value of the input variable (either a scientifically meaningful value, or the mean).
+
+\subsection{Direct effects}
 
@@
 \fref{uniMulti} shows an example, based on simulated data with an interaction between correlated variables.
-
-\subsection{Multi-parameter predictors}
-
-\subsection{Direct effects}
-
-\begin{figure}
-\begin{center}
-\includegraphics{eco.uniMulti.Rout.pdf}
-\end{center}
-\caption{Outcome plots with effect- (dotted straight lines) and prediction-style (dashed curved lines).
-The panel on the left shows the results of a univariate model fit; the panel on the right shows a multivariate fit (and effect-style lines only, see text.}
-\flab{uniMulti}
-\end{figure}
-%%
-\jd{Still working on this. One panel or two?}
-\fref{uniMulti} compares direct and indirect effects using univariate vs.~multivariate models
 

### 2023-02-07 paper/meanAnchor.R

@@ -1,0 +1,11 @@
+library(varpred)
+
+library(shellpipes)
+loadEnvironments()
+
+anchor <- mean(nitro)
+
+vpred <- as.data.frame(varpred(mod, "nitro", isolate=FALSE, steps=200, isolate.value=anchor))
+veff <- as.data.frame(varpred(mod, "nitro", steps=200, isolate.value=anchor))
+
+saveEnvironment()

### 2023-02-07 paper/paper.tex
homing (台大)

@@
 Prediction-style confidence intervals do not depend on an anchor choice, and correspond to \CIs\ for the predicted mean value of Biomass for the given value of Nitrogen (note that these differ from the broader \emph{prediction intervals} which give \CIs\ for observations rather than the predicted means, and which are not discussed further here).
-Effect-style confidence intervals show the range of slopes corresponding to the model's \CIs\ for the \emph{direct} effect of the focal predictor.
+Effect-style confidence intervals show the range of slopes corresponding to the model's \CIs\ for the uncertainty due specifically to the effect of the focal predictor.
 In this simple case, effect-style CIs, unlike prediction-style CIs, have a clear visual correspondence with statistical clarity: both CI lines slope in the same direction (both up, or both down) exactly when the P value for the effect of the predictor is below the threshold.
@@
 The natural alternative is to pick a value of the input variable (either a scientifically meaningful value, or the mean).
+
+\begin{figure}
+\begin{center}
+\includegraphics{eco.quadComp.Rout.pdf}
+\end{center}
+\caption{Outcome plots for a model with a non-linear response to the focal variable.
+The left plot is anchored at the model center, which does not correspond to a possible value for the focal variable.
+The right-hand plot is anchored at the mean value of the input variable (nitrogen).
+}
+\flab{quadComp}
+\end{figure}
 

### 2023-02-08 paper/paper.tex
Ready for evening session

@@
 \end{center}
-\caption{Outcome plots with effect- (dotted straight lines) and prediction-style (dashed curved lines).
+\caption{Outcome plots with effect- (dotted straight lines) and prediction-style (dashed curved lines) \CIs.
 The panel on the left anchors the effect-style intervals at the model center; the panel on the right has a specific anchor chosen for biological reasons.
-Anchor point does not affect prediction-style confidence intervals}
+Anchor point does not affect prediction-style confidence intervals.
+Points show the fitted data; the large grey point shows the mean values of predictor and the response.}
 \flab{splash}
@@
 
-\subsection{The model center}
+\subsection{Direct and indirect effects}
 
-To make an outcome plot, we vary the value of the focal predictor, but we need to choose values for other predictors.
+\begin{figure}
+\begin{center}
+\includegraphics{eco.uniMulti.Rout.pdf}
+%%
+\caption{Outcome plots with effect- (dotted straight lines) and prediction-style (dashed curved lines).
+The panel on the left shows the results of a univariate model fit; the panel on the right shows a multivariate fit (and effect-style lines only, see text.}
+%%
+\flab{uniMulti}
+\end{center}
+\end{figure}
+%%
+\jd{Still working on this. One panel or two?}
+
+\fref{uniMulti} shows an example.
+
+\subsection{Reference points}
+
+To make an outcome plot, we vary the value of the focal predictor, but we need to choose values for other predictors -- the reference point for the plot.
 Our default is to do this using the \emph{model center}.
@@
 Black lines show an effect plot (mean estimate and CIs) made using the model center (average of linear predictor variables) as a reference point.
-Blue lines show the same approach made by centering based on input variables; due to the interaction this model does not go through the model center.}
+Blue lines show the same approach made by centering based on input variables; due to the interaction this model does not go through the center point.}
 \flab{centers}
@@
 
-\subsection{Direct effects}
-
-\begin{figure}
-\begin{center}
-\includegraphics{eco.uniMulti.Rout.pdf}
-\end{center}
-\caption{Outcome plots with effect- (dotted straight lines) and prediction-style (dashed curved lines).
-The panel on the left anchors the effect-style intervals at the model center; the panel on the right has a specific anchor chosen for biological reasons.
-Anchor point does not affect prediction-style confidence intervals}
-\flab{uniMulti}
-\end{figure}
-
-\fref{uniMulti} shows an example, based on simulated data with an interaction between correlated variables.
-
+An example is shown in \fref{quadComp}.
+We fit a univariate, \emph{quadratic} response to nitrogen for the biomass.
+Both panels show the same model, and have the same response curves and prediction CIs.
+The panel on the left shows the default effect CIs, anchored at the model center.
+This does not correspond to any real value of nitrogen, because the mean of the quadratic values is taken separately, and is not the value that would correspond in that column to the mean of the original values.
+FOOTNOTE about Jensen, or what?
+The panel on the right calculates effect CIs starting from the point that would correspond to an actual nitrogen value (in this case, the mean across the sampled values).
+The x-axis also reflects actual possible values of nitrogen; this is why the effect uncertainty goes to zero (the anchor point is visible) in the right panel but not the left panel.
+This is also why the response curves do not pass through the model-center point.
 \section{Generalized linear models}

### 2023-02-08 paper/centers.R

@@ -14,7 +14,7 @@ points <- (ggplot(dat)
 	+ geom_point()
 	+ xlab("Nitrogen")
 	+ ylab("Biomass")
-	+ geom_point(aes(mean(nitro), mean(mass)), size=4)
+	+ geom_point(aes(mean(nitro), mean(mass)), size=4, color="grey")
 )
 
 vpoints <- (points

### 2023-02-14 paper/paper.tex
Comments on MS

@@
 Effect-style confidence intervals show the range of slopes corresponding to the model's \CIs\ for the uncertainty due specifically to the effect of the focal predictor.
-In this simple case, effect-style CIs, unlike prediction-style CIs, have a clear visual correspondence with statistical clarity: both CI lines slope in the same direction (both up, or both down) exactly when the P value for the effect of the predictor is below the threshold.
+In this simple case, effect-style CIs, unlike prediction-style CIs, have a clear visual correspondence with statistical clarity: both CI lines slope in the same direction (both up, or both down) exactly when the P value for the effect of the predictor is below the threshold \BC{What would be the threshold in this case?
+The model center?}.
 
@@
 \caption{Outcome plots with effect- (dotted straight lines) and prediction-style (dashed curved lines).
-The panel on the left shows the results of a univariate model fit; the panel on the right shows a multivariate fit (and effect-style lines only, see text.}
+The panel on the left shows the results of a univariate model fit; the panel on the right shows a multivariate fit (and effect-style lines only, see text).
+\BC{We have already compared effect- and prediction-styled plots in \fref{splash}.
+Here, we could do one panel showing direct and indirect effects only.}}
 %%
@@
 %%
-\jd{Still working on this. One panel or two?}
+\jd{Still working on this. One panel or two?} \BC{See above.}
 
@@
 The predicted outcome at the model center, however, is always the same as the mean predicted outcome (in a generalized linear model, this is true on the link scale where the model operates, not the original scale of the outcome variable).
-For this reason, using the model center tends to make outcome plots match the data better than other approaches, including that of taking the mean of input variables only.
+For this reason, using the model center tends to make outcome plots match the data better than other approaches, including the most common approach of taking the mean of input variables only.
 
@@
 Black lines show an effect plot (mean estimate and CIs) made using the model center (average of linear predictor variables) as a reference point.
-Blue lines show the same approach made by centering based on input variables; due to the interaction this model does not go through the center point.}
+Blue lines show the same approach made by centering based on input variables; due to the interaction the central prediction line does not go through the center point.}
 \flab{centers}
@@
 \fref{centers} shows an example, based on simulated data with an interaction between correlated variables.
-The two approaches correctly show the same information about effects of the focal variable, but the one using the model center as a reference point (the \pkg{varpred} default) is centered with respect to the data, while the one that centers based on input variables alone (the most common default, including \pkg{emmeans} and \pkg{effects} is offset vertically.
+The two approaches correctly show the same information about effects of the focal variable, but the one using the model center as a reference point (the \pkg{varpred} default) is centered with respect to the data, while the one that centers based on input variables alone (the most common default, including \pkg{emmeans} and \pkg{effects}) is offset vertically.
 This is due to the interaction term not being centered when the input variables are centered; these two approaches give identical results for the simpler model in \fref{splash}.
@@
 FOOTNOTE about Jensen, or what?
-The panel on the right calculates effect CIs starting from the point that would correspond to an actual nitrogen value (in this case, the mean across the sampled values).
+\BC{Not at this point.
+Jensen has to do with the curvature of the link function.
+Here, we could simply say the model center does not correspond to a single input value but rather multiple predictor variables??} The panel on the right calculates effect CIs starting from the point that would correspond to an actual nitrogen value (in this case, the mean across the sampled values).
 The x-axis also reflects actual possible values of nitrogen; this is why the effect uncertainty goes to zero (the anchor point is visible) in the right panel but not the left panel.

### 2023-02-14 paper/check.R

@@ -5,40 +5,11 @@ library(emmeans)
 library(shellpipes)
 loadEnvironments()
 
-mm <- model.matrix(mod)
-cm <- colMeans(mm)
-bb <- coefficients(mod)
 
-print(matrix(cm, nrow=1) %*% bb)
-mean(dat$mass)
+anchor <- mean(dat$nitro)
 
-veff <- varpred(mod, "nitro", at=list(nitro=dat$nitro), returnall=TRUE)
-
-# varpred
-## center
-vmm <- veff$raw$model.matrix
-head(vmm)
-
-cmm <- colMeans(vmm)
-
-print(matrix(cmm, nrow=1) %*% bb)
-
-## means
-colMeans(vmm)
-
-# data
-## means
-print(cm)
-
-l1 <- sweep(vmm, 2, cmm, FUN="/")
-l1 <- sweep(l1, 2, cm, FUN="*")
-print(l1)
-print(matrix(colMeans(l1), nrow=1) %*% bb)
-
-pred <- as.vector(l1 %*% bb)
-mean(pred)
-
-head(veff$m2)
-all.equal(veff$m2, vmm)
-saveEnvironment()
+vpred <- varpred(mod, "nitro", isolate=FALSE, steps=200, isolate.value=anchor)
+plot(vpred)
+veff <- varpred(mod, "nitro", steps=200, isolate.value=anchor)
+plot(veff)
 

### 2023-02-20 paper/paper.tex
modified:   pabs.tex (getting rid of this file)

@@
+% Please keep the abstract below 300 words
+\section*{Abstract}
+
+\jd{I haven't looked at the Abstract for a long time; will get back to it. Moving it now to the main document file.}
+
+In generalized linear (mixed) models that involve complex multiplicative interactions, multi-parameter variables, additional non-focal predictors or a non-linear link function, outcome plots (prediction and effect plots) can aid in understanding difficult-to-interpret coefficient estimates.
+Outcome plots depend on the choices we make about the non-focal predictors.
+The most common approach is to generate estimates (central estimates, predictions and effects) at a reference point, usually the mean of non-focal predictor.
+We call this mean-based approach and it estimates effect of an average case in the population.
+In the presence of additional non-focal predictors, non-linear link functions, random effect terms, etc., mean-based approach generates estimates that are biased and may not be consistent with the observed quantities.
+An alternative is the observed-value-based approach which estimates the average effect in the population.
+Moreover, effect-styled confidence intervals provides an alternative and a more clear way to describe uncertainty associated with the focal predictor.
+In addition to theoretical and methodical comparison, using simulation, we illustrate the two approaches and show that they can produce substantially different results and that observed-value-based approach can not only produce estimates consistent with the observed values, but also appropriate for bias correction.
+We also present an alternative way, effect-styled confidence intervals, to describe uncertainties associated with the central estimates.
+
+\BC{LET US DISCUSS: What do we want to focus on? common way of generating the plots (input variables) vs model center, prediction- vs effect-style or bias correction, or everything?}
+\jd{Certainly the first two. Do you think bias should be a separate paper?}
+
 \section{Introduction}
@@
 
-We endeavor here to make a conceptual distinction between two functions of 
-outcome plots: prediction and effects. 
+We endeavor here to make a conceptual distinction between two functions of outcome plots: prediction and effects. 
 The primary difference lies in how we describe the uncertainties around the central estimate. 

### 2023-02-20 paper/paper.tex
homing (台大)

@@
 Effect-style confidence intervals show the range of slopes corresponding to the model's \CIs\ for the uncertainty due specifically to the effect of the focal predictor.
-In this simple case, effect-style CIs, unlike prediction-style CIs, have a clear visual correspondence with statistical clarity: both CI lines slope in the same direction (both up, or both down) exactly when the P value for the effect of the predictor is below the threshold \BC{What would be the threshold in this case?
-The model center?}.
+In this simple case, effect-style CIs, unlike prediction-style CIs, have a clear visual correspondence with statistical clarity: both CI lines slope in the same direction (both up, or both down) exactly when the P value for the effect of the predictor is below the threshold for statistical clarity.
 
 \subsection{Direct and indirect effects}
+
+In a model with more than one predictor, we may be interested in an estimate of the \emph{direct} effects of one predictor (after controlling for others).
+If predictors are correlated, the direct effect of a predictor may not reflect the visible relationship between the predictor and the outcome variable.
 
@@
 \caption{Outcome plots with effect- (dotted straight lines) and prediction-style (dashed curved lines).
-The panel on the left shows the results of a univariate model fit; the panel on the right shows a multivariate fit (and effect-style lines only, see text).
-\BC{We have already compared effect- and prediction-styled plots in \fref{splash}.
-Here, we could do one panel showing direct and indirect effects only.}}
+The panel on the left shows the results of a univariate model fit; the panel on the right shows a multivariate fit (and effect-style lines only, see text).}
+\jd{How about now (see text also)?}
 %%
@@
 %%
-\jd{Still working on this. One panel or two?} \BC{See above.}
 
 \fref{uniMulti} shows an example.
+The predicted \emph{overall} relationship between nitrogen and biomass is shown in the left panel.
+This reflects not only direct and indirect effects of nitrogen on biomass, but also any correlations that may be driven by other causes.
+We may be interested in an effect-style plot here, if we think nitrogen is the main driver of observed effects and we want to estimate total (direct plus indirect) effects.
+Or we may be interested in a prediction-style plot of the expected value of biomass for a given \emph{observed} value of nitrogen.
+
+The predicted \emph{direct} relationship between nitrogen and biomass is shown in the right panel.
+This does not track the visible relationship in this case, but instead reflects the modeled \emph{direct} effect of nitrogen.
+The meaning of such a plot will depend on which variables we choose to control for.
+Multivariate plots are not expected to be useful for prediction-style plots, so here we show only effect-style intervals.
 

### 2023-02-22 paper/paper.tex
homing (台大)
ready to meet

@@
 The meaning of such a plot will depend on which variables we choose to control for.
-Multivariate plots are not expected to be useful for prediction-style plots, so here we show only effect-style intervals.
+Prediction-style plots are less likely to be useful for multivariate models, so here we show only effect-style intervals.
 
@@
 \BC{Not at this point.
-Jensen has to do with the curvature of the link function.
-Here, we could simply say the model center does not correspond to a single input value but rather multiple predictor variables??} The panel on the right calculates effect CIs starting from the point that would correspond to an actual nitrogen value (in this case, the mean across the sampled values).
+Jensen has to do with the curvature of the link function.} \jd{Jensen is also how you show that the column means can never correspond to a real value of the parameter: the mean of the squares always exceeds the square of the mean.} \BC{Here, we could simply say the model center does not correspond to a single input value but rather multiple predictor variables??} The panel on the right calculates effect CIs starting from the point that would correspond to an actual nitrogen value (in this case, the mean across the sampled values).
 The x-axis also reflects actual possible values of nitrogen; this is why the effect uncertainty goes to zero (the anchor point is visible) in the right panel but not the left panel.
 This is also why the response curves do not pass through the model-center point.
+
+In neither case is the picture as clear to interpret as the simplest case (\fref{splash}).
+We argue though that the picture on the left is a good way to show the shape of the model fit, while focusing on the uncertainty due to the parameter values due to the focal variable alone.
+
 \section{Generalized linear models}

### 2023-02-22 paper/high.R

@@ -1,0 +1,24 @@
+library(shellpipes)
+
+N <- 25 
+seed <- 31
+pot_mean <- 8
+pot_spread <- 0.4
+phos_mean <- 22
+phos_spread <- 0.4
+nitro_mean <- 33
+nitro_spread <- 0.4
+mass_mean <- 500
+mass_spread <- 0.4
+
+# nitrogen, phosphorous, kpotassium
+beta_np <- 0.5
+beta_pm <- 0.8
+beta_nm <- 0.9
+beta_km <- 0.2
+beta_nk <- 0
+beta_pk <- 0.4
+beta_pkm <- 0.1
+beta_nnm <- -0.625
+
+saveEnvironment()

### 2023-03-08 paper/focal_check.rmd
Adding base R example

---
title: "Focal interaction for multi-parameter input variables"
author: Bicko, Jonathan & Ben
date: "2023 Mar 08 (Wed)"
compact-title: false
output: pdf_document
---

```{r setup, include=FALSE, echo=FALSE}
library(shellpipes)
startGraphics(height=4)

library(dplyr)
library(ggplot2); theme_set(theme_bw(base_size=16))

## Binning function
binfun <- function(mod, focal, bins=50, groups=NULL, ...) {
	if (!is.null(groups)) {
		bins_all <- c(groups, "bin")
	} else {
		bins_all <- "bin"
	}
	if (!inherits(mod, "data.frame")) {
		mf <- model.frame(mod)
	} else {
		mf <- mod
	}
	check_df <- (mf
		%>% arrange_at(focal)
		%>% mutate(bin=ceiling(row_number()*bins/nrow(.)))
		%>% group_by_at(bins_all)
		%>% summarise_all(mean)
		%>% mutate(model="binned")
	)
	return(check_df)
}

df <- rdsRead()
```

The aim is to use a simple example to show what `varpred` does (there is more to that though in the package code) in the case of multi-parameter focal interaction.

Consider a model with the focal multi-parameter:

\[
mass = \beta_0 + \beta_1{nitro} + \beta_2{nitro}^2 + \beta_3{phos} + \epsilon
\]

```{r model}
mod <- lm(mass ~ poly(nitro, 2) + phos, df)
summary(mod)
```

Picking _nitro_ as the focal predictor, the non-focal predictor (_phos_) is set at its model center reference value. For the _focal_ predictor, we simply choose representative values from the input variable. Thus the central prediction is:

\[
\hat{mass} =\hat{\beta_0} + \hat{\beta_1}{nitro} + \hat{\beta_2}{nitro}^2 + \hat{\beta_3}{\bar{phos}}
\]

Let us try to do this the base R way:

```{r baseR}
focal <- df$nitro

beta <- coef(mod)
mod_mat <- model.matrix(mod)

non_focal_mean <- mean(mod_mat[, "phos"])

## Set non-focal model center reference value
mod_mat[, "phos"] <- non_focal_mean
head(mod_mat)

## Generate central predictions
base_pred <- as.vector(mod_mat %*% beta)
```

This averages to the observed average biomass:

```{r compare_baseR}
base <- c(observed=mean(df$mass), base=mean(base_pred))
print(base)
```

But the curve (which averages to the observed) does not pass through this point (black dot)

```{r base_plot}
df$base_pred <- base_pred
base_plot <- (ggplot(df, aes(x=nitro, y=base_pred))
	+ geom_line()
	+ geom_point(aes(y=mass), colour="grey")
	+ geom_hline(yintercept=base[["observed"]], lty=2, colour="blue")
	+ geom_hline(yintercept=base[["base"]], lty=3, colour="red")
	+ geom_point(data=data.frame(mass=base[["observed"]], nitro=mean(focal))
		, aes(x=nitro, y=mass)
		, size=4
		, colour="black"
	)
)
print(base_plot)
```

Let us add binned observations (dark red dots)

```{r binned}
binned_df <-  binfun(df, "nitro", bins=8)
base_plot <- (base_plot
	+ geom_point(data=binned_df, aes(x=nitro, y=mass), size=2, colour="darkred")
)
print(base_plot)
```

**Summary**

- The reference point for the non-focal predictor is chosen at the model center
- All the predictor variables corresponding to the multi-parameter focal variable are evaluated using the chosen focal values
- The average of the predictions is identical to the average of the observed biomass but the curve does not pass through the point (assuming the center of the multi-parameter focal variable is the mean of nitro -- in real sense we don't know the this.)
- These are based on simple base R implementation but agrees with what varpred does.
- **What are we missing in dealing with the multi-parameter focal variables?**

### 2023-03-16 paper/paper.tex
Ready to flash??

@@
 
-\item [Focal predictors] The input variable on the x-axis of an outcome plot.
+\item [Focal predictor] The input variable on the x-axis of an outcome plot.
 
@@
 Each input variable contributes to one or more linear predictor variables.
-N, P and K in our ecological data set are all linear predictor variables, but combinations like P:K or N$\mbox{}^2$ when they appear in the model.
+N, P and K in our ecological data set are all linear predictor variables; so are combinations like P:K or N$\mbox{}^2$ when they appear in the model.
 
-\item[Multi-parameter variables (MPVs)] Input variables which correspond to more than one linear predictor variable -- for example, variables with more than two categories (religion in our first example), or variables with spline or polynomial response.
+\item[Multi-parameter variables (MPVs)] Input variables which correspond to more than one linear predictor variable -- for example, variables with more than two categories, or variables with spline or polynomial response.
 
@@
 The panel on the left shows the default effect CIs, anchored at the model center.
-This does not correspond to any real value of nitrogen, because the mean of the quadratic values is taken separately, and is not the value that would correspond in that column to the mean of the original values.
-FOOTNOTE about Jensen, or what?
-\BC{Not at this point.
-Jensen has to do with the curvature of the link function.} \jd{Jensen is also how you show that the column means can never correspond to a real value of the parameter: the mean of the squares always exceeds the square of the mean.} \BC{Here, we could simply say the model center does not correspond to a single input value but rather multiple predictor variables??} The panel on the right calculates effect CIs starting from the point that would correspond to an actual nitrogen value (in this case, the mean across the sampled values).
-The x-axis also reflects actual possible values of nitrogen; this is why the effect uncertainty goes to zero (the anchor point is visible) in the right panel but not the left panel.
-This is also why the response curves do not pass through the model-center point.
+The model center does not correspond to any real value of nitrogen, because we average separately over the two different linear predictor variables (specifically, the mean of the squares is always greater than the square of the means).
+Thus, no anchor point where the effect-style CIs come together is seen on the plot.
+This is also why no model-center point is seen on the plot.
+The panel on the right is anchored at a point that does correspond to a real nitrogen value (in particular, the mean across the sampled values), and we see the effect-style CIs coming together at this point.
 

### 2023-03-29 paper/paper.tex

@@
 \caption{Outcome plots for a model with a non-linear response to the focal variable.
-The left plot is anchored at the model center, which does not correspond to a possible value for the focal variable.
+The left plot is anchored at the model center, which does not correspond to a possible value for the focal input variable (nitrogen).
 The right-hand plot is anchored at the mean value of the input variable (nitrogen).
@@
 The panel on the left shows the default effect CIs, anchored at the model center.
-The model center does not correspond to any real value of nitrogen, because we average separately over the two different linear predictor variables (specifically, the mean of the squares is always greater than the square of the means).
-Thus, no anchor point where the effect-style CIs come together is seen on the plot.
-This is also why no model-center point is seen on the plot.
-The panel on the right is anchored at a point that does correspond to a real nitrogen value (in particular, the mean across the sampled values), and we see the effect-style CIs coming together at this point.
+Because we average separately over the two different linear predictor variables, 
+the model center does not correspond to a real value of nitrogen and is therefore not seen on the plot. 
+Thus, no anchor point where the effect-style CIs come together is visible. 
+The panel on the right is anchored at a real nitrogen value (in particular, the mean across the sampled values), and we see the effect-style CIs coming together at this point.
 
@@
 We argue though that the picture on the left is a good way to show the shape of the model fit, while focusing on the uncertainty due to the parameter values due to the focal variable alone.
+The physically interpretable anchor on the right, even though chosen to correspond to the mean value of nitrogen, is fair from the model center and generates effect-style CIs that are at some points wider than the prediction-style CIs.
 

### 2023-05-23 paper/paper.tex
new file:   outputs/high.multiglm.obsRef.rnitro.glmPlot.Rout.pdf

@@
 The panel on the left shows the results of a univariate model fit; the panel on the right shows a multivariate fit (and effect-style lines only, see text).}
-\jd{How about now (see text also)?}
+\jd{Need to play with parameters or something; slopes should differ sharply.}
 %%
@@
 The predicted outcome at the model center, however, is always the same as the mean predicted outcome (in a generalized linear model, this is true on the link scale where the model operates, not the original scale of the outcome variable).
+In an ordinary (but not a generalized) model, this mean predicted outcome is also the same as the mean observed outcome.
 For this reason, using the model center tends to make outcome plots match the data better than other approaches, including the most common approach of taking the mean of input variables only.
@@
 We argue though that the picture on the left is a good way to show the shape of the model fit, while focusing on the uncertainty due to the parameter values due to the focal variable alone.
-The physically interpretable anchor on the right, even though chosen to correspond to the mean value of nitrogen, is fair from the model center and generates effect-style CIs that are at some points wider than the prediction-style CIs.
+The physically interpretable anchor on the right, even though chosen to correspond to the mean value of nitrogen, is far from the model center in other dimensions, and generates effect-style CIs that are at some points wider than the prediction-style CIs.
 

### 2023-05-23 paper/high.R

@@ -18,12 +18,13 @@ beta_nm <- 0.9
 beta_km <- 0.2
 beta_nk <- 0
 beta_pk <- 0.4
-beta_pkm <- 0.1
+beta_pkm <- 2
 beta_nnm <- -0.625
+beta_nnm <- 0
 
 # Extra params for binomial outcome
-binSize <- 20
-beta_r <- 3
+binSize <- 100
+beta_r <- 4
 beta_mr <- 0.004
 
 saveEnvironment()

### 2023-06-14 paper/paper.tex
Leave 台大
Friday shutdown
Working on citations

@@
 \section*{Abstract}
-
-\jd{I haven't looked at the Abstract for a long time; will get back to it. Moving it now to the main document file.}
 
@@
 Plots of predicted values of an outcome against predictors (often called effect plots or prediction plots, generalized here as “outcome plots”) are often a useful way to summarize the results of a statistical model.
-These can be used to illustrate the shape of model response and associated uncertainties, and may be easier to interpret than tables of coefficient estimates \citep{brambor_understanding_2006, berry_improving_2012, leeper2017interpreting}.
+These can be used to illustrate the shape of model response and associated uncertainties, and may be easier to interpret than tables of coefficient estimates \citep{searle_population_1980, brambor_understanding_2006, schielzeth2010simple, berry_improving_2012, leeper2017interpreting}.
 
@@
 If our goal is to \emph{predict} what we learn about the outcome variable by measuring the focal predictor, then we may want to capture a variety of sources of uncertainty, including that due to the intercept, focal and non-focal predictors, and random effects.
+This is a common approach and is implemented as a default in many software packages for outcome plots \citep{fox2009effect, leeper2017package, lenth2018package,marginaleffects2023package}.
 Conversely, if we wish to focus on the \emph{effect} of a focal predictor only, we might want to isolate uncertainty due to coefficients associated with that predictor.
@@
 We discuss the idea of a model “center” and show how it simplifies thinking about outcome plot choices, particularly in the case of ordinary linear models.
-We also discuss biases that arise from non-linear link functions in generalized linear models, and how such biases can be addressed.
+We also discuss biases that arise from non-linear link functions in generalized linear models \citep{duursma_bias_2003,duursma2003bias,hanmer2013behind}, and how such biases can be addressed.
+The proposed approach and R software package will complement the existing ones, \citep{fox2009effect, leeper2017package, lenth2018package,marginaleffects2023package}, by providing a straightforward way to generate effects plots (in our sense), and correct for bias in generalized linear models.
 
@@
 \begin{center}
-\includegraphics{eco.splash.Rout.pdf}
+\includegraphics{eco.jd.splash.Rout.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{eco.uniMulti.Rout.pdf}
+\includegraphics{eco.jd.uniMulti.Rout.pdf}
 %%
@@
 \begin{center}
-\includegraphics{eco.multiPK.defAnchor.centers.plot.Rout.pdf}
+\includegraphics{eco.jd.multiPK.defAnchor.centers.plot.Rout.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{eco.quadComp.Rout.pdf}
+\includegraphics{eco.jd.quadComp.Rout.pdf}
 \end{center}
